### 2022-06-16 manuscript/manuscript.Rmd
added manuscript template

---
title: "Article title goes here"
author: 
  - Kilian Eichenseer^[Department of Earth Sciences, Durham University, South Road, DH1 3LE, Durham, United Kingdom] and Lewis A. Jones^[Centro de Investigación Mariña, Grupo de Ecoloxía Animal, Universidade de Vigo, 36310 Vigo, Spain]
output: 
  word_document:
    reference_docx: "template.docx"
bibliography: references.bib
csl: palaeontology.csl
---

Abstract
===============================================================================

Here goes the abstract

# Introduction

Example reference [@R-base]

# Materials & Methods

# Results

# Discussion

# Conclusions

# Acknowledgements

# Author contributions

# Data accessibility

# References

### 2022-07-04 manuscript/manuscript.Rmd
thoughts on title and figures

@@
 ---
-title: "Article title goes here"
+title: "Modelling latitudinal temperature gradients from geological proxies: An early Eocene example"
 author: 
@@
 
+## Fig 1 - Model structure and proof of concept with modern data
+## Fig 2 - Eocene example
+## Fig 3 - Showing influence of changing model assumptions
+
 # Discussion

### 2022-07-04 R/subscripts/ClimateAllStages.R

@@ -92,7 +92,8 @@ for(st in 10) {
                                location = 22.8,
                                scale = 10,
                                shape = 4,
-                               distribution = "skew-normal")
+                               distribution = "skew-normal",
+                               proxy = "coral")
  } else coral_distrmat <- data.frame(NULL)
   
   # run model in parallel chains

### 2022-07-08 manuscript/manuscript.Rmd
prior density and plotting

@@
 ---
-title: "Modelling latitudinal temperature gradients from geological proxies: An early Eocene example"
+title: "Eocene temperature gradients from integrated multiple proxies"
 author: 
@@
 ---
+
+<!-- discarded titles: 
+Modelling latitudinal temperature gradients from geological proxies: An early Eocene example
+-->
 

### 2022-07-08 R/subscripts/AuxiliaryFunctions.R

@@ -158,6 +158,18 @@ plot_chains <- function(mod, params = 1:4, nthin = NULL, logQ = TRUE) {
   par(op)
 }
 
-prior_dens <- function(x,priorindex,priorvec) {
+prior_dens <- function(x,priorvec,priorindex) {
+  x = x
   exp(eval(parse(text = priorvec[priorindex])))
+}
+
+plot_dens <- function(x,dens,xlim = NULL,ylim = NULL,col = rgb(0,0.5,0.7,0.25),
+                      xlab = "", ylab = "density") {
+  
+  if(is.null(ylim)) ylim = c(0,max(dens))
+  if(is.null(xlim)) xlim = range(x)
+  
+  plot(0,0,type = "n", xlim = xlim, ylim = ylim, xlab = xlab, ylab = ylab)
+  polygon(c(x[1],x,x[length(x)]),c(0,dens,0), col = col, border = NA)
+  
 }

### 2022-07-08 R/subscripts/ClimateGradientModelSimple.R

@@ -99,7 +99,9 @@ dsnorm <- function(x,location,scale,alpha, log = TRUE) {
   return(out)
 }
 logposterior_s <- function(x, y, coeff, sdy){
+  if(!is.null(y)) {
   return (loglik_s(x, y, coeff, sdy) + logprior_custom(coeff)) # + 2*(coeff[4])
+  } else return(logprior_custom(coeff))
 }
 
 

### 2022-07-08 R/subscripts/ModernClimate.R

@@ -13,12 +13,13 @@ fastRandomPoints <- function(r, n) {
 }
 fastRandomPoints_lat <- function(r, n, min, max) {
   #if(raster::nlayers(r) > 1) r <- r[[1]]
-  row = raster::rowFromY(r, c(min,max))
-  row = row[1]:row[2]
-  cells = raster::cellFromRow(r, row)
-  r <- raster::rasterFromCells(r, cells, values = FALSE)
+  #row = raster::rowFromY(r, c(min,max))
+  #row = row[1]:row[2]
+  #cells = raster::cellFromRow(r, row)
+  #r <- raster::rasterFromCells(r, cells, values = FALSE)
+  extent <- c(-180,180,min,max)
+  r <- raster::crop(r,extent)
   v <- raster::getValues(r)
-  raster::cel
   v.notNA <- which(!is.na(v))
   x <- sample(v.notNA, n)
   pts <- raster::xyFromCell(r, x)
@@ -29,11 +30,11 @@ map2color<-function(x,pal,limits=NULL){
   pal[findInterval(x,seq(limits[1],limits[2],length.out=length(pal)+1), all.inside=TRUE)]
 }
 
-coords <- fastRandomPoints_lat(sstm$BO21_tempmean_ss,10,-25,25)
+coords <- fastRandomPoints_lat(sstm$BO21_tempmean_ss,1000,-90,90)
 
 sstr <- raster::extract(sstm,coords)
 
-plot(coords[,2],sstr, pch = 19
+plot(abs(coords[,2]),sstr, pch = 19
      , col = rgb(0,0,0,0.1))
      
 ### Test modern data
@@ -50,20 +51,55 @@ c("dsnorm(x,location = -2.7, scale = 16, alpha = 16, log = TRUE)", # prior on A
 "dlnorm(x, -2.2, 0.8, log = TRUE)") # prior on Q
 
 
+priorvec <- 
+  c("dnorm(x,10,15,log=TRUE)", # prior on A
+    "dnorm(x,10,15,log = TRUE)", # prior on DKA
+    "dnorm(x, 45, 25, log = TRUE)", # prior on M
+    "dlnorm(x, -2, 1, log = TRUE)") # prior on Q
+
+
+
+latx <- seq(-40,60,0.1)
+dens <- prior_dens(latx,priorvec,1)
+plot_dens(latx,dens,xlab="A")
+
[…]

### 2022-07-14 manuscript/manuscript.Rmd
Update manuscript.Rmd

@@
 ---
-title: "Eocene temperature gradients from integrated multiple proxies"
+title: "The Early Eocene Climate Optimum reconstructed with Bayesian multi-proxy integration"
 author: 
@@
 # Introduction
+
+The early Eocene climate optimum (EECO, ~ 53 - 49 million years ago) is thought to have been the warmest episode ever since.
+The EECO has received much attention due to its use as an analogue for extreme warming scenarios, depicting a possible future high-CO2 world.
+To reconstruct the climate of EECO, a variety of proxy data and Earth System Models have been used.
+Early modellers have soon noticed that their models struggle to reproduce the climate indicated by proxy data: Without unrealistically high concentrations of atmospheric CO2, early climate models struggled to reproduce the equitable polar climates indicated by early Eocene proxy data, e.g.
+the occurrence of mangroves in polar latitudes, and d18O data from x indicates a temperature of ~ y at latitude z.
 

### 2022-07-19 manuscript/manuscript.Rmd
synthesize Hollis 2019 data

@@
 To reconstruct the climate of EECO, a variety of proxy data and Earth System Models have been used.
-Early modellers have soon noticed that their models struggle to reproduce the climate indicated by proxy data: Without unrealistically high concentrations of atmospheric CO2, early climate models struggled to reproduce the equitable polar climates indicated by early Eocene proxy data, e.g.
+Early modelers have soon noticed that their models struggle to reproduce the climate indicated by proxy data: Without unrealistically high concentrations of atmospheric CO2, early climate models struggled to reproduce the equitable polar climates indicated by early Eocene proxy data, e.g.
 the occurrence of mangroves in polar latitudes, and d18O data from x indicates a temperature of ~ y at latitude z.
@@
 # Materials & Methods
+
+- alternative to current implementation: prior on (A+DKA) instead of on DKA?
+Then constrain DKA to be positive by proposing it in the log-space?
+This would allow for putting a direct prior on the upper temperature limit
 
@@
 
+- x1 Draw samples from the posterior gradient to get a latitudinal gradient of deg C / deg lat with uncertainty (intercept gradient at 0 and 90 deg lat)
+
+- x2 generate global average temperature, and tropical, temperate, polar temperatures in a similar way, accounting for area
+
 # Discussion
+
+- x1 compare lat gradient to literature estimates, e.g. Zhang2019
 

### 2022-07-19 R/subscripts/process_data/process_hollis2019.R

@@ -79,8 +79,8 @@ for(i in 1:length(datr)) {
   datr[[i]] <- datr[[i]][1:(which(sapply(1:nrow(datr[[i]]), function(x) all(is.na(datr[[i]][x,]))))[1]-1),]
 }
 
-dat <- data.frame(matrix(NA,nrow = sum(sapply(datr,nrow)),ncol = 7))
-colnames(dat) <- c("Timeslice", "Age", "MgCa", "SST_median", "SST_025", "SST_975", "Species")
+dat <- data.frame(matrix(NA,nrow = sum(sapply(datr,nrow)),ncol = 8))
+colnames(dat) <- c("Timeslice", "Age", "MgCa", "SST_median", "SST_025", "SST_975", "Species", "Preservation")
 ind <- c(0,cumsum(sapply(datr,nrow)))
 for(i in 1:length(datr)) {
   if("Age" %in% colnames(datr[[i]])) {
@@ -95,6 +95,7 @@ for(i in 1:length(datr)) {
   dat$SST_025[(ind[i]+1):(ind[i+1])] <- datr[[i]]$`2.5`
   dat$SST_975[(ind[i]+1):(ind[i+1])] <- datr[[i]]$`97.5`
   dat$Species[(ind[i]+1):(ind[i+1])] <- datr[[i]]$Species
+  dat$Preservation[(ind[i]+1):(ind[i+1])] <- datr[[i]]$Preservation
   
 }  
 
@@ -214,7 +215,8 @@ for(i in 1:nrow(locpar)) {
 ### EECO could be 53.3 - 49.2 Ma (according to the figures in Hollis et al 2019)
 ### However, data between range(as.numeric(dat$Age[which(dat$Timeslice=="EECO")]),na.rm=T)
 ### i.e. from 48.57246 to 53.48746 has been assigned to EECO in their database. Hence, we use that
-dat$Timeslice[which(is.na(dat$Timeslice) & (as.numeric(dat$Age) >= 48.57246 | as.numeric(dat$Age) <= 53.48746))] <- "EECO"
+#dat$Timeslice[which(is.na(dat$Timeslice) & (as.numeric(dat$Age) >= 48.57246 | as.numeric(dat$Age) <= 53.48746))] <- "EECO"
+# DON'T DO THIS IT IS WRONG, WILL BE TAKEN CARE OF IN THE SYNTHESIZE SCRIPT
 
 saveRDS(dat,"data/processed/Hollis_2019_TEX86_2022_07_18.rds")
 

### 2022-07-19 R/subscripts/process_data/synthesize_hollis2019.R

@@ -1,0 +1,95 @@
+d18O <- readRDS("data/processed/Hollis_2019_dO18_2022_07_15.rds")
+MgCa <- readRDS("data/processed/Hollis_2019_MgCa_2022_07_17.rds")
+d47 <- readRDS("data/processed/Hollis_2019_D47_2022_07_18.rds")
+TEX86 <- readRDS("data/processed/Hollis_2019_TEX86_2022_07_18.rds")
+
+head(d18O)
+
+sum(table(d18O$Timeslice))/nrow(d18O)
+sum(table(MgCa$Timeslice))/nrow(MgCa)
+sum(table(d47$Age))/nrow(d47)
+sum(table(TEX86$Timeslice))/nrow(TEX86)
+
+range(as.numeric(d18O$GTS2012[which(d18O$Timeslice=="EECO?")]),na.rm=T)
+
+### assign d18O data to EECO if they are within the range of 49.11000 53.76049
+### simplify to 49.1 - 53.8
+d18O$EECO <- rep(0,nrow(d18O))
+d18O$EECO[which(d18O$GTS2012 >= 49.1 & d18O$GTS2012 <=53.8 &!is.na(as.numeric(d18O$GTS2012)))] <- 1
+table(d18O$Timeslice[which(is.na(d18O$GTS2012))])
+## we should have captured all EECO data
+
+### assign MgCa data to EECO if they are within the range of 49.1 - 53.8
+MgCa$EECO <- rep(0,nrow(MgCa))
+MgCa$EECO[which(MgCa$Age >= 49.1 & MgCa$Age <=53.8 & !(is.na(as.numeric(MgCa$Age))))] <- 1
+table(MgCa$Timeslice[which(is.na(MgCa$Age))])
+## we should have captured all EECO data
+
+### assign d47 data to EECO if they are within the range of 49.1 - 53.8
+d47$EECO <- rep(0,nrow(d47))
+d47$EECO[which(d47$Age >= 49.1 & d47$Age <=53.8)] <- 1
+table(MgCa$Timeslice[which(is.na(MgCa$Age))])
+## all of them are EECO
+
+### assign TEX86 data to EECO if they are within the range of 49.1 - 53.8
+TEX86$EECO <- rep(0,nrow(TEX86))
+TEX86$EECO[which(TEX86$Age >= 49.1 & TEX86$Age <=53.8 & !is.na(as.numeric(TEX86$Age)))] <- 1
[…]

### 2022-07-22 manuscript/manuscript.Rmd
add notes to ms
describe mangroves as a proxy

@@
 
+In order to properly compare model output to proxy data, latitudinal temperature patterns need to be reconstructed from noisy data (impossible to properly compare model gradients with scattered proxy points "by eye")
+
 Example reference [@R-base]
@@
 # Materials & Methods
+
+We chose a range of proxies for marine, shallow water temperatures
+proxies: d18O, D47, Mg/Ca, TEX86
+mangroves, corals, sharks?
+
+## early Eocene SST proxies ##
+
+**d18O.**
+
+**d47.**
+
+**Mg/Ca.**
+
+**TEX86.**
+
+**Mangroves.** Recent mangroves predominantly occur in the tropics - subtropics.
+Although many factors besides temperature determine the extent of mangrove distributions, empirical temperature limits have been established for the genera *Avicennia* and *Rhizophora*, with lower mean annual SST limits of 15.6 degC and 20.8 degC, respectively [@Quisthoudt2012].
+Both *Avicennia* and members of the Rhizophoraceae family were widespread in the early Eocene, but only *Avicennia* occurred in polar latitudes [@Suan2017; @Popescu2021].
+Assuming that Eocene members of these mangrove taxa observed similar climatic requirements as their modern relatives, the presence and absence of *Avicennia* and Rhizophoraceae pollen can be used as a temperature proxy.
+To translate mangrove occurrences from the compilation by @Popescu2021 to a quantitative temperature proxy, we allocate locations at which *Avicennia* pollen, but no Rhizophoraceae pollen were found, a mean temperature between the lower *Avicennia* and Rhizophoraceae limit (18.2 degC), with a standard deviation of 1.33 degC, thus placing 95 % of the probability density within that interval (see Fig.
+1x).
+For locations at which both *Avicennia* and Rhizophoraceae pollen were found, the range of plausible temperatures is very wide and was not used in the analysis, as both mangrove taxa occur in temperatures > 32 deC in the modern.
+
+**Corals.**
+
+[Fig 1: In this part we can show a figure with the proxy data and the distributions of temperature proxies we specify, like mangroves]
+
+## A hierarchical model of latitudinal temperature gradients ##
+
+[Fig 2: In this part we can show a figure with the model structure and the priors we put on the parameters. Also, try it on the modern data? Alternatively, this figure goes to the SI]
 
@@
 
-## Fig 1 - Model structure and proof of concept with modern data
-## Fig 2 - Eocene example
-## Fig 3 - Showing influence of changing model assumptions
+## Fig 3 - Eocene gradient
+## Fig 4 - compare Eocene gradient with previous reconstructions / ES models
 

### 2022-07-22 R/subscripts/AuxiliaryFunctions.R

@@ -99,9 +99,13 @@ plot_posterior <- function(mod,burnin = NULL, lat = seq(0,90,0.2), confint_n = N
   if(!(is.null(mod$sdyest))) {
   nobsloc <- dim(mod$sdyest)[2]
   samples <- 1:dim(mod$sdyest)[2]
-  invisible(sapply(samples,function(x) points(mod$lat[x], mean(mod$yestimate[burnin:nIter,x]), pch = pch, col = col_obs, cex = cex)))
+  if(length(col_obs)!=nobsloc)  col_obs <- rep(col_obs[1],nobsloc)
+  
+  invisible(sapply(samples,function(x) points(mod$lat[x], mean(mod$yestimate[burnin:nIter,x]), pch = pch, col = col_obs[x], cex = cex)))
+  
   invisible(sapply(samples,function(x) points(rep(mod$lat[x],2), quantile(mod$yestimate[burnin:nIter,x], probs = c(0.05,0.95)), 
-                                    type = "l", col = col_obs)))
+                                    type = "l", col = col_obs[x])))
+ 
   } else nobsloc <- 0
   
   if(dim(mod$yestimate)[2] > nobsloc) {

### 2022-07-22 R/tests/Hollis_2019_full_test.R

@@ -1,7 +1,9 @@
+
+
 dat <- readRDS("data/processed/Hollis_processed_2022_07_19.rds")
 dat$proxy_value <- as.numeric(dat$proxy_value)
 hist(dat$temperature,100)
-
+dat$preservation[which(dat$preservation=="Recrystallized")] <- "recrystallised"
 
 dat_sub <- subset(dat,proxy=="d18O")
 dat_sub$mineral = "calcite"
@@ -29,8 +31,10 @@ plot(grosman_t,dat_sub$temperature)
                        (is.na(depth_habitat) | depth_habitat %in% c("Mixed-layer",     "Mixed layer")))
   data_sub <- data_sub[with(data_sub, order(abs(paleolat_Meredith), longitude)),]
   
-  data_sub <- subset(data_sub,proxy=="d18O")
+  #data_sub <- subset(data_sub,proxy=="MgCa")
   plot(data_sub$latitude,data_sub$temperature,xlim=c(-90,90), pch = 17, col = rgb(1,0,0,0.5), ylim = c(0,45))
+  plot(data_sub$latitude,data_sub$temperature,xlim=c(-90,90), pch = 17, col = rgb(data_sub$preservation=="recrystallised",0,0,0.5), ylim = c(0,45))
+
   points(data_sub$latitude[which(data_sub$preservation %in% c("recrystallised","Recrystallized"))],
          data_sub$temperature[which(data_sub$preservation %in% c("recrystallised","Recrystallized"))],
          col = NA, pch = 21, bg = rgb(1,0,0,0.5))
@@ -40,11 +44,43 @@ plot(grosman_t,dat_sub$temperature)
   # prepare for use in the model
   obsmat <- data.frame(sample = (paste(abs(data_sub$paleolat_Meredith),data_sub$longitude)),
                          latitude = abs(data_sub$paleolat_Meredith), temperature = data_sub$temperature,
-                         sd = data_sub$temperature_sd)
+                         sd = data_sub$temperature_sd,
+                       proxy = data_sub$proxy)
+  proxy_location_ordered <- sapply(unique(obsmat$sample), function(x) unique(obsmat$proxy[which(obsmat$sample==x)]))
[…]

### 2022-07-22 manuscript/manuscript.Rmd
minor corrections on mangrove paragraph

@@
 Both *Avicennia* and members of the Rhizophoraceae family were widespread in the early Eocene, but only *Avicennia* occurred in polar latitudes [@Suan2017; @Popescu2021].
-Assuming that Eocene members of these mangrove taxa observed similar climatic requirements as their modern relatives, the presence and absence of *Avicennia* and Rhizophoraceae pollen can be used as a temperature proxy.
+Assuming that Eocene members of these mangrove taxa had similar climatic requirements as their modern relatives, the presence and absence of *Avicennia* and Rhizophoraceae pollen can be used as a temperature proxy.
 To translate mangrove occurrences from the compilation by @Popescu2021 to a quantitative temperature proxy, we allocate locations at which *Avicennia* pollen, but no Rhizophoraceae pollen were found, a mean temperature between the lower *Avicennia* and Rhizophoraceae limit (18.2 degC), with a standard deviation of 1.33 degC, thus placing 95 % of the probability density within that interval (see Fig.
 1x).
-For locations at which both *Avicennia* and Rhizophoraceae pollen were found, the range of plausible temperatures is very wide and was not used in the analysis, as both mangrove taxa occur in temperatures > 32 deC in the modern.
+The range of temperatures in which both taxa occur in the modern is very wide (20.8 and 28 deC), hence occurrences of both *Avicennia* and Rhizophoraceae pollen were not used in the analysis.
 

### 2022-08-09 manuscript/manuscript.Rmd
modern run completed, first full Eocene run

@@
 
-We chose a range of proxies for marine, shallow water temperatures
+## early Eocene SST proxies
+
+Proxy-based reconstructions of Cenozoic shallow water temperatures chose a range of proxies for marine, shallow water temperatures
 proxies: d18O, D47, Mg/Ca, TEX86
 mangroves, corals, sharks?
-
-## early Eocene SST proxies ##
 
@@
 
-**Mangroves.** Recent mangroves predominantly occur in the tropics - subtropics.
+**Coral reefs.** Coral reefs built by zooxanthellate-bearing, shallow water corals are confined to the tropics - substropics. The coldest
+
+**Mangroves.** Recent mangroves occur in the tropics - subtropics.
 Although many factors besides temperature determine the extent of mangrove distributions, empirical temperature limits have been established for the genera *Avicennia* and *Rhizophora*, with lower mean annual SST limits of 15.6 degC and 20.8 degC, respectively [@Quisthoudt2012].
@@
 The range of temperatures in which both taxa occur in the modern is very wide (20.8 and 28 deC), hence occurrences of both *Avicennia* and Rhizophoraceae pollen were not used in the analysis.
-
-**Corals.**
 

### 2022-08-09 R/run-EECO-gradient.R

@@ -17,28 +17,36 @@ obsmat <- data.frame(sample = (paste(abs(data_sub$paleolat_Meredith),data_sub$lo
                      proxy = data_sub$proxy)
 
 ### Mangrove and Coral data
-# for now use this placeholder
-
+# read and assign proxy type
 bioprox <- readRDS("data/processed/bio_proxies_2022_08_08.RDS")
+bioprox$proxy <- rep("Reef",nrow(bioprox))
+bioprox$proxy[which(bioprox$taxa=="Avicennia")] <- "Avicennia"
+bioprox$proxy[which(bioprox$type=="mangrove" & bioprox$taxa!="Avicennia")] <- "Avicennia-Rhizophoraceae"
 
-distrmat = data.frame(latitude = 79.4 , ## palaeorotated from Faddeevsky Island: palaeorotate(data.frame(lat = 75.5, lng = 144, age = 52))
-                      location = mean(c(15.6,20.8)),
-                      scale = 1.33,
-                      shape = NA,
-                      distribution = "normal")
+# define proxy distributions
+proxy_distributions <- data.frame(name = c("Avicennia", "Avicennia-Rhizophoraceae", "Reef"),
+               distribution = c("normal", "normal", "normal"),
+                mean = c(mean(c(15.6,22.5)), mean(c(20.7,29.5)), 27.6),
+                sd = c((22.5-15.6)/4, c(29.5-20.7)/4, (29.5-21)/4),
+                shape = rep(NA,3))
+
+# create distribution matrix for use with model
+proxy_index <- sapply(bioprox$proxy, function(f) which(proxy_distributions$name==f))
+
+distrmat = data.frame(latitude = abs(bioprox$lat), 
+                      location = proxy_distributions$mean[proxy_index],
+                      scale = proxy_distributions$sd[proxy_index],
+                      shape = proxy_distributions$shape[proxy_index],
+                      distribution = proxy_distributions$distribution[proxy_index])
 
 ### Define the priors
-prior_fun <- list(  
-  f1 = function(x,log) dsnorm(x,location = -2.66, scale = 20, alpha = 20, log = log), # prior on A (lower asymptote)
-  f2 = function(x,lower,log) dtnorm(x, lower, upper = Inf, mean = 30, sd = 10, log = log), # prior on upper asymptote
[…]

### 2022-08-09 R/subscripts/AuxiliaryFunctions.R

@@ -39,7 +39,7 @@ gradient <- function(x, coeff, sdy) { # parametrise with difference between cold
 
 
 plot_gradient <- function(model_out, burnin = NULL, lat = seq(0,90,0.2), confint_n = NULL, add = F,
-                          ylim = NULL, line_col = "black", confint_col = rgb(0,0,0,0.2)) {
+                          ylim = NULL, line_col = "black", confint_col = rgb(0,0,0,0.2), lwd = 2) {
   # select only params
   if("params" %in% names(model_out)) model_out <- model_out$params
   
@@ -58,13 +58,13 @@ plot_gradient <- function(model_out, burnin = NULL, lat = seq(0,90,0.2), confint
   
   if(add == F) {
   plot(lat, med_grad, ylim = ylim,
-       type = "l", lwd = 2, xlab = "|latitude|", ylab = "temperature", col = line_col)
+       type = "l", lwd = lwd, xlab = "|latitude|", ylab = "temperature", col = line_col)
   error_polygon(lat,grad_q["2.5%",],grad_q["97.5%",], col = confint_col)
   }
   if(add == T) {
     error_polygon(lat,grad_q["2.5%",],grad_q["97.5%",], col = confint_col)
     points(lat, med_grad,
-           type = "l", lwd = 2, col = line_col)
+           type = "l", lwd = lwd, col = line_col)
   }
 }
 

### 2022-08-09 R/subscripts/ModernClimate.R

@@ -1,26 +1,26 @@
 ### load climate data
 # Mean annual sea surface temperatures (Bio-Oracle)
+set.seed(1)
 
 sst <- raster("./data/raw/climate/BioOracle_20220711/Present.Surface.Temperature.Mean.asc")
 
 # sample points randomly from the raster (within specified latitudinal limits) 
 source("R/subscripts/AuxiliaryFunctions.R")
-coords <- fastRandomPoints_lat(sst$Present.Surface.Temperature.Mean,2000,-90,90)
+coords <- fastRandomPoints_lat(sst$Present.Surface.Temperature.Mean,10000,-90,90)
 
 # extract temperatures from the sampled points
 sstr <- raster::extract(sst,coords)
 
 lat <- abs(coords[,2])
 
-plot(lat,sstr, pch = 19
-     , col = rgb(0,0,0,0.1))
+plot(lat,sstr, pch = 21,col=NA, bg = rgb(0,0,0,0.1))
 
 ### Define the priors
 prior_fun <- list(  
-  f1 = function(x,log) dsnorm(x,location = -2.66, scale = 20, alpha = 20, log = log), # prior on A (lower asymptote)
-  f2 = function(x,lower,log) dtnorm(x, lower, upper = Inf, mean = 30, sd = 10, log = log), # prior on upper asymptote
-  f3 = function(x,log) dnorm(x, mean = 45, sd = 12, log = log), # prior on M
-  f4 = function(x,log) dlnorm(x, mean = -2.4, sd = 0.6, log = log)# prior on Q
+  f1 = function(x,log) dsnorm(x,location = -2.93, scale = 12, alpha = 30, log = log), # prior on A (lower asymptote)
+  f2 = function(x,lower,log) dtnorm(x, lower, upper = Inf, mean = 28.3, sd = 10, log = log), # prior on upper asymptote
+  f3 = function(x,log) dnorm(x, mean = 42.1, sd = 12, log = log), # prior on M
+  f4 = function(x,log) dgamma(x, shape = 3.2, rate = 20, log = log)# prior on Q
 )
 
 # set up clusters for parallel computing
@@ -37,7 +37,7 @@ system.time({modm <- climate_simple_parallel(nChains = nClust, nIter = 100000, n
                                 x = lat, y = sstr, 
                                 coeff_inits = NULL, sdy_init = NULL, 
[…]

### 2022-08-09 R/tests/Hollis_2019_full_test.R

@@ -63,7 +63,9 @@ dat <- readRDS("data/processed/Hollis_processed_2022_07_19.rds")
   proxy_type_loc[which(apply(proxy_type_loc[,c(1,2,4)],1,function(x) all(x==1*0.9))),] <- c(.7,.6,0,.2)
   
   proxy_to_col <- apply(proxy_type_loc,1,function(x) rgb(x[1],x[2],x[4],0.75))
-##  
+  
+
+  ##  
   
   ### Add Siberian mangroves (Suan et al 2017)
   # test some distributions -- M
@@ -167,7 +169,7 @@ mod2<- run_MCMC_sd_obs(nIter = nIter, obsmat = NULL, distrmat = NULL, coeff_init
   plot_data(obsmat,distrmat, add = T,col=rgb(0,0,0,0.2))
   plot_distr(distrmat,col=rgb(0,0.7,0.7,0.3))
   
-  plot_posterior(mod1,col_obs = proxy_to_col, col_dist = rgb(0,0.6,0,0.7))
+  plot_posterior(mod2[[1]],col_obs = proxy_to_col, col_dist = rgb(0,0.6,0,0.7))
   
   plot_posterior(mod2,col_obs = rgb(0.5,0,0.8,0.7), col_dist = rgb(0.5,0,0.8,0.7))
   

### 2022-08-10 R/run-EECO-gradient.R
Cleaning repo

@@ -6,8 +6,8 @@ dat <- readRDS("data/processed/Hollis_processed_EECO_2022_07_19.rds")
 
 
 # prepare for use in the model
-obsmat <- data.frame(sample = (paste(abs(data_sub$paleolat_Meredith),data_sub$longitude, data_sub$proxy)),
-                     latitude = abs(data_sub$paleolat_Meredith), temperature = data_sub$temperature,
+obsmat <- data.frame(sample = (paste(abs(data_sub$p_lat),data_sub$longitude, data_sub$proxy)),
+                     latitude = abs(data_sub$p_lat), temperature = data_sub$temperature,
                      sd = data_sub$temperature_sd,
                      proxy = data_sub$proxy)
 
@@ -28,7 +28,7 @@ proxy_distributions <- data.frame(name = c("Avicennia", "Avicennia-Rhizophoracea
 # create distribution matrix for use with model
 proxy_index <- sapply(bioprox$proxy, function(f) which(proxy_distributions$name==f))
 
-distrmat = data.frame(latitude = abs(bioprox$lat), 
+distrmat = data.frame(latitude = abs(bioprox$p_lat), 
                       location = proxy_distributions$mean[proxy_index],
                       scale = proxy_distributions$sd[proxy_index],
                       shape = proxy_distributions$shape[proxy_index],
@@ -54,11 +54,25 @@ cl <- parallel::makeCluster(nClust)
 doParallel::registerDoParallel(cl)
 
 ### Run model
-mod2 <- climate_parallel_sd(nChains = nClust, nIter = 25000, nThin = 5, obsmat = obsmat, distrmat = distrmat, 
+mode <- climate_parallel_sd(nChains = nClust, nIter = 25000, nThin = 5, obsmat = obsmat, distrmat = distrmat, 
                             coeff_inits = NULL, sdy_init = NULL, 
                             yest_inits = NULL, sdyest_inits = NULL, prior_fun = prior_fun,
                             proposal_var_inits = c(3,3,3,0.2), adapt_sd = 2500,
                             adapt_sd_decay = 100, start_adapt = 101, quiet = FALSE) 
+# Run model without bio proxies
+mode_nobio <- climate_parallel_sd(nChains = nClust, nIter = 25000, nThin = 5, obsmat = obsmat, distrmat = NULL, 
[…]

### 2022-08-10 R/subscripts/AuxiliaryFunctions.R

@@ -152,29 +152,6 @@ plot_posterior <- function(mod,burnin = NULL, lat = seq(0,90,0.2), confint_n = N
   }
 }
 
-coral_distrmat <- function(data,stage) { # for PARED
-  
-  data_sub <- subset(data,early_stage == stage & !(is.na(pal_lat_scotese)))#table(iso$stage_2020)
-  data_sub <- data_sub[with(data_sub, order(abs(pal_lat_scotese), longit)),]
-  # prepare for use in the model
-  if(nrow(data_sub) >= 1) {data_distrmat <- data.frame(latitude = abs(data_sub$pal_lat_scotese),
-                                                         location = 22.8,
-                                                         scale = 10,
-                                                         shape = 4,
-                                                         distribution = "skew-normal")
-  } else data_distrmat <- data.frame(NULL)
-  
-}
-  
-iso_obsmat <- function(data, stage) { # for StabisoDB
-  # select data from one stage to test, exclude NA data
-  data_sub <- subset(data,stage_2020 == stage & !(is.na(paleolat)) & !(is.na(temperature)))
-  data_sub <- data_sub[with(data_sub, order(abs(paleolat), longitude)),]
-  # prepare for use in the model
-  data_mod <- data.frame(sample = (paste(abs(data_sub$paleolat),data_sub$longitude)),
-                        latitude = abs(data_sub$paleolat), temperature = data_sub$temperature)
-  return(data_mod)
-}
 
 
 plot_chains <- function(mod, params = 1:4, nthin = NULL, logQ = TRUE) {
@@ -230,28 +207,6 @@ plot_dens <- function(x,dens,xlim = NULL,ylim = NULL,col = rgb(0,0.5,0.7,0.25),
   
 }
 
-fastRandomPoints <- function(r, n) {
-  if(raster::nlayers(r) > 1) r <- r[[1]]
-  v <- raster::getValues(r)
-  v.notNA <- which(!is.na(v))
-  x <- sample(v.notNA, n)
-  pts <- raster::xyFromCell(r, x)
-  return(pts)
-}
-fastRandomPoints_lat <- function(r, n, min, max) {
-  #if(raster::nlayers(r) > 1) r <- r[[1]]
-  #row = raster::rowFromY(r, c(min,max))
-  #row = row[1]:row[2]
[…]

### 2022-08-10 R/subscripts/ClimateAllStages.R

@@ -1,116 +1,0 @@
-### Function to generate climate gradient across all stages
-
-### Does not quite work yet
-
-## vector with stages
-###### 
-stages <- c("Pre-Cambrian boundary", "Fortunian", "Stage 2", "Stage 3", "Stage 4", "Wuliuan", "Drumian", 
-            "Guzhangian", "Paibian", "Jiangshanian", "Stage 10", "Tremadocian", 
-            "Floian", "Dapingian", "Darriwilian", "Sandbian", "Katian", "Hirnantian", 
-            "Rhuddanian", "Aeronian", "Telychian", "Sheinwoodian", "Homerian", 
-            "Gorstian", "Ludfordian", "Pridoli", "Lochkovian", "Pragian", 
-            "Emsian", "Eifelian", "Givetian", "Frasnian", "Famennian", "Tournaisian", 
-            "Visean", "Serpukhovian", "Bashkirian", "Moscovian", "Kasimovian", 
-            "Gzhelian", "Asselian", "Sakmarian", "Artinskian", "Kungurian", 
-            "Roadian", "Wordian", "Capitanian", "Wuchiapingian", "Changhsingian", 
-            "Induan", "Olenekian", "Anisian", "Ladinian", "Carnian", "Norian", 
-            "Rhaetian", "Hettangian", "Sinemurian", "Pliensbachian", "Toarcian", 
-            "Aalenian", "Bajocian", "Bathonian", "Callovian", "Oxfordian", 
-            "Kimmeridgian", "Tithonian", "Berriasian", "Valanginian", "Hauterivian", 
-            "Barremian", "Aptian", "Albian", "Cenomanian", "Turonian", "Coniacian", 
-            "Santonian", "Campanian", "Maastrichtian", "Danian", "Selandian", 
-            "Thanetian", "Ypresian", "Lutetian", "Bartonian", "Priabonian", 
-            "Rupelian", "Chattian", "Aquitanian", "Burdigalian", "Langhian", 
-            "Serravallian", "Tortonian", "Messinian", "Zanclean", "Piacenzian", 
-            "Gelasian", "Calabrian", "Chibanian", "Upper Pleistocene", "Holocene")
-######
-
-# load isotope data
-iso <- readRDS("data/processed/StabisoDB_processed_26_06_2022.rds")
-
-# load coral reef data
-coral <- readRDS("data/processed/PARED_coralreefs_processed_02_07_2022.rds")
-
-
-# prepare parallel running
[…]

### 2022-08-10 R/subscripts/ClimateGradientModelwithSDonObs.R

@@ -207,6 +207,9 @@ run_MCMC_sd_obs <- function(nIter = 1000, nThin = 1, obsmat = NULL, distrmat = N
   accept = rep(NA,nIter)
   x = NULL
   if(!is.null(obsmat)) {
+    if(is.list(sapply(unique(obsmat$sample), function(x) unique(obsmat$latitude[which(obsmat$sample == x)])))) warning(
+      "something is wrong with the observation matrix. Do all obs from one sample have the same latitude?")
+    
     ylist <- lapply(unique(obsmat$sample), function(x) obsmat$temperature[which(obsmat$sample == x)])
     x <- as.numeric(sapply(unique(obsmat$sample), function(x) unique(obsmat$latitude[which(obsmat$sample == x)])))
     n_p <- length(ylist)

### 2022-08-10 R/subscripts/ModernClimate.R

@@ -2,18 +2,21 @@
 # Mean annual sea surface temperatures (Bio-Oracle)
 set.seed(1)
 
-sst <- raster("./data/raw/climate/BioOracle_20220711/Present.Surface.Temperature.Mean.asc")
+sst <- raster::raster("./data/raw/climate/BioOracle_20220711/Present.Surface.Temperature.Mean.asc")
 
 # sample points randomly from the raster (within specified latitudinal limits) 
 source("R/subscripts/AuxiliaryFunctions.R")
-coords <- fastRandomPoints_lat(sst$Present.Surface.Temperature.Mean,10000,-90,90)
 
-# extract temperatures from the sampled points
-sstr <- raster::extract(sst,coords)
+coords <- raster::sampleRandom(x = sst, 
+                                            size = 1000,
+                                            na.rm = TRUE,
+                                            xy=T)
+# temperatures from the sampled points
+ssts <- coords[,3]
 
 lat <- abs(coords[,2])
 
-plot(lat,sstr, pch = 21,col=NA, bg = rgb(0,0,0,0.1))
+plot(lat,ssts, pch = 21,col=NA, bg = rgb(0,0,0,0.1))
 
 ### Define the priors
 prior_fun <- list(  
@@ -34,7 +37,7 @@ doParallel::registerDoParallel(cl)
 source("R/subscripts/ClimateParallelSimple.R")
 #run
 system.time({modm <- climate_simple_parallel(nChains = nClust, nIter = 100000, nThin = 10,
-                                x = lat, y = sstr, 
+                                x = lat, y = ssts, 
                                 coeff_inits = NULL, sdy_init = NULL, 
                                 prior_fun = prior_fun,
                                 proposal_var_inits = c(2,2,2,0.2), adapt_sd = 2500,
@@ -48,7 +51,7 @@ parallel::stopCluster(cl)
 plot_chains(modm)
 plot_gradient(modm[[1]],burnin = 1000, ylim = c(-2,30))
 plot_gradient(modm[[2]],burnin = 1000, add = T, confint_col = rgb(0,0.5,1,0.2), line_col = rgb(0,0.5,1,1))
-points(lat,sstr, pch = 21, bg = rgb(0,0,0,0.2), col = NA)
[…]

### 2022-08-10 R/subscripts/PARED.R

@@ -1,245 +1,0 @@
-setwd("C://Users/keichenseer/OneDrive - University of Plymouth/data/reefzones")
-setwd("D://OneDrive - Durham University/Plymouth/data/reefzones")
-
-reef <- read.csv("pared_2021_06_12.csv") # reed the PARED data
-reef_text <- read.csv("pared_2021_06_12_text.csv") # reed the PARED data
-
-timebins <- read.csv(file="clean_timebins.csv", header=TRUE, sep=",") # read Phanerozoic international stratigraphy (Gradstein 2016)
-other_intervals <- read.csv(file="other_intervals.csv", header=TRUE, sep = ";") # read list of other stratigraphic intervals
-trim.leading <- function (x)  sub("^\\s+", "", x) # get rid of leading spaces
-other_intervals$early_stage <- trim.leading(other_intervals$early_stage)
-other_intervals$late_stage <- trim.leading(other_intervals$late_stage)
-
-length(table(reef$intervall))
-
-reef$intervall <- gsub("\\?","",reef$intervall) # remove the question marks from the intervall column
-reef$intervall <- gsub("\\(","",reef$intervall) # remove the parentheses from the intervall column
-reef$intervall <- gsub("\\)","",reef$intervall) # # remove the parentheses from the intervall column
-
-reef$intervall <- gsub("\\/","-",reef$intervall) # transform "/" into "-"
-reef$intervall <- gsub("\\ to ","-",reef$intervall) # transform " to " into "-"
-reef$intervall <- gsub("\\-"," ",reef$intervall) # transform "-" to " "
-reef$intervall <- gsub("\\  "," ",reef$intervall) # transform "  " to " "
-reef$intervall <- gsub("\\  "," ",reef$intervall) # transform "   " to " "
-reef$intervall <- gsub("\\Upper","upper",reef$intervall) # standardise capitalisation
-reef$intervall <- gsub("\\Middle","middle",reef$intervall) # standardise capitalisation
-reef$intervall <- gsub("\\Lower","lower",reef$intervall) # standardise capitalisation
[…]

### 2022-08-10 R/subscripts/StabisoDB.R

@@ -1,148 +1,0 @@
-###
-## Transfer function for calcite from Kim and O'Neil (1997),
-#  re-expressed in Leng and Marshall (2004):
-#
-calcite_temp <- function(d18O_calcite,d18O_sw) {
-  13.8 - 4.58 * (d18O_calcite - d18O_sw) + 0.08 * (d18O_calcite - d18O_sw)^2
-}
-
-calcite_temp(-1.6,-0.78)
-calcite_temp(-1.17,-0.67)
-
-## belemnite correction: -1.5
-## aragonite correction: +0.6
-
-## latitude correction
-lat_correction <- function(lat, d18O_sw_ice) {
-  lat = abs(lat)
-  d18O_sw_lat_mod = -6.650 * 10^(-4)*lat^2 + 3.363 * 10^(-2) * lat + 0.174
-  d18O_sw_lat_pg = -4.944 * 10^(-4)*lat^2 + 2.492 * 10^(-2) * lat + 0.102
-  out = d18O_sw_lat_mod * (d18O_sw_ice--1.08)/(0.45--1.08) + d18O_sw_lat_pg * (1-(d18O_sw_ice--1.08)/(0.45--1.08))
-  return(out)
-}
-
-lat_correction(40.21,-1.08)
-# test - seems to be roughly correct, according to the table from Grossman and Joachmiski 2022
-
-###
-## Transfer function for phosphate from Puceat et al. (2010),
-#  which version?
-#
-apatite_temp <- function(d18O_apatite,d18O_sw) {
-118.7 - 4.22 * (d18O_apatite - d18O_sw)
-}
-# do we need to correct for standard d18O in the apatite samples? Ask Grossman / Joachimski
-
-get_temperature <- function(d18O_sample, d18O_sw_ice, lat, mineral, fossil_group) {
-  out = rep(NA_real_,length(d18O_sample))
-  
-  index1 <- which(mineral == "calcite" & fossil_group != "Belemnite")
-  if(length(index1)>=1) out[index1] <- calcite_temp(
-    d18O_sample[index1], d18O_sw_ice[index1] + lat_correction(lat[index1],d18O_sw_ice[index1]))
-  
-  index2 <- which(mineral == "calcite" & fossil_group == "Belemnite")
-  if(length(index2)>=1) out[index2] <- calcite_temp(
-    d18O_sample[index2]- 1.5, d18O_sw_ice[index2] + lat_correction(lat[index2],d18O_sw_ice[index2]))
-  
[…]

### 2022-08-10 R/tests/model_test_with_stabisodb.R

@@ -1,77 +1,0 @@
-### A test with StabisoDB data
-### 26/06/2022
-###
-
-# load data
-dat <- readRDS("data/processed/StabisoDB_processed_26_06_2022.rds")
-
-# select data from one stage to test, exclude NA data
-dat_sub <- subset(dat,stage_2020 == "Rhuddanian" & !(is.na(paleolat)) & !(is.na(temperature)))#table(dat$stage_2020)
-dat_sub <- dat_sub[with(dat_sub, order(abs(paleolat), longitude)),]
-# prepare for use in the model
-dat_mod <- data.frame(sample = (paste(abs(dat_sub$paleolat),dat_sub$longitude)),
-                       latitude = abs(dat_sub$paleolat), temperature = dat_sub$temperature)
-samples1 <- unique(dat_mod$sample)
-for(i in 1:length(samples1)) dat_mod$sample[which(dat_mod$sample==samples1[i])] <- i
-
-samples <-  1:length(samples1)
-sample_lats <- sapply(samples, function(x) unique(dat_mod$lat[which(dat_mod$sample==x)]))
-  
-# visualise data
-plot(abs(dat_mod$latitude),dat_mod$temperature)
-
-# wait a second - how do we deal with situationsn with 1 value per locality?
-## ok, this is taken care of by the prior on yest ... although it is counterintuitive, as yest deviates less 
-## with larger priors on sdyest. Definetly some prior sensitivity tests are in order. 
-### resolved, it is an inverse gamma distribution, hence larger scale leads to smaller deviations, all good.
-
-# call model functions
-source("R/subscripts/ClimateGradientModel.R")
-
-# initial temperature values for the sample locations
-yest_inits <- sapply(unique(dat_mod$sample), function(x) mean(dat_mod$temperature[which(dat_mod$sample == x)]))
-
-nIter <- 20000
-
-
-# run the model
-mod <- run_MCMC(nIter = nIter, obsmat = dat_mod, distrmat = NULL, coeff_inits = c(10,20,45,0.1), sdy_init = 1, 
-         yest_inits = yest_inits, sdyest_inits = rep(2,length(yest_inits)),
-                     proposal_var_inits = c(2,2,2,0.2))
-
-
-burnin <- 4000+1 # burnin - to be discarded for the plots
-
-# function for plotting the 95 % CI shading
[…]

### 2022-08-10 R/tests/model_test_with_stabisodb_and_coralreefs.R

@@ -1,166 +1,0 @@
-
-### A test with StabisoDB and coral reef data
-### 02/07/2022
-###
-
-mystage <- "Turonian"
-
-# load isotope data
-iso <- readRDS("data/processed/StabisoDB_processed_26_06_2022.rds")
-
-# select data from one stage to test, exclude NA data
-iso_sub <- subset(iso,stage_2020 == mystage & !(is.na(paleolat)) & !(is.na(temperature)))#table(iso$stage_2020)
-iso_sub <- iso_sub[with(iso_sub, order(abs(paleolat), longitude)),]
-# prepare for use in the model
-iso_mod <- data.frame(sample = (paste(abs(iso_sub$paleolat),iso_sub$longitude)),
-                      latitude = abs(iso_sub$paleolat), temperature = iso_sub$temperature)
-samples1 <- unique(iso_mod$sample)
-for(i in 1:length(samples1)) iso_mod$sample[which(iso_mod$sample==samples1[i])] <- i
-
-nsamples <- length(samples1)
-samples <-  1:(max(1,nsamples))
-sample_lats <- sapply(samples, function(x) unique(iso_mod$lat[which(iso_mod$sample==x)]))
-
-
-# load coral reef data
-coral <- readRDS("data/processed/PARED_coralreefs_processed_02_07_2022.rds")
-
-coral_sub <- subset(coral,early_stage == mystage & !(is.na(pal_lat_scotese)))#table(iso$stage_2020)
-coral_sub <- coral_sub[with(coral_sub, order(abs(pal_lat_scotese), longit)),]
-# prepare for use in the model
-if(nrow(coral_sub) >= 1) {coral_distrmat <- data.frame(latitude = abs(coral_sub$pal_lat_scotese),
-                                                       location = 22.8,
-                                                       scale = 10,
-                                                       shape = 4,
-                                                       distribution = "skew-normal")
-} else coral_distrmat <- data.frame(NULL)
-
-# visualise data
-plot(abs(iso_mod$latitude),iso_mod$temperature, xlim = range(c(iso_mod$latitude, coral_distrmat$latitude)),
-     ylim = c(0,50))
-
-z1 <- list(NULL)
[…]

### 2022-08-10 R/tests/test_with_SD_on_observations.R

@@ -6,8 +6,10 @@ source("R/subscripts/AuxiliaryFunctions.R")
 obsmat = data.frame(
   sample = c(1,1,1,2,2,2,3,3,3,4,4,4,5,5,6,7,7),
   latitude = c(9,9,9,11,11,11,13,13,13,30,30,30,60,60,40,70,70),
-  temperature = c(30,33,30,36,39,36,29,30,29.5,22,24,23,8,9,19,6,5),
-  sd = c(NA,NA,NA,2,3,2,1,1,1,1,1,NA,1,1,2,1,NA))
+  temperature = c(30,33,30,38.3,38.5,38.7,29,30,29.5,22,24,23,8,9,19,6,5),
+  sd = c(NA,NA,NA,3,3,3,1,1,1,1,1,NA,1,1,2,1,NA))
+#obsmat$sd = NA
+plot(obsmat$latitude,obsmat$temperature)
 
 distrmat = NULL
 
@@ -37,14 +39,22 @@ if((!(is.null(distrmat)) | !(is.null(obsmat))) & is.null(yest_inits)) {
                   distrmat$location + distrmat$scale * sqrt(2/pi) * 
                     distrmat$shape/sqrt(1+distrmat$shape^2) )
 }
+
 if(!(is.null(obsmat)) & is.null(sdyest_inits)) sdyest_inits <- rep(2,length(unique(obsmat$sample)))
+prior_fun <- readRDS("results/modern/prior_from_modern_gradient.RDS")
+logprior <- write_logprior(prior_fun,log = TRUE)
 
 source("R/subscripts/ClimateGradientModelwithSDonObs.R")
-test4.4sd <- run_MCMC_sd_obs(nIter = 20000, obsmat = obsmat, distrmat = NULL, coeff_inits, sdy_init, yest_inits, sdyest_inits,
+test4.5sd <- run_MCMC_sd_obs(nIter = 20000, nThin = 1, obsmat = obsmat, distrmat = NULL, coeff_inits = coeff_inits, 
+                             sdy_init = sdy_init, yest_inits = yest_inits, sdyest_inits = sdyest_inits, logprior = logprior,
                                proposal_var_inits = c(2,2,2,0.2), adapt_sd = floor(0.2 * nIter),
                                adapt_sd_decay = max(floor(0.005*nIter),1),start_adapt = start_adapt, quiet = FALSE)
 
+plot_gradient(test4.4sd)
+plot_posterior(test4.4sd)
 
[…]

### 2022-08-11 manuscript/manuscript.Rmd
Cleaning repo
manuscript notes, option to fix sdy

@@
 
-The early Eocene climate optimum (EECO, ~ 53 - 49 million years ago) is thought to have been the warmest episode ever since.
-The EECO has received much attention due to its use as an analogue for extreme warming scenarios, depicting a possible future high-CO2 world.
+The early Eocene climate optimum (EECO, ~ 53 - 49 million years ago) is thought to have been the warmest episode of [...].
+The EECO is of particular interest due to its use as an analogue for extreme warming scenarios, depicting a possible future high-CO2 world.
 To reconstruct the climate of EECO, a variety of proxy data and Earth System Models have been used.
-Early modelers have soon noticed that their models struggle to reproduce the climate indicated by proxy data: Without unrealistically high concentrations of atmospheric CO2, early climate models struggled to reproduce the equitable polar climates indicated by early Eocene proxy data, e.g.
+Early modelers have soon noticed that their models struggle to reproduce the climate indicated by proxy data: Without unrealistically high concentrations of atmospheric CO2, early climate models were unable to reproduce the equitable polar climates indicated by early Eocene proxy data, e.g.
 the occurrence of mangroves in polar latitudes, and d18O data from x indicates a temperature of ~ y at latitude z.
 
+Contrast between complex, process-driven Earth system models and very simplistic use of the proxy data as local temperature estimates or global averages.
+Here, we introduce a proxy-driven modelling approach of relatively low complexity, which places the available chemical and biological proxy data in a single climate model.
 In order to properly compare model output to proxy data, latitudinal temperature patterns need to be reconstructed from noisy data (impossible to properly compare model gradients with scattered proxy points "by eye")

### 2022-08-11 R/subscripts/EECO_gradient.R

@@ -1,7 +1,7 @@
 
 ### Hollis data
 # 
-# dat <- readRDS("data/processed/Hollis_processed_2022_07_19.rds")
+ dat <- readRDS("data/processed/Hollis_processed_EECO_2022_07_19.rds")
 # # select data from one stage to test, exclude NA data
 # data_sub <- subset(dat,EECO == 1 & !(is.na(temperature)) & !(is.na(paleolat_Meredith)) &
 #                      (is.na(depth_habitat) | depth_habitat %in% c("Mixed-layer",     "Mixed layer")))
@@ -71,7 +71,7 @@ modeold <- foreach(pc = 1:nChains) %dopar% {
   hierarchical_model(n_iter = 30000, n_thin = 5,
                   obsmat = obsmat, distrmat = distrmat, 
                   logprior_input = prior_fun,adapt_sd = 3000,
-                  adapt_sd_decay = 100, A_sdy = 1, B_sdy = 1)
+                  adapt_sd_decay = 100, A_sdy = 1, B_sdy = 1, sdy_fixed = 2.2)
 }
 
 ### stop cluster
@@ -82,7 +82,7 @@ parallel::stopCluster(cl)
 #
 # Assess output
 
-plot_chains(mode1)
+plot_chains(modeold)
 plot(test$params$sdy)
 
 mode_allold <- combine_posterior(modeold,4000)
@@ -94,7 +94,7 @@ mode_all_nosd <- combine_posterior(mode_nosd,4000)
 
 
 mcmcse::multiESS(mode_all31[,1:4])
-plot_gradient(mode_all11,ylim = c(13,39))
+plot_gradient(mode_allold,ylim = c(13,39))
 plot_gradient(mode_allold,ylim = c(10,37), confint_col = rgb(0,.7,0.5,0.2),line_col = rgb(0,.7,0.5,1), add = T)
 plot_gradient(mode_all1,ylim = c(10,37), confint_col = rgb(0.7,0,0.5,0.2),line_col = rgb(0.7,0,0.5,1), add = T)
 

### 2022-08-11 R/subscripts/ModernClimate.R

@@ -115,7 +115,7 @@ abline(h=0)
 prior_fun <- list(  
   f1 = function(x,log) dsnorm(x,location = -3.03, scale = 12, alpha = 30, log = log), # prior on A (lower asymptote)
   f2 = function(x,lower,log) dtnorm(x, lower, upper = Inf, mean = 28.3, sd = 10, log = log), # prior on upper asymptote
-  f3 = function(x,log) dnorm(x, mean = 42.1, sd = 12, log = log), # prior on M
+  f3 = function(x,log) dnorm(x, mean = 42.1, sd = 10, log = log), # prior on M
   f4 = function(x,log) dgamma(x, shape = 3.2, rate = 20, log = log)# prior on Q
 )
 

### 2022-08-11 R/subscripts/models/EECO_climate_model.R

@@ -17,7 +17,7 @@ hierarchical_model <- function(n_iter = 1000, n_thin = 1, obsmat = NULL,
                                yest_inits = NULL, sdyest_inits = NULL,
                      proposal_var_inits = NULL, adapt_sd = NULL,
                      adapt_sd_decay = NULL,  start_adapt = NULL, 
-                     A_sdy = 1, B_sdy = 1,
+                     A_sdy = 1, B_sdy = 1, sdy_fixed = FALSE,
                      quiet = FALSE){
   
   ### Load functions
@@ -39,8 +39,8 @@ hierarchical_model <- function(n_iter = 1000, n_thin = 1, obsmat = NULL,
   coeff_inits[3] = rnorm(1,45,7.5)
   coeff_inits[4] = exp(rnorm(1,-2.3,0.25))
   }
-  if(is.null(sdy_init)) sdy_init = exp(rnorm(1,0.7,0.25))
-  
+  if(is.null(sdy_init) & sdy_fixed == FALSE ) sdy_init = exp(rnorm(1,0.7,0.25))
+  if(is.numeric(sdy_fixed)) sdy_init = sdy_fixed
   # deterministic setting of initial values for the temperature and temperature sd estimates
   if((!(is.null(distrmat)) | !(is.null(obsmat))) & is.null(yest_inits)) {
     yest_inits <- c(unlist(sapply(unique(obsmat$sample), function(x) mean(obsmat$temperature[which(obsmat$sample == x)]))),
@@ -249,9 +249,10 @@ hierarchical_model <- function(n_iter = 1000, n_thin = 1, obsmat = NULL,
     }
     
     ### 3. Gibbs step to estimate sdy
-    sdy[i] = sqrt(1/rgamma(1,
+    if(sdy_fixed == FALSE) {sdy[i] = sqrt(1/rgamma(1,
                            shape_sdy,
                            (B_sdy+0.5*sum((yestimate[i,]-pred)^2))))
+    } else sdy[i] = sdy[i-1]
     
     ## 4. Metropolis-Hastings step to estimate the regression coefficients
     # 4.0: create proposal innovations

### 2022-08-13 manuscript/manuscript.Rmd
Update manuscript.Rmd

@@
 ===============================================================================
-
-Here goes the abstract
+Geochemical data such as oxygen isotopes (d18O), tetraether indices (TEX86) and magnesium calcium ratios (Mg/Ca) are routinely used to reconstruct ocean temperatures in deep time.
+However, limited spatial coverage and disagreements between geochemical proxies compromise the accuracy of these reconstructions.
+Geological data such as coral reefs, mangroves, and evaporites have the potential to supplement geochemical data by improving spatial coverage and constraining temperature estimates.
+Historically, these data have been used to qualitatively inform upon palaeoclimatic conditions at broad spatial scales, yet no transfer functions exist to convert geological data into temperature estimates.
+Quantitative temperature reconstructions have therefore not made use of these data.
+Here, we present a Bayesian hierarchical model to integrate geological data–with established temperature proxies–into a unified quantitative framework.
+We apply this approach to the early Eocene climatic optimum (EECO), the interval with the warmest sustained temperatures of the Cenozoic.
+Assuming the conservation of thermal tolerances of modern coral reefs and mangrove taxa, we establish broad sea surface temperature ranges for EECO coral reef and mangrove sites.
+We integrate these temperature estimates with the EECO geochemical shallow marine proxy record to model the latitudinal temperature gradient and its uncertainty.
+Our results confirm the presence of a flattened latitudinal temperature gradient and unusually high polar temperatures during the EECO.
+We show that the inclusion of ecological data can substantially reduce the uncertainty on temperature estimates in climate zones lacking geochemical data.
 
@@
 
-The early Eocene climate optimum (EECO, ~ 53 - 49 million years ago) is thought to have been the warmest episode of [...].
+The early Eocene climate optimum (EECO, ~ 53--49 million years ago) is thought to have been the warmest episode of [...].
 The EECO is of particular interest due to its use as an analogue for extreme warming scenarios, depicting a possible future high-CO2 world.
@@
 
-Example reference [@R-base]
+Hook -> Generic problem:
+- Reconstructing past climate important for understanding current/future global warming
+- Two main pathways: highly complex ESM <--> proxy compilations
+- Problems with proxy data: only provides a local picture (few local sites interpreted as global signal)
+- Problems with ESM (long time to run, very complex, a lot of assumptions on processes/doesn't consider how they may change, need to be constrained with proxy data)
+- Lack of intermediate-complexity models to make full use of the proxy data
+
+Specific problem:
+- exemplified with the EECO: proxies and ESM hard to reconcile, EECO important climate analogue
+- a lot of proxies recrystallised <--> unreliable data or lack of data
+
+Solution:
+- Proxy-driven model (no strong assumptions on processes)
+- Integrate more data (quantify biological / geological information)
+- Integrate additional knowledge (constraints on such gradients)
+
+Application:
+- EECO time of extreme warming
+- problem of the Early Eocene:
+- apply solution to EECO to constrain EECO lat. T gradient
+
+Here:
+
+Here, we present a Bayesian hierarchical model to integrate geological data---with established temperature proxies---into a unified quantitative framework.
+We apply this approach to the early Eocene climatic optimum (EECO), the interval with the warmest sustained temperatures of the Cenozoic.
+Assuming the conservation of thermal tolerances of modern coral reefs and mangrove taxa, we establish broad sea surface temperature ranges for EECO coral reef and mangrove sites.
+We integrate these temperature estimates with the EECO geochemical shallow marine proxy record to model the latitudinal temperature gradient and its uncertainty.
+
+- EECO specific information
+- Disagreement between GCM + geochem data
 
@@
 
-## early Eocene SST proxies
-
-Proxy-based reconstructions of Cenozoic shallow water temperatures chose a range of proxies for marine, shallow water temperatures
-proxies: d18O, D47, Mg/Ca, TEX86
-mangroves, corals, sharks?
+## Geochemical data
 
@@
 **TEX86.**
+
+## Ecological data
 
@@
 The range of temperatures in which both taxa occur in the modern is very wide (20.8 and 28 deC), hence occurrences of both *Avicennia* and Rhizophoraceae pollen were not used in the analysis.
+
+[Fig. 1]
+
+## Bayesian framework
+- logistic regression
+- hierarchical model structure
+-
+
+[Fig. 2]
+## Model validation
+
+[Fig. 3]
+
+- Modern data
+- Sample based on Eocene localities
+
+## Application to the EECO
+
+[Fig. 4]
 

### 2022-08-13 manuscript/manuscript.Rmd
Update manuscript.Rmd

@@
 
-[Fig. 1]
+[Fig 1: Map and distribution of proxy data and ecological data]
 
@@
 - hierarchical model structure
--
+- choice of priors (based on modern, see below)
 
-[Fig. 2]
+[Fig. 2: Model structure and visualisation of parameters and priors]
+
 ## Model validation
@@
 [Fig. 3]
-
 - Modern data
@@
 
-## Application to the EECO
+## Model application
+- Calculate lat gradient
+- Calculate global T
 
@@
 
-[Fig 1: In this part we can show a figure with the proxy data and the distributions of temperature proxies we specify, like mangroves]
-
-## A hierarchical model of latitudinal temperature gradients ##
-
-[Fig 2: In this part we can show a figure with the model structure and the priors we put on the parameters. Also, try it on the modern data? Alternatively, this figure goes to the SI]
-
-- alternative to current implementation: prior on (A+DKA) instead of on DKA?
-Then constrain DKA to be positive by proposing it in the log-space?
-This would allow for putting a direct prior on the upper temperature limit
-
 # Results
 
-## Fig 3 - Eocene gradient
+## Model validation
+- Show that the modern gradient can be reproduced with Eocene sampling distribution
+[Fig 3 - Modern gradient]
+
+## EECO reconstruction
+- Show Eocene gradient 
+- Global t
+- report lat gradient
+
 ## Fig 4 - compare Eocene gradient with previous reconstructions / ES models

### 2022-08-16 manuscript/manuscript.Rmd
manuscript update

@@
 title: "The Early Eocene Climate Optimum reconstructed with Bayesian multi-proxy integration"
+"Improving palaeoclimatic reconstructions with ecological data: How hot was the early Eocene?"
+"Integrating ecological and geochemical proxies for palaeoclimatic reconstruction: An early Eocene example"
+
 author:
@@
 # Introduction
-
-The early Eocene climate optimum (EECO, ~ 53--49 million years ago) is thought to have been the warmest episode of [...].
-The EECO is of particular interest due to its use as an analogue for extreme warming scenarios, depicting a possible future high-CO2 world.
-To reconstruct the climate of EECO, a variety of proxy data and Earth System Models have been used.
-Early modelers have soon noticed that their models struggle to reproduce the climate indicated by proxy data: Without unrealistically high concentrations of atmospheric CO2, early climate models were unable to reproduce the equitable polar climates indicated by early Eocene proxy data, e.g.
-the occurrence of mangroves in polar latitudes, and d18O data from x indicates a temperature of ~ y at latitude z.
-
-Contrast between complex, process-driven Earth system models and very simplistic use of the proxy data as local temperature estimates or global averages.
-Here, we introduce a proxy-driven modelling approach of relatively low complexity, which places the available chemical and biological proxy data in a single climate model.
-In order to properly compare model output to proxy data, latitudinal temperature patterns need to be reconstructed from noisy data (impossible to properly compare model gradients with scattered proxy points "by eye")
+The geological record holds evidence for episodes in Earth history with fundamentally different climates than our current.
+Some episodes, for example the early Eocene, are thought to have been significantly warmer than the modern, potentially representing an analogue for extreme climate warming scenarios.
+Reconstructions of palaeoclimate commonly follow one of two approaches.
+Geochemical proxies such as oxygen isotopes (d18O), tetraether indices (TEX86) and magnesium calcium ratios (Mg/Ca) are used to derive palaeotemperature estimates for individual locations using experimentally derived transfer functions.
+Multiple local estimates are sometimes summarised to generate regional or global estimates, but this approach is prone to sampling bias [@Jones2022].
+Earth system models take a fundamentally different approach by modelling the climate-generating processes directly, reconstructing spatially resolved climate estimates for the entire glob.e This approach is, however, relying on proxy data to constrain the large number of possible model setups and ground truth the model output.
 
@@
 
+<!--
+
+The early Eocene climate optimum (EECO, ~ 53--49 million years ago) is thought to have been the warmest episode of [...].
+The EECO is of particular interest due to its use as an analogue for extreme warming scenarios, depicting a possible future high-CO2 world.
+To reconstruct the climate of EECO, a variety of proxy data and Earth System Models have been used.
+Early modelers have soon noticed that their models struggle to reproduce the climate indicated by proxy data: Without unrealistically high concentrations of atmospheric CO2, early climate models were unable to reproduce the equitable polar climates indicated by early Eocene proxy data, e.g.
+the occurrence of mangroves in polar latitudes, and d18O data from x indicates a temperature of ~ y at latitude z.
+
+Contrast between complex, process-driven Earth system models and very simplistic use of the proxy data as local temperature estimates or global averages.
+Here, we introduce a proxy-driven modelling approach of relatively low complexity, which places the available chemical and biological proxy data in a single climate model.
+In order to properly compare model output to proxy data, latitudinal temperature patterns need to be reconstructed from noisy data (impossible to properly compare model gradients with scattered proxy points "by eye")
+-->
 # Materials & Methods

### 2022-08-16 R/tests/seperate_gradients_north_south/function_tests_seperate_gradients.R

@@ -2,8 +2,21 @@ gradient1 <- function(lat,A,DKA,M,B) A + DKA/((1+(exp(B*(lat-M)))))
 gradient2 <- function(lat,A,DKA,M,B) A + DKA/(-(1+(exp(B*(lat-M)))))
 
 gradient_two_A <- function(lat,A1,A2,DKA,Q1,M1,Q2,M2) A1 + (DKA-A1)/((1+(exp(Q1*(lat-M1))))) + (DKA-A2)/(-(1+(exp(Q2*(lat-M2)))))
-gradient_one_A <- function(lat,A,DKA,Q1,M1,Q2,M2) A + (DKA-A)/((1+(exp(Q1*(lat-M1))))) + (DKA-A)/(-(1+(exp(Q2*(lat-M2)))))
+gradient_one_A <- function(lat,A,DKA,M1,M2,Q1,Q2) A + (DKA-A)/((1+(exp(Q1*(lat-M1))))) + (DKA-A)/(-(1+(exp(Q2*(lat-M2)))))
 gradient_one_AMQ <- function(lat,A,DKA,M,Q) A + (DKA-A)/((1+(exp(Q*(lat-M))))) + (DKA-A)/(-(1+(exp(Q*(lat-M)))))
+
+A = 5
+DKA = 20
+M1 = 45
+M2 = 30
+Q1 = 0.1
+Q2 = 0.033
+
+gradient_one_A <- function(lat,A,DKA,M1,M2,Q1,Q2) A + (2*DKA-A)/((1+(exp(Q1*(lat-M1))))+(1+(exp(Q2*(-lat-M2)))))
+grad1a <- gradient_one_A(lat,A,DKA,M1,M2,Q1,Q2)
+lat[which.max(grad1a)]
+#grad3 <- gradient_one_AMQ(lat,0,30,-45,0.1)
+plot(lat,grad1a,type = "l", ylim = c(-5,30),xlim = c(-90,90))
 
 DKA <- 30
 A1 <- 10
@@ -20,10 +33,11 @@ lat <- seq(-90,90,1)
 grad1 <- gradient1(lat,0,30,45,0.1)
 grad2 <- gradient2(lat,0,30,45,0.1)
 grad2a <- gradient_two_A(lat,10,-20,30,.1,35,.1,-55)
-grad1a <- gradient_one_A(lat,-10,30,.1,45,.1,-35)
+grad1a <- gradient_one_A(lat,0,30,45,-50,0.01,0.1)
 
[…]

### 2022-09-02 manuscript/manuscript.Rmd
model structure manuscript

@@
 title: "The Early Eocene Climate Optimum reconstructed with Bayesian multi-proxy integration"
-"Improving palaeoclimatic reconstructions with ecological data: How hot was the early Eocene?"
-"Integrating ecological and geochemical proxies for palaeoclimatic reconstruction: An early Eocene example"
 
@@
     reference_docx: "template.docx"
+
 bibliography: references.bib
 csl: palaeontology.csl
+
 ---
@@
 Modelling latitudinal temperature gradients from geological proxies: An early Eocene example
+"Improving palaeoclimatic reconstructions with ecological data: How hot was the early Eocene?"
+"Integrating ecological and geochemical proxies for palaeoclimatic reconstruction: An early Eocene example"
+
 -->
@@
 ## Bayesian framework
+
+**Model structure.** We model the mean temperature ($\mu$) at location $j$ as a function of absolute latitude ($abs(l)$) with a logistic regression (`growth curve` or `Richard's curve`) of the form:
+
+$$\mu_{j}  \sim N(\nu_j, \sigma), $$
+
+$$\nu_{j} = A~+\frac{K-A}{e^{B(abs(l_j)-M)}}~, ~~~~~ j = 1,...,n,  $$
+where $A$ and $K$ denote the lower and upper asymptote, respectively, $M$ specifies the latitude of maximal growth, the most quickly with latitude, $B$ denotes the growth rate (Fig.
+2), $\sigma$ denotes the residual standard deviation, and $n$ denotes the number of locations.
+
+We take a hierarchical approach and infer $\mu_j$ from $m$ individual temperature observations $t_{i = 1,...m}$ at location $j$ as
+
+$$t_{i,j}  \sim N(\mu_j, \sigma_j), ~~~~~ i = 1,...,m,  $$
+where $m$ is the number of observations at each location, and $\sigma_j$ is the estimated standard deviation of the temperatures at location $j$.
+
+This structure implies that $\mu_j$ is not fixed at the mean of the temperatures $t_{i=1,...,m,~j}$ of location $j$, but is drawn (`regularised`) towards the overall logistic regression curve.
+The pull towards the regression curve tends to be high when $m$ is low, when the observations $t_{i=1,...,m,~j}$ are scattered, and/or when the overall standard deviation $\sigma$ is low.
+In practice, this has the desirable consequence that locations with few observations and large temperature differences between observations have less influence on the regression curve than well-sampled locations with consistent temperature observations.
+
+**Priors.** In the Bayesian framework, priors need to be placed on the unknown parameters of a model.
+For $\sigma$ and $\sigma_{j=1,..n}$ we use weakly informative inverse-gamma priors with shape parameter $alpha$ = ...
+and scale parameter $beta$ = ...
+.
+Moderately changing these priors has little effect on the model outcome, which is desirable as we have little a-priori knowledge of these parameters.
+In contrast, we put informative priors on the regression coefficients $A$, $K$, $M$ and $B$, based on basic physical principles.
+Specifically, we put a skew-normal prior on the lower asymptote $A$, with ...
+.
+This constrains $A$ to not be $<< -2^\circ C$, which is the freezing point of sea water.
+We chose this distribution such that its mode is $~ 0^\circ C$, slightly higher than the modern...
+
 - logistic regression

### 2022-09-03 manuscript/manuscript.Rmd
manuscript methods: priors. small update to gradient: sdy = 0 (default)

@@
 The pull towards the regression curve tends to be high when $m$ is low, when the observations $t_{i=1,...,m,~j}$ are scattered, and/or when the overall standard deviation $\sigma$ is low.
-In practice, this has the desirable consequence that locations with few observations and large temperature differences between observations have less influence on the regression curve than well-sampled locations with consistent temperature observations.
+In practice, this has the desirable consequence that locations with few observations and large temperature differences between observations have less influence on the overall regression than well-sampled locations with consistent temperature observations.
 
-**Priors.** In the Bayesian framework, priors need to be placed on the unknown parameters of a model.
-For $\sigma$ and $\sigma_{j=1,..n}$ we use weakly informative inverse-gamma priors with shape parameter $alpha$ = ...
-and scale parameter $beta$ = ...
-.
-Moderately changing these priors has little effect on the model outcome, which is desirable as we have little a-priori knowledge of these parameters.
-In contrast, we put informative priors on the regression coefficients $A$, $K$, $M$ and $B$, based on basic physical principles.
-Specifically, we put a skew-normal prior on the lower asymptote $A$, with ...
+**Priors.** In the Bayesian framework, priors need to be placed on the unknown parameters of a model. We placed weakly informative, conjugate inverse-gamma priors on $\sigma$ and $\sigma_{j=1,..n}$:
+$$\sigma  \sim \sqrt{Inv{\text -}Gamma(\alpha+\frac{n}{2},\beta + 0.5\times(\mu_j-\nu_j))}, ~~~~~ j = 1,...,n,  $$
+$$\sigma_j  \sim \sqrt{Inv{\text -}Gamma(\alpha+\frac{m}{2},\beta + 0.5\times(t_{i,j}-\mu_j))}, ~~~~~ i = 1,...,m,~~~~~j = 1,...,n. $$
+We set $\alpha = 1$ and $\beta = 1$, allowing these priors to be quickly overwhelmed by the data as $n$ and $m$ increase, which is desirable as we have little a-priori knowledge of these parameters.
+
+In contrast, we put informative priors on the regression coefficients $A$, $K$, $M$ and $B$, based on physical principles and knowledge from the modern climate system.
+Specifically, we used the priors to express the following constraints and prior knowledge:
+1) Predicted seawater surface temperatures is not allowed to be $<< -2^\circ C$, the freezing point of sea water.
+2) Input of solar energy decreases from the tropics to the poles.
+Hence, the latitudinal temperature gradient is generally negative, i.e.
+temperature decreases with absolute latitude.
+This is achieved by constraining $K-A$ to be positive.
+3) The steepness of the gradient is likely highest in mid-latitudes.
+4) The 
+a skew-normal prior on the lower asymptote $A$, with ...
 .

### 2022-09-03 R/functions/model_components/gradient.R

@@ -1,6 +1,6 @@
 # logistic function to generate latitudinal temperature gradients based on four parameters
-
-gradient <- function(x, coeff, sdy) { # parametrised with difference between cold and hot
+# accepts a vector of length 4 or a dataframe with four columns
+gradient <- function(x, coeff, sdy = 0) { # parametrised with difference between cold and hot
   if(is.list(coeff) & !(is.data.frame(coeff) | is.matrix(coeff))) coeff = unlist(coeff)
   if(is.data.frame(coeff) | is.matrix(coeff)) {
     A = coeff[,1]

### 2022-09-03 manuscript/manuscript.Rmd
Update manuscript.Rmd

@@
 
-In contrast, we put informative priors on the regression coefficients $A$, $K$, $M$ and $B$, based on physical principles and knowledge from the modern climate system.
-Specifically, we used the priors to express the following constraints and prior knowledge:
+In contrast, we put informative priors on the regression coefficients $A$, $K$, $M$ and $B$, based on physical principles and knowledge from the modern climate system:
 1) Predicted seawater surface temperatures is not allowed to be $<< -2^\circ C$, the freezing point of sea water.

### 2022-09-03 manuscript/manuscript.Rmd
methods update

@@
 
-We take a hierarchical approach and infer $\mu_j$ from $m$ individual temperature observations $t_{i = 1,...m}$ at location $j$ as
+We infer $\mu_j$ from $m$ individual temperature observations $t_{i = 1,...m}$, derived from geochemical data, at location $j$ as
 
@@
 
-This structure implies that $\mu_j$ is not fixed at the mean of the temperatures $t_{i=1,...,m,~j}$ of location $j$, but is drawn (`regularised`) towards the overall logistic regression curve.
+Similarly, $\mu_j$ is inferred for locations with ecological proxies from the associated normal temperature distributions with a given mean and standard deviation, $mean_j$ and $sd_j$, as
+
+$$mean_j  \sim N(\, \mu_j,sd_j).  $$
+
+This structure implies that $\mu_j$ is not fixed at the mean proxy temperature at location $j$, <!--of the temperatures $t_{i=1,...,m,~j}$ of location $j$-->, but is drawn (`regularised`) towards the overall logistic regression curve.
 The pull towards the regression curve tends to be high when $m$ is low, when the observations $t_{i=1,...,m,~j}$ are scattered, and/or when the overall standard deviation $\sigma$ is low.

### 2022-09-05 manuscript/manuscript.Rmd
methods

@@
 
-**Model structure.** We model the mean temperature ($\mu$) at location $j$ as a function of absolute latitude ($abs(l)$) with a logistic regression (`growth curve` or `Richard's curve`) of the form:
+**Model structure (Fig 2).** We model the mean temperature ($\mu$) at location $j$ as a function of absolute latitude ($abs(l)$) with a logistic regression (`growth curve` or `Richard's curve`) of the form:
 
@@
 
-Similarly, $\mu_j$ is inferred for locations with ecological proxies from the associated normal temperature distributions with a given mean and standard deviation, $mean_j$ and $sd_j$, as
+Similarly, $\mu_j$ is inferred for locations with ecological proxies from the associated normal temperature distributions with a given mean and standard deviation, $t_{\mu,j}$ and $t_{\sigma,j}$, as
 
-$$mean_j  \sim N(\, \mu_j,sd_j).  $$
+$$t_{\mu,j}  \sim N(\mu_j,t_{\sigma,j}).  $$
 
-This structure implies that $\mu_j$ is not fixed at the mean proxy temperature at location $j$, <!--of the temperatures $t_{i=1,...,m,~j}$ of location $j$-->, but is drawn (`regularised`) towards the overall logistic regression curve.
-The pull towards the regression curve tends to be high when $m$ is low, when the observations $t_{i=1,...,m,~j}$ are scattered, and/or when the overall standard deviation $\sigma$ is low.
+This structure implies that $\mu_j$ is not fixed at the mean proxy temperature at location $j$, <!--of the temperatures $t_{i=1,...,m,~j}$ of location $j$-->, but is drawn towards the overall logistic regression curve, i.e.
+towards $\nu_j$.
+The pull towards $\nu_j$ tends to be high when $m$ is low, when the observations $t_{i=1,...,m,~j}$ are scattered, i.e.
+$\sigma_j$ is high, and/or when the overall standard deviation $\sigma$ is low.
 In practice, this has the desirable consequence that locations with few observations and large temperature differences between observations have less influence on the overall regression than well-sampled locations with consistent temperature observations.
 
-**Priors.** In the Bayesian framework, priors need to be placed on the unknown parameters of a model. We placed weakly informative, conjugate inverse-gamma priors on $\sigma$ and $\sigma_{j=1,..n}$:
-$$\sigma  \sim \sqrt{Inv{\text -}Gamma(\alpha+\frac{n}{2},\beta + 0.5\times(\mu_j-\nu_j))}, ~~~~~ j = 1,...,n,  $$
-$$\sigma_j  \sim \sqrt{Inv{\text -}Gamma(\alpha+\frac{m}{2},\beta + 0.5\times(t_{i,j}-\mu_j))}, ~~~~~ i = 1,...,m,~~~~~j = 1,...,n. $$
-We set $\alpha = 1$ and $\beta = 1$, allowing these priors to be quickly overwhelmed by the data as $n$ and $m$ increase, which is desirable as we have little a-priori knowledge of these parameters.
+**Priors (Fig2x).** In the Bayesian framework, priors need to be placed on the unknown parameters of a model.
+We placed weakly informative, conjugate inverse-gamma priors on $\sigma$ and $\sigma_{j=1,..n}$:
+$$\sigma  \sim \sqrt{Inv-Gamma(\alpha+\frac{n}{2},\beta + 0.5\times(\mu_j-\nu_j))}, ~~~~~ j = 1,...,n,  $$
+$$\sigma_j  \sim \sqrt{Inv-Gamma(\alpha+\frac{m}{2},\beta + 0.5\times(t_{i,j}-\mu_j))}, ~~~~~ i = 1,...,m,~~~~~j = 1,...,n. $$
+We set $\alpha = \beta = 1$, allowing these priors to be quickly overwhelmed by the data as $n$ and $m$ increase, as we have little *a priori* knowledge of these parameters.
 
-In contrast, we put informative priors on the regression coefficients $A$, $K$, $M$ and $B$, based on physical principles and knowledge from the modern climate system:
-1) Predicted seawater surface temperatures is not allowed to be $<< -2^\circ C$, the freezing point of sea water.
-2) Input of solar energy decreases from the tropics to the poles.
-Hence, the latitudinal temperature gradient is generally negative, i.e.
+In contrast, we put informative priors on the regression coefficients $A$, $K$, $M$ and $B$, based on physical principles, and vaguely based on the modern climate system:
+
+**A.** Predicted seawater surface temperatures are not allowed to be $<< -2^\circ C$, the freezing point of sea water.
+The highest prior density of $A$ is placed around $0^\circ C$, and it slowly tapers off towards higher temperatures.
+This shape is achieved by placing a skew-normal prior on the lower asymptote, specified as
+$$A \sim SN(\xi=-3.0,\omega=12,\alpha_{SN}=30),$$
+where $\xi$, $\omega$, and $\alpha_{SN}$ are the location, scale and shape parameters.
+
+**K.** Input of solar energy decreases from the tropics to the poles.
+Hence, the latitudinal temperature gradient is broadly negative, i.e.
 temperature decreases with absolute latitude.
-This is achieved by constraining $K-A$ to be positive.
-3) The steepness of the gradient is likely highest in mid-latitudes.
-4) The 
-a skew-normal prior on the lower asymptote $A$, with ...
-.
-This constrains $A$ to not be $<< -2^\circ C$, which is the freezing point of sea water.
-We chose this distribution such that its mode is $~ 0^\circ C$, slightly higher than the modern...
+This is achieved by setting $K \ge A$.
+The prior on the upper asymptote $K$ is a truncated normal distribution with the mean set to $K$ of the modern SST gradient, with a broad standard deviation:
+$$K \sim TN(\mu_{TN} = 28, \sigma_{TN} = 10, \alpha_{TN} = A, \beta_{TN}= \infty)$$
+The distribution is truncated to the left at $\alpha_{TN}=A$, but not truncated to the right ($\beta_{TN}$).
 
-- logistic regression
-- hierarchical model structure
-- choice of priors (based on modern, see below)
+**M.** The steepness of the gradient is presumed to be highest in mid-latitudes; this is expressed with a normal prior on $M$ with the mean set to $M$ of the modern SST gradient: 
+$$M \sim N(42,10)$$
+**B.** The steepness or growth rate $B$ of the gradient is constrained to be $\ge0$ and to not be exceedingly high, as oceanic and atmospheric heat transfer is bound to prevent very abrupt SST changes across latitudes.
+A gamma-distributed prior of the form
+$$B \sim Gamma(\alpha_G=4.3,\beta_G=30)$$
+was placed on $B$. The shape and rate parameters $\alpha_G$ and $\beta_G$ were chosen such that the highest prior density is at $B$ of the modern SST gradient, $0.11$.
 

### 2022-09-05 R/options.R

@@ -14,9 +14,9 @@ parallel = TRUE # should the chains run in parallel (using doParallel and foreac
 # 
 # Define the priors
 priors <- list(  # a list with one function per parameter of the logistic regression
-  f1 = function(x,log) dsnorm(x,location = -3.03, scale = 12, alpha = 30, log = log), # prior on A (lower asymptote)
-  f2 = function(x,lower,log) dtnorm(x, lower, upper = Inf, mean = 28.3, sd = 10, log = log), # prior on K (upper asymptote)
-  f3 = function(x,log) dnorm(x, mean = 42.1, sd = 10, log = log), # prior on M (latitude of steepest point of the gradient)
+  f1 = function(x,log) dsnorm(x,location = -3.0, scale = 12, alpha = 30, log = log), # prior on A (lower asymptote)
+  f2 = function(x,lower,log) dtnorm(x, lower, upper = Inf, mean = 28, sd = 10, log = log), # prior on K (upper asymptote)
+  f3 = function(x,log) dnorm(x, mean = 42, sd = 10, log = log), # prior on M (latitude of steepest point of the gradient)
   f4 = function(x,log) dgamma(x, shape = 4.3, rate = 30, log = log)# prior on B (growth rate)
 )
 #

### 2022-09-27 manuscript/manuscript.Rmd
title change and model assessment fix

@@
 ---
-title: "The Early Eocene Climate Optimum reconstructed with Bayesian multi-proxy integration"
+title: "An Early Eocene temperature gradient modelled with geological and ecological proxies"
 
@@
 "Integrating ecological and geochemical proxies for palaeoclimatic reconstruction: An early Eocene example"
-
+"The Early Eocene Climate Optimum reconstructed with Bayesian multi-proxy integration"
 -->

### 2022-10-07 manuscript/manuscript.Rmd
test commit

@@
 
-**Coral reefs.** Coral reefs built by zooxanthellate-bearing, shallow water corals are confined to the tropics - substropics. The coldest
+**Coral reefs.** Today, shallow warm-water coral reefs are restricted to tropical-to-subtropical latitudes.
+Coral reefs built by zooxanthellate-bearing, shallow water corals are confined to the tropics - substropics. The coldest
 
@@
 ## Model validation
+- Show that the 4-parameter logistic function can adequatly describe the modern temperature gradient
 - Show that the modern gradient can be reproduced with Eocene sampling distribution
 [Fig 3 - Modern gradient]
+
+The average modern temperature gradient can be closely approximated with our model when using the full modern SST dataset (Fig.
+3a).
+Degrading the amount and distribution of data points to match the early Eocene sampling distribution still results in a close match of the modelled gradient with the empirical gradient (Fig 3b).
+A noticable, but minor offset is apparent in high latitudes, where sampling is most incomplete (Fig 3b).
 

### 2022-10-07 manuscript/manuscript.Rmd
new, longer model run, manuscript edits

--- 2022-09-27 manuscript/manuscript.Rmd
@@
 
-**Coral reefs.** Coral reefs built by zooxanthellate-bearing, shallow water corals are confined to the tropics - substropics. The coldest
+**Coral reefs.** Today, shallow warm-water coral reefs are limited to tropical and subtropical latitudes (~34°N–32°S), with minimum sea surface temperature (~18 °C) tolerances being the primary constraint on this distribution [@johannes1983; @kleypas1999].
+As coral reefs reside at the upper thermal limit of the oceans today, their maximum sea surface temperature tolerance is less well-constrained.
+However,
 
@@
 ## Model validation
+- Show that the 4-parameter logistic function can adequatly describe the modern temperature gradient
 - Show that the modern gradient can be reproduced with Eocene sampling distribution
 [Fig 3 - Modern gradient]
-
+The average modern temperature gradient can be closely approximated with our model when using the full modern SST dataset (Fig.
+3a).
+Degrading the amount and distribution of data points to match the early Eocene sampling distribution still results in a close match of the modelled gradient with the empirical gradient (Fig 3b).
+A noticable, but minor offset is apparent in high latitudes, where sampling is most incomplete (Fig 3b).
 ## EECO reconstruction

### 2022-10-07 manuscript/manuscript.Rmd

@@
 
-**Coral reefs.** Today, shallow warm-water coral reefs are limited to tropical and subtropical latitudes (~34°N–32°S), with minimum sea surface temperature (~18 °C) tolerances being the primary constraint on this distribution [@johannes1983; @kleypas1999].
-As coral reefs reside at the upper thermal limit of the oceans today, their maximum sea surface temperature tolerance is less well-constrained.
-However,
+**Coral reefs.** Today, shallow warm-water coral reefs are limited to tropical and subtropical latitudes (~34°N–32°S), with minimum sea surface temperature (~18 °C) tolerances being the primary constraint on this distribution [@johannes1983; @kleypas1999; @yamano2001].
+However, as coral reefs reside at the upper thermal limit of the oceans today, their maximum sea surface temperature tolerance is less well-constrained with some studies suggesting up to 35.6 degC in the geological past [@jones2022reef].
+During the Eocene, coral communities and reefs occupied
 
@@
 
+## Parameter estimation
+We estimated the posterior distributions of the model parameters using a custom-made Markov chain Monte Carlo (MCMC) algorithm, written in R.
+Specifically, we sampled the unknown parameters $A$, $K$, $M$ and $B$ using a Metropolis-Hastings step, and used Gibbs sampling to estimate all other unknown parameters.
+Posterior inference on the modern gradient is based on four chains with 60,000 iterations each, 10,000 of which were discarded as burn-in, and keeping every 10th iteration, resulting in a total of 20,000 iterations.
+For the Eocene model, we increased these numbers by a factor of 10, i.e.
+running four chains with 600,000 iterations each, keeping every 100th iteration, as the hierarchical model structure results in higher autocorrelation of the chains.
+
 # Results

### 2022-10-07 manuscript/manuscript.Rmd
manuscript and ref updates, adjusting iteration number of EECO run

@@
 However, as coral reefs reside at the upper thermal limit of the oceans today, their maximum sea surface temperature tolerance is less well-constrained with some studies suggesting up to 35.6 degC in the geological past [@jones2022reef].
-During the Eocene, coral communities and reefs occupied
+During the Eocene, coral communities and reefs occupied expanded
 
@@
 Posterior inference on the modern gradient is based on four chains with 60,000 iterations each, 10,000 of which were discarded as burn-in, and keeping every 10th iteration, resulting in a total of 20,000 iterations.
-For the Eocene model, we increased these numbers by a factor of 10, i.e.
-running four chains with 600,000 iterations each, keeping every 100th iteration, as the hierarchical model structure results in higher autocorrelation of the chains.
+The modern temperature gradient with the Eocene sampling distribution was estimated in one chain with 25,000 iterations for each of the 100 temperature samples.
+5,000 iterations each were discarded as burn-in, and every 10th iteration was kept, resulting in a total of 200,000 iterations across all 100 model runs.
+For the Eocene model, we ran four chains with 600,000 iterations each, discarding 100,000 as burn-in and keeping every 100th iteration, as the hierarchical model structure results in higher autocorrelation of the chains.
+The Eocene posterior inference is thus based on a total of 20,000 iterations with low autocorrelation (effective multivariate sample size for $A$, $K$, $M$ and $B$ is > 18,000).
 

### 2022-10-07 manuscript/manuscript.Rmd
Eocene model run with new rotations

@@
 However, as coral reefs reside at the upper thermal limit of the oceans today, their maximum sea surface temperature tolerance is less well-constrained with some studies suggesting up to 35.6 degC in the geological past [@jones2022reef].
-During the Eocene, coral communities and reefs occupied expanded
+During the Eocene, coral communities and reefs persisted outside of the tropics, with some communities expanding up to 43 degC N [@zamagni2012]
 
@@
 We estimated the posterior distributions of the model parameters using a custom-made Markov chain Monte Carlo (MCMC) algorithm, written in R.
-Specifically, we sampled the unknown parameters $A$, $K$, $M$ and $B$ using a Metropolis-Hastings step, and used Gibbs sampling to estimate all other unknown parameters.
+Specifically, we sampled the unknown parameters $A$, $K$, $M$ and $B$ using Metropolis-Hastings, and used Gibbs sampling to estimate all other unknown parameters.
 Posterior inference on the modern gradient is based on four chains with 60,000 iterations each, 10,000 of which were discarded as burn-in, and keeping every 10th iteration, resulting in a total of 20,000 iterations.

### 2022-10-07 manuscript/manuscript.Rmd
data methods and eeco results

@@
 
-**d18O.**
+Geochemical climate proxy data were extracted from a latest Paleocene and early Eocene compilation [@hollis2019].
+For our analyses, this dataset was restricted to the EECO and contains four commonly used geochemical climate proxies: d180, d47, Mg/Ca and TEX86 (REFS).
+
+Do we need a description of each?
+
+**d18O.** Measurements of the ratio of stable isotope 18O and 16O from fossils provide
 
@@
 **Coral reefs.** Today, shallow warm-water coral reefs are limited to tropical and subtropical latitudes (~34°N–32°S), with minimum sea surface temperature (~18 °C) tolerances being the primary constraint on this distribution [@johannes1983; @kleypas1999; @yamano2001].
-However, as coral reefs reside at the upper thermal limit of the oceans today, their maximum sea surface temperature tolerance is less well-constrained with some studies suggesting up to 35.6 degC in the geological past [@jones2022reef].
-During the Eocene, coral communities and reefs persisted outside of the tropics, with some communities expanding up to 43 degC N [@zamagni2012]
+As coral reefs reside at the upper thermal limit of the oceans today, their maximum sea surface temperature tolerance is less well-constrained with some studies suggesting up to 35.6 degC in the geological past [@jones2022reef].
+Nevertheless, the distribution of coral reefs have frequently been recognised as tracers of past (sub-)tropical conditions [@ziegler1984; @kiessling2001].
+During the Eocene, coral communities and reefs expanded across tropical and temperate latitudes, with some communities found up to palaeolatitudes of 43 degC N [@zamagni2012].
+Using a compilation of Paleocene--early Eocene coral reefs and community localities [@zamagni2012], we generate quantitative sea surface temperature estimates for the ECCO.
+To do so, we extract localities inferred to be Ilerdian (early Eocene) coral reefs from the compilation, excluding coral knobs and coral-bearing mounds which might have broader climatic limits than coral reef ecosystems.
+This filtering resulted in four unique coral reef localities remaining for the early Eocene.
+Subsequently, we use statistically derived temperature limits (minimum = 21 degC, average = 27.6 degC, maximum = 29.5 degC) from the published literature [@kleypas1999] to define a normal probability distribution of potential temperature values for coral reef localities.
+This normal probability distribution was defined with a mean of 27.6 and a standard deviation of 2.125, placing 95% of the probability density within the minimum and maximum interval.
+[this is not quite accurate, as the mean is not exactly between 21 and 29.5, but higher.
+There is only 81 % probabil ]
 
-**Mangroves.** Recent mangroves occur in the tropics - subtropics.
+**Mangroves.** Mangroves are distributed throughout the tropics and subtropics today.
 Although many factors besides temperature determine the extent of mangrove distributions, empirical temperature limits have been established for the genera *Avicennia* and *Rhizophora*, with lower mean annual SST limits of 15.6 degC and 20.8 degC, respectively [@Quisthoudt2012].
@@
 Assuming that Eocene members of these mangrove taxa had similar climatic requirements as their modern relatives, the presence and absence of *Avicennia* and Rhizophoraceae pollen can be used as a temperature proxy.
-To translate mangrove occurrences from the compilation by @Popescu2021 to a quantitative temperature proxy, we allocate locations at which *Avicennia* pollen, but no Rhizophoraceae pollen were found, a mean temperature between the lower *Avicennia* and Rhizophoraceae limit (18.2 degC), with a standard deviation of 1.33 degC, thus placing 95 % of the probability density within that interval (see Fig.
+To translate mangrove occurrences from the compilation by @Popescu2021 to a quantitative temperature proxy, we allocate locations at which *Avicennia* pollen, but no Rhizophoraceae pollen were found, a mean temperature of 19.05 degC, between the lower *Avicennia* limit (15.6 degC) and the temperature at which *Rhizophora* commonly occurs (22.5 degC), deliberately wider than indicated by the lower *Rhizophora* limit.
+We set the standard deviation to 1.725 such that 95 % of the probability density are within 15.6 and 22.5 degC, placing 5% probability on the possibility that an Eocene *Avicennia*-only mangrove record occurred outside of this range.
+We define a 95% probability density for locations with *Avicennia* and Rhizophoraceae in a similar way.
+To allow for greater uncertainty at the lower temperature limit, we use the empirical 20.7 degC limit.
+For the upper limit, we use the maximal coral reef temperature of 29.5 degC from @kleypas1999, as the compendium of @Quisthoudt2012 does not include the warmest, equatorial regions.
+The mean and standard deviation of locations with *Avicennia* and Rhizophoraceae are thus set to 25.1 and 2.2, respectively.
+
+ Rhizophoraceae commonlylimit (19.05 degC), with a standard deviation of 1.725 degC, thus placing 95 % of the probability density within that interval (see Fig.
 1x).
-The range of temperatures in which both taxa occur in the modern is very wide (20.8 and 28 deC), hence occurrences of both *Avicennia* and Rhizophoraceae pollen were not used in the analysis.
+The range of temperatures in which both taxa occur in the modern besides the outlier at 20.8 degC is upwards of 22.5 degC [@Quisthoudt2012].
+As the study of @Quisthoudt2012 has not sufficiently represented equatorial regions, we use the upper coral temperature limit of 29.5 degC and set the 95 % probability interval for mangroves to 22.5 - 29.5  hence occurrences of both *Avicennia* and Rhizophoraceae pollen were not used in the analysis.
 
 [Fig 1: Map and distribution of proxy data and ecological data]
+
+## Palaeogeographic reconstruction
+
+The palaeogeographic distribution of geochemical and ecological data was reconstructed using the @merdith2021 plate rotation model via the palaeoverse R package ver.
+XXX (REF).
+The midpoint age of the EECO (51.5 Ma) coupled with the present-day coordinates of geochemical and ecological data.
 
@@
 A noticable, but minor offset is apparent in high latitudes, where sampling is most incomplete (Fig 3b).
+
 ## EECO reconstruction
@@
 - report lat gradient
+
+The modelled Eocene temperature gradient is starkly different from the modern (Fig 4).
+Modelled, median equatorial temperatures are 4.2 (95% CI: 0.2 - 8.3) degC higher for the Eocene, and polar temperatures are higher by 25.0 (95% CI: 17.0 - 29.1) degC.
+This results in a strongly flattened latitudinal temperature gradient of 9.0  (95% CI: 2.5 - 17.8) degC for the Eocene, as opposed to 29.6 degC for the modern.
+The high variability of early Eocene palaeotemperature proxies, particularly in the mid-latitudes, and the scarcity of high-latitude data, results in substantial uncertainties in the modelled temperature gradient, but ...
+
+The global Eocene temperatures
 

### 2022-10-07 manuscript/manuscript.Rmd
mangroves, corals, methods

@@
 Geochemical climate proxy data were extracted from a latest Paleocene and early Eocene compilation [@hollis2019].
-For our analyses, this dataset was restricted to the EECO and contains four commonly used geochemical climate proxies: d180, d47, Mg/Ca and TEX86 (REFS).
+This compilation provides  For our analyses, this dataset was restricted to the EECO and contains four commonly used geochemical climate proxies: d180, d47, Mg/Ca and TEX86 (REFS).
+For a detailed description of each see [@hollis2019].
 
-Do we need a description of each?
+- palaeoenvironment of data (which filters were applied)
+- recrystallisation (excluding most d18O)
 
-**d18O.** Measurements of the ratio of stable isotope 18O and 16O from fossils provide
-
-**d47.**
-
-**Mg/Ca.**
-
-**TEX86.**
+Do we need a description of each? Not in detail, 1 paragraph in total is enough
+ Measurements of the ratio of stable isotope 18O and 16O from fossils provide
 
@@
 Using a compilation of Paleocene--early Eocene coral reefs and community localities [@zamagni2012], we generate quantitative sea surface temperature estimates for the ECCO.
-To do so, we extract localities inferred to be Ilerdian (early Eocene) coral reefs from the compilation, excluding coral knobs and coral-bearing mounds which might have broader climatic limits than coral reef ecosystems.
-This filtering resulted in four unique coral reef localities remaining for the early Eocene.
+To do so, we extract localities inferred to be Ilerdian (early Eocene) coral reefs from the compilation, and that can be confidently assigned to the EECO.
+We exclude coral knobs and coral-bearing mounds which might have broader climatic limits than coral reef ecosystems.
+This filtering resulted in four unique coral reef localities remaining for the EECO, all of which conform to the modern latitudinal range of coral reefs (<34 deg N).
 Subsequently, we use statistically derived temperature limits (minimum = 21 degC, average = 27.6 degC, maximum = 29.5 degC) from the published literature [@kleypas1999] to define a normal probability distribution of potential temperature values for coral reef localities.
-This normal probability distribution was defined with a mean of 27.6 and a standard deviation of 2.125, placing 95% of the probability density within the minimum and maximum interval.
-[this is not quite accurate, as the mean is not exactly between 21 and 29.5, but higher.
-There is only 81 % probabil ]
+This normal probability distribution was defined with a mean of 27.6 and a standard deviation of 2.125, placing 97.5% of the probability density above the minimum.
+As the distribution of modern corals is skewed towards warmer temperatures, this approach results in 16.5% of the probability being placed on temperatures > 29.5 degC, allowing for the possibilty that Eocene coral reefs were adapted to warmer conditions than modern-day coral reefs.
 
 **Mangroves.** Mangroves are distributed throughout the tropics and subtropics today.
-Although many factors besides temperature determine the extent of mangrove distributions, empirical temperature limits have been established for the genera *Avicennia* and *Rhizophora*, with lower mean annual SST limits of 15.6 degC and 20.8 degC, respectively [@Quisthoudt2012].
-Both *Avicennia* and members of the Rhizophoraceae family were widespread in the early Eocene, but only *Avicennia* occurred in polar latitudes [@Suan2017; @Popescu2021].
-Assuming that Eocene members of these mangrove taxa had similar climatic requirements as their modern relatives, the presence and absence of *Avicennia* and Rhizophoraceae pollen can be used as a temperature proxy.
+While factors beside SST influence the distribution of mangroves, empirical, lower temperature limits have been established for the genera *Avicennia* (15.6 degC) and *Rhizophora* (20.7 degC) [@Quisthoudt2012].
+Both *Avicennia* and members of the Rhizophoraceae family were widespread in the early Eocene, but only *Avicennia* occurred at polar latitudes [@Suan2017; @Popescu2021].
+Assuming that Eocene members of these mangrove taxa conform to the same climatic requirements as their modern relatives, the presence and absence of *Avicennia* and Rhizophoraceae pollen can be used as a palaeotemperature indicator.
+For this analysis, published mangrove occurrence data were taken from @Popescu2021, and converted to quantitative temperature estimates.
+From this data, we identify two types of pollen assemblages which we ascribe different temperature distributions:
+
+1) *Avicennia*-only assemblages: the absence of Rhizophoraceae is indicative of temperatures being between 15.6 degC (lower temperature limt of *Avicennia*) and 20.7 (lower temperature limit of *Rhizophora*).
+However, a value of 22.5 degC is ascribed as the upper temperature limit here as *Rhizophora* is rare below this temperature.
+We define the *Avicennia*-only temperature distribution as a normal distribution with a mean of 19.05 degC and a standard deviation of 1.725, resulting in 95% of the probability density being placed within the temperature limits.
+
+2) *Avicennia* and Rhizophoraceae assemblages: the presence of both groups suggests that the locality should have a minimum temperature of 20.7 degC (lower temperature limit of *Rhizophora*).
+As the upper thermal limits of *Aviciennia* and *Rhizophora* are not well established in @Quisthoudt2012, we assign the same maximum temperature limits (29.5 degC) as coral reef localities as mangroves are also widely distributed throughout tropical regions.
+Consequently, we define the temperature distribution for this locality as a normal distribution with a mean of 25.1 and a standard deviation of 2.2, with 95% probability density within the temperature limits.
+
+ Mangrove localities containing *Avicennia* pollen, but no Rhizophoraceae pollen, a mean temperature of 19.05 degC was ascribed.
+The absence of Rhizophoraceae indicates that sea surface temperature values
+
+This value is between *Avicennia*'s lower temperature limit (15.6 degC) and the temperature at which *Rhizophora* commonly occurs (22.5 degC), deliberately wider than indicated by the lower *Rhizophora* limit.
+
 To translate mangrove occurrences from the compilation by @Popescu2021 to a quantitative temperature proxy, we allocate locations at which *Avicennia* pollen, but no Rhizophoraceae pollen were found, a mean temperature of 19.05 degC, between the lower *Avicennia* limit (15.6 degC) and the temperature at which *Rhizophora* commonly occurs (22.5 degC), deliberately wider than indicated by the lower *Rhizophora* limit.
@@
 
- Rhizophoraceae commonlylimit (19.05 degC), with a standard deviation of 1.725 degC, thus placing 95 % of the probability density within that interval (see Fig.
-1x).
-The range of temperatures in which both taxa occur in the modern besides the outlier at 20.8 degC is upwards of 22.5 degC [@Quisthoudt2012].
-As the study of @Quisthoudt2012 has not sufficiently represented equatorial regions, we use the upper coral temperature limit of 29.5 degC and set the 95 % probability interval for mangroves to 22.5 - 29.5  hence occurrences of both *Avicennia* and Rhizophoraceae pollen were not used in the analysis.
-
 [Fig 1: Map and distribution of proxy data and ecological data]
@@
 XXX (REF).
-The midpoint age of the EECO (51.5 Ma) coupled with the present-day coordinates of geochemical and ecological data.
+The midpoint age of the EECO (51.5 Ma), along with the present-day coordinates of geochemical and ecological data, were used for palaeorotation.
 

### 2022-10-07 manuscript/manuscript.Rmd
manuscript

@@
 This compilation provides  For our analyses, this dataset was restricted to the EECO and contains four commonly used geochemical climate proxies: d180, d47, Mg/Ca and TEX86 (REFS).
-For a detailed description of each see [@hollis2019].
+For a detailed description of each proxy see [@hollis2019].
 
@@
 
- Mangrove localities containing *Avicennia* pollen, but no Rhizophoraceae pollen, a mean temperature of 19.05 degC was ascribed.
-The absence of Rhizophoraceae indicates that sea surface temperature values
-
-This value is between *Avicennia*'s lower temperature limit (15.6 degC) and the temperature at which *Rhizophora* commonly occurs (22.5 degC), deliberately wider than indicated by the lower *Rhizophora* limit.
-
-To translate mangrove occurrences from the compilation by @Popescu2021 to a quantitative temperature proxy, we allocate locations at which *Avicennia* pollen, but no Rhizophoraceae pollen were found, a mean temperature of 19.05 degC, between the lower *Avicennia* limit (15.6 degC) and the temperature at which *Rhizophora* commonly occurs (22.5 degC), deliberately wider than indicated by the lower *Rhizophora* limit.
-We set the standard deviation to 1.725 such that 95 % of the probability density are within 15.6 and 22.5 degC, placing 5% probability on the possibility that an Eocene *Avicennia*-only mangrove record occurred outside of this range.
-We define a 95% probability density for locations with *Avicennia* and Rhizophoraceae in a similar way.
-To allow for greater uncertainty at the lower temperature limit, we use the empirical 20.7 degC limit.
-For the upper limit, we use the maximal coral reef temperature of 29.5 degC from @kleypas1999, as the compendium of @Quisthoudt2012 does not include the warmest, equatorial regions.
-The mean and standard deviation of locations with *Avicennia* and Rhizophoraceae are thus set to 25.1 and 2.2, respectively.
-
 [Fig 1: Map and distribution of proxy data and ecological data]
@@
 XXX (REF).
-The midpoint age of the EECO (51.5 Ma), along with the present-day coordinates of geochemical and ecological data, were used for palaeorotation.
+The midpoint age of the EECO (51.2 Ma), along with the present-day coordinates of geochemical and ecological data, were used for palaeorotation.
 
@@
 - Show Eocene gradient
-- Global t
+- Global t : use raster package for calculation of areal weights
 - report lat gradient

### 2022-10-09 manuscript/manuscript.Rmd
lat band segment areal weight solution

@@
 ## Parameter estimation
-We estimated the posterior distributions of the model parameters using a custom-made Markov chain Monte Carlo (MCMC) algorithm, written in R.
-Specifically, we sampled the unknown parameters $A$, $K$, $M$ and $B$ using Metropolis-Hastings, and used Gibbs sampling to estimate all other unknown parameters.
+We estimated the posterior distributions of the model parameters using a Markov chain Monte Carlo (MCMC) algorithm, written in R.
+Specifically, we sampled the unknown parameters $A$, $K$, $M$ and $B$ with Metropolis-Hastings, and used Gibbs sampling to estimate all other unknown parameters.
 Posterior inference on the modern gradient is based on four chains with 60,000 iterations each, 10,000 of which were discarded as burn-in, and keeping every 10th iteration, resulting in a total of 20,000 iterations.
@@
 
-The global Eocene temperatures
+The global Eocene mean sea surface temperature is estimated at 28.8 (95% CI: 25.7 - 31.7) degC, significantly higher than the modern (?? degC).
 

### 2022-10-09 R/analyses/assess_EECO_climate_model_output.R

@@ -50,3 +50,10 @@ legend("topright",c("posterior mean (proxy obs.)", "posterior mean (distribution
 # modern_sample <- readRDS("./results/modern/modern_climate_model_output.rds")
 # points(0:90,apply(temp_from_gradient(0:90,modern_sample),1,median), type = "l", lwd = 2)
 
+# global average temperature
+alpha1 <- seq(1,90,1)
+alpha2 <- seq(0,89,1)
+latweight <- sin(pi*alpha1/180) - sin(pi*alpha2/180)
+eotemp <- gradient(seq(0.5,89.5,1), mode_all[,1:4],0)
+eotempq <- apply(eotemp,2,function(x) quantile(x, probs = c(0.025,0.5,0.975)))
+apply(eotempq,1, function(x) sum(x*latweight))

### 2022-10-13 manuscript/manuscript.Rmd
in-line code for global average T in manuscript

@@
 
+```{r results block 1 - for in-line code}
+setwd('..') # move up to parent folder of /manuscript
+source("R/functions/model_components/gradient.R")
+mode_all <- readRDS("results/eeco/eeco_climate_model_output_params.rds")
+modm_all <- readRDS("results/modern/modern_climate_model_output.rds")
+
+# latitudinal weights in 1 deg lat bands
+alpha1 <- seq(1,90,1)
+alpha2 <- seq(0,89,1)
+latweight <- sin(pi*alpha1/180) - sin(pi*alpha2/180) # sums to 1 - careful when using lat. subsets
+# eocene temperature gradient
+eotemp <- gradient(seq(0.5,89.5,1), mode_all[,1:4],0)
+# eocene temperature gradient quantiles
+eotempq <- apply(eotemp,2,function(x) quantile(x, probs = c(0.025,0.5,0.975)))
+# eocene global average temperature with 95% CI
+eotemp_global <- apply(eotempq,1, function(x) round(sum(x*latweight),1), simplify = FALSE)
+# modern temperature gradient
+motemp <- gradient(seq(0.5,89.5,1), modm_all[,1:4],0)
+# modern temperature gradient quantiles
+motempq <- apply(motemp,2,function(x) quantile(x, probs = c(0.025,0.5,0.975)))
+# modern global average temperature with 95% CI
+motemp_global <- apply(motempq,1, function(x) round(sum(x*latweight),1), simplify = FALSE)
+
+```
+
 The modelled Eocene temperature gradient is starkly different from the modern (Fig 4).
@@
 
-The global Eocene mean sea surface temperature is estimated at 28.8 (95% CI: 25.7 - 31.7) degC, significantly higher than the modern (?? degC).
+The global Eocene mean sea surface temperature is estimated at `r eotemp_global$"50%"` (95% CI: `r eotemp_global$"2.5%"` - `r eotemp_global$"97.5%"`) degC, significantly higher than the modern (`r motemp_global$"50%"` degC).
 

### 2022-10-13 R/analyses/assess_EECO_climate_model_output.R

@@ -23,7 +23,7 @@ plot_chains(mod, params = c("sdy", "logpost"))
 
 # combine chains, discard burnin
 mode_all <- combine_posterior(mod,100000)
-
+saveRDS(mode_all,"results/eeco/eeco_climate_model_output_params.rds")
 # check effective sample size for A, dKA, M, and B
 mcmcse::multiESS(mode_all[,1:4])
 

### 2022-10-14 manuscript/manuscript.Rmd
comment manuscript
prepare data for Fig. 4, re-ran eeco and modern with eeco climate models

@@
 Geochemical climate proxy data were extracted from a latest Paleocene and early Eocene compilation [@hollis2019].
-This compilation provides  For our analyses, this dataset was restricted to the EECO and contains four commonly used geochemical climate proxies: d180, d47, Mg/Ca and TEX86 (REFS).
+This compilation provides four different geochemical proxies for reconstructing seawater temperature: d180, d47, Mg/Ca and TEX86.
+For our analyses, this dataset was restricted to the EECO and contains four commonly used geochemical climate proxies: d180, d47, Mg/Ca and TEX_86_ (REFS).
 For a detailed description of each proxy see [@hollis2019].
@@
 
-**Coral reefs.** Today, shallow warm-water coral reefs are limited to tropical and subtropical latitudes (~34°N–32°S), with minimum sea surface temperature (~18 °C) tolerances being the primary constraint on this distribution [@johannes1983; @kleypas1999; @yamano2001].
+**Coral reefs.** Today, shallow warm-water coral reefs are limited to tropical and subtropical latitudes (~34°N–32 degS), with minimum sea surface temperature (~18 degC) tolerances being the primary constraint on this distribution [@johannes1983; @kleypas1999; @yamano2001].
 As coral reefs reside at the upper thermal limit of the oceans today, their maximum sea surface temperature tolerance is less well-constrained with some studies suggesting up to 35.6 degC in the geological past [@jones2022reef].
@@
 For the Eocene model, we ran four chains with 600,000 iterations each, discarding 100,000 as burn-in and keeping every 100th iteration, as the hierarchical model structure results in higher autocorrelation of the chains.
-The Eocene posterior inference is thus based on a total of 20,000 iterations with low autocorrelation (effective multivariate sample size for $A$, $K$, $M$ and $B$ is > 18,000).
+The Eocene posterior inference is thus based on a total of 210,000 iterations with low autocorrelation (effective multivariate sample size for $A$, $K$, $M$ and $B$ is > 18,000).
 
@@
 modm_all <- readRDS("results/modern/modern_climate_model_output.rds")
+
+## outsource to a seperate analysis script
 

### 2022-10-14 R/figures/fig_4.R

@@ -1,0 +1,20 @@
+source("./R/functions/model_processing/temp_from_gradient.R")
+source("./R/functions/model_processing/combine_posterior.R")
+# Load data -------------------------------------------------------------------
+eeco_mod <- readRDS("./results/eeco/eeco_climate_model_output.rds")
+ecol_data <- readRDS("./data/processed/distribution_matrix_for_EECO_model.rds")
+geochem_data <- readRDS("./data/processed/observation_matrix_for_EECO_model.rds")
+
+
+# Data preparation ------------------------------------------------------------
+eeco_mod <- combine_posterior(mod = eeco_mod, burnin = 100000)
+
+# add proxy type to ecol_data
+ecol_data$proxy <- rep("coral reef", 11)
+ecol_data$proxy[1:2] <- "Avicennia"
+ecol_data$proxy[3:7] <- "A. and Rhizophoraceae"
+
+# Calculate eeco temperatures
+lat <- seq(from = 0, to = 90, by = 0.1)
+eeco_temp <- temp_from_gradient(lat = lat, model_out = eeco_mod)
+

### 2022-10-14 manuscript/manuscript.Rmd
geochem update

@@
 Geochemical climate proxy data were extracted from a latest Paleocene and early Eocene compilation [@hollis2019].
-This compilation provides four different geochemical proxies for reconstructing seawater temperature: d180, d47, Mg/Ca and TEX86.
-For our analyses, this dataset was restricted to the EECO and contains four commonly used geochemical climate proxies: d180, d47, Mg/Ca and TEX_86_ (REFS).
+This compilation provides four different geochemical proxies for reconstructing seawater temperature: $\delta^18^O, \Delta$_47_, Mg/Ca and TEX_86_.
+For our analyses, this dataset was restricted to the EECO and samples from shallow.
+We also excluded recrystallised $\delta^18^O samples as secondary diagenetic calcite precipitated after deposition can bias isotope measurements and offset temperature values [@schrag1999].
+
  For a detailed description of each proxy see [@hollis2019].
@@
 
+To test whether our logistic regression model can adequately describe latitudinal temperature gradients, we applied a simplified version of the model to modern sea surface temperatures from Bio-Oracle [@...]
+
+the climate model is able to estimate latitudinal temperature gradients from limited data, we subsampled modern sea surface temperatures from the Bio-Or
+
 ## Model application

### 2022-10-14 manuscript/manuscript.Rmd
model validation, refs

@@
 This compilation provides four different geochemical proxies for reconstructing seawater temperature: $\delta^18^O, \Delta$_47_, Mg/Ca and TEX_86_.
-For our analyses, this dataset was restricted to the EECO and samples from shallow.
-We also excluded recrystallised $\delta^18^O samples as secondary diagenetic calcite precipitated after deposition can bias isotope measurements and offset temperature values [@schrag1999].
-
+For our analyses, this dataset was restricted to the EECO and samples from the continental shelf.
+Recrystallised $\delta^18^O samples were also excluded as secondary diagenetic calcite precipitated after deposition can bias isotope measurements and offset temperature values [@schrag1999].
+This resulted in most $\delta^18^O samples being excluded from the dataset (x out of x).
+After data filtering, x geochemical proxy samples remained.
 For a detailed description of each proxy see [@hollis2019].
-
-- palaeoenvironment of data (which filters were applied)
-- recrystallisation (excluding most d18O)
-
-Do we need a description of each? Not in detail, 1 paragraph in total is enough
- Measurements of the ratio of stable isotope 18O and 16O from fossils provide
 

### 2022-10-14 manuscript/manuscript.Rmd
model validation, refs saved

@@
 
-To test whether our logistic regression model can adequately describe latitudinal temperature gradients, we applied a simplified version of the model to modern sea surface temperatures from Bio-Oracle [@...]
-
-the climate model is able to estimate latitudinal temperature gradients from limited data, we subsampled modern sea surface temperatures from the Bio-Or
+To test whether our logistic regression model can adequately describe latitudinal temperature gradients, we applied a simplified version of the model to modern sea surface temperatures from Bio-Oracle [@...].
+As these modern, annual sea surface temperatures estimates are associated with little uncertainty, there is no need for a hierarchical model structure, and we substituted temperature ($t$) for $\mu$ in eq.
+?
+and ?.
+To verify that the modern gradient can be approximated with limited sampling, we resampled modern sea surface temperatures at modern latitudes corresponding the palaeolatitudes of the Eocene samples.
+This process was repeated 100 times, randomly chosing a longitude for each latitude in each repetition.
+The parameters of the non-hierarchical temperature model were estimated for each of the 100 samples, and the iterations after burn-in from the posterior of the parameters were pooled to generate the resulting median temperature gradient with 95 % credible intervals.
 

### 2022-10-14 manuscript/manuscript.Rmd
global mean T script

@@
 
-## Model application
-- Calculate lat gradient
-- Calculate global T
-
-[Fig. 4]
-
 ## Parameter estimation
@@
 For the Eocene model, we ran four chains with 600,000 iterations each, discarding 100,000 as burn-in and keeping every 100th iteration, as the hierarchical model structure results in higher autocorrelation of the chains.
-The Eocene posterior inference is thus based on a total of 210,000 iterations with low autocorrelation (effective multivariate sample size for $A$, $K$, $M$ and $B$ is > 18,000).
+The Eocene posterior inference is thus based on a total of 20,000 iterations with low autocorrelation (effective multivariate sample size for $A$, $K$, $M$ and $B$ is > 18,000).
+
+## Processing of model results
+- Calculate lat gradient
+- Calculate global T
+[Fig. 4]
+
+Modelled sea surface temperature estimates were generated with eq.
+?, calculating the sea surface temperatures at any latitude with the parameter estimates of each iteration from the posterior.
+The median, $2.5^{th}$ and $97.5^{th}$ percentile of temperatures where then taken from all temperature estimates obtained at the latitude of interest.
+
+Differences between Eocene and modern temperatures at a certain latitude were calculated by randomly pairing all iterations of the posterior from the Eocene and modern temperature gradient model, calculating the Eocene and modern temperature using the respective iterations, taking the difference, and then calculating the median (2.5th, 97.5th percentile) from all pairs of iterations.
+
+Global average temperatures were calculated by taking the weighted mean of the median (2.5th, 97.5th percentile) temperature estimates in 1 degree latitudinal bins.
+The weights were set to the proportion of global surface area in each latitudinal bin, i.e.
+decreasing with increasing latitude as:
+$$weights = sin(\alpha_{1,i}) - sin(\alpha_{2,i}) ,$$
+where $\alpha1$ is the upper, and $\alpha2$ is the lower latitudinal boundary of bin $i$.
 
@@
 
-```{r results block 1 - for in-line code}
+```{r results block 1 - for in-line code, echo = FALSE}
 setwd('..') # move up to parent folder of /manuscript
-source("R/functions/model_components/gradient.R")
-mode_all <- readRDS("results/eeco/eeco_climate_model_output_params.rds")
-modm_all <- readRDS("results/modern/modern_climate_model_output.rds")
-
-## outsource to a seperate analysis script
-
-# latitudinal weights in 1 deg lat bands
-alpha1 <- seq(1,90,1)
-alpha2 <- seq(0,89,1)
-latweight <- sin(pi*alpha1/180) - sin(pi*alpha2/180) # sums to 1 - careful when using lat. subsets
-# eocene temperature gradient
-eotemp <- gradient(seq(0.5,89.5,1), mode_all[,1:4],0)
-# eocene temperature gradient quantiles
-eotempq <- apply(eotemp,2,function(x) quantile(x, probs = c(0.025,0.5,0.975)))
-# eocene global average temperature with 95% CI
-eotemp_global <- apply(eotempq,1, function(x) round(sum(x*latweight),1), simplify = FALSE)
-# modern temperature gradient
-motemp <- gradient(seq(0.5,89.5,1), modm_all[,1:4],0)
-# modern temperature gradient quantiles
-motempq <- apply(motemp,2,function(x) quantile(x, probs = c(0.025,0.5,0.975)))
-# modern global average temperature with 95% CI
-motemp_global <- apply(motempq,1, function(x) round(sum(x*latweight),1), simplify = FALSE)
+motemp_global <- readRDS("results/modern/global_mean_modern.rds")
+eotemp_global <- readRDS("results/eeco/global_mean_eeco.rds")
 

### 2022-10-14 R/analyses/global_mean_temperatures.R

@@ -1,0 +1,26 @@
+# global mean temperatures for the modern and the Eocene
+
+source("R/functions/model_components/gradient.R")
+mode_all <- readRDS("results/eeco/eeco_climate_model_output_params.rds")
+modm_all <- readRDS("results/modern/modern_climate_model_output.rds")
+
+# latitudinal weights in 1 deg lat bands
+alpha1 <- seq(1,90,1)
+alpha2 <- seq(0,89,1)
+latweight <- sin(pi*alpha1/180) - sin(pi*alpha2/180) # sums to 1 - careful when using lat. subsets
+# eocene temperature gradient
+eotemp <- gradient(seq(0.5,89.5,1), mode_all[,1:4],0)
+# eocene temperature gradient quantiles
+eotempq <- apply(eotemp,2,function(x) quantile(x, probs = c(0.025,0.5,0.975)))
+# eocene global average temperature with 95% CI
+eotemp_global <- apply(eotempq,1, function(x) round(sum(x*latweight),1), simplify = FALSE)
+# modern temperature gradient
+motemp <- gradient(seq(0.5,89.5,1), modm_all[,1:4],0)
+# modern temperature gradient quantiles
+motempq <- apply(motemp,2,function(x) quantile(x, probs = c(0.025,0.5,0.975)))
+# modern global average temperature with 95% CI
+motemp_global <- apply(motempq,1, function(x) round(sum(x*latweight),1), simplify = FALSE)
+#
+# save results
+saveRDS(motemp_global,"results/modern/global_mean_modern.rds")
+saveRDS(eotemp_global,"results/eeco/global_mean_eeco.rds")

### 2022-10-14 manuscript/manuscript.Rmd
clean scripts and result names

@@
 This process was repeated 100 times, randomly chosing a longitude for each latitude in each repetition.
-The parameters of the non-hierarchical temperature model were estimated for each of the 100 samples, and the iterations after burn-in from the posterior of the parameters were pooled to generate the resulting median temperature gradient with 95 % credible intervals.
+The parameters of the non-hierarchical temperature model were estimated for each of the 100 samples, and the iterations after burn-in from the posterior of the parameters were pooled to generate the resulting median temperature gradient with 95 % credible intervals ($2.5^{th}$ to $97.5^{th}$ percentile).
 
@@
 
-Differences between Eocene and modern temperatures at a certain latitude were calculated by randomly pairing all iterations of the posterior from the Eocene and modern temperature gradient model, calculating the Eocene and modern temperature using the respective iterations, taking the difference, and then calculating the median (2.5th, 97.5th percentile) from all pairs of iterations.
+Differences between Eocene and modern temperatures at a certain latitude were calculated by randomly pairing all iterations of the posterior from the Eocene and modern temperature gradient model, calculating the Eocene and modern temperature using the respective iterations, taking the difference, and then calculating the median ($2.5^{th}$, $97.5^{th}$ percentile) from all pairs of iterations.
 
-Global average temperatures were calculated by taking the weighted mean of the median (2.5th, 97.5th percentile) temperature estimates in 1 degree latitudinal bins.
+Global average temperatures were calculated by taking the weighted mean of the median ($2.5^{th}$, $97.5^{th}$ percentile) temperature estimates in 1 degree latitudinal bins.
 The weights were set to the proportion of global surface area in each latitudinal bin, i.e.
@@
 
-```{r results block 1 - for in-line code, echo = FALSE}
+```{r results block - for in-line code, echo = FALSE}
 setwd('..') # move up to parent folder of /manuscript
+
+# add gradient results here
+
 motemp_global <- readRDS("results/modern/global_mean_modern.rds")

### 2022-10-14 R/analyses/assess_EECO_climate_model_output.R

@@ -50,10 +50,3 @@ legend("topright",c("posterior mean (proxy obs.)", "posterior mean (distribution
 # modern_sample <- readRDS("./results/modern/modern_climate_model_output.rds")
 # points(0:90,apply(temp_from_gradient(0:90,modern_sample),1,median), type = "l", lwd = 2)
 
-# global average temperature
-alpha1 <- seq(1,90,1)
-alpha2 <- seq(0,89,1)
-latweight <- sin(pi*alpha1/180) - sin(pi*alpha2/180)
-eotemp <- gradient(seq(0.5,89.5,1), mode_all[,1:4],0)
-eotempq <- apply(eotemp,2,function(x) quantile(x, probs = c(0.025,0.5,0.975)))
-apply(eotempq,1, function(x) sum(x*latweight))

### 2022-10-14 R/analyses/calculate_temperature_gradients_from_models.R

@@ -1,0 +1,32 @@
+### calculating results
+source("./R/functions/model_processing/temp_from_gradient.R")
+
+# modern
+modm <- readRDS("./results/modern/modern_climate_model_output.rds")
+
+# eocene
+mode <- readRDS("./results/eeco/eeco_climate_model_output_combined.rds")
+# combine chains, discard burnin
+mode <- combine_posterior(mode,100000)
+
+# calculate temperatures at lat 0 and lat 90 for every iteration
+modern_t0 <- gradient(0,modm[,1:4])
+eocene_t0 <- gradient(0,mode[,1:4])
+
+modern_t90 <- gradient(90,modm[,1:4])
+eocene_t90 <- gradient(90,mode[,1:4])
+
+n <- length(modern_t0)
+
+# randomly recombine iterations of modern and eocene gradient and take difference
+set.seed(1)
+quantile(eocene_t0[sample(1:n,n)] - modern_t0[sample(1:n,n)], probs = c(0.025,0.5,0.975))
+quantile(eocene_t90[sample(1:n,n)] - modern_t90[sample(1:n,n)], probs = c(0.025,0.5,0.975))
+
+# calculate overall gradient for modern and eocene (polar - equatorial difference)
+quantile(eocene_t0 - eocene_t90, probs = c(0.025,0.5,0.975))
+quantile(modern_t0 - modern_t90, probs = c(0.025,0.5,0.975))
+
+# calculate global average temperatures
+
+(2*pi*1^2)*(cos(90-40)-cos(90-0))

### 2022-10-14 R/analyses/global_mean_temperatures.R

@@ -1,7 +1,7 @@
 # global mean temperatures for the modern and the Eocene
 
 source("R/functions/model_components/gradient.R")
-mode_all <- readRDS("results/eeco/eeco_climate_model_output_params.rds")
+mode_all <- readRDS("results/eeco/eeco_climate_model_output_combined.rds")
 modm_all <- readRDS("results/modern/modern_climate_model_output.rds")
 
 # latitudinal weights in 1 deg lat bands

### 2022-10-14 R/analyses/run_EECO_climate_model.R

@@ -60,4 +60,9 @@ parallel::stopCluster(cl)
 ### save output
 saveRDS(mode, "results/eeco/eeco_climate_model_output.rds")
 
+# combine chains and discard burn-in
+mode_all <- combine_posterior(mode,100000)
+# save combined chains
+saveRDS(mode_all, "results/eeco/eeco_climate_model_output_combined.rds")
+
 ### To check results use the "assess_EECO-climate_model_output_script.R

### 2022-10-16 manuscript/manuscript.Rmd
sdy and gradient results

@@
 This normal probability distribution was defined with a mean of 27.6 and a standard deviation of 2.125, placing 97.5% of the probability density above the minimum.
-As the distribution of modern corals is skewed towards warmer temperatures, this approach results in 16.5% of the probability being placed on temperatures > 29.5 degC, allowing for the possibilty that Eocene coral reefs were adapted to warmer conditions than modern-day coral reefs.
+As the distribution of modern corals is skewed towards warmer temperatures, this approach results in 16.5% of the probability being placed on temperatures > 29.5 degC, allowing for the possibility that Eocene coral reefs were adapted to warmer conditions than modern-day coral reefs.
 
@@
 
-# add gradient results here
+# temperature gradients
+tdiff_lat0 <- readRDS("results/eeco/eocene_modern_difference_equator.rds")
+tdiff_lat90 <- readRDS("results/eeco/eocene_modern_difference_poles.rds")
+eocene_gradient <- readRDS("results/eeco/eocene_overall_gradient.rds")
+modern_gradient <- readRDS("results/modern/modern_overall_gradient.rds")
 
+# sdy
+eeco_sdy <- readRDS("results/eeco/eeco_sdy.rds")
+modern_eeco_sample_sdy <- readRDS("results/modern/modern_eeco_sample_sdy.rds")
+
+# global average temperatures
 motemp_global <- readRDS("results/modern/global_mean_modern.rds")
@@
 The modelled Eocene temperature gradient is starkly different from the modern (Fig 4).
-Modelled, median equatorial temperatures are 4.2 (95% CI: 0.2 - 8.3) degC higher for the Eocene, and polar temperatures are higher by 25.0 (95% CI: 17.0 - 29.1) degC.
-This results in a strongly flattened latitudinal temperature gradient of 9.0  (95% CI: 2.5 - 17.8) degC for the Eocene, as opposed to 29.6 degC for the modern.
-The high variability of early Eocene palaeotemperature proxies, particularly in the mid-latitudes, and the scarcity of high-latitude data, results in substantial uncertainties in the modelled temperature gradient, but ...
+Modelled, median equatorial temperatures are `r tdiff_lat0$"50%"` (95% CI: `r tdiff_lat0$"2.5%"` - `r tdiff_lat0$"97.5%"`) degC higher for the Eocene, and polar temperatures are higher by `r tdiff_lat90$"50%"` (95% CI: `r tdiff_lat90$"2.5%"` - `r tdiff_lat90$"97.5%"`) degC.
+This results in a strongly flattened latitudinal temperature gradient of `r eocene_gradient$"50%"`  (95% CI: `r eocene_gradient$"2.5%"` - `r eocene_gradient$"97.5%"`) degC for the Eocene, as opposed to `r modern_gradient$"50%"` degC for the modern.
+
+The high variability of early Eocene palaeotemperature proxies, particularly in the mid-latitudes, and the scarcity of high-latitude data, results in substantial uncertainties in the modeled temperature gradient.
+This is reflected in the residual standard deviation ($\sigma$) of the early Eocene gradient, `r eeco_sdy$"50%"` (95% CI: `r eeco_sdy$"2.5%"` - `r eeco_sdy$"97.5%"`) degC, more than twice as high than  $\sigma$ of the modern gradient, resampled at early Eocene latitudes, `r modern_eeco_sample_sdy$"50%"` (95% CI: `r modern_eeco_sample_sdy$"2.5%"` - `r modern_eeco_sample_sdy$"97.5%"`) degC.
 
@@
 - x1 compare lat gradient to literature estimates, e.g. Zhang2019
+- compare lat gradient to climate models and the "early EECO problem"
+- discuss insufficiency of proxies (high variability, ...)
+- discuss utility of the model and future applications
 

### 2022-10-16 R/analyses/assess_sdy.R

@@ -1,0 +1,12 @@
+# compare gradient sdy
+source("./R/functions/model_processing/combine_posterior.R")
+modern_gradient_eeco_lat <- readRDS("./results/modern/modern_sample_eocene_p_lat_gradient.RDS")
+modern_gradient_eeco_lat <- combine_posterior(mod = modern_gradient_eeco_lat, burnin = 5000)
+eeco_gradient <- readRDS("./results/eeco/eeco_climate_model_output_combined.rds")
+
+modern_eeco_sample_sdy <- as.list(format(round(quantile(eocene_sample$sdy,probs = c(.025,.5,.975)),1),nsmall = 1))
+
+eeco_sdy <- as.list(format(round(quantile(mode_all$sdy,probs = c(.025,.5,.975)),1),nsmall = 1))
+
+saveRDS(modern_eeco_sample_sdy, "results/modern/modern_eeco_sample_sdy.rds")
+saveRDS(eeco_sdy, "results/eeco/eeco_sdy.rds")

### 2022-10-16 R/analyses/calculate_temperature_gradients_from_models.R

@@ -20,15 +20,15 @@ n <- length(modern_t0)
 
 # randomly recombine iterations of modern and eocene gradient and take difference
 set.seed(1)
-tdiff_lat0 <- as.list(quantile(eocene_t0[sample(1:n,n)] - modern_t0[sample(1:n,n)], probs = c(0.025,0.5,0.975)))
-tdiff_lat90 <- as.list(quantile(eocene_t90[sample(1:n,n)] - modern_t90[sample(1:n,n)], probs = c(0.025,0.5,0.975)))
+tdiff_lat0 <- as.list(format(round(quantile(eocene_t0[sample(1:n,n)] - modern_t0[sample(1:n,n)], probs = c(0.025,0.5,0.975)),1),nsmall = 1))
+tdiff_lat90 <- as.list(format(round(quantile(eocene_t90[sample(1:n,n)] - modern_t90[sample(1:n,n)], probs = c(0.025,0.5,0.975)),1),nsmall = 1))
 
 # calculate overall gradient for modern and eocene (polar - equatorial difference)
-eocene_gradient <- as.list(quantile(eocene_t0 - eocene_t90, probs = c(0.025,0.5,0.975)))
-modern_gradient <- as.list(quantile(modern_t0 - modern_t90, probs = c(0.025,0.5,0.975)))
+eocene_gradient <- as.list(format(round(quantile(eocene_t0 - eocene_t90, probs = c(0.025,0.5,0.975)),1),nsmall = 1))
+modern_gradient <- as.list(format(round(quantile(modern_t0 - modern_t90, probs = c(0.025,0.5,0.975)),1),nsmall = 1))
 
 # save results
-saveRDS(tdiff_lat0, "eocene_modern_difference_equator.rds")
-saveRDS(tdiff_lat90, "eocene_modern_difference_poles.rds")
-saveRDS(eocene_gradient, "eocene_overall_gradient.rds")
-saveRDS(modern_gradient, "modern_overall_gradient.rds")
+saveRDS(tdiff_lat0, "./results/eeco/eocene_modern_difference_equator.rds")
+saveRDS(tdiff_lat90, "./results/eeco/eocene_modern_difference_poles.rds")
+saveRDS(eocene_gradient, "./results/eeco/eocene_overall_gradient.rds")
+saveRDS(modern_gradient, "./results/modern/modern_overall_gradient.rds")

### 2022-12-02 manuscript/manuscript.Rmd
introduction

@@
 ---
-title: "An Early Eocene temperature gradient modelled with geological and ecological proxies"
+title: "Combinig geological and ecological proxies to model latitudinal temperature gradients in the early Eocene greenhouse"
 
@@
 "The Early Eocene Climate Optimum reconstructed with Bayesian multi-proxy integration"
+"An Early Eocene temperature gradient modelled with geological and ecological proxies"
 -->
@@
 # Introduction
-The geological record holds evidence for episodes in Earth history with fundamentally different climates than our current.
-Some episodes, for example the early Eocene, are thought to have been significantly warmer than the modern, potentially representing an analogue for extreme climate warming scenarios.
+The geological record holds evidence for episodes in Earth history with fundamentally different climates than our current [@Scotese2021].
+Some episodes, for example the early Eocene, are thought to have been significantly warmer than the modern [@Pross2012], potentially representing an analogue for extreme climate warming scenarios [@Burke2018].
 Reconstructions of palaeoclimate commonly follow one of two approaches.
-Geochemical proxies such as oxygen isotopes (d18O), tetraether indices (TEX86) and magnesium calcium ratios (Mg/Ca) are used to derive palaeotemperature estimates for individual locations using experimentally derived transfer functions.
-Multiple local estimates are sometimes summarised to generate regional or global estimates, but this approach is prone to sampling bias [@Jones2022].
-Earth system models take a fundamentally different approach by modelling the climate-generating processes directly, reconstructing spatially resolved climate estimates for the entire glob.e This approach is, however, relying on proxy data to constrain the large number of possible model setups and ground truth the model output.
+Geochemical proxies such as oxygen isotopes (d18O), tetraether indices (TEX86) and magnesium calcium ratios (Mg/Ca) are used to derive palaeotemperature estimates for individual locations using experimentally derived transfer functions [e.g.
+@Liu2009].
+To generate regional or global average estimates, the mean of multiple local estimates is taken [e.g.
+@Veizer2015].
+This approach has the advantage of being transparent and easy to reproduce, modify or expand with new data, but is prone to sampling bias [@Jones2022].
+Earth system models take a fundamentally different approach by modelling the climate-generating processes directly, reconstructing spatially resolved climate estimates for the entire globe [@???].
+However, this approach still relies on proxy data to constrain the large number of possible model setups and ground truth the model output [@???] It also requires substantial computing resources and technical expert knowledge, and is thus difficult to implement or modify for most palaeoclimate researchers.
+
+Climate reconstructions of low - intermediate complexity provide an alternative that can be employed more readily by a broad range of users.
+Examples of this approach include proxy compilations that take into account the spatial nature of the data, e.g.
+@Hollis2019 investigated latitudinal temperature trends across the late Palaeocene - early Eocene with a spatially resolved proxy record.
+Expanding from this approach, some proxy-based reconstructions use statistical modelling to infer climatic patterns, and to bridge data gaps, e.g.
+using polynomials to reconstruct latitudinal gradients [@Bijl2009], or Gaussian process regression for a 2D-reconstruction of surface temperatures [@Inglis2020].
+These latter approaches take advantage of the autoregressive nature
 
@@
 In order to properly compare model output to proxy data, latitudinal temperature patterns need to be reconstructed from noisy data (impossible to properly compare model gradients with scattered proxy points "by eye")
+
+Reconstructing Earth's climate history is fundamental for understanding the evolution of the Earth system and its consequences for life in the past and future.
+The geological record, the only empirical source of information on the palaeoclimate of the deep past, holds evidence.
+
 -->

### 2022-12-02 manuscript/manuscript.Rmd
introduction writing

@@
 Some episodes, for example the early Eocene, are thought to have been significantly warmer than the modern [@Pross2012], potentially representing an analogue for extreme climate warming scenarios [@Burke2018].
-Reconstructions of palaeoclimate commonly follow one of two approaches.
+Reconstructions of palaeoclimate commonly use 1) geochemical proxy data, 2) ecological or lithological climate indicators, or 3) earth system modelling.
+
 Geochemical proxies such as oxygen isotopes (d18O), tetraether indices (TEX86) and magnesium calcium ratios (Mg/Ca) are used to derive palaeotemperature estimates for individual locations using experimentally derived transfer functions [e.g.
@@
 This approach has the advantage of being transparent and easy to reproduce, modify or expand with new data, but is prone to sampling bias [@Jones2022].
+Ecological and lithological climate data provide an alternative method of reconstructing palaeotemperatures, using e.g.
+leaf physiognomy and floral assemblages to derive palaeotemperature proxies [@Greenwood2003], or using lithological indicators to determine the range of palaeoclimatic belts [@Scotese2021].
+Despite some recent advances [@Chandra2021], ecological and lithological climate indicators are still rarely used in quantitative palaeoclimate reconstructions, particularly for the marine realm.
 Earth system models take a fundamentally different approach by modelling the climate-generating processes directly, reconstructing spatially resolved climate estimates for the entire globe [@???].
@@
 
-Climate reconstructions of low - intermediate complexity provide an alternative that can be employed more readily by a broad range of users.
-Examples of this approach include proxy compilations that take into account the spatial nature of the data, e.g.
-@Hollis2019 investigated latitudinal temperature trends across the late Palaeocene - early Eocene with a spatially resolved proxy record.
-Expanding from this approach, some proxy-based reconstructions use statistical modelling to infer climatic patterns, and to bridge data gaps, e.g.
-using polynomials to reconstruct latitudinal gradients [@Bijl2009], or Gaussian process regression for a 2D-reconstruction of surface temperatures [@Inglis2020].
-These latter approaches take advantage of the autoregressive nature
+Spatially resolved climate reconstructions provide an alternative that can be employed more readily by a broad range of users.
+Examples of this approach include the visualisation of proxy compilations along a latitudinal gradient [@Hollis2019, @Vickers2021].
+Interpolation is sometimes used to bridge spatial gaps in palaeoclimate data [@Taylor2004], taking advantage of the autoregressive nature of climatic data: much of the information on the climate of any given location is contained in the climate data of nearby locations [@Reynolds1994].
+Adding to this, some proxy-based reconstructions use statistical modelling to infer climatic patterns.
+For example, polynomial regression has been used to reconstruct latitudinal temperature gradients [@Bijl2009], and 2D-reconstructions of surface temperatures have been created with Gaussian process regression [@Inglis2020].
+
+Here, we present a Bayesian, hierarchical model for inferring sea surface temperatures (SST) that expands upon existing, spatial reconstructions of palaeoclimate by allowing for the integration of 1) prior information based on physical principles and the observed, modern SST distribution, and of 2) geochemical and ecological climate proxies in a common, quantitative framework.
+We apply this approach to the early Eocene climatic optimum (EECO), the interval with the warmest sustained temperatures of the Cenozoic.
+Assuming the conservation of thermal tolerances of modern coral reefs and mangrove taxa, we establish broad temperature ranges for EECO coral reef and mangrove sites.
+We integrate these temperature estimates with the EECO geochemical, shallow marine proxy record to model the latitudinal SST gradient and its uncertainty.
+
+<!--
 
@@
 - Reconstructing past climate important for understanding current/future global warming
-- Two main pathways: highly complex ESM <--> proxy compilations
+- Two main pathways: highly complex ESM <-> proxy compilations
 - Problems with proxy data: only provides a local picture (few local sites interpreted as global signal)
@@
 - exemplified with the EECO: proxies and ESM hard to reconcile, EECO important climate analogue
-- a lot of proxies recrystallised <--> unreliable data or lack of data
+- a lot of proxies recrystallised <-> unreliable data or lack of data
 
@@
 
-Here, we present a Bayesian hierarchical model to integrate geological data---with established temperature proxies---into a unified quantitative framework.
-We apply this approach to the early Eocene climatic optimum (EECO), the interval with the warmest sustained temperatures of the Cenozoic.
-Assuming the conservation of thermal tolerances of modern coral reefs and mangrove taxa, we establish broad sea surface temperature ranges for EECO coral reef and mangrove sites.
-We integrate these temperature estimates with the EECO geochemical shallow marine proxy record to model the latitudinal temperature gradient and its uncertainty.
-
 - EECO specific information
 - Disagreement between GCM + geochem data
+-->
 
@@
 
+Here, we present a Bayesian hierarchical model to integrate geological data---with established temperature proxies---into a unified quantitative framework.
+We apply this approach to the early Eocene climatic optimum (EECO), the interval with the warmest sustained temperatures of the Cenozoic.
+Assuming the conservation of thermal tolerances of modern coral reefs and mangrove taxa, we establish broad sea surface temperature ranges for EECO coral reef and mangrove sites.
+We integrate these temperature estimates with the EECO geochemical shallow marine proxy record to model the latitudinal temperature gradient and its uncertainty.
+
+Our hierarchical approach accounts for different sample sizes between and uncertainties within proxy sites, and
+
 -->
+
 # Materials & Methods
@@
 [Fig. 3]
+<!--
 - Modern data
 - Sample based on Eocene localities
+-->
 
@@
 ## Processing of model results
+<!--
 - Calculate lat gradient
@@
 [Fig. 4]
-
+-->
 Modelled sea surface temperature estimates were generated with eq.
@@
 ## Model validation
-- Show that the 4-parameter logistic function can adequatly describe the modern temperature gradient
+<!-- - Show that the 4-parameter logistic function can adequatly describe the modern temperature gradient
 - Show that the modern gradient can be reproduced with Eocene sampling distribution
+-->
 [Fig 3 - Modern gradient]
@@
 ## EECO reconstruction
-- Show Eocene gradient
+<!-- - Show Eocene gradient
 - Global t : use raster package for calculation of areal weights
 - report lat gradient
-
+-->
 ```{r results block - for in-line code, echo = FALSE}
@@
 
-## Fig 4 - compare Eocene gradient with previous reconstructions / ES models
+## Fig 4 - compare Eocene gradient with modern gradient, also with previous reconstructions / ES models?
 

### 2022-12-04 manuscript/manuscript.Rmd
discussion paragraph 1

@@
 ---
-title: "Combinig geological and ecological proxies to model latitudinal temperature gradients in the early Eocene greenhouse"
+title: "Bayesian multi-proxy reconstruction of the early Eocene latitudinal temperature gradient"
 
@@
 <!-- discarded titles:
+Combining geochemical and ecological proxies...
 Modelling latitudinal temperature gradients from geological proxies: An early Eocene example
@@
 
-Spatially resolved climate reconstructions provide an alternative that can be employed more readily by a broad range of users.
+Spatially resolved climate reconstructions provide an alternative that can be employed more readily by a broad range of workers.
 Examples of this approach include the visualisation of proxy compilations along a latitudinal gradient [@Hollis2019, @Vickers2021].
@@
 
-1) *Avicennia*-only assemblages: the absence of Rhizophoraceae is indicative of temperatures being between 15.6 degC (lower temperature limt of *Avicennia*) and 20.7 (lower temperature limit of *Rhizophora*).
+1) *Avicennia*-only assemblages ($n=2$): the absence of Rhizophoraceae is indicative of temperatures being between 15.6 degC (lower temperature limt of *Avicennia*) and 20.7 (lower temperature limit of *Rhizophora*).
 However, a value of 22.5 degC is ascribed as the upper temperature limit here as *Rhizophora* is rare below this temperature.
@@
 
-2) *Avicennia* and Rhizophoraceae assemblages: the presence of both groups suggests that the locality should have a minimum temperature of 20.7 degC (lower temperature limit of *Rhizophora*).
+2) *Avicennia* and Rhizophoraceae assemblages ($n=5$): the presence of both groups suggests that the locality should have a minimum temperature of 20.7 degC (lower temperature limit of *Rhizophora*).
 As the upper thermal limits of *Aviciennia* and *Rhizophora* are not well established in @Quisthoudt2012, we assign the same maximum temperature limits (29.5 degC) as coral reef localities as mangroves are also widely distributed throughout tropical regions.
@@
 3a).
-Degrading the amount and distribution of data points to match the early Eocene sampling distribution still results in a close match of the modelled gradient with the empirical gradient (Fig 3b).
+Degrading the amount and distribution of data points to match the early Eocene sampling distribution ($n = 34$) still results in a close match of the modelled gradient with the empirical gradient (Fig 3b).
 A noticable, but minor offset is apparent in high latitudes, where sampling is most incomplete (Fig 3b).
@@
 This is reflected in the residual standard deviation ($\sigma$) of the early Eocene gradient, `r eeco_sdy$"50%"` (95% CI: `r eeco_sdy$"2.5%"` - `r eeco_sdy$"97.5%"`) degC, more than twice as high than  $\sigma$ of the modern gradient, resampled at early Eocene latitudes, `r modern_eeco_sample_sdy$"50%"` (95% CI: `r modern_eeco_sample_sdy$"2.5%"` - `r modern_eeco_sample_sdy$"97.5%"`) degC.
+The inclusion of ecological proxy data only slightly reduces the overall model standard deviation, but helps constrain polar temperature estimates (Supplementary materials).
 
@@
 # Discussion
+
+**Comparison to existing methods.** Latitudinal temperature gradients have been estimated with a variety of parametric and non-parametric statistical techniques.
+Parametric methods include polynomials [@Greenwood1995b; @Bij2009] and a cosine function [@Inglis2020].
+Whilst some of those functions are flexible enough to model a range of empirical temperature gradients, the logistic function used herein has the added advantage of an easily understandable relationship between each of the model parameters and the shape of the latitudinal gradient (Fig.
+2).
+Non-parametric models can very flexibly follow the shape of an unknown temperature gradient [@Jones2022, but see @Zhang2019], but have the disadvantage of having no model parameters that can be compared between data sets.
+Parametric methods should therefore be preferred to non-parametric temperature gradient models, as long as very broad prior information on the shape of the gradients exists.
+The ease with which such prior information can be integrated is perhaps the major advantage of our method: The Bayesian framework allows to specify arbitrary priors on the four parameters that control the upper and lower temperature limits, the steepness of the gradient, and the latitudinal position of the steepest point in the gradient (see Methods).
+A further improvement to existing methods is that the hierarchical structure of our model explicitly accounts for variation of temperature estimates at individual localities, and for uncertainty associated with the ecological temperature proxies.
+An extension of the model additionally accounts for measurement uncertainty of the geochemical proxies (Supplementary Materials).
+
+**Comparison to Eocene proxy and modelling studies**.
+
+A major theme of the discussions on the early Eocene greenhouse climate has been the discrepancy between proxy records, particularly ecological climate indicators such as macrofloral assemblage data, have
 

### 2022-12-04 R/analyses/assess_sdy.R

@@ -10,3 +10,8 @@ eeco_sdy <- as.list(format(round(quantile(mode_all$sdy,probs = c(.025,.5,.975)),
 
 saveRDS(modern_eeco_sample_sdy, "results/modern/modern_eeco_sample_sdy.rds")
 saveRDS(eeco_sdy, "results/eeco/eeco_sdy.rds")
+
+# just proxy sdy
+eeco_prox <- readRDS("results/eeco/eeco_climate_model_output_just_proxy.rds")
+eeco_prox <- combine_posterior(eeco_prox, burnin = 100000)
+eeco_prox_sdy <- as.list(format(round(quantile(eeco_prox$sdy,probs = c(.025,.5,.975)),1),nsmall = 1))

### 2022-12-08 manuscript/manuscript.Rmd
update references
discussion changes

@@
 Some episodes, for example the early Eocene, are thought to have been significantly warmer than the modern [@Pross2012], potentially representing an analogue for extreme climate warming scenarios [@Burke2018].
-Reconstructions of palaeoclimate commonly use 1) geochemical proxy data, 2) ecological or lithological climate indicators, or 3) earth system modelling.
+Reconstructions of palaeoclimate commonly use geochemical proxy data, palaeobiological or lithological climate indicators, or earth system modelling.
 
@@
 @Liu2009].
-To generate regional or global average estimates, the mean of multiple local estimates is taken [e.g.
-@Veizer2015].
-This approach has the advantage of being transparent and easy to reproduce, modify or expand with new data, but is prone to sampling bias [@Jones2022].
-Ecological and lithological climate data provide an alternative method of reconstructing palaeotemperatures, using e.g.
+Palaeobiological and lithological climate data provide a different method of reconstructing palaeotemperatures, using e.g.
 leaf physiognomy and floral assemblages to derive palaeotemperature proxies [@Greenwood2003], or using lithological indicators to determine the range of palaeoclimatic belts [@Scotese2021].
-Despite some recent advances [@Chandra2021], ecological and lithological climate indicators are still rarely used in quantitative palaeoclimate reconstructions, particularly for the marine realm.
+To capture complex dependencies between proxies and climate data, machine learning approaches [@Chandra2021] or articifical neural networks [@Malmgren1997; @Lauchstedt2017] are sometimes used.
 Earth system models take a fundamentally different approach by modelling the climate-generating processes directly, reconstructing spatially resolved climate estimates for the entire globe [@???].
@@
 
-Spatially resolved climate reconstructions provide an alternative that can be employed more readily by a broad range of workers.
+To make inferences on a regional or global scale, proxy-based climate reconstructions need some way of extrapolating a generally sparse record of local findings.
+Taking the mean of multiple local estimates to summarise regional or global climate variables [e.g.
+@Veizer2015] is a straightforward approach, but is likely to result in biased climate reconstructions [@Jones2022].
+Reconstructions that take into account the spatial distribution of climate data are...
+### HERE###
+provide an alternative that can be employed more readily by a broad range of workers.
 Examples of this approach include the visualisation of proxy compilations along a latitudinal gradient [@Hollis2019, @Vickers2021].
@@
 **Comparison to existing methods.** Latitudinal temperature gradients have been estimated with a variety of parametric and non-parametric statistical techniques.
-Parametric methods include polynomials [@Greenwood1995b; @Bij2009] and a cosine function [@Inglis2020].
+Parametric methods include polynomials [@Greenwood1995; @Bijl2009] and a cosine function [@Inglis2020].
 Whilst some of those functions are flexible enough to model a range of empirical temperature gradients, the logistic function used herein has the added advantage of an easily understandable relationship between each of the model parameters and the shape of the latitudinal gradient (Fig.

### 2023-01-15 manuscript/manuscript.Rmd
move Bayesian model description to SM

@@
 ## Bayesian framework
-
-**Model structure (Fig 2).** We model the mean temperature ($\mu$) at location $j$ as a function of absolute latitude ($abs(l)$) with a logistic regression (`growth curve` or `Richard's curve`) of the form:
-
-$$\mu_{j}  \sim N(\nu_j, \sigma), $$
-
+We model the relationship between sea surface temperature and latitude with a generalised logistic regression (`growth curve` or `Richard's curve`).
+Specifically, we modelled the estimate of the mean temperature $\mu_{j}$ at location $j$ as  $\mu_{j}  \sim N(\nu_j, \sigma)$, with
 $$\nu_{j} = A~+\frac{K-A}{e^{B(abs(l_j)-M)}}~, ~~~~~ j = 1,...,n,  $$
@@
 2), $\sigma$ denotes the residual standard deviation, and $n$ denotes the number of locations.
-
-We infer $\mu_j$ from $m$ individual temperature observations $t_{i = 1,...m}$, derived from geochemical data, at location $j$ as
-
-$$t_{i,j}  \sim N(\mu_j, \sigma_j), ~~~~~ i = 1,...,m,  $$
-where $m$ is the number of observations at each location, and $\sigma_j$ is the estimated standard deviation of the temperatures at location $j$.
-
-Similarly, $\mu_j$ is inferred for locations with ecological proxies from the associated normal temperature distributions with a given mean and standard deviation, $t_{\mu,j}$ and $t_{\sigma,j}$, as
-
-$$t_{\mu,j}  \sim N(\mu_j,t_{\sigma,j}).  $$
-
-This structure implies that $\mu_j$ is not fixed at the mean proxy temperature at location $j$, <!--of the temperatures $t_{i=1,...,m,~j}$ of location $j$-->, but is drawn towards the overall logistic regression curve, i.e.
-towards $\nu_j$.
-The pull towards $\nu_j$ tends to be high when $m$ is low, when the observations $t_{i=1,...,m,~j}$ are scattered, i.e.
-$\sigma_j$ is high, and/or when the overall standard deviation $\sigma$ is low.
-In practice, this has the desirable consequence that locations with few observations and large temperature differences between observations have less influence on the overall regression than well-sampled locations with consistent temperature observations.
-
-**Priors (Fig2x).** In the Bayesian framework, priors need to be placed on the unknown parameters of a model.
-We placed weakly informative, conjugate inverse-gamma priors on $\sigma$ and $\sigma_{j=1,..n}$:
-$$\sigma  \sim \sqrt{Inv-Gamma(\alpha+\frac{n}{2},\beta + 0.5\times(\mu_j-\nu_j))}, ~~~~~ j = 1,...,n,  $$
-$$\sigma_j  \sim \sqrt{Inv-Gamma(\alpha+\frac{m}{2},\beta + 0.5\times(t_{i,j}-\mu_j))}, ~~~~~ i = 1,...,m,~~~~~j = 1,...,n. $$
-We set $\alpha = \beta = 1$, allowing these priors to be quickly overwhelmed by the data as $n$ and $m$ increase, as we have little *a priori* knowledge of these parameters.
-
-In contrast, we put informative priors on the regression coefficients $A$, $K$, $M$ and $B$, based on physical principles, and vaguely based on the modern climate system:
-
-**A.** Predicted seawater surface temperatures are not allowed to be $<< -2^\circ C$, the freezing point of sea water.
-The highest prior density of $A$ is placed around $0^\circ C$, and it slowly tapers off towards higher temperatures.
-This shape is achieved by placing a skew-normal prior on the lower asymptote, specified as
-$$A \sim SN(\xi=-3.0,\omega=12,\alpha_{SN}=30),$$
-where $\xi$, $\omega$, and $\alpha_{SN}$ are the location, scale and shape parameters.
-
-**K.** Input of solar energy decreases from the tropics to the poles.
-Hence, the latitudinal temperature gradient is broadly negative, i.e.
-temperature decreases with absolute latitude.
-This is achieved by setting $K \ge A$.
-The prior on the upper asymptote $K$ is a truncated normal distribution with the mean set to $K$ of the modern SST gradient, with a broad standard deviation:
-$$K \sim TN(\mu_{TN} = 28, \sigma_{TN} = 10, \alpha_{TN} = A, \beta_{TN}= \infty)$$
-The distribution is truncated to the left at $\alpha_{TN}=A$, but not truncated to the right ($\beta_{TN}$).
-
-**M.** The steepness of the gradient is presumed to be highest in mid-latitudes; this is expressed with a normal prior on $M$ with the mean set to $M$ of the modern SST gradient:
-$$M \sim N(42,10)$$
-**B.** The steepness or growth rate $B$ of the gradient is constrained to be $\ge0$ and to not be exceedingly high, as oceanic and atmospheric heat transfer is bound to prevent very abrupt SST changes across latitudes.
-A gamma-distributed prior of the form
-$$B \sim Gamma(\alpha_G=4.3,\beta_G=30)$$
-was placed on $B$. The shape and rate parameters $\alpha_G$ and $\beta_G$ were chosen such that the highest prior density is at $B$ of the modern SST gradient, $0.11$.
+The hierarchical model structure implies that $\mu_{j}$ reflects the temperature proxies at location $j$, but is also affected by the fit to the overall latitudinal gradient.
+We placed priors on the unknown parameters $A$,$K$,$M$, and $B$ that were vaguely based on the modern, and observing general physical principles.
+For details on the model structure and the choice of priors, see the *Supplementary Material*.
 
@@
 ## Model validation
-
-[Fig. 3]
-<!--
-- Modern data
-- Sample based on Eocene localities
--->
-
 To test whether our logistic regression model can adequately describe latitudinal temperature gradients, we applied a simplified version of the model to modern sea surface temperatures from Bio-Oracle [@...].

### 2023-01-15 manuscript/supplementary_material.Rmd
move Bayesian model description to SM

---
title: "Supplementary material for  \nBayesian multi-proxy reconstruction of the early Eocene latitudinal temperature gradient"
author:
  - Kilian Eichenseer^[Department of Earth Sciences, Durham University, South Road, DH1 3LE, Durham, United Kingdom] and Lewis A. Jones^[Centro de Investigación Mariña, Grupo de Ecoloxía Animal, Universidade de Vigo, 36310 Vigo, Spain]
output:
  word_document:
    reference_docx: "template.docx"

bibliography: references.bib
csl: palaeontology.csl

---
## Hierarchical, Bayesian model for latitudinal temperature gradients

**Model structure (Fig 2).** We model the mean temperature ($\mu$) at location $j$ as a function of absolute latitude ($abs(l)$) with a logistic regression (`growth curve` or `Richard's curve`) of the form:

$$\mu_{j}  \sim N(\nu_j, \sigma), $$

$$\nu_{j} = A~+\frac{K-A}{e^{B(abs(l_j)-M)}}~, ~~~~~ j = 1,...,n,  $$
where $A$ and $K$ denote the lower and upper asymptote, respectively, $M$ specifies the latitude of maximal growth, the most quickly with latitude, $B$ denotes the growth rate (Fig. 2), $\sigma$ denotes the residual standard deviation, and $n$ denotes the number of locations.

We infer $\mu_j$ from $m$ individual temperature observations $t_{i = 1,...m}$, derived from geochemical data, at location $j$ as

$$t_{i,j}  \sim N(\mu_j, \sigma_j), ~~~~~ i = 1,...,m,  $$
where $m$ is the number of observations at each location, and $\sigma_j$ is the estimated standard deviation of the temperatures at location $j$.

Similarly, $\mu_j$ is inferred for locations with ecological proxies from the associated normal temperature distributions with a given mean and standard deviation, $t_{\mu,j}$ and $t_{\sigma,j}$, as

$$t_{\mu,j}  \sim N(\mu_j,t_{\sigma,j}).  $$

This structure implies that $\mu_j$ is not fixed at the mean proxy temperature at location $j$, <!--of the temperatures $t_{i=1,...,m,~j}$ of location $j$-->, but is drawn towards the overall logistic regression curve, i.e. towards $\nu_j$. The pull towards $\nu_j$ tends to be high when $m$ is low, when the observations $t_{i=1,...,m,~j}$ are scattered, i.e. $\sigma_j$ is high, and/or when the overall standard deviation $\sigma$ is low. In practice, this has the desirable consequence that locations with few observations and large temperature differences between observations have less influence on the overall regression than well-sampled locations with consistent temperature observations.

**Priors (Fig2x).** In the Bayesian framework, priors need to be placed on the unknown parameters of a model. We placed weakly informative, conjugate inverse-gamma priors on $\sigma$ and $\sigma_{j=1,..n}$:
$$\sigma  \sim \sqrt{Inv-Gamma(\alpha+\frac{n}{2},\beta + 0.5\times(\mu_j-\nu_j))}, ~~~~~ j = 1,...,n,  $$
$$\sigma_j  \sim \sqrt{Inv-Gamma(\alpha+\frac{m}{2},\beta + 0.5\times(t_{i,j}-\mu_j))}, ~~~~~ i = 1,...,m,~~~~~j = 1,...,n. $$
We set $\alpha = \beta = 1$, allowing these priors to be quickly overwhelmed by the data as $n$ and $m$ increase, as we have little *a priori* knowledge of these parameters.

In contrast, we put informative priors on the regression coefficients $A$, $K$, $M$ and $B$, based on physical principles, and vaguely based on the modern climate system:

**A.** Predicted seawater surface temperatures are not allowed to be $<< -2^\circ C$, the freezing point of sea water. The highest prior density of $A$ is placed around $0^\circ C$, and it slowly tapers off towards higher temperatures. This shape is achieved by placing a skew-normal prior on the lower asymptote, specified as
$$A \sim SN(\xi=-3.0,\omega=12,\alpha_{SN}=30),$$
where $\xi$, $\omega$, and $\alpha_{SN}$ are the location, scale and shape parameters.

**K.** Input of solar energy decreases from the tropics to the poles. Hence, the latitudinal temperature gradient is broadly negative, i.e. temperature decreases with absolute latitude. This is achieved by setting $K \ge A$. The prior on the upper asymptote $K$ is a truncated normal distribution with the mean set to $K$ of the modern SST gradient, with a broad standard deviation:
$$K \sim TN(\mu_{TN} = 28, \sigma_{TN} = 10, \alpha_{TN} = A, \beta_{TN}= \infty)$$
The distribution is truncated to the left at $\alpha_{TN}=A$, but not truncated to the right ($\beta_{TN}$).

**M.** The steepness of the gradient is presumed to be highest in mid-latitudes; this is expressed with a normal prior on $M$ with the mean set to $M$ of the modern SST gradient:
$$M \sim N(42,10)$$
**B.** The steepness or growth rate $B$ of the gradient is constrained to be $\ge0$ and to not be exceedingly high, as oceanic and atmospheric heat transfer is bound to prevent very abrupt SST changes across latitudes. A gamma-distributed prior of the form
$$B \sim Gamma(\alpha_G=4.3,\beta_G=30)$$
was placed on $B$. The shape and rate parameters $\alpha_G$ and $\beta_G$ were chosen such that the highest prior density is at $B$ of the modern SST gradient, $0.11$.

# References

### 2023-01-15 manuscript/manuscript.Rmd
disussion change

@@
 
-**Comparison to existing methods.** Latitudinal temperature gradients have been estimated with a variety of parametric and non-parametric statistical techniques.
+**Eocene temperature gradients.**
+
+**Methodological advances.** Latitudinal temperature gradients have been estimated with a variety of parametric and non-parametric statistical techniques.
 Parametric methods include polynomials [@Greenwood1995; @Bijl2009] and a cosine function [@Inglis2020].

### 2023-01-21 manuscript/discards.rmd
put full methods back in

--- 2023-01-15 manuscript/manuscript.Rmd
@@
 ---
-title: "Bayesian multi-proxy reconstruction of the early Eocene latitudinal temperature gradient"
-
-author:
-  - Kilian Eichenseer^[Department of Earth Sciences, Durham University, South Road, DH1 3LE, Durham, United Kingdom] and Lewis A.
-Jones^[Centro de Investigación Mariña, Grupo de Ecoloxía Animal, Universidade de Vigo, 36310 Vigo, Spain]
-output:
-  word_document:
-    reference_docx: "template.docx"
-
-bibliography: references.bib
-csl: palaeontology.csl
-
+title: "Discards"
+output: html_document
+date: "2023-01-21"
 ---
 
-<!-- discarded titles:
-Combining geochemical and ecological proxies...
-Modelling latitudinal temperature gradients from geological proxies: An early Eocene example
-"Improving palaeoclimatic reconstructions with ecological data: How hot was the early Eocene?"
-"Integrating ecological and geochemical proxies for palaeoclimatic reconstruction: An early Eocene example"
-"The Early Eocene Climate Optimum reconstructed with Bayesian multi-proxy integration"
-"An Early Eocene temperature gradient modelled with geological and ecological proxies"
--->
-
-Abstract
-===============================================================================
-Geochemical data such as oxygen isotopes (d18O), tetraether indices (TEX86) and magnesium calcium ratios (Mg/Ca) are routinely used to reconstruct ocean temperatures in deep time.
-However, limited spatial coverage and disagreements between geochemical proxies compromise the accuracy of these reconstructions.
-Geological data such as coral reefs, mangroves, and evaporites have the potential to supplement geochemical data by improving spatial coverage and constraining temperature estimates.
-Historically, these data have been used to qualitatively inform upon palaeoclimatic conditions at broad spatial scales, yet no transfer functions exist to convert geological data into temperature estimates.
-Quantitative temperature reconstructions have therefore not made use of these data.
-Here, we present a Bayesian hierarchical model to integrate geological data–with established temperature proxies–into a unified quantitative framework.
-We apply this approach to the early Eocene climatic optimum (EECO), the interval with the warmest sustained temperatures of the Cenozoic.
-Assuming the conservation of thermal tolerances of modern coral reefs and mangrove taxa, we establish broad sea surface temperature ranges for EECO coral reef and mangrove sites.
-We integrate these temperature estimates with the EECO geochemical shallow marine proxy record to model the latitudinal temperature gradient and its uncertainty.
-Our results confirm the presence of a flattened latitudinal temperature gradient and unusually high polar temperatures during the EECO.
-We show that the inclusion of ecological data can substantially reduce the uncertainty on temperature estimates in climate zones lacking geochemical data.
-
-# Introduction
-The geological record holds evidence for episodes in Earth history with fundamentally different climates than our current [@Scotese2021].
-Some episodes, for example the early Eocene, are thought to have been significantly warmer than the modern [@Pross2012], potentially representing an analogue for extreme climate warming scenarios [@Burke2018].
-Reconstructions of palaeoclimate commonly use geochemical proxy data, palaeobiological or lithological climate indicators, or earth system modelling.
-
-Geochemical proxies such as oxygen isotopes (d18O), tetraether indices (TEX86) and magnesium calcium ratios (Mg/Ca) are used to derive palaeotemperature estimates for individual locations using experimentally derived transfer functions [e.g.
-@Liu2009].
-Palaeobiological and lithological climate data provide a different method of reconstructing palaeotemperatures, using e.g.
-leaf physiognomy and floral assemblages to derive palaeotemperature proxies [@Greenwood2003], or using lithological indicators to determine the range of palaeoclimatic belts [@Scotese2021].
-To capture complex dependencies between proxies and climate data, machine learning approaches [@Chandra2021] or articifical neural networks [@Malmgren1997; @Lauchstedt2017] are sometimes used.
-Earth system models take a fundamentally different approach by modelling the climate-generating processes directly, reconstructing spatially resolved climate estimates for the entire globe [@???].
-However, this approach still relies on proxy data to constrain the large number of possible model setups and ground truth the model output [@???] It also requires substantial computing resources and technical expert knowledge, and is thus difficult to implement or modify for most palaeoclimate researchers.
-
-To make inferences on a regional or global scale, proxy-based climate reconstructions need some way of extrapolating a generally sparse record of local findings.
-Taking the mean of multiple local estimates to summarise regional or global climate variables [e.g.
-@Veizer2015] is a straightforward approach, but is likely to result in biased climate reconstructions [@Jones2022].
-Reconstructions that take into account the spatial distribution of climate data are...
-### HERE###
-provide an alternative that can be employed more readily by a broad range of workers.
-Examples of this approach include the visualisation of proxy compilations along a latitudinal gradient [@Hollis2019, @Vickers2021].
-Interpolation is sometimes used to bridge spatial gaps in palaeoclimate data [@Taylor2004], taking advantage of the autoregressive nature of climatic data: much of the information on the climate of any given location is contained in the climate data of nearby locations [@Reynolds1994].
-Adding to this, some proxy-based reconstructions use statistical modelling to infer climatic patterns.
-For example, polynomial regression has been used to reconstruct latitudinal temperature gradients [@Bijl2009], and 2D-reconstructions of surface temperatures have been created with Gaussian process regression [@Inglis2020].
-
-Here, we present a Bayesian, hierarchical model for inferring sea surface temperatures (SST) that expands upon existing, spatial reconstructions of palaeoclimate by allowing for the integration of 1) prior information based on physical principles and the observed, modern SST distribution, and of 2) geochemical and ecological climate proxies in a common, quantitative framework.
-We apply this approach to the early Eocene climatic optimum (EECO), the interval with the warmest sustained temperatures of the Cenozoic.
-Assuming the conservation of thermal tolerances of modern coral reefs and mangrove taxa, we establish broad temperature ranges for EECO coral reef and mangrove sites.
-We integrate these temperature estimates with the EECO geochemical, shallow marine proxy record to model the latitudinal SST gradient and its uncertainty.
-
-<!--
-
-Hook -> Generic problem:
-- Reconstructing past climate important for understanding current/future global warming
-- Two main pathways: highly complex ESM <-> proxy compilations
-- Problems with proxy data: only provides a local picture (few local sites interpreted as global signal)
-- Problems with ESM (long time to run, very complex, a lot of assumptions on processes/doesn't consider how they may change, need to be constrained with proxy data)
-- Lack of intermediate-complexity models to make full use of the proxy data
-
-Specific problem:
-- exemplified with the EECO: proxies and ESM hard to reconcile, EECO important climate analogue
-- a lot of proxies recrystallised <-> unreliable data or lack of data
-
-Solution:
-- Proxy-driven model (no strong assumptions on processes)
-- Integrate more data (quantify biological / geological information)
-- Integrate additional knowledge (constraints on such gradients)
-
-Application:
-- EECO time of extreme warming
-- problem of the Early Eocene:
-- apply solution to EECO to constrain EECO lat. T gradient
-
-Here:
-
-- EECO specific information
-- Disagreement between GCM + geochem data
--->
-
-<!--
-
-The early Eocene climate optimum (EECO, ~ 53--49 million years ago) is thought to have been the warmest episode of [...].
-The EECO is of particular interest due to its use as an analogue for extreme warming scenarios, depicting a possible future high-CO2 world.
-To reconstruct the climate of EECO, a variety of proxy data and Earth System Models have been used.
-Early modelers have soon noticed that their models struggle to reproduce the climate indicated by proxy data: Without unrealistically high concentrations of atmospheric CO2, early climate models were unable to reproduce the equitable polar climates indicated by early Eocene proxy data, e.g.
-the occurrence of mangroves in polar latitudes, and d18O data from x indicates a temperature of ~ y at latitude z.
-
-Contrast between complex, process-driven Earth system models and very simplistic use of the proxy data as local temperature estimates or global averages.
-Here, we introduce a proxy-driven modelling approach of relatively low complexity, which places the available chemical and biological proxy data in a single climate model.
-In order to properly compare model output to proxy data, latitudinal temperature patterns need to be reconstructed from noisy data (impossible to properly compare model gradients with scattered proxy points "by eye")
-
-Reconstructing Earth's climate history is fundamental for understanding the evolution of the Earth system and its consequences for life in the past and future.
-The geological record, the only empirical source of information on the palaeoclimate of the deep past, holds evidence.
-
-Here, we present a Bayesian hierarchical model to integrate geological data---with established temperature proxies---into a unified quantitative framework.
-We apply this approach to the early Eocene climatic optimum (EECO), the interval with the warmest sustained temperatures of the Cenozoic.
-Assuming the conservation of thermal tolerances of modern coral reefs and mangrove taxa, we establish broad sea surface temperature ranges for EECO coral reef and mangrove sites.
-We integrate these temperature estimates with the EECO geochemical shallow marine proxy record to model the latitudinal temperature gradient and its uncertainty.
-
-Our hierarchical approach accounts for different sample sizes between and uncertainties within proxy sites, and
-
--->
-
-# Materials & Methods
-
-## Geochemical data
-
-Geochemical climate proxy data were extracted from a latest Paleocene and early Eocene compilation [@hollis2019].
-This compilation provides four different geochemical proxies for reconstructing seawater temperature: $\delta^18^O, \Delta$_47_, Mg/Ca and TEX_86_.
-For our analyses, this dataset was restricted to the EECO and samples from the continental shelf.
-Recrystallised $\delta^18^O samples were also excluded as secondary diagenetic calcite precipitated after deposition can bias isotope measurements and offset temperature values [@schrag1999].
-This resulted in most $\delta^18^O samples being excluded from the dataset (x out of x).
-After data filtering, x geochemical proxy samples remained.
-For a detailed description of each proxy see [@hollis2019].
-
-## Ecological data
-
-**Coral reefs.** Today, shallow warm-water coral reefs are limited to tropical and subtropical latitudes (~34°N–32 degS), with minimum sea surface temperature (~18 degC) tolerances being the primary constraint on this distribution [@johannes1983; @kleypas1999; @yamano2001].
-As coral reefs reside at the upper thermal limit of the oceans today, their maximum sea surface temperature tolerance is less well-constrained with some studies suggesting up to 35.6 degC in the geological past [@jones2022reef].
-Nevertheless, the distribution of coral reefs have frequently been recognised as tracers of past (sub-)tropical conditions [@ziegler1984; @kiessling2001].
-During the Eocene, coral communities and reefs expanded across tropical and temperate latitudes, with some communities found up to palaeolatitudes of 43 degC N [@zamagni2012].
-Using a compilation of Paleocene--early Eocene coral reefs and community localities [@zamagni2012], we generate quantitative sea surface temperature estimates for the ECCO.
-To do so, we extract localities inferred to be Ilerdian (early Eocene) coral reefs from the compilation, and that can be confidently assigned to the EECO.
-We exclude coral knobs and coral-bearing mounds which might have broader climatic limits than coral reef ecosystems.
-This filtering resulted in four unique coral reef localities remaining for the EECO, all of which conform to the modern latitudinal range of coral reefs (<34 deg N).
-Subsequently, we use statistically derived temperature limits (minimum = 21 degC, average = 27.6 degC, maximum = 29.5 degC) from the published literature [@kleypas1999] to define a normal probability distribution of potential temperature values for coral reef localities.
-This normal probability distribution was defined with a mean of 27.6 and a standard deviation of 2.125, placing 97.5% of the probability density above the minimum.
-As the distribution of modern corals is skewed towards warmer temperatures, this approach results in 16.5% of the probability being placed on temperatures > 29.5 degC, allowing for the possibility that Eocene coral reefs were adapted to warmer conditions than modern-day coral reefs.
-
-**Mangroves.** Mangroves are distributed throughout the tropics and subtropics today.
-While factors beside SST influence the distribution of mangroves, empirical, lower temperature limits have been established for the genera *Avicennia* (15.6 degC) and *Rhizophora* (20.7 degC) [@Quisthoudt2012].
-Both *Avicennia* and members of the Rhizophoraceae family were widespread in the early Eocene, but only *Avicennia* occurred at polar latitudes [@Suan2017; @Popescu2021].
-Assuming that Eocene members of these mangrove taxa conform to the same climatic requirements as their modern relatives, the presence and absence of *Avicennia* and Rhizophoraceae pollen can be used as a palaeotemperature indicator.
-For this analysis, published mangrove occurrence data were taken from @Popescu2021, and converted to quantitative temperature estimates.
-From this data, we identify two types of pollen assemblages which we ascribe different temperature distributions:
-
-1) *Avicennia*-only assemblages ($n=2$): the absence of Rhizophoraceae is indicative of temperatures being between 15.6 degC (lower temperature limt of *Avicennia*) and 20.7 (lower temperature limit of *Rhizophora*).
-However, a value of 22.5 degC is ascribed as the upper temperature limit here as *Rhizophora* is rare below this temperature.
-We define the *Avicennia*-only temperature distribution as a normal distribution with a mean of 19.05 degC and a standard deviation of 1.725, resulting in 95% of the probability density being placed within the temperature limits.
-
-2) *Avicennia* and Rhizophoraceae assemblages ($n=5$): the presence of both groups suggests that the locality should have a minimum temperature of 20.7 degC (lower temperature limit of *Rhizophora*).
-As the upper thermal limits of *Aviciennia* and *Rhizophora* are not well established in @Quisthoudt2012, we assign the same maximum temperature limits (29.5 degC) as coral reef localities as mangroves are also widely distributed throughout tropical regions.
-Consequently, we define the temperature distribution for this locality as a normal distribution with a mean of 25.1 and a standard deviation of 2.2, with 95% probability density within the temperature limits.
-
-[Fig 1: Map and distribution of proxy data and ecological data]
-
-## Palaeogeographic reconstruction
-
-The palaeogeographic distribution of geochemical and ecological data was reconstructed using the @merdith2021 plate rotation model via the palaeoverse R package ver.
-XXX (REF).
-The midpoint age of the EECO (51.2 Ma), along with the present-day coordinates of geochemical and ecological data, were used for palaeorotation.
-
-## Bayesian framework
+## Condensed methods Bayesian framework
 We model the relationship between sea surface temperature and latitude with a generalised logistic regression (`growth curve` or `Richard's curve`).
@@
 For details on the model structure and the choice of priors, see the *Supplementary Material*.
-
-[Fig. 2: Model structure and visualisation of parameters and priors]
-
-## Model validation
-To test whether our logistic regression model can adequately describe latitudinal temperature gradients, we applied a simplified version of the model to modern sea surface temperatures from Bio-Oracle [@...].
-As these modern, annual sea surface temperatures estimates are associated with little uncertainty, there is no need for a hierarchical model structure, and we substituted temperature ($t$) for $\mu$ in eq.
-?
-and ?.
-To verify that the modern gradient can be approximated with limited sampling, we resampled modern sea surface temperatures at modern latitudes corresponding the palaeolatitudes of the Eocene samples.
-This process was repeated 100 times, randomly chosing a longitude for each latitude in each repetition.
-The parameters of the non-hierarchical temperature model were estimated for each of the 100 samples, and the iterations after burn-in from the posterior of the parameters were pooled to generate the resulting median temperature gradient with 95 % credible intervals ($2.5^{th}$ to $97.5^{th}$ percentile).
-
-## Parameter estimation
-We estimated the posterior distributions of the model parameters using a Markov chain Monte Carlo (MCMC) algorithm, written in R.
-Specifically, we sampled the unknown parameters $A$, $K$, $M$ and $B$ with Metropolis-Hastings, and used Gibbs sampling to estimate all other unknown parameters.
-Posterior inference on the modern gradient is based on four chains with 60,000 iterations each, 10,000 of which were discarded as burn-in, and keeping every 10th iteration, resulting in a total of 20,000 iterations.
-The modern temperature gradient with the Eocene sampling distribution was estimated in one chain with 25,000 iterations for each of the 100 temperature samples.
-5,000 iterations each were discarded as burn-in, and every 10th iteration was kept, resulting in a total of 200,000 iterations across all 100 model runs.
-For the Eocene model, we ran four chains with 600,000 iterations each, discarding 100,000 as burn-in and keeping every 100th iteration, as the hierarchical model structure results in higher autocorrelation of the chains.
-The Eocene posterior inference is thus based on a total of 20,000 iterations with low autocorrelation (effective multivariate sample size for $A$, $K$, $M$ and $B$ is > 18,000).
-
-## Processing of model results
-<!--
-- Calculate lat gradient
-- Calculate global T
-[Fig. 4]
--->
-Modelled sea surface temperature estimates were generated with eq.
-?, calculating the sea surface temperatures at any latitude with the parameter estimates of each iteration from the posterior.
-The median, $2.5^{th}$ and $97.5^{th}$ percentile of temperatures where then taken from all temperature estimates obtained at the latitude of interest.
-
-Differences between Eocene and modern temperatures at a certain latitude were calculated by randomly pairing all iterations of the posterior from the Eocene and modern temperature gradient model, calculating the Eocene and modern temperature using the respective iterations, taking the difference, and then calculating the median ($2.5^{th}$, $97.5^{th}$ percentile) from all pairs of iterations.
-
-Global average temperatures were calculated by taking the weighted mean of the median ($2.5^{th}$, $97.5^{th}$ percentile) temperature estimates in 1 degree latitudinal bins.
-The weights were set to the proportion of global surface area in each latitudinal bin, i.e.
-decreasing with increasing latitude as:
-$$weights = sin(\alpha_{1,i}) - sin(\alpha_{2,i}) ,$$
-where $\alpha1$ is the upper, and $\alpha2$ is the lower latitudinal boundary of bin $i$.
-
-# Results
-
-## Model validation
-<!-- - Show that the 4-parameter logistic function can adequatly describe the modern temperature gradient
-- Show that the modern gradient can be reproduced with Eocene sampling distribution
--->
-[Fig 3 - Modern gradient]
-The average modern temperature gradient can be closely approximated with our model when using the full modern SST dataset (Fig.
-3a).
-Degrading the amount and distribution of data points to match the early Eocene sampling distribution ($n = 34$) still results in a close match of the modelled gradient with the empirical gradient (Fig 3b).
-A noticable, but minor offset is apparent in high latitudes, where sampling is most incomplete (Fig 3b).
-
-## EECO reconstruction
-<!-- - Show Eocene gradient
-- Global t : use raster package for calculation of areal weights
-- report lat gradient
--->
-```{r results block - for in-line code, echo = FALSE}
-setwd('..') # move up to parent folder of /manuscript
-
-# temperature gradients
-tdiff_lat0 <- readRDS("results/eeco/eocene_modern_difference_equator.rds")
-tdiff_lat90 <- readRDS("results/eeco/eocene_modern_difference_poles.rds")
-eocene_gradient <- readRDS("results/eeco/eocene_overall_gradient.rds")
-modern_gradient <- readRDS("results/modern/modern_overall_gradient.rds")
-
-# sdy
-eeco_sdy <- readRDS("results/eeco/eeco_sdy.rds")
-modern_eeco_sample_sdy <- readRDS("results/modern/modern_eeco_sample_sdy.rds")
-
-# global average temperatures
-motemp_global <- readRDS("results/modern/global_mean_modern.rds")
-eotemp_global <- readRDS("results/eeco/global_mean_eeco.rds")
-
-```
-
-The modelled Eocene temperature gradient is starkly different from the modern (Fig 4).
-Modelled, median equatorial temperatures are `r tdiff_lat0$"50%"` (95% CI: `r tdiff_lat0$"2.5%"` - `r tdiff_lat0$"97.5%"`) degC higher for the Eocene, and polar temperatures are higher by `r tdiff_lat90$"50%"` (95% CI: `r tdiff_lat90$"2.5%"` - `r tdiff_lat90$"97.5%"`) degC.
-This results in a strongly flattened latitudinal temperature gradient of `r eocene_gradient$"50%"`  (95% CI: `r eocene_gradient$"2.5%"` - `r eocene_gradient$"97.5%"`) degC for the Eocene, as opposed to `r modern_gradient$"50%"` degC for the modern.
-
-The high variability of early Eocene palaeotemperature proxies, particularly in the mid-latitudes, and the scarcity of high-latitude data, results in substantial uncertainties in the modeled temperature gradient.
-This is reflected in the residual standard deviation ($\sigma$) of the early Eocene gradient, `r eeco_sdy$"50%"` (95% CI: `r eeco_sdy$"2.5%"` - `r eeco_sdy$"97.5%"`) degC, more than twice as high than  $\sigma$ of the modern gradient, resampled at early Eocene latitudes, `r modern_eeco_sample_sdy$"50%"` (95% CI: `r modern_eeco_sample_sdy$"2.5%"` - `r modern_eeco_sample_sdy$"97.5%"`) degC.
-The inclusion of ecological proxy data only slightly reduces the overall model standard deviation, but helps constrain polar temperature estimates (Supplementary materials).
-
-The global Eocene mean sea surface temperature is estimated at `r eotemp_global$"50%"` (95% CI: `r eotemp_global$"2.5%"` - `r eotemp_global$"97.5%"`) degC, significantly higher than the modern (`r motemp_global$"50%"` degC).
-
-## Fig 4 - compare Eocene gradient with modern gradient, also with previous reconstructions / ES models?
-
-- x1 Draw samples from the posterior gradient to get a latitudinal gradient of deg C / deg lat with uncertainty (intercept gradient at 0 and 90 deg lat)
-
-- x2 generate global average temperature, and tropical, temperate, polar temperatures in a similar way, accounting for area
-
-# Discussion
-
-**Comparison to existing methods.** Latitudinal temperature gradients have been estimated with a variety of parametric and non-parametric statistical techniques.
-Parametric methods include polynomials [@Greenwood1995; @Bijl2009] and a cosine function [@Inglis2020].
-Whilst some of those functions are flexible enough to model a range of empirical temperature gradients, the logistic function used herein has the added advantage of an easily understandable relationship between each of the model parameters and the shape of the latitudinal gradient (Fig.
-2).
-Non-parametric models can very flexibly follow the shape of an unknown temperature gradient [@Jones2022, but see @Zhang2019], but have the disadvantage of having no model parameters that can be compared between data sets.
-Parametric methods should therefore be preferred to non-parametric temperature gradient models, as long as very broad prior information on the shape of the gradients exists.
-The ease with which such prior information can be integrated is perhaps the major advantage of our method: The Bayesian framework allows to specify arbitrary priors on the four parameters that control the upper and lower temperature limits, the steepness of the gradient, and the latitudinal position of the steepest point in the gradient (see Methods).
-A further improvement to existing methods is that the hierarchical structure of our model explicitly accounts for variation of temperature estimates at individual localities, and for uncertainty associated with the ecological temperature proxies.
-An extension of the model additionally accounts for measurement uncertainty of the geochemical proxies (Supplementary Materials).
-
-**Comparison to Eocene proxy and modelling studies**.
-
-A major theme of the discussions on the early Eocene greenhouse climate has been the discrepancy between proxy records, particularly ecological climate indicators such as macrofloral assemblage data, have
-
-- x1 compare lat gradient to literature estimates, e.g. Zhang2019
-- compare lat gradient to climate models and the "early EECO problem"
-- discuss insufficiency of proxies (high variability, ...)
-- discuss utility of the model and future applications
-
-# Conclusions
-
-# Acknowledgements
-
-# Author contributions
-
-# Data accessibility
-
-# References

### 2023-01-21 manuscript/manuscript.Rmd
put full methods back in

--- 2022-12-08 manuscript/manuscript.Rmd
@@
 ---
-title: "Bayesian multi-proxy reconstruction of the early Eocene latitudinal temperature gradient"
+title: "Bayesian multi-proxy reconstruction of latitudinal temperature gradients for the Early Eocene"
 
@@
 <!-- discarded titles:
+Bayesian multi-proxy reconstruction of the early Eocene latitudinal temperature gradient
 Combining geochemical and ecological proxies...
@@
 ## Bayesian framework
-
 **Model structure (Fig 2).** We model the mean temperature ($\mu$) at location $j$ as a function of absolute latitude ($abs(l)$) with a logistic regression (`growth curve` or `Richard's curve`) of the form:

### 2023-01-23 manuscript/manuscript.Rmd
SM and results

@@
 The Eocene posterior inference is thus based on a total of 20,000 iterations with low autocorrelation (effective multivariate sample size for $A$, $K$, $M$ and $B$ is > 18,000).
+Trace plots were used to assess the convergence and mixing of the MCMC chains (Supplementary Materials S1).
 
@@
 # Results
-
-## Model validation
-<!-- - Show that the 4-parameter logistic function can adequatly describe the modern temperature gradient
-- Show that the modern gradient can be reproduced with Eocene sampling distribution
--->
-[Fig 3 - Modern gradient]
-The average modern temperature gradient can be closely approximated with our model when using the full modern SST dataset (Fig.
-3a).
-Degrading the amount and distribution of data points to match the early Eocene sampling distribution ($n = 34$) still results in a close match of the modelled gradient with the empirical gradient (Fig 3b).
-A noticable, but minor offset is apparent in high latitudes, where sampling is most incomplete (Fig 3b).
-
-## EECO reconstruction
-<!-- - Show Eocene gradient
-- Global t : use raster package for calculation of areal weights
-- report lat gradient
--->
 ```{r results block - for in-line code, echo = FALSE}
@@
 modern_gradient <- readRDS("results/modern/modern_overall_gradient.rds")
+modern_eocene_gradient <- readRDS("results/modern/modern_eocene_sampling_overall_gradient.rds")
+modern_eocene_gradient_diff <- readRDS("results/modern/modern_eocene_sampling_overall_gradient_diff.rds")
 
@@
 eotemp_global <- readRDS("results/eeco/global_mean_eeco.rds")
+moeotemp_global <- readRDS("results/eeco/global_mean_modern_eocene_sampling_eeco.rds")
+moeotemp_global_diff <- readRDS("results/eeco/global_mean_modern_eocene_sampling_difference.rds")
 
 ```
+## Model validation
+<!-- - Show that the 4-parameter logistic function can adequatly describe the modern temperature gradient
+- Show that the modern gradient can be reproduced with Eocene sampling distribution
+-->
+[Fig 3 - Modern gradient]
+The average modern temperature gradient can be closely approximated with our model when using the full modern SST dataset (Fig.
+3a).
+Reducing the amount and the spatial coverage of data points to match the early Eocene sampling distribution ($n = 34$) still results in a close match of the modeled gradient with the empirical gradient (Fig.
+3b).
+The 95% credible interval of reconstructed gradients (Fig.
+3b) is noticeably narrower than the spread of empirical temperature values (Fig.
+3a).
+The modern, latitudinal gradient, modeled with all data, is `r modern_gradient$"50%"` degC, and has a median of `r modern_eocene_gradient$"50%"`  (95% CI: `r modern_eocene_gradient$"2.5%"` - `r modern_eocene_gradient$"97.5%"`) degC when modelled with only the early Eocene sampling locations.
+A random sampling of modern temperatures at latitudes corresponding to the sampled Eocene palaeolatitudes would thus allow the reconstruction of the modern latitudinal temperature gradient with good accuracy, with expected deviations from the actual gradient as low as `r modern_eocene_gradient_diff$"2.5%"` to `r modern_eocene_gradient_diff$"97.5%"` degC (95 % CI).
+
+The modeled, latitudinal temperature gradient facilitates accurate reconstruction of global average temperatures.
+The modern, global mean sea surface temperature estimate is `r motemp_global$"50%"` degC when using the full, modern data set.
+When reduced to the early Eocene sampling distribution, the modern global mean sea surface temperature is estimated at `r moeotemp_global$"50%"` (95% CI: `r moeotemp_global$"2.5%"` - `r moeotemp_global$"97.5%"`) degC.
+The deviations from the full modern estimate are again very low, with a 95 % credible interval of `r moeotemp_global_diff$"2.5%"` to `r moeotemp_global_diff$"97.5%"` degC.
+
+## EECO reconstruction
+<!-- - Show Eocene gradient
+- Global t : use raster package for calculation of areal weights
+- report lat gradient
+-->
 
@@
 The inclusion of ecological proxy data only slightly reduces the overall model standard deviation, but helps constrain polar temperature estimates (Supplementary materials).
+### Continue here.
 

### 2023-01-23 manuscript/supplementary_material.Rmd
update SM
SM and results

@@
 ---
-## Hierarchical, Bayesian model for latitudinal temperature gradients
+```{r, setup, include=FALSE}
+#knitr::opts_knit$set(root.dir = '..')
+```
 
-**Model structure (Fig 2).** We model the mean temperature ($\mu$) at location $j$ as a function of absolute latitude ($abs(l)$) with a logistic regression (`growth curve` or `Richard's curve`) of the form:
+## S1. Convergence checks
+To ensure the convergence and mixing of the chains, we inspected traceplots of all the unknown parameters of the model.
 
-$$\mu_{j}  \sim N(\nu_j, \sigma), $$
+```{r out.width='100%', echo = FALSE, fig.cap = "Fig S1.
+Traceplots of a subset of the unknown model parameters.
+The four colours correspond to the four independent model runs.
+a) Traceplot of A, K-A, M, B and $\\sigma$; b) Traceplot of five selected $\\mu_j$; c) Traceplot of five selected $\\sigma_j$."}
+setwd('..') # move up to parent folder of /manuscript
+knitr::include_graphics("D://Github/PalaeoClimateGradient/figures/SM/FigS1-Traceplot.jpg", error = FALSE)
+```
 
-$$\nu_{j} = A~+\frac{K-A}{e^{B(abs(l_j)-M)}}~, ~~~~~ j = 1,...,n,  $$
-where $A$ and $K$ denote the lower and upper asymptote, respectively, $M$ specifies the latitude of maximal growth, the most quickly with latitude, $B$ denotes the growth rate (Fig.
-2), $\sigma$ denotes the residual standard deviation, and $n$ denotes the number of locations.
+## S2. Gradient with just the geochemical proxy data
 
-We infer $\mu_j$ from $m$ individual temperature observations $t_{i = 1,...m}$, derived from geochemical data, at location $j$ as
+```{r out.width='100%', echo = FALSE, fig.cap = "Fig S2.
+Latitudinal temperature gradient with just the geochemical proxies (orange), showing the median (line) and 95% credibal interval (shading).
+Symbols with vertical lines show the median and 95% credible intervals of $\\mu$.
+The grey line and shading in the background show the latitudinal temperature gradient with the geochemical and ecological proxy data."}
+setwd('..') # move up to parent folder of /manuscript
+knitr::include_graphics("D://Github/PalaeoClimateGradient/figures/SM/FigS2_geochem_eeco.png", error = FALSE)
+```
 
-$$t_{i,j}  \sim N(\mu_j, \sigma_j), ~~~~~ i = 1,...,m,  $$
-where $m$ is the number of observations at each location, and $\sigma_j$ is the estimated standard deviation of the temperatures at location $j$.
+## S3. Seperate gradients on the northen and southern hemisphere
 
-Similarly, $\mu_j$ is inferred for locations with ecological proxies from the associated normal temperature distributions with a given mean and standard deviation, $t_{\mu,j}$ and $t_{\sigma,j}$, as
-
-$$t_{\mu,j}  \sim N(\mu_j,t_{\sigma,j}).  $$
-
-This structure implies that $\mu_j$ is not fixed at the mean proxy temperature at location $j$, <!--of the temperatures $t_{i=1,...,m,~j}$ of location $j$-->, but is drawn towards the overall logistic regression curve, i.e.
-towards $\nu_j$.
-The pull towards $\nu_j$ tends to be high when $m$ is low, when the observations $t_{i=1,...,m,~j}$ are scattered, i.e.
-$\sigma_j$ is high, and/or when the overall standard deviation $\sigma$ is low.
-In practice, this has the desirable consequence that locations with few observations and large temperature differences between observations have less influence on the overall regression than well-sampled locations with consistent temperature observations.
-
-**Priors (Fig2x).** In the Bayesian framework, priors need to be placed on the unknown parameters of a model.
-We placed weakly informative, conjugate inverse-gamma priors on $\sigma$ and $\sigma_{j=1,..n}$:
-$$\sigma  \sim \sqrt{Inv-Gamma(\alpha+\frac{n}{2},\beta + 0.5\times(\mu_j-\nu_j))}, ~~~~~ j = 1,...,n,  $$
-$$\sigma_j  \sim \sqrt{Inv-Gamma(\alpha+\frac{m}{2},\beta + 0.5\times(t_{i,j}-\mu_j))}, ~~~~~ i = 1,...,m,~~~~~j = 1,...,n. $$
-We set $\alpha = \beta = 1$, allowing these priors to be quickly overwhelmed by the data as $n$ and $m$ increase, as we have little *a priori* knowledge of these parameters.
-
-In contrast, we put informative priors on the regression coefficients $A$, $K$, $M$ and $B$, based on physical principles, and vaguely based on the modern climate system:
-
-**A.** Predicted seawater surface temperatures are not allowed to be $<< -2^\circ C$, the freezing point of sea water.
-The highest prior density of $A$ is placed around $0^\circ C$, and it slowly tapers off towards higher temperatures.
-This shape is achieved by placing a skew-normal prior on the lower asymptote, specified as
-$$A \sim SN(\xi=-3.0,\omega=12,\alpha_{SN}=30),$$
-where $\xi$, $\omega$, and $\alpha_{SN}$ are the location, scale and shape parameters.
-
-**K.** Input of solar energy decreases from the tropics to the poles.
-Hence, the latitudinal temperature gradient is broadly negative, i.e.
-temperature decreases with absolute latitude.
-This is achieved by setting $K \ge A$.
-The prior on the upper asymptote $K$ is a truncated normal distribution with the mean set to $K$ of the modern SST gradient, with a broad standard deviation:
-$$K \sim TN(\mu_{TN} = 28, \sigma_{TN} = 10, \alpha_{TN} = A, \beta_{TN}= \infty)$$
-The distribution is truncated to the left at $\alpha_{TN}=A$, but not truncated to the right ($\beta_{TN}$).
-
-**M.** The steepness of the gradient is presumed to be highest in mid-latitudes; this is expressed with a normal prior on $M$ with the mean set to $M$ of the modern SST gradient:
-$$M \sim N(42,10)$$
-**B.** The steepness or growth rate $B$ of the gradient is constrained to be $\ge0$ and to not be exceedingly high, as oceanic and atmospheric heat transfer is bound to prevent very abrupt SST changes across latitudes.
-A gamma-distributed prior of the form
-$$B \sim Gamma(\alpha_G=4.3,\beta_G=30)$$
-was placed on $B$. The shape and rate parameters $\alpha_G$ and $\beta_G$ were chosen such that the highest prior density is at $B$ of the modern SST gradient, $0.11$.
-
-# References
+```{r out.width='100%', echo = FALSE, fig.cap = "Fig S3.
+Latitudinal temperature gradient in seperate hemispheres, showing the medians (lines) and 95% credibal intervals (shadings) in the southern (red) and northern hemisphere (blue).
+Symbols with vertical lines show the median and 95% credible intervals of $\\mu$.
+Turquoise symbols in the northern hemisphere highlight the ecological proxy data.
+The grey line and shading in the background show the latitudinal temperature gradient with the data from both hemispheres combined, plotted in both hemispheres."}
+setwd('..') # move up to parent folder of /manuscript
+knitr::include_graphics("D://Github/PalaeoClimateGradient/figures/SM/FigS3_north_south_eeco.png", error = FALSE)
+```

### 2023-01-23 R/analyses/SM/run_EECO_climate_model_hemispheres.R

@@ -1,0 +1,135 @@
+
+###
+### seperate gradients- northern hemisphere
+###
+
+### Read data
+## Hollis data
+dat <- readRDS("data/processed/Hollis_processed_EECO_2022_07_19.rds")
+#
+# prepare for use in the model
+obsmat <- data.frame(sample = (paste((dat$p_lat),dat$longitude, dat$proxy)),
+                     p_lat = (dat$p_lat), temperature = dat$temperature,
+                     sd = dat$temperature_sd,
+                     proxy = dat$proxy)
+# order observation matrix for easily keeping tracks of mean estimates
+obsmat <- obsmat[with(obsmat, order(p_lat, sample)),]
+obsmat <- obsmat[which(obsmat$p_lat>=0),]
+
+### Mangrove and Coral data
+# read and assign proxy type
+bioprox <- readRDS("data/processed/bio_proxies_2022_08_08.RDS")
+bioprox$proxy <- rep("Reef",nrow(bioprox))
+bioprox$proxy[which(bioprox$taxa=="Avicennia")] <- "Avicennia"
+bioprox$proxy[which(bioprox$type=="mangrove" & bioprox$taxa!="Avicennia")] <- 
+  "Avicennia-Rhizophoraceae"
+
+# create distribution matrix for use with model (distributions defined in options)
+proxy_index <- sapply(bioprox$proxy, function(f) which(proxy_distributions$name==f))
+
+distrmat = data.frame(p_lat = (bioprox$p_lat), 
+                      mu = proxy_distributions$mean[proxy_index],
+                      scale = proxy_distributions$sd[proxy_index],
+                      shape = proxy_distributions$shape[proxy_index],
+                      distribution = proxy_distributions$distribution[proxy_index])
+distrmat <- distrmat[which(distrmat$p_lat>=0),]
+
+### Prepare for model run
+# set up clusters for parallel computing
+library(doParallel)
+nClust <- 4
+cl <- parallel::makeCluster(nClust)
+doParallel::registerDoParallel(cl)
+
+nChains = nClust # run 1 chain per cluster
+### Run model
+# source script
+mode_north <- foreach(pc = 1:nChains) %dopar% {
+  # call model functions
+  source("R/functions/model_components/climate_model_EECO.R")
+  # set random seed
+  set.seed(pc)
+  hierarchical_model(n_iter = 600000, n_thin = 100,
+                     obsmat = obsmat, distrmat = distrmat, 
[…]

### 2023-01-23 R/analyses/global_mean_temperatures.R

@@ -3,6 +3,9 @@
 source("R/functions/model_components/gradient.R")
 mode_all <- readRDS("results/eeco/eeco_climate_model_output_combined.rds")
 modm_all <- readRDS("results/modern/modern_climate_model_output.rds")
+# modern with eocene sampling
+modem_list <- readRDS("./results/modern/modern_sample_eocene_p_lat_gradient.rds")
+modem_all <- combine_posterior(modem_list,5000)
 
 # latitudinal weights in 1 deg lat bands
 alpha1 <- seq(1,90,1)
@@ -20,7 +23,22 @@ motemp <- gradient(seq(0.5,89.5,1), modm_all[,1:4],0)
 motempq <- apply(motemp,2,function(x) quantile(x, probs = c(0.025,0.5,0.975)))
 # modern global average temperature with 95% CI
 motemp_global <- apply(motempq,1, function(x) round(sum(x*latweight),1), simplify = FALSE)
+# modern eocene sampled temperature gradient
+moeotemp <- gradient(seq(0.5,89.5,1), modem_all[,1:4],0)
+# modern eocene sampled temperature gradient quantiles
+moeotempq <- apply(moeotemp,2,function(x) quantile(x, probs = c(0.025,0.5,0.975)))
+# modern eocene sampled global average temperature with 95% CI
+moeotemp_global <- apply(moeotempq,1, function(x) round(sum(x*latweight),1), simplify = FALSE)
 #
+
+moeotemp_global_all <- apply(moeotemp,1, function(x) (sum(x*latweight)), simplify = T)
+motemp_global_all <- apply(motemp,1, function(x) (sum(x*latweight)), simplify = T)
+
+# modern eocene sample deviation of global average
+modern_eocene_gradient_deviation <- as.list(format(round(quantile(moeotemp_global_all-motemp_global_all, probs = c(0.025,0.5,0.975)),1),nsmall = 1))
+
 # save results
 saveRDS(motemp_global,"results/modern/global_mean_modern.rds")
 saveRDS(eotemp_global,"results/eeco/global_mean_eeco.rds")
+saveRDS(moeotemp_global,"results/eeco/global_mean_modern_eocene_sampling_eeco.rds")
+saveRDS(modern_eocene_gradient_deviation,"results/eeco/global_mean_modern_eocene_sampling_difference.rds")

### 2023-01-23 R/figures/SM/FigS1.R

@@ -3,25 +3,25 @@ eeco_mod <- readRDS("./results/eeco/eeco_climate_model_output.rds")
 source("R/functions/auxiliary_functions.R")
 
 # 1. Traceplot of A, K, M and log(B)
-png("figures/SM/S1_traceplot_eeco.png", width = 4.5, height = 5, units = "in", res = 400)
+png("figures/SM/S1_traceplot_eeco.png", width = 3.5, height = 4.9, units = "in", res = 400)
 plot_chains(eeco_mod, select = list("params", 1:5), xlabel = expression("iteration"), ylabel = c("A", "K - A", "M", "B", expression(sigma)),
             xaxt = "n", xaxsat = c(1,c(1,2,3,4,5,6)*10^5), xaxslab = c(1,"100,000","200,000","300,000",
                                                                        "400,000","500,000","600,000"))
 dev.off()
 
 # 2. Traceplot of some of the location mu
-png("figures/SM/S1_traceplot_eeco_yestimate.png", width = 4.5, height = 5, units = "in", res = 400)
-plot_chains(eeco_mod, select = list("yestimate", c(1,8,25,30)), xlabel = expression("iteration"), ylabel = unlist(sapply(c(1,8,25,30), function(x) 
+png("figures/SM/S1_traceplot_eeco_muj.png", width = 3.5, height = 4.9, units = "in", res = 400)
+plot_chains(eeco_mod, select = list("yestimate", c(1,8,10,25,30)), xlabel = expression("iteration"), ylabel = unlist(sapply(c(1,8,10,25,30), function(x) 
   bquote(mu[.(x)]))),
             xaxt = "n", xaxsat = c(1,c(1,2,3,4,5,6)*10^5), xaxslab = c(1,"100,000","200,000","300,000",
                                                                        "400,000","500,000","600,000"))
 dev.off()
 
-## comment: these are oddly large in some cases
+## comment: large values of sdyest occur when yestimate is far from the empirical mean
 # 3. Traceplot of some of the location sd
-png("figures/SM/S1_traceplot_eeco_yestimate.png", width = 4.5, height = 5, units = "in", res = 400)
-plot_chains(eeco_mod, select = list("sdyest", c(1,8,12,22)), xlabel = expression("iteration"), ylabel = unlist(sapply(c(1,8,25,30), function(x) 
-  bquote(mu[.(x)]))),
[…]

### 2023-01-23 R/figures/SM/FigS2.R

@@ -1,0 +1,68 @@
+# Supplementary Figure S2 - gradient on Northern and Southern hemisphere
+
+# load figures for plotting, distributions ect.
+source("./R/functions/auxiliary_functions.R")
+source("./R/functions/model_components/gradient.R")
+# read model output
+mode <- readRDS("./results/eeco/eeco_climate_model_output.rds")
+mode_all <- readRDS("./results/eeco/eeco_climate_model_output_combined.rds")
+mode_prox <- readRDS("./results/eeco/eeco_climate_model_output_just_proxy.rds")
+mode_all_prox <- combine_posterior(mode_prox, burnin = 100000)
+
+# read the input data
+obsmat <- readRDS("./data/processed/observation_matrix_for_EECO_model.rds")
+distrmat <- readRDS("./data/processed/distribution_matrix_for_EECO_model.rds")
+
+# calculate location means of observation input data for easier plotting
+ylist <- lapply(unique(obsmat$sample), function(x) obsmat$temperature[which(obsmat$sample == x)])
+yobs_mean= c(sapply(ylist,mean))
+n_obs <- length(yobs_mean)
+# add distribution means
+mean_t <- c(yobs_mean,distrmat$mu)
+lats <- mode[[1]]$lat # latitudes of locations is saved in model output
+
+
+cex1 <- c(1.05,1,1.15,0.92)
+alpha2 <- 0.75
+pch1 <- c(16,17,18,15)
+proxind1 <- sapply(obsmat$proxy, function(x) which(x == c("d18O","MgCa","TEX86","d47")))
+alpha3 <- 0.9
+col_proxgrad <- rgb(0.85,0.6,0,0.9)
+col_proxgrad_shade <- rgb(0.85,0.6,0,0.27)
+xlab1 <- expression("|latitude| ("*degree*")")
+ylab1 <- expression("temperature ("*degree*"C)")
+lat_obs <- mode[[1]]$lat
+
+eecoprox <- readRDS("results/eeco/eeco_climate_model_output_just_proxy.rds")
+eecoproxcomb <- combine_posterior(eecoprox, burnin = 100000)
+
+
+mu_obs <- apply(eecoprox[[1]]$yestimate[1001:6000,],2,mean)
[…]

### 2023-01-23 manuscript/manuscript.Rmd
results

@@
 modern_eeco_sample_sdy <- readRDS("results/modern/modern_eeco_sample_sdy.rds")
+modern_sdy <- readRDS("results/modern/modern_sdy.rds")
 
@@
 ## EECO reconstruction
-<!-- - Show Eocene gradient
-- Global t : use raster package for calculation of areal weights
-- report lat gradient
--->
 
-The modelled Eocene temperature gradient is starkly different from the modern (Fig 4).
+The modeled Eocene temperature gradient is starkly different from the modern (Fig 4).
 Modelled, median equatorial temperatures are `r tdiff_lat0$"50%"` (95% CI: `r tdiff_lat0$"2.5%"` - `r tdiff_lat0$"97.5%"`) degC higher for the Eocene, and polar temperatures are higher by `r tdiff_lat90$"50%"` (95% CI: `r tdiff_lat90$"2.5%"` - `r tdiff_lat90$"97.5%"`) degC.
@@
 
-The high variability of early Eocene palaeotemperature proxies, particularly in the mid-latitudes, and the scarcity of high-latitude data, results in substantial uncertainties in the modeled temperature gradient.
-This is reflected in the residual standard deviation ($\sigma$) of the early Eocene gradient, `r eeco_sdy$"50%"` (95% CI: `r eeco_sdy$"2.5%"` - `r eeco_sdy$"97.5%"`) degC, more than twice as high than  $\sigma$ of the modern gradient, resampled at early Eocene latitudes, `r modern_eeco_sample_sdy$"50%"` (95% CI: `r modern_eeco_sample_sdy$"2.5%"` - `r modern_eeco_sample_sdy$"97.5%"`) degC.
-The inclusion of ecological proxy data only slightly reduces the overall model standard deviation, but helps constrain polar temperature estimates (Supplementary materials).
-### Continue here.
+The high variability of early Eocene palaeotemperature proxies, particularly in the mid-latitudes, and the scarcity of high-latitude data, result in substantial uncertainties in the modeled temperature gradient.
+This is reflected in the residual standard deviation ($\sigma$) of the early Eocene gradient, `r eeco_sdy$"50%"` (95% CI: `r eeco_sdy$"2.5%"` - `r eeco_sdy$"97.5%"`) degC, more than twice as high than  $\sigma$ of the modern gradient, `r modern_sdy$"50%"`.
 
 The global Eocene mean sea surface temperature is estimated at `r eotemp_global$"50%"` (95% CI: `r eotemp_global$"2.5%"` - `r eotemp_global$"97.5%"`) degC, significantly higher than the modern (`r motemp_global$"50%"` degC).
+
+For comparison, a model run excluding the ecological proxies results in a global mean temperature of  
+The inclusion of ecological proxy data only slightly reduces the overall model standard deviation (Supplementary materials).
 

### 2023-01-23 R/analyses/assess_sdy.R

@@ -3,13 +3,18 @@ source("./R/functions/model_processing/combine_posterior.R")
 modern_gradient_eeco_lat <- readRDS("./results/modern/modern_sample_eocene_p_lat_gradient.RDS")
 modern_gradient_eeco_lat <- combine_posterior(mod = modern_gradient_eeco_lat, burnin = 5000)
 eeco_gradient <- readRDS("./results/eeco/eeco_climate_model_output_combined.rds")
+modern_gradient<- readRDS("results/modern/modern_climate_model_output.rds")
 
-modern_eeco_sample_sdy <- as.list(format(round(quantile(eocene_sample$sdy,probs = c(.025,.5,.975)),1),nsmall = 1))
+modern_eeco_sample_sdy <- as.list(format(round(quantile(modern_gradient_eeco_lat$sdy,probs = c(.025,.5,.975)),1),nsmall = 1))
 
-eeco_sdy <- as.list(format(round(quantile(mode_all$sdy,probs = c(.025,.5,.975)),1),nsmall = 1))
+eeco_sdy <- as.list(format(round(quantile(eeco_gradient$sdy,probs = c(.025,.5,.975)),1),nsmall = 1))
+
+modern_sdy <- as.list(format(round(quantile(modern_gradient$sdy,probs = c(.025,.5,.975)),1),nsmall = 1))
+
 
 saveRDS(modern_eeco_sample_sdy, "results/modern/modern_eeco_sample_sdy.rds")
 saveRDS(eeco_sdy, "results/eeco/eeco_sdy.rds")
+saveRDS(modern_sdy, "results/modern/modern_sdy.rds")
 
 # just proxy sdy
 eeco_prox <- readRDS("results/eeco/eeco_climate_model_output_just_proxy.rds")

### 2023-01-24 manuscript/manuscript.Rmd
results updates

@@
 eocene_gradient <- readRDS("results/eeco/eocene_overall_gradient.rds")
+eocene_proxy_gradient <- readRDS("results/eeco/eocene_proxy_gradient.rds")
+
 modern_gradient <- readRDS("results/modern/modern_overall_gradient.rds")
@@
 modern_eocene_gradient_diff <- readRDS("results/modern/modern_eocene_sampling_overall_gradient_diff.rds")
+proxy_uncertain_diff_percent <- readRDS("results/eeco/proxy_vs_full_uncertain_diff_percent.rds")
 
@@
 moeotemp_global_diff <- readRDS("results/eeco/global_mean_modern_eocene_sampling_difference.rds")
+eoceneproxy_global_diff <- readRDS("results/eeco/global_mean_eocene_eoceneproxy_difference.rds")
 
@@
 3b).
-The 95% credible interval of reconstructed gradients (Fig.
+The 95% credible interval (CI) of reconstructed gradients (Fig.
 3b) is noticeably narrower than the spread of empirical temperature values (Fig.
 3a).
-The modern, latitudinal gradient, modeled with all data, is `r modern_gradient$"50%"` degC, and has a median of `r modern_eocene_gradient$"50%"`  (95% CI: `r modern_eocene_gradient$"2.5%"` - `r modern_eocene_gradient$"97.5%"`) degC when modelled with only the early Eocene sampling locations.
+The modern, latitudinal gradient, modeled with all data, spans `r modern_gradient$"50%"` degC from the equator to the poles.
+When modeled with only the early Eocene sampling locations, the median temeprature gradient is `r modern_eocene_gradient$"50%"`  (95% CI: `r modern_eocene_gradient$"2.5%"` - `r modern_eocene_gradient$"97.5%"`) degC.
 A random sampling of modern temperatures at latitudes corresponding to the sampled Eocene palaeolatitudes would thus allow the reconstruction of the modern latitudinal temperature gradient with good accuracy, with expected deviations from the actual gradient as low as `r modern_eocene_gradient_diff$"2.5%"` to `r modern_eocene_gradient_diff$"97.5%"` degC (95 % CI).
 
-The modeled, latitudinal temperature gradient facilitates accurate reconstruction of global average temperatures.
+Modeling the latitudinal temperature gradient facilitates accurate reconstruction of global average temperatures.
 The modern, global mean sea surface temperature estimate is `r motemp_global$"50%"` degC when using the full, modern data set.
@@
 Modelled, median equatorial temperatures are `r tdiff_lat0$"50%"` (95% CI: `r tdiff_lat0$"2.5%"` - `r tdiff_lat0$"97.5%"`) degC higher for the Eocene, and polar temperatures are higher by `r tdiff_lat90$"50%"` (95% CI: `r tdiff_lat90$"2.5%"` - `r tdiff_lat90$"97.5%"`) degC.
-This results in a strongly flattened latitudinal temperature gradient of `r eocene_gradient$"50%"`  (95% CI: `r eocene_gradient$"2.5%"` - `r eocene_gradient$"97.5%"`) degC for the Eocene, as opposed to `r modern_gradient$"50%"` degC for the modern.
+This results in a flattened latitudinal temperature gradient of `r eocene_gradient$"50%"`  (95% CI: `r eocene_gradient$"2.5%"` - `r eocene_gradient$"97.5%"`) degC for the Eocene, as opposed to `r modern_gradient$"50%"` degC for the modern.
 
@@
 This is reflected in the residual standard deviation ($\sigma$) of the early Eocene gradient, `r eeco_sdy$"50%"` (95% CI: `r eeco_sdy$"2.5%"` - `r eeco_sdy$"97.5%"`) degC, more than twice as high than $\sigma$ of the modern gradient, `r modern_sdy$"50%"`.
+This signifies that the early Eocene data fit less well to the logistic, latitudinal gradient model, which can also be seen from the drastic departure of some of the proxy data from the gradient estimates (Fig.
+4).
 
-The global Eocene mean sea surface temperature is estimated at `r eotemp_global$"50%"` (95% CI: `r eotemp_global$"2.5%"` - `r eotemp_global$"97.5%"`) degC, significantly higher than the modern (`r motemp_global$"50%"` degC).
+The global early Eocene mean sea surface temperature is estimated at `r eotemp_global$"50%"` (95% CI: `r eotemp_global$"2.5%"` - `r eotemp_global$"97.5%"`) degC, significantly higher than the modern (`r motemp_global$"50%"` degC).
 
-For comparison, a model run excluding the ecological proxies results in a global mean temperature of  
-The inclusion of ecological proxy data only slightly reduces the overall model standard deviation (Supplementary materials).
+A model run excluding the ecological proxies results in a similar median latitudinal gradient (`r eocene_proxy_gradient$"50%"` degC), but with a `r proxy_uncertain_diff_percent`% wider 95% CI (Supplementary materials).
+This indicates that the ecological proxy data are broadly in agreement with the geochemical proxy data, while providing additional constraints on the shape of the early Eocene temperature gradient.
+
+Due to the limited spatial coverage of the early Eocene proxy record, and due to the added model complexity of simultaneously estimating a model across both hemispheres, we pooled the proxy data across both hemispheres.
+Applying the model seperately to each hemispheres results in substantial differences in hemispherical, average temperatures, with the southern hemisphere being warmer by ...
 

### 2023-01-24 R/analyses/calculate_temperature_gradients_from_models.R

@@ -11,15 +11,22 @@ mode <- readRDS("./results/eeco/eeco_climate_model_output_combined.rds")
 # modern with eocene sampling
 modem_list <- readRDS("./results/modern/modern_sample_eocene_p_lat_gradient.rds")
 modem <- combine_posterior(modem_list,5000)
-  
+
+# just proxy
+eeco_prox <- readRDS("results/eeco/eeco_climate_model_output_just_proxy.rds")
+eeco_prox <- combine_posterior(eeco_prox, burnin = 100000)
+
+
 # calculate temperatures at lat 0 and lat 90 for every iteration
 modern_t0 <- gradient(0,modm[,1:4])
 modern_eocene_t0 <- gradient(0,modem[,1:4])
 eocene_t0 <- gradient(0,mode[,1:4])
+eeco_prox_t0 <- gradient(0,eeco_prox[,1:4])
 
 modern_t90 <- gradient(90,modm[,1:4])
 modern_eocene_t90 <- gradient(90,modm[,1:4])
 eocene_t90 <- gradient(90,mode[,1:4])
+eeco_prox_t90 <- gradient(90,eeco_prox[,1:4])
 
 n <- length(modern_t0)
 
@@ -32,6 +39,12 @@ tdiff_lat90 <- as.list(format(round(quantile(eocene_t90[sample(1:n,n)] - modern_
 eocene_gradient <- as.list(format(round(quantile(eocene_t0 - eocene_t90, probs = c(0.025,0.5,0.975)),1),nsmall = 1))
 modern_gradient <- as.list(format(round(quantile(modern_t0 - modern_t90, probs = c(0.025,0.5,0.975)),1),nsmall = 1))
 modern_eocene_gradient <- as.list(format(round(quantile(modern_eocene_t0 - modern_eocene_t90, probs = c(0.025,0.5,0.975)),1),nsmall = 1))
+eocene_proxy_gradient <- as.list(format(round(quantile(eeco_prox_t0 - eeco_prox_t90, probs = c(0.025,0.5,0.975)),1),nsmall = 1))
+
+eocene_gradient_uncertainty <- quantile(eocene_t0 - eocene_t90, probs = c(0.975))-quantile(eocene_t0 - eocene_t90, probs = c(0.025))
+eocene_proxy_gradient_uncertainty <- quantile(eeco_prox_t0 - eeco_prox_t90, probs = c(0.975))-quantile(eeco_prox_t0 - eeco_prox_t90, probs = c(0.025))
+
[…]

### 2023-01-24 R/analyses/global_mean_temperatures.R

@@ -1,11 +1,16 @@
 # global mean temperatures for the modern and the Eocene
 
 source("R/functions/model_components/gradient.R")
+source("R/functions/auxiliary_functions.R")
+
 mode_all <- readRDS("results/eeco/eeco_climate_model_output_combined.rds")
 modm_all <- readRDS("results/modern/modern_climate_model_output.rds")
 # modern with eocene sampling
 modem_list <- readRDS("./results/modern/modern_sample_eocene_p_lat_gradient.rds")
 modem_all <- combine_posterior(modem_list,5000)
+# just proxy
+eeco_prox <- readRDS("results/eeco/eeco_climate_model_output_just_proxy.rds")
+eeco_prox <- combine_posterior(eeco_prox, burnin = 100000)
 
 # latitudinal weights in 1 deg lat bands
 alpha1 <- seq(1,90,1)
@@ -30,15 +35,23 @@ moeotempq <- apply(moeotemp,2,function(x) quantile(x, probs = c(0.025,0.5,0.975)
 # modern eocene sampled global average temperature with 95% CI
 moeotemp_global <- apply(moeotempq,1, function(x) round(sum(x*latweight),1), simplify = FALSE)
 #
+# eocene proxy only temperature gradient
+eotempprox <- gradient(seq(0.5,89.5,1), eeco_prox[,1:4],0)
+
 
 moeotemp_global_all <- apply(moeotemp,1, function(x) (sum(x*latweight)), simplify = T)
 motemp_global_all <- apply(motemp,1, function(x) (sum(x*latweight)), simplify = T)
-
+eotemp_global_all <- apply(eotemp,1, function(x) (sum(x*latweight)), simplify = T)
+eotempprox_global_all <- apply(eotempprox,1, function(x) (sum(x*latweight)), simplify = T)
 # modern eocene sample deviation of global average
 modern_eocene_gradient_deviation <- as.list(format(round(quantile(moeotemp_global_all-motemp_global_all, probs = c(0.025,0.5,0.975)),1),nsmall = 1))
+
+# eocene proxy deviation of all eocene average
+eocene_eoceneproxy_gradient_deviation <- as.list(format(round(quantile(eotempprox_global_all-eotemp_global_all, probs = c(0.025,0.5,0.975)),1),nsmall = 1))
 
 # save results
 saveRDS(motemp_global,"results/modern/global_mean_modern.rds")
[…]

### 2023-02-02 manuscript/manuscript.Rmd
results and intro updates

@@
 Quantitative temperature reconstructions have therefore not made use of these data.
+Likewise, the large spatial gaps in existing proxy records have often been bridged by simple averaging, without taking into consideration the spatial distribution of the samples.
 Here, we present a Bayesian hierarchical model to integrate geological data–with established temperature proxies–into a unified quantitative framework.
@@
 # Introduction
+The increasing availability of compilations and databases of deep-time climate data now offers the opportunity for empirical palaeoclimate reconstructions of unprecedented scope [e.g.
+@Veizer2015; @Hollis2019; @Grossman2022; @Judd2022].
+In order for these reconstructions to be effective, two fundamental challenges need to be overcome: 1) disparate types of proxy data need to be integrated to inform a single climate reconstruction, and 2) large-scale scale climate patterns need to be inferred from localised, often sparse observations.
+
+For the marine realm, most palaeoclimate reconstructions use geochemical proxy data
+
+The first challenge requires that all proxy data is converted to a common scale.
+
+A fundamental challenge in reconstructing Earth's past climate is to infer large-scale climatic patterns from local observations, derived from disparate proxies.  In order for the
+
+Proxy-derived palaeoclimate reconstructions offer an important alternative to and a test of Earth system models, and as such are fundamental for reconstructing Earth's past climate
+
 The geological record holds evidence for episodes in Earth history with fundamentally different climates than our current [@Scotese2021].
@@
 The Eocene posterior inference is thus based on a total of 20,000 iterations with low autocorrelation (effective multivariate sample size for $A$, $K$, $M$ and $B$ is > 18,000).
-Trace plots were used to assess the convergence and mixing of the MCMC chains (Supplementary Materials S1).
+Trace plots of the MCMC chains indicate convergence and good mixing of the chains (Supplementary Figure S1).
 
@@
 
+eocene_north_south_gradient_diff <- readRDS("./results/SM/eocene_north_south_gradient_diff.rds")
+
 # sdy
@@
 eoceneproxy_global_diff <- readRDS("results/eeco/global_mean_eocene_eoceneproxy_difference.rds")
+eeco_north_global <- readRDS("results/SM/global_mean_northern_eeco.rds")
+eeco_south_global <- readRDS("results/SM/global_mean_southern_eeco.rds")
+eeco_south_north_global_diff <- readRDS("results/SM/global_mean_diff_south_north_eeco.rds")
 
@@
 The modern, global mean sea surface temperature estimate is `r motemp_global$"50%"` degC when using the full, modern data set.
-When reduced to the early Eocene sampling distribution, the modern global mean sea surface temperature is estimated at `r moeotemp_global$"50%"` (95% CI: `r moeotemp_global$"2.5%"` - `r moeotemp_global$"97.5%"`) degC.
+When reduced to the early Eocene sampling distribution, the modern global mean sea surface temperature is estimated at `r moeotemp_global$"50%"` (`r moeotemp_global$"2.5%"` - `r moeotemp_global$"97.5%"`) degC.
 The deviations from the full modern estimate are again very low, with a 95 % credible interval of `r moeotemp_global_diff$"2.5%"` to `r moeotemp_global_diff$"97.5%"` degC.
@@
 The modeled Eocene temperature gradient is starkly different from the modern (Fig 4).
-Modelled, median equatorial temperatures are `r tdiff_lat0$"50%"` (95% CI: `r tdiff_lat0$"2.5%"` - `r tdiff_lat0$"97.5%"`) degC higher for the Eocene, and polar temperatures are higher by `r tdiff_lat90$"50%"` (95% CI: `r tdiff_lat90$"2.5%"` - `r tdiff_lat90$"97.5%"`) degC.
-This results in a flattened latitudinal temperature gradient of `r eocene_gradient$"50%"`  (95% CI: `r eocene_gradient$"2.5%"` - `r eocene_gradient$"97.5%"`) degC for the Eocene, as opposed to `r modern_gradient$"50%"` degC for the modern.
+Modelled, median equatorial temperatures are `r tdiff_lat0$"50%"` (`r tdiff_lat0$"2.5%"` - `r tdiff_lat0$"97.5%"`) degC higher for the Eocene, and polar temperatures are higher by `r tdiff_lat90$"50%"` (`r tdiff_lat90$"2.5%"` - `r tdiff_lat90$"97.5%"`) degC.
+This results in a flattened latitudinal temperature gradient of `r eocene_gradient$"50%"`  (`r eocene_gradient$"2.5%"` - `r eocene_gradient$"97.5%"`) degC for the Eocene, as opposed to `r modern_gradient$"50%"` degC for the modern.
 
 The high variability of early Eocene palaeotemperature proxies, particularly in the mid-latitudes, and the scarcity of high-latitude data, result in substantial uncertainties in the modeled temperature gradient.
-This is reflected in the residual standard deviation ($\sigma$) of the early Eocene gradient, `r eeco_sdy$"50%"` (95% CI: `r eeco_sdy$"2.5%"` - `r eeco_sdy$"97.5%"`) degC, more than twice as high than $\sigma$ of the modern gradient, `r modern_sdy$"50%"`.
+This is reflected in the residual standard deviation ($\sigma$) of the early Eocene gradient, `r eeco_sdy$"50%"` (`r eeco_sdy$"2.5%"` - `r eeco_sdy$"97.5%"`) degC, more than twice as high than $\sigma$ of the modern gradient, `r modern_sdy$"50%"`.
 This signifies that the early Eocene data fit less well to the logistic, latitudinal gradient model, which can also be seen from the drastic departure of some of the proxy data from the gradient estimates (Fig.
@@
 
-The global early Eocene mean sea surface temperature is estimated at `r eotemp_global$"50%"` (95% CI: `r eotemp_global$"2.5%"` - `r eotemp_global$"97.5%"`) degC, significantly higher than the modern (`r motemp_global$"50%"` degC).
+The global early Eocene mean sea surface temperature is estimated at `r eotemp_global$"50%"` (`r eotemp_global$"2.5%"` - `r eotemp_global$"97.5%"`) degC, significantly higher than the modern (`r motemp_global$"50%"` degC).
 
-A model run excluding the ecological proxies results in a similar median latitudinal gradient (`r eocene_proxy_gradient$"50%"` degC), but with a `r proxy_uncertain_diff_percent`% wider 95% CI (Supplementary materials).
-This indicates that the ecological proxy data are broadly in agreement with the geochemical proxy data, while providing additional constraints on the shape of the early Eocene temperature gradient.
+A model run excluding the ecological proxies results in a similar median latitudinal gradient, `r eocene_proxy_gradient$"50%"` degC, but with a `r proxy_uncertain_diff_percent`% wider 95% CI (Supplementary materials).
+This indicates that the ecological proxy data are broadly in agreement with the geochemical proxies, while providing additional constraints on the shape of the early Eocene temperature gradient.
 
 Due to the limited spatial coverage of the early Eocene proxy record, and due to the added model complexity of simultaneously estimating a model across both hemispheres, we pooled the proxy data across both hemispheres.
-Applying the model seperately to each hemispheres results in substantial differences in hemispherical, average temperatures, with the southern hemisphere being warmer by ...
+Applying the model separately to each hemispheres results in substantial differences in hemispherical, average temperatures, with the southern hemisphere being warmer by `r eeco_south_north_global_diff$"50%"` (`r eeco_south_north_global_diff$"2.5%"` - `r eeco_south_north_global_diff$"97.5%"`) degC.
+The inferred latitudinal gradient is somewhat steeper (`r eocene_north_south_gradient_diff$"50%"` degC, although the 95% CI spans `r eocene_north_south_gradient_diff$"2.5%"` to `r eocene_north_south_gradient_diff$"97.5%"` degC) in the northern hemisphere, but the large uncertainties associated with both gradients, and the lack of polar proxy data in the southern hemisphere preclude a more precise statement (Fig.
+S3).)
 

### 2023-02-02 manuscript/supplementary_material.Rmd
results and intro updates

@@
 ## S1. Convergence checks
-To ensure the convergence and mixing of the chains, we inspected traceplots of all the unknown parameters of the model.
 
@@
 The four colours correspond to the four independent model runs.
-a) Traceplot of A, K-A, M, B and $\\sigma$; b) Traceplot of five selected $\\mu_j$; c) Traceplot of five selected $\\sigma_j$."}
+a) Traceplot of A, K-A, M, B and $\\sigma$; b) Traceplot of five selected $\\mu_j$; c) Traceplot of five selected $\\sigma_j$.
+All traceplots display mixing of the chains, and relatively quick convergence."}
 setwd('..') # move up to parent folder of /manuscript
@@
 
-## S3. Seperate gradients on the northen and southern hemisphere
+## S3. Seperate gradients on the northern and southern hemisphere
 

### 2023-02-02 R/analyses/global_mean_temperatures.R

@@ -12,6 +12,10 @@ modem_all <- combine_posterior(modem_list,5000)
 eeco_prox <- readRDS("results/eeco/eeco_climate_model_output_just_proxy.rds")
 eeco_prox <- combine_posterior(eeco_prox, burnin = 100000)
 
+# eeco northern and southern
+eeco_north <- readRDS("results/SM/eeco_climate_model_northern_hemisphere_output_combined.rds")
+eeco_south <- readRDS("results/SM/eeco_climate_model_southern_hemisphere_output_combined.rds")
+
 # latitudinal weights in 1 deg lat bands
 alpha1 <- seq(1,90,1)
 alpha2 <- seq(0,89,1)
@@ -39,10 +43,31 @@ moeotemp_global <- apply(moeotempq,1, function(x) round(sum(x*latweight),1), sim
 eotempprox <- gradient(seq(0.5,89.5,1), eeco_prox[,1:4],0)
 
 
+# eocene northern temperature gradient
+eeconorthtemp <- gradient(seq(0.5,89.5,1), eeco_north[,1:4],0)
+# eocene northern temperature gradient quantiles
+eeconorthtempq <- apply(eeconorthtemp,2,function(x) quantile(x, probs = c(0.025,0.5,0.975)))
+# eocene northern global average temperature with 95% CI
+eeconorthtemp_global <- apply(eeconorthtempq,1, function(x) round(sum(x*latweight),1), simplify = FALSE)
+#
+# eocene southern temperature gradient
+eecosouthtemp <- gradient(seq(0.5,89.5,1), eeco_south[,1:4],0)
+# eocene southern temperature gradient quantiles
+eecosouthtempq <- apply(eecosouthtemp,2,function(x) quantile(x, probs = c(0.025,0.5,0.975)))
+# eocene southern global average temperature with 95% CI
+eecosouthtemp_global <- apply(eecosouthtempq,1, function(x) round(sum(x*latweight),1), simplify = FALSE)
+#
+
 moeotemp_global_all <- apply(moeotemp,1, function(x) (sum(x*latweight)), simplify = T)
 motemp_global_all <- apply(motemp,1, function(x) (sum(x*latweight)), simplify = T)
 eotemp_global_all <- apply(eotemp,1, function(x) (sum(x*latweight)), simplify = T)
 eotempprox_global_all <- apply(eotempprox,1, function(x) (sum(x*latweight)), simplify = T)
[…]

### 2023-02-04 manuscript/manuscript.Rmd
Update manuscript.Rmd

@@
 # Introduction
-The increasing availability of compilations and databases of deep-time climate data now offers the opportunity for empirical palaeoclimate reconstructions of unprecedented scope [e.g.
-@Veizer2015; @Hollis2019; @Grossman2022; @Judd2022].
-In order for these reconstructions to be effective, two fundamental challenges need to be overcome: 1) disparate types of proxy data need to be integrated to inform a single climate reconstruction, and 2) large-scale scale climate patterns need to be inferred from localised, often sparse observations.
+The increasing availability of large compilations and databases of deep-time climate data now offers the opportunity for empirical palaeoclimate reconstructions of unprecedented scope [e.g.
+@Veizer2015; @Hollis2019; @Scotese2021; @Grossman2022; @Judd2022].
+To exploit the full potential these data sets offer, two fundamental challenges need to be overcome: Disparate types of proxy data need to be integrated to inform a single climate reconstruction, and large-scale scale climate patterns need to be inferred from localised, often sparse observations.
+To address these problems, we present a novel Bayesian method that combines quantitative proxies and palaeoecological constraints into a fully quantitative model of the latitudinal temperature gradient, bridging spatial gaps in sparsely sampled climate data.
 
-For the marine realm, most palaeoclimate reconstructions use geochemical proxy data
+Existing palaeoclimate reconstructions are generally based on one of three approaches: 1) Quantitative, geochemical proxy data, 2) palaeobiological or geological climate indicators that provide only broad climate constraints which are hard to quantify, and 3) earth system modelling [@Burgener2023], with the latter ultimately relying on constraints provided by the geological and fossil record [@Lunt2021].
+Combining quantitative proxy data with other climate constraints has the potential to improve upon climate reconstructions based on a single type of proxy [@Burgener2023], but is rarely employed in deep-time palaeoclimatic studies.
+
+Climate data not based on geochemical, quantitative proxies generally provide broad categorical information, e.g.
+"tropical" or "dry", or broad quantitative constraints, e.g.
+"> 20 degC mean annual temperature".
+Integrating this information into fully quantitative climate reconstructions requires transforming it to a numerical scale.
+This approach is showcased by nearest living relative methods, commonly employed in terrestrial, Cenozoic climate reconstructions, which work on the premise that the climatic requirements of ancient plant taxa were similar to those of there nearest modern relatives [@Kvacek2007].
+A plethora of methods have been developed to derive numerical climate values from taxonomic data [@Faucette2007; @...].
+Approaches based on probability density functions have the advantage of naturally capturing the uncertainty associated with the climatic occupancy of modern nearest relatives [@Greenwood2017], and probability distributions can be readily integrated in a quantiative modelling framework.
+
+Deep-time palaeoclimate reconstructions use geochemical proxy data
 

### 2023-02-04 manuscript/manuscript.Rmd
reordering introduction

@@
 # Introduction
+Existing palaeoclimate reconstructions are generally based on one of three approaches: 1) Quantitative, geochemical proxy data, 2) palaeobiological or geological climate indicators that provide only broad climate constraints which are hard to quantify, and 3) earth system modelling [@Burgener2023], with the latter ultimately relying on constraints provided by the geological and fossil record [@Lunt2021].
 The increasing availability of large compilations and databases of deep-time climate data now offers the opportunity for empirical palaeoclimate reconstructions of unprecedented scope [e.g.
@@
 To exploit the full potential these data sets offer, two fundamental challenges need to be overcome: Disparate types of proxy data need to be integrated to inform a single climate reconstruction, and large-scale scale climate patterns need to be inferred from localised, often sparse observations.
-To address these problems, we present a novel Bayesian method that combines quantitative proxies and palaeoecological constraints into a fully quantitative model of the latitudinal temperature gradient, bridging spatial gaps in sparsely sampled climate data.
+To address these problems, we present a novel Bayesian method that combines quantitative proxies and palaeoecological constraints into a fully quantitative model of the latitudinal gradient of sea surface temperatures, bridging spatial gaps in sparsely sampled climate data.
 
-Existing palaeoclimate reconstructions are generally based on one of three approaches: 1) Quantitative, geochemical proxy data, 2) palaeobiological or geological climate indicators that provide only broad climate constraints which are hard to quantify, and 3) earth system modelling [@Burgener2023], with the latter ultimately relying on constraints provided by the geological and fossil record [@Lunt2021].
-Combining quantitative proxy data with other climate constraints has the potential to improve upon climate reconstructions based on a single type of proxy [@Burgener2023], but is rarely employed in deep-time palaeoclimatic studies.
+For the marine realm, Geochemical proxies such as oxygen isotopes (d18O), tetraether indices (TEX86) and magnesium calcium ratios (Mg/Ca) are used to derive palaeotemperature estimates for individual locations using experimentally derived transfer functions [e.g.
+@Liu2009].
 
@@
 A plethora of methods have been developed to derive numerical climate values from taxonomic data [@Faucette2007; @...].
-Approaches based on probability density functions have the advantage of naturally capturing the uncertainty associated with the climatic occupancy of modern nearest relatives [@Greenwood2017], and probability distributions can be readily integrated in a quantiative modelling framework.
+Approaches based on probability density functions have the advantage of naturally capturing the uncertainty associated with the climatic occupancy of modern nearest relatives [@Greenwood2017], and probability distributions can be readily integrated in a quantitative modelling framework.
+Here, we employ a similar approach, constructing probability distributions broadly reflecting the thermal preferences of modern coral reefs and mangrove taxa [@...], which we assign to fossil occurrences of coral reefs and mangroves.
+
+Combining quantitative proxy data with other climate constraints has the potential to improve upon climate reconstructions based on a single type of proxy [@Burgener2023], but is rarely employed in deep-time palaeoclimatic studies.
 

### 2023-02-04 manuscript/manuscript.Rmd
rewriting introduction

@@
 # Introduction
-Existing palaeoclimate reconstructions are generally based on one of three approaches: 1) Quantitative, geochemical proxy data, 2) palaeobiological or geological climate indicators that provide only broad climate constraints which are hard to quantify, and 3) earth system modelling [@Burgener2023], with the latter ultimately relying on constraints provided by the geological and fossil record [@Lunt2021].
 The increasing availability of large compilations and databases of deep-time climate data now offers the opportunity for empirical palaeoclimate reconstructions of unprecedented scope [e.g.
@@
 To exploit the full potential these data sets offer, two fundamental challenges need to be overcome: Disparate types of proxy data need to be integrated to inform a single climate reconstruction, and large-scale scale climate patterns need to be inferred from localised, often sparse observations.
-To address these problems, we present a novel Bayesian method that combines quantitative proxies and palaeoecological constraints into a fully quantitative model of the latitudinal gradient of sea surface temperatures, bridging spatial gaps in sparsely sampled climate data.
+To address these challenges, we present a novel Bayesian method that combines quantitative proxies and palaeoecological constraints into a fully quantitative model of the latitudinal gradient of sea surface temperatures, bridging spatial gaps in sparsely sampled climate data.
 
+Existing palaeoclimate reconstructions are generally based on one of three approaches.
+Quantitative, geochemical proxies such as oxygen isotopes (d18O), tetraether indices (TEX86) and magnesium calcium ratios (Mg/Ca), and sometimes biological proxies such as leaf physiognomy, are used to derive palaeotemperature estimates using experimentally derived transfer functions [e.g.
+@Liu2009; @Royer2012].
+These data can be readily processed for regional or global-scale palaeoclimate reconstructions [@...].
+Palaeoecological and geological proxy data can provide broad climate constraints, but no established transfer functions exist to directly infer quantitative climate parameters from these records [e.g.
+@Kvacek2007, @Cao2019].
+Earth system models offer the most highly resolved and complete climate reconstructions, but ultimately rely on constraints provided by the geological and fossil record [@Lunt2021].
+
+Marine, proxy-based, deep-time palaeoclimate reconstructions commonly rely solely on geochemical proxies [@Veizer2015;@Grossman2022].
+Despite the impressive size of these archives large spatial gaps bias global temperature reconstructions based on those records [@Jones2022].
+In contrast, studies on the terrestrial and coastal climate record have long used ecological and lithological data for quantitative climate reconstructions [@].
+For example, a plethora of "nearest living relative" methods exists for deriving numerical climate estmates from taxonomic plant data for the Cenozoic  [@Faucette2007; @...].
+These methods work on the the premise that the climatic requirements of ancient plant taxa were similar to those of their nearest modern relatives [@Kvacek2007; @Greenwood2017].
+
+Even when using the full range of available records, any proxy-based climate reconstructions that goes back beyond the Holocene is inevitably very patchy.
+To infer regional or global-scale patterns from an incomplete record requires the bridging of spatial gaps.
+This can involve interpolation between data points, or extrapolation if entire climate zones remain unsampled.
+
+Here, we apply broadly analogous method to the marine record, deriving probability densities broadly reflecting the thermal preferences of modern coral reefs and mangrove taxa [@...], which we assign to fossil occurrences of coral reefs and mangroves.
+
+Combining quantitative proxy data with not fully quantified climate constraints, e.g.
+the presence of tropical weathering products such as bauxite and laterite, has the potential to improve upon existing, purely proxy-based reconstructions [@Scotese2021; @Burgener2023].
+Terrestrial and
+
+approach is still rarely employed in deep-time palaeoclimatic studies, as it This requires transforming ordinal or interval-bound data into numerical estimates, which is common
+
+Integrating this type of information into fully quantitative climate reconstructions requires
+
+data provide a different method of reconstructing palaeotemperatures
+Palaeobiological or geological climate indicators that provide only broad climate constraints which are hard to quantify, and 3) 
+[@Burgener2023]
 For the marine realm, Geochemical proxies such as oxygen isotopes (d18O), tetraether indices (TEX86) and magnesium calcium ratios (Mg/Ca) are used to derive palaeotemperature estimates for individual locations using experimentally derived transfer functions [e.g.

### 2023-02-05 manuscript/manuscript.Rmd
introduction updates

@@
 
-Existing palaeoclimate reconstructions are generally based on one of three approaches.
-Quantitative, geochemical proxies such as oxygen isotopes (d18O), tetraether indices (TEX86) and magnesium calcium ratios (Mg/Ca), and sometimes biological proxies such as leaf physiognomy, are used to derive palaeotemperature estimates using experimentally derived transfer functions [e.g.
-@Liu2009; @Royer2012].
+Existing palaeoclimate reconstructions are generally based on one of three approaches [see @Burgener2023]: 1) Quantitative, geochemical proxies such as oxygen isotopes (d18O), tetraether indices (TEX86) and magnesium calcium ratios (Mg/Ca) are used to derive palaeotemperature estimates using experimentally derived transfer functions [e.g.
+@Liu2009].
 These data can be readily processed for regional or global-scale palaeoclimate reconstructions [@...].
-Palaeoecological and geological proxy data can provide broad climate constraints, but no established transfer functions exist to directly infer quantitative climate parameters from these records [e.g.
-@Kvacek2007, @Cao2019].
-Earth system models offer the most highly resolved and complete climate reconstructions, but ultimately rely on constraints provided by the geological and fossil record [@Lunt2021].
+2) Palaeobiological, ecological and geological proxy data can provide broad climate constraints, but no experimentally determined transfer functions exist to directly infer quantitative climate parameters from these records [e.g.
+@Kvacek2007; @Cao2019; but see @Royer2012].
+Instead, palaeoclimate signals are inferred based on the climatic conditions of analogous, modern sediments, or based on the premise that the climatic requirements of ancient taxa, plant traits or ecological communities were similar to those of their nearest modern relatives [@Kvacek2007; @Greenwood2017].
+3) Earth system models offer the most highly resolved and complete climate reconstructions, but ultimately rely on constraints provided by the geological and fossil record [@Lunt2021].
 
-Marine, proxy-based, deep-time palaeoclimate reconstructions commonly rely solely on geochemical proxies [@Veizer2015;@Grossman2022].
-Despite the impressive size of these archives large spatial gaps bias global temperature reconstructions based on those records [@Jones2022].
-In contrast, studies on the terrestrial and coastal climate record have long used ecological and lithological data for quantitative climate reconstructions [@].
-For example, a plethora of "nearest living relative" methods exists for deriving numerical climate estmates from taxonomic plant data for the Cenozoic  [@Faucette2007; @...].
-These methods work on the the premise that the climatic requirements of ancient plant taxa were similar to those of their nearest modern relatives [@Kvacek2007; @Greenwood2017].
+Even when using the full range of available records, proxy-based climate reconstructions beyond the Holocene are inevitably patchy [e.g.
+@Hollis2019], and global temperature reconstructions based on those records alone may be biased [@Jones2022].
+To infer regional or global-scale patterns from an incomplete record relies on the bridging of spatial gaps.
+A common approach that requires no additional computation is the spatial visualisation of proxy-derived temperatures against latitude, showing broad latitudinal temperature trends [@Hollis2019, @Vickers2021].
+Interpolation is sometimes used to bridge spatial gaps in palaeoclimate data [@Taylor2004], taking advantage of the autoregressive nature of climatic data: much of the information on the climate of any given location is contained in the climate data of nearby locations [@Reynolds1994].
+Adding to this, some proxy-based reconstructions use statistical modelling to infer climatic patterns.
+For example, polynomial regression has been used to reconstruct latitudinal temperature gradients [@Bijl2009], and 2D-reconstructions of surface temperatures and precipitation have been created with Gaussian process regression [@Inglis2020; @Chandra2021].
 
-Even when using the full range of available records, any proxy-based climate reconstructions that goes back beyond the Holocene is inevitably very patchy.
-To infer regional or global-scale patterns from an incomplete record requires the bridging of spatial gaps.
-This can involve interpolation between data points, or extrapolation if entire climate zones remain unsampled.
+Here, we present a Bayesian, hierarchical model for inferring sea surface temperatures (SST) that expands upon existing, spatial reconstructions of palaeoclimate by allowing for the integration of 1) prior information based on physical principles and the observed, modern SST distribution, and of 2) geochemical and ecological climate proxies in a common, quantitative framework.
+nearest living relative...
 
-Here, we apply broadly analogous method to the marine record, deriving probability densities broadly reflecting the thermal preferences of modern coral reefs and mangrove taxa [@...], which we assign to fossil occurrences of coral reefs and mangroves.
+We apply this approach to the early Eocene climatic optimum (EECO), the interval with the warmest sustained temperatures of the Cenozoic.
+Assuming the conservation of thermal tolerances of modern coral reefs and mangrove taxa, we establish broad temperature ranges for EECO coral reef and mangrove sites.
+We integrate these temperature estimates with the EECO geochemical, shallow marine proxy record to model the latitudinal SST gradient and its uncertainty.
 
+We apply this model to the early Eocene climate optimum,...
+
+"nearest living relative" methods are used for deriving numerical climate estimates from taxonomic plant data for the Cenozoic  [@Faucette2007; @Chevalier2022].
+These methods work on the the premise that .
+Proxy-based, deep-time palaeoclimate reconstructions relying solely on geochemical proxies [e.g.
+@Veizer2015; @Grossman2022] have significant spatial gaps, and global temperature reconstructions based on those records alone may be biased [@Jones2022].
+Ecological and lithological data are sometimes used for quantitative climate reconstructions [@Uhl2009;], but are rarely used in conjunction with geochemical proxies [@Burgener2023].
+
+For example, a plethora of
+
+Here, we combine sea surface temperatures from geochemical proxies with numerical temperature estimated from mangroves and coral reefs, derived with a nearest living relative
+
+apply broadly analogous method to the marine record, deriving probability densities broadly reflecting the thermal preferences of modern coral reefs and mangrove taxa [@...], which we assign to fossil occurrences of coral reefs and mangroves.
+
+, and sometimes biological proxies such as leaf physiognomy,
+. This can involve interpolation between data points, or extrapolation if entire climate zones remain unsampled. Examples of this approach include t
 Combining quantitative proxy data with not fully quantified climate constraints, e.g.
@@
 provide an alternative that can be employed more readily by a broad range of workers.
-Examples of this approach include the visualisation of proxy compilations along a latitudinal gradient [@Hollis2019, @Vickers2021].
-Interpolation is sometimes used to bridge spatial gaps in palaeoclimate data [@Taylor2004], taking advantage of the autoregressive nature of climatic data: much of the information on the climate of any given location is contained in the climate data of nearby locations [@Reynolds1994].
-Adding to this, some proxy-based reconstructions use statistical modelling to infer climatic patterns.
-For example, polynomial regression has been used to reconstruct latitudinal temperature gradients [@Bijl2009], and 2D-reconstructions of surface temperatures have been created with Gaussian process regression [@Inglis2020].
 

### 2023-02-05 manuscript/manuscript.Rmd
Update manuscript.Rmd

@@
 
-Here, we present a Bayesian, hierarchical model for inferring sea surface temperatures (SST) that expands upon existing, spatial reconstructions of palaeoclimate by allowing for the integration of 1) prior information based on physical principles and the observed, modern SST distribution, and of 2) geochemical and ecological climate proxies in a common, quantitative framework.
+Here, we present a Bayesian, hierarchical model for inferring sea surface temperatures (SST) that expands upon existing, spatial reconstructions of palaeoclimate by allowing for the integration of 1) prior information based on physical principles and on the observed, modern SST distribution, and of 2) geochemical and ecological climate proxies in a common, quantitative framework.
 nearest living relative...
 
-We apply this approach to the early Eocene climatic optimum (EECO), the interval with the warmest sustained temperatures of the Cenozoic.
+We apply this approach to the early Eocene climatic optimum (EECO), the interval with the warmest sustained temperatures of the Cenozoic [@Pross2012].
+The early Eocene is the focus of a substantial body of palaeoclimate literature [@Berggren1998; @Thomas2000; @Hyland2017; @Lunt2021], as it potentially represents an analogue for extreme climate warming scenarios [@Burke2018].
+For palaeoclimate modellers, the early Eocene presents a challenge: Geochemical proxy data [@Hollis2019] and ecological data [@Marwick1994; @Greenwood1995] indicate tropical to subtropical conditions up to high latitudes, but early climate models have not been able to model warm polar temperatures and a shallow latitudinal temperature gradient under realistic conditions [@Sloan1990; ...].
+More recent climate models have been able to align their simulations with the proxy record [@Huber2013;...].
+A comparison of multiple proxy-based early Eocene reconstructions of global average temperatures @Inglis2020
+
 Assuming the conservation of thermal tolerances of modern coral reefs and mangrove taxa, we establish broad temperature ranges for EECO coral reef and mangrove sites.

### 2023-02-05 manuscript/manuscript.Rmd
update introduction

--- 2023-02-05 manuscript/manuscript.Rmd
@@
 
-Here, we present a Bayesian, hierarchical model for inferring sea surface temperatures (SST) that expands upon existing, spatial reconstructions of palaeoclimate by allowing for the integration of 1) prior information based on physical principles and the observed, modern SST distribution, and of 2) geochemical and ecological climate proxies in a common, quantitative framework.
+Here, we present a Bayesian, hierarchical model for inferring sea surface temperatures (SSTs) that expands upon existing, spatial reconstructions of palaeoclimate by allowing for the integration of 1) prior information based on physical principles and on the observed, modern SST distribution, and of 2) geochemical and ecological climate proxies in a common, quantitative framework.
 nearest living relative...
 
-We apply this approach to the early Eocene climatic optimum (EECO), the interval with the warmest sustained temperatures of the Cenozoic.
+We apply this approach to the record of the early Eocene climatic optimum (EECO), the interval with the warmest sustained temperatures of the Cenozoic [@Pross2012].
+The early Eocene is the focus of a substantial body of palaeoclimate literature [@Berggren1998; @Thomas2000; @Hyland2017; @Lunt2021], as it potentially represents an analogue for extreme climate warming scenarios [@Burke2018].
+For palaeoclimate modelers, the early Eocene presents a challenge: Geochemical proxy data [@Hollis2019] and ecological data [@Marwick1994; @Greenwood1995] indicate tropical to subtropical conditions up to high latitudes, but early climate models have not been able to model warm polar temperatures and a shallow latitudinal temperature gradient under realistic conditions [@Sloan1990; ...].
+Although some more recent climate models have been able to align their simulations with the proxy record [@Huber2013;...], the nature of the early Eocene temperature gradient remains a matter of debate.
+A relatively recent, proxy-based studies inferred a SST gradient that was shallower than the modern by 32+-10 % [@Evans2018], which would correspond to a gradient of ~20+-3 degC.
+Later studies based on new proxy compilations only estimate temperature gradients up to mid-latitudes [@Hollis2019], or combine SST proxies with other data [@Inglis2020], making it difficult to infer a meaningful estimate of the SST gradient.
+
 Assuming the conservation of thermal tolerances of modern coral reefs and mangrove taxa, we establish broad temperature ranges for EECO coral reef and mangrove sites.
@@
 
+-GP regression and interpolation can give rise to inverse gradients ect. --> our method nicely prevents this.
+
 **Comparison to Eocene proxy and modelling studies**.

### 2023-02-06 manuscript/discards.rmd
introduction and testing function on different shapes

@@
 ---
+
+## Introduction discards
+
+A relatively recent, proxy-based studies inferred a SST gradient that was shallower than the modern by 32+-10 % [@Evans2018], which would correspond to a gradient of ~20+-3 degC.
+Later studies based on new proxy compilations only estimate temperature gradients up to mid-latitudes [@Hollis2019], or combine SST proxies with other data [@Inglis2020], making it difficult to infer a meaningful estimate of the SST gradient.
+We therefore provide The early Eocene climatic optimum (EECO) showcases the
+
+is an .
+
+Assuming the conservation of thermal tolerances of modern coral reefs and mangrove taxa, we establish broad temperature ranges for EECO coral reef and mangrove sites.
+We integrate these temperature estimates with the EECO geochemical, shallow marine proxy record to model the latitudinal SST gradient and its uncertainty.
+
+We apply this model to the early Eocene climate optimum,...
+
+"nearest living relative" methods are used for deriving numerical climate estimates from taxonomic plant data for the Cenozoic  [@Faucette2007; @Chevalier2022].
+These methods work on the the premise that .
+Proxy-based, deep-time palaeoclimate reconstructions relying solely on geochemical proxies [e.g.
+@Veizer2015; @Grossman2022] have significant spatial gaps, and global temperature reconstructions based on those records alone may be biased [@Jones2022].
+Ecological and lithological data are sometimes used for quantitative climate reconstructions [@Uhl2009;], but are rarely used in conjunction with geochemical proxies [@Burgener2023].
+
+For example, a plethora of
+
+Here, we combine sea surface temperatures from geochemical proxies with numerical temperature estimated from mangroves and coral reefs, derived with a nearest living relative
+
+apply broadly analogous method to the marine record, deriving probability densities broadly reflecting the thermal preferences of modern coral reefs and mangrove taxa [@...], which we assign to fossil occurrences of coral reefs and mangroves.
+
+, and sometimes biological proxies such as leaf physiognomy,
+. This can involve interpolation between data points, or extrapolation if entire climate zones remain unsampled. Examples of this approach include t
+Combining quantitative proxy data with not fully quantified climate constraints, e.g.
+the presence of tropical weathering products such as bauxite and laterite, has the potential to improve upon existing, purely proxy-based reconstructions [@Scotese2021; @Burgener2023].
+Terrestrial and
+
+approach is still rarely employed in deep-time palaeoclimatic studies, as it This requires transforming ordinal or interval-bound data into numerical estimates, which is common
+
+Integrating this type of information into fully quantitative climate reconstructions requires
+
+data provide a different method of reconstructing palaeotemperatures
+Palaeobiological or geological climate indicators that provide only broad climate constraints which are hard to quantify, and 3) 
+[@Burgener2023]
+For the marine realm, Geochemical proxies such as oxygen isotopes (d18O), tetraether indices (TEX86) and magnesium calcium ratios (Mg/Ca) are used to derive palaeotemperature estimates for individual locations using experimentally derived transfer functions [e.g.
+@Liu2009].
+
+Climate data not based on geochemical, quantitative proxies generally provide broad categorical information, e.g.
+"tropical" or "dry", or broad quantitative constraints, e.g.
+"> 20 degC mean annual temperature".
+Integrating this information into fully quantitative climate reconstructions requires transforming it to a numerical scale.
+This approach is showcased by nearest living relative methods, commonly employed in terrestrial, Cenozoic climate reconstructions, which work on the premise that the climatic requirements of ancient plant taxa were similar to those of there nearest modern relatives [@Kvacek2007].
+A plethora of methods have been developed to derive numerical climate values from taxonomic data [@Faucette2007; @...].
+Approaches based on probability density functions have the advantage of naturally capturing the uncertainty associated with the climatic occupancy of modern nearest relatives [@Greenwood2017], and probability distributions can be readily integrated in a quantitative modelling framework.
+Here, we employ a similar approach, constructing probability distributions broadly reflecting the thermal preferences of modern coral reefs and mangrove taxa [@...], which we assign to fossil occurrences of coral reefs and mangroves.
+
+Combining quantitative proxy data with other climate constraints has the potential to improve upon climate reconstructions based on a single type of proxy [@Burgener2023], but is rarely employed in deep-time palaeoclimatic studies.
+
+Deep-time palaeoclimate reconstructions use geochemical proxy data
+
+The first challenge requires that all proxy data is converted to a common scale.
+
+A fundamental challenge in reconstructing Earth's past climate is to infer large-scale climatic patterns from local observations, derived from disparate proxies.  In order for the
+
+Proxy-derived palaeoclimate reconstructions offer an important alternative to and a test of Earth system models, and as such are fundamental for reconstructing Earth's past climate
+
+The geological record holds evidence for episodes in Earth history with fundamentally different climates than our current [@Scotese2021].
+Some episodes, for example the early Eocene, are thought to have been significantly warmer than the modern [@Pross2012], potentially representing an analogue for extreme climate warming scenarios [@Burke2018].
+Reconstructions of palaeoclimate commonly use geochemical proxy data, palaeobiological or lithological climate indicators, or earth system modelling.
+
+Geochemical proxies such as oxygen isotopes (d18O), tetraether indices (TEX86) and magnesium calcium ratios (Mg/Ca) are used to derive palaeotemperature estimates for individual locations using experimentally derived transfer functions [e.g.
+@Liu2009].
+Palaeobiological and lithological climate data provide a different method of reconstructing palaeotemperatures, using e.g.
+leaf physiognomy and floral assemblages to derive palaeotemperature proxies [@Greenwood2003], or using lithological indicators to determine the range of palaeoclimatic belts [@Scotese2021].
+To capture complex dependencies between proxies and climate data, machine learning approaches [@Chandra2021] or articifical neural networks [@Malmgren1997; @Lauchstedt2017] are sometimes used.
+Earth system models take a fundamentally different approach by modelling the climate-generating processes directly, reconstructing spatially resolved climate estimates for the entire globe [@???].
+However, this approach still relies on proxy data to constrain the large number of possible model setups and ground truth the model output [@???] It also requires substantial computing resources and technical expert knowledge, and is thus difficult to implement or modify for most palaeoclimate researchers.
+
+To make inferences on a regional or global scale, proxy-based climate reconstructions need some way of extrapolating a generally sparse record of local findings.
+Taking the mean of multiple local estimates to summarise regional or global climate variables [e.g.
+@Veizer2015] is a straightforward approach, but is likely to result in biased climate reconstructions [@Jones2022].
+Reconstructions that take into account the spatial distribution of climate data are...
+### HERE###
+provide an alternative that can be employed more readily by a broad range of workers.
+
+Here, we present a Bayesian, hierarchical model for inferring sea surface temperatures (SST) that expands upon existing, spatial reconstructions of palaeoclimate by allowing for the integration of 1) prior information based on physical principles and the observed, modern SST distribution, and of 2) geochemical and ecological climate proxies in a common, quantitative framework.
+We apply this approach to the early Eocene climatic optimum (EECO), the interval with the warmest sustained temperatures of the Cenozoic.
+Assuming the conservation of thermal tolerances of modern coral reefs and mangrove taxa, we establish broad temperature ranges for EECO coral reef and mangrove sites.
+We integrate these temperature estimates with the EECO geochemical, shallow marine proxy record to model the latitudinal SST gradient and its uncertainty.
+
+<!--
+
+Hook -> Generic problem:
+- Reconstructing past climate important for understanding current/future global warming
+- Two main pathways: highly complex ESM <-> proxy compilations
+- Problems with proxy data: only provides a local picture (few local sites interpreted as global signal)
+- Problems with ESM (long time to run, very complex, a lot of assumptions on processes/doesn't consider how they may change, need to be constrained with proxy data)
+- Lack of intermediate-complexity models to make full use of the proxy data
+
+Specific problem:
+- exemplified with the EECO: proxies and ESM hard to reconcile, EECO important climate analogue
+- a lot of proxies recrystallised <-> unreliable data or lack of data
+
+Solution:
+- Proxy-driven model (no strong assumptions on processes)
+- Integrate more data (quantify biological / geological information)
+- Integrate additional knowledge (constraints on such gradients)
+
+Application:
+- EECO time of extreme warming
+- problem of the Early Eocene:
+- apply solution to EECO to constrain EECO lat. T gradient
+
+Here:
+
+- EECO specific information
+- Disagreement between GCM + geochem data
+-->
+
+<!--
+
+The early Eocene climate optimum (EECO, ~ 53--49 million years ago) is thought to have been the warmest episode of [...].
+The EECO is of particular interest due to its use as an analogue for extreme warming scenarios, depicting a possible future high-CO2 world.
+To reconstruct the climate of EECO, a variety of proxy data and Earth System Models have been used.
+Early modelers have soon noticed that their models struggle to reproduce the climate indicated by proxy data: Without unrealistically high concentrations of atmospheric CO2, early climate models were unable to reproduce the equitable polar climates indicated by early Eocene proxy data, e.g.
+the occurrence of mangroves in polar latitudes, and d18O data from x indicates a temperature of ~ y at latitude z.
+
+Contrast between complex, process-driven Earth system models and very simplistic use of the proxy data as local temperature estimates or global averages.
+Here, we introduce a proxy-driven modelling approach of relatively low complexity, which places the available chemical and biological proxy data in a single climate model.
+In order to properly compare model output to proxy data, latitudinal temperature patterns need to be reconstructed from noisy data (impossible to properly compare model gradients with scattered proxy points "by eye")
+
+Reconstructing Earth's climate history is fundamental for understanding the evolution of the Earth system and its consequences for life in the past and future.
+The geological record, the only empirical source of information on the palaeoclimate of the deep past, holds evidence.
+
+Here, we present a Bayesian hierarchical model to integrate geological data---with established temperature proxies---into a unified quantitative framework.
+We apply this approach to the early Eocene climatic optimum (EECO), the interval with the warmest sustained temperatures of the Cenozoic.
+Assuming the conservation of thermal tolerances of modern coral reefs and mangrove taxa, we establish broad sea surface temperature ranges for EECO coral reef and mangrove sites.
+We integrate these temperature estimates with the EECO geochemical shallow marine proxy record to model the latitudinal temperature gradient and its uncertainty.
+
+Our hierarchical approach accounts for different sample sizes between and uncertainties within proxy sites, and
+
+-->
 

### 2023-02-06 manuscript/manuscript.Rmd
introduction and testing function on different shapes

--- 2023-02-05 manuscript/manuscript.Rmd
@@
 To exploit the full potential these data sets offer, two fundamental challenges need to be overcome: Disparate types of proxy data need to be integrated to inform a single climate reconstruction, and large-scale scale climate patterns need to be inferred from localised, often sparse observations.
-To address these challenges, we present a novel Bayesian method that combines quantitative proxies and palaeoecological constraints into a fully quantitative model of the latitudinal gradient of sea surface temperatures, bridging spatial gaps in sparsely sampled climate data.
+To address these challenges, we present a novel Bayesian model that combines quantitative proxies and palaeoecological constraints into a fully quantitative model of the latitudinal gradient of sea surface temperatures, bridging spatial gaps in sparsely sampled climate data.
 
-Existing palaeoclimate reconstructions are generally based on one of three approaches [see @Burgener2023]: 1) Quantitative, geochemical proxies such as oxygen isotopes (d18O), tetraether indices (TEX86) and magnesium calcium ratios (Mg/Ca) are used to derive palaeotemperature estimates using experimentally derived transfer functions [e.g.
+Existing palaeoclimate reconstructions are generally based on one of three approaches [see @Burgener2023]: 1) Quantitative, geochemical proxies such as oxygen isotopes (d18O), tetraether indices (TEX86) and magnesium calcium ratios (Mg/Ca) are used to infer palaeotemperature estimates using experimentally derived transfer functions [e.g.
 @Liu2009].
@@
 2) Palaeobiological, ecological and geological proxy data can provide broad climate constraints, but no experimentally determined transfer functions exist to directly infer quantitative climate parameters from these records [e.g.
-@Kvacek2007; @Cao2019; but see @Royer2012].
-Instead, palaeoclimate signals are inferred based on the climatic conditions of analogous, modern sediments, or based on the premise that the climatic requirements of ancient taxa, plant traits or ecological communities were similar to those of their nearest modern relatives [@Kvacek2007; @Greenwood2017].
+@Kvacek2007; @Cao2019].
+Instead, palaeoclimate signals are inferred based on the climatic conditions of analogous, modern sediments, or based on the premise that the climatic requirements of ancient taxa, plant traits or ecological communities were similar to those of their nearest modern relatives [@Kvacek2007; @Royer2012].
 3) Earth system models offer the most highly resolved and complete climate reconstructions, but ultimately rely on constraints provided by the geological and fossil record [@Lunt2021].
@@
 Even when using the full range of available records, proxy-based climate reconstructions beyond the Holocene are inevitably patchy [e.g.
-@Hollis2019], and global temperature reconstructions based on those records alone may be biased [@Jones2022].
-To infer regional or global-scale patterns from an incomplete record relies on the bridging of spatial gaps.
-A common approach that requires no additional computation is the spatial visualisation of proxy-derived temperatures against latitude, showing broad latitudinal temperature trends [@Hollis2019, @Vickers2021].
+@Hollis2019], and global temperature reconstructions based on those records alone are prone to bias [@Jones2022].
+Inferring regional or global-scale patterns from an incomplete record relies on the bridging of spatial gaps.
+A common approach that requires no additional computation is the spatial visualisation of proxy-derived temperatures against latitude, showing broad latitudinal temperature trends [@Hollis2019; @Vickers2021].
 Interpolation is sometimes used to bridge spatial gaps in palaeoclimate data [@Taylor2004], taking advantage of the autoregressive nature of climatic data: much of the information on the climate of any given location is contained in the climate data of nearby locations [@Reynolds1994].
@@
 
-Here, we present a Bayesian, hierarchical model for inferring sea surface temperatures (SST) that expands upon existing, spatial reconstructions of palaeoclimate by allowing for the integration of 1) prior information based on physical principles and the observed, modern SST distribution, and of 2) geochemical and ecological climate proxies in a common, quantitative framework.
-nearest living relative...
+Here, we present a Bayesian, hierarchical model for inferring the latitudinal gradient of sea surface temperatures (SSTs) that expands upon existing, spatially explicit reconstructions of palaeoclimate by allowing for the integration of 1) prior information based on physical principles and on the observed, modern SST distribution, and of 2) geochemical and ecological climate proxies in a common, quantitative framework.
+We use a generalised logistic function to infer the shape of the temperature gradient despite a patchy latitudinal coverage, and test the robustness of this method using a down-sampled record of modern SSTs.
+We apply the model to the record of the early Eocene climatic optimum (EECO), combining a compilation of geochemical proxies [@Hollis2019], mangrove communities [@Popescu2021], and coral reefs [@Zamagni], using a nearest-living-relative approach [see e.g.
+@Greenwood2017] to establish broad temperature ranges for the ecological data.
 
-We apply this approach to the early Eocene climatic optimum (EECO), the interval with the warmest sustained temperatures of the Cenozoic.
-Assuming the conservation of thermal tolerances of modern coral reefs and mangrove taxa, we establish broad temperature ranges for EECO coral reef and mangrove sites.
-We integrate these temperature estimates with the EECO geochemical, shallow marine proxy record to model the latitudinal SST gradient and its uncertainty.
-
-We apply this model to the early Eocene climate optimum,...
-
-"nearest living relative" methods are used for deriving numerical climate estimates from taxonomic plant data for the Cenozoic  [@Faucette2007; @Chevalier2022].
-These methods work on the the premise that .
-Proxy-based, deep-time palaeoclimate reconstructions relying solely on geochemical proxies [e.g.
-@Veizer2015; @Grossman2022] have significant spatial gaps, and global temperature reconstructions based on those records alone may be biased [@Jones2022].
-Ecological and lithological data are sometimes used for quantitative climate reconstructions [@Uhl2009;], but are rarely used in conjunction with geochemical proxies [@Burgener2023].
-
-For example, a plethora of
-
-Here, we combine sea surface temperatures from geochemical proxies with numerical temperature estimated from mangroves and coral reefs, derived with a nearest living relative
-
-apply broadly analogous method to the marine record, deriving probability densities broadly reflecting the thermal preferences of modern coral reefs and mangrove taxa [@...], which we assign to fossil occurrences of coral reefs and mangroves.
-
-, and sometimes biological proxies such as leaf physiognomy,
-. This can involve interpolation between data points, or extrapolation if entire climate zones remain unsampled. Examples of this approach include t
-Combining quantitative proxy data with not fully quantified climate constraints, e.g.
-the presence of tropical weathering products such as bauxite and laterite, has the potential to improve upon existing, purely proxy-based reconstructions [@Scotese2021; @Burgener2023].
-Terrestrial and
-
-approach is still rarely employed in deep-time palaeoclimatic studies, as it This requires transforming ordinal or interval-bound data into numerical estimates, which is common
-
-Integrating this type of information into fully quantitative climate reconstructions requires
-
-data provide a different method of reconstructing palaeotemperatures
-Palaeobiological or geological climate indicators that provide only broad climate constraints which are hard to quantify, and 3) 
-[@Burgener2023]
-For the marine realm, Geochemical proxies such as oxygen isotopes (d18O), tetraether indices (TEX86) and magnesium calcium ratios (Mg/Ca) are used to derive palaeotemperature estimates for individual locations using experimentally derived transfer functions [e.g.
-@Liu2009].
-
-Climate data not based on geochemical, quantitative proxies generally provide broad categorical information, e.g.
-"tropical" or "dry", or broad quantitative constraints, e.g.
-"> 20 degC mean annual temperature".
-Integrating this information into fully quantitative climate reconstructions requires transforming it to a numerical scale.
-This approach is showcased by nearest living relative methods, commonly employed in terrestrial, Cenozoic climate reconstructions, which work on the premise that the climatic requirements of ancient plant taxa were similar to those of there nearest modern relatives [@Kvacek2007].
-A plethora of methods have been developed to derive numerical climate values from taxonomic data [@Faucette2007; @...].
-Approaches based on probability density functions have the advantage of naturally capturing the uncertainty associated with the climatic occupancy of modern nearest relatives [@Greenwood2017], and probability distributions can be readily integrated in a quantitative modelling framework.
-Here, we employ a similar approach, constructing probability distributions broadly reflecting the thermal preferences of modern coral reefs and mangrove taxa [@...], which we assign to fossil occurrences of coral reefs and mangroves.
-
-Combining quantitative proxy data with other climate constraints has the potential to improve upon climate reconstructions based on a single type of proxy [@Burgener2023], but is rarely employed in deep-time palaeoclimatic studies.
-
-Deep-time palaeoclimate reconstructions use geochemical proxy data
-
-The first challenge requires that all proxy data is converted to a common scale.
-
-A fundamental challenge in reconstructing Earth's past climate is to infer large-scale climatic patterns from local observations, derived from disparate proxies.  In order for the
-
-Proxy-derived palaeoclimate reconstructions offer an important alternative to and a test of Earth system models, and as such are fundamental for reconstructing Earth's past climate
-
-The geological record holds evidence for episodes in Earth history with fundamentally different climates than our current [@Scotese2021].
-Some episodes, for example the early Eocene, are thought to have been significantly warmer than the modern [@Pross2012], potentially representing an analogue for extreme climate warming scenarios [@Burke2018].
-Reconstructions of palaeoclimate commonly use geochemical proxy data, palaeobiological or lithological climate indicators, or earth system modelling.
-
-Geochemical proxies such as oxygen isotopes (d18O), tetraether indices (TEX86) and magnesium calcium ratios (Mg/Ca) are used to derive palaeotemperature estimates for individual locations using experimentally derived transfer functions [e.g.
-@Liu2009].
-Palaeobiological and lithological climate data provide a different method of reconstructing palaeotemperatures, using e.g.
-leaf physiognomy and floral assemblages to derive palaeotemperature proxies [@Greenwood2003], or using lithological indicators to determine the range of palaeoclimatic belts [@Scotese2021].
-To capture complex dependencies between proxies and climate data, machine learning approaches [@Chandra2021] or articifical neural networks [@Malmgren1997; @Lauchstedt2017] are sometimes used.
-Earth system models take a fundamentally different approach by modelling the climate-generating processes directly, reconstructing spatially resolved climate estimates for the entire globe [@???].
-However, this approach still relies on proxy data to constrain the large number of possible model setups and ground truth the model output [@???] It also requires substantial computing resources and technical expert knowledge, and is thus difficult to implement or modify for most palaeoclimate researchers.
-
-To make inferences on a regional or global scale, proxy-based climate reconstructions need some way of extrapolating a generally sparse record of local findings.
-Taking the mean of multiple local estimates to summarise regional or global climate variables [e.g.
-@Veizer2015] is a straightforward approach, but is likely to result in biased climate reconstructions [@Jones2022].
-Reconstructions that take into account the spatial distribution of climate data are...
-### HERE###
-provide an alternative that can be employed more readily by a broad range of workers.
-
-Here, we present a Bayesian, hierarchical model for inferring sea surface temperatures (SST) that expands upon existing, spatial reconstructions of palaeoclimate by allowing for the integration of 1) prior information based on physical principles and the observed, modern SST distribution, and of 2) geochemical and ecological climate proxies in a common, quantitative framework.
-We apply this approach to the early Eocene climatic optimum (EECO), the interval with the warmest sustained temperatures of the Cenozoic.
-Assuming the conservation of thermal tolerances of modern coral reefs and mangrove taxa, we establish broad temperature ranges for EECO coral reef and mangrove sites.
-We integrate these temperature estimates with the EECO geochemical, shallow marine proxy record to model the latitudinal SST gradient and its uncertainty.
-
-<!--
-
-Hook -> Generic problem:
-- Reconstructing past climate important for understanding current/future global warming
-- Two main pathways: highly complex ESM <-> proxy compilations
-- Problems with proxy data: only provides a local picture (few local sites interpreted as global signal)
-- Problems with ESM (long time to run, very complex, a lot of assumptions on processes/doesn't consider how they may change, need to be constrained with proxy data)
-- Lack of intermediate-complexity models to make full use of the proxy data
-
-Specific problem:
-- exemplified with the EECO: proxies and ESM hard to reconcile, EECO important climate analogue
-- a lot of proxies recrystallised <-> unreliable data or lack of data
-
-Solution:
-- Proxy-driven model (no strong assumptions on processes)
-- Integrate more data (quantify biological / geological information)
-- Integrate additional knowledge (constraints on such gradients)
-
-Application:
-- EECO time of extreme warming
-- problem of the Early Eocene:
-- apply solution to EECO to constrain EECO lat. T gradient
-
-Here:
-
-- EECO specific information
-- Disagreement between GCM + geochem data
--->
-
-<!--
-
-The early Eocene climate optimum (EECO, ~ 53--49 million years ago) is thought to have been the warmest episode of [...].
-The EECO is of particular interest due to its use as an analogue for extreme warming scenarios, depicting a possible future high-CO2 world.
-To reconstruct the climate of EECO, a variety of proxy data and Earth System Models have been used.
-Early modelers have soon noticed that their models struggle to reproduce the climate indicated by proxy data: Without unrealistically high concentrations of atmospheric CO2, early climate models were unable to reproduce the equitable polar climates indicated by early Eocene proxy data, e.g.
-the occurrence of mangroves in polar latitudes, and d18O data from x indicates a temperature of ~ y at latitude z.
-
-Contrast between complex, process-driven Earth system models and very simplistic use of the proxy data as local temperature estimates or global averages.
-Here, we introduce a proxy-driven modelling approach of relatively low complexity, which places the available chemical and biological proxy data in a single climate model.
-In order to properly compare model output to proxy data, latitudinal temperature patterns need to be reconstructed from noisy data (impossible to properly compare model gradients with scattered proxy points "by eye")
-
-Reconstructing Earth's climate history is fundamental for understanding the evolution of the Earth system and its consequences for life in the past and future.
-The geological record, the only empirical source of information on the palaeoclimate of the deep past, holds evidence.
-
-Here, we present a Bayesian hierarchical model to integrate geological data---with established temperature proxies---into a unified quantitative framework.
-We apply this approach to the early Eocene climatic optimum (EECO), the interval with the warmest sustained temperatures of the Cenozoic.
-Assuming the conservation of thermal tolerances of modern coral reefs and mangrove taxa, we establish broad sea surface temperature ranges for EECO coral reef and mangrove sites.
-We integrate these temperature estimates with the EECO geochemical shallow marine proxy record to model the latitudinal temperature gradient and its uncertainty.
-
-Our hierarchical approach accounts for different sample sizes between and uncertainties within proxy sites, and
-
--->
+The early Eocene is the focus of a substantial body of palaeoclimate literature [@Berggren1998; @Thomas2000; @Hyland2017; @Lunt2021], as it is the interval with the warmest sustained temperatures of the Cenozoic [@Pross2012], and potentially represents an analogue for extreme climate warming scenarios [@Burke2018].
+For palaeoclimate modelers, the early Eocene presents a challenge: Geochemical proxy data [@Hollis2019] and ecological data [@Marwick1994; @Greenwood1995] indicate tropical to subtropical conditions up to high latitudes, but early climate models have not been able to model warm polar temperatures and a shallow latitudinal temperature gradient under realistic conditions [@Sloan1990; ...].
+Although some more recent climate models have been able to align their simulations with the proxy record [@Huber2013;...], the nature of the early Eocene temperature gradient remains a matter of debate.
+The inferred latitudinal gradients depend on the selection of proxies and on the method for dealing with spatial gaps and uneven sampling, and the inclusion or exclusion of terrestrial, shallow and deep marine data, and the differential reporting of the results complicate comparisons of different approaches [@Evans2018; @Hollis2019; @Inglis2020].
+Our goal is to provide a robust method for inferring temperature gradients that can be readily used with different proxy types and with very patchy records, and that fully quantifies the modeled temperature gradient with uncertainty.
+Applying this method to the EECO proxy record, we quantify the latitudinal SST gradient and the global mean SST.
 
 # Materials & Methods
+
+## Modern sea surface temperatures
 
@@
 
-**Eocene temperature gradients.**
-
 **Methodological advances.** Latitudinal temperature gradients have been estimated with a variety of parametric and non-parametric statistical techniques.
@@
 
+-GP regression and interpolation can give rise to inverse gradients ect. --> our method nicely prevents this.
+
+**Eocene temperature gradients.**
 **Comparison to Eocene proxy and modelling studies**.

### 2023-02-06 R/figures/SM/figS4.R

@@ -1,0 +1,90 @@
+### Supplementary figure: different gradient shapes
+source("R/functions/model_components/climate_model_modern.R")
+source("R/functions/model_components/gradient.R")
+source("R/functions/model_processing/combine_posterior.R")
+source("R/functions/auxiliary_functions.R")
+
+# source options for analysis
+source("R/options.R")
+#
+
+ylim = c(-2,32)
+
+## 1. Linear
+set.seed(1)
+lat <- runif(30,0,90)
+temp <- 30-lat/3
+m1 <- run_MCMC_simple(n_iter = 20000,
+                n_thin = 10,
+                x = lat,
+                y = temp,
+                prior_input = priors,
+                adapt_sd = NULL)
+r1 <- combine_posterior(list(m1))
+plot_chains(list(m1))
+par(mfrow=c(1,1))
+plot(lat,temp, xlab = expression("|latitude| ("*degree*")"), 
+     ylab = expression("temperature ("*degree*"C)"),
+     pch = 21, col = NA, bg = rgb(0,0,0,0.5), ylim = ylim)
+points(c(0,90),c(30,0),type = "l", lwd = 2)
+plot_gradient(r1,add = T, line_col = "red", confint_col = rgb(1,0,0,0.2))
+
+## 2. Flat
+set.seed(1)
+lat <- runif(30,0,90)
+temp <- rep(25,30)
+m1 <- run_MCMC_simple(n_iter = 20000,
+                      n_thin = 10,
+                      x = lat,
+                      y = temp,
+                      prior_input = priors,
+                      adapt_sd = NULL)
+r1 <- combine_posterior(list(m1))
+plot_chains(list(m1))
+par(mfrow=c(1,1))
+plot(lat,temp, xlab = expression("|latitude| ("*degree*")"), 
+     ylab = expression("temperature ("*degree*"C)"),
+     pch = 21, col = NA, bg = rgb(0,0,0,0.5), ylim = c(23,27))
+points(c(0,90),c(25,25),type = "l", lwd = 2)
+plot_gradient(r1,add = T, line_col = "red", confint_col = rgb(1,0,0,0.2))
+
+## 3. Quadratic
+set.seed(1)
+lat <- runif(30,0,90)
+temp <- 30-lat^2/270
+m1 <- run_MCMC_simple(n_iter = 20000,
+                      n_thin = 10,
+                      x = lat,
[…]

### 2023-02-06 manuscript/supplementary_material.Rmd
S4 figure

@@
 ```
+
+## S4. Different shapes of temperature gradients
+
+```{r out.width='100%', echo = FALSE, fig.cap = "Fig S4.
+Simulated latitudinal temperature gradients sampled at the absolute latitudes of the EECO data.
+The dark grey points show the simulated temperature data, the black line shows the shape of the true gradient, and the medians (lines) and 95% credibal intervals (shadings) of the inferred gradients using the model are shown in red.
+a) linear gradient; b) flat gradient; c) quadratic gradient; d) logit-shaped gradient.
+Except for d), which is very different from the modern and from reconstructed gradients, all latitudinal temperature gradients are well captured by the model in the latitudinal range for which data is provided."}
+setwd('..') # move up to parent folder of /manuscript
+knitr::include_graphics("D://Github/PalaeoClimateGradient/figures/SM/FigS4_different_gradients.png", error = FALSE)
+```

### 2023-02-06 R/figures/SM/figS4.R

@@ -8,83 +8,97 @@ source("R/functions/auxiliary_functions.R")
 source("R/options.R")
 #
 
-ylim = c(-2,32)
+ylim = c(0,30)
+xlim = c(0,90)
+lat <- abs(c(-59.42, -59.10, -54.13, -47.84, -47.84, -47.47, -46.14, -45.80, -45.80, -24.25,  -8.57,  -2.05,  -1.86,   0.49,   4.29,  14.67,
+  27.44,  31.00,  32.50,  32.80,  32.90,  35.54,  36.22,  39.02,  40.39,  40.63,  40.89,  43.79,  44.11,  50.04,  50.39,  58.45,
+  76.45,  77.13))
 
 ## 1. Linear
 set.seed(1)
-lat <- runif(30,0,90)
-temp <- 30-lat/3
+#lat <- runif(30,0,90)
+temp1 <- 30-lat/3
 m1 <- run_MCMC_simple(n_iter = 20000,
                 n_thin = 10,
                 x = lat,
-                y = temp,
+                y = temp1,
                 prior_input = priors,
                 adapt_sd = NULL)
 r1 <- combine_posterior(list(m1))
 plot_chains(list(m1))
-par(mfrow=c(1,1))
-plot(lat,temp, xlab = expression("|latitude| ("*degree*")"), 
-     ylab = expression("temperature ("*degree*"C)"),
-     pch = 21, col = NA, bg = rgb(0,0,0,0.5), ylim = ylim)
-points(c(0,90),c(30,0),type = "l", lwd = 2)
-plot_gradient(r1,add = T, line_col = "red", confint_col = rgb(1,0,0,0.2))
+
 
 ## 2. Flat
 set.seed(1)
-lat <- runif(30,0,90)
-temp <- rep(25,30)
-m1 <- run_MCMC_simple(n_iter = 20000,
+#lat <- runif(30,0,90)
+temp2 <- rep(25,length(lat))
+m2 <- run_MCMC_simple(n_iter = 20000,
                       n_thin = 10,
                       x = lat,
-                      y = temp,
+                      y = temp2,
                       prior_input = priors,
                       adapt_sd = NULL)
-r1 <- combine_posterior(list(m1))
+r2 <- combine_posterior(list(m2))
 plot_chains(list(m1))
-par(mfrow=c(1,1))
[…]

### 2023-02-07 manuscript/manuscript.Rmd
S4 figure
intro improvements and results on fig s4

@@
 ---
-title: "Bayesian multi-proxy reconstruction of latitudinal temperature gradients for the Early Eocene"
+title: "A Bayesian method for latitudinal temperature gradients in the early Eocene"
 
@@
 <!-- discarded titles:
+Bayesian multi-proxy reconstruction of latitudinal temperature gradients for the Early Eocene
 Bayesian multi-proxy reconstruction of the early Eocene latitudinal temperature gradient
@@
 Adding to this, some proxy-based reconstructions use statistical modelling to infer climatic patterns.
-For example, polynomial regression has been used to reconstruct latitudinal temperature gradients [@Bijl2009], and 2D-reconstructions of surface temperatures and precipitation have been created with Gaussian process regression [@Inglis2020; @Chandra2021].
+For example, polynomial regression [@Bijl2009] and cosine functions [@Inglis2020] have been used to reconstruct latitudinal temperature gradients, and 2D-reconstructions of surface temperatures have been created with Gaussian process regression [@Inglis2020].
+These approaches work well for interpolating relatively well-sampled data, but the absence of constraints on the modeled parameters means that such models can produce unrealistic temperature estimates when extrapolating from sparsely sampled data.
+Statistical modelling in the Bayesian framework helps overcome this problem by requiring the explicit specification of priors for the model parameters, which can be used to express physical constraints [@Chandra2021].
 
@@
 We use a generalised logistic function to infer the shape of the temperature gradient despite a patchy latitudinal coverage, and test the robustness of this method using a down-sampled record of modern SSTs.
-We apply the model to the record of the early Eocene climatic optimum (EECO), combining a compilation of geochemical proxies [@Hollis2019], mangrove communities [@Popescu2021], and coral reefs [@Zamagni], using a nearest-living-relative approach [see e.g.
+We apply the model to the record of the early Eocene climatic optimum (EECO), combining a compilation of geochemical proxies [@Hollis2019], mangrove communities [@Popescu2021], and coral reefs [@Zamagni2012], using a nearest-living-relative approach [see e.g.
 @Greenwood2017] to establish broad temperature ranges for the ecological data.
@@
 For palaeoclimate modelers, the early Eocene presents a challenge: Geochemical proxy data [@Hollis2019] and ecological data [@Marwick1994; @Greenwood1995] indicate tropical to subtropical conditions up to high latitudes, but early climate models have not been able to model warm polar temperatures and a shallow latitudinal temperature gradient under realistic conditions [@Sloan1990; ...].
-Although some more recent climate models have been able to align their simulations with the proxy record [@Huber2013;...], the nature of the early Eocene temperature gradient remains a matter of debate.
-The inferred latitudinal gradients depend on the selection of proxies and on the method for dealing with spatial gaps and uneven sampling, and the inclusion or exclusion of terrestrial, shallow and deep marine data, and the differential reporting of the results complicate comparisons of different approaches [@Evans2018; @Hollis2019; @Inglis2020].
-Our goal is to provide a robust method for inferring temperature gradients that can be readily used with different proxy types and with very patchy records, and that fully quantifies the modeled temperature gradient with uncertainty.
-Applying this method to the EECO proxy record, we quantify the latitudinal SST gradient and the global mean SST.
+Although some more recent climate models have been able to better align their simulations with the proxy record [@Huber2013;...], the nature of the early Eocene temperature gradient remains a matter of debate [@Tierney2017].
+The shape and magnitude of the inferred EECO latitudinal gradient depends on the selection of proxies, on the method for dealing with spatial gaps and uneven sampling, on the inclusion or exclusion of terrestrial, shallow and deep marine data, while differential reporting of the results complicates comparisons of different approaches [@Evans2018; @Hollis2019; @Inglis2020].
+Our goal is to provide a robust and relatively simple method for inferring and comparing temperature gradients that can be readily used with different proxy types and with patchy records.
+We use this model to quantify the latitudinal SST gradient and the global mean SST of the early Eocene, but the unspecific design of our method means it can be readily applied to other time intervals and proxy records.
 
@@
 
+Additional model runs with simulated SST data show that latitudinal gradients following a flat, linear or quadratic function can also be adequately reconstructed by our model (Fig.
+S4).
+The generalised logistic function underlying the reconstruction leads to the inferred gradient being pushed towards a sigmoidal shape in the absence of proxies (see very high latitudes in Fig S4a,c).
+This is intended, as in the absence of data, the shape of the reconstructed gradient should be broadly similar to the sigmoidal shape of the modern SST gradient, as it is the only empirically observed SST gradient.
+
 Modeling the latitudinal temperature gradient facilitates accurate reconstruction of global average temperatures.

### 2023-02-07 manuscript/manuscript.Rmd
add R2 modern, discussion first part

@@
 The shape and magnitude of the inferred EECO latitudinal gradient depends on the selection of proxies, on the method for dealing with spatial gaps and uneven sampling, on the inclusion or exclusion of terrestrial, shallow and deep marine data, while differential reporting of the results complicates comparisons of different approaches [@Evans2018; @Hollis2019; @Inglis2020].
-Our goal is to provide a robust and relatively simple method for inferring and comparing temperature gradients that can be readily used with different proxy types and with patchy records.
+Our goal is to provide a robust, yet simple method for inferring and comparing temperature gradients that can be readily used with different proxy types and with patchy records.
 We use this model to quantify the latitudinal SST gradient and the global mean SST of the early Eocene, but the unspecific design of our method means it can be readily applied to other time intervals and proxy records.
@@
 
+# modern R2
+R2_modern <- readRDS("results/modern/modern_model_empirical_R2.rds")
+
 # global average temperatures
@@
 The average modern temperature gradient can be closely approximated with our model when using the full modern SST dataset (Fig.
-3a).
+3a); almost all of the variation in the empirical median temperatures in bins of 1 deg absolute latitude is explained by the modeled gradient ($R^2 = `r R2_modern`$).
 Reducing the amount and the spatial coverage of data points to match the early Eocene sampling distribution ($n = 34$) still results in a close match of the modeled gradient with the empirical gradient (Fig.
@@
 S4).
-The generalised logistic function underlying the reconstruction leads to the inferred gradient being pushed towards a sigmoidal shape in the absence of proxies (see very high latitudes in Fig S4a,c).
+The generalised logistic function underlying the reconstruction leads to the inferred gradient being pushed towards a sigmoidal shape in the absence of proxies (see very high latitudes in Fig.
+S4a,c).
 This is intended, as in the absence of data, the shape of the reconstructed gradient should be broadly similar to the sigmoidal shape of the modern SST gradient, as it is the only empirically observed SST gradient.
@@
 # Discussion
+The Bayesian model presented herein accurately reconstructs the modern, latitudinal SST gradient from a patchy and relatively small (n = 34) sample of temperature data.
+This is an advancement over previously used linear or quadratic approximations [e.g.
+@Refs], which are inherently unable to capture the sigmoidal shape of the modern SST gradient, although our generalised logistic function is flexible enough to approximate those shapes, too.
+As such, our model presents an alternative to non-parametric methods for inferring latitudinal temperature gradients.
+Non-parametric models can very flexibly follow the shape of an unknown temperature gradient [e.g.
+@Zhang2019; @Jones2022], but have the disadvantage of having no model parameters that can be readily compared between data sets.
+In contrast, the four parameters that describe the shape of our generalised logistic model clearly relate to the magnitude, steepness and the latitude of greatest steepness of the modeled gradient.
+Given that at least vague prior information on the shape of latitudinal temperature gradients on Earth should exist for any geological time period, we argue that parametric methods should be preferred for reconstructing those gradients.
+
+The ease with which such prior information can be integrated is a major advantage of our method: The Bayesian framework allows to specify arbitrary priors on the four parameters that control the shape of the gradient (see Methods).
+A further improvement to existing methods is that the hierarchical structure of our model explicitly accounts for variation of temperature estimates at individual localities, and for uncertainty associated with the ecological temperature proxies.
+An extension of the model additionally accounts for measurement uncertainty of the geochemical proxies (Supplementary Materials).
 

### 2023-02-07 manuscript/supplementary_material.Rmd
add R2 modern, discussion first part

--- 2023-02-02 manuscript/supplementary_material.Rmd
@@
 Symbols with vertical lines show the median and 95% credible intervals of $\\mu$.
-The grey line and shading in the background show the latitudinal temperature gradient with the geochemical and ecological proxy data."}
+The blue line and shading in the background show the latitudinal temperature gradient with the geochemical and ecological proxy data."}
 setwd('..') # move up to parent folder of /manuscript
@@
 ```
+
+## S4. Different shapes of temperature gradients
+
+```{r out.width='100%', echo = FALSE, fig.cap = "Fig S4.
+Simulated latitudinal temperature gradients sampled at the absolute latitudes of the EECO data.
+The dark grey points show the simulated temperature data, the black line shows the shape of the true gradient, and the medians (lines) and 95% credibal intervals (shadings) of the inferred gradients using the model are shown in red.
+\\\n a) linear gradient; $SST = 30-|latitude|/3$ \\\n b) flat gradient; $SST = 25$ \\\n c) quadratic gradient; $SST = 30-|latitude|^2/270$ \\\n d) logit-shaped gradient; $SST = 14-3*log(|latitude|/(90-|latitude|))$ \\\n  Except for d), which is very unlike the modern gradient or reconstructed gradients, all latitudinal temperature gradients are well captured by the model in the latitudinal range for which data is provided."}
+setwd('..') # move up to parent folder of /manuscript
+knitr::include_graphics("D://Github/PalaeoClimateGradient/figures/SM/FigS4_different_gradients.png", error = FALSE)
+```

### 2023-02-07 R/analyses/assess_sdy.R

@@ -1,5 +1,8 @@
 # compare gradient sdy
 source("./R/functions/model_processing/combine_posterior.R")
+source("./R/functions/model_processing/temp_from_gradient.R")
+
+
 modern_gradient_eeco_lat <- readRDS("./results/modern/modern_sample_eocene_p_lat_gradient.RDS")
 modern_gradient_eeco_lat <- combine_posterior(mod = modern_gradient_eeco_lat, burnin = 5000)
 eeco_gradient <- readRDS("./results/eeco/eeco_climate_model_output_combined.rds")
@@ -20,3 +23,14 @@ saveRDS(modern_sdy, "results/modern/modern_sdy.rds")
 eeco_prox <- readRDS("results/eeco/eeco_climate_model_output_just_proxy.rds")
 eeco_prox <- combine_posterior(eeco_prox, burnin = 100000)
 eeco_prox_sdy <- as.list(format(round(quantile(eeco_prox$sdy,probs = c(.025,.5,.975)),1),nsmall = 1))
+
+# residual sum of squares modern gradient modeled vs zonal means
+mod_emp_med <- readRDS("results/modern/empirical_median_1-deg-lat.rds")
+
+lat <- seq(0.5,89.5,1)
+mod_med <- (temp_from_gradient(lat,modern_gradient))$median
+
+# coefficient of determination
+R2_modern <- 1-sum((mod_emp_med$SST - mod_med)^2)/sum((mod_emp_med$SST-mean(mod_emp_med$SST))^2)
+R2_modern <- format(round(R2_modern,3),nsmall = 1)
+saveRDS(R2_modern,"results/modern/modern_model_empirical_R2.rds")

### 2023-02-07 R/figures/SM/FigS2.R

@@ -32,7 +32,8 @@ col_proxgrad_shade <- rgb(0.85,0.6,0,0.27)
 xlab1 <- expression("|latitude| ("*degree*")")
 ylab1 <- expression("temperature ("*degree*"C)")
 lat_obs <- mode[[1]]$lat
-
+alpha2 <- 0.75
+col1 <- c(rgb(.9,.8,.5,alpha2), rgb(.8,.7,.5,alpha2), rgb(.7,.6,.5,alpha2), rgb(.65,.6,.6,alpha2))
 eecoprox <- readRDS("results/eeco/eeco_climate_model_output_just_proxy.rds")
 eecoproxcomb <- combine_posterior(eecoprox, burnin = 100000)
 
@@ -50,9 +51,8 @@ par(mar=c(4.2,4.2,0.5,0.5), mgp = c(2.5,0.8,0)
     ,las = 1)
 plot(obsmat$p_lat,obsmat$temperature, pch = pch1[proxind1], col = col1[proxind1], xlim = c(-2,92), ylim = c(14,41.5),
      xlab = xlab1, ylab = ylab1, cex = cex1[proxind1], type = "n", xaxs = "i", yaxs = "i")
-alpha2 <- 0.75
-col1 <- c(rgb(.9,.8,.5,alpha2), rgb(.8,.7,.5,alpha2), rgb(.7,.6,.5,alpha2), rgb(.65,.6,.6,alpha2))
-plot_gradient(mode_all,add=T, line_col = rgb(0,0.1,0.1,0.5), confint_col = rgb(0,0.1,0.1,0.15))
+
+plot_gradient(mode_all,add=T, line_col = rgb(0,0.3,0.8,0.5), confint_col = rgb(0,0.3,0.8,0.15))
 
 plot_gradient(eecoproxcomb,add=T, line_col = col_proxgrad, confint_col = col_proxgrad_shade)
 
@@ -63,6 +63,6 @@ points(lat_obs[1:23], mu_obs[1:23], col = col1[ind1_mu], pch = pch1[ind1_mu], lw
 sapply(1:23, function(x) points(rep(lat_obs[x],2), c(low_obs[x],upp_obs[x]), type = "l", lwd = 2, col = col1[ind1_mu[x]]))
 
 text(15,39.5,"only geochemical proxies", col = col_proxgrad, adj = 0)
[…]

### 2023-02-07 R/figures/fig_3.R

@@ -28,7 +28,8 @@ temp$bin <- ceiling(temp$lat) - 0.5
 grad <- tapply(temp$SST, temp$bin, median)
 grad <- data.frame(lat = as.numeric(names(grad)), 
                        SST = as.vector(grad))
-
+# save for later use
+saveRDS(grad,"results/modern/empirical_median_1-deg-lat.rds")
 # Calculate modern temperatures from sample
 lat <- seq(from = 0, to = 90, by = 0.1)
 modern_sample <- temp_from_gradient(lat = lat, model_out = modern_sample)

### 2023-02-07 manuscript/manuscript.Rmd
update discussion

@@
 This is an advancement over previously used linear or quadratic approximations [e.g.
-@Refs], which are inherently unable to capture the sigmoidal shape of the modern SST gradient, although our generalised logistic function is flexible enough to approximate those shapes, too.
+@Refs], which are inherently unable to capture the sigmoidal shape of the modern SST gradient.
+Our generalised logistic function is flexible enough to approximate the shapes of all of those functions.
 As such, our model presents an alternative to non-parametric methods for inferring latitudinal temperature gradients.
 Non-parametric models can very flexibly follow the shape of an unknown temperature gradient [e.g.
-@Zhang2019; @Jones2022], but have the disadvantage of having no model parameters that can be readily compared between data sets.
+@Zhang2019; @Jones2022], but are unsuitable for extrapolation, and have no model parameters that can be readily compared between data sets.
+When used for interpolation, non-paremetric methods tend to strictly respond to the data, even if the resulting patterns are unlike those predicted by physical processes [].
 In contrast, the four parameters that describe the shape of our generalised logistic model clearly relate to the magnitude, steepness and the latitude of greatest steepness of the modeled gradient.
@@
 The ease with which such prior information can be integrated is a major advantage of our method: The Bayesian framework allows to specify arbitrary priors on the four parameters that control the shape of the gradient (see Methods).
+In practice, those priors can be informed by physical constraints, e.g.
+restricting water temperatures to above the freezing point.
 A further improvement to existing methods is that the hierarchical structure of our model explicitly accounts for variation of temperature estimates at individual localities, and for uncertainty associated with the ecological temperature proxies.
-An extension of the model additionally accounts for measurement uncertainty of the geochemical proxies (Supplementary Materials).
 
@@
 -GP regression and interpolation can give rise to inverse gradients ect. --> our method nicely prevents this.
+
+An extension of the model additionally accounts for measurement uncertainty of the geochemical proxies (maybe cut, otherwise put something to prove that in SM).
 

### 2023-02-08 manuscript/manuscript.Rmd
discussion update

@@
 Our generalised logistic function is flexible enough to approximate the shapes of all of those functions.
-As such, our model presents an alternative to non-parametric methods for inferring latitudinal temperature gradients.
-Non-parametric models can very flexibly follow the shape of an unknown temperature gradient [e.g.
-@Zhang2019; @Jones2022], but are unsuitable for extrapolation, and have no model parameters that can be readily compared between data sets.
-When used for interpolation, non-paremetric methods tend to strictly respond to the data, even if the resulting patterns are unlike those predicted by physical processes [].
-In contrast, the four parameters that describe the shape of our generalised logistic model clearly relate to the magnitude, steepness and the latitude of greatest steepness of the modeled gradient.
-Given that at least vague prior information on the shape of latitudinal temperature gradients on Earth should exist for any geological time period, we argue that parametric methods should be preferred for reconstructing those gradients.
+As such, our model presents an alternative to non-parametric methods for inferring latitudinal temperature gradients, which are sometimes favoured as they can very flexibly follow the shape of an unknown temperature gradient [e.g.
+@Zhang2019; @Jones2022].
+However, when used for interpolation or prediction outside the proxy range, non-parametric methods such as Gaussian process regression strictly respond to the data [e.g.
+@Inglis2020], and the resulting large-scale patterns are informed by the idiosyncracies of a patchy proxy record.
 
+In parametric models, placing informative priors on parameters can improve interpolation and prediction.
+For example, the greater amount of solar radiation per unit area in low latitudes causes Earth's latitudinal temperature gradient to be broadly negative [@Beer2008].
+Prior information on the shape of latitudinal temperature gradients on Earth thus exists for all geological time periods, and should be used in any palaeoclimate reconstruction.
 The ease with which such prior information can be integrated is a major advantage of our method: The Bayesian framework allows to specify arbitrary priors on the four parameters that control the shape of the gradient (see Methods).
-In practice, those priors can be informed by physical constraints, e.g.
+In practice, those priors are informed by physical constraints, e.g.
 restricting water temperatures to above the freezing point.
 A further improvement to existing methods is that the hierarchical structure of our model explicitly accounts for variation of temperature estimates at individual localities, and for uncertainty associated with the ecological temperature proxies.
+
+Modelling the latitudinal temperature gradient from the equator to the poles facilitates reconstruction of the global mean temperature.
+
+Our parametric model has four parameters that describe the shape of the modeled gradient, and they clearly relate to its magnitude, steepness and the latitude of its greatest steepness.
 
@@
 An extension of the model additionally accounts for measurement uncertainty of the geochemical proxies (Supplementary Materials).
-
+ Disadvanand have no model parameters that can be readily compared between data sets. 
 -GP regression and interpolation can give rise to inverse gradients ect. --> our method nicely prevents this.

### 2023-02-08 manuscript/manuscript.Rmd
update discussion

@@
 Differences between Eocene and modern temperatures at a certain latitude were calculated by randomly pairing all iterations of the posterior from the Eocene and modern temperature gradient model, calculating the Eocene and modern temperature using the respective iterations, taking the difference, and then calculating the median ($2.5^{th}$, $97.5^{th}$ percentile) from all pairs of iterations.
-
 Global average temperatures were calculated by taking the weighted mean of the median ($2.5^{th}$, $97.5^{th}$ percentile) temperature estimates in 1 degree latitudinal bins.
@@
 
-Modeling the latitudinal temperature gradient facilitates accurate reconstruction of global average temperatures.
-The modern, global mean sea surface temperature estimate is `r motemp_global$"50%"` degC when using the full, modern data set.
-When reduced to the early Eocene sampling distribution, the modern global mean sea surface temperature is estimated at `r moeotemp_global$"50%"` (`r moeotemp_global$"2.5%"` - `r moeotemp_global$"97.5%"`) degC.
+Modeling the latitudinal temperature gradient facilitates accurate reconstruction of global mean surface temperatures (GMST).
+The modern, global mean sea surface temperature (GMsST) estimate is `r motemp_global$"50%"` degC when using the full, modern data set.
+When reduced to the early Eocene sampling distribution, the modern GMsST is estimated at `r moeotemp_global$"50%"` (`r moeotemp_global$"2.5%"` - `r moeotemp_global$"97.5%"`) degC.
 The deviations from the full modern estimate are again very low, with a 95 % credible interval of `r moeotemp_global_diff$"2.5%"` to `r moeotemp_global_diff$"97.5%"` degC.
@@
 
-The global early Eocene mean sea surface temperature is estimated at `r eotemp_global$"50%"` (`r eotemp_global$"2.5%"` - `r eotemp_global$"97.5%"`) degC, significantly higher than the modern (`r motemp_global$"50%"` degC).
+The early Eocene GMsST is estimated at `r eotemp_global$"50%"` (`r eotemp_global$"2.5%"` - `r eotemp_global$"97.5%"`) degC,  `r  round(as.numeric(eotemp_global$"50%") - as.numeric(motemp_global$"50%"),1)` degC  higher than the modern.
 
@@
 
-Modelling the latitudinal temperature gradient from the equator to the poles facilitates reconstruction of the global mean temperature.
+Palaeoclimate reconstructions are often summarised as global mean surface temperatures, to provide a standardised metric for characterising the state of the Earth's climate [@Royer2004;@Inglis2020].
+The calculation of global mean surface temperatures from sparse proxy data is susceptible to bias [@Jones2022].
+By modeling the temperature variation across latitudes, a complete temperature distribution is obtained, filling in gaps in the proxy record through inter- or extrapolation, eliminating the problem that specific climate zones may dominate the proxy record.
+Intersecting the modeled temperature gradient at narrow latitudinal intervals and accounting for the varying sizes of latitudinal bands by weighting provides unbiased GMST estimates.
+Recent deep-time GMST reconstructions are sometimes accounting for different sampling by calculating zonal averages, but assum [@Inglis2020].
+We anticipate that applying this method to intervals with a small or biased sample of temperature proxies will change Phanerozoic, proxy-based temperature curves [@Royer2004;@Veizer2015;@Jones2022].
+
+Estimating the GMST without bias is achieved by intersecting the modeled temperature gradient at narrow latitudinal intervals and adjusting for the differing sizes of latitudinal bands through weighting.
+
+Estimating the GMST from sparse proxies is prone to bias [@Jones2022], and
+
+Modeling the latitudinal temperature gradient provides a comprehensive temperature distribution across all latitudes, which allows to estimate GMST even
+
+Modeling the latitudinal temperature gradient provides a comprehensive temperature distribution across all latitudes, which can be used to improve estimates of the global mean temperature.
+Even if the underlying proxy data is sparse or incomplete, the model allows for effective utilization of data through interpolation and extrapolation techniques.
+The model also takes into account the different areas represented by each latitude band, weighting the temperatures accordingly, resulting in a more accurate estimate of the global mean temperature.
+
+popularity of reporting global mean temperature in paleoclimate studies is due to its utility as a comprehensive and standardized metric for characterizing and comparing the state of the Earth's climate and evaluating climate models.
+
+Modeling the latitudinal temperature gradient provides a comprehensive temperature distribution across all latitudes.
+This information can be used to improve estimates of the global mean temperature.
+Even if the underlying proxy data is sparse or incomplete, the model allows for effective utilization of data through interpolation and extrapolation techniques.
+The model also takes into account the different areas represented by each latitude band, weighting the temperatures accordingly, resulting in a more accurate estimate of the global mean temperature.
+
+Modeling the latitudinal temperature gradient provides a comprehensive representation of the temperature distribution across different latitudes.
+This information can be used to make more informed estimates of the global mean temperature.
+Even if the data is sparse or incomplete, the model allows for effective utilization of data through interpolation and extrapolation techniques.
+The model also takes into account the different areas represented by each latitude band, weighting the temperatures accordingly, resulting in a more accurate estimate of the global mean temperature.
 

### 2023-02-08 manuscript/manuscript.Rmd
update discussion

@@
 ---
-title: "A Bayesian method for latitudinal temperature gradients in the early Eocene"
+title: "A Bayesian model of latitudinal temperature gradients applied to the Early Eocene"
 
@@
 # Discussion
+## Improved reconstruction of temperature gradients and global average temperatures
 The Bayesian model presented herein accurately reconstructs the modern, latitudinal SST gradient from a patchy and relatively small (n = 34) sample of temperature data.
@@
 
-In parametric models, placing informative priors on parameters can improve interpolation and prediction.
+In our Bayesian, parametric model, informative priors on the model parameters improve the prediction of sea surface temperatures in the unsampled, very high latitudes: Notice that the upper limit of the credible interval does not increase beyond the range of the data, whereas unconstrained approaches such as splines, Gaussian processes or even standard linear regression could lead to unrealistically high upper bounds in this case [see @Rasmussen2008].
+Prior information on the shape of latitudinal temperature gradients on Earth exists for all geological time periods, and should be used in any palaeoclimate reconstruction.
 For example, the greater amount of solar radiation per unit area in low latitudes causes Earth's latitudinal temperature gradient to be broadly negative [@Beer2008].
-Prior information on the shape of latitudinal temperature gradients on Earth thus exists for all geological time periods, and should be used in any palaeoclimate reconstruction.
-The ease with which such prior information can be integrated is a major advantage of our method: The Bayesian framework allows to specify arbitrary priors on the four parameters that control the shape of the gradient (see Methods).
-In practice, those priors are informed by physical constraints, e.g.
-restricting water temperatures to above the freezing point.
-A further improvement to existing methods is that the hierarchical structure of our model explicitly accounts for variation of temperature estimates at individual localities, and for uncertainty associated with the ecological temperature proxies.
+The ease with which such prior information can be integrated is a major advantage of our method, as the shape of the modeled gradient is controlled by four parameters which clearly relate to its magnitude, steepness and the latitude of its greatest steepness.
 
-Palaeoclimate reconstructions are often summarised as global mean surface temperatures, to provide a standardised metric for characterising the state of the Earth's climate [@Royer2004;@Inglis2020].
-The calculation of global mean surface temperatures from sparse proxy data is susceptible to bias [@Jones2022].
-By modeling the temperature variation across latitudes, a complete temperature distribution is obtained, filling in gaps in the proxy record through inter- or extrapolation, eliminating the problem that specific climate zones may dominate the proxy record.
-Intersecting the modeled temperature gradient at narrow latitudinal intervals and accounting for the varying sizes of latitudinal bands by weighting provides unbiased GMST estimates.
-Recent deep-time GMST reconstructions are sometimes accounting for different sampling by calculating zonal averages, but assum [@Inglis2020].
-We anticipate that applying this method to intervals with a small or biased sample of temperature proxies will change Phanerozoic, proxy-based temperature curves [@Royer2004;@Veizer2015;@Jones2022].
+Palaeoclimate reconstructions are often summarised as global mean surface temperatures (GMST), providing a standardised metric for characterising the state of the Earth's climate [@Royer2004;@Inglis2020].
+However, the calculation of global mean surface temperatures from sparse proxy data is susceptible to bias [@Jones2022].
+By modeling the temperature variation across latitudes, a complete temperature distribution is obtained, filling in gaps in the proxy record through inter- or extrapolation.
+This eliminates the problem that specific climate zones may dominate the proxy record, as intersecting the modeled temperature gradient at narrow latitudinal intervals and accounting for the varying sizes of latitudinal bands by weighting provides unbiased GMST estimates.
+Some deep-time GMST reconstructions account for differential sampling by calculating zonal averages, but do not use the full latitudinal range of temperatures [@Inglis2020].
+We anticipate that applying this improved will significantly alter Phanerozoic, proxy-based temperature curves, particularly when applied to intervals with small or biased samples of temperature proxies.
+[@Royer2004;@Veizer2015;@Jones2022].
+
+## Incorporating ecological constraints in palaeoclimate reconstructions
+
+Our results show that incorporating quantified ecological temperature constraints can improve temperature reconstructions base on geochemical proxies alone.
+Given the significant body of research on the climatic significance of e.g.
+plants [@ @ @], reptiles [@ ] and lithology types [@ ], there is great potential for improving quantitative palaeoclimate reconstructions across the Phanerozoic.
+Although the quantification of each climate constraint needs to account for specific information, our modelling framework offers a straightforward way of integrating the quantified information with other proxy data: The hierarchical model structure accounts for variation of temperature estimates from proxies at individual localities, which is treated equivalent to the uncertainty associated with the ecological temperature proxies.
+A local temperature estimate, based on multiple geochemical proxies, thus has the same weight as a local temperature estimate obtained from the occurrence of a climate-sensitive plant community, whilst preserving the uncertainty associated with each estimate.
+
+As with other nearest-living relative methods, our ecological temperature estimates come with limitations: 1)
+
+However, our simple approach to constructing probability distributions for the thermal range of coral reefs has limitations.
+The use of only modern coral reefs and mangrove taxa as analogues for their fossil counterparts assumes that the thermal preferences of these taxa have remained unchanged over time, which may not always be the case.
+In addition, the use of only nearest living relatives as proxies assumes that past reefs may have thrived under the same temperatures as those observed today, which may not always be accurate.
+Climate change and shifting biogeographical patterns may have resulted in different thermal conditions for ancient reefs.
+Therefore, our approach should be supplemented with other methods that take into account changes in the thermal tolerance of reef taxa over time.
+
+Ecological data and other semi-quantiative or qualitative climate constraints are readily incorporated in this modelling framework: The hierarchical structure accounts for variation of temperature estimates from proxies at individual localities, which is treated equivalent to the uncertainty associated with the ecological temperature proxies.
+Existing palaeoclimate studies have mostly
+
+## Early Eocene climate
 
@@
 popularity of reporting global mean temperature in paleoclimate studies is due to its utility as a comprehensive and standardized metric for characterizing and comparing the state of the Earth's climate and evaluating climate models.
+
+: The Bayesian framework allows to specify arbitrary priors on the four parameters that control the shape of the gradient (see Methods).
+In practice, those priors are informed by physical constraints, e.g.
+restricting water temperatures to above the freezing point.
+A further improvement to existing methods is that the hierarchical structure of our model explicitly accounts for variation of temperature estimates at individual localities, and for uncertainty associated with the ecological temperature proxies.
 
@@
 # Conclusions
-
+In conclusion, the Bayesian hierarchical model presented in this study represents a significant improvement over existing methods for reconstructing past temperatures by incorporating both geochemical and ecological data.
+The use of a Bayesian framework allows for the specification of prior information based on physical constraints, which can improve interpolation and prediction.
+By modeling the temperature variation across latitudes, a complete temperature distribution is obtained, filling in gaps in the proxy record and reducing the risk of bias in global mean surface temperature (GMST) estimates.
+The hierarchical structure of the model accounts for the variation of temperature estimates at individual localities and uncertainty associated with the ecological temperature proxies.
+This improved method is expected to significantly alter the proxy-based temperature reconstructions of the past, particularly for intervals with small or biased samples of temperature proxies.
 # Acknowledgements

### 2023-02-09 manuscript/discards.rmd
discussion eco prox

@@
 
+## Discussion discards
+
+or even adjacent time intervals 
+the fossil record indicates that many taxa
+
+ This approach is showcased by nearest living relative methods, commonly employed in terrestrial, Cenozoic climate reconstructions, which work on the premise that the climatic requirements of ancient plant taxa were similar to those of there nearest modern relatives [@Kvacek2007].
+A plethora of methods have been developed to derive numerical climate values from taxonomic data [@Faucette2007; @...].
+Approaches based on probability density functions have the advantage of naturally capturing the uncertainty associated with the climatic occupancy of modern nearest relatives [@Greenwood2017], and probability distributions can be readily integrated in a quantitative modelling framework.
+Here, we employ a similar approach, constructing probability distributions broadly reflecting the thermal preferences of modern coral reefs and mangrove taxa [@...], which we assign to fossil occurrences of coral reefs and mangroves.
+
+Quantifying climatic constraints provided by fossil organisms relies on the assumption that the climatic conditions under which they lived are similar to those occupied by their modern relatives.
+
+ Similarly
+
+Our approach for quantifying climatic constraints as normal distributions broadely based on their modern, climatic equivalents, deviated from recent
+
+tQuantifying climatic constraints provided by fossil organisms
+
+relies on the assumption that the climatic conditions under which they lived are similar to those occupied by their modern relatives. Similarly
+
+Whilst our approach provides normal probability distributions for deriving temperature estimates of corals and mangrove communities,
+
+As with other nearest-living relative methods, our ecological temperature estimates come with limitations: 1)
+
+Although the quantification of each climate constraint needs to account for specific information, o
+
+However, our simple approach to constructing probability distributions for the thermal range of coral reefs has limitations.
+The use of only modern coral reefs and mangrove taxa as analogues for their fossil counterparts assumes that the thermal preferences of these taxa have remained unchanged over time, which may not always be the case.
+In addition, the use of only nearest living relatives as proxies assumes that past reefs may have thrived under the same temperatures as those observed today, which may not always be accurate.
+Climate change and shifting biogeographical patterns may have resulted in different thermal conditions for ancient reefs.
+Therefore, our approach should be supplemented with other methods that take into account changes in the thermal tolerance of reef taxa over time.
+
+Ecological data and other semi-quantiative or qualitative climate constraints are readily incorporated in this modelling framework: The hierarchical structure accounts for variation of temperature estimates from proxies at individual localities, which is treated equivalent to the uncertainty associated with the ecological temperature proxies.
+Existing palaeoclimate studies have mostly
+
 ## Condensed methods Bayesian framework

### 2023-02-09 manuscript/manuscript.Rmd
discussion eco prox

@@
 
-Our results show that incorporating quantified ecological temperature constraints can improve temperature reconstructions base on geochemical proxies alone.
-Given the significant body of research on the climatic significance of e.g.
-plants [@ @ @], reptiles [@ ] and lithology types [@ ], there is great potential for improving quantitative palaeoclimate reconstructions across the Phanerozoic.
-Although the quantification of each climate constraint needs to account for specific information, our modelling framework offers a straightforward way of integrating the quantified information with other proxy data: The hierarchical model structure accounts for variation of temperature estimates from proxies at individual localities, which is treated equivalent to the uncertainty associated with the ecological temperature proxies.
+Our results show that incorporating quantified ecological temperature constraints can provide more precise temperature reconstructions than geochemical proxies alone.
+Climate-sensitive plant communities [@ @ @], reptiles [@ ], leaf shapes [@], and lithologies [@ ] offer great potential for improving quantitative palaeoclimate reconstructions across the Phanerozoic.
+Our modelling framework offers a straightforward way of integrating the quantified information with other proxy data: The hierarchical model structure accounts for variation of temperature estimates from proxies at individual localities, which is treated equivalent to the uncertainty associated with the ecological temperature proxies.
 A local temperature estimate, based on multiple geochemical proxies, thus has the same weight as a local temperature estimate obtained from the occurrence of a climate-sensitive plant community, whilst preserving the uncertainty associated with each estimate.
 
-As with other nearest-living relative methods, our ecological temperature estimates come with limitations: 1)
-
-However, our simple approach to constructing probability distributions for the thermal range of coral reefs has limitations.
-The use of only modern coral reefs and mangrove taxa as analogues for their fossil counterparts assumes that the thermal preferences of these taxa have remained unchanged over time, which may not always be the case.
-In addition, the use of only nearest living relatives as proxies assumes that past reefs may have thrived under the same temperatures as those observed today, which may not always be accurate.
-Climate change and shifting biogeographical patterns may have resulted in different thermal conditions for ancient reefs.
-Therefore, our approach should be supplemented with other methods that take into account changes in the thermal tolerance of reef taxa over time.
-
-Ecological data and other semi-quantiative or qualitative climate constraints are readily incorporated in this modelling framework: The hierarchical structure accounts for variation of temperature estimates from proxies at individual localities, which is treated equivalent to the uncertainty associated with the ecological temperature proxies.
-Existing palaeoclimate studies have mostly
+Our approach for deriving fully quantitative climate reconstructions from ecological data is borrowed from nearest living relative methods, commonly employed in terrestrial, Cenozoic climate reconstructions [@Faucette2007; @Kvacek2007].
+These methods have the problem that the thermal preferences of taxa may have changed over time.
+More significantly, in the early Eocene, sea surface temperatures may have reached heights unknown in the modern world, and nearest living relative methods based on the modern are inherently unable to predict such elevated temperatures.
+This is especially true for taxa that inhabit the warmest part of the ocean today, e.g.
+coral reefs [@].
+Although coral reefs are threatened by warming sea surface temperatures today [@], it is conceivable that Eocene reef corals were adapted to a warmer climate.
+The fossil record indicates that reef development may have been stunted in the early Eocene, but the few early Eocene coral reefs occur in low latitudes [@Zamagni2012].
+Tropical temperatures predicted by the geochemical record indicate hotter-than-modern tropical temperatures (Fig.
+S2), suggesting that the modern climate range of coral reefs may underestimate the early Eocene thermal niche for coral reefs.
+We have tried to account for that possibility by widening the temperature probability distribution for reef corals, but the predicted temperatures for the reef and mangrove sites still lie below the temperatures indicated by the geochemical proxy record (Fig.
+4, Fig.
+S2).
+Future studies may try to resolve this dilemma by applying machine learning approaches to inferring quanitative temperatures from ecological and lithological proxies (e.g.
+@Salonen2019; @Wei2020; @Chandra2021) in novel way: If geochemical proxies and ecological or lithological data are derived from the same location and time interval, the climatic requirements of these taxa or sediments can be learned by using palaeotemperatures from that time interval, instead of or in addition to modern modern temperatures.
+This approach would ultimately rely on geochemical proxies providing absolute temperature estimates, but the resulting information could be exported to other locations lacking a good coverage of geochemical proxies via their fossil or sedimentary record.
 

### 2023-02-09 manuscript/manuscript.Rmd
further discussion

@@
 
+The latitudinal gradient is calculated as the difference between the modeled temperature at the equator (0 deg latitude) and at the poles (90 deg absolute latitude).
+Given the sigmoidal shape of the modern as well as the Eocene gradient (see Fig.
+4), this results in only slightly higher estimates of the gradient than when comparing e.g.
+the zonal average of equatorial and high-latitude temperatures, as is done in some earlier studies [e.g.
+@Evans2018].
+
 Differences between Eocene and modern temperatures at a certain latitude were calculated by randomly pairing all iterations of the posterior from the Eocene and modern temperature gradient model, calculating the Eocene and modern temperature using the respective iterations, taking the difference, and then calculating the median ($2.5^{th}$, $97.5^{th}$ percentile) from all pairs of iterations.
@@
 eocene_proxy_gradient <- readRDS("results/eeco/eocene_proxy_gradient.rds")
+eocene_gradient_polar_circle <- readRDS("./results/eeco/eocene_overall_gradient_polar_circle.rds")
 
@@
 The modeled Eocene temperature gradient is starkly different from the modern (Fig 4).
-Modelled, median equatorial temperatures are `r tdiff_lat0$"50%"` (`r tdiff_lat0$"2.5%"` - `r tdiff_lat0$"97.5%"`) degC higher for the Eocene, and polar temperatures are higher by `r tdiff_lat90$"50%"` (`r tdiff_lat90$"2.5%"` - `r tdiff_lat90$"97.5%"`) degC.
-This results in a flattened latitudinal temperature gradient of `r eocene_gradient$"50%"`  (`r eocene_gradient$"2.5%"` - `r eocene_gradient$"97.5%"`) degC for the Eocene, as opposed to `r modern_gradient$"50%"` degC for the modern.
+Modelled, median equatorial temperatures are `r tdiff_lat0$"50%"` (`r tdiff_lat0$"2.5%"` - `r tdiff_lat0$"97.5%"`) degC higher for the EECO, and polar temperatures are higher by `r tdiff_lat90$"50%"` (`r tdiff_lat90$"2.5%"` - `r tdiff_lat90$"97.5%"`) degC.
+This results in a flattened latitudinal temperature gradient of `r eocene_gradient$"50%"`  (`r eocene_gradient$"2.5%"` - `r eocene_gradient$"97.5%"`) degC for the EECO, as opposed to `r modern_gradient$"50%"` degC for the modern.
+To facilitate the comparison with latitudinal gradients reported in the literature, which sometimes do not report temperatures in very high latitudes, we report also the EECO gradient between the equator and the modern-day polar circle (66.6 deg), which is slightly lower at `r eocene_gradient_polar_circle$"50%"`  (`r eocene_gradient_polar_circle$"2.5%"` - `r eocene_gradient_polar_circle$"97.5%"`).
 
-The high variability of early Eocene palaeotemperature proxies, particularly in the mid-latitudes, and the scarcity of high-latitude data, result in substantial uncertainties in the modeled temperature gradient.
-This is reflected in the residual standard deviation ($\sigma$) of the early Eocene gradient, `r eeco_sdy$"50%"` (`r eeco_sdy$"2.5%"` - `r eeco_sdy$"97.5%"`) degC, more than twice as high than $\sigma$ of the modern gradient, `r modern_sdy$"50%"`.
+The high variability of EECO palaeotemperature proxies, particularly in the mid-latitudes, and the scarcity of high-latitude data, result in substantial uncertainties in the modeled temperature gradient.
+This is reflected in the residual standard deviation ($\sigma$) of the EECO gradient, `r eeco_sdy$"50%"` (`r eeco_sdy$"2.5%"` - `r eeco_sdy$"97.5%"`) degC, more than twice as high than $\sigma$ of the modern gradient, `r modern_sdy$"50%"`.
 This signifies that the early Eocene data fit less well to the logistic, latitudinal gradient model, which can also be seen from the drastic departure of some of the proxy data from the gradient estimates (Fig.
@@
 The fossil record indicates that reef development may have been stunted in the early Eocene, but the few early Eocene coral reefs occur in low latitudes [@Zamagni2012].
-Tropical temperatures predicted by the geochemical record indicate hotter-than-modern tropical temperatures (Fig.
+Tropical temperatures predicted by the geochemical proxy record indicate hotter-than-modern tropical temperatures (Fig.
 S2), suggesting that the modern climate range of coral reefs may underestimate the early Eocene thermal niche for coral reefs.
@@
 S2).
-Future studies may try to resolve this dilemma by applying machine learning approaches to inferring quanitative temperatures from ecological and lithological proxies (e.g.
+
+This dilemma could be resolved in future studies by applying machine learning approaches to inferring quanitative temperatures from ecological and lithological proxies (e.g.
 @Salonen2019; @Wei2020; @Chandra2021) in novel way: If geochemical proxies and ecological or lithological data are derived from the same location and time interval, the climatic requirements of these taxa or sediments can be learned by using palaeotemperatures from that time interval, instead of or in addition to modern modern temperatures.
@@
 ## Early Eocene climate
+The geochemical proxy record and ecological data agree that the latitudinal SST gradient of the early Eocene climatic optimum was significantly shallower than the modern [@Huber2011], but beyond that, there is little agreement.
+@Evans2018 estimate the SST gradient at 32+-10% of the modern, which would correspond to ~`r round( as.numeric(modern_gradient$"50%") - 0.32*as.numeric(modern_gradient$"50%"),0)`+- `r format(round( 0.10*as.numeric(modern_gradient$"50%"),0),nsmall=0)`degC, using the difference between the mean tropical and deep ocean data.
+The modeled SST of @Tierney2017 using TEX86 is ~ `r round(35.8-23.8,0)` degC, when taking the difference between temperatures at the polar circles and the equator.
+@Inglis2020 include both terrestrial air temperature and SST estimates in their model of the latitudinal temperature gradient in their Figure S4a; they model a polar circle to equatorial gradient of ~ `r round(30.7-(20.2+14.3)/2,0)`degC.
+All of those estimates are significantly lower than the SST gradient predicted by an EECO climate model ensemble (@Lunt2012, as shown in @Tierney2017), which predicts a polar circle to equatorial gradient of  `r round(31.5-5.5,0)`degC.
+Our polar circle to equatorial gradient is the lowest at `r eocene_gradient_polar_circle$"50%"`degC, although the 95% credible interval overlaps with the TEX86-based estimates of @Tierney2017 and the composite estimate of @Inglis2020.
+
+In latitudes beyond the polar circle, the discrepancy between our model predictions and those of earlier, proxy-based models increases, as those predict almost linearly decreasing SSTs towards the poles, whereas our median model prediction suggests only a slight decrease beyond the polar circle.
+However, the scarcity of temperature records in this range leads to widening credible intervals, including the possibility of stronger temperature decreases.
+Polar temperature estimates from our model thus conservative in that they admit large uncertainty where data is absent, which is desirable.
+However, the presence of high proxy-derived temperature estimates at ~ 60 deg latitudes forces the modeled median temperature curve to be too high at ~ 24 degC, relative to the temperatures indicated by the high-latitude mangrove communities (15.6 - 22.5 degC).
+
+These high-t proxies at high latitudes, combined with the overall high variability of the proxy record, is also the cause f...
+
+as shown in their Supplementary figure S4a would be ~ `r round(30.7-(13.3+5.0)/2,1)`degC when taking difference between temperatures at the equator and at the poles, but only ~ `r round(30.7-(20.2+14.3)/2,1)`degC when taking the difference between temperatures at the polar circles and the equator.
+This stark difference is due the almost linear gradient in mid - high latitudes modeled by the authors.
+
+on the exact nature of the EECO SST gradientProxy-based estimates put the EECO SST gradient at ~ `r round(30.7-(13.3+5.0)/2,1)` [@Inglis2020], at ~ `r as.numeric(modern_gradient$"50%") - 0.32*as.numeric(modern_gradient$"50%")` [@Evans2018]
 

### 2023-02-09 R/analyses/calculate_temperature_gradients_from_models.R

@@ -32,6 +32,12 @@ modern_eocene_t90 <- gradient(90,modm[,1:4])
 eocene_t90 <- gradient(90,mode[,1:4])
 eeco_prox_t90 <- gradient(90,eeco_prox[,1:4])
 
+eocene_t66.6 <- gradient(66.6,mode[,1:4])
+
+#quantile(eocene_t0-eocene_t66.6,probs = c(0.025,0.5,0.975))
+#quantile(eocene_t0-eocene_t90,probs = c(0.025,0.5,0.975))
+
+
 eeco_north_t0 <- gradient(0,eeco_north[,1:4])
 eeco_south_t0 <- gradient(0,eeco_south[,1:4])
 eeco_north_t90 <- gradient(90,eeco_north[,1:4])
@@ -50,6 +56,9 @@ eocene_gradient <- as.list(format(round(quantile(eocene_t0 - eocene_t90, probs =
 modern_gradient <- as.list(format(round(quantile(modern_t0 - modern_t90, probs = c(0.025,0.5,0.975)),1),nsmall = 1))
 modern_eocene_gradient <- as.list(format(round(quantile(modern_eocene_t0 - modern_eocene_t90, probs = c(0.025,0.5,0.975)),1),nsmall = 1))
 eocene_proxy_gradient <- as.list(format(round(quantile(eeco_prox_t0 - eeco_prox_t90, probs = c(0.025,0.5,0.975)),1),nsmall = 1))
+
+eocene_gradient_polar_circle <- as.list(format(round(quantile(eocene_t0 - eocene_t66.6, probs = c(0.025,0.5,0.975)),1),nsmall = 1))
+
 
 eocene_gradient_uncertainty <- quantile(eocene_t0 - eocene_t90, probs = c(0.975))-quantile(eocene_t0 - eocene_t90, probs = c(0.025))
 eocene_proxy_gradient_uncertainty <- quantile(eeco_prox_t0 - eeco_prox_t90, probs = c(0.975))-quantile(eeco_prox_t0 - eeco_prox_t90, probs = c(0.025))
@@ -74,6 +83,8 @@ saveRDS(tdiff_lat0, "./results/eeco/eocene_modern_difference_equator.rds")
 saveRDS(tdiff_lat90, "./results/eeco/eocene_modern_difference_poles.rds")
 saveRDS(eocene_gradient, "./results/eeco/eocene_overall_gradient.rds")
[…]

### 2023-02-09 R/analyses/run_EECO_climate_model.R

@@ -66,3 +66,33 @@ mode_all <- combine_posterior(mode,100000)
 saveRDS(mode_all, "results/eeco/eeco_climate_model_output_combined.rds")
 
 ### To check results use the "assess_EECO-climate_model_output_script.R
+
+
+### Run a model on just the priors
+nClust <- 4
+cl <- parallel::makeCluster(nClust)
+doParallel::registerDoParallel(cl)
+nChains = nClust # run 1 chain per cluster
+
+# source script
+modp <- foreach(pc = 1:nChains) %dopar% {
+  # call model functions
+  source("R/functions/model_components/climate_model_EECO.R")
+  # set random seed
+  set.seed(pc)
+  hierarchical_model(n_iter = 60000, n_thin = 10,
+                     obsmat = NULL, distrmat = NULL, 
+                     prior_input = priors,adapt_sd = 5000)
+}
+
+### stop cluster
+parallel::stopCluster(cl)
+
+### save output
+saveRDS(modp, "results/eeco/just_prior_model_output.rds")
+
+# combine chains and discard burn-in
+modp_all <- combine_posterior(modp,10000)
+# save combined chains
+saveRDS(modp_all, "results/eeco/just_prior_model_output.rds_combined.rds")
+

### 2023-02-09 R/figures/prior_posterior-GMST-Gradient.R

@@ -1,0 +1,78 @@
+# Supplementary Figure S2 - gradient on Northern and Southern hemisphere
+
+# load figures for plotting, distributions ect.
+source("./R/functions/auxiliary_functions.R")
+source("./R/functions/model_components/gradient.R")
+# read model output
+# Eocene
+mode_all <- readRDS("./results/eeco/eeco_climate_model_output_combined.rds")
+# prior
+modp_all <- readRDS("./results/eeco/just_prior_model_output.rds_combined.rds")
+# modern with eocene sampling
+modem_list <- readRDS("./results/modern/modern_sample_eocene_p_lat_gradient.rds")
+modem_all <- combine_posterior(modem_list,5000)
+
+
+
+# calculate temperatures at lat 0 and lat 90 for every iteration
+eo_t0 <- gradient(0,mode_all[,1:4])
+prior_t0 <- gradient(0,modp_all[,1:4])
+moeo_t0 <- gradient(0,modem_all[,1:4])
+
+eo_t90 <- gradient(90,mode_all[,1:4])
+prior_t90 <- gradient(90,modp_all[,1:4])
+moeo_t90 <- gradient(90,modem_all[,1:4])
+
+# gradient
+eo_grad <- eo_t0 - eo_t90
+prior_grad <- prior_t0 - prior_t90
+moeo_grad <- moeo_t0 - moeo_t90
+
+# densities
+d_eo <- density(eo_grad, n=length(eo_grad))
+d_prior <- density(prior_grad, n=length(prior_grad))
+d_moeo <- density(moeo_grad, n=length(moeo_grad))
+
+plot(0,0,type = "n", xlim = c(0,50), ylim = c(0,0.26), xaxs = "i", yaxs = "i", ylab = "posterior density")
+error_polygon(d_prior$x,d_prior$y,rep(0,length(d_prior$y)))
+points(d_prior$x, type = "l", col = rgb(0,0,0,0.4), lwd = 2)
+error_polygon(d_eo$x,d_eo$y,rep(0,length(d_eo$y)), col = rgb(1,0,0,0.2))
+points(d_eo, type = "l", col = rgb(1,0,0,0.4), lwd = 2)
+error_polygon(d_moeo$x,d_moeo$y,rep(0,length(d_moeo$y)), col = rgb(0,0.5,1,0.2))
+points(d_moeo, type = "l", col = rgb(0,0.5,1,0.4), lwd = 2)
+
+
[…]

### 2023-02-12 manuscript/manuscript.Rmd
discussion

@@
 However, the scarcity of temperature records in this range leads to widening credible intervals, including the possibility of stronger temperature decreases.
-Polar temperature estimates from our model thus conservative in that they admit large uncertainty where data is absent, which is desirable.
+Polar temperature estimates from our model are thus conservative in that they admit large uncertainty where data is absent, which is desirable.
 However, the presence of high proxy-derived temperature estimates at ~ 60 deg latitudes forces the modeled median temperature curve to be too high at ~ 24 degC, relative to the temperatures indicated by the high-latitude mangrove communities (15.6 - 22.5 degC).
 
-These high-t proxies at high latitudes, combined with the overall high variability of the proxy record, is also the cause f...
+The very high variability of the proxy record in mid-latitudes results in large uncertainties on the shape of temperature gradient and on the GMSsT, and demands closer scrutiny.
+Biases and errors in the proxy reconstructions likely contribute to the observed variability.
+A host of factors influences reconstructed temperatures from geochemical proxies [@Hollis2019].
+We tried to mitigate biases by excluding d18O measurements from recrystallised fossils, and by using the screened data set of @Hollis2019, but systematic offsets remain between mostly warm temperatures derived from TEX86, and cooler temperatures derived from d18O and delta47.
+Temporal averaging also likely contribute to the large variability of the EECO proxy data.
+The EECO spans ~ 4 million years, and
 

### 2023-02-12 manuscript/manuscript.Rmd
more discussion

--- 2023-02-09 manuscript/manuscript.Rmd
@@
 The early Eocene GMsST is estimated at `r eotemp_global$"50%"` (`r eotemp_global$"2.5%"` - `r eotemp_global$"97.5%"`) degC,  `r  round(as.numeric(eotemp_global$"50%") - as.numeric(motemp_global$"50%"),1)` degC  higher than the modern.
-
-A model run excluding the ecological proxies results in a similar median latitudinal gradient, `r eocene_proxy_gradient$"50%"` degC, but with a `r proxy_uncertain_diff_percent`% wider 95% CI (Supplementary materials).
+A model run excluding the ecological proxies results in a GMsST that is higher by `r eoceneproxy_global_diff$"50%"` (`r eoceneproxy_global_diff$"2.5%"` - `r eoceneproxy_global_diff$"97.5%"`) degC.
+The median latitudinal gradient is similar when excluding the ecological proxies, with a median of `r eocene_proxy_gradient$"50%"` degC, but with a `r proxy_uncertain_diff_percent`% wider 95% CI (Supplementary materials).
 This indicates that the ecological proxy data are broadly in agreement with the geochemical proxies, while providing additional constraints on the shape of the early Eocene temperature gradient.
@@
 
+[add table on lat gradients]
+
 In latitudes beyond the polar circle, the discrepancy between our model predictions and those of earlier, proxy-based models increases, as those predict almost linearly decreasing SSTs towards the poles, whereas our median model prediction suggests only a slight decrease beyond the polar circle.
 However, the scarcity of temperature records in this range leads to widening credible intervals, including the possibility of stronger temperature decreases.
-Polar temperature estimates from our model thus conservative in that they admit large uncertainty where data is absent, which is desirable.
+Polar temperature estimates from our model are thus conservative in that they admit large uncertainty where data is absent, which is desirable.
 However, the presence of high proxy-derived temperature estimates at ~ 60 deg latitudes forces the modeled median temperature curve to be too high at ~ 24 degC, relative to the temperatures indicated by the high-latitude mangrove communities (15.6 - 22.5 degC).
+The very high variability of the proxy record in mid-latitudes results in large uncertainties on the shape of temperature gradient and on the GMSsT.
+Biases and errors in the proxy reconstructions likely contribute to the observed variability, as geochemical proxies reflect many other factors besides seawater temperature [@Hollis2019].
+Despite excluding d18O measurements from recrystallised fossils, systematic offsets remain between mostly warm temperatures derived from TEX86, and cooler temperatures derived from d18O, delta47, and the ecological proxies.
+Seasonality [@Keating-Bitoni] and temporal changes within the EECO [@Westerhold2018] may also contribute to the large variability of the EECO proxy data.
 
-These high-t proxies at high latitudes, combined with the overall high variability of the proxy record, is also the cause f...
+GMSsT
+
+28 degC from bottom water temperatures
 
@@
 # Conclusions
+
+- focus on gradient rather than gmst - gmst can easily be calculated from gradient.
+
 In conclusion, the Bayesian hierarchical model presented in this study represents a significant improvement over existing methods for reconstructing past temperatures by incorporating both geochemical and ecological data.

### 2023-02-12 R/analyses/global_mean_temperatures.R

@@ -25,22 +25,26 @@ eotemp <- gradient(seq(0.5,89.5,1), mode_all[,1:4],0)
 # eocene temperature gradient quantiles
 eotempq <- apply(eotemp,2,function(x) quantile(x, probs = c(0.025,0.5,0.975)))
 # eocene global average temperature with 95% CI
-eotemp_global <- apply(eotempq,1, function(x) round(sum(x*latweight),1), simplify = FALSE)
+eotemp_global <- as.list(format(round(quantile(apply(eotemp,1, function(x) sum(x*latweight), simplify = TRUE),probs = c(0.025,0.5,0.975)),1),nsmall = 1))
 # modern temperature gradient
 motemp <- gradient(seq(0.5,89.5,1), modm_all[,1:4],0)
 # modern temperature gradient quantiles
 motempq <- apply(motemp,2,function(x) quantile(x, probs = c(0.025,0.5,0.975)))
 # modern global average temperature with 95% CI
-motemp_global <- apply(motempq,1, function(x) round(sum(x*latweight),1), simplify = FALSE)
+motemp_global <- as.list(format(round(quantile(apply(motemp,1, function(x) sum(x*latweight), simplify = TRUE),probs = c(0.025,0.5,0.975)),1),nsmall = 1))
 # modern eocene sampled temperature gradient
 moeotemp <- gradient(seq(0.5,89.5,1), modem_all[,1:4],0)
 # modern eocene sampled temperature gradient quantiles
 moeotempq <- apply(moeotemp,2,function(x) quantile(x, probs = c(0.025,0.5,0.975)))
 # modern eocene sampled global average temperature with 95% CI
-moeotemp_global <- apply(moeotempq,1, function(x) round(sum(x*latweight),1), simplify = FALSE)
-#
+moeotemp_global <- as.list(format(round(quantile(apply(moeotemp,1, function(x) sum(x*latweight), simplify = TRUE),probs = c(0.025,0.5,0.975)),1),nsmall = 1))
+
+
 # eocene proxy only temperature gradient
 eotempprox <- gradient(seq(0.5,89.5,1), eeco_prox[,1:4],0)
+# eocene proxy only global average temperature with 95% CI
+eotempprox_global <- as.list(format(round(quantile(apply(eotempprox,1, function(x) sum(x*latweight), simplify = TRUE),probs = c(0.025,0.5,0.975)),1),nsmall = 1))
+
 
 
[…]

### 2023-02-12 manuscript/manuscript.Rmd
SM fig update

@@
 All of those estimates are significantly lower than the SST gradient predicted by an EECO climate model ensemble (@Lunt2012, as shown in @Tierney2017), which predicts a polar circle to equatorial gradient of  `r round(31.5-5.5,0)`degC.
-Our polar circle to equatorial gradient is the lowest at `r eocene_gradient_polar_circle$"50%"`degC, although the 95% credible interval overlaps with the TEX86-based estimates of @Tierney2017 and the composite estimate of @Inglis2020.
+@Keating-Bitoni show a proxy-based SST gradient of ~ `r round(30-(19+15.6)/2,0)`degC, and @Bij2009 model a gradient of ~`r round(36-27,0)`degC in the northern and only ~`r 36-31`degC in the southern hemisphere.
+Our polar circle to equatorial gradient is lower than most previous estimates at `r eocene_gradient_polar_circle$"50%"`degC, although the 95% credible interval overlap with several of the listed gradient estimates.
 
@@
 However, the presence of high proxy-derived temperature estimates at ~ 60 deg latitudes forces the modeled median temperature curve to be too high at ~ 24 degC, relative to the temperatures indicated by the high-latitude mangrove communities (15.6 - 22.5 degC).
+In contrast, the extrapolated polar temperatures of most previous proxy-based models are likely too low, given the abundance of ecological data indicating temperate or subtropical high-latitude climates during the EECO [@ @]
+
 The very high variability of the proxy record in mid-latitudes results in large uncertainties on the shape of temperature gradient and on the GMSsT.

### 2023-02-12 manuscript/supplementary_material.Rmd
SM fig update

@@
 The dark grey points show the simulated temperature data, the black line shows the shape of the true gradient, and the medians (lines) and 95% credibal intervals (shadings) of the inferred gradients using the model are shown in red.
-\\\n a) linear gradient; $SST = 30-|latitude|/3$ \\\n b) flat gradient; $SST = 25$ \\\n c) quadratic gradient; $SST = 30-|latitude|^2/270$ \\\n d) logit-shaped gradient; $SST = 14-3*log(|latitude|/(90-|latitude|))$ \\\n  Except for d), which is very unlike the modern gradient or reconstructed gradients, all latitudinal temperature gradients are well captured by the model in the latitudinal range for which data is provided."}
+\\\n a) linear gradient; $SST = 30-|latitude|/3$ \\\n b) flat gradient; $SST = 25$ \\\n c) quadratic gradient; $SST = 30-|latitude|^2/270$ \\\n d) as c), but the priors on A, M, and B were relaxed to allow for a better fit to the quadratic gradient in high latitudes \\\n  All latitudinal temperature gradients are well captured by the model in the latitudinal range for which data is provided.
+If the priors, particularly the prior on the lower limit of the gradient (A), are relaxed (d), the model can approximate a quadratic or linear function even in high latitudes and at cold temperatures."}
 setwd('..') # move up to parent folder of /manuscript

### 2023-02-12 R/figures/SM/figS4.R

@@ -54,11 +54,14 @@ m3 <- run_MCMC_simple(n_iter = 20000,
 r3 <- combine_posterior(list(m3))
 plot_chains(list(m1))
 
-## 4. Inverse Sigmoidal
+## 4. Quadratic, relaxed A prior
+# relax prior on A
+priors$f1 <- function(x,log) dnorm(x,0,12,log)
+priors$f3 <- function(x,log) dnorm(x,45,25,log)
+priors$f4 <- function(x,log) dgamma(x,shape = 3,rate = 21,log=log)
 set.seed(1)
 #lat <- runif(30,0,90)
-temp4 <- 14-3*log(lat/(90-lat))
-plot(lat,temp4)
+temp4 <- 30-lat^2/270
 m4 <- run_MCMC_simple(n_iter = 20000,
                       n_thin = 10,
                       x = lat,
@@ -66,7 +69,22 @@ m4 <- run_MCMC_simple(n_iter = 20000,
                       prior_input = priors,
                       adapt_sd = NULL)
 r4 <- combine_posterior(list(m4))
-plot_chains(list(m4))
+plot_chains(list(m1))
+
+
+# ## 4. Inverse Sigmoidal
+# set.seed(1)
+# #lat <- runif(30,0,90)
+# temp4 <- 14-3*log(lat/(90-lat))
+# plot(lat,temp4)
+# m4 <- run_MCMC_simple(n_iter = 20000,
+#                       n_thin = 10,
+#                       x = lat,
+#                       y = temp4,
+#                       prior_input = priors,
+#                       adapt_sd = NULL)
+# r4 <- combine_posterior(list(m4))
+# plot_chains(list(m4))
 
 
 png("figures/SM/FigS4_different_gradients.png",width = 5,height = 4.5,unit = "in", res = 300)
@@ -76,14 +94,14 @@ plot(lat,temp1, xlab = expression("|latitude| ("*degree*")"),
      ylab = expression("temperature ("*degree*"C)"),
      pch = 21, col = NA, bg = rgb(0,0,0,0.5), ylim = ylim, xlim = xlim, cex = 1.4)
 points(c(0,90),c(30,0),type = "l", lwd = 2)
-plot_gradient(r1,add = T, line_col = "red", confint_col = rgb(1,0,0,0.2))
[…]

### 2023-02-23 manuscript/manuscript.Rmd
Update manuscript.Rmd

@@
 Applying the model separately to each hemispheres results in substantial differences in hemispherical, average temperatures, with the southern hemisphere being warmer by `r eeco_south_north_global_diff$"50%"` (`r eeco_south_north_global_diff$"2.5%"` - `r eeco_south_north_global_diff$"97.5%"`) degC.
-The inferred latitudinal gradient is somewhat steeper (`r eocene_north_south_gradient_diff$"50%"` degC, although the 95% CI spans `r eocene_north_south_gradient_diff$"2.5%"` to `r eocene_north_south_gradient_diff$"97.5%"` degC) in the northern hemisphere, but the large uncertainties associated with both gradients, and the lack of polar proxy data in the southern hemisphere preclude a more precise statement (Fig.
+The inferred latitudinal gradient is somewhat steeper in the northern hemisphere (`r eocene_north_south_gradient_diff$"50%"` degC, although the 95% CI spans `r eocene_north_south_gradient_diff$"2.5%"` to `r eocene_north_south_gradient_diff$"97.5%"` degC), but the large uncertainties associated with both gradients, and the lack of polar proxy data in the southern hemisphere preclude a more precise statement (Fig.
 S3).)
@@
 ## Early Eocene climate
-The geochemical proxy record and ecological data agree that the latitudinal SST gradient of the early Eocene climatic optimum was significantly shallower than the modern [@Huber2011], but beyond that, there is little agreement.
+The geochemical proxy record and ecological data indicate that the latitudinal SST gradient of the early Eocene climatic optimum was significantly shallower than the modern [@Huber2011], but beyond that, there is little agreement.
 @Evans2018 estimate the SST gradient at 32+-10% of the modern, which would correspond to ~`r round( as.numeric(modern_gradient$"50%") - 0.32*as.numeric(modern_gradient$"50%"),0)`+- `r format(round( 0.10*as.numeric(modern_gradient$"50%"),0),nsmall=0)`degC, using the difference between the mean tropical and deep ocean data.
@@
 
-GMSsT
+Our GMSsT estimate is
 

### 2023-02-24 manuscript/discards.rmd
gmst discussion

@@
 For details on the model structure and the choice of priors, see the *Supplementary Material*.
+
+## Discussion discards
+
+28 degC from bottom water temperatures
+
+as shown in their Supplementary figure S4a would be ~ `r round(30.7-(13.3+5.0)/2,1)`degC when taking difference between temperatures at the equator and at the poles, but only ~ `r round(30.7-(20.2+14.3)/2,1)`degC when taking the difference between temperatures at the polar circles and the equator.
+This stark difference is due the almost linear gradient in mid - high latitudes modeled by the authors.
+
+on the exact nature of the EECO SST gradientProxy-based estimates put the EECO SST gradient at ~ `r round(30.7-(13.3+5.0)/2,1)` [@Inglis2020], at ~ `r as.numeric(modern_gradient$"50%") - 0.32*as.numeric(modern_gradient$"50%")` [@Evans2018]
+
+Estimating the GMST without bias is achieved by intersecting the modeled temperature gradient at narrow latitudinal intervals and adjusting for the differing sizes of latitudinal bands through weighting.
+
+Estimating the GMST from sparse proxies is prone to bias [@Jones2022], and
+
+Modeling the latitudinal temperature gradient provides a comprehensive temperature distribution across all latitudes, which allows to estimate GMST even
+
+Modeling the latitudinal temperature gradient provides a comprehensive temperature distribution across all latitudes, which can be used to improve estimates of the global mean temperature.
+Even if the underlying proxy data is sparse or incomplete, the model allows for effective utilization of data through interpolation and extrapolation techniques.
+The model also takes into account the different areas represented by each latitude band, weighting the temperatures accordingly, resulting in a more accurate estimate of the global mean temperature.
+
+popularity of reporting global mean temperature in paleoclimate studies is due to its utility as a comprehensive and standardized metric for characterizing and comparing the state of the Earth's climate and evaluating climate models.
+
+: The Bayesian framework allows to specify arbitrary priors on the four parameters that control the shape of the gradient (see Methods).
+In practice, those priors are informed by physical constraints, e.g.
+restricting water temperatures to above the freezing point.
+A further improvement to existing methods is that the hierarchical structure of our model explicitly accounts for variation of temperature estimates at individual localities, and for uncertainty associated with the ecological temperature proxies.
+
+Modeling the latitudinal temperature gradient provides a comprehensive temperature distribution across all latitudes.
+This information can be used to improve estimates of the global mean temperature.
+Even if the underlying proxy data is sparse or incomplete, the model allows for effective utilization of data through interpolation and extrapolation techniques.
+The model also takes into account the different areas represented by each latitude band, weighting the temperatures accordingly, resulting in a more accurate estimate of the global mean temperature.
+
+Modeling the latitudinal temperature gradient provides a comprehensive representation of the temperature distribution across different latitudes.
+This information can be used to make more informed estimates of the global mean temperature.
+Even if the data is sparse or incomplete, the model allows for effective utilization of data through interpolation and extrapolation techniques.
+The model also takes into account the different areas represented by each latitude band, weighting the temperatures accordingly, resulting in a more accurate estimate of the global mean temperature.
+
+Our parametric model has four parameters that describe the shape of the modeled gradient, and they clearly relate to its magnitude, steepness and the latitude of its greatest steepness.
+
+**Methodological advances.** Latitudinal temperature gradients have been estimated with a variety of parametric and non-parametric statistical techniques.
+Parametric methods include polynomials [@Greenwood1995; @Bijl2009] and a cosine function [@Inglis2020].
+Whilst some of those functions are flexible enough to model a range of empirical temperature gradients, the logistic function used herein has the added advantage of an easily understandable relationship between each of the model parameters and the shape of the latitudinal gradient (Fig.
+2).
+Non-parametric models can very flexibly follow the shape of an unknown temperature gradient [@Jones2022, but see @Zhang2019], but have the disadvantage of having no model parameters that can be compared between data sets.
+Parametric methods should therefore be preferred to non-parametric temperature gradient models, as long as very broad prior information on the shape of the gradients exists.
+The ease with which such prior information can be integrated is perhaps the major advantage of our method: The Bayesian framework allows to specify arbitrary priors on the four parameters that control the upper and lower temperature limits, the steepness of the gradient, and the latitudinal position of the steepest point in the gradient (see Methods).
+A further improvement to existing methods is that the hierarchical structure of our model explicitly accounts for variation of temperature estimates at individual localities, and for uncertainty associated with the ecological temperature proxies.
+An extension of the model additionally accounts for measurement uncertainty of the geochemical proxies (Supplementary Materials).
+ Disadvanand have no model parameters that can be readily compared between data sets. 
+-GP regression and interpolation can give rise to inverse gradients ect. --> our method nicely prevents this.
+
+An extension of the model additionally accounts for measurement uncertainty of the geochemical proxies (maybe cut, otherwise put something to prove that in SM).
+
+**Eocene temperature gradients.**
+**Comparison to Eocene proxy and modelling studies**.
+
+A major theme of the discussions on the early Eocene greenhouse climate has been the discrepancy between proxy records, particularly ecological climate indicators such as macrofloral assemblage data, have
+
+- x1 compare lat gradient to literature estimates, e.g. Zhang2019
+- compare lat gradient to climate models and the "early EECO problem"
+- discuss insufficiency of proxies (high variability, ...)
+- discuss utility of the model and future applications

### 2023-02-24 manuscript/manuscript.Rmd
gmst discussion

@@
 All of those estimates are significantly lower than the SST gradient predicted by an EECO climate model ensemble (@Lunt2012, as shown in @Tierney2017), which predicts a polar circle to equatorial gradient of  `r round(31.5-5.5,0)`degC.
-@Keating-Bitoni show a proxy-based SST gradient of ~ `r round(30-(19+15.6)/2,0)`degC, and @Bij2009 model a gradient of ~`r round(36-27,0)`degC in the northern and only ~`r 36-31`degC in the southern hemisphere.
+@Keating-Bitoni show a proxy-based SST gradient of ~ `r round(30-(19+15.6)/2,0)`degC, and @Bij2009 model a gradient of ~`r round(36-27,0)`degC in the northern and only ~`r 36-31`degC in the southern hemisphere for the early Eocene.
 Our polar circle to equatorial gradient is lower than most previous estimates at `r eocene_gradient_polar_circle$"50%"`degC, although the 95% credible interval overlap with several of the listed gradient estimates.
@@
 
-Our GMSsT estimate is
-
-28 degC from bottom water temperatures
-
-as shown in their Supplementary figure S4a would be ~ `r round(30.7-(13.3+5.0)/2,1)`degC when taking difference between temperatures at the equator and at the poles, but only ~ `r round(30.7-(20.2+14.3)/2,1)`degC when taking the difference between temperatures at the polar circles and the equator.
-This stark difference is due the almost linear gradient in mid - high latitudes modeled by the authors.
-
-on the exact nature of the EECO SST gradientProxy-based estimates put the EECO SST gradient at ~ `r round(30.7-(13.3+5.0)/2,1)` [@Inglis2020], at ~ `r as.numeric(modern_gradient$"50%") - 0.32*as.numeric(modern_gradient$"50%")` [@Evans2018]
-
-Estimating the GMST without bias is achieved by intersecting the modeled temperature gradient at narrow latitudinal intervals and adjusting for the differing sizes of latitudinal bands through weighting.
-
-Estimating the GMST from sparse proxies is prone to bias [@Jones2022], and
-
-Modeling the latitudinal temperature gradient provides a comprehensive temperature distribution across all latitudes, which allows to estimate GMST even
-
-Modeling the latitudinal temperature gradient provides a comprehensive temperature distribution across all latitudes, which can be used to improve estimates of the global mean temperature.
-Even if the underlying proxy data is sparse or incomplete, the model allows for effective utilization of data through interpolation and extrapolation techniques.
-The model also takes into account the different areas represented by each latitude band, weighting the temperatures accordingly, resulting in a more accurate estimate of the global mean temperature.
-
-popularity of reporting global mean temperature in paleoclimate studies is due to its utility as a comprehensive and standardized metric for characterizing and comparing the state of the Earth's climate and evaluating climate models.
-
-: The Bayesian framework allows to specify arbitrary priors on the four parameters that control the shape of the gradient (see Methods).
-In practice, those priors are informed by physical constraints, e.g.
-restricting water temperatures to above the freezing point.
-A further improvement to existing methods is that the hierarchical structure of our model explicitly accounts for variation of temperature estimates at individual localities, and for uncertainty associated with the ecological temperature proxies.
-
-Modeling the latitudinal temperature gradient provides a comprehensive temperature distribution across all latitudes.
-This information can be used to improve estimates of the global mean temperature.
-Even if the underlying proxy data is sparse or incomplete, the model allows for effective utilization of data through interpolation and extrapolation techniques.
-The model also takes into account the different areas represented by each latitude band, weighting the temperatures accordingly, resulting in a more accurate estimate of the global mean temperature.
-
-Modeling the latitudinal temperature gradient provides a comprehensive representation of the temperature distribution across different latitudes.
-This information can be used to make more informed estimates of the global mean temperature.
-Even if the data is sparse or incomplete, the model allows for effective utilization of data through interpolation and extrapolation techniques.
-The model also takes into account the different areas represented by each latitude band, weighting the temperatures accordingly, resulting in a more accurate estimate of the global mean temperature.
-
-Our parametric model has four parameters that describe the shape of the modeled gradient, and they clearly relate to its magnitude, steepness and the latitude of its greatest steepness.
-
-**Methodological advances.** Latitudinal temperature gradients have been estimated with a variety of parametric and non-parametric statistical techniques.
-Parametric methods include polynomials [@Greenwood1995; @Bijl2009] and a cosine function [@Inglis2020].
-Whilst some of those functions are flexible enough to model a range of empirical temperature gradients, the logistic function used herein has the added advantage of an easily understandable relationship between each of the model parameters and the shape of the latitudinal gradient (Fig.
-2).
-Non-parametric models can very flexibly follow the shape of an unknown temperature gradient [@Jones2022, but see @Zhang2019], but have the disadvantage of having no model parameters that can be compared between data sets.
-Parametric methods should therefore be preferred to non-parametric temperature gradient models, as long as very broad prior information on the shape of the gradients exists.
-The ease with which such prior information can be integrated is perhaps the major advantage of our method: The Bayesian framework allows to specify arbitrary priors on the four parameters that control the upper and lower temperature limits, the steepness of the gradient, and the latitudinal position of the steepest point in the gradient (see Methods).
-A further improvement to existing methods is that the hierarchical structure of our model explicitly accounts for variation of temperature estimates at individual localities, and for uncertainty associated with the ecological temperature proxies.
-An extension of the model additionally accounts for measurement uncertainty of the geochemical proxies (Supplementary Materials).
- Disadvanand have no model parameters that can be readily compared between data sets. 
--GP regression and interpolation can give rise to inverse gradients ect. --> our method nicely prevents this.
-
-An extension of the model additionally accounts for measurement uncertainty of the geochemical proxies (maybe cut, otherwise put something to prove that in SM).
-
-**Eocene temperature gradients.**
-**Comparison to Eocene proxy and modelling studies**.
-
-A major theme of the discussions on the early Eocene greenhouse climate has been the discrepancy between proxy records, particularly ecological climate indicators such as macrofloral assemblage data, have
-
-- x1 compare lat gradient to literature estimates, e.g. Zhang2019
-- compare lat gradient to climate models and the "early EECO problem"
-- discuss insufficiency of proxies (high variability, ...)
-- discuss utility of the model and future applications
+Recent GMSsT estimates of the EECO and of the early Eocene range from 23.4 to 37.1 degC, with the lowest GMSTs being derived from d18O, and the higher estimates including TEX86 [@Inglis2020].
+Many studies include both marine and terrestrial proxies to derive GMST estimates, but despite great differences in proxy selection and in the calculation of global average temperatures, most estimates fall in the range of 27 - 29.5 degC [@@@@], similar to our GMSsT estimate of `r eotemp_global$"50%"` degC.
 

### 2023-02-28 manuscript/manuscript.Rmd
add references to bib

@@
 
-Existing palaeoclimate reconstructions are generally based on one of three approaches [see @Burgener2023]: 1) Quantitative, geochemical proxies such as oxygen isotopes (d18O), tetraether indices (TEX86) and magnesium calcium ratios (Mg/Ca) are used to infer palaeotemperature estimates using experimentally derived transfer functions [e.g.
-@Liu2009].
-These data can be readily processed for regional or global-scale palaeoclimate reconstructions [@...].
-2) Palaeobiological, ecological and geological proxy data can provide broad climate constraints, but no experimentally determined transfer functions exist to directly infer quantitative climate parameters from these records [e.g.
-@Kvacek2007; @Cao2019].
-Instead, palaeoclimate signals are inferred based on the climatic conditions of analogous, modern sediments, or based on the premise that the climatic requirements of ancient taxa, plant traits or ecological communities were similar to those of their nearest modern relatives [@Kvacek2007; @Royer2012].
+Existing palaeoclimate reconstructions are generally based on one of three approaches [see @Burgener2023]: 1) Geochemical proxies such as oxygen isotopes (d18O), tetraether indices (TEX86) and magnesium calcium ratios (Mg/Ca) provide quantitative palaeotemperature estimates using transfer functions based on experimental work and field observations [e.g.
+@Liu2009; @Wuchter2004].
+2) Palaeobiological, ecological and lithological proxy data  provide climate constraints, but no experimentally determined transfer functions exist to directly infer quantitative climate parameters from these records [e.g.
+@Wolfe1995; @Kvavcek2007; @Cao2019].
+Instead, palaeoclimate signals are inferred based on the climatic conditions of analogous, modern sediments, or based on the premise that the climatic requirements of ancient taxa, plant traits or ecological communities were similar to those of their nearest modern relatives [@Peppe2011; @Royer2012].
 3) Earth system models offer the most highly resolved and complete climate reconstructions, but ultimately rely on constraints provided by the geological and fossil record [@Lunt2021].
@@
 We use a generalised logistic function to infer the shape of the temperature gradient despite a patchy latitudinal coverage, and test the robustness of this method using a down-sampled record of modern SSTs.
-We apply the model to the record of the early Eocene climatic optimum (EECO), combining a compilation of geochemical proxies [@Hollis2019], mangrove communities [@Popescu2021], and coral reefs [@Zamagni2012], using a nearest-living-relative approach [see e.g.
+We apply this model to the record of the early Eocene climatic optimum (EECO), combining a compilation of geochemical proxies [@Hollis2019], mangrove communities [@Popescu2021], and coral reefs [@Zamagni2012], using a nearest-living-relative approach [see e.g.
 @Greenwood2017] to establish broad temperature ranges for the ecological data.
@@
 The early Eocene is the focus of a substantial body of palaeoclimate literature [@Berggren1998; @Thomas2000; @Hyland2017; @Lunt2021], as it is the interval with the warmest sustained temperatures of the Cenozoic [@Pross2012], and potentially represents an analogue for extreme climate warming scenarios [@Burke2018].
-For palaeoclimate modelers, the early Eocene presents a challenge: Geochemical proxy data [@Hollis2019] and ecological data [@Marwick1994; @Greenwood1995] indicate tropical to subtropical conditions up to high latitudes, but early climate models have not been able to model warm polar temperatures and a shallow latitudinal temperature gradient under realistic conditions [@Sloan1990; ...].
-Although some more recent climate models have been able to better align their simulations with the proxy record [@Huber2013;...], the nature of the early Eocene temperature gradient remains a matter of debate [@Tierney2017].
+For palaeoclimate modelers, the early Eocene presents a challenge: Geochemical proxy data [@Hollis2019] and ecological data [@Markwick1994; @Greenwood1995] indicate tropical to subtropical conditions up to high latitudes, but early climate models have not been able to model warm polar temperatures and a shallow latitudinal temperature gradient under realistic conditions [@Sloan1990].
+Although more recent climate models have been able to better align their simulations with the proxy record [@Huber2011], the nature of the early Eocene temperature gradient remains a matter of debate [@Tierney2017].
 The shape and magnitude of the inferred EECO latitudinal gradient depends on the selection of proxies, on the method for dealing with spatial gaps and uneven sampling, on the inclusion or exclusion of terrestrial, shallow and deep marine data, while differential reporting of the results complicates comparisons of different approaches [@Evans2018; @Hollis2019; @Inglis2020].
@@
 
-Geochemical climate proxy data were extracted from a latest Paleocene and early Eocene compilation [@hollis2019].
+Geochemical climate proxy data were extracted from a latest Paleocene and early Eocene compilation [@Hollis2019].
 This compilation provides four different geochemical proxies for reconstructing seawater temperature: $\delta^18^O, \Delta$_47_, Mg/Ca and TEX_86_.
@@
 After data filtering, x geochemical proxy samples remained.
-For a detailed description of each proxy see [@hollis2019].
+For a detailed description of each proxy see [@Hollis2019].
 
@@
 Nevertheless, the distribution of coral reefs have frequently been recognised as tracers of past (sub-)tropical conditions [@ziegler1984; @kiessling2001].
-During the Eocene, coral communities and reefs expanded across tropical and temperate latitudes, with some communities found up to palaeolatitudes of 43 degC N [@zamagni2012].
-Using a compilation of Paleocene--early Eocene coral reefs and community localities [@zamagni2012], we generate quantitative sea surface temperature estimates for the ECCO.
+During the Eocene, coral communities and reefs expanded across tropical and temperate latitudes, with some communities found up to palaeolatitudes of 43 degC N [@Zamagni2012].
+Using a compilation of Paleocene--early Eocene coral reefs and community localities [@Zamagni2012], we generate quantitative sea surface temperature estimates for the ECCO.
 To do so, we extract localities inferred to be Ilerdian (early Eocene) coral reefs from the compilation, and that can be confidently assigned to the EECO.
@@
 
-In our Bayesian, parametric model, informative priors on the model parameters improve the prediction of sea surface temperatures in the unsampled, very high latitudes: Notice that the upper limit of the credible interval does not increase beyond the range of the data, whereas unconstrained approaches such as splines, Gaussian processes or even standard linear regression could lead to unrealistically high upper bounds in this case [see @Rasmussen2008].
+In our Bayesian, parametric model, informative priors on the model parameters improve the prediction of sea surface temperatures in the unsampled, very high latitudes: Notice that the upper limit of the credible interval does not increase beyond the range of the data, whereas unconstrained approaches such as splines, Gaussian processes or even standard linear regression could lead to unrealistically high upper bounds in this case [see @Rasmussen2004].
 Prior information on the shape of latitudinal temperature gradients on Earth exists for all geological time periods, and should be used in any palaeoclimate reconstruction.
@@
 
-Our approach for deriving fully quantitative climate reconstructions from ecological data is borrowed from nearest living relative methods, commonly employed in terrestrial, Cenozoic climate reconstructions [@Faucette2007; @Kvacek2007].
+Our approach for deriving fully quantitative climate reconstructions from ecological data is borrowed from nearest living relative methods, commonly employed in terrestrial, Cenozoic climate reconstructions [@Fauquette2007; @Pross2012].
 These methods have the problem that the thermal preferences of taxa may have changed over time.

### 2023-02-28 manuscript/manuscript.Rmd
corrections and refs

@@
 
-Our results show that incorporating quantified ecological temperature constraints can provide more precise temperature reconstructions than geochemical proxies alone.
-Climate-sensitive plant communities [@ @ @], reptiles [@ ], leaf shapes [@], and lithologies [@ ] offer great potential for improving quantitative palaeoclimate reconstructions across the Phanerozoic.
-Our modelling framework offers a straightforward way of integrating the quantified information with other proxy data: The hierarchical model structure accounts for variation of temperature estimates from proxies at individual localities, which is treated equivalent to the uncertainty associated with the ecological temperature proxies.
+Our results exemplify how incorporating quantified ecological temperature constraints can provide more precise temperature reconstructions than geochemical proxies alone.
+Combining climate-sensitive plant communities [@ @ @], reptiles [@ ] and leaf shapes [@] with geochemical proxies offers substantial potential for improving quantitative palaeoclimate reconstructions across the Phanerozoic.
+Our modelling framework offers a straightforward way of integrating ecological climate data with other proxy data: The hierarchical model structure accounts for variation of temperature estimates from proxies at individual localities, which is treated equivalent to the uncertainty associated with the ecological temperature proxies.
 A local temperature estimate, based on multiple geochemical proxies, thus has the same weight as a local temperature estimate obtained from the occurrence of a climate-sensitive plant community, whilst preserving the uncertainty associated with each estimate.
@@
 
-This dilemma could be resolved in future studies by applying machine learning approaches to inferring quanitative temperatures from ecological and lithological proxies (e.g.
-@Salonen2019; @Wei2020; @Chandra2021) in novel way: If geochemical proxies and ecological or lithological data are derived from the same location and time interval, the climatic requirements of these taxa or sediments can be learned by using palaeotemperatures from that time interval, instead of or in addition to modern modern temperatures.
-This approach would ultimately rely on geochemical proxies providing absolute temperature estimates, but the resulting information could be exported to other locations lacking a good coverage of geochemical proxies via their fossil or sedimentary record.
+This dilemma could be resolved in future studies by applying machine learning approaches to inferring quantitative temperatures from ecological and lithological proxies (e.g.
+@Salonen2019; @Wei2020; @Chandra2021) in a novel way: If geochemical proxies and ecological or lithological data are derived from the same location and time interval, the climatic requirements of these taxa or sediments can be learned by using palaeotemperatures from that time interval, instead of or in addition to modern modern temperatures.
+This approach would ultimately rely on geochemical proxies providing absolute temperature estimates, but the resulting information could be exported to other locations lacking good coverage of geochemical proxies via their fossil or sedimentary record.
 
@@
 @Inglis2020 include both terrestrial air temperature and SST estimates in their model of the latitudinal temperature gradient in their Figure S4a; they model a polar circle to equatorial gradient of ~ `r round(30.7-(20.2+14.3)/2,0)`degC.
-All of those estimates are significantly lower than the SST gradient predicted by an EECO climate model ensemble (@Lunt2012, as shown in @Tierney2017), which predicts a polar circle to equatorial gradient of  `r round(31.5-5.5,0)`degC.
-@Keating-Bitoni show a proxy-based SST gradient of ~ `r round(30-(19+15.6)/2,0)`degC, and @Bij2009 model a gradient of ~`r round(36-27,0)`degC in the northern and only ~`r 36-31`degC in the southern hemisphere for the early Eocene.
+All of those estimates are significantly lower than the SST gradient predicted by an EECO climate model ensemble [@Lunt2012, as shown in @Tierney2017], which predicts a polar circle to equatorial gradient of  `r round(31.5-5.5,0)`degC.
+@Keating-Bitonti2011 show a proxy-based SST gradient of ~ `r round(30-(19+15.6)/2,0)`degC, and @Bijl2009 model a gradient of ~`r round(36-27,0)`degC in the northern and only ~`r 36-31`degC in the southern hemisphere for the early Eocene.
 Our polar circle to equatorial gradient is lower than most previous estimates at `r eocene_gradient_polar_circle$"50%"`degC, although the 95% credible interval overlap with several of the listed gradient estimates.
@@
 Despite excluding d18O measurements from recrystallised fossils, systematic offsets remain between mostly warm temperatures derived from TEX86, and cooler temperatures derived from d18O, delta47, and the ecological proxies.
-Seasonality [@Keating-Bitoni] and temporal changes within the EECO [@Westerhold2018] may also contribute to the large variability of the EECO proxy data.
+Seasonality [@Keating-Bitonti2011 and temporal changes within the EECO [@Westerhold2018] may also contribute to the large variability of the EECO proxy data.
 

### 2023-03-01 manuscript/manuscript.Rmd
manuscript and ref corrections

@@
 ---
-title: "A Bayesian model of latitudinal temperature gradients applied to the Early Eocene"
+title: "Bayesian multi-proxy reconstruction of latitudinal temperature gradients applied to the Early Eocene"
 
@@
 <!-- discarded titles:
+A Bayesian model of latitudinal temperature gradients applied to the Early Eocene
 Bayesian multi-proxy reconstruction of latitudinal temperature gradients for the Early Eocene
@@
 ===============================================================================
-Geochemical data such as oxygen isotopes (d18O), tetraether indices (TEX86) and magnesium calcium ratios (Mg/Ca) are routinely used to reconstruct ocean temperatures in deep time.
+Accurately reconstructing large-scale palaeoclimate patterns from sparse local records is critical for understanding the evolution of Earth's climate.
+Challenges arise particularly from the patchiness and uneven sampling of palaeoclimate data, and from the disparate nature of the proxy data.
+Geochemical data usually provide temperature estimates via transfer functions informed by experiments, and transfer functions based on the climatic requirements of modern taxa exist for some fossil plant assemblages.
+In contrast, most ecological and lithological data, e.g.
+coral reefs and evaporites, only convey information on broad climatic requirements.
+Historically, most large-scale proxy-based reconstructions have been based on either geochemical or ecological data, but few approaches exist for combining multiple proxy types into a single, quantitative reconstruction.
+Large spatial gaps in existing proxy records have often been bridged by simple averaging, without taking into consideration the spatial distribution of the samples, leading to biased temperature reconstructions.
+Here, we present a Bayesian hierarchical model to integrate ecological data with established geochemical proxies into a unified quantitative framework, bridging gaps in the latitudinal coverage of proxy data.
+We apply this approach to the early Eocene climatic optimum (EECO), the interval with the warmest sustained temperatures of the Cenozoic.
+Assuming the conservation of thermal tolerances of modern coral reefs and mangrove taxa, we establish broad sea surface temperature ranges for EECO coral reef and mangrove sites.
+We integrate these temperature estimates with the EECO geochemical shallow marine proxy record to model the latitudinal sea surface temperature gradient and global average temperatures.
+Our results confirm the presence of a flattened latitudinal temperature gradient and unusually high polar temperatures during the EECO, which is supported by high-latitude ecological data.
+We show that the inclusion of ecological data and adequate prior information can substantially reduce uncertainty in palaeoclimate reconstructions, and allow for unbiased temperature estimates from sparse data.
+
+<!--Geochemical data such as oxygen isotopes (d18O), tetraether indices (TEX86) and magnesium calcium ratios (Mg/Ca) are routinely used to reconstruct ocean temperatures in deep time.
 However, limited spatial coverage and disagreements between geochemical proxies compromise the accuracy of these reconstructions.
@@
 We show that the inclusion of ecological data can substantially reduce the uncertainty on temperature estimates in climate zones lacking geochemical data.
+-->
 
@@
 @Wolfe1995; @Kvavcek2007; @Cao2019].
-Instead, palaeoclimate signals are inferred based on the climatic conditions of analogous, modern sediments, or based on the premise that the climatic requirements of ancient taxa, plant traits or ecological communities were similar to those of their nearest modern relatives [@Peppe2011; @Royer2012].
+Instead, palaeoclimate signals are inferred based on the climatic conditions of analogous, modern sediments [@Chandra2021], or based on the premise that the climatic requirements of ancient taxa, plant traits or ecological communities were similar to those of their nearest modern relatives [@Peppe2011; @Royer2012; @Salonen2019].
 3) Earth system models offer the most highly resolved and complete climate reconstructions, but ultimately rely on constraints provided by the geological and fossil record [@Lunt2021].
@@
 
-The early Eocene is the focus of a substantial body of palaeoclimate literature [@Berggren1998; @Thomas2000; @Hyland2017; @Lunt2021], as it is the interval with the warmest sustained temperatures of the Cenozoic [@Pross2012], and potentially represents an analogue for extreme climate warming scenarios [@Burke2018].
+The early Eocene is the focus of a substantial body of palaeoclimate literature [e.g.
+@Berggren1998; @Thomas2000; @Hyland2017; @Lunt2021], as it is the interval with the warmest sustained temperatures of the Cenozoic [@Pross2012], and potentially represents an analogue for extreme climate warming scenarios [@Burke2018].
 For palaeoclimate modelers, the early Eocene presents a challenge: Geochemical proxy data [@Hollis2019] and ecological data [@Markwick1994; @Greenwood1995] indicate tropical to subtropical conditions up to high latitudes, but early climate models have not been able to model warm polar temperatures and a shallow latitudinal temperature gradient under realistic conditions [@Sloan1990].
@@
 Geochemical climate proxy data were extracted from a latest Paleocene and early Eocene compilation [@Hollis2019].
-This compilation provides four different geochemical proxies for reconstructing seawater temperature: $\delta^18^O, \Delta$_47_, Mg/Ca and TEX_86_.
+This compilation provides four different geochemical proxies for reconstructing seawater temperature: $\delta^{18}O$, $\Delta_{47}$, Mg/Ca and $TEX_{86}$.
 For our analyses, this dataset was restricted to the EECO and samples from the continental shelf.
-Recrystallised $\delta^18^O samples were also excluded as secondary diagenetic calcite precipitated after deposition can bias isotope measurements and offset temperature values [@schrag1999].
+Recrystallised $\delta^18^O$ samples were also excluded as secondary diagenetic calcite precipitated after deposition can bias isotope measurements and offset temperature values [@Schrag1999].
 This resulted in most $\delta^18^O samples being excluded from the dataset (x out of x).
@@
 
-Our results exemplify how incorporating quantified ecological temperature constraints can provide more precise temperature reconstructions than geochemical proxies alone.
-Combining climate-sensitive plant communities [@ @ @], reptiles [@ ] and leaf shapes [@] with geochemical proxies offers substantial potential for improving quantitative palaeoclimate reconstructions across the Phanerozoic.
+Our results exemplify how incorporating quantified ecological temperature constraints can provide more precise temperature reconstructions than geochemical proxies alone, adding to the advances in palaeoclimate reconstructions achieved by integrating lithological data [@Scotese2021].
+Combining the occurrences of climate-sensitive plant communities [@Greenwood1995], reptiles [@Marwick2007], leaf shapes [@Peppe2011], with geochemical proxies offers substantial potential for improving quantitative palaeoclimate reconstructions across the Phanerozoic.
 Our modelling framework offers a straightforward way of integrating ecological climate data with other proxy data: The hierarchical model structure accounts for variation of temperature estimates from proxies at individual localities, which is treated equivalent to the uncertainty associated with the ecological temperature proxies.
@@
 
-This dilemma could be resolved in future studies by applying machine learning approaches to inferring quantitative temperatures from ecological and lithological proxies (e.g.
-@Salonen2019; @Wei2020; @Chandra2021) in a novel way: If geochemical proxies and ecological or lithological data are derived from the same location and time interval, the climatic requirements of these taxa or sediments can be learned by using palaeotemperatures from that time interval, instead of or in addition to modern modern temperatures.
+This dilemma could be resolved in future studies by applying machine learning approaches to inferring quantitative temperatures from ecological and lithological proxies: If geochemical proxies and ecological or lithological data are derived from the same location and time interval, the climatic requirements of these taxa or sediments can be learned by using palaeotemperatures from that time interval, instead of or in addition to modern modern temperatures.
 This approach would ultimately rely on geochemical proxies providing absolute temperature estimates, but the resulting information could be exported to other locations lacking good coverage of geochemical proxies via their fossil or sedimentary record.
+
+<!-- This could be a neat future manuscript (although time...) so not sure if we state this idea, or just cut the previous paragraph -->
 
@@
 
-[add table on lat gradients]
+[Table 1: Early Eocene latitudinal temperature gradients from the literature]
 
@@
 However, the presence of high proxy-derived temperature estimates at ~ 60 deg latitudes forces the modeled median temperature curve to be too high at ~ 24 degC, relative to the temperatures indicated by the high-latitude mangrove communities (15.6 - 22.5 degC).
-In contrast, the extrapolated polar temperatures of most previous proxy-based models are likely too low, given the abundance of ecological data indicating temperate or subtropical high-latitude climates during the EECO [@ @]
+In contrast, the extrapolated polar temperatures of most previous proxy-based models are likely too low, given the abundance of ecological data indicating temperate or subtropical high-latitude climates during the EECO [@Pross2012; @Popescu2021]
 
-The very high variability of the proxy record in mid-latitudes results in large uncertainties on the shape of temperature gradient and on the GMSsT.
+The very high variability of the proxy record in mid-latitudes results in large uncertainties on the shape of temperature gradient and on the GMsST.
 Biases and errors in the proxy reconstructions likely contribute to the observed variability, as geochemical proxies reflect many other factors besides seawater temperature [@Hollis2019].
@@
 
-Recent GMSsT estimates of the EECO and of the early Eocene range from 23.4 to 37.1 degC, with the lowest GMSTs being derived from d18O, and the higher estimates including TEX86 [@Inglis2020].
-Many studies include both marine and terrestrial proxies to derive GMST estimates, but despite great differences in proxy selection and in the calculation of global average temperatures, most estimates fall in the range of 27 - 29.5 degC [@@@@], similar to our GMSsT estimate of `r eotemp_global$"50%"` degC.
+Recent GMsST estimates of the EECO and of the early Eocene range from 23.4 to 37.1 degC, with the lowest GMSTs being derived from d18O, and the higher estimates including TEX86 [@Inglis2020].
+Many studies include both marine and terrestrial proxies to derive GMST estimates, but despite great differences in proxy selection and in the calculation of global average temperatures, many recent estimates fall in the range of 27 - 29.5 degC [@Hansen2013; @Caballero2013; @Cramwinckel2018; @Zhu2019], similar to our GMsST estimate of `r eotemp_global$"50%"` degC.
 
@@
 
-- focus on gradient rather than gmst - gmst can easily be calculated from gradient.
+The Bayesian hierarchical model presented herein is able to reconstruct latitudinal gradients from both geochemical and ecological proxy data, whilst reflecting the uncertainty associated with the ecological temperature proxies, and accounts for the variation of multiple temperature estimates at individual localities.
+Using informative prior information allows for accurate temperature reconstructions from records with geographically incomplete sampling.
+By providing temperature estimates across the entire latitudinal range, this method also facilitates the reconstruction of unbiased global average temperatures.
+Application of our model to the EECO confirms the existence of latitudinal temperature gradients that were flatter than predicted by most Earth system models and by the majority of proxy-based work.
+High-latitude pollen records support this interpretation.
+Our GMsST estimate is in good agreement with most existing estimates, indicating that broadly accurate GMsST reconstructions are possible even with substantial deviations in the shape of the latitudinal temperature gradient.
+Our new method opens the door for improving the accuracy of proxy-based palaeoclimate reoncstructions and Phanerozoic temperature curves, particularly in intervals with a patchy record, and mitigates the biases incurred from uneven sampling.
 
-In conclusion, the Bayesian hierarchical model presented in this study represents a significant improvement over existing methods for reconstructing past temperatures by incorporating both geochemical and ecological data.
+<!-- for reconstructing past temperatures by incorporating both geochemical and ecological data.
 The use of a Bayesian framework allows for the specification of prior information based on physical constraints, which can improve interpolation and prediction.
@@
 This improved method is expected to significantly alter the proxy-based temperature reconstructions of the past, particularly for intervals with small or biased samples of temperature proxies.
+-->
+
 # Acknowledgements

### 2023-03-09 manuscript/main.qmd
clean up and update to use quarto

--- 2023-03-01 manuscript/manuscript.Rmd
@@
 title: "Bayesian multi-proxy reconstruction of latitudinal temperature gradients applied to the Early Eocene"
-
 author:
-  - Kilian Eichenseer^[Department of Earth Sciences, Durham University, South Road, DH1 3LE, Durham, United Kingdom] and Lewis A.
-Jones^[Centro de Investigación Mariña, Grupo de Ecoloxía Animal, Universidade de Vigo, 36310 Vigo, Spain]
-output:
-  word_document:
-    reference_docx: "template.docx"
-
+  - "---"
+  - Kilian Eichenseer^1^ and Lewis A. Jones^2^
+  - "---"
+  - "^1^Department of Earth Sciences, Durham University, South Road, DH1 3LE, Durham, United Kingdom"
+  - "^2^Centro de Investigación Mariña, Grupo de Ecoloxía Animal, Departamento de Ecoloxía e Bioloxía Animal, Universidade de Vigo, 36310 Vigo, Spain."
+  - "---"
+  - "**Corresponding author:** kilian.eichenseer@durham.ac.uk"
+abstract: \newpage Abstract
+format:
+  docx:
+    reference-doc: template.docx
+    highlight-style: github
+    fig-dpi: 900
+  pdf:
+    documentclass: scrartcl
+    papersize: letter
+    geometry:
+      - top=25.4mm
+      - bottom=25.4mm
+      - left=25.4mm
+      - right=25.4mm
+      - heightrounded
 bibliography: references.bib
-csl: palaeontology.csl
-
+csl: geology.csl
 ---
@@
 <!-- discarded titles:
+Bayesian multi-proxy reconstruction of latitudinal temperature gradients applied to the Early Eocene
 A Bayesian model of latitudinal temperature gradients applied to the Early Eocene
@@
 
-Abstract
-===============================================================================
 Accurately reconstructing large-scale palaeoclimate patterns from sparse local records is critical for understanding the evolution of Earth's climate.
@@
 We show that the inclusion of ecological data and adequate prior information can substantially reduce uncertainty in palaeoclimate reconstructions, and allow for unbiased temperature estimates from sparse data.
+
+# Keywords
 

### 2023-03-09 manuscript/supplementary.qmd
clean up and update to use quarto

--- 2023-02-12 manuscript/supplementary_material.Rmd
@@
 ---
-title: "Supplementary material for  \nBayesian multi-proxy reconstruction of the early Eocene latitudinal temperature gradient"
+title: "Supplementary Material: Bayesian multi-proxy reconstruction of latitudinal temperature gradients applied to the Early Eocene"
 author:
-  - Kilian Eichenseer^[Department of Earth Sciences, Durham University, South Road, DH1 3LE, Durham, United Kingdom] and Lewis A.
-Jones^[Centro de Investigación Mariña, Grupo de Ecoloxía Animal, Universidade de Vigo, 36310 Vigo, Spain]
-output:
-  word_document:
-    reference_docx: "template.docx"
+  - "---"
+  - Kilian Eichenseer^1^ and Lewis A. Jones^2^
+  - "---"
+  - "^1^Department of Earth Sciences, Durham University, South Road, DH1 3LE, Durham, United Kingdom"
+  - "^2^Centro de Investigación Mariña, Grupo de Ecoloxía Animal, Departamento de Ecoloxía e Bioloxía Animal, Universidade de Vigo, 36310 Vigo, Spain."
+  - "---"
+  - "**Corresponding author:** kilian.eichenseer@durham.ac.uk"
+format:
+  docx:
+    reference-doc: template.docx
+    highlight-style: github
+    fig-dpi: 900
+  pdf:
+    documentclass: scrartcl
+    papersize: letter
+    geometry:
+      - top=25.4mm
+      - bottom=25.4mm
+      - left=25.4mm
+      - right=25.4mm
+      - heightrounded
+bibliography: references.bib
+csl: geology.csl
+---
 
-bibliography: references.bib
-csl: palaeontology.csl
-
----
 ```{r, setup, include=FALSE}

### 2023-03-10 manuscript/main.qmd
Figure changes

--- 2023-03-01 manuscript/manuscript.Rmd
@@
 ---
-title: "Bayesian multi-proxy reconstruction of latitudinal temperature gradients applied to the Early Eocene"
-
+title: "Integrating geochemical and geoecological data to estimate latitudinal temperature gradients: an early Eocene example"
 author:
-  - Kilian Eichenseer^[Department of Earth Sciences, Durham University, South Road, DH1 3LE, Durham, United Kingdom] and Lewis A.
-Jones^[Centro de Investigación Mariña, Grupo de Ecoloxía Animal, Universidade de Vigo, 36310 Vigo, Spain]
-output:
-  word_document:
-    reference_docx: "template.docx"
-
+  - "---"
+  - Kilian Eichenseer^1^ and Lewis A. Jones^2^
+  - "---"
+  - "^1^Department of Earth Sciences, Durham University, South Road, DH1 3LE, Durham, United Kingdom"
+  - "^2^Centro de Investigación Mariña, Grupo de Ecoloxía Animal, Departamento de Ecoloxía e Bioloxía Animal, Universidade de Vigo, 36310 Vigo, Spain."
+  - "---"
+  - "**Corresponding author:** kilian.eichenseer@durham.ac.uk"
+abstract: \newpage Abstract
+format:
+  docx:
+    reference-doc: template.docx
+    highlight-style: github
+    fig-dpi: 900
+  pdf:
+    documentclass: scrartcl
+    papersize: letter
+    geometry:
+      - top=25.4mm
+      - bottom=25.4mm
+      - left=25.4mm
+      - right=25.4mm
+      - heightrounded
 bibliography: references.bib
-csl: palaeontology.csl
-
+csl: climate-of-the-past.csl
 ---
 
+```{=html}
 <!-- discarded titles:
+Bayesian multi-proxy reconstruction of latitudinal temperature gradients applied to the Early Eocene
 A Bayesian model of latitudinal temperature gradients applied to the Early Eocene
@@
 -->
-
-Abstract
-===============================================================================
+```
 Accurately reconstructing large-scale palaeoclimate patterns from sparse local records is critical for understanding the evolution of Earth's climate.
@@
 # Introduction
+
 The increasing availability of large compilations and databases of deep-time climate data now offers the opportunity for empirical palaeoclimate reconstructions of unprecedented scope [e.g.
@@
 Recrystallised $\delta^18^O$ samples were also excluded as secondary diagenetic calcite precipitated after deposition can bias isotope measurements and offset temperature values [@Schrag1999].
-This resulted in most $\delta^18^O samples being excluded from the dataset (x out of x).
+This resulted in most \$\delta^18^O samples being excluded from the dataset (x out of x).
 After data filtering, x geochemical proxy samples remained.
@@
 
-## Ecological data
+## Geoecological data
 
-**Coral reefs.** Today, shallow warm-water coral reefs are limited to tropical and subtropical latitudes (~34°N–32 degS), with minimum sea surface temperature (~18 degC) tolerances being the primary constraint on this distribution [@johannes1983; @kleypas1999; @yamano2001].
+**Coral reefs.** Today, shallow warm-water coral reefs are limited to tropical and subtropical latitudes (\~34°N--32 degS), with minimum sea surface temperature (\~18 degC) tolerances being the primary constraint on this distribution [@johannes1983; @kleypas1999; @yamano2001].
 As coral reefs reside at the upper thermal limit of the oceans today, their maximum sea surface temperature tolerance is less well-constrained with some studies suggesting up to 35.6 degC in the geological past [@jones2022reef].
@@
 We exclude coral knobs and coral-bearing mounds which might have broader climatic limits than coral reef ecosystems.
-This filtering resulted in four unique coral reef localities remaining for the EECO, all of which conform to the modern latitudinal range of coral reefs (<34 deg N).
+This filtering resulted in four unique coral reef localities remaining for the EECO, all of which conform to the modern latitudinal range of coral reefs (\<34 deg N).
 Subsequently, we use statistically derived temperature limits (minimum = 21 degC, average = 27.6 degC, maximum = 29.5 degC) from the published literature [@kleypas1999] to define a normal probability distribution of potential temperature values for coral reef localities.
 This normal probability distribution was defined with a mean of 27.6 and a standard deviation of 2.125, placing 97.5% of the probability density above the minimum.
-As the distribution of modern corals is skewed towards warmer temperatures, this approach results in 16.5% of the probability being placed on temperatures > 29.5 degC, allowing for the possibility that Eocene coral reefs were adapted to warmer conditions than modern-day coral reefs.
+As the distribution of modern corals is skewed towards warmer temperatures, this approach results in 16.5% of the probability being placed on temperatures \> 29.5 degC, allowing for the possibility that Eocene coral reefs were adapted to warmer conditions than modern-day coral reefs.
 
@@
 
-[Fig 1: Map and distribution of proxy data and ecological data]
+::: {#fig-1}
+![](../figures/fig_1.png){width="15cm"}
+
+Palaeogeographic distribution of the geochemical and geoecological data compilation used in this study.
+Palaeocoordinates were estimated via the palaeoverse R package ver.
+1.2.0 [@jones2023].
+To do so, the @merdith2021 Global Plate Model was used to palaeorotate present-day coordinates to 51.2 Ma.
+Map is presented in the Robinson projection (ESRI:54030).
+:::
 
@@
 The palaeogeographic distribution of geochemical and ecological data was reconstructed using the @merdith2021 plate rotation model via the palaeoverse R package ver.
-XXX (REF).
-The midpoint age of the EECO (51.2 Ma), along with the present-day coordinates of geochemical and ecological data, were used for palaeorotation.
+1.2.0 (@jones2023).
+The midpoint age of the EECO (51.2 Ma), along with the present-day coordinates of geochemical and geoecological data, were used for palaeorotation.
 
 ## Bayesian framework
+
 **Model structure (Fig 2).** We model the mean temperature ($\mu$) at location $j$ as a function of absolute latitude ($abs(l)$) with a logistic regression (`growth curve` or `Richard's curve`) of the form:
@@
 
-$$\nu_{j} = A~+\frac{K-A}{e^{B(abs(l_j)-M)}}~, ~~~~~ j = 1,...,n,  $$
-where $A$ and $K$ denote the lower and upper asymptote, respectively, $M$ specifies the latitude of maximal growth, the most quickly with latitude, $B$ denotes the growth rate (Fig.
+$$\nu_{j} = A~+\frac{K-A}{e^{B(abs(l_j)-M)}}~, ~~~~~ j = 1,...,n,  $$ where $A$ and $K$ denote the lower and upper asymptote, respectively, $M$ specifies the latitude of maximal growth, the most quickly with latitude, $B$ denotes the growth rate (Fig.
 2), $\sigma$ denotes the residual standard deviation, and $n$ denotes the number of locations.
@@
 
-$$t_{i,j}  \sim N(\mu_j, \sigma_j), ~~~~~ i = 1,...,m,  $$
-where $m$ is the number of observations at each location, and $\sigma_j$ is the estimated standard deviation of the temperatures at location $j$.
+$$t_{i,j}  \sim N(\mu_j, \sigma_j), ~~~~~ i = 1,...,m,  $$ where $m$ is the number of observations at each location, and $\sigma_j$ is the estimated standard deviation of the temperatures at location $j$.
 
@@
 **Priors (Fig2x).** In the Bayesian framework, priors need to be placed on the unknown parameters of a model.
-We placed weakly informative, conjugate inverse-gamma priors on $\sigma$ and $\sigma_{j=1,..n}$:
-$$\sigma  \sim \sqrt{Inv-Gamma(\alpha+\frac{n}{2},\beta + 0.5\times(\mu_j-\nu_j))}, ~~~~~ j = 1,...,n,  $$
-$$\sigma_j  \sim \sqrt{Inv-Gamma(\alpha+\frac{m}{2},\beta + 0.5\times(t_{i,j}-\mu_j))}, ~~~~~ i = 1,...,m,~~~~~j = 1,...,n. $$
-We set $\alpha = \beta = 1$, allowing these priors to be quickly overwhelmed by the data as $n$ and $m$ increase, as we have little *a priori* knowledge of these parameters.
+We placed weakly informative, conjugate inverse-gamma priors on $\sigma$ and $\sigma_{j=1,..n}$: $$\sigma  \sim \sqrt{Inv-Gamma(\alpha+\frac{n}{2},\beta + 0.5\times(\mu_j-\nu_j))}, ~~~~~ j = 1,...,n,  $$ $$\sigma_j  \sim \sqrt{Inv-Gamma(\alpha+\frac{m}{2},\beta + 0.5\times(t_{i,j}-\mu_j))}, ~~~~~ i = 1,...,m,~~~~~j = 1,...,n.
+$$ We set $\alpha = \beta = 1$, allowing these priors to be quickly overwhelmed by the data as $n$ and $m$ increase, as we have little *a priori* knowledge of these parameters.
 
@@
 The highest prior density of $A$ is placed around $0^\circ C$, and it slowly tapers off towards higher temperatures.
-This shape is achieved by placing a skew-normal prior on the lower asymptote, specified as
-$$A \sim SN(\xi=-3.0,\omega=12,\alpha_{SN}=30),$$
-where $\xi$, $\omega$, and $\alpha_{SN}$ are the location, scale and shape parameters.
+This shape is achieved by placing a skew-normal prior on the lower asymptote, specified as $$A \sim SN(\xi=-3.0,\omega=12,\alpha_{SN}=30),$$ where $\xi$, $\omega$, and $\alpha_{SN}$ are the location, scale and shape parameters.
 
@@
 This is achieved by setting $K \ge A$.
-The prior on the upper asymptote $K$ is a truncated normal distribution with the mean set to $K$ of the modern SST gradient, with a broad standard deviation:
-$$K \sim TN(\mu_{TN} = 28, \sigma_{TN} = 10, \alpha_{TN} = A, \beta_{TN}= \infty)$$
-The distribution is truncated to the left at $\alpha_{TN}=A$, but not truncated to the right ($\beta_{TN}$).
+The prior on the upper asymptote $K$ is a truncated normal distribution with the mean set to $K$ of the modern SST gradient, with a broad standard deviation: $$K \sim TN(\mu_{TN} = 28, \sigma_{TN} = 10, \alpha_{TN} = A, \beta_{TN}= \infty)$$ The distribution is truncated to the left at $\alpha_{TN}=A$, but not truncated to the right ($\beta_{TN}$).
 
-**M.** The steepness of the gradient is presumed to be highest in mid-latitudes; this is expressed with a normal prior on $M$ with the mean set to $M$ of the modern SST gradient:
-$$M \sim N(42,10)$$
-**B.** The steepness or growth rate $B$ of the gradient is constrained to be $\ge0$ and to not be exceedingly high, as oceanic and atmospheric heat transfer is bound to prevent very abrupt SST changes across latitudes.
-A gamma-distributed prior of the form
-$$B \sim Gamma(\alpha_G=4.3,\beta_G=30)$$
-was placed on $B$. The shape and rate parameters $\alpha_G$ and $\beta_G$ were chosen such that the highest prior density is at $B$ of the modern SST gradient, $0.11$.
+**M.** The steepness of the gradient is presumed to be highest in mid-latitudes; this is expressed with a normal prior on $M$ with the mean set to $M$ of the modern SST gradient: $$M \sim N(42,10)$$ **B.** The steepness or growth rate $B$ of the gradient is constrained to be $\ge0$ and to not be exceedingly high, as oceanic and atmospheric heat transfer is bound to prevent very abrupt SST changes across latitudes.
+A gamma-distributed prior of the form $$B \sim Gamma(\alpha_G=4.3,\beta_G=30)$$ was placed on $B$.
+The shape and rate parameters $\alpha_G$ and $\beta_G$ were chosen such that the highest prior density is at $B$ of the modern SST gradient, $0.11$.
 
-[Fig. 2: Model structure and visualisation of parameters and priors]
+\[Fig. 2: Model structure and visualisation of parameters and priors\]
 
 ## Model validation
-To test whether our logistic regression model can adequately describe latitudinal temperature gradients, we applied a simplified version of the model to modern sea surface temperatures from Bio-Oracle [@...].
+
+To test whether our logistic regression model can adequately describe latitudinal temperature gradients, we applied a simplified version of the model to modern sea surface temperatures from Bio-Oracle \[\@...\].
 As these modern, annual sea surface temperatures estimates are associated with little uncertainty, there is no need for a hierarchical model structure, and we substituted temperature ($t$) for $\mu$ in eq.
@@
 ## Parameter estimation
+
 We estimated the posterior distributions of the model parameters using a Markov chain Monte Carlo (MCMC) algorithm, written in R.
@@
 For the Eocene model, we ran four chains with 600,000 iterations each, discarding 100,000 as burn-in and keeping every 100th iteration, as the hierarchical model structure results in higher autocorrelation of the chains.
-The Eocene posterior inference is thus based on a total of 20,000 iterations with low autocorrelation (effective multivariate sample size for $A$, $K$, $M$ and $B$ is > 18,000).
+The Eocene posterior inference is thus based on a total of 20,000 iterations with low autocorrelation (effective multivariate sample size for $A$, $K$, $M$ and $B$ is \> 18,000).
 Trace plots of the MCMC chains indicate convergence and good mixing of the chains (Supplementary Figure S1).
@@
 ## Processing of model results
+
+```{=html}
 <!--
@@
 -->
+```
+
 Modelled sea surface temperature estimates were generated with eq.
@@
 
-Differences between Eocene and modern temperatures at a certain latitude were calculated by randomly pairing all iterations of the posterior from the Eocene and modern temperature gradient model, calculating the Eocene and modern temperature using the respective iterations, taking the difference, and then calculating the median ($2.5^{th}$, $97.5^{th}$ percentile) from all pairs of iterations.
+Differences between Eocene and modern temperatures at a certain latitude were calculated by randomly pairing all iterations of the posterior from the Eocene and modern temperature gradient model, calculating the Eocene and modern temperature using the respective iterations, taking the difference, and then calculating the median ($2.5^{th}$, $97.5^{th}$ percentile) from all pairs of iterations.\
 Global average temperatures were calculated by taking the weighted mean of the median ($2.5^{th}$, $97.5^{th}$ percentile) temperature estimates in 1 degree latitudinal bins.
 The weights were set to the proportion of global surface area in each latitudinal bin, i.e.
-decreasing with increasing latitude as:
-$$weights = sin(\alpha_{1,i}) - sin(\alpha_{2,i}) ,$$
-where $\alpha1$ is the upper, and $\alpha2$ is the lower latitudinal boundary of bin $i$.
+decreasing with increasing latitude as: $$weights = sin(\alpha_{1,i}) - sin(\alpha_{2,i}) ,$$ where $\alpha1$ is the upper, and $\alpha2$ is the lower latitudinal boundary of bin $i$.
 
 # Results
+
 ```{r results block - for in-line code, echo = FALSE}
@@
 ```
+
 ## Model validation
+
+```{=html}
 <!-- - Show that the 4-parameter logistic function can adequatly describe the modern temperature gradient
@@
 -->
-[Fig 3 - Modern gradient]
+```
+\[Fig 3 - Modern gradient\]
+
+::: {#fig-3}
+![](../figures/fig_3.png){width="10cm"}
+
+Estimates of the present-day latitudinal temperature gradient using the Bayesian hierarchical model.
+(A) The present-day empirical (black line) latitudinal temperature gradient (mean sea surface temperature) and that estimated by the Bayesian hierarchical model (green line) when using all cell values from the Bio-ORACLE grid of mean sea surface temperature.
+Grey points depict the individual cell values of the Bio-ORACLE grid.
+(B) The present-day empirical (black line) latitudinal temperature gradient and that estimated by the Bayesian hierarchical model (purple line) when limiting cell values by the palaeogeographic distribution of Eocene samples.
+To do so, present-day temperature values were extracted based on the palaeolatitudes of Eocene samples with randomly generated longitudes.
+This process was repeated 100 times, randomly choosing longitude for each sample in each repetition to capture the potential longitudinal variability in sea surface temperature.
+Purple points depict the median sea surface temperature value for each sample and the purple bars illustrate the 95% credible intervals.
+:::
+
 The average modern temperature gradient can be closely approximated with our model when using the full modern SST dataset (Fig.
@@
 
+::: {#fig-4}
+![](../figures/fig_4.png){width="10cm"}
+
+Estimates of the present-day and early Eocene climatic optimum latitudinal temperature gradient using the Bayesian hierarchical model.
+The estimated present-day (green line) and early Eocene climatic optimum (purple line) latitudinal temperature gradient.
+The respective ribbons (green and purple shading) depict the 95% credible intervals of the estimated gradients.
+The yellow points depict the palaeolatitudinal distribution of geochemical data (e.g.
+$\delta$^18^O) and their respective point estimate temperature values.
+The green squares depict the palaeolatitudinal distribution of geoecological data (i.e.
+coral reefs and mangroves assemblages) and their mean temperature values sampled from their respective normal distributions.
+:::
+
 -   x1 Draw samples from the posterior gradient to get a latitudinal gradient of deg C / deg lat with uncertainty (intercept gradient at 0 and 90 deg lat)
@@
 # Discussion
+
 ## Improved reconstruction of temperature gradients and global average temperatures
+
 The Bayesian model presented herein accurately reconstructs the modern, latitudinal SST gradient from a patchy and relatively small (n = 34) sample of temperature data.
@@
 
-Palaeoclimate reconstructions are often summarised as global mean surface temperatures (GMST), providing a standardised metric for characterising the state of the Earth's climate [@Royer2004;@Inglis2020].
+Palaeoclimate reconstructions are often summarised as global mean surface temperatures (GMST), providing a standardised metric for characterising the state of the Earth's climate [@Royer2004; @Inglis2020].
 However, the calculation of global mean surface temperatures from sparse proxy data is susceptible to bias [@Jones2022].
@@
 We anticipate that applying this improved will significantly alter Phanerozoic, proxy-based temperature curves, particularly when applied to intervals with small or biased samples of temperature proxies.
-[@Royer2004;@Veizer2015;@Jones2022].
+[@Royer2004; @Veizer2015; @Jones2022].
 
@@
 This is especially true for taxa that inhabit the warmest part of the ocean today, e.g.
-coral reefs [@].
-Although coral reefs are threatened by warming sea surface temperatures today [@], it is conceivable that Eocene reef corals were adapted to a warmer climate.
+coral reefs \[\@\].
+Although coral reefs are threatened by warming sea surface temperatures today \[\@\], it is conceivable that Eocene reef corals were adapted to a warmer climate.
 The fossil record indicates that reef development may have been stunted in the early Eocene, but the few early Eocene coral reefs occur in low latitudes [@Zamagni2012].
@@
 ## Early Eocene climate
+
 The geochemical proxy record and ecological data indicate that the latitudinal SST gradient of the early Eocene climatic optimum was significantly shallower than the modern [@Huber2011], but beyond that, there is little agreement.
-@Evans2018 estimate the SST gradient at 32+-10% of the modern, which would correspond to ~`r round( as.numeric(modern_gradient$"50%") - 0.32*as.numeric(modern_gradient$"50%"),0)`+- `r format(round( 0.10*as.numeric(modern_gradient$"50%"),0),nsmall=0)`degC, using the difference between the mean tropical and deep ocean data.
-The modeled SST of @Tierney2017 using TEX86 is ~ `r round(35.8-23.8,0)` degC, when taking the difference between temperatures at the polar circles and the equator.
-@Inglis2020 include both terrestrial air temperature and SST estimates in their model of the latitudinal temperature gradient in their Figure S4a; they model a polar circle to equatorial gradient of ~ `r round(30.7-(20.2+14.3)/2,0)`degC.
+@Evans2018 estimate the SST gradient at 32+-10% of the modern, which would correspond to \~`r round( as.numeric(modern_gradient$"50%") - 0.32*as.numeric(modern_gradient$"50%"),0)`+- `r format(round( 0.10*as.numeric(modern_gradient$"50%"),0),nsmall=0)`degC, using the difference between the mean tropical and deep ocean data.
+The modeled SST of @Tierney2017 using TEX86 is \~ `r round(35.8-23.8,0)` degC, when taking the difference between temperatures at the polar circles and the equator.
+@Inglis2020 include both terrestrial air temperature and SST estimates in their model of the latitudinal temperature gradient in their Figure S4a; they model a polar circle to equatorial gradient of \~ `r round(30.7-(20.2+14.3)/2,0)`degC.
 All of those estimates are significantly lower than the SST gradient predicted by an EECO climate model ensemble [@Lunt2012, as shown in @Tierney2017], which predicts a polar circle to equatorial gradient of `r round(31.5-5.5,0)`degC.
-@Keating-Bitonti2011 show a proxy-based SST gradient of ~ `r round(30-(19+15.6)/2,0)`degC, and @Bijl2009 model a gradient of ~`r round(36-27,0)`degC in the northern and only ~`r 36-31`degC in the southern hemisphere for the early Eocene.
+@Keating-Bitonti2011 show a proxy-based SST gradient of \~ `r round(30-(19+15.6)/2,0)`degC, and @Bijl2009 model a gradient of \~`r round(36-27,0)`degC in the northern and only \~`r 36-31`degC in the southern hemisphere for the early Eocene.
 Our polar circle to equatorial gradient is lower than most previous estimates at `r eocene_gradient_polar_circle$"50%"`degC, although the 95% credible interval overlap with several of the listed gradient estimates.
 
-[Table 1: Early Eocene latitudinal temperature gradients from the literature]
+\[Table 1: Early Eocene latitudinal temperature gradients from the literature\]
 
@@
 Polar temperature estimates from our model are thus conservative in that they admit large uncertainty where data is absent, which is desirable.
-However, the presence of high proxy-derived temperature estimates at ~ 60 deg latitudes forces the modeled median temperature curve to be too high at ~ 24 degC, relative to the temperatures indicated by the high-latitude mangrove communities (15.6 - 22.5 degC).
+However, the presence of high proxy-derived temperature estimates at \~ 60 deg latitudes forces the modeled median temperature curve to be too high at \~ 24 degC, relative to the temperatures indicated by the high-latitude mangrove communities (15.6 - 22.5 degC).
 In contrast, the extrapolated polar temperatures of most previous proxy-based models are likely too low, given the abundance of ecological data indicating temperate or subtropical high-latitude climates during the EECO [@Pross2012; @Popescu2021]

### 2023-03-10 R/figures/fig_1.R

@@ -1,25 +1,19 @@
-# Load libraries --------------------------------------------------------------
-library(raster)
-library(rgdal)
-library(broom)
+# Load libraries ---------------------------------------------------------
+library(sf)
 library(ggplot2)
-# Read data -------------------------------------------------------------------
+library(ggthemes)
+# Read data --------------------------------------------------------------
 chem <- readRDS("./data/processed/Hollis_processed_EECO_2022_07_19.rds")
 bio <- readRDS("./data/processed/bio_proxies_2022_08_08.RDS")
 # Merdith2021 shapefiles
-coastline <-
-  "http://gws.gplates.org/reconstruct/coastlines/?time=51&model=MERDITH2021"
-coastline <- rgdal::readOGR(coastline)
-coast_poly <- broom::tidy(coastline)
-#Data prep -------------------------------------------------------------------
+coastline <- "https://gws.gplates.org/reconstruct/coastlines/?time=51&model=MERDITH2021"
+coastline <- sf::read_sf(coastline)
+#Data prep ---------------------------------------------------------------
 # Filter data
 chem <- chem[, c("p_lng", "p_lat", "proxy")]
 chem$shape <- 1
 bio <- bio[, c("p_lng", "p_lat", "type")]
 bio$shape <- 2
-# Rename data
-bio[which(bio$type == "reef"), c("type")] <- "Coral reef"
-bio[which(bio$type == "mangrove"), c("type")] <- "Mangrove"
 # Rename cols 
 names(chem) <- c("p_lng", "p_lat", "proxy", "shape")
 names(bio) <- c("p_lng", "p_lat", "proxy", "shape")
@@ -28,77 +22,48 @@ bio <- unique(bio)
 chem <- unique(chem)
 # Bind data
 locs <- rbind.data.frame(chem, bio)
-locs$unique <- paste(locs$p_lng, locs$p_lat)
-filt <- names(which(table(locs$uni) > 1))
-locs[which(locs$uni %in% filt), c("proxy")] <- c("Various")
-locs[which(locs$uni %in% filt), c("shape")] <- 3
 
 # Set factors
-locs[, c("proxy")] <- factor(locs[, c("proxy")],
-                                  levels = c("Coral reef", "Mangrove", "d18O",
-                                             "d47", "TEX86","MgCa", "Various"))
-# Shift point for visualisation
-locs[which.max(locs$p_lat), c("p_lat")] <- 86
+levels <- c("reef", "mangrove", "d18O", "d47", "TEX86", "MgCa")
[…]

### 2023-03-10 R/figures/fig_3.R

@@ -1,21 +1,20 @@
-# Load libraries --------------------------------------------------------------
-library(raster)
+# Load libraries ---------------------------------------------------------
+library(terra)
 library(ggplot2)
 library(ggpubr)
 source("./R/functions/model_processing/temp_from_gradient.R")
 source("./R/functions/model_processing/combine_posterior.R")
-# Load data -------------------------------------------------------------------
-temp <- raster(
-  "./data/raw/climate/BioOracle_20220711/Present.Surface.Temperature.Mean.asc")
+# Load data --------------------------------------------------------------
+temp <- rast("./data/raw/climate/BioOracle_20220711/Present.Surface.Temperature.Mean.asc")
 modern_sample <- readRDS("./results/modern/modern_climate_model_output.rds")
 eocene_sample <- readRDS("./results/modern/modern_sample_eocene_p_lat_gradient.RDS")
 eocene_reps <- readRDS("./results/modern/modern_sample_summary_stats.RDS")
-# Data preparation ------------------------------------------------------------
+# Data preparation -------------------------------------------------------
 # Reduce resolution of raster
-r <- raster(res = 1)
+r <- rast(res = 1)
 temp <- resample(x = temp, y = r)
 # Extract temperature data
-temp <- raster::as.data.frame(x = temp, xy = TRUE, centroids = TRUE)
+temp <- terra::as.data.frame(x = temp, xy = TRUE)
 colnames(temp) <- c("lng", "lat", "SST")
 # Remove data from continents
 temp <- na.omit(temp)
@@ -24,10 +23,9 @@ temp$lat <- abs(temp$lat)
 eocene_reps$p_lat <- abs(eocene_reps$p_lat)
 # Add 1 deg latitudinal bins
 temp$bin <- ceiling(temp$lat) - 0.5
-# Calculate mean per lat bin
+# Calculate median per lat bin
 grad <- tapply(temp$SST, temp$bin, median)
-grad <- data.frame(lat = as.numeric(names(grad)), 
-                       SST = as.vector(grad))
+grad <- data.frame(lat = as.numeric(names(grad)), SST = as.vector(grad))
 # save for later use
 saveRDS(grad,"results/modern/empirical_median_1-deg-lat.rds")
 # Calculate modern temperatures from sample
@@ -40,98 +38,63 @@ eocene_sample <- temp_from_gradient(lat = lat, model_out = eocene_sample)
 # Modern gradient
 p1 <- ggplot() + 
   geom_point(data = temp,
-             aes(x = lat,
[…]

### 2023-03-10 R/figures/fig_4.R

@@ -1,43 +1,24 @@
-# Load libraries --------------------------------------------------------------
+# Load libraries ---------------------------------------------------------
 library(ggplot2)
 library(ggpubr)
 source("./R/functions/model_processing/temp_from_gradient.R")
 source("./R/functions/model_processing/combine_posterior.R")
-# Load data -------------------------------------------------------------------
+# Load data --------------------------------------------------------------
 eeco_mod <- readRDS("./results/eeco/eeco_climate_model_output.rds")
 bio <- readRDS("./data/processed/distribution_matrix_for_EECO_model.rds")
 chem <- readRDS("./data/processed/observation_matrix_for_EECO_model.rds")
 modern_sample <- readRDS("./results/modern/modern_climate_model_output.rds")
 
-# Data preparation ------------------------------------------------------------
-# Filter data
-chem <- chem[, c("p_lat", "temperature", "sd", "proxy")]
+# Data preparation -------------------------------------------------------
+# Filter chem data
+chem <- chem[, c("p_lat", "temperature", "sd")]
 chem$type <- "Geochemical"
-chem$col <- "#fdbf6f"
-
-# add proxy type to ecol_data
-bio$proxy <- rep("Coral reef", 11)
-bio$proxy[1:2] <- "Avicennia"
-bio$proxy[3:7] <- "Avicennia and Rhizophoraceae"
-bio <- bio[, c("p_lat", "mu", "scale", "proxy")]
-names(bio) <- c("p_lat", "temperature", "scale", "proxy")
-bio$type <- "Ecological"
-bio$col <- "#33a02c"
-# Unique data
-bio <- unique(bio)
-chem <- unique(chem)
+# Filter eco data
+bio <- bio[, c("p_lat", "mu", "scale")]
+names(bio) <- c("p_lat", "temperature", "sd")
+bio$type <- "Geoecological"
 # Bind data
 locs <- rbind.data.frame(chem, bio)
-
-# Set factors
-locs[, c("proxy")] <- factor(locs[, c("proxy")],
-                             levels = c("Coral reef",
-                                        "Avicennia",
-                                        "Avicennia and Rhizophoraceae",
-                                        "d18O",
-                                        "d47",
-                                        "TEX86",
-                                        "MgCa"))
 
 # eeco model
 eeco_mod <- combine_posterior(mod = eeco_mod, burnin = 100000)
[…]

### 2023-03-10 figures/fig_2.R

@@ -1,0 +1,28 @@
+source("R/options.R")
+source("R/functions/model_components/dsnorm.R")
+source("R/functions/model_components/dtnorm.R")
+
+# Priors
+n <- 1000
+xval = list(A = seq(-6, 36, length.out = n),
+            K = seq(-5, 65, length.out = n),
+            M = seq(9, 80, length.out = n),
+            B = seq(0, 0.4, length.out = n)
+)
+A <- priors$f1(xval[["A"]], log = FALSE)
+A <- A / max(A)
+K <- priors$f2(xval[["K"]], xval[["K"]][which.max(A)], log = FALSE)
+K <- K / max(K)
+M <- priors$f3(xval[["M"]], log = FALSE)
+M <- M / max(M)
+B <- priors$f4(xval[["B"]], log = FALSE)
+B <- B / max(B)
+
+# Set up dataframe
+df <- data.frame(x = c(xval[["A"]], xval[["K"]], xval[["M"]], xval[["B"]]),
+                 y = c(A, K, M, B),
+                 prior = rep(c("A", "K", "M", "B"), each = n))
+
+ggplot(data = df, aes(y = y)) +
+  geom_density(fill="royalblue3") +
+  facet_wrap(~prior, scales = "free")

### 2023-03-20 manuscript/main.qmd
Work in progress...

@@
 ---
-title: "Integrating geochemical and geoecological data to estimate latitudinal temperature gradients: an early Eocene example"
+title: "Bayesian multi-proxy reconstruction of latitudinal temperature gradients: An early Eocene example"
 author:
@@
 
+Palaeoclimate, latitudinal temperature gradients, temperature proxies, Eocene, spatial bias
+
 <!--Geochemical data such as oxygen isotopes (d18O), tetraether indices (TEX86) and magnesium calcium ratios (Mg/Ca) are routinely used to reconstruct ocean temperatures in deep time.
@@
 
-The increasing availability of large compilations and databases of deep-time climate data now offers the opportunity for empirical palaeoclimate reconstructions of unprecedented scope [e.g.
+Reconstructing Earth's past climate is fundamental to understanding how ecosystems respond to climatic and environmental change.
+To reconstruct palaeoclimatic conditions, researchers use a range of approaches including geochemical proxies, geological evidence, and Earth System Models [see @Burgener2023].
+Geochemical proxies such as oxygen isotopes (𝛅^18^O) are used to provide quantitative palaeotemperature estimates by applying transfer functions derived from experimental work and field observations [e.g.
+@Liu2009; @Wuchter2004].
+Geological evidence such as lithological and fossil data are used to qualitatively infer past climatic conditions based on analogous modern sediments [@Chandra2021], or the premise that the climatic requirements of ancient taxa, biological traits, or ecological communities were similar to those of their nearest modern relatives [@Peppe2011; @Royer2012; @Salonen2019].
+Finally, Earth System Models (ESMs)--such as the HadCM3 climate model [@valdes2017]--are used to model the main processes of the Earth system to provide spatially and temporally explicit quantitative estimates of palaeoclimatic conditions.
+However, ESMs ultimately rely on constraints provided by the geological and fossil record [@Lunt2021].
+
+The recent increasing availability of large-scale compilations of deep-time climate proxy data provide an unprecedented opportunity to reconstruct palaeoclimatic conditions [e.g.
 @Veizer2015; @Hollis2019; @Scotese2021; @Grossman2022; @Judd2022].
-To exploit the full potential these data sets offer, two fundamental challenges need to be overcome: Disparate types of proxy data need to be integrated to inform a single climate reconstruction, and large-scale scale climate patterns need to be inferred from localised, often sparse observations.
+However, two fundamental challenges must still be addressed to exploit the full potential of these data sets.
+First, disparate types of proxy data need to be integrated in order to generate unified palaeoclimatic reconstructions, making use of all available data.
+Second, the impact of incomplete preservation and uneven sampling of proxy records needs to be addressed [@Jones2022].
+
 To address these challenges, we present a novel Bayesian model that combines quantitative proxies and palaeoecological constraints into a fully quantitative model of the latitudinal gradient of sea surface temperatures, bridging spatial gaps in sparsely sampled climate data.
 
-Existing palaeoclimate reconstructions are generally based on one of three approaches [see @Burgener2023]: 1) Geochemical proxies such as oxygen isotopes (d18O), tetraether indices (TEX86) and magnesium calcium ratios (Mg/Ca) provide quantitative palaeotemperature estimates using transfer functions based on experimental work and field observations [e.g.
-@Liu2009; @Wuchter2004].
-2) Palaeobiological, ecological and lithological proxy data provide climate constraints, but no experimentally determined transfer functions exist to directly infer quantitative climate parameters from these records [e.g.
+Routinely, one of three approaches are used to reconstruct past conditions one of three approaches
+
+constrain palaeoclimatic conditions, but no experimentally determined transfer functions exist to directly infer quantitative climate parameters from these records [e.g.
 @Wolfe1995; @Kvavcek2007; @Cao2019].
-Instead, palaeoclimate signals are inferred based on the climatic conditions of analogous, modern sediments [@Chandra2021], or based on the premise that the climatic requirements of ancient taxa, plant traits or ecological communities were similar to those of their nearest modern relatives [@Peppe2011; @Royer2012; @Salonen2019].
-3) Earth system models offer the most highly resolved and complete climate reconstructions, but ultimately rely on constraints provided by the geological and fossil record [@Lunt2021].
+Instead, palaeoclimate signals (3)
 
@@
 ```
-
 Modelled sea surface temperature estimates were generated with eq.
@@
 The yellow points depict the palaeolatitudinal distribution of geochemical data (e.g.
-$\delta$^18^O) and their respective point estimate temperature values.
+\$\\delta\$^18^O) and their respective point estimate temperature values.
 The green squares depict the palaeolatitudinal distribution of geoecological data (i.e.

### 2023-03-20 manuscript/supplementary.qmd
Work in progress...

@@
 bibliography: references.bib
-csl: geology.csl
+csl: climate-of-the-past.csl
 ---

### 2023-03-22 manuscript/main.qmd
Add figure 2

@@
 
+::: {#fig-2}
+![](../figures/fig_2.png){width="15cm"}
+
+Model reconstructions of simulated latitudinal temperature gradients at various sample sizes.
+Each column depicts a different reconstruction for a given sample size (5, 10, 20, and 50).
+Each row depicts a different simulated latitudinal temperature gradient that represents idealised climatic states: extreme greenhouse (flat), greenhouse (quadratic), icehouse (linear), extreme icehouse (steep).
+The black line illustrates the simulated gradient.
+The blue line depicts the reconstructed gradient represented by the median sea surface temperature value estimated from 1000 model runs.
+The blue shading depicts the 90%, 95%, and 99% credible intervals.
+Each gradient is depicted in absolute latitude.
+
+:::
+
 ```{=html}

### 2023-03-22 R/figures/fig_2.R

@@ -1,0 +1,32 @@
+# Load libraries --------------------------------------------------------
+library(ggplot2)
+library(ggthemes)
+# Read data -------------------------------------------------------------
+df <- readRDS("./results/simulation/simulated_gradients_temperature_estimates.rds")
+# Categorise gradients
+df$gradient_type[which(df$gradient_type == "flat")] <- "Extreme Greenhouse"
+df$gradient_type[which(df$gradient_type == "quadratic")] <- "Greenhouse"
+df$gradient_type[which(df$gradient_type == "linear")] <- "Icehouse"
+df$gradient_type[which(df$gradient_type == "steep")] <- "Extreme Icehouse"
+# Set factor levels
+df$gradient_type <- factor(x = df$gradient_type, levels = c("Extreme Icehouse",
+                                                            "Icehouse",
+                                                            "Greenhouse",
+                                                            "Extreme Greenhouse"))
+# Plot data -------------------------------------------------------------
+col <- "#02818a"
+ggplot(data = df, aes(x = lat, y = median)) +
+  geom_ribbon(aes(x = lat, ymin = p_100, ymax = p_900),
+              fill = col, alpha = 0.25) +
+  geom_ribbon(aes(x = lat, ymin = p_005, ymax = p_995),
+              fill = col, alpha = 0.25) +
+  geom_ribbon(aes(x = lat, ymin = p_025, ymax = p_975),
+              fill = col, alpha = 0.25) +
+  geom_line(aes(x = lat, y = known), colour = "black", size = 0.75, alpha = 1) +
+  geom_line(colour = col, size = 0.75, alpha = 1) +
+  facet_grid(gradient_type~sample_size) +
+  labs(y = "Sea surface temperature (\u00B0C)", x = "Abs. latitude (\u00B0)") +
+  theme_bw()
+# Save plot -------------------------------------------------------------
+ggsave("./figures/fig_2.png",
+       units = "mm", width = 200, height = 200, dpi = 600)

### 2023-03-25 manuscript/main.qmd
Work in progress...

@@
 
-Reconstructing Earth's past climate is fundamental to understanding how ecosystems respond to climatic and environmental change.
+Reconstructing past climate is fundamental to understanding the long-term evolution of the Earth's climate system, including natural climate variability [@zachos2008], the impacts of external forcings such as impact events [@chiarenza2020], and changes in greenhouse gas concentrations [@foster2017].
+Moreover, palaeoclimatic reconstructions are vital for contextualising current global warming, enabling evaluation of the magnitude and significance in a broader historical perspective [@royer2004; @Burke2018], as well as understanding ecosystem response to environmental change [@jones2022reef; @woodhouse2023].
+
+Recent advances in deep-time climate reconstructions [e.g.
+@Burgener2023; @chevalier2022], including improvements in proxy data quantification [e.g.
+@bao2023] and data synthesis [e.g.
+@Veizer2015; @Hollis2019; @Scotese2021; @Grossman2022; @Judd2022], are enabling researchers to better understand the complex and dynamic nature of the Earth's climate system over geologic time.
+However, a fundamental challenge remains which must be addressed to exploit the full potential of these advances: the proxy record of past climate is spatially biased, impacted by incomplete preservation and uneven sampling [@judd2020; @Jones2022; @Judd2022].
+Recent work has demonstrated that this bias can lead to spurious estimates of latitudinal temperature gradients and global temperatures [@judd2020; @Jones2022].
+
+Here, we present a novel Bayesian hierarchical model that combines quantitative proxies and ecological constraints into a fully quantitative model of the latitudinal gradient of sea surface temperatures, bridging spatial gaps in sparsely sampled climate data.
+This model expands upon existing, spatially explicit reconstructions of palaeoclimate by allowing for the integration of 1) prior information based on physical principles and on the observed, modern sea surface temperature distribution, and of 2) geochemical and ecological climate proxies in a common, quantitative framework.
+We use a generalised logistic function to infer the shape of the temperature gradient despite a patchy latitudinal coverage, and test the robustness of this method using a down-sampled record of modern SSTs.
+We apply this model to the record of the early Eocene climatic optimum (EECO), combining a compilation of geochemical proxies [@Hollis2019], mangrove communities [@Popescu2021], and coral reefs [@Zamagni2012], using a nearest-living-relative approach [see e.g.
+@Greenwood2017] to establish broad temperature ranges for the ecological data.
+
+----
+
+The early Eocene is the focus of a substantial body of palaeoclimate literature [e.g.
+@Berggren1998; @Thomas2000; @Hyland2017; @Lunt2021], as it is the interval with the warmest sustained temperatures of the Cenozoic [@Pross2012], and potentially represents an analogue for extreme climate warming scenarios [@Burke2018].
+For palaeoclimate modelers, the early Eocene presents a challenge: Geochemical proxy data [@Hollis2019] and ecological data [@Markwick1994; @Greenwood1995] indicate tropical to subtropical conditions up to high latitudes, but early climate models have not been able to model warm polar temperatures and a shallow latitudinal temperature gradient under realistic conditions [@Sloan1990].
+Although more recent climate models have been able to better align their simulations with the proxy record [@Huber2011], the nature of the early Eocene temperature gradient remains a matter of debate [@Tierney2017].
+The shape and magnitude of the inferred EECO latitudinal gradient depends on the selection of proxies, on the method for dealing with spatial gaps and uneven sampling, on the inclusion or exclusion of terrestrial, shallow and deep marine data, while differential reporting of the results complicates comparisons of different approaches [@Evans2018; @Hollis2019; @Inglis2020].
+Our goal is to provide a robust, yet simple method for inferring and comparing temperature gradients that can be readily used with different proxy types and with patchy records.
+We use this model to quantify the latitudinal SST gradient and the global mean SST of the early Eocene, but the unspecific design of our method means it can be readily applied to other time intervals and proxy records.
+
+how ecosystems respond to climatic and environmental change.
 To reconstruct palaeoclimatic conditions, researchers use a range of approaches including geochemical proxies, geological evidence, and Earth System Models [see @Burgener2023].
@@
 Second, the impact of incomplete preservation and uneven sampling of proxy records needs to be addressed [@Jones2022].
-
-To address these challenges, we present a novel Bayesian model that combines quantitative proxies and palaeoecological constraints into a fully quantitative model of the latitudinal gradient of sea surface temperatures, bridging spatial gaps in sparsely sampled climate data.
-
-Routinely, one of three approaches are used to reconstruct past conditions one of three approaches
 
@@
 Statistical modelling in the Bayesian framework helps overcome this problem by requiring the explicit specification of priors for the model parameters, which can be used to express physical constraints [@Chandra2021].
-
-Here, we present a Bayesian, hierarchical model for inferring the latitudinal gradient of sea surface temperatures (SSTs) that expands upon existing, spatially explicit reconstructions of palaeoclimate by allowing for the integration of 1) prior information based on physical principles and on the observed, modern SST distribution, and of 2) geochemical and ecological climate proxies in a common, quantitative framework.
-We use a generalised logistic function to infer the shape of the temperature gradient despite a patchy latitudinal coverage, and test the robustness of this method using a down-sampled record of modern SSTs.
-We apply this model to the record of the early Eocene climatic optimum (EECO), combining a compilation of geochemical proxies [@Hollis2019], mangrove communities [@Popescu2021], and coral reefs [@Zamagni2012], using a nearest-living-relative approach [see e.g.
-@Greenwood2017] to establish broad temperature ranges for the ecological data.
-
-The early Eocene is the focus of a substantial body of palaeoclimate literature [e.g.
-@Berggren1998; @Thomas2000; @Hyland2017; @Lunt2021], as it is the interval with the warmest sustained temperatures of the Cenozoic [@Pross2012], and potentially represents an analogue for extreme climate warming scenarios [@Burke2018].
-For palaeoclimate modelers, the early Eocene presents a challenge: Geochemical proxy data [@Hollis2019] and ecological data [@Markwick1994; @Greenwood1995] indicate tropical to subtropical conditions up to high latitudes, but early climate models have not been able to model warm polar temperatures and a shallow latitudinal temperature gradient under realistic conditions [@Sloan1990].
-Although more recent climate models have been able to better align their simulations with the proxy record [@Huber2011], the nature of the early Eocene temperature gradient remains a matter of debate [@Tierney2017].
-The shape and magnitude of the inferred EECO latitudinal gradient depends on the selection of proxies, on the method for dealing with spatial gaps and uneven sampling, on the inclusion or exclusion of terrestrial, shallow and deep marine data, while differential reporting of the results complicates comparisons of different approaches [@Evans2018; @Hollis2019; @Inglis2020].
-Our goal is to provide a robust, yet simple method for inferring and comparing temperature gradients that can be readily used with different proxy types and with patchy records.
-We use this model to quantify the latitudinal SST gradient and the global mean SST of the early Eocene, but the unspecific design of our method means it can be readily applied to other time intervals and proxy records.
 
@@
 
-To test whether our logistic regression model can adequately describe latitudinal temperature gradients, we applied a simplified version of the model to modern sea surface temperatures from Bio-Oracle \[\@...\].
+To test whether our logistic regression model can adequately describe different latitudinal temperature gradients at various sample sizes, we generated four idealised gradients that emulate potential climatic states throughout Earth's geological history: extreme icehouse, icehouse, greenhouse and extreme greenhouse (REF).
+We then randomly sampled (1000 iterations) these gradients using a decreasing sample size (50, 20, 10, and 5) and reconstructed the latitudinal temperature gradient using our model for each of these sample sizes and gradient types (**Should we be using some kind of statistic here to say that the curves are similar, perhaps the Kolmogorov-Smirnoff (K-S) Statistic?**).
+Subsequently, we test whether our model could accurately reconstruct latitudinal temperature gradients using the palaeogeographic distribution of Eocene samples, providing an empirical example distribution that captures both limited sample size and skewed geographic distributions.
+
+To verify that the modern gradient can be approximated with limited sampling, we resampled modern sea surface temperatures at modern latitudes corresponding the palaeolatitudes of the Eocene samples.
+This process was repeated 100 times, randomly chosing a longitude for each latitude in each repetition.
+The parameters of the non-hierarchical temperature model were estimated for each of the 100 samples, and the iterations after burn-in from the posterior of the parameters were pooled to generate the resulting median temperature gradient with 95 % credible intervals ($2.5^{th}$ to $97.5^{th}$ percentile).
+
+we applied a simplified version of the model to modern sea surface temperatures from Bio-Oracle \[\@...\].
 As these modern, annual sea surface temperatures estimates are associated with little uncertainty, there is no need for a hierarchical model structure, and we substituted temperature ($t$) for $\mu$ in eq.
@@
 Each gradient is depicted in absolute latitude.
-
 :::

### 2023-03-25 R/figures/fig_2.R

@@ -4,10 +4,10 @@ library(ggthemes)
 # Read data -------------------------------------------------------------
 df <- readRDS("./results/simulation/simulated_gradients_temperature_estimates.rds")
 # Categorise gradients
-df$gradient_type[which(df$gradient_type == "flat")] <- "Extreme Greenhouse"
-df$gradient_type[which(df$gradient_type == "quadratic")] <- "Greenhouse"
-df$gradient_type[which(df$gradient_type == "linear")] <- "Icehouse"
-df$gradient_type[which(df$gradient_type == "steep")] <- "Extreme Icehouse"
+df$gradient_type[which(df$gradient_type == "extreme_greenhouse")] <- "Extreme Greenhouse"
+df$gradient_type[which(df$gradient_type == "greenhouse")] <- "Greenhouse"
+df$gradient_type[which(df$gradient_type == "icehouse")] <- "Icehouse"
+df$gradient_type[which(df$gradient_type == "extreme_icehouse")] <- "Extreme Icehouse"
 # Set factor levels
 df$gradient_type <- factor(x = df$gradient_type, levels = c("Extreme Icehouse",
                                                             "Icehouse",
@@ -17,15 +17,16 @@ df$gradient_type <- factor(x = df$gradient_type, levels = c("Extreme Icehouse",
 col <- "#02818a"
 ggplot(data = df, aes(x = lat, y = median)) +
   geom_ribbon(aes(x = lat, ymin = p_100, ymax = p_900),
-              fill = col, alpha = 0.25) +
+              fill = col, alpha = 0.2) +
   geom_ribbon(aes(x = lat, ymin = p_005, ymax = p_995),
-              fill = col, alpha = 0.25) +
+              fill = col, alpha = 0.2) +
   geom_ribbon(aes(x = lat, ymin = p_025, ymax = p_975),
-              fill = col, alpha = 0.25) +
-  geom_line(aes(x = lat, y = known), colour = "black", size = 0.75, alpha = 1) +
-  geom_line(colour = col, size = 0.75, alpha = 1) +
+              fill = col, alpha = 0.2) +
+  geom_line(aes(x = lat, y = known), linetype = 1, colour = "black", size = 0.75, alpha = 1) +
+  geom_line(colour = col, linetype = 1, size = 0.75, alpha = 0.9) +
   facet_grid(gradient_type~sample_size) +
[…]

### 2023-04-12 manuscript/main.qmd
abstract update

@@
 Our results confirm the presence of a flattened latitudinal temperature gradient and unusually high polar temperatures during the EECO, which is supported by high-latitude ecological data.
-We show that the inclusion of ecological data and adequate prior information can substantially reduce uncertainty in palaeoclimate reconstructions, and allow for unbiased temperature estimates from sparse data.
+Integrating ecological data and adequate prior information has the potential to substantially reduce uncertainty in palaeoclimate reconstructions, allowing for unbiased temperature estimates from sparse data.
 

### 2023-04-17 manuscript/discards.qmd
update intro/figure captions

--- 2023-02-24 manuscript/discards.rmd
@@
 is an .
+
+The early Eocene is the focus of a substantial body of palaeoclimate literature [e.g.
+@Berggren1998; @Thomas2000; @Hyland2017; @Lunt2021], as it is the interval with the warmest sustained temperatures of the Cenozoic [@Pross2012], and potentially represents an analogue for extreme climate warming scenarios [@Burke2018].
+For palaeoclimate modelers, the early Eocene presents a challenge: Geochemical proxy data [@Hollis2019] and ecological data [@Markwick1994; @Greenwood1995] indicate tropical to subtropical conditions up to high latitudes, but early climate models have not been able to model warm polar temperatures and a shallow latitudinal temperature gradient under realistic conditions [@Sloan1990].
+Although more recent climate models have been able to better align their simulations with the proxy record [@Huber2011], the nature of the early Eocene temperature gradient remains a matter of debate [@Tierney2017].
+The shape and magnitude of the inferred EECO latitudinal gradient depends on the selection of proxies, on the method for dealing with spatial gaps and uneven sampling, on the inclusion or exclusion of terrestrial, shallow and deep marine data, while differential reporting of the results complicates comparisons of different approaches [@Evans2018; @Hollis2019; @Inglis2020].
+Our goal is to provide a robust, yet simple method for inferring and comparing temperature gradients that can be readily used with different proxy types and with patchy records.
+We use this model to quantify the latitudinal SST gradient and the global mean SST of the early Eocene, but the unspecific design of our method means it can be readily applied to other time intervals and proxy records.
+
+constrain palaeoclimatic conditions, but no experimentally determined transfer functions exist to directly infer quantitative climate parameters from these records [e.g.
+@Wolfe1995; @Kvavcek2007; @Cao2019].
+Instead, palaeoclimate signals (3)
 

### 2023-04-17 manuscript/main.qmd
update intro/figure captions

@@
 
+To reconstruct palaeoclimatic conditions, researchers use a range of approaches including geochemical proxies, geological evidence, and Earth System Models [see @Burgener2023].
+Geochemical proxies such as oxygen isotopes ($\delta^18^O$) are used to provide quantitative palaeotemperature estimates by applying transfer functions derived from experimental work and field observations [e.g.
+@Liu2009; @Wuchter2004].
+Geological evidence such as lithological and fossil data are used to qualitatively infer past climatic conditions based on analogous modern sediments [@Chandra2021], or the premise that the climatic requirements of ancient taxa, biological traits, or ecological communities were similar to those of their nearest modern relatives [@Peppe2011; @Royer2012; @Salonen2019].
+Finally, Earth System Models (ESMs)--such as the HadCM3 climate model [@valdes2017]--are used to model the main processes of the Earth system to provide spatially and temporally explicit quantitative estimates of palaeoclimatic conditions.
+Nevertheless, ESMs ultimately rely on constraints provided by the geological and fossil record [@Lunt2021].
+
 Recent advances in deep-time climate reconstructions [e.g.
@@
 
-Here, we present a novel Bayesian hierarchical model that combines quantitative proxies and ecological constraints into a fully quantitative model of the latitudinal gradient of sea surface temperatures, bridging spatial gaps in sparsely sampled climate data.
-This model expands upon existing, spatially explicit reconstructions of palaeoclimate by allowing for the integration of 1) prior information based on physical principles and on the observed, modern sea surface temperature distribution, and of 2) geochemical and ecological climate proxies in a common, quantitative framework.
-We use a generalised logistic function to infer the shape of the temperature gradient despite a patchy latitudinal coverage, and test the robustness of this method using a down-sampled record of modern SSTs.
-We apply this model to the record of the early Eocene climatic optimum (EECO), combining a compilation of geochemical proxies [@Hollis2019], mangrove communities [@Popescu2021], and coral reefs [@Zamagni2012], using a nearest-living-relative approach [see e.g.
-@Greenwood2017] to establish broad temperature ranges for the ecological data.
-
-----
-
-The early Eocene is the focus of a substantial body of palaeoclimate literature [e.g.
-@Berggren1998; @Thomas2000; @Hyland2017; @Lunt2021], as it is the interval with the warmest sustained temperatures of the Cenozoic [@Pross2012], and potentially represents an analogue for extreme climate warming scenarios [@Burke2018].
-For palaeoclimate modelers, the early Eocene presents a challenge: Geochemical proxy data [@Hollis2019] and ecological data [@Markwick1994; @Greenwood1995] indicate tropical to subtropical conditions up to high latitudes, but early climate models have not been able to model warm polar temperatures and a shallow latitudinal temperature gradient under realistic conditions [@Sloan1990].
-Although more recent climate models have been able to better align their simulations with the proxy record [@Huber2011], the nature of the early Eocene temperature gradient remains a matter of debate [@Tierney2017].
-The shape and magnitude of the inferred EECO latitudinal gradient depends on the selection of proxies, on the method for dealing with spatial gaps and uneven sampling, on the inclusion or exclusion of terrestrial, shallow and deep marine data, while differential reporting of the results complicates comparisons of different approaches [@Evans2018; @Hollis2019; @Inglis2020].
-Our goal is to provide a robust, yet simple method for inferring and comparing temperature gradients that can be readily used with different proxy types and with patchy records.
-We use this model to quantify the latitudinal SST gradient and the global mean SST of the early Eocene, but the unspecific design of our method means it can be readily applied to other time intervals and proxy records.
-
-how ecosystems respond to climatic and environmental change.
-To reconstruct palaeoclimatic conditions, researchers use a range of approaches including geochemical proxies, geological evidence, and Earth System Models [see @Burgener2023].
-Geochemical proxies such as oxygen isotopes (𝛅^18^O) are used to provide quantitative palaeotemperature estimates by applying transfer functions derived from experimental work and field observations [e.g.
-@Liu2009; @Wuchter2004].
-Geological evidence such as lithological and fossil data are used to qualitatively infer past climatic conditions based on analogous modern sediments [@Chandra2021], or the premise that the climatic requirements of ancient taxa, biological traits, or ecological communities were similar to those of their nearest modern relatives [@Peppe2011; @Royer2012; @Salonen2019].
-Finally, Earth System Models (ESMs)--such as the HadCM3 climate model [@valdes2017]--are used to model the main processes of the Earth system to provide spatially and temporally explicit quantitative estimates of palaeoclimatic conditions.
-However, ESMs ultimately rely on constraints provided by the geological and fossil record [@Lunt2021].
-
-The recent increasing availability of large-scale compilations of deep-time climate proxy data provide an unprecedented opportunity to reconstruct palaeoclimatic conditions [e.g.
-@Veizer2015; @Hollis2019; @Scotese2021; @Grossman2022; @Judd2022].
-However, two fundamental challenges must still be addressed to exploit the full potential of these data sets.
-First, disparate types of proxy data need to be integrated in order to generate unified palaeoclimatic reconstructions, making use of all available data.
-Second, the impact of incomplete preservation and uneven sampling of proxy records needs to be addressed [@Jones2022].
-
-constrain palaeoclimatic conditions, but no experimentally determined transfer functions exist to directly infer quantitative climate parameters from these records [e.g.
-@Wolfe1995; @Kvavcek2007; @Cao2019].
-Instead, palaeoclimate signals (3)
-
-Even when using the full range of available records, proxy-based climate reconstructions beyond the Holocene are inevitably patchy [e.g.
-@Hollis2019], and global temperature reconstructions based on those records alone are prone to bias [@Jones2022].
 Inferring regional or global-scale patterns from an incomplete record relies on the bridging of spatial gaps.
@@
 Statistical modelling in the Bayesian framework helps overcome this problem by requiring the explicit specification of priors for the model parameters, which can be used to express physical constraints [@Chandra2021].
+
+Here, we present a novel Bayesian hierarchical model that combines quantitative proxies and ecological constraints into a fully quantitative model of the latitudinal gradient of sea surface temperatures, bridging spatial gaps in sparsely sampled climate data.
+This model expands upon existing, spatially explicit reconstructions of palaeoclimate by allowing for the integration of 1) prior information based on physical principles and on the observed, modern sea surface temperature distribution, and of 2) geochemical and ecological climate proxies in a common, quantitative framework.
+We use a generalised logistic function to infer the shape of the temperature gradient despite a patchy latitudinal coverage, and test the robustness of this method using a down-sampled record of modern SSTs.
+We apply this model to the record of the early Eocene climatic optimum (EECO), combining a compilation of geochemical proxies [@Hollis2019], mangrove communities [@Popescu2021], and coral reefs [@Zamagni2012], using a nearest-living-relative approach [see e.g.
+@Greenwood2017] to establish broad temperature ranges for the ecological data.
 
@@
 Model reconstructions of simulated latitudinal temperature gradients at various sample sizes.
-Each column depicts a different reconstruction for a given sample size (5, 10, 20, and 50).
-Each row depicts a different simulated latitudinal temperature gradient that represents idealised climatic states: extreme greenhouse (flat), greenhouse (quadratic), icehouse (linear), extreme icehouse (steep).
+Each column depicts a different reconstruction for given sample sizes: 5, 10, 20, and 34 (sample size for the EECO).
+Each row depicts a different simulated latitudinal temperature gradient that represents idealised climatic states: extreme greenhouse, greenhouse, icehouse, extreme icehouse.
 The black line illustrates the simulated gradient.
@@
 
-Estimates of the present-day latitudinal temperature gradient using the Bayesian hierarchical model.
-(A) The present-day empirical (black line) latitudinal temperature gradient (mean sea surface temperature) and that estimated by the Bayesian hierarchical model (green line) when using all cell values from the Bio-ORACLE grid of mean sea surface temperature.
+Estimate of the present-day latitudinal temperature gradient using the Bayesian hierarchical model.
+The present-day empirical (black line) latitudinal temperature gradient (mean sea surface temperature) and that estimated by the Bayesian hierarchical model (green line) when using all cell values from the Bio-ORACLE grid of mean sea surface temperature.
 Grey points depict the individual cell values of the Bio-ORACLE grid.
-(B) The present-day empirical (black line) latitudinal temperature gradient and that estimated by the Bayesian hierarchical model (purple line) when limiting cell values by the palaeogeographic distribution of Eocene samples.
-To do so, present-day temperature values were extracted based on the palaeolatitudes of Eocene samples with randomly generated longitudes.
-This process was repeated 100 times, randomly choosing longitude for each sample in each repetition to capture the potential longitudinal variability in sea surface temperature.
-Purple points depict the median sea surface temperature value for each sample and the purple bars illustrate the 95% credible intervals.
 :::
@@
 The respective ribbons (green and purple shading) depict the 95% credible intervals of the estimated gradients.
-The yellow points depict the palaeolatitudinal distribution of geochemical data (e.g.
-\$\\delta\$^18^O) and their respective point estimate temperature values.
-The green squares depict the palaeolatitudinal distribution of geoecological data (i.e.
-coral reefs and mangroves assemblages) and their mean temperature values sampled from their respective normal distributions.
+Points within the plot depict the palaeolatitudinal distribution of geochemical (e.g.
+$\delta^18^O$) and ecological (e.g.
+mangroves) data.
+Geochemical data are plotted by their point estimate temperature value.
+Ecological data are plotted by their mean temperature values sampled from their respective normal distributions.
 :::
@@
 
+The authors are grateful to all those who have enabled this work by collecting, measuring, collating and screening geochemical and fossil data.
+The contribution of K.E.
+was supported by...
+The contribution of L.A.J.
+was supported by a Juan de la Cierva-formación 2021 fellowship (FJC2021-046695-I/MCIN/AEI/10.13039/501100011033) from the European Union "NextGenerationEU"/PRTR.
+
 # Author contributions
+
+Both authors designed the study and carried out data preparation.
+K.E.
+programmed the model and conducted the analyses.
+L.A.J.
+generated the figures.
+Both authors contributed to the writing of the manuscript.
+
+# Competing Interests
+
+The authors declare that they have no conflicts of interest.
 
@@
 
+The data and code used to produce the results of this study are available via GitHub (<https://github.com/KEichenseer/PalaeoClimateGradient>) and the linked Zenodo repository (XXX).
+
 # References

### 2023-04-19 manuscript/main.qmd
upload r squared results and methods changes

@@
 To test whether our logistic regression model can adequately describe different latitudinal temperature gradients at various sample sizes, we generated four idealised gradients that emulate potential climatic states throughout Earth's geological history: extreme icehouse, icehouse, greenhouse and extreme greenhouse (REF).
-We then randomly sampled (1000 iterations) these gradients using a decreasing sample size (50, 20, 10, and 5) and reconstructed the latitudinal temperature gradient using our model for each of these sample sizes and gradient types (**Should we be using some kind of statistic here to say that the curves are similar, perhaps the Kolmogorov-Smirnoff (K-S) Statistic?**).
-Subsequently, we test whether our model could accurately reconstruct latitudinal temperature gradients using the palaeogeographic distribution of Eocene samples, providing an empirical example distribution that captures both limited sample size and skewed geographic distributions.
-
-To verify that the modern gradient can be approximated with limited sampling, we resampled modern sea surface temperatures at modern latitudes corresponding the palaeolatitudes of the Eocene samples.
-This process was repeated 100 times, randomly chosing a longitude for each latitude in each repetition.
-The parameters of the non-hierarchical temperature model were estimated for each of the 100 samples, and the iterations after burn-in from the posterior of the parameters were pooled to generate the resulting median temperature gradient with 95 % credible intervals ($2.5^{th}$ to $97.5^{th}$ percentile).
+We then randomly sampled (1000 iterations) these gradients using increasing sample sizes (5, 10, and 20) and reconstructed the latitudinal temperature gradient using our model for each of these sample sizes and gradient types (**Should we be using some kind of statistic here to say that the curves are similar, perhaps the Kolmogorov-Smirnoff (K-S) Statistic?**).
+Using the same idealised gradients, we also tested whether our model could accurately reconstruct latitudinal temperature gradients using the palaeogeographic distribution of Eocene samples (n = 34), providing an empirical example distribution that captures both limited sample size and skewed geographic distributions.
 
@@
 and ?.
-To verify that the modern gradient can be approximated with limited sampling, we resampled modern sea surface temperatures at modern latitudes corresponding the palaeolatitudes of the Eocene samples.
-This process was repeated 100 times, randomly chosing a longitude for each latitude in each repetition.
-The parameters of the non-hierarchical temperature model were estimated for each of the 100 samples, and the iterations after burn-in from the posterior of the parameters were pooled to generate the resulting median temperature gradient with 95 % credible intervals ($2.5^{th}$ to $97.5^{th}$ percentile).
 
@@
 Posterior inference on the modern gradient is based on four chains with 60,000 iterations each, 10,000 of which were discarded as burn-in, and keeping every 10th iteration, resulting in a total of 20,000 iterations.
-The modern temperature gradient with the Eocene sampling distribution was estimated in one chain with 25,000 iterations for each of the 100 temperature samples.
-5,000 iterations each were discarded as burn-in, and every 10th iteration was kept, resulting in a total of 200,000 iterations across all 100 model runs.
+The modern simulated gradients were estimated in one chain with 10,000 iterations for each of the 1,000 random samples.
+5,000 iterations each were discarded as burn-in, and every 25th iteration was kept, resulting in a total of 200,000 iterations across all 1,000 model runs.
+For the simulated gradients with an Eocene sampling distribution, a single chain with 250,000 iterations was used, thinned to 10,000 iterations after burn-in.
 For the Eocene model, we ran four chains with 600,000 iterations each, discarding 100,000 as burn-in and keeping every 100th iteration, as the hierarchical model structure results in higher autocorrelation of the chains.
@@
 The average modern temperature gradient can be closely approximated with our model when using the full modern SST dataset (Fig.
-3a); almost all of the variation in the empirical median temperatures in bins of 1 deg absolute latitude is explained by the modeled gradient ($R^2 = `r R2_modern`$).
+3a); almost all of the variation in the empirical median temperatures in bins of 1 deg absolute latitude is explained by the modelled gradient ($R^2 = `r R2_modern`$).
 Reducing the amount and the spatial coverage of data points to match the early Eocene sampling distribution ($n = 34$) still results in a close match of the modeled gradient with the empirical gradient (Fig.

### 2023-04-19 R/analyses/simulated_gradients_R_squared.R

@@ -1,0 +1,141 @@
+### goodness of fit of the modelled, simulated gradients
+
+library(dplyr)
+
+source("R/functions/model_components/gradient.R")
+source("R/functions/auxiliary_functions.R")
+
+# read simulated gradient models
+four_gradients_model <- readRDS("D://OneDrive - Durham University/projects/four_gradients_model_proc.rds")
+
+lats <- seq(0,90,1)
+grad <- gradient(lats, four_gradients_model[[1]][[1]])
+
+gradient_formulas <- rev(list( function(x) 30-x^2/1080, 
+                               function(x) 30-x^2/360, 
+                               function(x)  11.25*atan(-0.1*x+3.5)+15.25, 
+                               function(x) 8.75*atan(-0.5*x+15)+12))
+
+plot(lats,gradient_formulas[[1]](lats), type = "l")
+points(lats, grad[30000,], col = "orange")
+
+
+# basic R squared
+
+R_squared <- list(NULL)
+
+for(g in 1:4) {
+  R_squared[[g]] <- list(NULL)
+  for(s in 1:4) {
+    grad <- gradient(lats, four_gradients_model[[g]][[s]])
+    true_grad <- gradient_formulas[[g]](lats)
+    SSres <- apply(grad,1,function(x) sum((x-true_grad)^2))
+    SStot <- sum((true_grad-mean(true_grad))^2)
+    R_squared[[g]][[s]] <- 1-SSres/SStot
+  }
+}
+
+### Try weighted R squared (by latitudinal area)
+
+lats <- seq(0.5,89.5,1)
+alpha1 <- seq(1,90,1)
+alpha2 <- seq(0,89,1)
+latweight <- sin(pi*alpha1/180) - sin(pi*alpha2/180) # sums to 1 - careful when using lat. subsets
+
+R_squared_weighted <- list(NULL)
+
+for(g in 1:4) {
+  R_squared_weighted[[g]] <- list(NULL)
+  for(s in 1:4) {
+    grad <- gradient(lats, four_gradients_model[[g]][[s]])
+    true_grad <- gradient_formulas[[g]](lats)
+    SSres <- apply(grad,1,function(x) sum(latweight*(x-true_grad)^2))
+    SStot <- sum(latweight*(true_grad-mean(true_grad))^2)
+    R_squared_weighted[[g]][[s]] <- 1-SSres/SStot
+    print(c(g,s))
+  }
+}
+
[…]

### 2023-04-23 manuscript/main.qmd
work on intro and update fig. 2

@@
 
-Reconstructing past climate is fundamental to understanding the long-term evolution of the Earth's climate system, including natural climate variability [@zachos2008], the impacts of external forcings such as impact events [@chiarenza2020], and changes in greenhouse gas concentrations [@foster2017].
-Moreover, palaeoclimatic reconstructions are vital for contextualising current global warming, enabling evaluation of the magnitude and significance in a broader historical perspective [@royer2004; @Burke2018], as well as understanding ecosystem response to environmental change [@jones2022reef; @woodhouse2023].
+Reconstructing past climate is essential to understanding the long-term evolution of the Earth's climate system and contextualising current global warming [@royer2004; @Burke2018].
+Geochemical proxies, such as oxygen isotopes ($\delta^18^O$), are regularly used to provide quantitative palaeotemperature estimates for the geological past at a range of temporal and spatial scales [e.g.
+@Liu2009; @Wuchter2004].
+Recent advances in the synthesis of such data [e.g.
+@Veizer2015; @Hollis2019; @Scotese2021; @Grossman2022; @Judd2022] are enabling a better understanding of the complex and dynamic nature of the Earth's climate system.
+However, a fundamental challenge remains which must be addressed to exploit the full potential of these advances; the proxy record of past climate is spatially incomplete and biased by imperfect preservation and uneven sampling [@judd2020; @Jones2022; @Judd2022].
 
-To reconstruct palaeoclimatic conditions, researchers use a range of approaches including geochemical proxies, geological evidence, and Earth System Models [see @Burgener2023].
-Geochemical proxies such as oxygen isotopes ($\delta^18^O$) are used to provide quantitative palaeotemperature estimates by applying transfer functions derived from experimental work and field observations [e.g.
-@Liu2009; @Wuchter2004].
-Geological evidence such as lithological and fossil data are used to qualitatively infer past climatic conditions based on analogous modern sediments [@Chandra2021], or the premise that the climatic requirements of ancient taxa, biological traits, or ecological communities were similar to those of their nearest modern relatives [@Peppe2011; @Royer2012; @Salonen2019].
-Finally, Earth System Models (ESMs)--such as the HadCM3 climate model [@valdes2017]--are used to model the main processes of the Earth system to provide spatially and temporally explicit quantitative estimates of palaeoclimatic conditions.
-Nevertheless, ESMs ultimately rely on constraints provided by the geological and fossil record [@Lunt2021].
-
-Recent advances in deep-time climate reconstructions [e.g.
-@Burgener2023; @chevalier2022], including improvements in proxy data quantification [e.g.
-@bao2023] and data synthesis [e.g.
-@Veizer2015; @Hollis2019; @Scotese2021; @Grossman2022; @Judd2022], are enabling researchers to better understand the complex and dynamic nature of the Earth's climate system over geologic time.
-However, a fundamental challenge remains which must be addressed to exploit the full potential of these advances: the proxy record of past climate is spatially biased, impacted by incomplete preservation and uneven sampling [@judd2020; @Jones2022; @Judd2022].
-Recent work has demonstrated that this bias can lead to spurious estimates of latitudinal temperature gradients and global temperatures [@judd2020; @Jones2022].
-
-Inferring regional or global-scale patterns from an incomplete record relies on the bridging of spatial gaps.
-A common approach that requires no additional computation is the spatial visualisation of proxy-derived temperatures against latitude, showing broad latitudinal temperature trends [@Hollis2019; @Vickers2021].
-Interpolation is sometimes used to bridge spatial gaps in palaeoclimate data [@Taylor2004], taking advantage of the autoregressive nature of climatic data: much of the information on the climate of any given location is contained in the climate data of nearby locations [@Reynolds1994].
+Whilst geochemical proxy data can provide robust estimates of palaeotemperature at local scales (e.g.
+REF), recent work has demonstrated that spatial biases in the geochemical proxy record can lead to spurious estimates of regional (e.g.
+latitudinal temperature gradients) and global temperatures [@judd2020; @Jones2022].
+At the most fundamental level, this can be driven by two factors: (1) missing data for some regions (e.g.
+no high-latitude data); or (2) overrepresentation of some regions (e.g.
+a high proportion of samples from tropical regions).
+The latter can be addressed through the down-sampling of data or restricting analyses to specifics regions (e.g.
+REFs).
+However, in order to robustly infer regional or global-scale patterns from an incomplete record, spatial gaps must ultimately be bridged.
+One common approach, that requires no additional computation, is the spatial visualisation of proxy-derived temperatures against latitude, showing broad latitudinal temperature trends [@Hollis2019; @Vickers2021].
+Interpolation is also sometimes used to bridge spatial gaps in palaeoclimate data [@Taylor2004], taking advantage of the autoregressive nature of climatic data: much of the information on the climate of any given location is contained in the climate data of nearby locations [@Reynolds1994].
 Adding to this, some proxy-based reconstructions use statistical modelling to infer climatic patterns.
 For example, polynomial regression [@Bijl2009] and cosine functions [@Inglis2020] have been used to reconstruct latitudinal temperature gradients, and 2D-reconstructions of surface temperatures have been created with Gaussian process regression [@Inglis2020].
-These approaches work well for interpolating relatively well-sampled data, but the absence of constraints on the modeled parameters means that such models can produce unrealistic temperature estimates when extrapolating from sparsely sampled data.
-Statistical modelling in the Bayesian framework helps overcome this problem by requiring the explicit specification of priors for the model parameters, which can be used to express physical constraints [@Chandra2021].
+These approaches work well for interpolating relatively well-sampled data, but the absence of constraints on the modelled parameters means that such models can produce unrealistic temperature estimates when extrapolating from sparsely sampled data.
+Statistical modelling in a Bayesian framework can help overcome this problem by requiring the explicit specification of priors for the model parameters, which can be used to express physical constraints [@Chandra2021].
+
+Spatial gaps in the record of past climate can also be addressed through the integration of additional data.
+For example, lithological and fossil data can be used to infer past climatic conditions based on analogous modern sediments [@Chandra2021], or the premise that the climatic requirements of ancient taxa, biological traits, or ecological communities were similar to those of their nearest modern relatives [@Peppe2011; @Royer2012; @Salonen2019].
+Despite this potential, the integration of geochemical proxy data with other sources of information (e.g.
+ecological data) is rare due to the absence of a quantitative framework [@Burgener2023].
 
 Here, we present a novel Bayesian hierarchical model that combines quantitative proxies and ecological constraints into a fully quantitative model of the latitudinal gradient of sea surface temperatures, bridging spatial gaps in sparsely sampled climate data.
-This model expands upon existing, spatially explicit reconstructions of palaeoclimate by allowing for the integration of 1) prior information based on physical principles and on the observed, modern sea surface temperature distribution, and of 2) geochemical and ecological climate proxies in a common, quantitative framework.
+This model expands upon existing, spatially explicit reconstructions of palaeoclimate by allowing for the integration of (1) prior information based on physical principles and on the observed, modern sea surface temperature distribution, and of (2) geochemical and ecological climate proxies in a common, quantitative framework.
 We use a generalised logistic function to infer the shape of the temperature gradient despite a patchy latitudinal coverage, and test the robustness of this method using a down-sampled record of modern SSTs.
@@
 @Greenwood2017] to establish broad temperature ranges for the ecological data.
+Whilst we primarily do so to demonstrate an application of the model, the EECO represents an interval with the warmest sustained temperatures of the Cenozoic [@Pross2012], and a potential analogue for extreme climate warming scenarios [@Burke2018].
 
@@
 The blue shading depicts the 90%, 95%, and 99% credible intervals.
+Bold black text within each panel depicts the coefficient of determination (*R*^2^ value) for estimating goodness of fit between the simulated and modelled gradient.
+The median (50%) *R*^2^ value along with the 2.5% and 97.5% confidence intervals from the 1000 model runs are shown.
 Each gradient is depicted in absolute latitude.

### 2023-04-23 R/figures/fig_2.R

@@ -20,6 +20,29 @@ df$gradient_type <- factor(x = df$gradient_type, levels = c("Extreme Icehouse",
                                                             "Greenhouse",
                                                             "Extreme Greenhouse"))
 df$sample_size <- factor(x = df$sample_size, levels = c(5, 10, 20, "34 (Eocene samples)"))
+
+# Get R2 values
+r2 <- readRDS("./results/simulation/simulated_gradients_r_squared_gelman_unweighted.rds")
+# Rename gradients
+r2$gradient_type[which(r2$gradient_type == "extreme_greenhouse")] <- "Extreme Greenhouse"
+r2$gradient_type[which(r2$gradient_type == "greenhouse")] <- "Greenhouse"
+r2$gradient_type[which(r2$gradient_type == "icehouse")] <- "Icehouse"
+r2$gradient_type[which(r2$gradient_type == "extreme_icehouse")] <- "Extreme Icehouse"
+# Set factor levels
+r2$gradient_type <- factor(x = r2$gradient_type, levels = c("Extreme Icehouse",
+                                                            "Icehouse",
+                                                            "Greenhouse",
+                                                            "Extreme Greenhouse"))
+# Rename sample size
+r2$sample_size[which(r2$sample_size == "eocene")] <- "34 (Eocene samples)"
+r2$sample_size <- factor(x = r2$sample_size, levels = c(5, 10, 20, "34 (Eocene samples)"))
+# Round values
+r2$median <- round(r2$median, 3)
+r2$p_025 <- round(r2$p_025, 3)
+r2$p_975 <- round(r2$p_975, 3)
+r2$value <- paste0("2.5% = ", r2$p_025, "\n", 
+                   "50% = ", r2$median, "\n", 
+                   "97.5% = ", r2$p_975, "")
 # Plot data -------------------------------------------------------------
 col <- "#02818a"
 ggplot(data = df, aes(x = lat, y = median)) +
@@ -31,6 +54,7 @@ ggplot(data = df, aes(x = lat, y = median)) +
               fill = col, alpha = 0.2) +
   geom_line(aes(x = lat, y = known), linetype = 1, colour = "black", size = 0.75, alpha = 1) +
   geom_line(colour = col, linetype = 1, size = 0.75, alpha = 0.9) +
+  geom_text(data = r2, aes(x = 55, y = 38, label = value), hjust = 0, fontface = "bold", size = 2.5) +
   facet_grid(gradient_type~sample_size) +
[…]

### 2023-04-24 manuscript/main.qmd
Update intro

@@
 csl: climate-of-the-past.csl
+editor: source
 ---
@@
 Reconstructing past climate is essential to understanding the long-term evolution of the Earth's climate system and contextualising current global warming [@royer2004; @Burke2018].
-Geochemical proxies, such as oxygen isotopes ($\delta^18^O$), are regularly used to provide quantitative palaeotemperature estimates for the geological past at a range of temporal and spatial scales [e.g.
+Geochemical proxies, such as oxygen isotopes (&delta;^18^O), are regularly used to provide quantitative palaeotemperature estimates for the geological past at a range of temporal and spatial scales [e.g.
 @Liu2009; @Wuchter2004].
 Recent advances in the synthesis of such data [e.g.
-@Veizer2015; @Hollis2019; @Scotese2021; @Grossman2022; @Judd2022] are enabling a better understanding of the complex and dynamic nature of the Earth's climate system.
+@Veizer2015; @Hollis2019; @song2019; @Grossman2022; @Judd2022] are enabling a better understanding of the complex and dynamic nature of the Earth's climate system.
 However, a fundamental challenge remains which must be addressed to exploit the full potential of these advances; the proxy record of past climate is spatially incomplete and biased by imperfect preservation and uneven sampling [@judd2020; @Jones2022; @Judd2022].
 
-Whilst geochemical proxy data can provide robust estimates of palaeotemperature at local scales (e.g.
-REF), recent work has demonstrated that spatial biases in the geochemical proxy record can lead to spurious estimates of regional (e.g.
+Whilst geochemical proxy data can provide robust estimates of palaeotemperature at local scales, recent work has demonstrated that spatial biases in the geochemical proxy record can lead to spurious estimates of regional (e.g.
 latitudinal temperature gradients) and global temperatures [@judd2020; @Jones2022].
-At the most fundamental level, this can be driven by two factors: (1) missing data for some regions (e.g.
+At the most basic level, this can be driven by two factors: (1) missing data for some regions (e.g.
 no high-latitude data); or (2) overrepresentation of some regions (e.g.
 a high proportion of samples from tropical regions).
-The latter can be addressed through the down-sampling of data or restricting analyses to specifics regions (e.g.
-REFs).
+The latter can be addressed through the down-sampling of data or restricting analyses to specific regions (e.g.
+@song2019).
 However, in order to robustly infer regional or global-scale patterns from an incomplete record, spatial gaps must ultimately be bridged.
@@
 Interpolation is also sometimes used to bridge spatial gaps in palaeoclimate data [@Taylor2004], taking advantage of the autoregressive nature of climatic data: much of the information on the climate of any given location is contained in the climate data of nearby locations [@Reynolds1994].
-Adding to this, some proxy-based reconstructions use statistical modelling to infer climatic patterns.
+Adding to this, some proxy-based reconstructions use statistical modelling to infer palaeoclimatic patterns.
 For example, polynomial regression [@Bijl2009] and cosine functions [@Inglis2020] have been used to reconstruct latitudinal temperature gradients, and 2D-reconstructions of surface temperatures have been created with Gaussian process regression [@Inglis2020].
@@
 Geochemical climate proxy data were extracted from a latest Paleocene and early Eocene compilation [@Hollis2019].
-This compilation provides four different geochemical proxies for reconstructing seawater temperature: $\delta^{18}O$, $\Delta_{47}$, Mg/Ca and $TEX_{86}$.
+This compilation provides four different geochemical proxies for reconstructing seawater temperature: &delta;^18^O, &Delta;47, Mg/Ca and $TEX_{86}$.
 For our analyses, this dataset was restricted to the EECO and samples from the continental shelf.
-Recrystallised $\delta^18^O$ samples were also excluded as secondary diagenetic calcite precipitated after deposition can bias isotope measurements and offset temperature values [@Schrag1999].
-This resulted in most \$\delta^18^O samples being excluded from the dataset (x out of x).
+Recrystallised &delta;^18^O samples were also excluded as secondary diagenetic calcite precipitated after deposition can bias isotope measurements and offset temperature values [@Schrag1999].
+This resulted in most &delta;^18^O samples being excluded from the dataset (x out of x).
 After data filtering, x geochemical proxy samples remained.
@@
 Points within the plot depict the palaeolatitudinal distribution of geochemical (e.g.
-$\delta^18^O$) and ecological (e.g.
+&delta;^18^O) and ecological (e.g.
 mangroves) data.
@@
 Biases and errors in the proxy reconstructions likely contribute to the observed variability, as geochemical proxies reflect many other factors besides seawater temperature [@Hollis2019].
-Despite excluding d18O measurements from recrystallised fossils, systematic offsets remain between mostly warm temperatures derived from TEX86, and cooler temperatures derived from d18O, delta47, and the ecological proxies.
+Despite excluding &delta;^18^O measurements from recrystallised fossils, systematic offsets remain between mostly warm temperatures derived from TEX86, and cooler temperatures derived from &delta;^18^O, &Delta;~47~, and the ecological proxies.
 Seasonality [@Keating-Bitonti2011 and temporal changes within the EECO [@Westerhold2018] may also contribute to the large variability of the EECO proxy data.

### 2023-05-11 manuscript/main.qmd
methods updates

@@
 
+```{r methods block - for in-line code, echo = FALSE}
+setwd('..') # move up to parent folder of /manuscript
+
+# data
+obsmat <- readRDS("data/processed/observation_matrix_for_EECO_model.rds")
+n_geochem <- nrow(obsmat)
+```
 ## Modern sea surface temperatures
@@
 Geochemical climate proxy data were extracted from a latest Paleocene and early Eocene compilation [@Hollis2019].
-This compilation provides four different geochemical proxies for reconstructing seawater temperature: &delta;^18^O, &Delta;47, Mg/Ca and $TEX_{86}$.
+This compilation provides data on four different geochemical proxies for reconstructing seawater temperature: &delta;^18^O, &Delta;47, Mg/Ca and $TEX_{86}$.
 For our analyses, this dataset was restricted to the EECO and samples from the continental shelf.
 Recrystallised &delta;^18^O samples were also excluded as secondary diagenetic calcite precipitated after deposition can bias isotope measurements and offset temperature values [@Schrag1999].
-This resulted in most &delta;^18^O samples being excluded from the dataset (x out of x).
-After data filtering, x geochemical proxy samples remained.
+This filtering resulted in most &delta;^18^O samples being excluded from the dataset (retaining 8 out of 152).
+After data filtering, `r n_geochem` geochemical proxy samples remained.
 For a detailed description of each proxy see [@Hollis2019].
 
-## Geoecological data
+## Ecological data
 
@@
 As coral reefs reside at the upper thermal limit of the oceans today, their maximum sea surface temperature tolerance is less well-constrained with some studies suggesting up to 35.6 degC in the geological past [@jones2022reef].
-Nevertheless, the distribution of coral reefs have frequently been recognised as tracers of past (sub-)tropical conditions [@ziegler1984; @kiessling2001].
+Nevertheless, coral reefs have frequently been recognised as tracers of past (sub-)tropical conditions [@ziegler1984; @kiessling2001].
 During the Eocene, coral communities and reefs expanded across tropical and temperate latitudes, with some communities found up to palaeolatitudes of 43 degC N [@Zamagni2012].
 Using a compilation of Paleocene--early Eocene coral reefs and community localities [@Zamagni2012], we generate quantitative sea surface temperature estimates for the ECCO.
-To do so, we extract localities inferred to be Ilerdian (early Eocene) coral reefs from the compilation, and that can be confidently assigned to the EECO.
+To do so, we extract localities from the compilation that are inferred to be Ilerdian (early Eocene) coral reefs, and that can be confidently assigned to the EECO.
 We exclude coral knobs and coral-bearing mounds which might have broader climatic limits than coral reef ecosystems.
@@
 This normal probability distribution was defined with a mean of 27.6 and a standard deviation of 2.125, placing 97.5% of the probability density above the minimum.
-As the distribution of modern corals is skewed towards warmer temperatures, this approach results in 16.5% of the probability being placed on temperatures \> 29.5 degC, allowing for the possibility that Eocene coral reefs were adapted to warmer conditions than modern-day coral reefs.
+As the distribution of modern corals is skewed towards warmer temperatures, this approach results in 16.5% of the probability being placed on temperatures \> 29.5 degC, allowing for the possibility that Eocene coral reefs were adapted to warmer conditions than present-day coral reefs.
 
@@
 
-Palaeogeographic distribution of the geochemical and geoecological data compilation used in this study.
-Palaeocoordinates were estimated via the palaeoverse R package ver.
+Palaeogeographic distribution of the geochemical and ecological data compilation used in this study.
+Palaeocoordinates of samples were estimated via the palaeoverse R package ver.
 1.2.0 [@jones2023].
@@
 1.2.0 (@jones2023).
-The midpoint age of the EECO (51.2 Ma), along with the present-day coordinates of geochemical and geoecological data, were used for palaeorotation.
+The midpoint age of the EECO (51.2 Ma), along with the present-day coordinates of geochemical and ecological data, were used for palaeogeographic reconstruction.
 
@@
 
-**Model structure (Fig 2).** We model the mean temperature ($\mu$) at location $j$ as a function of absolute latitude ($abs(l)$) with a logistic regression (`growth curve` or `Richard's curve`) of the form:
+**Model structure (Fig 2).** We model the mean temperature ($\mu$) at location $j$ as a function of absolute latitude ($abs(l)$) with a logistic regression (also known as growth curve or Richard's curve) of the form:
 
-$$\mu_{j}  \sim N(\nu_j, \sigma), $$
+$$\mu_{j}  \sim N(\nu_j, \sigma), $${#eq-mu}
 
-$$\nu_{j} = A~+\frac{K-A}{e^{B(abs(l_j)-M)}}~, ~~~~~ j = 1,...,n,  $$ where $A$ and $K$ denote the lower and upper asymptote, respectively, $M$ specifies the latitude of maximal growth, the most quickly with latitude, $B$ denotes the growth rate (Fig.
+$$\nu_{j} = A~+\frac{K-A}{e^{B(abs(l_j)-M)}}~, ~~~~~ j = 1,...,n,  $${#eq-nu} where $A$ and $K$ denote the lower and upper asymptote, respectively, $M$ specifies the latitude of maximal growth, i.e.
+the latitude around which temperature falls most steeply with latitude, $B$ denotes the growth rate (Fig.
 2), $\sigma$ denotes the residual standard deviation, and $n$ denotes the number of locations.
@@
 
-$$t_{i,j}  \sim N(\mu_j, \sigma_j), ~~~~~ i = 1,...,m,  $$ where $m$ is the number of observations at each location, and $\sigma_j$ is the estimated standard deviation of the temperatures at location $j$.
+$$t_{i,j}  \sim N(\mu_j, \sigma_j), ~~~~~ i = 1,...,m,  $${#eq-obs} where $m$ is the number of observations at each location, and $\sigma_j$ is the estimated standard deviation of the temperatures at location $j$.
 
@@
 
-$$t_{\mu,j}  \sim N(\mu_j,t_{\sigma,j}).  $$
+$$t_{\mu,j}  \sim N(\mu_j,t_{\sigma,j}).  $${#eq-dist}
 
@@
 
-**Priors (Fig2x).** In the Bayesian framework, priors need to be placed on the unknown parameters of a model.
-We placed weakly informative, conjugate inverse-gamma priors on $\sigma$ and $\sigma_{j=1,..n}$: $$\sigma  \sim \sqrt{Inv-Gamma(\alpha+\frac{n}{2},\beta + 0.5\times(\mu_j-\nu_j))}, ~~~~~ j = 1,...,n,  $$ $$\sigma_j  \sim \sqrt{Inv-Gamma(\alpha+\frac{m}{2},\beta + 0.5\times(t_{i,j}-\mu_j))}, ~~~~~ i = 1,...,m,~~~~~j = 1,...,n.
-$$ We set $\alpha = \beta = 1$, allowing these priors to be quickly overwhelmed by the data as $n$ and $m$ increase, as we have little *a priori* knowledge of these parameters.
+**Priors.** In the Bayesian framework, priors need to be placed on the unknown parameters of a model.
+We placed weakly informative, conjugate inverse-gamma priors on $\sigma$ and $\sigma_{j=1,..n}$: $$\sigma  \sim \sqrt{Inv-Gamma(\alpha+\frac{n}{2},\beta + 0.5\times(\mu_j-\nu_j))}, ~~~~~ j = 1,...,n,  $${#eq-sigma}
+$$\sigma_j  \sim \sqrt{Inv-Gamma(\alpha+\frac{m}{2},\beta + 0.5\times(t_{i,j}-\mu_j))}, ~~~~~ i = 1,...,m,~~~~~j = 1,...,n. $${#eq-sigmaj}
+We set $\alpha = \beta = 1$, allowing these priors to be quickly overwhelmed by the data as $n$ and $m$ increase, as we have little *a priori* knowledge of these parameters.
 
@@
 The highest prior density of $A$ is placed around $0^\circ C$, and it slowly tapers off towards higher temperatures.
-This shape is achieved by placing a skew-normal prior on the lower asymptote, specified as $$A \sim SN(\xi=-3.0,\omega=12,\alpha_{SN}=30),$$ where $\xi$, $\omega$, and $\alpha_{SN}$ are the location, scale and shape parameters.
+This shape is achieved by placing a skew-normal prior on the lower asymptote, specified as $$A \sim SN(\xi=-3.0,\omega=12,\alpha_{SN}=30),$${#eq-A}
+where $\xi$, $\omega$, and $\alpha_{SN}$ are the location, scale and shape parameters.
 
@@
 This is achieved by setting $K \ge A$.
-The prior on the upper asymptote $K$ is a truncated normal distribution with the mean set to $K$ of the modern SST gradient, with a broad standard deviation: $$K \sim TN(\mu_{TN} = 28, \sigma_{TN} = 10, \alpha_{TN} = A, \beta_{TN}= \infty)$$ The distribution is truncated to the left at $\alpha_{TN}=A$, but not truncated to the right ($\beta_{TN}$).
+The prior on the upper asymptote $K$ is a truncated normal distribution with the mean set to $K$ of the modern SST gradient, with a broad standard deviation: $$K \sim TN(\mu_{TN} = 28, \sigma_{TN} = 10, \alpha_{TN} = A, \beta_{TN}= \infty)$${#eq-K}
+The distribution is truncated to the left at $\alpha_{TN}=A$, but not truncated to the right ($\beta_{TN}$).
 
-**M.** The steepness of the gradient is presumed to be highest in mid-latitudes; this is expressed with a normal prior on $M$ with the mean set to $M$ of the modern SST gradient: $$M \sim N(42,10)$$ **B.** The steepness or growth rate $B$ of the gradient is constrained to be $\ge0$ and to not be exceedingly high, as oceanic and atmospheric heat transfer is bound to prevent very abrupt SST changes across latitudes.
-A gamma-distributed prior of the form $$B \sim Gamma(\alpha_G=4.3,\beta_G=30)$$ was placed on $B$.
-The shape and rate parameters $\alpha_G$ and $\beta_G$ were chosen such that the highest prior density is at $B$ of the modern SST gradient, $0.11$.
+**M.** The steepness of the gradient is presumed to be highest in mid-latitudes; this is expressed with a normal prior on $M$ with the mean set to $M$ of the modern SST gradient: $$M \sim N(42,10)$${#eq-M}
 
-\[Fig. 2: Model structure and visualisation of parameters and priors\]
+**B.** The steepness or growth rate $B$ of the gradient is constrained to be $\ge0$ and to not be exceedingly high, as oceanic and atmospheric heat transfer is bound to prevent very abrupt SST changes across latitudes.
+A gamma-distributed prior of the form $$B \sim Gamma(\alpha_G=4.3,\beta_G=30)$${#eq-B}
+was placed on $B$. The shape and rate parameters $\alpha_G$ and $\beta_G$ were chosen such that the highest prior density is at $B$ of the modern SST gradient, $0.11$.
 
@@
 To test whether our logistic regression model can adequately describe different latitudinal temperature gradients at various sample sizes, we generated four idealised gradients that emulate potential climatic states throughout Earth's geological history: extreme icehouse, icehouse, greenhouse and extreme greenhouse (REF).
-We then randomly sampled (1000 iterations) these gradients using increasing sample sizes (5, 10, and 20) and reconstructed the latitudinal temperature gradient using our model for each of these sample sizes and gradient types (**Should we be using some kind of statistic here to say that the curves are similar, perhaps the Kolmogorov-Smirnoff (K-S) Statistic?**).
+We then randomly sampled (1000 iterations) these gradients using increasing sample sizes (5, 10, and 20) and reconstructed the latitudinal temperature gradient using our model for each of these sample sizes and gradient types.
 Using the same idealised gradients, we also tested whether our model could accurately reconstruct latitudinal temperature gradients using the palaeogeographic distribution of Eocene samples (n = 34), providing an empirical example distribution that captures both limited sample size and skewed geographic distributions.
 
-we applied a simplified version of the model to modern sea surface temperatures from Bio-Oracle \[\@...\].
-As these modern, annual sea surface temperatures estimates are associated with little uncertainty, there is no need for a hierarchical model structure, and we substituted temperature ($t$) for $\mu$ in eq.
-?
-and ?.
+We also used our model to estimate the modern sea surface temperature gradient, using annual sea surface mean temperatures from Bio-Oracle [@Assis2018], down-sampled to a $1^\circ \times 1 ^\circ$ raster (n = 46,131)
+
+To reconstruct the idealised gradients and the modern gradient, we used a simplified, non-hierarchical version of our model, as every location is associated with only one temperature value.
+We thus substituted temperature ($t_j$) for $\mu_j$ in [@eq-mu] and [@eq-sigma].
 
@@
 We estimated the posterior distributions of the model parameters using a Markov chain Monte Carlo (MCMC) algorithm, written in R.
-Specifically, we sampled the unknown parameters $A$, $K$, $M$ and $B$ with Metropolis-Hastings, and used Gibbs sampling to estimate all other unknown parameters.
+Specifically, we sampled the unknown parameters $A$, $K$, $M$ and $B$ with Metropolis-Hastings, and used Gibbs sampling to estimate all other unknown parameters [see @Gilks1995; @Gelman2013].
 Posterior inference on the modern gradient is based on four chains with 60,000 iterations each, 10,000 of which were discarded as burn-in, and keeping every 10th iteration, resulting in a total of 20,000 iterations.
-The modern simulated gradients were estimated in one chain with 10,000 iterations for each of the 1,000 random samples.
+The re-sampled, simulated gradients were estimated in one chain with 10,000 iterations for each of the 1,000 random samples.
 5,000 iterations each were discarded as burn-in, and every 25th iteration was kept, resulting in a total of 200,000 iterations across all 1,000 model runs.
@@
 ```
-Modelled sea surface temperature estimates were generated with eq.
-?, calculating the sea surface temperatures at any latitude with the parameter estimates of each iteration from the posterior.
-The median, $2.5^{th}$ and $97.5^{th}$ percentile of temperatures where then taken from all temperature estimates obtained at the latitude of interest.
+Modeled sea surface temperature estimates were generated with [@eq-nu], calculating the sea surface temperatures at any latitude with the parameter estimates of each iteration from the posterior.
+The median, $2.5^{th}$ and $97.5^{th}$ percentiles of temperatures where then taken from all temperature estimates obtained at the latitude of interest.
 
@@
 
-Differences between Eocene and modern temperatures at a certain latitude were calculated by randomly pairing all iterations of the posterior from the Eocene and modern temperature gradient model, calculating the Eocene and modern temperature using the respective iterations, taking the difference, and then calculating the median ($2.5^{th}$, $97.5^{th}$ percentile) from all pairs of iterations.\
+Differences between Eocene and modern temperatures at a certain latitude were calculated by randomly pairing all iterations of the posterior from the Eocene and modern temperature gradient model, calculating the Eocene and modern temperature using the respective iterations, taking the difference, and then calculating the median ($2.5^{th}$, $97.5^{th}$ percentile) from all pairs of iterations.
+
 Global average temperatures were calculated by taking the weighted mean of the median ($2.5^{th}$, $97.5^{th}$ percentile) temperature estimates in 1 degree latitudinal bins.
@@
 The modern, latitudinal gradient, modeled with all data, spans `r modern_gradient$"50%"` degC from the equator to the poles.
-When modeled with only the early Eocene sampling locations, the median temeprature gradient is `r modern_eocene_gradient$"50%"` (95% CI: `r modern_eocene_gradient$"2.5%"` - `r modern_eocene_gradient$"97.5%"`) degC.
+When modeled with only the early Eocene sampling locations, the median temperature gradient is `r modern_eocene_gradient$"50%"` (95% CI: `r modern_eocene_gradient$"2.5%"` - `r modern_eocene_gradient$"97.5%"`) degC.
 A random sampling of modern temperatures at latitudes corresponding to the sampled Eocene palaeolatitudes would thus allow the reconstruction of the modern latitudinal temperature gradient with good accuracy, with expected deviations from the actual gradient as low as `r modern_eocene_gradient_diff$"2.5%"` to `r modern_eocene_gradient_diff$"97.5%"` degC (95 % CI).

### 2023-05-12 manuscript/main.qmd
update results, reformat suppl. material

@@
       - heightrounded
+header-includes:
+  - \usepackage{newunicodechar}
+  - \newcommand{\$^\circ$C}{\textdegree C}
 bibliography: references.bib
@@
 
-Reconstructing past climate is essential to understanding the long-term evolution of the Earth's climate system and contextualising current global warming [@royer2004; @Burke2018].
+Reconstructing past climate is essential to understanding the long-term evolution of the Earth's climate system and contextualising current global warming [@Royer2004; @Burke2018].
 Geochemical proxies, such as oxygen isotopes (&delta;^18^O), are regularly used to provide quantitative palaeotemperature estimates for the geological past at a range of temporal and spatial scales [e.g.
@@
 @Veizer2015; @Hollis2019; @song2019; @Grossman2022; @Judd2022] are enabling a better understanding of the complex and dynamic nature of the Earth's climate system.
-However, a fundamental challenge remains which must be addressed to exploit the full potential of these advances; the proxy record of past climate is spatially incomplete and biased by imperfect preservation and uneven sampling [@judd2020; @Jones2022; @Judd2022].
+However, a fundamental challenge remains which must be addressed to exploit the full potential of these advances: the proxy record of past climate is spatially incomplete and biased by imperfect preservation and uneven sampling [@judd2020; @Jones2022; @Judd2022].
 
@@
 Interpolation is also sometimes used to bridge spatial gaps in palaeoclimate data [@Taylor2004], taking advantage of the autoregressive nature of climatic data: much of the information on the climate of any given location is contained in the climate data of nearby locations [@Reynolds1994].
-Adding to this, some proxy-based reconstructions use statistical modelling to infer palaeoclimatic patterns.
+Adding to this, some proxy-based reconstructions use statistical modeling to infer palaeoclimatic patterns.
 For example, polynomial regression [@Bijl2009] and cosine functions [@Inglis2020] have been used to reconstruct latitudinal temperature gradients, and 2D-reconstructions of surface temperatures have been created with Gaussian process regression [@Inglis2020].
 These approaches work well for interpolating relatively well-sampled data, but the absence of constraints on the modelled parameters means that such models can produce unrealistic temperature estimates when extrapolating from sparsely sampled data.
-Statistical modelling in a Bayesian framework can help overcome this problem by requiring the explicit specification of priors for the model parameters, which can be used to express physical constraints [@Chandra2021].
+Statistical modeling in a Bayesian framework can help overcome this problem by requiring the explicit specification of priors for the model parameters, which can be used to express physical constraints [@Chandra2021].
 
@@
 
-**Coral reefs.** Today, shallow warm-water coral reefs are limited to tropical and subtropical latitudes (\~34°N--32 degS), with minimum sea surface temperature (\~18 degC) tolerances being the primary constraint on this distribution [@johannes1983; @kleypas1999; @yamano2001].
-As coral reefs reside at the upper thermal limit of the oceans today, their maximum sea surface temperature tolerance is less well-constrained with some studies suggesting up to 35.6 degC in the geological past [@jones2022reef].
+**Coral reefs.** Today, shallow warm-water coral reefs are limited to tropical and subtropical latitudes (\~34$^\circ$N--32$^\circ$S), with minimum sea surface temperature (\~18$^\circ$C) tolerances being the primary constraint on this distribution [@johannes1983; @kleypas1999; @yamano2001].
+As coral reefs reside at the upper thermal limit of the oceans today, their maximum sea surface temperature tolerance is less well-constrained with some studies suggesting up to 35.6$^\circ$C in the geological past [@jones2022reef].
 Nevertheless, coral reefs have frequently been recognised as tracers of past (sub-)tropical conditions [@ziegler1984; @kiessling2001].
-During the Eocene, coral communities and reefs expanded across tropical and temperate latitudes, with some communities found up to palaeolatitudes of 43 degC N [@Zamagni2012].
+During the Eocene, coral communities and reefs expanded across tropical and temperate latitudes, with some communities found up to palaeolatitudes of 43 $^\circ$N [@Zamagni2012].
 Using a compilation of Paleocene--early Eocene coral reefs and community localities [@Zamagni2012], we generate quantitative sea surface temperature estimates for the ECCO.
@@
 We exclude coral knobs and coral-bearing mounds which might have broader climatic limits than coral reef ecosystems.
-This filtering resulted in four unique coral reef localities remaining for the EECO, all of which conform to the modern latitudinal range of coral reefs (\<34 deg N).
-Subsequently, we use statistically derived temperature limits (minimum = 21 degC, average = 27.6 degC, maximum = 29.5 degC) from the published literature [@kleypas1999] to define a normal probability distribution of potential temperature values for coral reef localities.
+This filtering resulted in four unique coral reef localities remaining for the EECO, all of which conform to the modern latitudinal range of coral reefs (\<34$^\circ$ N).
+Subsequently, we use statistically derived temperature limits (minimum = 21$^\circ$C, average = 27.6$^\circ$C, maximum = 29.5$^\circ$C) from the published literature [@kleypas1999] to define a normal probability distribution of potential temperature values for coral reef localities.
 This normal probability distribution was defined with a mean of 27.6 and a standard deviation of 2.125, placing 97.5% of the probability density above the minimum.
-As the distribution of modern corals is skewed towards warmer temperatures, this approach results in 16.5% of the probability being placed on temperatures \> 29.5 degC, allowing for the possibility that Eocene coral reefs were adapted to warmer conditions than present-day coral reefs.
+As the distribution of modern corals is skewed towards warmer temperatures, this approach results in 16.5% of the probability being placed on temperatures \> 29.5$^\circ$C, allowing for the possibility that Eocene coral reefs were adapted to warmer conditions than present-day coral reefs.
 
 **Mangroves.** Mangroves are distributed throughout the tropics and subtropics today.
-While factors beside SST influence the distribution of mangroves, empirical, lower temperature limits have been established for the genera *Avicennia* (15.6 degC) and *Rhizophora* (20.7 degC) [@Quisthoudt2012].
+While factors beside SST influence the distribution of mangroves, empirical, lower temperature limits have been established for the genera *Avicennia* (15.6$^\circ$C) and *Rhizophora* (20.7$^\circ$C) [@Quisthoudt2012].
 Both *Avicennia* and members of the Rhizophoraceae family were widespread in the early Eocene, but only *Avicennia* occurred at polar latitudes [@Suan2017; @Popescu2021].
@@
 
-1)  *Avicennia*-only assemblages ($n=2$): the absence of Rhizophoraceae is indicative of temperatures being between 15.6 degC (lower temperature limt of *Avicennia*) and 20.7 (lower temperature limit of *Rhizophora*).
-However, a value of 22.5 degC is ascribed as the upper temperature limit here as *Rhizophora* is rare below this temperature.
-We define the *Avicennia*-only temperature distribution as a normal distribution with a mean of 19.05 degC and a standard deviation of 1.725, resulting in 95% of the probability density being placed within the temperature limits.
+1)  *Avicennia*-only assemblages ($n=2$): the absence of Rhizophoraceae is indicative of temperatures being between 15.6$^\circ$C (lower temperature limt of *Avicennia*) and 20.7 (lower temperature limit of *Rhizophora*).
+However, a value of 22.5$^\circ$C is ascribed as the upper temperature limit here as *Rhizophora* is rare below this temperature.
+We define the *Avicennia*-only temperature distribution as a normal distribution with a mean of 19.05$^\circ$C and a standard deviation of 1.725, resulting in 95% of the probability density being placed within the temperature limits.
 
-2)  *Avicennia* and Rhizophoraceae assemblages ($n=5$): the presence of both groups suggests that the locality should have a minimum temperature of 20.7 degC (lower temperature limit of *Rhizophora*).
-As the upper thermal limits of *Aviciennia* and *Rhizophora* are not well established in @Quisthoudt2012, we assign the same maximum temperature limits (29.5 degC) as coral reef localities as mangroves are also widely distributed throughout tropical regions.
+2)  *Avicennia* and Rhizophoraceae assemblages ($n=5$): the presence of both groups suggests that the locality should have a minimum temperature of 20.7$^\circ$C (lower temperature limit of *Rhizophora*).
+As the upper thermal limits of *Aviciennia* and *Rhizophora* are not well established in @Quisthoudt2012, we assign the same maximum temperature limits (29.5$^\circ$C) as coral reef localities as mangroves are also widely distributed throughout tropical regions.
 Consequently, we define the temperature distribution for this locality as a normal distribution with a mean of 25.1 and a standard deviation of 2.2, with 95% probability density within the temperature limits.
@@
 Using the same idealised gradients, we also tested whether our model could accurately reconstruct latitudinal temperature gradients using the palaeogeographic distribution of Eocene samples (n = 34), providing an empirical example distribution that captures both limited sample size and skewed geographic distributions.
+To assess how well the modeled gradients reconstructed from limited sampling fit the idealised gradients, we calculated an R^2 for Bayesian regression models [@Gelman2019].
+For every iteration from the posterior, we intercepted the modeled and the idealised gradient in intervals of 1$^\circ$ latitude, and calculated the R^2 based on these values.
+We report the median, and 2.5th and 97.5th percentile of the resulting R^2 measures.
 
-We also used our model to estimate the modern sea surface temperature gradient, using annual sea surface mean temperatures from Bio-Oracle [@Assis2018], down-sampled to a $1^\circ \times 1 ^\circ$ raster (n = 46,131)
+To test whether out model can accurately model the the modern sea surface temperature gradient, and to facilitate comparison with the Eocene gradient, we applied our model to annual sea surface mean temperatures from Bio-Oracle [@Assis2018], down-sampled to a $1^\circ \times 1 ^\circ$ raster (n = 46,131).
+The R^2 for the modern gradient was calculated as described above, comparing the modeled gradient and the empirical temperature averages in 1$^\circ$ latitude bins.
+Only the medians are reported for the modern gradient, as the 95% credible intervals are extremely narrow due to the high precision of the posterior estimates.
 
-To reconstruct the idealised gradients and the modern gradient, we used a simplified, non-hierarchical version of our model, as every location is associated with only one temperature value.
+To reconstruct the idealised gradients and the modern gradient, we used a simplified, non-hierarchical version of our model, as every location is associated with only one temperature value, making the hierarchical structure superfluous.
 We thus substituted temperature ($t_j$) for $\mu_j$ in [@eq-mu] and [@eq-sigma].
@@
 Specifically, we sampled the unknown parameters $A$, $K$, $M$ and $B$ with Metropolis-Hastings, and used Gibbs sampling to estimate all other unknown parameters [see @Gilks1995; @Gelman2013].
-Posterior inference on the modern gradient is based on four chains with 60,000 iterations each, 10,000 of which were discarded as burn-in, and keeping every 10th iteration, resulting in a total of 20,000 iterations.
-The re-sampled, simulated gradients were estimated in one chain with 10,000 iterations for each of the 1,000 random samples.
+Posterior inference on the modern gradient is based on four chains with 60,000 iterations each, 10,000 of which were discarded as burn-in.
+Every 10th iteration was retained, resulting in a total of 20,000 iterations with low autocorrelation.
+The re-sampled, simulated gradients were modeled in one chain with 10,000 iterations for each of the 1,000 random samples.
 5,000 iterations each were discarded as burn-in, and every 25th iteration was kept, resulting in a total of 200,000 iterations across all 1,000 model runs.
@@
 
-The latitudinal gradient is calculated as the difference between the modeled temperature at the equator (0 deg latitude) and at the poles (90 deg absolute latitude).
+The latitudinal gradient is calculated as the difference between the modeled temperature at the equator (0$^\circ$ latitude) and at the poles (90$^\circ$ absolute latitude).
 Given the sigmoidal shape of the modern as well as the Eocene gradient (see Fig.
@@
 
-Global average temperatures were calculated by taking the weighted mean of the median ($2.5^{th}$, $97.5^{th}$ percentile) temperature estimates in 1 degree latitudinal bins.
+Global average temperatures were calculated by taking the weighted mean of the median ($2.5^{th}$, $97.5^{th}$ percentile) temperature estimates in 1$^\circ$ latitudinal bins.
 The weights were set to the proportion of global surface area in each latitudinal bin, i.e.
-decreasing with increasing latitude as: $$weights = sin(\alpha_{1,i}) - sin(\alpha_{2,i}) ,$$ where $\alpha1$ is the upper, and $\alpha2$ is the lower latitudinal boundary of bin $i$.
+decreasing with increasing latitude as: $$weights = sin(\alpha_{1,i}) - sin(\alpha_{2,i}) ,$${#eq-weights} where $\alpha_1$ is the upper, and $\alpha_2$ is the lower latitudinal boundary of bin $i$.
 
@@
 
+modern_gradient_empirical <- readRDS("results/modern/modern_empirical_gradient.rds")
+
 # sdy
@@
 # modern R2
-R2_modern <- readRDS("results/modern/modern_model_empirical_R2.rds")
+R2_modern <- readRDS("results/modern/modern_model_empirical_R2_gelman.rds")
 
@@
 
-## Model validation
+## Model validation and modern gradient
 
@@
 Model reconstructions of simulated latitudinal temperature gradients at various sample sizes.
-Each column depicts a different reconstruction for given sample sizes: 5, 10, 20, and 34 (sample size for the EECO).
+Each column depicts a different reconstruction for given sample sizes: 5, 10, 20, and 34 (latitudes of EECO samples).
 Each row depicts a different simulated latitudinal temperature gradient that represents idealised climatic states: extreme greenhouse, greenhouse, icehouse, extreme icehouse.
 The black line illustrates the simulated gradient.
-The blue line depicts the reconstructed gradient represented by the median sea surface temperature value estimated from 1000 model runs.
-The blue shading depicts the 90%, 95%, and 99% credible intervals.
-Bold black text within each panel depicts the coefficient of determination (*R*^2^ value) for estimating goodness of fit between the simulated and modelled gradient.
+The blue line depicts the reconstructed gradient represented by the median sea surface temperature value estimated from 1000 model runs with different random samples (first three columns), and a single run with the EECO latitudinal sampling distribution (fourth column).
+The blue shadings depict the 90%, 95%, and 99% credible intervals.
+Bold black text within each panel depicts the coefficient of determination (*R*^2) for estimating goodness of fit between the simulated and modeled gradient.
 The median (50%) *R*^2^ value along with the 2.5% and 97.5% confidence intervals from the 1000 model runs are shown.
@@
 ```
-\[Fig 3 - Modern gradient\]
 
@@
 Estimate of the present-day latitudinal temperature gradient using the Bayesian hierarchical model.
-The present-day empirical (black line) latitudinal temperature gradient (mean sea surface temperature) and that estimated by the Bayesian hierarchical model (green line) when using all cell values from the Bio-ORACLE grid of mean sea surface temperature.
+The present-day empirical (black line) latitudinal temperature gradient (mean sea surface temperature) and that estimated by the Bayesian hierarchical model (green line) when using all cell values from the Bio-ORACLE grid of mean sea surface temperatures.
 Grey points depict the individual cell values of the Bio-ORACLE grid.
@@
 
-The average modern temperature gradient can be closely approximated with our model when using the full modern SST dataset (Fig.
-3a); almost all of the variation in the empirical median temperatures in bins of 1 deg absolute latitude is explained by the modelled gradient ($R^2 = `r R2_modern`$).
+Our Bayesian model is able to accurately model a range of idealised temperature gradients, ranging from extreme icehouse to "super greenhouse" scenarios (Fig.
+2).
+Random latitudinal sampling results in highly accurate reconstructions already at a sample size of 10 for the icehouse scenarios (95 % CI of R^2 > 0.9).
+Greenhouse scenarios require more random samples to accurately predict high-latitude temperatures.
+This is because in the absence of high-latitude samples, the modeled gradient is heavily influenced by the priors, which we based on the modern, the only empirically known latitudinal temperature gradient.
+A sampling distribution resembling that of the early Eocene data set used in this study allows for a highly accurate reconstruction of even the extreme greenhouse scenario (95 % CI of R^2 > 0.95).
+
+The average, modern temperature gradient can be closely approximated with our model when using the full modern SST dataset (Fig.
+3); almost all of the variation in the empirical median temperatures in bins of 1$^\circ$ absolute latitude is explained by the modeled gradient ($R^2 = `r R2_modern`$).
+The empirical gradient spans `r modern_gradient_empirical`$^\circ$C from the equator to the poles, the modeled gradient is only very slightly higher at `r modern_gradient$"50%"`$^\circ$C.
+The modern, global men sea surface temperature based on our modeled gradient is `r motemp_global$"50%"`$^\circ$C.
+
+<!-- REMOVED
 Reducing the amount and the spatial coverage of data points to match the early Eocene sampling distribution ($n = 34$) still results in a close match of the modeled gradient with the empirical gradient (Fig.
@@
 3a).
-The modern, latitudinal gradient, modeled with all data, spans `r modern_gradient$"50%"` degC from the equator to the poles.
-When modeled with only the early Eocene sampling locations, the median temperature gradient is `r modern_eocene_gradient$"50%"` (95% CI: `r modern_eocene_gradient$"2.5%"` - `r modern_eocene_gradient$"97.5%"`) degC.
-A random sampling of modern temperatures at latitudes corresponding to the sampled Eocene palaeolatitudes would thus allow the reconstruction of the modern latitudinal temperature gradient with good accuracy, with expected deviations from the actual gradient as low as `r modern_eocene_gradient_diff$"2.5%"` to `r modern_eocene_gradient_diff$"97.5%"` degC (95 % CI).
+The modern, latitudinal gradient, modeled with all data, spans `r modern_gradient$"50%"`$^\circ$C from the equator to the poles.
+When modeled with only the early Eocene sampling locations, the median temperature gradient is `r modern_eocene_gradient$"50%"` (95% CI: `r modern_eocene_gradient$"2.5%"` - `r modern_eocene_gradient$"97.5%"`)$^\circ$C.
+A random sampling of modern temperatures at latitudes corresponding to the sampled Eocene palaeolatitudes would thus allow the reconstruction of the modern latitudinal temperature gradient with good accuracy, with expected deviations from the actual gradient as low as `r modern_eocene_gradient_diff$"2.5%"` to `r modern_eocene_gradient_diff$"97.5%"`$^\circ$C (95 % CI).
 
@@
 Modeling the latitudinal temperature gradient facilitates accurate reconstruction of global mean surface temperatures (GMST).
-The modern, global mean sea surface temperature (GMsST) estimate is `r motemp_global$"50%"` degC when using the full, modern data set.
-When reduced to the early Eocene sampling distribution, the modern GMsST is estimated at `r moeotemp_global$"50%"` (`r moeotemp_global$"2.5%"` - `r moeotemp_global$"97.5%"`) degC.
-The deviations from the full modern estimate are again very low, with a 95 % credible interval of `r moeotemp_global_diff$"2.5%"` to `r moeotemp_global_diff$"97.5%"` degC.
+The modern, global mean sea surface temperature (GMsST) estimate is `r motemp_global$"50%"`$^\circ$C when using the full, modern data set.
+When reduced to the early Eocene sampling distribution, the modern GMsST is estimated at `r moeotemp_global$"50%"` (`r moeotemp_global$"2.5%"` - `r moeotemp_global$"97.5%"`)$^\circ$C.
+The deviations from the full modern estimate are again very low, with a 95 % credible interval of `r moeotemp_global_diff$"2.5%"` to `r moeotemp_global_diff$"97.5%"`$^\circ$C.
+-->
 
@@
 The modeled Eocene temperature gradient is starkly different from the modern (Fig 4).
-Modelled, median equatorial temperatures are `r tdiff_lat0$"50%"` (`r tdiff_lat0$"2.5%"` - `r tdiff_lat0$"97.5%"`) degC higher for the EECO, and polar temperatures are higher by `r tdiff_lat90$"50%"` (`r tdiff_lat90$"2.5%"` - `r tdiff_lat90$"97.5%"`) degC.
-This results in a flattened latitudinal temperature gradient of `r eocene_gradient$"50%"` (`r eocene_gradient$"2.5%"` - `r eocene_gradient$"97.5%"`) degC for the EECO, as opposed to `r modern_gradient$"50%"` degC for the modern.
-To facilitate the comparison with latitudinal gradients reported in the literature, which sometimes do not report temperatures in very high latitudes, we report also the EECO gradient between the equator and the modern-day polar circle (66.6 deg), which is slightly lower at `r eocene_gradient_polar_circle$"50%"` (`r eocene_gradient_polar_circle$"2.5%"` - `r eocene_gradient_polar_circle$"97.5%"`).
+Modelled, median equatorial temperatures are `r tdiff_lat0$"50%"` (`r tdiff_lat0$"2.5%"` - `r tdiff_lat0$"97.5%"`)$^\circ$C higher for the EECO, and polar temperatures are higher by `r tdiff_lat90$"50%"` (`r tdiff_lat90$"2.5%"` - `r tdiff_lat90$"97.5%"`)$^\circ$C.
+This results in a flattened latitudinal temperature gradient of `r eocene_gradient$"50%"` (`r eocene_gradient$"2.5%"` - `r eocene_gradient$"97.5%"`)$^\circ$C for the EECO, as opposed to `r modern_gradient$"50%"`$^\circ$C for the modern.
+To facilitate the comparison with latitudinal gradients reported in the literature, which sometimes do not report temperatures in very high latitudes, we report also the EECO gradient between the equator and the modern-day polar circle (66.6$^\circ$), which is slightly lower at `r eocene_gradient_polar_circle$"50%"` (`r eocene_gradient_polar_circle$"2.5%"` - `r eocene_gradient_polar_circle$"97.5%"`)$^\circ$C.
 
 The high variability of EECO palaeotemperature proxies, particularly in the mid-latitudes, and the scarcity of high-latitude data, result in substantial uncertainties in the modeled temperature gradient.
-This is reflected in the residual standard deviation ($\sigma$) of the EECO gradient, `r eeco_sdy$"50%"` (`r eeco_sdy$"2.5%"` - `r eeco_sdy$"97.5%"`) degC, more than twice as high than $\sigma$ of the modern gradient, `r modern_sdy$"50%"`.
+This is reflected in the residual standard deviation ($\sigma$) of the EECO gradient, `r eeco_sdy$"50%"` (`r eeco_sdy$"2.5%"` - `r eeco_sdy$"97.5%"`)$^\circ$C, more than twice as high than $\sigma$ of the modern gradient, `r modern_sdy$"50%"`.
 This signifies that the early Eocene data fit less well to the logistic, latitudinal gradient model, which can also be seen from the drastic departure of some of the proxy data from the gradient estimates (Fig.
@@
 
-The early Eocene GMsST is estimated at `r eotemp_global$"50%"` (`r eotemp_global$"2.5%"` - `r eotemp_global$"97.5%"`) degC, `r  round(as.numeric(eotemp_global$"50%") - as.numeric(motemp_global$"50%"),1)` degC higher than the modern.
-A model run excluding the ecological proxies results in a GMsST that is higher by `r eoceneproxy_global_diff$"50%"` (`r eoceneproxy_global_diff$"2.5%"` - `r eoceneproxy_global_diff$"97.5%"`) degC.
-The median latitudinal gradient is similar when excluding the ecological proxies, with a median of `r eocene_proxy_gradient$"50%"` degC, but with a `r proxy_uncertain_diff_percent`% wider 95% CI (Supplementary materials).
+The early Eocene GMsST is estimated at `r eotemp_global$"50%"` (`r eotemp_global$"2.5%"` - `r eotemp_global$"97.5%"`)$^\circ$C, `r  round(as.numeric(eotemp_global$"50%") - as.numeric(motemp_global$"50%"),1)`$^\circ$C higher than the modern.
+A model run excluding the ecological proxies increases the GMsST by `r eoceneproxy_global_diff$"50%"` (`r eoceneproxy_global_diff$"2.5%"` - `r eoceneproxy_global_diff$"97.5%"`)$^\circ$C.
+The median latitudinal gradient is similar when excluding the ecological proxies, with a median of `r eocene_proxy_gradient$"50%"`$^\circ$C, but with a `r proxy_uncertain_diff_percent`% wider 95% CI (Supplementary materials).
 This indicates that the ecological proxy data are broadly in agreement with the geochemical proxies, while providing additional constraints on the shape of the early Eocene temperature gradient.
@@
 Due to the limited spatial coverage of the early Eocene proxy record, and due to the added model complexity of simultaneously estimating a model across both hemispheres, we pooled the proxy data across both hemispheres.
-Applying the model separately to each hemispheres results in substantial differences in hemispherical, average temperatures, with the southern hemisphere being warmer by `r eeco_south_north_global_diff$"50%"` (`r eeco_south_north_global_diff$"2.5%"` - `r eeco_south_north_global_diff$"97.5%"`) degC.
-The inferred latitudinal gradient is somewhat steeper in the northern hemisphere (`r eocene_north_south_gradient_diff$"50%"` degC, although the 95% CI spans `r eocene_north_south_gradient_diff$"2.5%"` to `r eocene_north_south_gradient_diff$"97.5%"` degC), but the large uncertainties associated with both gradients, and the lack of polar proxy data in the southern hemisphere preclude a more precise statement (Fig.
-S3).)
-
-## Fig 4 - compare Eocene gradient with modern gradient, also with previous reconstructions / ES models?
+Applying the model separately to each hemispheres results in substantial differences in hemispherical, average temperatures, with the southern hemisphere being warmer by `r eeco_south_north_global_diff$"50%"` (`r eeco_south_north_global_diff$"2.5%"` - `r eeco_south_north_global_diff$"97.5%"`)$^\circ$C.
+The inferred latitudinal gradient is somewhat steeper in the northern hemisphere (`r eocene_north_south_gradient_diff$"50%"`$^\circ$C, although the 95% CI spans `r eocene_north_south_gradient_diff$"2.5%"` to `r eocene_north_south_gradient_diff$"97.5%"`$^\circ$C), but the large uncertainties associated with both gradients, and the lack of polar proxy data in the southern hemisphere preclude a more precise statement (Fig.
+S3).
 
@@
 
-Estimates of the present-day and early Eocene climatic optimum latitudinal temperature gradient using the Bayesian hierarchical model.
-The estimated present-day (green line) and early Eocene climatic optimum (purple line) latitudinal temperature gradient.
+Estimates of the early Eocene climatic optimum and the present-day latitudinal temperature gradient using the Bayesian model.
+The estimated present-day (turquoise line) and early Eocene climatic optimum (purple line) latitudinal temperature gradient.
 The respective ribbons (green and purple shading) depict the 95% credible intervals of the estimated gradients.
@@
 Geochemical data are plotted by their point estimate temperature value.
-Ecological data are plotted by their mean temperature values sampled from their respective normal distributions.
+Ecological data are plotted at the mean temperature values of their respective normal distributions.
 :::
-
--   x1 Draw samples from the posterior gradient to get a latitudinal gradient of deg C / deg lat with uncertainty (intercept gradient at 0 and 90 deg lat)
-
--   x2 generate global average temperature, and tropical, temperate, polar temperatures in a similar way, accounting for area
 
@@
 Combining the occurrences of climate-sensitive plant communities [@Greenwood1995], reptiles [@Marwick2007], leaf shapes [@Peppe2011], with geochemical proxies offers substantial potential for improving quantitative palaeoclimate reconstructions across the Phanerozoic.
-Our modelling framework offers a straightforward way of integrating ecological climate data with other proxy data: The hierarchical model structure accounts for variation of temperature estimates from proxies at individual localities, which is treated equivalent to the uncertainty associated with the ecological temperature proxies.
+Our modeling framework offers a straightforward way of integrating ecological climate data with other proxy data: The hierarchical model structure accounts for variation of temperature estimates from proxies at individual localities, which is treated equivalent to the uncertainty associated with the ecological temperature proxies.
 A local temperature estimate, based on multiple geochemical proxies, thus has the same weight as a local temperature estimate obtained from the occurrence of a climate-sensitive plant community, whilst preserving the uncertainty associated with each estimate.
@@
 The geochemical proxy record and ecological data indicate that the latitudinal SST gradient of the early Eocene climatic optimum was significantly shallower than the modern [@Huber2011], but beyond that, there is little agreement.
-@Evans2018 estimate the SST gradient at 32+-10% of the modern, which would correspond to \~`r round( as.numeric(modern_gradient$"50%") - 0.32*as.numeric(modern_gradient$"50%"),0)`+- `r format(round( 0.10*as.numeric(modern_gradient$"50%"),0),nsmall=0)`degC, using the difference between the mean tropical and deep ocean data.
-The modeled SST of @Tierney2017 using TEX86 is \~ `r round(35.8-23.8,0)` degC, when taking the difference between temperatures at the polar circles and the equator.
-@Inglis2020 include both terrestrial air temperature and SST estimates in their model of the latitudinal temperature gradient in their Figure S4a; they model a polar circle to equatorial gradient of \~ `r round(30.7-(20.2+14.3)/2,0)`degC.
-All of those estimates are significantly lower than the SST gradient predicted by an EECO climate model ensemble [@Lunt2012, as shown in @Tierney2017], which predicts a polar circle to equatorial gradient of `r round(31.5-5.5,0)`degC.
-@Keating-Bitonti2011 show a proxy-based SST gradient of \~ `r round(30-(19+15.6)/2,0)`degC, and @Bijl2009 model a gradient of \~`r round(36-27,0)`degC in the northern and only \~`r 36-31`degC in the southern hemisphere for the early Eocene.
-Our polar circle to equatorial gradient is lower than most previous estimates at `r eocene_gradient_polar_circle$"50%"`degC, although the 95% credible interval overlap with several of the listed gradient estimates.
+@Evans2018 estimate the SST gradient at 32+-10% of the modern, which would correspond to \~`r round( as.numeric(modern_gradient$"50%") - 0.32*as.numeric(modern_gradient$"50%"),0)`+- `r format(round( 0.10*as.numeric(modern_gradient$"50%"),0),nsmall=0)`$^\circ$C, using the difference between the mean tropical and deep ocean data.
+The modeled SST of @Tierney2017 using TEX86 is \~ `r round(35.8-23.8,0)`$^\circ$C, when taking the difference between temperatures at the polar circles and the equator.
+@Inglis2020 include both terrestrial air temperature and SST estimates in their model of the latitudinal temperature gradient in their Figure S4a; they model a polar circle to equatorial gradient of \~ `r round(30.7-(20.2+14.3)/2,0)`$^\circ$C.
+All of those estimates are significantly lower than the SST gradient predicted by an EECO climate model ensemble [@Lunt2012, as shown in @Tierney2017], which predicts a polar circle to equatorial gradient of `r round(31.5-5.5,0)`$^\circ$C.
+@Keating-Bitonti2011 show a proxy-based SST gradient of \~ `r round(30-(19+15.6)/2,0)`$^\circ$C, and @Bijl2009 model a gradient of \~`r round(36-27,0)`$^\circ$C in the northern and only \~`r 36-31`$^\circ$C in the southern hemisphere for the early Eocene.
+Our polar circle to equatorial gradient is lower than most previous estimates at `r eocene_gradient_polar_circle$"50%"`$^\circ$C, although the 95% credible interval overlap with several of the listed gradient estimates.
 
@@
 Polar temperature estimates from our model are thus conservative in that they admit large uncertainty where data is absent, which is desirable.
-However, the presence of high proxy-derived temperature estimates at \~ 60 deg latitudes forces the modeled median temperature curve to be too high at \~ 24 degC, relative to the temperatures indicated by the high-latitude mangrove communities (15.6 - 22.5 degC).
+However, the presence of high proxy-derived temperature estimates at \~ 60$^\circ$ latitudes forces the modeled median temperature curve to be too high at \~ 24$^\circ$C, relative to the temperatures indicated by the high-latitude mangrove communities (15.6 - 22.5$^\circ$C).
 In contrast, the extrapolated polar temperatures of most previous proxy-based models are likely too low, given the abundance of ecological data indicating temperate or subtropical high-latitude climates during the EECO [@Pross2012; @Popescu2021]
@@
 
-Recent GMsST estimates of the EECO and of the early Eocene range from 23.4 to 37.1 degC, with the lowest GMSTs being derived from d18O, and the higher estimates including TEX86 [@Inglis2020].
-Many studies include both marine and terrestrial proxies to derive GMST estimates, but despite great differences in proxy selection and in the calculation of global average temperatures, many recent estimates fall in the range of 27 - 29.5 degC [@Hansen2013; @Caballero2013; @Cramwinckel2018; @Zhu2019], similar to our GMsST estimate of `r eotemp_global$"50%"` degC.
+Recent GMsST estimates of the EECO and of the early Eocene range from 23.4 to 37.1$^\circ$C, with the lowest GMSTs being derived from d18O, and the higher estimates including TEX86 [@Inglis2020].
+Many studies include both marine and terrestrial proxies to derive GMST estimates, but despite great differences in proxy selection and in the calculation of global average temperatures, many recent estimates fall in the range of 27 - 29.5$^\circ$C [@Hansen2013; @Caballero2013; @Cramwinckel2018; @Zhu2019], similar to our GMsST estimate of `r eotemp_global$"50%"`$^\circ$C.
 

### 2023-05-12 manuscript/supplementary.qmd
update results, reformat suppl. material

--- 2023-03-09 manuscript/supplementary.qmd
@@
       - heightrounded
+crossref:
+  fig-title: Supplementary Figure
+  fig-prefix: Supplementary Figure
 bibliography: references.bib
-csl: geology.csl
+csl: climate-of-the-past.csl
 ---
@@
 
-```{r out.width='100%', echo = FALSE, fig.cap = "Fig S1.
+::: {#fig-S1}
+![](../figures/SM/FigS1-Traceplot.jpg){width="16cm"}
+
 Traceplots of a subset of the unknown model parameters.
 The four colours correspond to the four independent model runs.
-a) Traceplot of A, K-A, M, B and $\\sigma$; b) Traceplot of five selected $\\mu_j$; c) Traceplot of five selected $\\sigma_j$.
-All traceplots display mixing of the chains, and relatively quick convergence."}
-setwd('..') # move up to parent folder of /manuscript
-knitr::include_graphics("D://Github/PalaeoClimateGradient/figures/SM/FigS1-Traceplot.jpg", error = FALSE)
-```
+a) Traceplot of A, K-A, M, B and $\sigma$; b) Traceplot of five selected $\mu_j$; c) Traceplot of five selected $\sigma_j$.
+All traceplots display mixing of the chains, and relatively quick convergence."
+:::
 
 ## S2. Gradient with just the geochemical proxy data
+::: {#fig-S2}
+![](../figures/SM/FigS2_geochem_eeco.png){width="10.5cm"}
 
-```{r out.width='100%', echo = FALSE, fig.cap = "Fig S2.
 Latitudinal temperature gradient with just the geochemical proxies (orange), showing the median (line) and 95% credibal interval (shading).
-Symbols with vertical lines show the median and 95% credible intervals of $\\mu$.
-The blue line and shading in the background show the latitudinal temperature gradient with the geochemical and ecological proxy data."}
-setwd('..') # move up to parent folder of /manuscript
-knitr::include_graphics("D://Github/PalaeoClimateGradient/figures/SM/FigS2_geochem_eeco.png", error = FALSE)
-```
+Symbols with vertical lines show the median and 95% credible intervals of $\mu_j$.
+The blue line and shading in the background show the latitudinal temperature gradient with the geochemical and ecological proxy data, as in Figure 4.
+:::
 
-## S3. Seperate gradients on the northern and southern hemisphere
+## S3. Separate gradients on the northern and southern hemisphere
+::: {#fig-S3}
+![](../figures/SM/FigS3_north_south_eeco.png){width="12cm"}
 
-```{r out.width='100%', echo = FALSE, fig.cap = "Fig S3.
 Latitudinal temperature gradient in seperate hemispheres, showing the medians (lines) and 95% credibal intervals (shadings) in the southern (red) and northern hemisphere (blue).
-Symbols with vertical lines show the median and 95% credible intervals of $\\mu$.
+Symbols with vertical lines show the median and 95% credible intervals of $\mu_j$.
 Turquoise symbols in the northern hemisphere highlight the ecological proxy data.
-The grey line and shading in the background show the latitudinal temperature gradient with the data from both hemispheres combined, plotted in both hemispheres."}
-setwd('..') # move up to parent folder of /manuscript
-knitr::include_graphics("D://Github/PalaeoClimateGradient/figures/SM/FigS3_north_south_eeco.png", error = FALSE)
-```
-
-## S4. Different shapes of temperature gradients
-
+The grey line and shading in the background show the latitudinal temperature gradient with the data from both hemispheres combined, plotted in both hemispheres.
+:::
+<!-- REMOVED
 ```{r out.width='100%', echo = FALSE, fig.cap = "Fig S4.
@@
 ```
+-->

### 2023-05-12 R/analyses/assess_sdy.R

@@ -34,3 +34,18 @@ mod_med <- (temp_from_gradient(lat,modern_gradient))$median
 R2_modern <- 1-sum((mod_emp_med$SST - mod_med)^2)/sum((mod_emp_med$SST-mean(mod_emp_med$SST))^2)
 R2_modern <- format(round(R2_modern,3),nsmall = 1)
 saveRDS(R2_modern,"results/modern/modern_model_empirical_R2.rds")
+
+# Gelman's R2
+lats <- seq(0.5,89.5,1)
+
+mod_t <- gradient(lats,modern_gradient)
+
+var_fit <- apply(mod_t,1,function(x) var(x))
+var_res <- apply(mod_t,1,function(x) var(x-mod_emp_med$SST))
+
+R_squared_gelman_modern <- var_fit/(var_fit+var_res)
+quantile(R_squared_gelman_modern, probs = c(0.025,0.5,0.975))
+saveRDS(format(round(median(R_squared_gelman_modern),3),nsmall = 1) ,"results/modern/modern_model_empirical_R2_gelman.rds")
+
+# modern empirical gradient
+saveRDS(format(round(mod_emp_med$SST[1]-mod_emp_med$SST[90],1),nsmall = 1),"results/modern/modern_empirical_gradient.rds")

### 2023-05-22 manuscript/main.qmd
introduction small changes, table

@@
 
-Reconstructing past climate is essential to understanding the long-term evolution of the Earth's climate system and contextualising current global warming [@Royer2004; @Burke2018].
+Understanding the long-term evolution of Earth's climate system and contextualising current global warming relies on accurate reconstructions of past climates [@Royer2004; @Burke2018].
+Recent advances in the synthesis of palaeoclimate data [e.g.
+@Veizer2015; @Hollis2019; @song2019; @Grossman2022; @Judd2022] are offering unprecedented insights into the complex and dynamic nature of the Earth's climate system, but a fundamental challenge remains: the proxy record of past climate is spatially incomplete and biased by imperfect preservation and uneven sampling [@judd2020; @Jones2022; @Judd2022].
+
 Geochemical proxies, such as oxygen isotopes (&delta;^18^O), are regularly used to provide quantitative palaeotemperature estimates for the geological past at a range of temporal and spatial scales [e.g.
 @Liu2009; @Wuchter2004].
-Recent advances in the synthesis of such data [e.g.
-@Veizer2015; @Hollis2019; @song2019; @Grossman2022; @Judd2022] are enabling a better understanding of the complex and dynamic nature of the Earth's climate system.
 However, a fundamental challenge remains which must be addressed to exploit the full potential of these advances: the proxy record of past climate is spatially incomplete and biased by imperfect preservation and uneven sampling [@judd2020; @Jones2022; @Judd2022].
@@
 However, in order to robustly infer regional or global-scale patterns from an incomplete record, spatial gaps must ultimately be bridged.
-One common approach, that requires no additional computation, is the spatial visualisation of proxy-derived temperatures against latitude, showing broad latitudinal temperature trends [@Hollis2019; @Vickers2021].
+One common approach, which requires no additional computation, is the spatial visualisation of proxy-derived temperatures against latitude, showing broad latitudinal temperature trends [@Hollis2019; @Vickers2021].
 Interpolation is also sometimes used to bridge spatial gaps in palaeoclimate data [@Taylor2004], taking advantage of the autoregressive nature of climatic data: much of the information on the climate of any given location is contained in the climate data of nearby locations [@Reynolds1994].
@@
 For example, polynomial regression [@Bijl2009] and cosine functions [@Inglis2020] have been used to reconstruct latitudinal temperature gradients, and 2D-reconstructions of surface temperatures have been created with Gaussian process regression [@Inglis2020].
-These approaches work well for interpolating relatively well-sampled data, but the absence of constraints on the modelled parameters means that such models can produce unrealistic temperature estimates when extrapolating from sparsely sampled data.
+These approaches work well for interpolating relatively densely-sampled data, but the absence of constraints on the modeled parameters means that such models can produce unrealistic temperature estimates when extrapolating from sparse data.
 Statistical modeling in a Bayesian framework can help overcome this problem by requiring the explicit specification of priors for the model parameters, which can be used to express physical constraints [@Chandra2021].
@@
 Spatial gaps in the record of past climate can also be addressed through the integration of additional data.
-For example, lithological and fossil data can be used to infer past climatic conditions based on analogous modern sediments [@Chandra2021], or the premise that the climatic requirements of ancient taxa, biological traits, or ecological communities were similar to those of their nearest modern relatives [@Peppe2011; @Royer2012; @Salonen2019].
+For example, lithological and fossil data can be used to infer past climatic conditions based on analogous modern sediments [@Chandra2021], or based on the premise that the climatic requirements of ancient taxa, biological traits, or ecological communities were similar to those of their nearest modern relatives [@Peppe2011; @Royer2012; @Salonen2019].
 Despite this potential, the integration of geochemical proxy data with other sources of information (e.g.
-ecological data) is rare due to the absence of a quantitative framework [@Burgener2023].
+ecological data) has rarely been realised in a rigorous, quantitative framework [@Burgener2023].
 
@@
 This model expands upon existing, spatially explicit reconstructions of palaeoclimate by allowing for the integration of (1) prior information based on physical principles and on the observed, modern sea surface temperature distribution, and of (2) geochemical and ecological climate proxies in a common, quantitative framework.
-We use a generalised logistic function to infer the shape of the temperature gradient despite a patchy latitudinal coverage, and test the robustness of this method using a down-sampled record of modern SSTs.
+We use a generalised logistic function to infer the shape of the temperature gradient despite a patchy latitudinal coverage, and test the robustness of this method using downsampled, simulated temperature gradients.
+
 We apply this model to the record of the early Eocene climatic optimum (EECO), combining a compilation of geochemical proxies [@Hollis2019], mangrove communities [@Popescu2021], and coral reefs [@Zamagni2012], using a nearest-living-relative approach [see e.g.
 @Greenwood2017] to establish broad temperature ranges for the ecological data.
-Whilst we primarily do so to demonstrate an application of the model, the EECO represents an interval with the warmest sustained temperatures of the Cenozoic [@Pross2012], and a potential analogue for extreme climate warming scenarios [@Burke2018].
+We chose the EECO to demonstrate the application of the model due to its significance as the interval with the warmest sustained temperatures of the Cenozoic [@Pross2012], rendering it a potential analogue for extreme climate warming scenarios [@Burke2018].
+Our integrative approach allows us to shed new light on the long-standing dispute on the shape of the early Eocene temperature gradient between climate modellers and researchers working with the geological and fossil record [@Sloan1990; @Markwick1994; @Huber2011; @Tierney2017; @Inglis2020].
+
+::: {#table-1}
+![](../figures/Table1.png){width="15cm"}
+Table of...
+:::
 

### 2023-05-24 manuscript/main.qmd
discussion part 1
still check use of "EECO" and "early Eocene"

@@
 ::: {#table-1}
-![](../figures/Table1.png){width="15cm"}
-Table of...
+![](../figures/Table1.png){width="16.25cm"}
+Table 1: Inferred latitudinal SST gradients for the early Eocene (EE) or the EECO, as shown in earlier, proxy-based studies.
+For comparison, a gradient derived from a climate model ensemble is also shown.
+
 :::
@@
 
-## Improved reconstruction of temperature gradients and global average temperatures
+## Improved estimation of latitudinal and global palaeotemperatures
 
-The Bayesian model presented herein accurately reconstructs the modern, latitudinal SST gradient from a patchy and relatively small (n = 34) sample of temperature data.
-This is an advancement over previously used linear or quadratic approximations [e.g.
-@Refs], which are inherently unable to capture the sigmoidal shape of the modern SST gradient.
-Our generalised logistic function is flexible enough to approximate the shapes of all of those functions.
+Our results show that relatively small (*n* = 10 - 20) and patchy samples of temperature data are sufficient to reconstruct greenhouse- and icehouse-type, latitudinal SST gradients, using our Bayesian model.
+This is an advancement over previously used linear, quadratic or Gaussian approximations [e.g.
+@Bijl2009; @Tierney2017], which can fit only specific types of gradients.
 As such, our model presents an alternative to non-parametric methods for inferring latitudinal temperature gradients, which are sometimes favoured as they can very flexibly follow the shape of an unknown temperature gradient [e.g.
@@
 However, when used for interpolation or prediction outside the proxy range, non-parametric methods such as Gaussian process regression strictly respond to the data [e.g.
-@Inglis2020], and the resulting large-scale patterns are informed by the idiosyncracies of a patchy proxy record.
+@Inglis2020].
+This means that the idiosyncrasies of a patchy proxy record, potentially afflicted with measurement errors, calibration errors, and palaeogeographic and temporal uncertainty, directly generates the resulting large-scale climate patterns.
 
-In our Bayesian, parametric model, informative priors on the model parameters improve the prediction of sea surface temperatures in the unsampled, very high latitudes: Notice that the upper limit of the credible interval does not increase beyond the range of the data, whereas unconstrained approaches such as splines, Gaussian processes or even standard linear regression could lead to unrealistically high upper bounds in this case [see @Rasmussen2004].
-Prior information on the shape of latitudinal temperature gradients on Earth exists for all geological time periods, and should be used in any palaeoclimate reconstruction.
+In contrast, our Bayesian, parametric model, allows for the inclusion of informative priors on the model parameters.
+The modeled sea surface temperature gradient thus does not strictly follow the proxy data, but instead represents a compromise between the data and our prior knowledge.
+In the EECO example (Fig.
+4), the inclusion of informative priors improves the prediction of sea surface temperatures in the unsampled, very high latitudes: Notice that the upper limit of the credible interval does not increase beyond the range of the data, whereas unconstrained approaches such as splines, Gaussian processes or even standard linear regression could lead to unrealistically high upper bounds in this case [see @Rasmussen2004].
+Prior information on the shape of latitudinal temperature gradients on Earth exists for all geological time periods.
 For example, the greater amount of solar radiation per unit area in low latitudes causes Earth's latitudinal temperature gradient to be broadly negative [@Beer2008].
@@
 However, the calculation of global mean surface temperatures from sparse proxy data is susceptible to bias [@Jones2022].
-By modeling the temperature variation across latitudes, a complete temperature distribution is obtained, filling in gaps in the proxy record through inter- or extrapolation.
-This eliminates the problem that specific climate zones may dominate the proxy record, as intersecting the modeled temperature gradient at narrow latitudinal intervals and accounting for the varying sizes of latitudinal bands by weighting provides unbiased GMST estimates.
-Some deep-time GMST reconstructions account for differential sampling by calculating zonal averages, but do not use the full latitudinal range of temperatures [@Inglis2020].
-We anticipate that applying this improved will significantly alter Phanerozoic, proxy-based temperature curves, particularly when applied to intervals with small or biased samples of temperature proxies.
-[@Royer2004; @Veizer2015; @Jones2022].
+By modeling the temperature variation across latitudes, a complete temperature distribution along a latitudinal axis can be obtained, filling in gaps in the proxy record through inter- or extrapolation.
+This eliminates the common problem that specific climate zones dominate the proxy record.
+Reconstructing the GMST directly from the proxies would lead to an estimate biased towards the well-sampled latitudes.
+Calculating zonal averages alleviates this problem, but this method relies on comprehensive latitudinal coverage [@Inglis2020].
+Instead, our method allows for intersecting the modeled temperature gradient at narrow latitudinal intervals, even when significant latitudinal gaps exist.
+Weighting the temperatures of those latitudinal intervals by area then provides unbiased GMST estimates.
+We anticipate that this improved method may significantly alter Phanerozoic, proxy-based temperature curves, which have often been directly calculated from the proxy record [@Royer2004; @Veizer2015].
+This is particularly relevant for the early Mesozoic and older intervals, for which the spatial coverage is generally poor due to the absence of data from ocean drilling sites [@Jones2022].
 
-## Incorporating ecological constraints in palaeoclimate reconstructions
+## The role of ecological constraints in palaeoclimate reconstructions
 
-Our results exemplify how incorporating quantified ecological temperature constraints can provide more precise temperature reconstructions than geochemical proxies alone, adding to the advances in palaeoclimate reconstructions achieved by integrating lithological data [@Scotese2021].
-Combining the occurrences of climate-sensitive plant communities [@Greenwood1995], reptiles [@Marwick2007], leaf shapes [@Peppe2011], with geochemical proxies offers substantial potential for improving quantitative palaeoclimate reconstructions across the Phanerozoic.
-Our modeling framework offers a straightforward way of integrating ecological climate data with other proxy data: The hierarchical model structure accounts for variation of temperature estimates from proxies at individual localities, which is treated equivalent to the uncertainty associated with the ecological temperature proxies.
+Our results further exemplify how incorporating quantified ecological temperature constraints can provide more precise temperature reconstructions than geochemical proxies alone, adding to the advances in palaeoclimate reconstructions achieved by integrating lithological data [@Scotese2021].
+Combining the occurrences of climate-sensitive plant communities [@Greenwood1995], reptiles [@Markwick2007], leaf shapes [@Peppe2011], with geochemical proxies offers substantial potential for improving quantitative palaeoclimate reconstructions across the Phanerozoic.
+Our modeling framework offers a straightforward, efficient way of integrating ecological climate data with other proxy data: The hierarchical model structure accounts for variation of temperature estimates from proxies at individual localities, which is treated equivalent to the uncertainty associated with the ecological temperature proxies.
 A local temperature estimate, based on multiple geochemical proxies, thus has the same weight as a local temperature estimate obtained from the occurrence of a climate-sensitive plant community, whilst preserving the uncertainty associated with each estimate.
+The model could easily be extended to include uncertainties on individual geochemical proxy data, or to differently weight different proxy records classified as more or less reliable.
 
 Our approach for deriving fully quantitative climate reconstructions from ecological data is borrowed from nearest living relative methods, commonly employed in terrestrial, Cenozoic climate reconstructions [@Fauquette2007; @Pross2012].
-These methods have the problem that the thermal preferences of taxa may have changed over time.
+One major limitation to these methods is that the thermal preferences of taxa may have changed over time.
 More significantly, in the early Eocene, sea surface temperatures may have reached heights unknown in the modern world, and nearest living relative methods based on the modern are inherently unable to predict such elevated temperatures.
 This is especially true for taxa that inhabit the warmest part of the ocean today, e.g.
-coral reefs \[\@\].
-Although coral reefs are threatened by warming sea surface temperatures today \[\@\], it is conceivable that Eocene reef corals were adapted to a warmer climate.
-The fossil record indicates that reef development may have been stunted in the early Eocene, but the few early Eocene coral reefs occur in low latitudes [@Zamagni2012].
-Tropical temperatures predicted by the geochemical proxy record indicate hotter-than-modern tropical temperatures (Fig.
+coral reefs [@kleypas1999].
+Although coral reefs are threatened by warming sea surface temperatures today [@Hoegh2011], it is conceivable that Eocene reef corals were adapted to a warmer climate.
+The fossil record indicates that reef development may have been stunted in the early Eocene, with few early Eocene coral reefs occurring in low latitudes [@Zamagni2012].
+The absence of coral reefs in higher latitudes in the early Eocene could be due to requirements in irrdiance, rather than temperature [@Muir2015].
+Tropical temperatures predicted by the geochemical proxy record indicate hotter-than-modern tropical temperatures for the early Eocene (Fig.
 S2), suggesting that the modern climate range of coral reefs may underestimate the early Eocene thermal niche for coral reefs.
@@
 
-This dilemma could be resolved in future studies by applying machine learning approaches to inferring quantitative temperatures from ecological and lithological proxies: If geochemical proxies and ecological or lithological data are derived from the same location and time interval, the climatic requirements of these taxa or sediments can be learned by using palaeotemperatures from that time interval, instead of or in addition to modern modern temperatures.
+<!-- This dilemma could be resolved in future studies by applying machine learning approaches to inferring quantitative temperatures from ecological and lithological proxies: If geochemical proxies and ecological or lithological data are derived from the same location and time interval, the climatic requirements of these taxa or sediments can be learned by using palaeotemperatures from that time interval, instead of or in addition to modern modern temperatures.
 This approach would ultimately rely on geochemical proxies providing absolute temperature estimates, but the resulting information could be exported to other locations lacking good coverage of geochemical proxies via their fossil or sedimentary record.
-
-<!-- This could be a neat future manuscript (although time...) so not sure if we state this idea, or just cut the previous paragraph -->
+-->
 

### 2023-05-24 manuscript/main.qmd
discussion part 2 and proofreading checks

@@
 ---
-title: "Bayesian multi-proxy reconstruction of latitudinal temperature gradients: An early Eocene example"
+title: "Bayesian multi-proxy reconstruction of early Eocene latitudinal temperature gradients"
 author:
@@
 Our results confirm the presence of a flattened latitudinal temperature gradient and unusually high polar temperatures during the EECO, which is supported by high-latitude ecological data.
-Integrating ecological data and adequate prior information has the potential to substantially reduce uncertainty in palaeoclimate reconstructions, allowing for unbiased temperature estimates from sparse data.
+Integrating multiple types of proxy data and adequate prior information has the potential to substantially reduce uncertainty in palaeoclimate reconstructions, allowing for unbiased temperature estimates from sparse data.
 
@@
 
-Palaeoclimate, latitudinal temperature gradients, temperature proxies, Eocene, spatial bias
+Palaeoclimate, latitudinal temperature gradients, temperature proxies, Eocene, spatial bias, Bayesian
 
@@
 Recent advances in the synthesis of palaeoclimate data [e.g.
-@Veizer2015; @Hollis2019; @song2019; @Grossman2022; @Judd2022] are offering unprecedented insights into the complex and dynamic nature of the Earth's climate system, but a fundamental challenge remains: the proxy record of past climate is spatially incomplete and biased by imperfect preservation and uneven sampling [@judd2020; @Jones2022; @Judd2022].
-
-Geochemical proxies, such as oxygen isotopes (&delta;^18^O), are regularly used to provide quantitative palaeotemperature estimates for the geological past at a range of temporal and spatial scales [e.g.
-@Liu2009; @Wuchter2004].
-However, a fundamental challenge remains which must be addressed to exploit the full potential of these advances: the proxy record of past climate is spatially incomplete and biased by imperfect preservation and uneven sampling [@judd2020; @Jones2022; @Judd2022].
+@Veizer2015; @Hollis2019; @song2019; @Grossman2022; @Judd2022] are offering unprecedented insights into the complex and dynamic nature of the Earth's climate system, but a fundamental challenge remains: the proxy record of past climates is spatially incomplete and afflicted by imperfect preservation and uneven sampling [@judd2020; @Jones2022; @Judd2022].
 
@@
 At the most basic level, this can be driven by two factors: (1) missing data for some regions (e.g.
-no high-latitude data); or (2) overrepresentation of some regions (e.g.
-a high proportion of samples from tropical regions).
+no high-latitude data); or (2) overrepresentation of other regions (e.g.
+a high proportion of samples from tropical areas).
 The latter can be addressed through the down-sampling of data or restricting analyses to specific regions (e.g.
@@
 These approaches work well for interpolating relatively densely-sampled data, but the absence of constraints on the modeled parameters means that such models can produce unrealistic temperature estimates when extrapolating from sparse data.
-Statistical modeling in a Bayesian framework can help overcome this problem by requiring the explicit specification of priors for the model parameters, which can be used to express physical constraints [@Chandra2021].
+Statistical modeling in the Bayesian framework can help overcome this problem by requiring the explicit specification of priors for the model parameters, which can be used to express physical constraints [@Chandra2021].
 
-Spatial gaps in the record of past climate can also be addressed through the integration of additional data.
+Spatial gaps in the palaeoclimate record can also be addressed through the integration of additional data.
 For example, lithological and fossil data can be used to infer past climatic conditions based on analogous modern sediments [@Chandra2021], or based on the premise that the climatic requirements of ancient taxa, biological traits, or ecological communities were similar to those of their nearest modern relatives [@Peppe2011; @Royer2012; @Salonen2019].
@@
 Here, we present a novel Bayesian hierarchical model that combines quantitative proxies and ecological constraints into a fully quantitative model of the latitudinal gradient of sea surface temperatures, bridging spatial gaps in sparsely sampled climate data.
-This model expands upon existing, spatially explicit reconstructions of palaeoclimate by allowing for the integration of (1) prior information based on physical principles and on the observed, modern sea surface temperature distribution, and of (2) geochemical and ecological climate proxies in a common, quantitative framework.
-We use a generalised logistic function to infer the shape of the temperature gradient despite a patchy latitudinal coverage, and test the robustness of this method using downsampled, simulated temperature gradients.
+This model expands upon existing, spatially explicit palaeoclimate reconstructions  by allowing for the integration of (1) prior information based on physical principles and on the observed, modern sea surface temperature distribution, and of (2) geochemical and ecological climate proxies in a common, quantitative framework.
+We use a generalised logistic function to accurately infer the shape of the temperature gradient despite a patchy latitudinal coverage, and test the robustness of this method using downsampled, simulated temperature gradients.
 
@@
 We chose the EECO to demonstrate the application of the model due to its significance as the interval with the warmest sustained temperatures of the Cenozoic [@Pross2012], rendering it a potential analogue for extreme climate warming scenarios [@Burke2018].
-Our integrative approach allows us to shed new light on the long-standing dispute on the shape of the early Eocene temperature gradient between climate modellers and researchers working with the geological and fossil record [@Sloan1990; @Markwick1994; @Huber2011; @Tierney2017; @Inglis2020].
+Our integrative approach allows us to shed new light on the long-standing dispute on the steepness of the early Eocene temperature gradient [@Sloan1990; @Markwick1994; @Huber2011; @Tierney2017; @Inglis2020].
 
@@
 ![](../figures/Table1.png){width="16.25cm"}
-Table 1: Inferred latitudinal SST gradients for the early Eocene (EE) or the EECO, as shown in earlier, proxy-based studies.
+Table 1: Inferred latitudinal sea surface temperature (SST) gradients for the early Eocene (EE) or the EECO, as shown in earlier, proxy-based studies.
 For comparison, a gradient derived from a climate model ensemble is also shown.
@@
 n_geochem <- nrow(obsmat)
+n_geochem_location <- length(unique(obsmat$sample))
+
 ```
@@
 This compilation provides data on four different geochemical proxies for reconstructing seawater temperature: &delta;^18^O, &Delta;47, Mg/Ca and $TEX_{86}$.
-For our analyses, this dataset was restricted to the EECO and samples from the continental shelf.
+For our analyses, this dataset was restricted to the EECO (defined as 49.1 - 53.8 Ma) and samples from the continental shelf.
 Recrystallised &delta;^18^O samples were also excluded as secondary diagenetic calcite precipitated after deposition can bias isotope measurements and offset temperature values [@Schrag1999].
 This filtering resulted in most &delta;^18^O samples being excluded from the dataset (retaining 8 out of 152).
-After data filtering, `r n_geochem` geochemical proxy samples remained.
-For a detailed description of each proxy see [@Hollis2019].
+After data filtering, `r n_geochem` geochemical proxy samples from `r n_geochem_location` locations remained.
+For a detailed description of each proxy see @Hollis2019.
 
@@
 
-**Coral reefs.** Today, shallow warm-water coral reefs are limited to tropical and subtropical latitudes (\~34$^\circ$N--32$^\circ$S), with minimum sea surface temperature (\~18$^\circ$C) tolerances being the primary constraint on this distribution [@johannes1983; @kleypas1999; @yamano2001].
-As coral reefs reside at the upper thermal limit of the oceans today, their maximum sea surface temperature tolerance is less well-constrained with some studies suggesting up to 35.6$^\circ$C in the geological past [@jones2022reef].
+**Coral reefs.** Today, shallow warm-water coral reefs are limited to tropical and subtropical latitudes (\~34$^\circ$N--32$^\circ$S), with minimum sea surface temperature tolerances (\~18$^\circ$C) being the primary constraint on this distribution [@johannes1983; @kleypas1999; @yamano2001].
+As coral reefs reside at the upper thermal limit of the oceans today, their maximum sea surface temperature tolerance is less well-constrained, with some studies suggesting up to 35.6$^\circ$C in the geological past [@jones2022reef].
 Nevertheless, coral reefs have frequently been recognised as tracers of past (sub-)tropical conditions [@ziegler1984; @kiessling2001].
 During the Eocene, coral communities and reefs expanded across tropical and temperate latitudes, with some communities found up to palaeolatitudes of 43 $^\circ$N [@Zamagni2012].
-Using a compilation of Paleocene--early Eocene coral reefs and community localities [@Zamagni2012], we generate quantitative sea surface temperature estimates for the ECCO.
-To do so, we extract localities from the compilation that are inferred to be Ilerdian (early Eocene) coral reefs, and that can be confidently assigned to the EECO.
-We exclude coral knobs and coral-bearing mounds which might have broader climatic limits than coral reef ecosystems.
+Using a compilation of Paleocene--early Eocene coral reefs and community localities [@Zamagni2012], we generated quantitative sea surface temperature estimates for the ECCO.
+To do so, we extracted localities from the compilation that are inferred to be Ilerdian (early Eocene) coral reefs, and that could be confidently assigned to the EECO.
+We excluded coral knobs and coral-bearing mounds which might have broader climatic limits than coral reef ecosystems.
 This filtering resulted in four unique coral reef localities remaining for the EECO, all of which conform to the modern latitudinal range of coral reefs (\<34$^\circ$ N).
-Subsequently, we use statistically derived temperature limits (minimum = 21$^\circ$C, average = 27.6$^\circ$C, maximum = 29.5$^\circ$C) from the published literature [@kleypas1999] to define a normal probability distribution of potential temperature values for coral reef localities.
+Subsequently, we used statistically derived temperature limits (minimum = 21$^\circ$C, average = 27.6$^\circ$C, maximum = 29.5$^\circ$C) from the published literature [@kleypas1999] to define a normal probability distribution of potential temperature values for coral reef localities.
 This normal probability distribution was defined with a mean of 27.6 and a standard deviation of 2.125, placing 97.5% of the probability density above the minimum.
@@
 **Mangroves.** Mangroves are distributed throughout the tropics and subtropics today.
-While factors beside SST influence the distribution of mangroves, empirical, lower temperature limits have been established for the genera *Avicennia* (15.6$^\circ$C) and *Rhizophora* (20.7$^\circ$C) [@Quisthoudt2012].
+While factors beside sea surface temperatures (SST) influence the distribution of mangroves, empirical, lower temperature limits have been established for the genera *Avicennia* (15.6$^\circ$C) and *Rhizophora* (20.7$^\circ$C) [@Quisthoudt2012].
 Both *Avicennia* and members of the Rhizophoraceae family were widespread in the early Eocene, but only *Avicennia* occurred at polar latitudes [@Suan2017; @Popescu2021].
-Assuming that Eocene members of these mangrove taxa conform to the same climatic requirements as their modern relatives, the presence and absence of *Avicennia* and Rhizophoraceae pollen can be used as a palaeotemperature indicator.
+Assuming that Eocene members of these mangrove taxa conform to similar climatic requirements as their modern relatives, the presence and absence of *Avicennia* and Rhizophoraceae pollen can be used as a palaeotemperature indicator.
 For this analysis, published mangrove occurrence data were taken from @Popescu2021, and converted to quantitative temperature estimates.
@@
 However, a value of 22.5$^\circ$C is ascribed as the upper temperature limit here as *Rhizophora* is rare below this temperature.
-We define the *Avicennia*-only temperature distribution as a normal distribution with a mean of 19.05$^\circ$C and a standard deviation of 1.725, resulting in 95% of the probability density being placed within the temperature limits.
+We define the *Avicennia*-only temperature distribution as a normal distribution with a mean of 19.05 and a standard deviation of 1.725, resulting in 95% of the probability density being placed within the temperature limits.
 
 2)  *Avicennia* and Rhizophoraceae assemblages ($n=5$): the presence of both groups suggests that the locality should have a minimum temperature of 20.7$^\circ$C (lower temperature limit of *Rhizophora*).
-As the upper thermal limits of *Aviciennia* and *Rhizophora* are not well established in @Quisthoudt2012, we assign the same maximum temperature limits (29.5$^\circ$C) as coral reef localities as mangroves are also widely distributed throughout tropical regions.
+As the upper thermal limits of *Aviciennia* and *Rhizophora* are not well established in @Quisthoudt2012, we assign the same maximum temperature limits (29.5$^\circ$C) as coral reef localities, because mangroves are also widely distributed throughout tropical regions.
 Consequently, we define the temperature distribution for this locality as a normal distribution with a mean of 25.1 and a standard deviation of 2.2, with 95% probability density within the temperature limits.
@@
 
-Palaeogeographic distribution of the geochemical and ecological data compilation used in this study.
-Palaeocoordinates of samples were estimated via the palaeoverse R package ver.
-1.2.0 [@jones2023].
+Palaeogeographic distribution of the geochemical and ecological data compilation used in this study. <!--  I cut repetition from the methods, but maybe cite where the map is coming from. 
+Palaeocoordinates of samples were estimated via the palaeoverse R package [version 1.2.0,@jones2023].
 To do so, the @merdith2021 Global Plate Model was used to palaeorotate present-day coordinates to 51.2 Ma.
-Map is presented in the Robinson projection (ESRI:54030).
+--> Map is presented in the Robinson projection (ESRI:54030).
 :::
@@
 
-The palaeogeographic distribution of geochemical and ecological data was reconstructed using the @merdith2021 plate rotation model via the palaeoverse R package ver.
-1.2.0 (@jones2023).
+The palaeogeographic distribution of geochemical and ecological data was reconstructed using the @merdith2021 plate rotation model via the palaeoverse R package [version 1.2.0, @jones2023].
 The midpoint age of the EECO (51.2 Ma), along with the present-day coordinates of geochemical and ecological data, were used for palaeogeographic reconstruction.
@@
 
-**Model structure (Fig 2).** We model the mean temperature ($\mu$) at location $j$ as a function of absolute latitude ($abs(l)$) with a logistic regression (also known as growth curve or Richard's curve) of the form:
+**Model structure (Fig 2).** We model the mean temperature ($\mu$) at location $j$ as a function of absolute latitude ($abs(l)$) with a logistic regression (also known as "growth curve" or "Richard's curve") of the form:
 
@@
 
-This structure implies that $\mu_j$ is not fixed at the mean proxy temperature at location $j$, <!--of the temperatures $t_{i=1,...,m,~j}$ of location $j$-->, but is drawn towards the overall logistic regression curve, i.e.
+This structure implies that $\mu_j$ is not fixed at the mean proxy temperature at location $j$, <!--of the temperatures $t_{i=1,...,m,~j}$ of location $j$--> but is drawn towards the overall logistic regression curve, i.e.
 towards $\nu_j$.
@@
 
-To test whether our logistic regression model can adequately describe different latitudinal temperature gradients at various sample sizes, we generated four idealised gradients that emulate potential climatic states throughout Earth's geological history: extreme icehouse, icehouse, greenhouse and extreme greenhouse (REF).
+To test whether our logistic regression model can adequately describe different latitudinal temperature gradients at various sample sizes, we generated four idealised gradients that emulate potential climatic states throughout Earth's geological history: extreme icehouse, icehouse, greenhouse and extreme greenhouse [@Frakes1992].
 We then randomly sampled (1000 iterations) these gradients using increasing sample sizes (5, 10, and 20) and reconstructed the latitudinal temperature gradient using our model for each of these sample sizes and gradient types.
-Using the same idealised gradients, we also tested whether our model could accurately reconstruct latitudinal temperature gradients using the palaeogeographic distribution of Eocene samples (n = 34), providing an empirical example distribution that captures both limited sample size and skewed geographic distributions.
-To assess how well the modeled gradients reconstructed from limited sampling fit the idealised gradients, we calculated an R^2 for Bayesian regression models [@Gelman2019].
-For every iteration from the posterior, we intercepted the modeled and the idealised gradient in intervals of 1$^\circ$ latitude, and calculated the R^2 based on these values.
-We report the median, and 2.5th and 97.5th percentile of the resulting R^2 measures.
+Using the same idealised gradients, we also tested whether our model could accurately reconstruct latitudinal temperature gradients using the palaeogeographic distribution of Eocene samples (*n* = 34), providing an empirical, examplary distribution that captures both limited sample size and skewed geographic origins of samples.
+To assess how well the modeled gradients, reconstructed from limited sampling, fit the idealised gradients, we calculated an $R^2$ for Bayesian regression models [@Gelman2019].
+For every iteration from the posterior, we intercepted the modeled and the idealised gradient in intervals of 1$^\circ$ latitude, and calculated the $R^2$ based on these values.
+We report the median, and $2.5^{th}$ and $97.5^{th}$ percentile of the resulting $R^2$ measures.
 
-To test whether out model can accurately model the the modern sea surface temperature gradient, and to facilitate comparison with the Eocene gradient, we applied our model to annual sea surface mean temperatures from Bio-Oracle [@Assis2018], down-sampled to a $1^\circ \times 1 ^\circ$ raster (n = 46,131).
-The R^2 for the modern gradient was calculated as described above, comparing the modeled gradient and the empirical temperature averages in 1$^\circ$ latitude bins.
+To test whether our model can accurately depict the shape of the modern sea surface temperature gradient, and to facilitate comparison with the Eocene gradient, we applied our model to annual sea surface mean temperatures from Bio-Oracle [@Assis2018], down-sampled to a $1^\circ \times 1 ^\circ$ raster (*n* = 46,131).
+The $R^2$ for the modern gradient was calculated as above [@Gelman2019], comparing the modeled gradient and the empirical temperature averages in 1$^\circ$ latitude bins.
 Only the medians are reported for the modern gradient, as the 95% credible intervals are extremely narrow due to the high precision of the posterior estimates.
@@
 To reconstruct the idealised gradients and the modern gradient, we used a simplified, non-hierarchical version of our model, as every location is associated with only one temperature value, making the hierarchical structure superfluous.
-We thus substituted temperature ($t_j$) for $\mu_j$ in [@eq-mu] and [@eq-sigma].
+To achieve this, we substituted temperature ($t_j$) for $\mu_j$ in [@eq-mu] and [@eq-sigma].
 
@@
 Modeled sea surface temperature estimates were generated with [@eq-nu], calculating the sea surface temperatures at any latitude with the parameter estimates of each iteration from the posterior.
-The median, $2.5^{th}$ and $97.5^{th}$ percentiles of temperatures where then taken from all temperature estimates obtained at the latitude of interest.
+The median, $2.5^{th}$ and $97.5^{th}$ percentiles of temperatures where then taken from all temperature estimates obtained at the latitudes of interest.
 
 The latitudinal gradient is calculated as the difference between the modeled temperature at the equator (0$^\circ$ latitude) and at the poles (90$^\circ$ absolute latitude).
+To facilitate comparison with earlier estimates, we also calculate the gradient with the temperature at the polar circle (66.6$^\circ$ absolute latitude) being used instead of at the poles.
 Given the sigmoidal shape of the modern as well as the Eocene gradient (see Fig.
-4), this results in only slightly higher estimates of the gradient than when comparing e.g.
-the zonal average of equatorial and high-latitude temperatures, as is done in some earlier studies [e.g.
-@Evans2018].
+4), these results are broadly comparable to a gradient inferred from the zonal average of equatorial and high-latitude temperatures, as is done in some earlier studies [@Evans2018].
 
@@
 
-Global average temperatures were calculated by taking the weighted mean of the median ($2.5^{th}$, $97.5^{th}$ percentile) temperature estimates in 1$^\circ$ latitudinal bins.
+Global average temperatures with 95% credible intervals were calculated by taking the weighted mean of the median ($2.5^{th}$, $97.5^{th}$ percentile) temperature estimates in 1$^\circ$ latitudinal bins.
 The weights were set to the proportion of global surface area in each latitudinal bin, i.e.
@@
 Each column depicts a different reconstruction for given sample sizes: 5, 10, 20, and 34 (latitudes of EECO samples).
-Each row depicts a different simulated latitudinal temperature gradient that represents idealised climatic states: extreme greenhouse, greenhouse, icehouse, extreme icehouse.
+Each row depicts a different simulated latitudinal temperature gradient that represents idealised climatic states: extreme icehouse, icehouse, greenhouse, extreme greenhouse.
 The black line illustrates the simulated gradient.
-The blue line depicts the reconstructed gradient represented by the median sea surface temperature value estimated from 1000 model runs with different random samples (first three columns), and a single run with the EECO latitudinal sampling distribution (fourth column).
+The blue line depicts the reconstructed gradient represented by the median sea surface temperature value estimated from 1,000 model runs with different random samples (first three columns), and a single run with the EECO latitudinal sampling distribution (fourth column).
 The blue shadings depict the 90%, 95%, and 99% credible intervals.
-Bold black text within each panel depicts the coefficient of determination (*R*^2) for estimating goodness of fit between the simulated and modeled gradient.
-The median (50%) *R*^2^ value along with the 2.5% and 97.5% confidence intervals from the 1000 model runs are shown.
+Bold black text within each panel depicts the coefficient of determination ($R^2$) for estimating goodness of fit between the simulated and modeled gradient.
+The median (50%) $R^2$ value along with the 2.5% and 97.5% confidence intervals from the 1,000 model runs are shown.
 Each gradient is depicted in absolute latitude.
@@
 
-Estimate of the present-day latitudinal temperature gradient using the Bayesian hierarchical model.
-The present-day empirical (black line) latitudinal temperature gradient (mean sea surface temperature) and that estimated by the Bayesian hierarchical model (green line) when using all cell values from the Bio-ORACLE grid of mean sea surface temperatures.
-Grey points depict the individual cell values of the Bio-ORACLE grid.
+Present-day latitudinal temperature gradient.
+The present-day empirical latitudinal temperature gradient (mean sea surface temperature) is depicted as a black line, and the gradient estimated by the Bayesian model is shown in turquoise.
+Grey points depict the individual cell values of the Bio-ORACLE grid of mean sea surface temperatures, which were used to infer the empirical and the modeled gradient.
 :::
@@
 2).
-Random latitudinal sampling results in highly accurate reconstructions already at a sample size of 10 for the icehouse scenarios (95 % CI of R^2 > 0.9).
+Random latitudinal sampling results in highly accurate reconstructions already at a sample size of 10 for the icehouse scenarios (95 % CI of $R^2$ > 0.9).
 Greenhouse scenarios require more random samples to accurately predict high-latitude temperatures.
 This is because in the absence of high-latitude samples, the modeled gradient is heavily influenced by the priors, which we based on the modern, the only empirically known latitudinal temperature gradient.
-A sampling distribution resembling that of the early Eocene data set used in this study allows for a highly accurate reconstruction of even the extreme greenhouse scenario (95 % CI of R^2 > 0.95).
+A sampling distribution resembling that of the early Eocene data set used in this study allows for a highly accurate reconstruction of even the extreme greenhouse scenario (95 % CI of $R^2$ > 0.95).
 
@@
 The early Eocene GMsST is estimated at `r eotemp_global$"50%"` (`r eotemp_global$"2.5%"` - `r eotemp_global$"97.5%"`)$^\circ$C, `r  round(as.numeric(eotemp_global$"50%") - as.numeric(motemp_global$"50%"),1)`$^\circ$C higher than the modern.
-A model run excluding the ecological proxies increases the GMsST by `r eoceneproxy_global_diff$"50%"` (`r eoceneproxy_global_diff$"2.5%"` - `r eoceneproxy_global_diff$"97.5%"`)$^\circ$C.
-The median latitudinal gradient is similar when excluding the ecological proxies, with a median of `r eocene_proxy_gradient$"50%"`$^\circ$C, but with a `r proxy_uncertain_diff_percent`% wider 95% CI (Supplementary materials).
+A model run excluding the ecological proxies increases the GMST by `r eoceneproxy_global_diff$"50%"` (`r eoceneproxy_global_diff$"2.5%"` - `r eoceneproxy_global_diff$"97.5%"`)$^\circ$C.
+The median latitudinal gradient is similar when excluding the ecological proxies, with a median of `r eocene_proxy_gradient$"50%"`$^\circ$C, but with a `r proxy_uncertain_diff_percent`% wider 95% CI (Fig.
+S2).
 This indicates that the ecological proxy data are broadly in agreement with the geochemical proxies, while providing additional constraints on the shape of the early Eocene temperature gradient.
@@
 Applying the model separately to each hemispheres results in substantial differences in hemispherical, average temperatures, with the southern hemisphere being warmer by `r eeco_south_north_global_diff$"50%"` (`r eeco_south_north_global_diff$"2.5%"` - `r eeco_south_north_global_diff$"97.5%"`)$^\circ$C.
-The inferred latitudinal gradient is somewhat steeper in the northern hemisphere (`r eocene_north_south_gradient_diff$"50%"`$^\circ$C, although the 95% CI spans `r eocene_north_south_gradient_diff$"2.5%"` to `r eocene_north_south_gradient_diff$"97.5%"`$^\circ$C), but the large uncertainties associated with both gradients, and the lack of polar proxy data in the southern hemisphere preclude a more precise statement (Fig.
+The inferred latitudinal gradient is somewhat steeper in the northern hemisphere (steeper by `r eocene_north_south_gradient_diff$"50%"`$^\circ$C, although the 95% CI spans `r eocene_north_south_gradient_diff$"2.5%"` to `r eocene_north_south_gradient_diff$"97.5%"`$^\circ$C), but the large uncertainties associated with both gradients, and the lack of polar proxy data in the southern hemisphere preclude a more precise statement (see Fig.
 S3).
@@
 
-Estimates of the early Eocene climatic optimum and the present-day latitudinal temperature gradient using the Bayesian model.
-The estimated present-day (turquoise line) and early Eocene climatic optimum (purple line) latitudinal temperature gradient.
-The respective ribbons (green and purple shading) depict the 95% credible intervals of the estimated gradients.
-Points within the plot depict the palaeolatitudinal distribution of geochemical (e.g.
+Estimates of the median, latitudinal SST gradients of the early Eocene climatic optimum (purple line) and of the present-day (turquoise), both estimated with the Bayesian model.
+The ribbons (purple and turquoise shadings) depict the 95% credible intervals of these gradients, respectively.
+Points within the plot depict the geochemical (e.g.
 &delta;^18^O) and ecological (e.g.
@@
 
-In contrast, our Bayesian, parametric model, allows for the inclusion of informative priors on the model parameters.
-The modeled sea surface temperature gradient thus does not strictly follow the proxy data, but instead represents a compromise between the data and our prior knowledge.
+In contrast, our Bayesian, parametric model allows for the inclusion of informative priors on the model parameters.
+The modeled sea surface temperature gradient thus does not strictly follow the proxy data, but instead represents a compromise between the data and prior knowledge.
 In the EECO example (Fig.
@@
 Palaeoclimate reconstructions are often summarised as global mean surface temperatures (GMST), providing a standardised metric for characterising the state of the Earth's climate [@Royer2004; @Inglis2020].
-However, the calculation of global mean surface temperatures from sparse proxy data is susceptible to bias [@Jones2022].
+The calculation of global mean surface temperatures directly from sparse proxy data is susceptible to bias [@Jones2022].
 By modeling the temperature variation across latitudes, a complete temperature distribution along a latitudinal axis can be obtained, filling in gaps in the proxy record through inter- or extrapolation.
@@
 Instead, our method allows for intersecting the modeled temperature gradient at narrow latitudinal intervals, even when significant latitudinal gaps exist.
-Weighting the temperatures of those latitudinal intervals by area then provides unbiased GMST estimates.
+Weighting the temperatures of those latitudinal intervals by area results in GMST estimates without intrinsic spatial biases.
 We anticipate that this improved method may significantly alter Phanerozoic, proxy-based temperature curves, which have often been directly calculated from the proxy record [@Royer2004; @Veizer2015].
@@
 
-Our results further exemplify how incorporating quantified ecological temperature constraints can provide more precise temperature reconstructions than geochemical proxies alone, adding to the advances in palaeoclimate reconstructions achieved by integrating lithological data [@Scotese2021].
+Our results further exemplify how incorporating quantified ecological temperature constraints can provide more precise temperature reconstructions than geochemical proxies alone, adding to the advances in palaeoclimate reconstructions achieved by integrating lithological data [@Scotese2021; @Burgener2023].
 Combining the occurrences of climate-sensitive plant communities [@Greenwood1995], reptiles [@Markwick2007], leaf shapes [@Peppe2011], with geochemical proxies offers substantial potential for improving quantitative palaeoclimate reconstructions across the Phanerozoic.
@@
 A local temperature estimate, based on multiple geochemical proxies, thus has the same weight as a local temperature estimate obtained from the occurrence of a climate-sensitive plant community, whilst preserving the uncertainty associated with each estimate.
-The model could easily be extended to include uncertainties on individual geochemical proxy data, or to differently weight different proxy records classified as more or less reliable.
+The model could easily be extended to include uncertainties on individual geochemical proxy data, or to differently weight proxy records classified as more or less reliable.
 
@@
 The fossil record indicates that reef development may have been stunted in the early Eocene, with few early Eocene coral reefs occurring in low latitudes [@Zamagni2012].
-The absence of coral reefs in higher latitudes in the early Eocene could be due to requirements in irrdiance, rather than temperature [@Muir2015].
+The absence of coral reefs in higher latitudes in the early Eocene could be due to requirements in irradiance, rather than temperature [@Muir2015].
 Tropical temperatures predicted by the geochemical proxy record indicate hotter-than-modern tropical temperatures for the early Eocene (Fig.
@@
 The geochemical proxy record and ecological data indicate that the latitudinal SST gradient of the early Eocene climatic optimum was significantly shallower than the modern [@Huber2011], but beyond that, there is little agreement.
-@Evans2018 estimate the SST gradient at 32+-10% of the modern, which would correspond to \~`r round( as.numeric(modern_gradient$"50%") - 0.32*as.numeric(modern_gradient$"50%"),0)`+- `r format(round( 0.10*as.numeric(modern_gradient$"50%"),0),nsmall=0)`$^\circ$C, using the difference between the mean tropical and deep ocean data.
+Earlier, reconstructed early Eocene and EECO SST gradients range from $7 - 21^\circ$C (Table 1); a more recent reconstruction that includes terrestrial air and sea surface temperatures arrives at a gradient of $13^\circ$C [@Inglis2020].
+Our polar circle to equatorial gradient estimate is lower than most previous estimates at $7.8^\circ$, although the 95% credible interval extends up to $13.7^\circ$ and thus overlaps earlier estimates based on shallow water proxies.
+The confirmation of a very flat gradient by both geochemical and ecological shallow water data indicates that inferred SST gradients based on tropical, shallow water and deep water samples [@Cramwinckel2018; @Evans2018] may overestimate the SST gradient of the early Eocene greenhouse world.
+
+<!-- @Evans2018 estimate the SST gradient at 32+-10% of the modern, which would correspond to \~`r round( as.numeric(modern_gradient$"50%") - 0.32*as.numeric(modern_gradient$"50%"),0)`+- `r format(round( 0.10*as.numeric(modern_gradient$"50%"),0),nsmall=0)`$^\circ$C, using the difference between the mean tropical and deep ocean data.
 The modeled SST of @Tierney2017 using TEX86 is \~ `r round(35.8-23.8,0)`$^\circ$C, when taking the difference between temperatures at the polar circles and the equator.
@@
 Our polar circle to equatorial gradient is lower than most previous estimates at `r eocene_gradient_polar_circle$"50%"`$^\circ$C, although the 95% credible interval overlap with several of the listed gradient estimates.
+-->
 
-\[Table 1: Early Eocene latitudinal temperature gradients from the literature\]
-
-In latitudes beyond the polar circle, the discrepancy between our model predictions and those of earlier, proxy-based models increases, as those predict almost linearly decreasing SSTs towards the poles, whereas our median model prediction suggests only a slight decrease beyond the polar circle.
-However, the scarcity of temperature records in this range leads to widening credible intervals, including the possibility of stronger temperature decreases.
+Discrepancies between earlier, proxy-based reconstructions and our modeling results are most pronounced in latitudes beyond the polar circle, as earlier approaches [e.g.
+@Tierney2017] predict almost linearly decreasing SSTs towards the poles, whereas our median prediction suggests only a slight decrease beyond the polar circle.
+The scarcity of temperature records in this range leads to widening credible intervals in our prediction, including the possibility of stronger temperature decreases.
 Polar temperature estimates from our model are thus conservative in that they admit large uncertainty where data is absent, which is desirable.
 However, the presence of high proxy-derived temperature estimates at \~ 60$^\circ$ latitudes forces the modeled median temperature curve to be too high at \~ 24$^\circ$C, relative to the temperatures indicated by the high-latitude mangrove communities (15.6 - 22.5$^\circ$C).
-In contrast, the extrapolated polar temperatures of most previous proxy-based models are likely too low, given the abundance of ecological data indicating temperate or subtropical high-latitude climates during the EECO [@Pross2012; @Popescu2021]
+In contrast, the extrapolated polar temperatures of most previous proxy-based models are likely too low, given the abundance of ecological data indicating temperate or subtropical high-latitude climates during the EECO [@Pross2012; @Popescu2021].
 
-The very high variability of the proxy record in mid-latitudes results in large uncertainties on the shape of temperature gradient and on the GMsST.
+The very high variability of the proxy record in mid-latitudes results in large uncertainties on the shape of temperature gradient and on the GMST.
 Biases and errors in the proxy reconstructions likely contribute to the observed variability, as geochemical proxies reflect many other factors besides seawater temperature [@Hollis2019].
-Despite excluding &delta;^18^O measurements from recrystallised fossils, systematic offsets remain between mostly warm temperatures derived from TEX86, and cooler temperatures derived from &delta;^18^O, &Delta;~47~, and the ecological proxies.
-Seasonality [@Keating-Bitonti2011 and temporal changes within the EECO [@Westerhold2018] may also contribute to the large variability of the EECO proxy data.
+Despite excluding &delta;^18^O measurements from recrystallised fossils, systematic offsets remain between mostly warm temperatures derived from $TEX_{86}$, and cooler temperatures derived from &delta;^18^O, &Delta;~47~, and the ecological proxies.
+Seasonality [@Keating-Bitonti2011] and temporal changes within the EECO [@Westerhold2018] may also contribute to the large variability of the EECO proxy data.
 
-Recent GMsST estimates of the EECO and of the early Eocene range from 23.4 to 37.1$^\circ$C, with the lowest GMSTs being derived from d18O, and the higher estimates including TEX86 [@Inglis2020].
-Many studies include both marine and terrestrial proxies to derive GMST estimates, but despite great differences in proxy selection and in the calculation of global average temperatures, many recent estimates fall in the range of 27 - 29.5$^\circ$C [@Hansen2013; @Caballero2013; @Cramwinckel2018; @Zhu2019], similar to our GMsST estimate of `r eotemp_global$"50%"`$^\circ$C.
+Recent, marine GMST estimates of the EECO and of the early Eocene range from 23.4 to 37.1$^\circ$C, with the lowest GMSTs being derived from &delta;^18^O, and the higher estimates including $TEX_{86}$ [@Inglis2020].
+Many studies include both marine and terrestrial proxies to derive GMST estimates, but despite great differences in proxy selection and in the calculation of global average temperatures, many recent estimates fall in the range of 27 - 29.5$^\circ$C [@Hansen2013; @Caballero2013; @Cramwinckel2018; @Zhu2019], similar to our median GMST estimate of `r eotemp_global$"50%"`$^\circ$C.
 
@@
 By providing temperature estimates across the entire latitudinal range, this method also facilitates the reconstruction of unbiased global average temperatures.
-Application of our model to the EECO confirms the existence of latitudinal temperature gradients that were flatter than predicted by most Earth system models and by the majority of proxy-based work.
+Application of our model to the EECO suggests that latitudinal sea surface temperature gradients were shallower than estimated by most earlier, proxy-based studies.
 High-latitude pollen records support this interpretation.
-Our GMsST estimate is in good agreement with most existing estimates, indicating that broadly accurate GMsST reconstructions are possible even with substantial deviations in the shape of the latitudinal temperature gradient.
+Our GMST estimate is in good agreement with most existing estimates, indicating that broadly accurate GMST reconstructions are possible even with substantial deviations in the shape of the latitudinal temperature gradient.
 Our new method opens the door for improving the accuracy of proxy-based palaeoclimate reoncstructions and Phanerozoic temperature curves, particularly in intervals with a patchy record, and mitigates the biases incurred from uneven sampling.
+The flexibility of this approach means that our estimates can be quickly and easily updated when new data is made available.
 
@@
 The authors are grateful to all those who have enabled this work by collecting, measuring, collating and screening geochemical and fossil data.
-The contribution of K.E.
-was supported by...
 The contribution of L.A.J.
 was supported by a Juan de la Cierva-formación 2021 fellowship (FJC2021-046695-I/MCIN/AEI/10.13039/501100011033) from the European Union "NextGenerationEU"/PRTR.
+<!-- regarding funding acknowledgement, I talked to my boss and he says I don't need to credit my funding as I have done this work largely in my spare time -->
 

### 2023-05-24 manuscript/supplementary.qmd
discussion part 2 and proofreading checks

@@
 ---
-title: "Supplementary Material: Bayesian multi-proxy reconstruction of latitudinal temperature gradients applied to the Early Eocene"
+title: "Supplementary Material: Bayesian multi-proxy reconstruction of early Eocene latitudinal temperature gradients"
 author:

### 2023-05-30 manuscript/main.qmd
Lewis comments addressed until end of results

@@
 Accurately reconstructing large-scale palaeoclimate patterns from sparse local records is critical for understanding the evolution of Earth's climate.
-Challenges arise particularly from the patchiness and uneven sampling of palaeoclimate data, and from the disparate nature of the proxy data.
-Geochemical data usually provide temperature estimates via transfer functions informed by experiments, and transfer functions based on the climatic requirements of modern taxa exist for some fossil plant assemblages.
-In contrast, most ecological and lithological data, e.g.
-coral reefs and evaporites, only convey information on broad climatic requirements.
-Historically, most large-scale proxy-based reconstructions have been based on either geochemical or ecological data, but few approaches exist for combining multiple proxy types into a single, quantitative reconstruction.
-Large spatial gaps in existing proxy records have often been bridged by simple averaging, without taking into consideration the spatial distribution of the samples, leading to biased temperature reconstructions.
+Particular challenges arise from the patchiness, uneven spatial distribution, and disparate nature of palaeoclimatic proxy records.
+Geochemical data typically provide temperature estimates via transfer functions derived from experiments.
+Similarly, transfer functions based on the climatic requirements of modern taxa exist for some fossil groups, such as pollen assemblages.
+In contrast, most ecological and lithological data (e.g.
+coral reefs and evaporites) only convey information on broad climatic requirements.
+Historically, most large-scale proxy-based reconstructions have used either geochemical or ecological data, but few studies have combined multiple proxy types into a single quantitative reconstruction.
+Large spatial gaps in existing proxy records have often been bridged by simple averaging, without taking into account the spatial distribution of samples, leading to biased temperature reconstructions.
 Here, we present a Bayesian hierarchical model to integrate ecological data with established geochemical proxies into a unified quantitative framework, bridging gaps in the latitudinal coverage of proxy data.
@@
 Assuming the conservation of thermal tolerances of modern coral reefs and mangrove taxa, we establish broad sea surface temperature ranges for EECO coral reef and mangrove sites.
-We integrate these temperature estimates with the EECO geochemical shallow marine proxy record to model the latitudinal sea surface temperature gradient and global average temperatures.
+We integrate these temperature estimates with the EECO geochemical shallow marine proxy record to model the latitudinal sea surface temperature gradient and global average temperatures of the EECO.
 Our results confirm the presence of a flattened latitudinal temperature gradient and unusually high polar temperatures during the EECO, which is supported by high-latitude ecological data.
-Integrating multiple types of proxy data and adequate prior information has the potential to substantially reduce uncertainty in palaeoclimate reconstructions, allowing for unbiased temperature estimates from sparse data.
+We show that integrating multiple types of proxy data, and adequate prior information, has the potential to substantially reduce uncertainty in palaeoclimate reconstructions, allowing for unbiased temperature estimates from sparse data.
 
@@
 Recent advances in the synthesis of palaeoclimate data [e.g.
-@Veizer2015; @Hollis2019; @song2019; @Grossman2022; @Judd2022] are offering unprecedented insights into the complex and dynamic nature of the Earth's climate system, but a fundamental challenge remains: the proxy record of past climates is spatially incomplete and afflicted by imperfect preservation and uneven sampling [@judd2020; @Jones2022; @Judd2022].
+@Veizer2015; @Hollis2019; @song2019; @Grossman2022; @Judd2022] are offering unprecedented insights into the complex and dynamic nature of the Earth's climate system, yet a fundamental challenge remains: the proxy record of past climates is spatially incomplete and afflicted by imperfect preservation and uneven sampling [@judd2020; @Jones2022; @Judd2022].
 
@@
 latitudinal temperature gradients) and global temperatures   [@judd2020; @Jones2022].
-At the most basic level, this can be driven by two factors: (1) missing data for some regions (e.g.
+Principally, this can be driven by two factors: (1) missing data for some regions (e.g.
 no high-latitude data); or (2) overrepresentation of other regions (e.g.
 a high proportion of samples from tropical areas).
-The latter can be addressed through the down-sampling of data or restricting analyses to specific regions (e.g.
-@song2019).
+The latter can be addressed through the down-sampling of data or restricting analyses to specific regions [e.g.
+@song2019].
 However, in order to robustly infer regional or global-scale patterns from an incomplete record, spatial gaps must ultimately be bridged.
@@
 Interpolation is also sometimes used to bridge spatial gaps in palaeoclimate data [@Taylor2004], taking advantage of the autoregressive nature of climatic data: much of the information on the climate of any given location is contained in the climate data of nearby locations [@Reynolds1994].
-Adding to this, some proxy-based reconstructions use statistical modeling to infer palaeoclimatic patterns.
+Adding to this, some proxy-based reconstructions use statistical modelling to infer palaeoclimatic patterns.
 For example, polynomial regression [@Bijl2009] and cosine functions [@Inglis2020] have been used to reconstruct latitudinal temperature gradients, and 2D-reconstructions of surface temperatures have been created with Gaussian process regression [@Inglis2020].
-These approaches work well for interpolating relatively densely-sampled data, but the absence of constraints on the modeled parameters means that such models can produce unrealistic temperature estimates when extrapolating from sparse data.
-Statistical modeling in the Bayesian framework can help overcome this problem by requiring the explicit specification of priors for the model parameters, which can be used to express physical constraints [@Chandra2021].
+These approaches work well for interpolating relatively densely-sampled data, but the absence of constraints on the modelled parameters means that such models can produce unrealistic temperature estimates when extrapolating from sparse data.
+Statistical modelling in a Bayesian framework can help overcome this problem by requiring the explicit specification of priors for the model parameters, which can be used to express physical constraints [@Chandra2021].
 
@@
 Here, we present a novel Bayesian hierarchical model that combines quantitative proxies and ecological constraints into a fully quantitative model of the latitudinal gradient of sea surface temperatures, bridging spatial gaps in sparsely sampled climate data.
-This model expands upon existing, spatially explicit palaeoclimate reconstructions  by allowing for the integration of (1) prior information based on physical principles and on the observed, modern sea surface temperature distribution, and of (2) geochemical and ecological climate proxies in a common, quantitative framework.
-We use a generalised logistic function to accurately infer the shape of the temperature gradient despite a patchy latitudinal coverage, and test the robustness of this method using downsampled, simulated temperature gradients.
+This model expands upon existing, spatially explicit palaeoclimatic reconstructions  by allowing for the integration of (1) prior information based on physical principles and the observed modern sea surface temperature distribution, and of (2) geochemical and ecological climate proxies in a common, quantitative framework.
+We use a generalised logistic function to accurately infer the shape of the temperature gradient despite a patchy latitudinal coverage, and test the robustness of this method using down-sampled, simulated temperature gradients.
 
-We apply this model to the record of the early Eocene climatic optimum (EECO), combining a compilation of geochemical proxies [@Hollis2019], mangrove communities [@Popescu2021], and coral reefs [@Zamagni2012], using a nearest-living-relative approach [see e.g.
+We apply this model to the record of the early Eocene climatic optimum (EECO), combining a compilation of geochemical proxies [@Hollis2019], mangrove communities [@Popescu2021], and coral reefs [@Zamagni2012], using a nearest-living-relative approach [e.g.
 @Greenwood2017] to establish broad temperature ranges for the ecological data.
-We chose the EECO to demonstrate the application of the model due to its significance as the interval with the warmest sustained temperatures of the Cenozoic [@Pross2012], rendering it a potential analogue for extreme climate warming scenarios [@Burke2018].
-Our integrative approach allows us to shed new light on the long-standing dispute on the steepness of the early Eocene temperature gradient [@Sloan1990; @Markwick1994; @Huber2011; @Tierney2017; @Inglis2020].
+We choose the EECO to demonstrate the application of the model due to its significance as the interval with the warmest sustained temperatures of the Cenozoic [@Pross2012], rendering it a potential analogue for extreme climate warming scenarios [@Burke2018].
+Our integrative approach allows us to shed new light on the long-standing dispute on the steepness of the early Eocene temperature gradient [Table 1; @Sloan1990; @Markwick1994; @Huber2011; @Tierney2017; @Inglis2020].
 
@@
 Table 1: Inferred latitudinal sea surface temperature (SST) gradients for the early Eocene (EE) or the EECO, as shown in earlier, proxy-based studies.
-For comparison, a gradient derived from a climate model ensemble is also shown.
+For comparison, a gradient derived from an atmosphere-ocean general circulation model ensemble is also shown.
 
@@
 ```
-## Modern sea surface temperatures
-
 ## Geochemical data
@@
 This compilation provides data on four different geochemical proxies for reconstructing seawater temperature: &delta;^18^O, &Delta;47, Mg/Ca and $TEX_{86}$.
-For our analyses, this dataset was restricted to the EECO (defined as 49.1 - 53.8 Ma) and samples from the continental shelf.
+For our analyses, this dataset was restricted to the EECO (defined as 53.8 -- 49.1 Ma) and samples from the continental shelf.
 Recrystallised &delta;^18^O samples were also excluded as secondary diagenetic calcite precipitated after deposition can bias isotope measurements and offset temperature values [@Schrag1999].
@@
 
-**Coral reefs.** Today, shallow warm-water coral reefs are limited to tropical and subtropical latitudes (\~34$^\circ$N--32$^\circ$S), with minimum sea surface temperature tolerances (\~18$^\circ$C) being the primary constraint on this distribution [@johannes1983; @kleypas1999; @yamano2001].
+**Coral reefs.** Today, shallow warm-water coral reefs are limited to tropical and subtropical latitudes (\~34$^\circ$ N -- 32$^\circ$ S), with minimum sea surface temperature tolerances (\~18$^\circ$C) being the primary constraint on this distribution [@johannes1983; @kleypas1999; @yamano2001].
 As coral reefs reside at the upper thermal limit of the oceans today, their maximum sea surface temperature tolerance is less well-constrained, with some studies suggesting up to 35.6$^\circ$C in the geological past [@jones2022reef].
 Nevertheless, coral reefs have frequently been recognised as tracers of past (sub-)tropical conditions [@ziegler1984; @kiessling2001].
-During the Eocene, coral communities and reefs expanded across tropical and temperate latitudes, with some communities found up to palaeolatitudes of 43 $^\circ$N [@Zamagni2012].
+During the Eocene, coral communities and reefs expanded across tropical and temperate latitudes, with communities found up to palaeolatitudes of 43 $^\circ$ N [@Zamagni2012].
 Using a compilation of Paleocene--early Eocene coral reefs and community localities [@Zamagni2012], we generated quantitative sea surface temperature estimates for the ECCO.
 To do so, we extracted localities from the compilation that are inferred to be Ilerdian (early Eocene) coral reefs, and that could be confidently assigned to the EECO.
-We excluded coral knobs and coral-bearing mounds which might have broader climatic limits than coral reef ecosystems.
+We excluded coral knobs and coral-bearing mounds which might have broader climatic limits than warm-water coral reef ecosystems.
 This filtering resulted in four unique coral reef localities remaining for the EECO, all of which conform to the modern latitudinal range of coral reefs (\<34$^\circ$ N).
@@
 **Mangroves.** Mangroves are distributed throughout the tropics and subtropics today.
-While factors beside sea surface temperatures (SST) influence the distribution of mangroves, empirical, lower temperature limits have been established for the genera *Avicennia* (15.6$^\circ$C) and *Rhizophora* (20.7$^\circ$C) [@Quisthoudt2012].
-Both *Avicennia* and members of the Rhizophoraceae family were widespread in the early Eocene, but only *Avicennia* occurred at polar latitudes [@Suan2017; @Popescu2021].
+While factors besides sea surface temperatures (SST) influence the distribution of mangroves, empirical, lower temperature limits have been established for the genera *Avicennia* (15.6$^\circ$C) and *Rhizophora* (20.7$^\circ$C) [@Quisthoudt2012].
+Both *Avicennia* and members of the Rhizophoraceae family were widespread and co-occurred across tropical and temperate latitudes in the early Eocene.
+Only *Avicennia*, however, occurred at polar latitudes [@Suan2017; @Popescu2021].
 Assuming that Eocene members of these mangrove taxa conform to similar climatic requirements as their modern relatives, the presence and absence of *Avicennia* and Rhizophoraceae pollen can be used as a palaeotemperature indicator.
@@
 
-1)  *Avicennia*-only assemblages ($n=2$): the absence of Rhizophoraceae is indicative of temperatures being between 15.6$^\circ$C (lower temperature limt of *Avicennia*) and 20.7 (lower temperature limit of *Rhizophora*).
+1)  *Avicennia*-only assemblages ($n=2$): the absence of Rhizophoraceae is indicative of temperatures being between 15.6$^\circ$C (lower temperature limt of *Avicennia*) and 20.7$^\circ$C (lower temperature limit of *Rhizophora*).
 However, a value of 22.5$^\circ$C is ascribed as the upper temperature limit here as *Rhizophora* is rare below this temperature.
@@
 
-**Model structure (Fig 2).** We model the mean temperature ($\mu$) at location $j$ as a function of absolute latitude ($abs(l)$) with a logistic regression (also known as "growth curve" or "Richard's curve") of the form:
+**Model structure.** We model the mean temperature ($\mu$) at location $j$ as a function of absolute latitude ($abs(l)$) with a logistic regression (also known as "growth curve" or "Richard's curve") of the form:
 
@@
 $$\nu_{j} = A~+\frac{K-A}{e^{B(abs(l_j)-M)}}~, ~~~~~ j = 1,...,n,  $${#eq-nu} where $A$ and $K$ denote the lower and upper asymptote, respectively, $M$ specifies the latitude of maximal growth, i.e.
-the latitude around which temperature falls most steeply with latitude, $B$ denotes the growth rate (Fig.
-2), $\sigma$ denotes the residual standard deviation, and $n$ denotes the number of locations.
+the latitude around which temperature falls most steeply with latitude, $B$ denotes the growth rate, $\sigma$ denotes the residual standard deviation, and $n$ denotes the number of locations.
 
@@
 
-**Priors.** In the Bayesian framework, priors need to be placed on the unknown parameters of a model.
+**Priors.** In a Bayesian framework, priors need to be placed on the unknown parameters of a model.
 We placed weakly informative, conjugate inverse-gamma priors on $\sigma$ and $\sigma_{j=1,..n}$: $$\sigma  \sim \sqrt{Inv-Gamma(\alpha+\frac{n}{2},\beta + 0.5\times(\mu_j-\nu_j))}, ~~~~~ j = 1,...,n,  $${#eq-sigma}
@@
 
-In contrast, we put informative priors on the regression coefficients $A$, $K$, $M$ and $B$, based on physical principles, and vaguely based on the modern climate system:
+In contrast, we put informative priors on the regression coefficients $A$, $K$, $M$ and $B$, based on physical principles, and loosely based on the modern climate system:
 
@@
 
-**M.** The steepness of the gradient is presumed to be highest in mid-latitudes; this is expressed with a normal prior on $M$ with the mean set to $M$ of the modern SST gradient: $$M \sim N(42,10)$${#eq-M}
+**M.** The steepness of the gradient is presumed to be highest in mid-latitudes; this is expressed with a normal prior on $M$ with the mean set to 42, i.e.
+$M$ of the modern SST gradient, and a moderately wide standard deviation of 10: $$M \sim N(42,10)$${#eq-M}
 
-**B.** The steepness or growth rate $B$ of the gradient is constrained to be $\ge0$ and to not be exceedingly high, as oceanic and atmospheric heat transfer is bound to prevent very abrupt SST changes across latitudes.
+**B.** The steepness or growth rate $B$ of the gradient is constrained to be $\ge0$ and to not be exceedingly high, as oceanic and atmospheric heat transfer is bound to limit very abrupt SST changes across latitudes on a global scale.
 A gamma-distributed prior of the form $$B \sim Gamma(\alpha_G=4.3,\beta_G=30)$${#eq-B}
-was placed on $B$. The shape and rate parameters $\alpha_G$ and $\beta_G$ were chosen such that the highest prior density is at $B$ of the modern SST gradient, $0.11$.
+was placed on $B$.
+The shape and rate parameters $\alpha_G$ and $\beta_G$ were chosen such that the highest prior density is at $B$ of the modern SST gradient, $0.11$.
+We informed the prior distributions on M and B based on a provisional model run with the modern SST data.
 
@@
 
-To test whether our logistic regression model can adequately describe different latitudinal temperature gradients at various sample sizes, we generated four idealised gradients that emulate potential climatic states throughout Earth's geological history: extreme icehouse, icehouse, greenhouse and extreme greenhouse [@Frakes1992].
+To test whether our logistic regression model can adequately describe different latitudinal temperature gradients at various sample sizes, we generated four idealised gradients that emulate potential climatic states throughout Earth's geological history: extreme icehouse, icehouse, greenhouse, and extreme greenhouse [@Frakes1992].
 We then randomly sampled (1000 iterations) these gradients using increasing sample sizes (5, 10, and 20) and reconstructed the latitudinal temperature gradient using our model for each of these sample sizes and gradient types.
-Using the same idealised gradients, we also tested whether our model could accurately reconstruct latitudinal temperature gradients using the palaeogeographic distribution of Eocene samples (*n* = 34), providing an empirical, examplary distribution that captures both limited sample size and skewed geographic origins of samples.
-To assess how well the modeled gradients, reconstructed from limited sampling, fit the idealised gradients, we calculated an $R^2$ for Bayesian regression models [@Gelman2019].
-For every iteration from the posterior, we intercepted the modeled and the idealised gradient in intervals of 1$^\circ$ latitude, and calculated the $R^2$ based on these values.
-We report the median, and $2.5^{th}$ and $97.5^{th}$ percentile of the resulting $R^2$ measures.
+Using the same idealised gradients, we also tested whether our model could accurately reconstruct latitudinal temperature gradients using the palaeogeographic distribution of Eocene samples (*n* = 34), providing an empirical, exemplary distribution that captures both limited sample size and skewed geographic origins of samples.
+To evaluate how well the model performed in reconstructing the idealised gradients from limited sampling, we calculated the coefficient of determination ($R^2$) for Bayesian regression models [@Gelman2019].
+For every iteration from the posterior, we intercepted the modelled and the idealised gradient in intervals of 1$^\circ$ latitude and calculated the $R^2$ based on these values.
+We report the median, and 95% credible intervals (CI) of the resulting $R^2$ values.
+Here and in all other instances, the 95% CI refer to the interval between the 2.5% point and the 97.5% point of the samples or sampled posterior distribution.
 
-To test whether our model can accurately depict the shape of the modern sea surface temperature gradient, and to facilitate comparison with the Eocene gradient, we applied our model to annual sea surface mean temperatures from Bio-Oracle [@Assis2018], down-sampled to a $1^\circ \times 1 ^\circ$ raster (*n* = 46,131).
-The $R^2$ for the modern gradient was calculated as above [@Gelman2019], comparing the modeled gradient and the empirical temperature averages in 1$^\circ$ latitude bins.
+To test whether our model can accurately depict the shape of the modern sea surface temperature gradient, and to facilitate comparison with the Eocene gradient, we applied our model to annual sea surface mean temperatures from Bio-Oracle [@Assis2018], aggregated to a $1^\circ \times 1 ^\circ$ raster (*n* = 46,131).
+The $R^2$ for the modern gradient was calculated as above [@Gelman2019], comparing the modelled gradient and the empirical temperature averages in 1$^\circ$ latitude bins.
 Only the medians are reported for the modern gradient, as the 95% credible intervals are extremely narrow due to the high precision of the posterior estimates.
@@
 Every 10th iteration was retained, resulting in a total of 20,000 iterations with low autocorrelation.
-The re-sampled, simulated gradients were modeled in one chain with 10,000 iterations for each of the 1,000 random samples.
+The re-sampled, simulated gradients were modelled in one chain with 10,000 iterations for each of the 1,000 random samples.
 5,000 iterations each were discarded as burn-in, and every 25th iteration was kept, resulting in a total of 200,000 iterations across all 1,000 model runs.
@@
 The Eocene posterior inference is thus based on a total of 20,000 iterations with low autocorrelation (effective multivariate sample size for $A$, $K$, $M$ and $B$ is \> 18,000).
-Trace plots of the MCMC chains indicate convergence and good mixing of the chains (Supplementary Figure S1).
+Trace plots of the MCMC chains indicate convergence and good mixing of the chains (Fig.
+S1).
 
@@
 ```
-Modeled sea surface temperature estimates were generated with [@eq-nu], calculating the sea surface temperatures at any latitude with the parameter estimates of each iteration from the posterior.
-The median, $2.5^{th}$ and $97.5^{th}$ percentiles of temperatures where then taken from all temperature estimates obtained at the latitudes of interest.
+modelled sea surface temperature estimates were generated with [@eq-nu], calculating the sea surface temperatures at any latitude with the parameter estimates of each iteration from the posterior.
+The median and 95% CI of temperatures where then taken from all temperature estimates obtained at the latitudes of interest.
 
-The latitudinal gradient is calculated as the difference between the modeled temperature at the equator (0$^\circ$ latitude) and at the poles (90$^\circ$ absolute latitude).
+The latitudinal gradient is calculated as the difference between the modelled temperature at the equator (0$^\circ$ latitude) and at the poles (90$^\circ$ absolute latitude).
 To facilitate comparison with earlier estimates, we also calculate the gradient with the temperature at the polar circle (66.6$^\circ$ absolute latitude) being used instead of at the poles.
@@
 
-Differences between Eocene and modern temperatures at a certain latitude were calculated by randomly pairing all iterations of the posterior from the Eocene and modern temperature gradient model, calculating the Eocene and modern temperature using the respective iterations, taking the difference, and then calculating the median ($2.5^{th}$, $97.5^{th}$ percentile) from all pairs of iterations.
+Differences between Eocene and modern temperatures at a certain latitude were calculated by randomly pairing all iterations of the posterior from the Eocene and modern temperature gradient model, calculating the Eocene and modern temperature using the respective iterations, taking the difference, and then calculating the median (95% CI) from all pairs of iterations.
 
-Global average temperatures with 95% credible intervals were calculated by taking the weighted mean of the median ($2.5^{th}$, $97.5^{th}$ percentile) temperature estimates in 1$^\circ$ latitudinal bins.
+Global average temperatures with 95% credible intervals were calculated by taking the weighted mean of the median (95% CI) of temperature estimates in 1$^\circ$ latitudinal bins.
 The weights were set to the proportion of global surface area in each latitudinal bin, i.e.
-decreasing with increasing latitude as: $$weights = sin(\alpha_{1,i}) - sin(\alpha_{2,i}) ,$${#eq-weights} where $\alpha_1$ is the upper, and $\alpha_2$ is the lower latitudinal boundary of bin $i$.
+decreasing with increasing latitude as: $$weights = sin(\alpha_{1,i}) - sin(\alpha_{2,i}) ,$${#eq-weights} where $\alpha_1$ is the upper, and $\alpha_2$ is the lower latitudinal boundary of bin $i$, i.e.
+we approximated the shape of the globe as a spheroid.
 
@@
 motemp_global <- readRDS("results/modern/global_mean_modern.rds")
+motemp_empirical_global <- readRDS("results/modern/global_mean_empirical_modern.rds")
+
 eotemp_global <- readRDS("results/eeco/global_mean_eeco.rds")
@@
 
-## Model validation and modern gradient
+## Model validation
 
@@
 Each column depicts a different reconstruction for given sample sizes: 5, 10, 20, and 34 (latitudes of EECO samples).
-Each row depicts a different simulated latitudinal temperature gradient that represents idealised climatic states: extreme icehouse, icehouse, greenhouse, extreme greenhouse.
+Each row depicts a different simulated latitudinal temperature gradient that represents idealised climatic states: extreme icehouse, icehouse, greenhouse, and extreme greenhouse.
 The black line illustrates the simulated gradient.
@@
 The blue shadings depict the 90%, 95%, and 99% credible intervals.
-Bold black text within each panel depicts the coefficient of determination ($R^2$) for estimating goodness of fit between the simulated and modeled gradient.
-The median (50%) $R^2$ value along with the 2.5% and 97.5% confidence intervals from the 1,000 model runs are shown.
+Bold black text within each panel depicts the coefficient of determination ($R^2$) for estimating goodness of fit between the simulated and modelled gradient.
+The median (50%) $R^2$ value along with the 2.5% and 97.5% credible intervals from the 1,000 model runs are shown.
 Each gradient is depicted in absolute latitude.
@@
 Present-day latitudinal temperature gradient.
-The present-day empirical latitudinal temperature gradient (mean sea surface temperature) is depicted as a black line, and the gradient estimated by the Bayesian model is shown in turquoise.
-Grey points depict the individual cell values of the Bio-ORACLE grid of mean sea surface temperatures, which were used to infer the empirical and the modeled gradient.
+The present-day empirical latitudinal temperature gradient (median sea surface temperature) is depicted as a black line, and the gradient estimated by the Bayesian model is shown in turquoise.
+Grey points depict the individual cell values of the Bio-ORACLE grid of mean sea surface temperatures, which were used to infer the empirical and the modelled gradient.
 :::
 
-Our Bayesian model is able to accurately model a range of idealised temperature gradients, ranging from extreme icehouse to "super greenhouse" scenarios (Fig.
+Our Bayesian model is able to accurately model a range of idealised temperature gradients, ranging from extreme icehouse to 'super greenhouse' scenarios (Fig.
 2).
-Random latitudinal sampling results in highly accurate reconstructions already at a sample size of 10 for the icehouse scenarios (95 % CI of $R^2$ > 0.9).
-Greenhouse scenarios require more random samples to accurately predict high-latitude temperatures.
-This is because in the absence of high-latitude samples, the modeled gradient is heavily influenced by the priors, which we based on the modern, the only empirically known latitudinal temperature gradient.
+Random latitudinal sampling results in highly accurate reconstructions at a sample sizes as low as 10 for the icehouse scenarios (95 % CI of $R^2$ > 0.9).
+Greenhouse scenarios require additional samples to accurately predict high-latitude temperatures.
+This is because in the absence of high-latitude samples, the modelled gradient is heavily influenced by the priors, which we based on the modern, the only empirically known latitudinal temperature gradient.
 A sampling distribution resembling that of the early Eocene data set used in this study allows for a highly accurate reconstruction of even the extreme greenhouse scenario (95 % CI of $R^2$ > 0.95).
@@
 The average, modern temperature gradient can be closely approximated with our model when using the full modern SST dataset (Fig.
-3); almost all of the variation in the empirical median temperatures in bins of 1$^\circ$ absolute latitude is explained by the modeled gradient ($R^2 = `r R2_modern`$).
-The empirical gradient spans `r modern_gradient_empirical`$^\circ$C from the equator to the poles, the modeled gradient is only very slightly higher at `r modern_gradient$"50%"`$^\circ$C.
-The modern, global men sea surface temperature based on our modeled gradient is `r motemp_global$"50%"`$^\circ$C.
+3); almost all of the variation in the empirical median temperatures in bins of 1$^\circ$ absolute latitude is explained by the modelled gradient ($R^2 = `r R2_modern`$).
+The empirical gradient spans `r modern_gradient_empirical`$^\circ$C from the equator to the poles, the modelled gradient is only very slightly higher at `r modern_gradient$"50%"`$^\circ$C.
+The modern, global mean temperature (GMST) based on our modelled, median gradient is `r motemp_global$"50%"`$^\circ$C, very similar to the GMST derived from the empirical median gradient (`r motemp_empirical_global`$^\circ$C).
 
 <!-- REMOVED
-Reducing the amount and the spatial coverage of data points to match the early Eocene sampling distribution ($n = 34$) still results in a close match of the modeled gradient with the empirical gradient (Fig.
+Reducing the amount and the spatial coverage of data points to match the early Eocene sampling distribution ($n = 34$) still results in a close match of the modelled gradient with the empirical gradient (Fig.
 3b).
@@
 3a).
-The modern, latitudinal gradient, modeled with all data, spans `r modern_gradient$"50%"`$^\circ$C from the equator to the poles.
-When modeled with only the early Eocene sampling locations, the median temperature gradient is `r modern_eocene_gradient$"50%"` (95% CI: `r modern_eocene_gradient$"2.5%"` - `r modern_eocene_gradient$"97.5%"`)$^\circ$C.
+The modern, latitudinal gradient, modelled with all data, spans `r modern_gradient$"50%"`$^\circ$C from the equator to the poles.
+When modelled with only the early Eocene sampling locations, the median temperature gradient is `r modern_eocene_gradient$"50%"` (95% CI: `r modern_eocene_gradient$"2.5%"` -- `r modern_eocene_gradient$"97.5%"`)$^\circ$C.
 A random sampling of modern temperatures at latitudes corresponding to the sampled Eocene palaeolatitudes would thus allow the reconstruction of the modern latitudinal temperature gradient with good accuracy, with expected deviations from the actual gradient as low as `r modern_eocene_gradient_diff$"2.5%"` to `r modern_eocene_gradient_diff$"97.5%"`$^\circ$C (95 % CI).
@@
 
-Modeling the latitudinal temperature gradient facilitates accurate reconstruction of global mean surface temperatures (GMST).
+modelling the latitudinal temperature gradient facilitates accurate reconstruction of global mean surface temperatures (GMST).
 The modern, global mean sea surface temperature (GMsST) estimate is `r motemp_global$"50%"`$^\circ$C when using the full, modern data set.
-When reduced to the early Eocene sampling distribution, the modern GMsST is estimated at `r moeotemp_global$"50%"` (`r moeotemp_global$"2.5%"` - `r moeotemp_global$"97.5%"`)$^\circ$C.
+When reduced to the early Eocene sampling distribution, the modern GMsST is estimated at `r moeotemp_global$"50%"` (`r moeotemp_global$"2.5%"` -- `r moeotemp_global$"97.5%"`)$^\circ$C.
 The deviations from the full modern estimate are again very low, with a 95 % credible interval of `r moeotemp_global_diff$"2.5%"` to `r moeotemp_global_diff$"97.5%"`$^\circ$C.
@@
 
-The modeled Eocene temperature gradient is starkly different from the modern (Fig 4).
-Modelled, median equatorial temperatures are `r tdiff_lat0$"50%"` (`r tdiff_lat0$"2.5%"` - `r tdiff_lat0$"97.5%"`)$^\circ$C higher for the EECO, and polar temperatures are higher by `r tdiff_lat90$"50%"` (`r tdiff_lat90$"2.5%"` - `r tdiff_lat90$"97.5%"`)$^\circ$C.
-This results in a flattened latitudinal temperature gradient of `r eocene_gradient$"50%"` (`r eocene_gradient$"2.5%"` - `r eocene_gradient$"97.5%"`)$^\circ$C for the EECO, as opposed to `r modern_gradient$"50%"`$^\circ$C for the modern.
-To facilitate the comparison with latitudinal gradients reported in the literature, which sometimes do not report temperatures in very high latitudes, we report also the EECO gradient between the equator and the modern-day polar circle (66.6$^\circ$), which is slightly lower at `r eocene_gradient_polar_circle$"50%"` (`r eocene_gradient_polar_circle$"2.5%"` - `r eocene_gradient_polar_circle$"97.5%"`)$^\circ$C.
+The modelled Eocene temperature gradient is starkly different from the modern (Fig 4).
+Modelled, median equatorial temperatures are `r tdiff_lat0$"50%"` (95% CI: `r tdiff_lat0$"2.5%"` -- `r tdiff_lat0$"97.5%"`)$^\circ$C higher for the EECO, and polar temperatures are `r tdiff_lat90$"50%"` (`r tdiff_lat90$"2.5%"` -- `r tdiff_lat90$"97.5%"`)$^\circ$C higher.
+This results in a flattened latitudinal temperature gradient of `r eocene_gradient$"50%"` (`r eocene_gradient$"2.5%"` -- `r eocene_gradient$"97.5%"`)$^\circ$C for the EECO, as opposed to `r modern_gradient$"50%"`$^\circ$C for the modern.
+To facilitate the comparison with latitudinal gradients reported in the literature, which sometimes do not report temperatures at very high latitudes, we report also the EECO gradient between the equator and the modern-day polar circle (66.6$^\circ$), which is slightly lower at `r eocene_gradient_polar_circle$"50%"` (`r eocene_gradient_polar_circle$"2.5%"` -- `r eocene_gradient_polar_circle$"97.5%"`)$^\circ$C.
 
-The high variability of EECO palaeotemperature proxies, particularly in the mid-latitudes, and the scarcity of high-latitude data, result in substantial uncertainties in the modeled temperature gradient.
-This is reflected in the residual standard deviation ($\sigma$) of the EECO gradient, `r eeco_sdy$"50%"` (`r eeco_sdy$"2.5%"` - `r eeco_sdy$"97.5%"`)$^\circ$C, more than twice as high than $\sigma$ of the modern gradient, `r modern_sdy$"50%"`.
-This signifies that the early Eocene data fit less well to the logistic, latitudinal gradient model, which can also be seen from the drastic departure of some of the proxy data from the gradient estimates (Fig.
+The high variability of EECO palaeotemperature proxies, particularly in the mid-latitudes, and the scarcity of high-latitude data, result in substantial uncertainties in the modelled temperature gradient.
+This is reflected in the residual standard deviation ($\sigma$) of the EECO gradient -- `r eeco_sdy$"50%"` (`r eeco_sdy$"2.5%"` -- `r eeco_sdy$"97.5%"`)$^\circ$C -- which is more than double the  $\sigma$ for the modern gradient, `r modern_sdy$"50%"`.
+This signifies that the early Eocene data does not fit as well to the logistic latitudinal gradient model, which can also be seen from the drastic departure of some of the proxy data from the gradient estimates (Fig.
 4).
 
-The early Eocene GMsST is estimated at `r eotemp_global$"50%"` (`r eotemp_global$"2.5%"` - `r eotemp_global$"97.5%"`)$^\circ$C, `r  round(as.numeric(eotemp_global$"50%") - as.numeric(motemp_global$"50%"),1)`$^\circ$C higher than the modern.
-A model run excluding the ecological proxies increases the GMST by `r eoceneproxy_global_diff$"50%"` (`r eoceneproxy_global_diff$"2.5%"` - `r eoceneproxy_global_diff$"97.5%"`)$^\circ$C.
+The early Eocene GMST is estimated at `r eotemp_global$"50%"` (`r eotemp_global$"2.5%"` -- `r eotemp_global$"97.5%"`)$^\circ$C, `r  round(as.numeric(eotemp_global$"50%") - as.numeric(motemp_global$"50%"),1)`$^\circ$C higher than the modern.
+A model run excluding the ecological proxies increases the GMST by `r eoceneproxy_global_diff$"50%"` (`r eoceneproxy_global_diff$"2.5%"` -- `r eoceneproxy_global_diff$"97.5%"`)$^\circ$C.
 The median latitudinal gradient is similar when excluding the ecological proxies, with a median of `r eocene_proxy_gradient$"50%"`$^\circ$C, but with a `r proxy_uncertain_diff_percent`% wider 95% CI (Fig.
@@
 Due to the limited spatial coverage of the early Eocene proxy record, and due to the added model complexity of simultaneously estimating a model across both hemispheres, we pooled the proxy data across both hemispheres.
-Applying the model separately to each hemispheres results in substantial differences in hemispherical, average temperatures, with the southern hemisphere being warmer by `r eeco_south_north_global_diff$"50%"` (`r eeco_south_north_global_diff$"2.5%"` - `r eeco_south_north_global_diff$"97.5%"`)$^\circ$C.
-The inferred latitudinal gradient is somewhat steeper in the northern hemisphere (steeper by `r eocene_north_south_gradient_diff$"50%"`$^\circ$C, although the 95% CI spans `r eocene_north_south_gradient_diff$"2.5%"` to `r eocene_north_south_gradient_diff$"97.5%"`$^\circ$C), but the large uncertainties associated with both gradients, and the lack of polar proxy data in the southern hemisphere preclude a more precise statement (see Fig.
+Applying the model separately within each hemisphere results in substantial differences in hemispherical, average temperatures, with the Southern Hemisphere being warmer by `r eeco_south_north_global_diff$"50%"` (`r eeco_south_north_global_diff$"2.5%"` -- `r eeco_south_north_global_diff$"97.5%"`)$^\circ$C.
+The inferred latitudinal gradient is somewhat steeper in the Northern Hemisphere (steeper by `r eocene_north_south_gradient_diff$"50%"`$^\circ$C, although the 95% CI spans `r eocene_north_south_gradient_diff$"2.5%"` -- `r eocene_north_south_gradient_diff$"97.5%"`$^\circ$C), but the large uncertainties associated with both gradients, and the lack of polar proxy data in the Southern Hemisphere preclude a more precise statement (see Fig.
 S3).
@@
 
-Estimates of the median, latitudinal SST gradients of the early Eocene climatic optimum (purple line) and of the present-day (turquoise), both estimated with the Bayesian model.
-The ribbons (purple and turquoise shadings) depict the 95% credible intervals of these gradients, respectively.
+Estimates of the median, latitudinal sea surface temperature gradients of the early Eocene climatic optimum (purple line) and of the present-day (turquoise), both estimated with the Bayesian model.
+The ribbons (purple and turquoise shading) depict the 95% credible intervals of these gradients, respectively.
 Points within the plot depict the geochemical (e.g.
@@
 In contrast, our Bayesian, parametric model allows for the inclusion of informative priors on the model parameters.
-The modeled sea surface temperature gradient thus does not strictly follow the proxy data, but instead represents a compromise between the data and prior knowledge.
+The modelled sea surface temperature gradient thus does not strictly follow the proxy data, but instead represents a compromise between the data and prior knowledge.
 In the EECO example (Fig.
@@
 For example, the greater amount of solar radiation per unit area in low latitudes causes Earth's latitudinal temperature gradient to be broadly negative [@Beer2008].
-The ease with which such prior information can be integrated is a major advantage of our method, as the shape of the modeled gradient is controlled by four parameters which clearly relate to its magnitude, steepness and the latitude of its greatest steepness.
+The ease with which such prior information can be integrated is a major advantage of our method, as the shape of the modelled gradient is controlled by four parameters which clearly relate to its magnitude, steepness and the latitude of its greatest steepness.
 
@@
 The calculation of global mean surface temperatures directly from sparse proxy data is susceptible to bias [@Jones2022].
-By modeling the temperature variation across latitudes, a complete temperature distribution along a latitudinal axis can be obtained, filling in gaps in the proxy record through inter- or extrapolation.
+By modelling the temperature variation across latitudes, a complete temperature distribution along a latitudinal axis can be obtained, filling in gaps in the proxy record through inter- or extrapolation.
 This eliminates the common problem that specific climate zones dominate the proxy record.
@@
 Calculating zonal averages alleviates this problem, but this method relies on comprehensive latitudinal coverage [@Inglis2020].
-Instead, our method allows for intersecting the modeled temperature gradient at narrow latitudinal intervals, even when significant latitudinal gaps exist.
+Instead, our method allows for intersecting the modelled temperature gradient at narrow latitudinal intervals, even when significant latitudinal gaps exist.
 Weighting the temperatures of those latitudinal intervals by area results in GMST estimates without intrinsic spatial biases.
@@
 Combining the occurrences of climate-sensitive plant communities [@Greenwood1995], reptiles [@Markwick2007], leaf shapes [@Peppe2011], with geochemical proxies offers substantial potential for improving quantitative palaeoclimate reconstructions across the Phanerozoic.
-Our modeling framework offers a straightforward, efficient way of integrating ecological climate data with other proxy data: The hierarchical model structure accounts for variation of temperature estimates from proxies at individual localities, which is treated equivalent to the uncertainty associated with the ecological temperature proxies.
+Our modelling framework offers a straightforward, efficient way of integrating ecological climate data with other proxy data: The hierarchical model structure accounts for variation of temperature estimates from proxies at individual localities, which is treated equivalent to the uncertainty associated with the ecological temperature proxies.
 A local temperature estimate, based on multiple geochemical proxies, thus has the same weight as a local temperature estimate obtained from the occurrence of a climate-sensitive plant community, whilst preserving the uncertainty associated with each estimate.
@@
 <!-- @Evans2018 estimate the SST gradient at 32+-10% of the modern, which would correspond to \~`r round( as.numeric(modern_gradient$"50%") - 0.32*as.numeric(modern_gradient$"50%"),0)`+- `r format(round( 0.10*as.numeric(modern_gradient$"50%"),0),nsmall=0)`$^\circ$C, using the difference between the mean tropical and deep ocean data.
-The modeled SST of @Tierney2017 using TEX86 is \~ `r round(35.8-23.8,0)`$^\circ$C, when taking the difference between temperatures at the polar circles and the equator.
+The modelled SST of @Tierney2017 using TEX86 is \~ `r round(35.8-23.8,0)`$^\circ$C, when taking the difference between temperatures at the polar circles and the equator.
 @Inglis2020 include both terrestrial air temperature and SST estimates in their model of the latitudinal temperature gradient in their Figure S4a; they model a polar circle to equatorial gradient of \~ `r round(30.7-(20.2+14.3)/2,0)`$^\circ$C.
@@
 
-Discrepancies between earlier, proxy-based reconstructions and our modeling results are most pronounced in latitudes beyond the polar circle, as earlier approaches [e.g.
+Discrepancies between earlier, proxy-based reconstructions and our modelling results are most pronounced in latitudes beyond the polar circle, as earlier approaches [e.g.
 @Tierney2017] predict almost linearly decreasing SSTs towards the poles, whereas our median prediction suggests only a slight decrease beyond the polar circle.
@@
 Polar temperature estimates from our model are thus conservative in that they admit large uncertainty where data is absent, which is desirable.
-However, the presence of high proxy-derived temperature estimates at \~ 60$^\circ$ latitudes forces the modeled median temperature curve to be too high at \~ 24$^\circ$C, relative to the temperatures indicated by the high-latitude mangrove communities (15.6 - 22.5$^\circ$C).
+However, the presence of high proxy-derived temperature estimates at \~ 60$^\circ$ latitudes forces the modelled median temperature curve to be too high at \~ 24$^\circ$C, relative to the temperatures indicated by the high-latitude mangrove communities (15.6 - 22.5$^\circ$C).
 In contrast, the extrapolated polar temperatures of most previous proxy-based models are likely too low, given the abundance of ecological data indicating temperate or subtropical high-latitude climates during the EECO [@Pross2012; @Popescu2021].
@@
 The use of a Bayesian framework allows for the specification of prior information based on physical constraints, which can improve interpolation and prediction.
-By modeling the temperature variation across latitudes, a complete temperature distribution is obtained, filling in gaps in the proxy record and reducing the risk of bias in global mean surface temperature (GMST) estimates.
+By modelling the temperature variation across latitudes, a complete temperature distribution is obtained, filling in gaps in the proxy record and reducing the risk of bias in global mean surface temperature (GMST) estimates.
 The hierarchical structure of the model accounts for the variation of temperature estimates at individual localities and uncertainty associated with the ecological temperature proxies.

### 2023-05-31 manuscript/main.qmd
Update main.qmd

@@
 
-Our results show that relatively small (*n* = 10 - 20) and patchy samples of temperature data are sufficient to reconstruct greenhouse- and icehouse-type, latitudinal SST gradients, using our Bayesian model.
+Our results show that our Bayesian model can be used to reconstruct different types of latitudinal SST gradients from proxy data, even with small sample sizes (*n* = 10 - 20) and patchy sampling distributions (Fig.
+2).
 This is an advancement over previously used linear, quadratic or Gaussian approximations [e.g.
 @Bijl2009; @Tierney2017], which can fit only specific types of gradients.
-As such, our model presents an alternative to non-parametric methods for inferring latitudinal temperature gradients, which are sometimes favoured as they can very flexibly follow the shape of an unknown temperature gradient [e.g.
+As such, our model presents an alternative to non-parametric methods for inferring latitudinal temperature gradients, which are sometimes favoured as they can flexibly follow the shape of an unknown temperature gradient [e.g.
 @Zhang2019; @Jones2022].
@@
 @Inglis2020].
-This means that the idiosyncrasies of a patchy proxy record, potentially afflicted with measurement errors, calibration errors, and palaeogeographic and temporal uncertainty, directly generates the resulting large-scale climate patterns.
+This means that the idiosyncrasies of a patchy proxy record, potentially afflicted with measurement errors, calibration errors, and palaeogeographic and temporal uncertainty, dictate the reconstruction of large-scale climate patterns without the option of including additional knowledge (e.g.
+that latitudinal temperature gradients should be broadly negative).
 

### 2023-06-01 manuscript/main.qmd
Update main.qmd
manuscript and cover letter updates

@@
   - \usepackage{newunicodechar}
-  - \newcommand{\$^\circ$C}{\textdegree C}
+  - \newcommand{\&deg;C}{\textdegree C}
 bibliography: references.bib
@@
 ::: {#table-1}
-![](../figures/Table1.png){width="16.25cm"}
 Table 1: Inferred latitudinal sea surface temperature (SST) gradients for the early Eocene (EE) or the EECO, as shown in earlier, proxy-based studies.
-For comparison, a gradient derived from an atmosphere-ocean general circulation model ensemble is also shown.
+For comparison, a gradient derived from an atmosphere-ocean general circulation model (GCM) ensemble is also shown.
 
+```{r, results='asis', echo = FALSE}
+library(knitr)
+
+data <- data.frame(
+  Source = c("Bijl et al.
+(2009)", "Keating-Bitoni et al.
+(2011)", "Tierney et al.
+(2017)", "Cramwinckel et al.
+(2018)", "Evans et al.
+(2018)", "Pross et al.
+(2012), as shown in Tierney et al.
+(2017)"),
+  Time = c("EE", "EECO", "EE", "EECO", "EE", "EE"),
+  Gradient = c("7", "13", "12", "21 (±1)", "20 (±3)", "26"),
+  Type_of_gradient = c("equator - polar circle", "equator - polar circle", "equator - polar circle", "equator - deep water", "tropics - deep water", "equator - polar circle"),
+  Model = c("$2^{nd}$ order polynomial",
+            "$2^{nd}$ order polynomial",
+            "Gaussian function",
+            "-",
+            "-",
+            "climate model ensemble (GCM)"),
+  Proxy_system = c("$TEX_{86}$, $UK_{37}^{K'}$",
+                   "$TEX_{86}$, MBT/CBT, $\\Delta_{47}$, Mg/Ca, $\\delta^{18}O$",
+                   "$TEX_{86}$",
+                   "$TEX_{86}$, $\\Delta_{47}$, Mg/Ca, $\\delta^{18}O$, deepwater $\\delta^{18}O$",
+                   "$\\Delta_{47}$, deepwater Mg/Ca",
+                   "none (GCM simulations)")
+)
+
+kable(data, format = "markdown")
+```
 :::
@@
 
-**Coral reefs.** Today, shallow warm-water coral reefs are limited to tropical and subtropical latitudes (\~34$^\circ$ N -- 32$^\circ$ S), with minimum sea surface temperature tolerances (\~18$^\circ$C) being the primary constraint on this distribution [@johannes1983; @kleypas1999; @yamano2001].
-As coral reefs reside at the upper thermal limit of the oceans today, their maximum sea surface temperature tolerance is less well-constrained, with some studies suggesting up to 35.6$^\circ$C in the geological past [@jones2022reef].
+**Coral reefs.** Today, shallow warm-water coral reefs are limited to tropical and subtropical latitudes (\~34&deg;&nbsp;N -- 32&deg;&nbsp;S), with minimum sea surface temperature tolerances (\~18&deg;C) being the primary constraint on this distribution [@johannes1983; @kleypas1999; @yamano2001].
+As coral reefs reside at the upper thermal limit of the oceans today, their maximum sea surface temperature tolerance is less well-constrained, with some studies suggesting up to 35.6&deg;C in the geological past [@jones2022reef].
 Nevertheless, coral reefs have frequently been recognised as tracers of past (sub-)tropical conditions [@ziegler1984; @kiessling2001].
-During the Eocene, coral communities and reefs expanded across tropical and temperate latitudes, with communities found up to palaeolatitudes of 43 $^\circ$ N [@Zamagni2012].
+During the Eocene, coral communities and reefs expanded across tropical and temperate latitudes, with communities found up to palaeolatitudes of 43 &deg;&nbsp;N [@Zamagni2012].
 Using a compilation of Paleocene--early Eocene coral reefs and community localities [@Zamagni2012], we generated quantitative sea surface temperature estimates for the ECCO.
@@
 We excluded coral knobs and coral-bearing mounds which might have broader climatic limits than warm-water coral reef ecosystems.
-This filtering resulted in four unique coral reef localities remaining for the EECO, all of which conform to the modern latitudinal range of coral reefs (\<34$^\circ$ N).
-Subsequently, we used statistically derived temperature limits (minimum = 21$^\circ$C, average = 27.6$^\circ$C, maximum = 29.5$^\circ$C) from the published literature [@kleypas1999] to define a normal probability distribution of potential temperature values for coral reef localities.
+This filtering resulted in four unique coral reef localities remaining for the EECO, all of which conform to the modern latitudinal range of coral reefs (\<34&deg;&nbsp;N).
+Subsequently, we used statistically derived temperature limits (minimum = 21&deg;C, average = 27.6&deg;C, maximum = 29.5&deg;C) from the published literature [@kleypas1999] to define a normal probability distribution of potential temperature values for coral reef localities.
 This normal probability distribution was defined with a mean of 27.6 and a standard deviation of 2.125, placing 97.5% of the probability density above the minimum.
-As the distribution of modern corals is skewed towards warmer temperatures, this approach results in 16.5% of the probability being placed on temperatures \> 29.5$^\circ$C, allowing for the possibility that Eocene coral reefs were adapted to warmer conditions than present-day coral reefs.
+As the distribution of modern corals is skewed towards warmer temperatures, this approach results in 16.5% of the probability being placed on temperatures \> 29.5&deg;C, allowing for the possibility that Eocene coral reefs were adapted to warmer conditions than present-day coral reefs.
 
 **Mangroves.** Mangroves are distributed throughout the tropics and subtropics today.
-While factors besides sea surface temperatures (SST) influence the distribution of mangroves, empirical, lower temperature limits have been established for the genera *Avicennia* (15.6$^\circ$C) and *Rhizophora* (20.7$^\circ$C) [@Quisthoudt2012].
+While factors besides sea surface temperatures (SST) influence the distribution of mangroves, empirical, lower temperature limits have been established for the genera *Avicennia* (15.6&deg;C) and *Rhizophora* (20.7&deg;C) [@Quisthoudt2012].
 Both *Avicennia* and members of the Rhizophoraceae family were widespread and co-occurred across tropical and temperate latitudes in the early Eocene.
@@
 
-1)  *Avicennia*-only assemblages ($n=2$): the absence of Rhizophoraceae is indicative of temperatures being between 15.6$^\circ$C (lower temperature limt of *Avicennia*) and 20.7$^\circ$C (lower temperature limit of *Rhizophora*).
-However, a value of 22.5$^\circ$C is ascribed as the upper temperature limit here as *Rhizophora* is rare below this temperature.
+1)  *Avicennia*-only assemblages ($n=2$): the absence of Rhizophoraceae is indicative of temperatures being between 15.6&deg;C (lower temperature limt of *Avicennia*) and 20.7&deg;C (lower temperature limit of *Rhizophora*).
+However, a value of 22.5&deg;C is ascribed as the upper temperature limit here as *Rhizophora* is rare below this temperature.
 We define the *Avicennia*-only temperature distribution as a normal distribution with a mean of 19.05 and a standard deviation of 1.725, resulting in 95% of the probability density being placed within the temperature limits.
 
-2)  *Avicennia* and Rhizophoraceae assemblages ($n=5$): the presence of both groups suggests that the locality should have a minimum temperature of 20.7$^\circ$C (lower temperature limit of *Rhizophora*).
-As the upper thermal limits of *Aviciennia* and *Rhizophora* are not well established in @Quisthoudt2012, we assign the same maximum temperature limits (29.5$^\circ$C) as coral reef localities, because mangroves are also widely distributed throughout tropical regions.
+2)  *Avicennia* and Rhizophoraceae assemblages ($n=5$): the presence of both groups suggests that the locality should have a minimum temperature of 20.7&deg;C (lower temperature limit of *Rhizophora*).
+As the upper thermal limits of *Aviciennia* and *Rhizophora* are not well established in @Quisthoudt2012, we assign the same maximum temperature limits (29.5&deg;C) as coral reef localities, because mangroves are also widely distributed throughout tropical regions.
 Consequently, we define the temperature distribution for this locality as a normal distribution with a mean of 25.1 and a standard deviation of 2.2, with 95% probability density within the temperature limits.
@@
 To evaluate how well the model performed in reconstructing the idealised gradients from limited sampling, we calculated the coefficient of determination ($R^2$) for Bayesian regression models [@Gelman2019].
-For every iteration from the posterior, we intercepted the modelled and the idealised gradient in intervals of 1$^\circ$ latitude and calculated the $R^2$ based on these values.
+For every iteration from the posterior, we intercepted the modelled and the idealised gradient in intervals of 1&deg; latitude and calculated the $R^2$ based on these values.
 We report the median, and 95% credible intervals (CI) of the resulting $R^2$ values.
@@
 To test whether our model can accurately depict the shape of the modern sea surface temperature gradient, and to facilitate comparison with the Eocene gradient, we applied our model to annual sea surface mean temperatures from Bio-Oracle [@Assis2018], aggregated to a $1^\circ \times 1 ^\circ$ raster (*n* = 46,131).
-The $R^2$ for the modern gradient was calculated as above [@Gelman2019], comparing the modelled gradient and the empirical temperature averages in 1$^\circ$ latitude bins.
+The $R^2$ for the modern gradient was calculated as above [@Gelman2019], comparing the modelled gradient and the empirical temperature averages in 1&deg; latitude bins.
 Only the medians are reported for the modern gradient, as the 95% credible intervals are extremely narrow due to the high precision of the posterior estimates.
@@
 
-The latitudinal gradient is calculated as the difference between the modelled temperature at the equator (0$^\circ$ latitude) and at the poles (90$^\circ$ absolute latitude).
-To facilitate comparison with earlier estimates, we also calculate the gradient with the temperature at the polar circle (66.6$^\circ$ absolute latitude) being used instead of at the poles.
+The latitudinal gradient is calculated as the difference between the modelled temperature at the equator (0&deg; latitude) and at the poles (90&deg; absolute latitude).
+To facilitate comparison with earlier estimates, we also calculate the gradient with the temperature at the polar circle (66.6&deg; absolute latitude) being used instead of at the poles.
 Given the sigmoidal shape of the modern as well as the Eocene gradient (see Fig.
@@
 
-Global average temperatures with 95% credible intervals were calculated by taking the weighted mean of the median (95% CI) of temperature estimates in 1$^\circ$ latitudinal bins.
+Global average temperatures with 95% credible intervals were calculated by taking the weighted mean of the median (95% CI) of temperature estimates in 1&deg; latitudinal bins.
 The weights were set to the proportion of global surface area in each latitudinal bin, i.e.
@@
 The average, modern temperature gradient can be closely approximated with our model when using the full modern SST dataset (Fig.
-3); almost all of the variation in the empirical median temperatures in bins of 1$^\circ$ absolute latitude is explained by the modelled gradient ($R^2 = `r R2_modern`$).
-The empirical gradient spans `r modern_gradient_empirical`$^\circ$C from the equator to the poles, the modelled gradient is only very slightly higher at `r modern_gradient$"50%"`$^\circ$C.
-The modern, global mean temperature (GMST) based on our modelled, median gradient is `r motemp_global$"50%"`$^\circ$C, very similar to the GMST derived from the empirical median gradient (`r motemp_empirical_global`$^\circ$C).
+3); almost all of the variation in the empirical median temperatures in bins of 1&deg; absolute latitude is explained by the modelled gradient ($R^2 = `r R2_modern`$).
+The empirical gradient spans `r modern_gradient_empirical`&deg;C from the equator to the poles, the modelled gradient is only very slightly higher at `r modern_gradient$"50%"`&deg;C.
+The modern, global mean temperature (GMST) based on our modelled, median gradient is `r motemp_global$"50%"`&deg;C, very similar to the GMST derived from the empirical median gradient (`r motemp_empirical_global`&deg;C).
 
@@
 3a).
-The modern, latitudinal gradient, modelled with all data, spans `r modern_gradient$"50%"`$^\circ$C from the equator to the poles.
-When modelled with only the early Eocene sampling locations, the median temperature gradient is `r modern_eocene_gradient$"50%"` (95% CI: `r modern_eocene_gradient$"2.5%"` -- `r modern_eocene_gradient$"97.5%"`)$^\circ$C.
-A random sampling of modern temperatures at latitudes corresponding to the sampled Eocene palaeolatitudes would thus allow the reconstruction of the modern latitudinal temperature gradient with good accuracy, with expected deviations from the actual gradient as low as `r modern_eocene_gradient_diff$"2.5%"` to `r modern_eocene_gradient_diff$"97.5%"`$^\circ$C (95 % CI).
+The modern, latitudinal gradient, modelled with all data, spans `r modern_gradient$"50%"`&deg;C from the equator to the poles.
+When modelled with only the early Eocene sampling locations, the median temperature gradient is `r modern_eocene_gradient$"50%"` (95% CI: `r modern_eocene_gradient$"2.5%"` -- `r modern_eocene_gradient$"97.5%"`)&deg;C.
+A random sampling of modern temperatures at latitudes corresponding to the sampled Eocene palaeolatitudes would thus allow the reconstruction of the modern latitudinal temperature gradient with good accuracy, with expected deviations from the actual gradient as low as `r modern_eocene_gradient_diff$"2.5%"` to `r modern_eocene_gradient_diff$"97.5%"`&deg;C (95 % CI).
 
@@
 modelling the latitudinal temperature gradient facilitates accurate reconstruction of global mean surface temperatures (GMST).
-The modern, global mean sea surface temperature (GMsST) estimate is `r motemp_global$"50%"`$^\circ$C when using the full, modern data set.
-When reduced to the early Eocene sampling distribution, the modern GMsST is estimated at `r moeotemp_global$"50%"` (`r moeotemp_global$"2.5%"` -- `r moeotemp_global$"97.5%"`)$^\circ$C.
-The deviations from the full modern estimate are again very low, with a 95 % credible interval of `r moeotemp_global_diff$"2.5%"` to `r moeotemp_global_diff$"97.5%"`$^\circ$C.
+The modern, global mean sea surface temperature (GMsST) estimate is `r motemp_global$"50%"`&deg;C when using the full, modern data set.
+When reduced to the early Eocene sampling distribution, the modern GMsST is estimated at `r moeotemp_global$"50%"` (`r moeotemp_global$"2.5%"` -- `r moeotemp_global$"97.5%"`)&deg;C.
+The deviations from the full modern estimate are again very low, with a 95 % credible interval of `r moeotemp_global_diff$"2.5%"` to `r moeotemp_global_diff$"97.5%"`&deg;C.
 -->
@@
 The modelled Eocene temperature gradient is starkly different from the modern (Fig 4).
-Modelled, median equatorial temperatures are `r tdiff_lat0$"50%"` (95% CI: `r tdiff_lat0$"2.5%"` -- `r tdiff_lat0$"97.5%"`)$^\circ$C higher for the EECO, and polar temperatures are `r tdiff_lat90$"50%"` (`r tdiff_lat90$"2.5%"` -- `r tdiff_lat90$"97.5%"`)$^\circ$C higher.
-This results in a flattened latitudinal temperature gradient of `r eocene_gradient$"50%"` (`r eocene_gradient$"2.5%"` -- `r eocene_gradient$"97.5%"`)$^\circ$C for the EECO, as opposed to `r modern_gradient$"50%"`$^\circ$C for the modern.
-To facilitate the comparison with latitudinal gradients reported in the literature, which sometimes do not report temperatures at very high latitudes, we report also the EECO gradient between the equator and the modern-day polar circle (66.6$^\circ$), which is slightly lower at `r eocene_gradient_polar_circle$"50%"` (`r eocene_gradient_polar_circle$"2.5%"` -- `r eocene_gradient_polar_circle$"97.5%"`)$^\circ$C.
+Modelled, median equatorial temperatures are `r tdiff_lat0$"50%"` (95% CI: `r tdiff_lat0$"2.5%"` -- `r tdiff_lat0$"97.5%"`)&deg;C higher for the EECO, and polar temperatures are `r tdiff_lat90$"50%"` (`r tdiff_lat90$"2.5%"` -- `r tdiff_lat90$"97.5%"`)&deg;C higher.
+This results in a flattened latitudinal temperature gradient of `r eocene_gradient$"50%"` (`r eocene_gradient$"2.5%"` -- `r eocene_gradient$"97.5%"`)&deg;C for the EECO, as opposed to `r modern_gradient$"50%"`&deg;C for the modern.
+To facilitate the comparison with latitudinal gradients reported in the literature, which sometimes do not report temperatures at very high latitudes, we report also the EECO gradient between the equator and the modern-day polar circle (66.6&deg;), which is slightly lower at `r eocene_gradient_polar_circle$"50%"` (`r eocene_gradient_polar_circle$"2.5%"` -- `r eocene_gradient_polar_circle$"97.5%"`)&deg;C.
 
 The high variability of EECO palaeotemperature proxies, particularly in the mid-latitudes, and the scarcity of high-latitude data, result in substantial uncertainties in the modelled temperature gradient.
-This is reflected in the residual standard deviation ($\sigma$) of the EECO gradient -- `r eeco_sdy$"50%"` (`r eeco_sdy$"2.5%"` -- `r eeco_sdy$"97.5%"`)$^\circ$C -- which is more than double the  $\sigma$ for the modern gradient, `r modern_sdy$"50%"`.
+This is reflected in the residual standard deviation ($\sigma$) of the EECO gradient -- `r eeco_sdy$"50%"` (`r eeco_sdy$"2.5%"` -- `r eeco_sdy$"97.5%"`)&deg;C -- which is more than double the  $\sigma$ for the modern gradient, `r modern_sdy$"50%"`.
 This signifies that the early Eocene data does not fit as well to the logistic latitudinal gradient model, which can also be seen from the drastic departure of some of the proxy data from the gradient estimates (Fig.
@@
 
-The early Eocene GMST is estimated at `r eotemp_global$"50%"` (`r eotemp_global$"2.5%"` -- `r eotemp_global$"97.5%"`)$^\circ$C, `r  round(as.numeric(eotemp_global$"50%") - as.numeric(motemp_global$"50%"),1)`$^\circ$C higher than the modern.
-A model run excluding the ecological proxies increases the GMST by `r eoceneproxy_global_diff$"50%"` (`r eoceneproxy_global_diff$"2.5%"` -- `r eoceneproxy_global_diff$"97.5%"`)$^\circ$C.
-The median latitudinal gradient is similar when excluding the ecological proxies, with a median of `r eocene_proxy_gradient$"50%"`$^\circ$C, but with a `r proxy_uncertain_diff_percent`% wider 95% CI (Fig.
+The early Eocene GMST is estimated at `r eotemp_global$"50%"` (`r eotemp_global$"2.5%"` -- `r eotemp_global$"97.5%"`)&deg;C, `r  round(as.numeric(eotemp_global$"50%") - as.numeric(motemp_global$"50%"),1)`&deg;C higher than the modern.
+A model run excluding the ecological proxies increases the GMST by `r eoceneproxy_global_diff$"50%"` (`r eoceneproxy_global_diff$"2.5%"` -- `r eoceneproxy_global_diff$"97.5%"`)&deg;C.
+The median latitudinal gradient is similar when excluding the ecological proxies, with a median of `r eocene_proxy_gradient$"50%"`&deg;C, but with a `r proxy_uncertain_diff_percent`% wider 95% CI (Fig.
 S2).
@@
 Due to the limited spatial coverage of the early Eocene proxy record, and due to the added model complexity of simultaneously estimating a model across both hemispheres, we pooled the proxy data across both hemispheres.
-Applying the model separately within each hemisphere results in substantial differences in hemispherical, average temperatures, with the Southern Hemisphere being warmer by `r eeco_south_north_global_diff$"50%"` (`r eeco_south_north_global_diff$"2.5%"` -- `r eeco_south_north_global_diff$"97.5%"`)$^\circ$C.
-The inferred latitudinal gradient is somewhat steeper in the Northern Hemisphere (steeper by `r eocene_north_south_gradient_diff$"50%"`$^\circ$C, although the 95% CI spans `r eocene_north_south_gradient_diff$"2.5%"` -- `r eocene_north_south_gradient_diff$"97.5%"`$^\circ$C), but the large uncertainties associated with both gradients, and the lack of polar proxy data in the Southern Hemisphere preclude a more precise statement (see Fig.
+Applying the model separately within each hemisphere results in substantial differences in hemispherical, average temperatures, with the Southern Hemisphere being warmer by `r eeco_south_north_global_diff$"50%"` (`r eeco_south_north_global_diff$"2.5%"` -- `r eeco_south_north_global_diff$"97.5%"`)&deg;C.
+The inferred latitudinal gradient is somewhat steeper in the Northern Hemisphere (steeper by `r eocene_north_south_gradient_diff$"50%"`&deg;C, although the 95% CI spans `r eocene_north_south_gradient_diff$"2.5%"` -- `r eocene_north_south_gradient_diff$"97.5%"`&deg;C), but the large uncertainties associated with both gradients, and the lack of polar proxy data in the Southern Hemisphere preclude a more precise statement (see Fig.
 S3).
@@
 A local temperature estimate, based on multiple geochemical proxies, thus has the same weight as a local temperature estimate obtained from the occurrence of a climate-sensitive plant community, whilst preserving the uncertainty associated with each estimate.
-The model could easily be extended to include uncertainties on individual geochemical proxy data, or to differently weight proxy records classified as more or less reliable.
+The model could easily be extended to include uncertainties on individual geochemical proxy data, or to variably weight proxy records classified as more or less reliable.
 
@@
 S2), suggesting that the modern climate range of coral reefs may underestimate the early Eocene thermal niche for coral reefs.
-We have tried to account for that possibility by widening the temperature probability distribution for reef corals, but the predicted temperatures for the reef and mangrove sites still lie below the temperatures indicated by the geochemical proxy record (Fig.
+We have tried to account for that possibility by widening the temperature probability distribution for coral reefs, but the predicted temperatures for the reef and mangrove sites still lie below the temperatures indicated by the geochemical proxy record (Fig.
 4, Fig.
@@
 The geochemical proxy record and ecological data indicate that the latitudinal SST gradient of the early Eocene climatic optimum was significantly shallower than the modern [@Huber2011], but beyond that, there is little agreement.
-Earlier, reconstructed early Eocene and EECO SST gradients range from $7 - 21^\circ$C (Table 1); a more recent reconstruction that includes terrestrial air and sea surface temperatures arrives at a gradient of $13^\circ$C [@Inglis2020].
-Our polar circle to equatorial gradient estimate is lower than most previous estimates at $7.8^\circ$, although the 95% credible interval extends up to $13.7^\circ$ and thus overlaps earlier estimates based on shallow water proxies.
+Earlier, reconstructed early Eocene and EECO SST gradients range from 7 -- 21&deg;C (Table 1); a more recent reconstruction that includes terrestrial air and sea surface temperatures arrives at a gradient of 13&deg;C [@Inglis2020].
+Our polar circle to equatorial gradient estimate is lower than most previous estimates at 7.8&deg;C, although the 95% credible interval extends up to 13.7&deg;C and thus overlaps earlier estimates based on shallow water proxies.
 The confirmation of a very flat gradient by both geochemical and ecological shallow water data indicates that inferred SST gradients based on tropical, shallow water and deep water samples [@Cramwinckel2018; @Evans2018] may overestimate the SST gradient of the early Eocene greenhouse world.
 
-<!-- @Evans2018 estimate the SST gradient at 32+-10% of the modern, which would correspond to \~`r round( as.numeric(modern_gradient$"50%") - 0.32*as.numeric(modern_gradient$"50%"),0)`+- `r format(round( 0.10*as.numeric(modern_gradient$"50%"),0),nsmall=0)`$^\circ$C, using the difference between the mean tropical and deep ocean data.
-The modelled SST of @Tierney2017 using TEX86 is \~ `r round(35.8-23.8,0)`$^\circ$C, when taking the difference between temperatures at the polar circles and the equator.
-@Inglis2020 include both terrestrial air temperature and SST estimates in their model of the latitudinal temperature gradient in their Figure S4a; they model a polar circle to equatorial gradient of \~ `r round(30.7-(20.2+14.3)/2,0)`$^\circ$C.
-All of those estimates are significantly lower than the SST gradient predicted by an EECO climate model ensemble [@Lunt2012, as shown in @Tierney2017], which predicts a polar circle to equatorial gradient of `r round(31.5-5.5,0)`$^\circ$C.
-@Keating-Bitonti2011 show a proxy-based SST gradient of \~ `r round(30-(19+15.6)/2,0)`$^\circ$C, and @Bijl2009 model a gradient of \~`r round(36-27,0)`$^\circ$C in the northern and only \~`r 36-31`$^\circ$C in the southern hemisphere for the early Eocene.
-Our polar circle to equatorial gradient is lower than most previous estimates at `r eocene_gradient_polar_circle$"50%"`$^\circ$C, although the 95% credible interval overlap with several of the listed gradient estimates.
+<!-- @Evans2018 estimate the SST gradient at 32+-10% of the modern, which would correspond to \~`r round( as.numeric(modern_gradient$"50%") - 0.32*as.numeric(modern_gradient$"50%"),0)`+- `r format(round( 0.10*as.numeric(modern_gradient$"50%"),0),nsmall=0)`&deg;C, using the difference between the mean tropical and deep ocean data.
+The modelled SST of @Tierney2017 using TEX86 is \~ `r round(35.8-23.8,0)`&deg;C, when taking the difference between temperatures at the polar circles and the equator.
+@Inglis2020 include both terrestrial air temperature and SST estimates in their model of the latitudinal temperature gradient in their Figure S4a; they model a polar circle to equatorial gradient of \~ `r round(30.7-(20.2+14.3)/2,0)`&deg;C.
+All of those estimates are significantly lower than the SST gradient predicted by an EECO climate model ensemble [@Lunt2012, as shown in @Tierney2017], which predicts a polar circle to equatorial gradient of `r round(31.5-5.5,0)`&deg;C.
+@Keating-Bitonti2011 show a proxy-based SST gradient of \~ `r round(30-(19+15.6)/2,0)`&deg;C, and @Bijl2009 model a gradient of \~`r round(36-27,0)`&deg;C in the northern and only \~`r 36-31`&deg;C in the southern hemisphere for the early Eocene.
+Our polar circle to equatorial gradient is lower than most previous estimates at `r eocene_gradient_polar_circle$"50%"`&deg;C, although the 95% credible interval overlap with several of the listed gradient estimates.
 -->
@@
 Polar temperature estimates from our model are thus conservative in that they admit large uncertainty where data is absent, which is desirable.
-However, the presence of high proxy-derived temperature estimates at \~ 60$^\circ$ latitudes forces the modelled median temperature curve to be too high at \~ 24$^\circ$C, relative to the temperatures indicated by the high-latitude mangrove communities (15.6 - 22.5$^\circ$C).
+However, the presence of high proxy-derived temperature estimates at \~ 60&deg; latitudes forces the modelled median temperature curve to be too high at \~ 24&deg;C, relative to the temperatures indicated by the high-latitude mangrove communities (15.6 - 22.5&deg;C).
 In contrast, the extrapolated polar temperatures of most previous proxy-based models are likely too low, given the abundance of ecological data indicating temperate or subtropical high-latitude climates during the EECO [@Pross2012; @Popescu2021].
@@
 Biases and errors in the proxy reconstructions likely contribute to the observed variability, as geochemical proxies reflect many other factors besides seawater temperature [@Hollis2019].
-Despite excluding &delta;^18^O measurements from recrystallised fossils, systematic offsets remain between mostly warm temperatures derived from $TEX_{86}$, and cooler temperatures derived from &delta;^18^O, &Delta;~47~, and the ecological proxies.
+Despite excluding &delta;^18^O measurements from recrystallised fossils, systematic offsets remain between mostly warm temperatures derived from TEX~86~, and cooler temperatures derived from &delta;^18^O, &Delta;~47~, and the ecological proxies.
 Seasonality [@Keating-Bitonti2011] and temporal changes within the EECO [@Westerhold2018] may also contribute to the large variability of the EECO proxy data.
 
-Recent, marine GMST estimates of the EECO and of the early Eocene range from 23.4 to 37.1$^\circ$C, with the lowest GMSTs being derived from &delta;^18^O, and the higher estimates including $TEX_{86}$ [@Inglis2020].
-Many studies include both marine and terrestrial proxies to derive GMST estimates, but despite great differences in proxy selection and in the calculation of global average temperatures, many recent estimates fall in the range of 27 - 29.5$^\circ$C [@Hansen2013; @Caballero2013; @Cramwinckel2018; @Zhu2019], similar to our median GMST estimate of `r eotemp_global$"50%"`$^\circ$C.
+Recent, marine GMST estimates of the EECO and of the early Eocene range from 23.4 -- 37.1&deg;C, with the lowest GMSTs being derived from &delta;^18^O, and the higher estimates including $TEX_{86}$ [@Inglis2020].
+Many studies include both marine and terrestrial proxies to derive GMST estimates, but despite great differences in proxy selection and in the calculation of global average temperatures, many recent estimates fall in the range of 27 - 29.5&deg;C [@Hansen2013; @Caballero2013; @Cramwinckel2018; @Zhu2019], similar to our median GMST estimate of `r eotemp_global$"50%"`&deg;C.
 
@@
 
-The Bayesian hierarchical model presented herein is able to reconstruct latitudinal gradients from both geochemical and ecological proxy data, whilst reflecting the uncertainty associated with the ecological temperature proxies, and accounts for the variation of multiple temperature estimates at individual localities.
+The Bayesian hierarchical model presented here is able to reconstruct latitudinal gradients from both geochemical and ecological proxy data, while reflecting the uncertainty associated with the ecological temperature proxies, and accounting for the variation of multiple temperature estimates at individual localities.
 Using informative prior information allows for accurate temperature reconstructions from records with geographically incomplete sampling.
 By providing temperature estimates across the entire latitudinal range, this method also facilitates the reconstruction of unbiased global average temperatures.
-Application of our model to the EECO suggests that latitudinal sea surface temperature gradients were shallower than estimated by most earlier, proxy-based studies.
+Application of our model to the EECO suggests that latitudinal sea surface temperature gradients were shallower than estimated by most previous proxy-based studies.
 High-latitude pollen records support this interpretation.
 Our GMST estimate is in good agreement with most existing estimates, indicating that broadly accurate GMST reconstructions are possible even with substantial deviations in the shape of the latitudinal temperature gradient.
-Our new method opens the door for improving the accuracy of proxy-based palaeoclimate reoncstructions and Phanerozoic temperature curves, particularly in intervals with a patchy record, and mitigates the biases incurred from uneven sampling.
-The flexibility of this approach means that our estimates can be quickly and easily updated when new data is made available.
+Our new method opens the door for improving the accuracy of proxy-based palaeoclimate reconstructions and Phanerozoic temperature curves, particularly in intervals with a patchy and unenvenly sampled record.
+Finally, the flexibility of our approach means that estimates can be efficiently updated when new data are made available.
 
@@
 L.A.J.
+and K.E.
 generated the figures.

### 2023-06-01 manuscript/main.qmd
final manuscript corrections

@@
 
-Understanding the long-term evolution of Earth's climate system and contextualising current global warming relies on accurate reconstructions of past climates [@Royer2004; @Burke2018].
+Understanding the long-term evolution of Earth's climate system and contextualising current global warming relies on accurate reconstructions of past climates [@Royer2004; @Burke2018; @Tierney2020].
 Recent advances in the synthesis of palaeoclimate data [e.g.
@@
 We choose the EECO to demonstrate the application of the model due to its significance as the interval with the warmest sustained temperatures of the Cenozoic [@Pross2012], rendering it a potential analogue for extreme climate warming scenarios [@Burke2018].
-Our integrative approach allows us to shed new light on the long-standing dispute on the steepness of the early Eocene temperature gradient [Table 1; @Sloan1990; @Markwick1994; @Huber2011; @Tierney2017; @Inglis2020].
+Our integrative approach allows us to shed new light on the long-standing dispute on the steepness of the early Eocene temperature gradient [Table 1\; @Sloan1990; @Markwick1994; @Huber2011; @Tierney2017; @Inglis2020].
 
@@
 Geochemical climate proxy data were extracted from a latest Paleocene and early Eocene compilation [@Hollis2019].
-This compilation provides data on four different geochemical proxies for reconstructing seawater temperature: &delta;^18^O, &Delta;47, Mg/Ca and $TEX_{86}$.
+This compilation provides data on four different geochemical proxies for reconstructing seawater temperature: &delta;^18^O, &Delta;~47~, Mg/Ca and TEX~86~.
 For our analyses, this dataset was restricted to the EECO (defined as 53.8 -- 49.1 Ma) and samples from the continental shelf.
@@
 During the Eocene, coral communities and reefs expanded across tropical and temperate latitudes, with communities found up to palaeolatitudes of 43 &deg;&nbsp;N [@Zamagni2012].
-Using a compilation of Paleocene--early Eocene coral reefs and community localities [@Zamagni2012], we generated quantitative sea surface temperature estimates for the ECCO.
+Using a compilation of Paleocene -- early Eocene coral reefs and community localities [@Zamagni2012], we generated quantitative sea surface temperature estimates for the ECCO.
 To do so, we extracted localities from the compilation that are inferred to be Ilerdian (early Eocene) coral reefs, and that could be confidently assigned to the EECO.
@@
 towards $\nu_j$.
-The pull towards $\nu_j$ tends to be high when $m$ is low, when the observations $t_{i=1,...,m,~j}$ are scattered, i.e.
+The pull towards $\nu_j$ tends to be strong when $m$ is low, when the observations $t_{i=1,...,m,~j}$ are scattered, i.e.
 $\sigma_j$ is high, and/or when the overall standard deviation $\sigma$ is low.
-In practice, this has the desirable consequence that locations with few observations and large temperature differences between observations have less influence on the overall regression than well-sampled locations with consistent temperature observations.
+In practice, this has the desirable consequence that locations with few observations and large temperature differences between observations have less influence on the overall regression than well-sampled locations with consistent reconstructed temperatures.
 
@@
 The shape and rate parameters $\alpha_G$ and $\beta_G$ were chosen such that the highest prior density is at $B$ of the modern SST gradient, $0.11$.
-We informed the prior distributions on M and B based on a provisional model run with the modern SST data.
+We informed the prior distributions on $M$ and $B$ based on a provisional model run with the modern SST data.
 
@@
 To test whether our logistic regression model can adequately describe different latitudinal temperature gradients at various sample sizes, we generated four idealised gradients that emulate potential climatic states throughout Earth's geological history: extreme icehouse, icehouse, greenhouse, and extreme greenhouse [@Frakes1992].
-We then randomly sampled (1000 iterations) these gradients using increasing sample sizes (5, 10, and 20) and reconstructed the latitudinal temperature gradient using our model for each of these sample sizes and gradient types.
+We then randomly sampled (1,000 iterations) these gradients using increasing sample sizes (5, 10, and 20) and reconstructed the latitudinal temperature gradient using our model for each of these sample sizes and gradient types.
 Using the same idealised gradients, we also tested whether our model could accurately reconstruct latitudinal temperature gradients using the palaeogeographic distribution of Eocene samples (*n* = 34), providing an empirical, exemplary distribution that captures both limited sample size and skewed geographic origins of samples.
@@
 The latitudinal gradient is calculated as the difference between the modelled temperature at the equator (0&deg; latitude) and at the poles (90&deg; absolute latitude).
-To facilitate comparison with earlier estimates, we also calculate the gradient with the temperature at the polar circle (66.6&deg; absolute latitude) being used instead of at the poles.
+To facilitate comparison with earlier estimates, we also calculate the gradient with the temperature at the polar circle (66.6&deg; absolute latitude) being used instead of the temperature at the poles.
 Given the sigmoidal shape of the modern as well as the Eocene gradient (see Fig.
-4), these results are broadly comparable to a gradient inferred from the zonal average of equatorial and high-latitude temperatures, as is done in some earlier studies [@Evans2018].
+4), these results are broadly comparable to a gradient inferred from the zonal average of equatorial and high-latitude temperatures, as has been done in some earlier studies [@Evans2018].
 
@@
 Bold black text within each panel depicts the coefficient of determination ($R^2$) for estimating goodness of fit between the simulated and modelled gradient.
-The median (50%) $R^2$ value along with the 2.5% and 97.5% credible intervals from the 1,000 model runs are shown.
+The median (50%) $R^2$ value along with the 95% credible intervals from all model runs are shown.
 Each gradient is depicted in absolute latitude.
@@
 3); almost all of the variation in the empirical median temperatures in bins of 1&deg; absolute latitude is explained by the modelled gradient ($R^2 = `r R2_modern`$).
-The empirical gradient spans `r modern_gradient_empirical`&deg;C from the equator to the poles, the modelled gradient is only very slightly higher at `r modern_gradient$"50%"`&deg;C.
+The empirical gradient spans `r modern_gradient_empirical`&deg;C from the equator to the poles, the modelled gradient is only slightly higher at `r modern_gradient$"50%"`&deg;C.
 The modern, global mean temperature (GMST) based on our modelled, median gradient is `r motemp_global$"50%"`&deg;C, very similar to the GMST derived from the empirical median gradient (`r motemp_empirical_global`&deg;C).
@@
 Estimates of the median, latitudinal sea surface temperature gradients of the early Eocene climatic optimum (purple line) and of the present-day (turquoise), both estimated with the Bayesian model.
-The ribbons (purple and turquoise shading) depict the 95% credible intervals of these gradients, respectively.
+The purple ribbon (shading) depict the 95% credible interval of the Eocene gradient, the uncertainty of the modern gradient is too low to be visible.
 Points within the plot depict the geochemical (e.g.
-&delta;^18^O) and ecological (e.g.
+TEX~86~) and ecological (e.g.
 mangroves) data.
@@
 
-Our results show that our Bayesian model can be used to reconstruct different types of latitudinal SST gradients from proxy data, even with small sample sizes (*n* = 10 - 20) and patchy sampling distributions (Fig.
+Our results show that our Bayesian model can be used to reconstruct different types of latitudinal SST gradients from proxy data, even with small sample sizes (*n* = 10 -- 20) and patchy sampling distributions (Fig.
 2).
@@
 @Inglis2020].
-This means that the idiosyncrasies of a patchy proxy record, potentially afflicted with measurement errors, calibration errors, and palaeogeographic and temporal uncertainty, dictate the reconstruction of large-scale climate patterns without the option of including additional knowledge (e.g.
+This means that the idiosyncrasies of a patchy proxy record, potentially afflicted with measurement errors, calibration errors, and palaeogeographic and temporal uncertainty, dictate the reconstruction of large-scale climate patterns, without the option of including additional knowledge (e.g.
 that latitudinal temperature gradients should be broadly negative).
@@
 The absence of coral reefs in higher latitudes in the early Eocene could be due to requirements in irradiance, rather than temperature [@Muir2015].
-Tropical temperatures predicted by the geochemical proxy record indicate hotter-than-modern tropical temperatures for the early Eocene (Fig.
-S2), suggesting that the modern climate range of coral reefs may underestimate the early Eocene thermal niche for coral reefs.
+Tropical temperatures predicted by the geochemical proxy record indicate hotter-than-modern tropical temperatures for the early Eocene (Fig.&nbsp;S2), suggesting that the modern climate range of coral reefs may underestimate the early Eocene thermal niche for coral reefs.
 We have tried to account for that possibility by widening the temperature probability distribution for coral reefs, but the predicted temperatures for the reef and mangrove sites still lie below the temperatures indicated by the geochemical proxy record (Fig.
@@
 The geochemical proxy record and ecological data indicate that the latitudinal SST gradient of the early Eocene climatic optimum was significantly shallower than the modern [@Huber2011], but beyond that, there is little agreement.
-Earlier, reconstructed early Eocene and EECO SST gradients range from 7 -- 21&deg;C (Table 1); a more recent reconstruction that includes terrestrial air and sea surface temperatures arrives at a gradient of 13&deg;C [@Inglis2020].
+Earlier, reconstructed early Eocene and EECO SST gradients range from 7 -- 21&deg;C (Table 1); a more recent reconstruction that includes terrestrial air and sea surface temperatures arrives at a gradient of ~13&deg;C [@Inglis2020].
 Our polar circle to equatorial gradient estimate is lower than most previous estimates at 7.8&deg;C, although the 95% credible interval extends up to 13.7&deg;C and thus overlaps earlier estimates based on shallow water proxies.
@@
 
-Recent, marine GMST estimates of the EECO and of the early Eocene range from 23.4 -- 37.1&deg;C, with the lowest GMSTs being derived from &delta;^18^O, and the higher estimates including $TEX_{86}$ [@Inglis2020].
+Recent, marine GMST estimates of the EECO and of the early Eocene range from 23.4 -- 37.1&deg;C, with the lowest GMSTs being derived from &delta;^18^O, and the higher estimates including TEX~86~ [@Inglis2020].
 Many studies include both marine and terrestrial proxies to derive GMST estimates, but despite great differences in proxy selection and in the calculation of global average temperatures, many recent estimates fall in the range of 27 - 29.5&deg;C [@Hansen2013; @Caballero2013; @Cramwinckel2018; @Zhu2019], similar to our median GMST estimate of `r eotemp_global$"50%"`&deg;C.

### 2023-06-01 manuscript/supplementary.qmd
SI, copyright statement in acknowledgements

@@
 crossref:
-  fig-title: Supplementary Figure
-  fig-prefix: Supplementary Figure
+  fig-title: "Figure S"
+  fig-prefix: "Figure S"
 bibliography: references.bib
@@
 
-## S1. Convergence checks
-
 ::: {#fig-S1}
@@
 a) Traceplot of A, K-A, M, B and $\sigma$; b) Traceplot of five selected $\mu_j$; c) Traceplot of five selected $\sigma_j$.
-All traceplots display mixing of the chains, and relatively quick convergence."
+All traceplots display mixing of the chains, and relatively quick convergence.
 :::
 
-## S2. Gradient with just the geochemical proxy data
 ::: {#fig-S2}
@@
 
-Latitudinal temperature gradient with just the geochemical proxies (orange), showing the median (line) and 95% credibal interval (shading).
+Estimated latitudinal temperature gradient using only the geochemical proxy data (orange), showing the median (line) and 95% credible interval (shading).
 Symbols with vertical lines show the median and 95% credible intervals of $\mu_j$.
@@
 
-## S3. Separate gradients on the northern and southern hemisphere
 ::: {#fig-S3}
@@
 
-Latitudinal temperature gradient in seperate hemispheres, showing the medians (lines) and 95% credibal intervals (shadings) in the southern (red) and northern hemisphere (blue).
+Estimated latitudinal temperature gradients for each hemisphere, showing the medians (lines) and 95% credible intervals (shadings) in the Southern (red) and Northern Hemisphere (blue).
 Symbols with vertical lines show the median and 95% credible intervals of $\mu_j$.
-Turquoise symbols in the northern hemisphere highlight the ecological proxy data.
+Turquoise symbols in the Northern Hemisphere highlight the ecological proxy data.
 The grey line and shading in the background show the latitudinal temperature gradient with the data from both hemispheres combined, plotted in both hemispheres.
@@
 Simulated latitudinal temperature gradients sampled at the absolute latitudes of the EECO data.
-The dark grey points show the simulated temperature data, the black line shows the shape of the true gradient, and the medians (lines) and 95% credibal intervals (shadings) of the inferred gradients using the model are shown in red.
+The dark grey points show the simulated temperature data, the black line shows the shape of the true gradient, and the medians (lines) and 95% credible intervals (shadings) of the inferred gradients using the model are shown in red.
 \\\n a) linear gradient; $SST = 30-|latitude|/3$ \\\n b) flat gradient; $SST = 25$ \\\n c) quadratic gradient; $SST = 30-|latitude|^2/270$ \\\n d) as c), but the priors on A, M, and B were relaxed to allow for a better fit to the quadratic gradient in high latitudes \\\n  All latitudinal temperature gradients are well captured by the model in the latitudinal range for which data is provided.

### 2023-06-01 R/figures/SM/FigS3.R

@@ -21,10 +21,10 @@ lats <- mode[[1]]$lat # latitudes of locations is saved in model output
 
 
 # read model output
-mode_north <- readRDS("./results/eeco/eeco_climate_model_northern_hemisphere_output")
-mode_all_north <- readRDS("./results/eeco/eeco_climate_model_northern_hemisphere_output_combined.rds")
-mode_south <- readRDS("./results/eeco/eeco_climate_model_southern_hemisphere_output")
-mode_all_south <- readRDS("./results/eeco/eeco_climate_model_southern_hemisphere_output_combined.rds")
+mode_north <- readRDS("./results/SM/eeco_climate_model_northern_hemisphere_output.rds")
+mode_all_north <- readRDS("./results/SM/eeco_climate_model_northern_hemisphere_output_combined.rds")
+mode_south <- readRDS("./results/SM/eeco_climate_model_southern_hemisphere_output.rds")
+mode_all_south <- readRDS("./results/SM/eeco_climate_model_southern_hemisphere_output_combined.rds")
 
 dat <- readRDS("data/processed/Hollis_processed_EECO_2022_07_19.rds")
 #
@@ -76,26 +76,26 @@ plot_gradient(mode_all_north, add = T, negative = F, line_col = rgb(0,0,0.8), co
 plot_gradient(mode_all_south, add = T, negative = T, line_col = rgb(0.8,0,0), confint_col = rgb(0.8,0,0,0.2))
 
 # add median of the poster estimates of location means
-plot_posterior(mode_north[[1]], col_obs = rgb(0,0,1), col_dist = rgb(0,.8,.8)) # select the first run, doesn't yet work for combined chains
+plot_posterior(mode_north[[1]], col_obs = rgb(0,0,1), col_dist = rgb(0,.75,.72)) # select the first run, doesn't yet work for combined chains
 plot_posterior(mode_south[[1]], col_obs = rgb(1,0,0), col_dist = rgb(1,0,0), negative = TRUE) # select the first run, doesn't yet work for combined chains
 
 axis(1,seq(-90,90,30))
 
-text(-87.5,38.25,"southern hemisphere", col = rgb(.8,0,0), adj = 0)
-text(87.5,38.25,"northern hemisphere", col = rgb(0,0,0.8), adj = 1)
+text(-87.5,38.25,"Southern Hemisphere", col = rgb(.8,0,0), adj = 0)
[…]

### 2023-06-01 manuscript/main.qmd
CC BY licence in acknowledgements
add zenodo link

@@
 was supported by a Juan de la Cierva-formación 2021 fellowship (FJC2021-046695-I/MCIN/AEI/10.13039/501100011033) from the European Union "NextGenerationEU"/PRTR.
+For the purpose of open access, the authors have applied a Creative Commons Attribution (CC BY) licence to any Author Accepted Manuscript version arising from this submission.
 <!-- regarding funding acknowledgement, I talked to my boss and he says I don't need to credit my funding as I have done this work largely in my spare time -->
@@
 
-The data and code used to produce the results of this study are available via GitHub (<https://github.com/KEichenseer/PalaeoClimateGradient>) and the linked Zenodo repository (XXX).
+The data and code used to produce the results of this study are available via GitHub (<https://github.com/KEichenseer/PalaeoClimateGradient>) and the linked Zenodo repository (<https://zenodo.org/record/7995969>).
 

### 2023-08-09 manuscript/main.qmd

@@
 Here, we present a novel Bayesian hierarchical model that combines quantitative proxies and ecological constraints into a fully quantitative model of the latitudinal gradient of sea surface temperatures, bridging spatial gaps in sparsely sampled climate data.
+The Bayesian approach offers a powerful framework for integrating various sources of uncertainty and modelling complex hierarchical relationships, and is increasingly used in palaeoclimatic reconstructions [e.g.
+@Weitzel2019; @Yang2022; @Burgener2023].
 This model expands upon existing, spatially explicit palaeoclimatic reconstructions  by allowing for the integration of (1) prior information based on physical principles and the observed modern sea surface temperature distribution, and of (2) geochemical and ecological climate proxies in a common, quantitative framework.
-We use a generalised logistic function to accurately infer the shape of the temperature gradient despite a patchy latitudinal coverage, and test the robustness of this method using down-sampled, simulated temperature gradients.
+We chose a generalised logistic function to accurately infer the shape of the temperature gradient despite a patchy latitudinal coverage.
+This choice is motivated by the flexibility and ability of this function to approximate a variety of nonlinear patterns in the potential, underlying temperature gradients, that other parametric approaches, such as lower order polynomials [e.g.
+@Keating-Bitoni2011; @Bijl2009], lack.
+We test the robustness of this method using down-sampled, simulated temperature gradients.
 
@@
 the latitude around which temperature falls most steeply with latitude, $B$ denotes the growth rate, $\sigma$ denotes the residual standard deviation, and $n$ denotes the number of locations.
+
+<!-- Add justification of logistic function -->
 

### 2023-08-09 manuscript/main.qmd
replies to reviewer comments, sensitivity ecological proxies

@@
 We chose a generalised logistic function to accurately infer the shape of the temperature gradient despite a patchy latitudinal coverage.
-This choice is motivated by the flexibility and ability of this function to approximate a variety of nonlinear patterns in the potential, underlying temperature gradients, that other parametric approaches, such as lower order polynomials [e.g.
-@Keating-Bitoni2011; @Bijl2009], lack.
+This choice is motivated by the flexibility and ability of this function to approximate a variety of nonlinear patterns in the underlying temperature gradients, that other parametric approaches, such as lower order polynomials [e.g.
+@Bijl2009; @Keating-Bitoni2011], lack.
 We test the robustness of this method using down-sampled, simulated temperature gradients.
@@
 
-$$\nu_{j} = A~+\frac{K-A}{e^{B(abs(l_j)-M)}}~, ~~~~~ j = 1,...,n,  $${#eq-nu} where $A$ and $K$ denote the lower and upper asymptote, respectively, $M$ specifies the latitude of maximal growth, i.e.
+$$\nu_{j} = A~+\frac{K-A}{1+e^{B(abs(l_j)-M)}}~, ~~~~~ j = 1,...,n,  $${#eq-nu} where $A$ and $K$ denote the lower and upper asymptote, respectively, $M$ specifies the latitude of maximal growth, i.e.
 the latitude around which temperature falls most steeply with latitude, $B$ denotes the growth rate, $\sigma$ denotes the residual standard deviation, and $n$ denotes the number of locations.
 
-<!-- Add justification of logistic function -->
+We use this generalised logistic function because it can follow the equatorial and polar asymptotes observed in the modern, latitudinal SST gradient, but can also accommodate a variety of other shapes, while consisting of only four shape parameters.
+This flexibility is primarily achieved by shifting the location of the curve along the latitudinal axis by varying $M$, and by altering the steepness of the curve by varying $B$.
+For example, one limb of a second-order polynomial as in @Bijl2009 can be approximated by increasing $M$ towards high latitudes, and decreasing $B$ to reduce the steepness of the curve.
 

### 2023-08-09 R/analyses/SM/run_EECO_climate_model_vary_ecological_constraints.R

@@ -1,0 +1,79 @@
+#### Run the climate model for the Early Eocene Climate Optimum
+####
+# source options for analysis
+source("./R/options.R")
+#
+### Read data
+## Hollis data
+dat <- readRDS("data/processed/Hollis_processed_EECO_2022_07_19.rds")
+#
+# prepare for use in the model
+obsmat <- data.frame(sample = (paste(abs(dat$p_lat),dat$longitude, dat$proxy)),
+                     p_lat = abs(dat$p_lat), temperature = dat$temperature,
+                     sd = dat$temperature_sd,
+                     proxy = dat$proxy)
+# order observation matrix for easily keeping tracks of mean estimates
+obsmat <- obsmat[with(obsmat, order(p_lat, sample)),]
+
+
+### Mangrove and Coral data
+# read and assign proxy type
+bioprox <- readRDS("data/processed/bio_proxies_2022_08_08.RDS")
+bioprox$proxy <- rep("Reef",nrow(bioprox))
+bioprox$proxy[which(bioprox$taxa=="Avicennia")] <- "Avicennia"
+bioprox$proxy[which(bioprox$type=="mangrove" & bioprox$taxa!="Avicennia")] <- 
+  "Avicennia-Rhizophoraceae"
+
+# create distribution matrix for use with model (distributions defined in options)
+proxy_index <- sapply(bioprox$proxy, function(f) which(proxy_distributions$name==f))
+
+  
+wider_coral_mean <- mean(c(35.6,16.0))
+wider_coral_sd <- round(1/(1.96*2) * (35.6-16.0),2)
+proxy_distributions$mean[3] <- wider_coral_mean
+proxy_distributions$sd[3] <- wider_coral_sd
+
+wider_avi_rhi_mean <- wider_coral_mean
+wider_avi_rhi_sd <- wider_coral_sd
+proxy_distributions$mean[2] <- wider_avi_rhi_mean
+proxy_distributions$sd[2] <- wider_avi_rhi_sd
+
+distrmat = data.frame(p_lat = abs(bioprox$p_lat), 
+                      mu = proxy_distributions$mean[proxy_index],
+                      scale = proxy_distributions$sd[proxy_index],
+                      shape = proxy_distributions$shape[proxy_index],
+                      distribution = proxy_distributions$distribution[proxy_index])
+
+### Prepare for model run
+# set up clusters for parallel computing
+library(doParallel)
+nClust <- n_chains # from the options script
[…]

### 2023-08-09 R/figures/SM/FigS2.R

@@ -1,4 +1,4 @@
-# Supplementary Figure S2 - gradient on Northern and Southern hemisphere
+# Supplementary Figure S2 - gradient with just geochem.
 
 # load figures for plotting, distributions ect.
 source("./R/functions/auxiliary_functions.R")

### 2023-08-09 R/figures/SM/figS-wider-ecol-limits.R

@@ -1,0 +1,88 @@
+# Supplementary Figure S2 - gradient with just geochem.
+source("./R/options.R")
+
+# load figures for plotting, distributions ect.
+source("./R/functions/auxiliary_functions.R")
+source("./R/functions/model_components/gradient.R")
+# read model output
+mode <- readRDS("./results/eeco/eeco_climate_model_output.rds")
+mode_all <- readRDS("./results/eeco/eeco_climate_model_output_combined.rds")
+mode_ecowide <- readRDS("./results/SM/eeco_climate_model_wider_ecol.rds")
+mode_all_ecowide <-  readRDS("./results/SM/eeco_climate_model_wider_ecol_combined.rds")
+
+# read the input data
+obsmat <- readRDS("./data/processed/observation_matrix_for_EECO_model.rds")
+distrmat <- readRDS("./data/processed/distribution_matrix_for_EECO_model.rds")
+
+
+wider_coral_mean <- mean(c(35.6,16.0))
+wider_coral_sd <- round(1/(1.96*2) * (35.6-16.0),2)
+proxy_distributions$mean[3] <- wider_coral_mean
+proxy_distributions$sd[3] <- wider_coral_sd
+
+wider_avi_rhi_mean <- wider_coral_mean
+wider_avi_rhi_sd <- wider_coral_sd
+proxy_distributions$mean[2] <- wider_avi_rhi_mean
+proxy_distributions$sd[2] <- wider_avi_rhi_sd
+
+distrmat2 = data.frame(p_lat = abs(bioprox$p_lat), 
+                      mu = proxy_distributions$mean[proxy_index],
+                      scale = proxy_distributions$sd[proxy_index],
+                      shape = proxy_distributions$shape[proxy_index],
+                      distribution = proxy_distributions$distribution[proxy_index])
+
+
+
+# calculate location means of observation input data for easier plotting
+ylist <- lapply(unique(obsmat$sample), function(x) obsmat$temperature[which(obsmat$sample == x)])
+yobs_mean= c(sapply(ylist,mean))
+n_obs <- length(yobs_mean)
+# add distribution means
[…]

### 2023-08-10 manuscript/main.qmd
more SM, reviewer replies

@@
 Geochemical climate proxy data were extracted from a latest Paleocene and early Eocene compilation [@Hollis2019].
-This compilation provides data on four different geochemical proxies for reconstructing seawater temperature: &delta;^18^O, &Delta;~47~, Mg/Ca and TEX~86~.
-For our analyses, this dataset was restricted to the EECO (defined as 53.8 -- 49.1 Ma) and samples from the continental shelf.
+This compilation provides sea surface temperature data on four different geochemical proxies for reconstructing seawater temperature: &delta;^18^O, &Delta;~47~, Mg/Ca and TEX~86~.
+For our analyses, this dataset was restricted to the EECO (defined as 53.8 -- 49.1 Ma) and samples originating from near the ocean surface or mixed layer.
+Samples labelled as "thermocline" or "sub-thermocline" were excluded.
 Recrystallised &delta;^18^O samples were also excluded as secondary diagenetic calcite precipitated after deposition can bias isotope measurements and offset temperature values [@Schrag1999].

### 2023-08-10 R/data_preparation/01_data_wrangle_hollis2019.R

@@ -215,8 +215,9 @@ for(i in 1:nrow(locpar)) {
 ### EECO could be 53.3 - 49.2 Ma (according to the figures in Hollis et al 2019)
 ### However, data between range(as.numeric(dat$Age[which(dat$Timeslice=="EECO")]),na.rm=T)
 ### i.e. from 48.57246 to 53.48746 has been assigned to EECO in their database. Hence, we use that
+# not needed anymore, age selection will be done in 02_prepare_hollis2019.R:
 #dat$Timeslice[which(is.na(dat$Timeslice) & (as.numeric(dat$Age) >= 48.57246 | as.numeric(dat$Age) <= 53.48746))] <- "EECO"
-# DON'T DO THIS IT IS WRONG, WILL BE TAKEN CARE OF IN THE SYNTHESIZE SCRIPT
+# 
 
 saveRDS(dat,"data/processed/Hollis_2019_TEX86_2022_07_18.rds")
 

### 2023-08-10 R/figures/SM/FigS-latitude_uncertainty.R

@@ -1,0 +1,18 @@
+### Read data
+## Hollis data
+dat <- readRDS("data/processed/Hollis_processed_EECO_2022_07_19.rds")
+#
+# prepare for use in the model
+obsmat <- data.frame(sample = (paste(abs(dat$p_lat),dat$longitude, dat$proxy)),
+                     p_lat = abs(dat$p_lat), temperature = dat$temperature,
+                     sd = dat$temperature_sd,
+                     proxy = dat$proxy)
+# order observation matrix for easily keeping tracks of mean estimates
+obsmat <- obsmat[with(obsmat, order(p_lat, sample)),]
+
+plot(obsmat$p_lat,obsmat$sd)
+summary(lm(obsmat$sd~obsmat$p_lat))
+
+loc_sd <- sapply(unique(obsmat$sample), function(x) sd(obsmat$temperature[which(obsmat$sample==x)]))
+loc_lat <- sapply(unique(obsmat$sample), function(x) unique(obsmat$p_lat[which(obsmat$sample==x)]))
+plot(loc_lat,loc_sd, ylab = "sd of temperature estimates", )

### 2023-08-13 manuscript/supplementary.qmd
revised simulations

@@
 
+```{r results block - for in-line code, echo = FALSE}
+setwd('..') # move up to parent folder of /manuscript
+
+# sdy
+eeco_sdy <- readRDS("results/eeco/eeco_sdy.rds")
+eeco_sd_sdy <- readRDS("results/SM/eeco_sd_sdy.rds")
+
+```
+
+# EECO gradient with uncertainty on proxy observations 
+Most of the geochemical proxy data used in the EECO analysis (300 of 308) come with uncertainties around their temperature estimates.
+Due to their negligible influence on the model results, we have not included them in the main analysis, but we show a comparison between the main EECO model output (see Fig.
+4) and an the results of an expanded model, including uncertainties, below.
+
+Specifically, we took the 95 or 90% confidence intervals of the temperature estimates to calculate the standard deviation representative of this uncertainty ($t_{\sigma,i,j}$), which we used for the analysis.
+Instead of using fixed mean proxy temperature observations $t_{i,j}$ directly in Equation 3, we let $t_{i,j}$ be the function of a normal distribution with mean $t_{\mu,i,j}$ and standard deviation $t_{\sigma,i,j}$, where $t_{\mu,i,j}$ and $t_{\sigma,i,j}$ are provided by the proxy data set:
+
+$$t_{\mu,i,j}  \sim N(t_{i,j}, t_{\sigma,i,j}), ~~~~~ i = 1,...,m, ~~~~ j = 1,...,n, $${#eq-obs} where $m$ is the number of observations at each location, and $n$ denotes the number of locations.
+
+The resulting gradient estimated with this expanded model is very similar to that estimated with the original model (Fig.
+S2).
+The uncertainty around the proxy temperatures allows the estimated location mean temperatures to be drawn closer to the gradient line, resulting in a slightly lower median residual standard deviation (`r eeco_sd_sdy$"50%"` as opposed to `r eeco_sdy$"50%"` in the original model).
+
+::: {#fig-S2}
+![](../figures/SM/FigS_sd_included.png){width="10.5cm"}
+
+Estimated latitudinal temperature gradient using only the geochemical proxy data (orange), showing the median (line) and 95% credible interval (shading).
+Symbols with vertical lines show the median and 95% credible intervals of $\mu_j$.
+The blue line and shading in the background show the latitudinal temperature gradient with the geochemical and ecological proxy data, as in Figure 4.
+:::
+
 ::: {#fig-S2}

### 2023-08-13 R/analyses/SM/run_EECO_climate_model_with_proxy_sd.R

@@ -67,3 +67,7 @@ mode_all_sd <- combine_posterior(mode_sd,100000)
 saveRDS(mode_all_sd, "results/SM/eeco_climate_model__with_observation_sd_output_combined.rds")
 
 ### To check results use the "processing/assess_EECO-climate_model_output_script.R
+
+# sdy
+eeco_sd_sdy <- as.list(format(round(quantile(mode_all_sd$sdy,probs = c(.025,.5,.975)),1),nsmall = 1))
+saveRDS(eeco_sd_sdy, "results/SM/eeco_sd_sdy.rds")

### 2023-08-13 R/analyses/processing/simulated_gradients_R_squared_revised.R

@@ -1,0 +1,156 @@
+### goodness of fit of the modelled, simulated gradients
+
+library(dplyr)
+
+source("R/functions/model_components/gradient.R")
+source("R/functions/auxiliary_functions.R")
+
+# read simulated gradient models
+four_gradients_model <- readRDS("D://OneDrive - Durham University/projects/four_gradients_revised_model_proc.rds")
+
+lats <- seq(0,90,1)
+grad <- gradient(lats, four_gradients_model[[1]][[1]])
+
+gradient_formulas_static <- rev(list( function(x) 30-x^2/1080, # extreme greenhouse
+                                      function(x) 30-x^2/360, 
+                                      function(x) {
+                                        x[which(x>=90)] <- 89.5
+                                        modern_1deg[sapply(floor(x)+0.5,function(y) which(y==modern_1deg[,"lat"])),2]
+                                      }, #gradient(x, c(-1.775, 30.115, 42.070,  0.106)), # now use modern gradient instead # 11.25*atan(-0.1*x+3.5)+15.25, 
+                                      function(x) 8.75*atan(-0.5*x+15)+12)) # extreme icehouse
+
+
+plot(lats,gradient_formulas_static[[1]](lats), type = "l")
+points(lats, grad[30000,], col = "orange")
+
+
+# basic R squared
+
+R_squared <- list(NULL)
+
+for(g in 1:4) {
+  R_squared[[g]] <- list(NULL)
+  for(s in 1:4) {
+    grad <- gradient(lats, four_gradients_model[[g]][[s]])
+    true_grad <- gradient_formulas_static[[g]](lats)
+    SSres <- apply(grad,1,function(x) sum((x-true_grad)^2))
+    SStot <- sum((true_grad-mean(true_grad))^2)
+    R_squared[[g]][[s]] <- 1-SSres/SStot
+  }
+}
+
+### Try weighted R squared (by latitudinal area)
+
+lats <- seq(0.5,89.5,1)
+alpha1 <- seq(1,90,1)
+alpha2 <- seq(0,89,1)
+latweight <- sin(pi*alpha1/180) - sin(pi*alpha2/180) # sums to 1 - careful when using lat. subsets
+
+R_squared_weighted <- list(NULL)
+
+for(g in 1:4) {
[…]

### 2023-08-13 R/analyses/run_simulated_gradients_revision2.R

@@ -1,0 +1,426 @@
+### Supplementary figure: different gradient shapes
+
+library(dplyr)
+
+# source functions
+source("R/functions/model_components/climate_model_EECO.R")
+source("R/functions/model_components/gradient.R")
+source("R/functions/model_processing/combine_posterior.R")
+source("R/functions/auxiliary_functions.R")
+# source options for analysis
+source("R/options.R")
+
+## Hollis data
+dat <- readRDS("data/processed/Hollis_processed_EECO_2022_07_19.rds")
+#
+# prepare for use in the model
+obsmat <- data.frame(sample = (paste(abs(dat$p_lat),dat$longitude, dat$proxy)),
+                     p_lat = abs(dat$p_lat), temperature = dat$temperature,
+                     sd = dat$temperature_sd,
+                     proxy = dat$proxy)
+# order observation matrix for easily keeping tracks of mean estimates
+obsmat <- obsmat[with(obsmat, order(p_lat, sample)),]
+
+location_sample_size <- as.numeric(c(table(obsmat$sample), c(rep(1,11))))
+
+eocene_lat <- c(-59.42, -59.10, -54.13, -47.84, -47.84, -47.47, -46.14, -45.80, -45.80, -24.25, -8.57, -2.05, -1.86, 0.49, 4.29, 14.67, 27.44, 31.00, 32.50, 32.80, 32.90, 35.54, 36.22, 39.02, 40.39, 40.63, 40.89, 43.79, 44.11, 50.04, 50.39, 58.45, 76.45, 77.13)
+# 
+generate_lat_samples <- function(n) 180/pi * asin(runif(n,0,1))
+
+# sample eocene lats
+
+generate_lat_samples <- function(n) sample(eocene_lat,n,replace = F)
+
+set.seed(1)
+
+
+
+gradient_formulas_static <- rev(list( function(x) 30-x^2/1080, # extreme greenhouse
+                               function(x) 30-x^2/360, 
+                               function(x) {
+                                 x[which(x>=90)] <- 89.5
+                                 modern_1deg[sapply(floor(x)+0.5,function(y) which(y==modern_1deg[,"lat"])),2]
[…]

### 2023-08-13 R/figures/SM/FigS-include-proxy-SD.R

@@ -17,24 +17,6 @@ obsmat <- readRDS("./data/processed/observation_matrix_for_EECO_model.rds")
 distrmat <- readRDS("./data/processed/distribution_matrix_for_EECO_model.rds")
 
 
-wider_coral_mean <- mean(c(35.6,16.0))
-wider_coral_sd <- round(1/(1.96*2) * (35.6-16.0),2)
-proxy_distributions$mean[3] <- wider_coral_mean
-proxy_distributions$sd[3] <- wider_coral_sd
-
-wider_avi_rhi_mean <- wider_coral_mean
-wider_avi_rhi_sd <- wider_coral_sd
-proxy_distributions$mean[2] <- wider_avi_rhi_mean
-proxy_distributions$sd[2] <- wider_avi_rhi_sd
-
-distrmat2 = data.frame(p_lat = abs(bioprox$p_lat), 
-                       mu = proxy_distributions$mean[proxy_index],
-                       scale = proxy_distributions$sd[proxy_index],
-                       shape = proxy_distributions$shape[proxy_index],
-                       distribution = proxy_distributions$distribution[proxy_index])
-
-
-
 # calculate location means of observation input data for easier plotting
 ylist <- lapply(unique(obsmat$sample), function(x) obsmat$temperature[which(obsmat$sample == x)])
 yobs_mean= c(sapply(ylist,mean))
@@ -74,7 +56,7 @@ plot_posterior(mode_sd[[1]], col_obs = rgb(0,0.6,1,1), col_dist = NA, add=T)# li
 plot_posterior(mode[[1]], col_obs = rgb(0,0,0,0.5), col_dist = NA, add=T,
                pch = 19)# line_col = rgb(0,0,1,0.5), confint_col = rgb(0,0,1,0.1))
 
-legend("topright", legend = c("original analysis", "with observation sd"), fill = c(rgb(0,0,0,0.75), rgb(0,0.6,1,1)),
-       bty = "n")
+legend("topright", legend = c("original analysis", "with proxy uncertainty"), fill = c(rgb(0,0,0,0.75), rgb(0,0.6,1,1)),
+       bty = "n", col = NA)
 
 dev.off()

### 2023-08-13 R/figures/fig_2_revised.R

@@ -1,0 +1,71 @@
+# Load libraries --------------------------------------------------------
+library(ggplot2)
+library(ggthemes)
+library(ggtext)
+# Read data -------------------------------------------------------------
+df <- readRDS("./results/simulation/simulated_gradients_temperature_estimates_with_noise_eocene_resampled.rds")
+
+# Rename sample size
+df$sample_size[which(df$sample_size == "eocene")] <- "34 (all Eocene samples)"
+# Rename gradients
+df$gradient_type[which(df$gradient_type == "extreme_greenhouse")] <- "Extreme Greenhouse"
+df$gradient_type[which(df$gradient_type == "greenhouse")] <- "Greenhouse"
+df$gradient_type[which(df$gradient_type == "icehouse (modern)")] <- "Icehouse (modern)"
+df$gradient_type[which(df$gradient_type == "extreme_icehouse")] <- "Extreme Icehouse"
+# Set factor levels
+df$gradient_type <- factor(x = df$gradient_type, levels = c("Extreme Icehouse",
+                                                            "Icehouse (modern)",
+                                                            "Greenhouse",
+                                                            "Extreme Greenhouse"))
+df$sample_size <- factor(x = df$sample_size, levels = c(5, 10, 20, "34 (all Eocene samples)"))
+
+# Get R2 values
+r2 <- readRDS("./results/simulation/simulated_gradients_r_squared_gelman_unweighted_revised.rds")
+# Rename gradients
+r2$gradient_type[which(r2$gradient_type == "extreme_greenhouse")] <- "Extreme Greenhouse"
+r2$gradient_type[which(r2$gradient_type == "greenhouse")] <- "Greenhouse"
+r2$gradient_type[which(r2$gradient_type == "icehouse")] <- "Icehouse (modern)"
+r2$gradient_type[which(r2$gradient_type == "extreme_icehouse")] <- "Extreme Icehouse"
+# Set factor levels
+r2$gradient_type <- factor(x = r2$gradient_type, levels = c("Extreme Icehouse",
+                                                            "Icehouse (modern)",
+                                                            "Greenhouse",
+                                                            "Extreme Greenhouse"))
+# Rename sample size
+r2$sample_size[which(r2$sample_size == 34)] <- "34 (all Eocene samples)"
[…]

### 2023-08-14 manuscript/main.qmd
review updates

@@
 
-To test whether our logistic regression model can adequately describe different latitudinal temperature gradients at various sample sizes, we generated four idealised gradients that emulate potential climatic states throughout Earth's geological history: extreme icehouse, icehouse, greenhouse, and extreme greenhouse [@Frakes1992].
-We then randomly sampled (1,000 iterations) these gradients using increasing sample sizes (5, 10, and 20) and reconstructed the latitudinal temperature gradient using our model for each of these sample sizes and gradient types.
-Using the same idealised gradients, we also tested whether our model could accurately reconstruct latitudinal temperature gradients using the palaeogeographic distribution of Eocene samples (*n* = 34), providing an empirical, exemplary distribution that captures both limited sample size and skewed geographic origins of samples.
+To test whether our logistic regression model can adequately describe different latitudinal temperature gradients at various sample sizes, we used the empirical, modern gradient, representative of an icehouse climate, and generated three idealised gradients that emulate potential climatic states throughout Earth's geological history: extreme icehouse, icehouse (modern), greenhouse, and extreme greenhouse [@Frakes1992].
+The idealised gradients serve to test whether our model setup is able to infer gradients that are strongly different from the modern from a varying amount of samples.
+
+We created test data from these gradients as follows: To capture both limited sample size and skewed geographic origins of samples, we randomly sampled (200 iterations) the latitudes of the EECO data set (n = 34) used in this study, at sample sizes of 5, 10 and 20, and also used the full set of 34 latitudes.
+For each latitude, we inferred a location mean temperature from the gradients, adding random noise from a normal distribution with a standard deviation of 2.2, which is the residual standard deviation of modern sea surface temperature averages from the modelled modern gradient (see below).
+We then randomly allocated a sample size $s$ from the sample sizes present in the EECO data set to each location, and randomly generated $s$ temperature data from a normal distribution with mean equal to the previously generated location mean temperatures, and a standard deviation of 3.8.
+This standard deviation is the average standard deviation of the EECO geochemical proxy data, and with that we aim to simulate randomly distributed errors in the proxy data.
+This approach should give a fair assessment of how the model setup can be affected by the known unknowns (quantifiable errors), but we acknowledge that it cannot quantify the potential impact of unknown, systematic offsets that may bias all proxy data from one location in the same direction.
+
 To evaluate how well the model performed in reconstructing the idealised gradients from limited sampling, we calculated the coefficient of determination ($R^2$) for Bayesian regression models [@Gelman2019].

### 2023-08-14 R/analyses/SM/run_EECO_climate_model_different_prior.R

@@ -1,0 +1,87 @@
+#### Run the climate model for the Early Eocene Climate Optimum
+####
+# source options for analysis
+source("./R/options.R")
+#
+# change prior on M
+priors$f3 <- function(x,log) dunif(x, min = 0, max = 90, log = log) # dnorm(x, mean = 42, sd = 25, log = log) #
+# change prior on K
+priors$f2 <- function(x,lower,log) dtnorm(x, lower, upper = Inf, mean = 28, sd = 20, log = log)
+
+### Read data
+## Hollis data
+dat <- readRDS("data/processed/Hollis_processed_EECO_2022_07_19.rds")
+#
+# prepare for use in the model
+obsmat <- data.frame(sample = (paste(abs(dat$p_lat),dat$longitude, dat$proxy)),
+                     p_lat = abs(dat$p_lat), temperature = dat$temperature,
+                     sd = dat$temperature_sd,
+                     proxy = dat$proxy)
+# order observation matrix for easily keeping tracks of mean estimates
+obsmat <- obsmat[with(obsmat, order(p_lat, sample)),]
+
+### Mangrove and Coral data
+# read and assign proxy type
+bioprox <- readRDS("data/processed/bio_proxies_2022_08_08.RDS")
+bioprox$proxy <- rep("Reef",nrow(bioprox))
+bioprox$proxy[which(bioprox$taxa=="Avicennia")] <- "Avicennia"
+bioprox$proxy[which(bioprox$type=="mangrove" & bioprox$taxa!="Avicennia")] <- 
+  "Avicennia-Rhizophoraceae"
+
+# create distribution matrix for use with model (distributions defined in options)
+proxy_index <- sapply(bioprox$proxy, function(f) which(proxy_distributions$name==f))
+
+distrmat = data.frame(p_lat = abs(bioprox$p_lat), 
+                      mu = proxy_distributions$mean[proxy_index],
+                      scale = proxy_distributions$sd[proxy_index],
+                      shape = proxy_distributions$shape[proxy_index],
+                      distribution = proxy_distributions$distribution[proxy_index])
+
+### Prepare for model run
+# set up clusters for parallel computing
+library(doParallel)
+nClust <- n_chains # from the options script
+cl <- parallel::makeCluster(nClust)
+doParallel::registerDoParallel(cl)
+
+### Run model
[…]

### 2023-08-14 R/analyses/SM/run_modern_climate_model_hemispheres.R

@@ -1,0 +1,87 @@
+#### Run a non-hierarchical climate model on the modern data
+#
+# source options for analysis
+source("R/options.R")
+#
+### load climate data
+# Mean annual sea surface temperatures (Bio-Oracle)
+set.seed(1)
+
+sst <- raster::raster("./data/raw/climate/BioOracle_20220711/Present.Surface.Temperature.Mean.asc")
+r <- raster::raster(res = 1)
+sst <- raster::resample(x = sst, y = r)
+
+sst <- raster::as.data.frame(x = sst, xy = TRUE, centroids = TRUE)
+sst <- na.omit(sst)
+
+temp <- sst$Present.Surface.Temperature.Mean
+
+# use absolute latitudes
+lat <- abs(sst$y)
+
+sst_north <- subset(sst,y > 0)
+sst_south <- subset(sst,y < 0)
+
+
+# set up clusters for parallel computing
+library(doParallel)
+nClust <- 4
+cl <- parallel::makeCluster(nClust)
+doParallel::registerDoParallel(cl)
+
+nChains = nClust
+### Run model
+# source script
+mod_north <- foreach(pc = 1:nChains) %dopar% {
+  # call model functions
+  source("R/functions/model_components/climate_model_modern.R")
+  # set random seed
+  set.seed(pc)
+  run_MCMC_simple(n_iter = 60000,
+                  n_thin = 10,
+                  x = sst_north$y,
+                  y = sst_north$Present.Surface.Temperature.Mean,
+                  prior_input = priors,
+                  adapt_sd = 5000)
+}
+                
+### stop cluster
+parallel::stopCluster(cl)
+
+cl <- parallel::makeCluster(nClust)
+doParallel::registerDoParallel(cl)
+
+nChains = nClust
+### Run model
+# source script
+mod_south <- foreach(pc = 1:nChains) %dopar% {
+  # call model functions
+  source("R/functions/model_components/climate_model_modern.R")
+  # set random seed
+  set.seed(pc)
+  run_MCMC_simple(n_iter = 60000,
+                  n_thin = 10,
+                  x = abs(sst_south$y),
+                  y = sst_south$Present.Surface.Temperature.Mean,
+                  prior_input = priors,
+                  adapt_sd = 5000)
+}
+
+### stop cluster
+parallel::stopCluster(cl)
+
+
+
+# discard 10k iterations as burnin each, and combine the results of the 4 chains
[…]

### 2023-08-18 manuscript/main.qmd
manuscript updates following reviewer comments

@@
 
-Whilst geochemical proxy data can provide robust estimates of palaeotemperature at local scales, recent work has demonstrated that spatial biases in the geochemical proxy record can lead to spurious estimates of regional (e.g.
+Acknowledging the assumptions and limitations inherent in geochemical temperature proxies, e.g.
+experimentally derived calibrations, influences from seasonality, dissolution effects and differential preservation [e.g.
+@Tierney2017; ], can allow for robust estimates of palaeotemperatures at a local scale.
+However, recent work has demonstrated that spatial biases in the geochemical proxy record can lead to spurious estimates of regional (e.g.
 latitudinal temperature gradients) and global temperatures [@judd2020; @Jones2022].
@@
 For example, one limb of a second-order polynomial as in @Bijl2009 can be approximated by increasing $M$ towards high latitudes, and decreasing $B$ to reduce the steepness of the curve.
+The model is designed for modelling the average gradient across both hemispheres, but can also be applied to individual hemispheres, to assess hemispherical differences (Supplementary Material).
 
@@
 
-The modelled Eocene temperature gradient is starkly different from the modern (Fig 4).
+The Eocene temperature gradient reconstructed with our Bayesian model is starkly different from the modern (Fig 4).
 Modelled, median equatorial temperatures are `r tdiff_lat0$"50%"` (95% CI: `r tdiff_lat0$"2.5%"` -- `r tdiff_lat0$"97.5%"`)&deg;C higher for the EECO, and polar temperatures are `r tdiff_lat90$"50%"` (`r tdiff_lat90$"2.5%"` -- `r tdiff_lat90$"97.5%"`)&deg;C higher.

### 2023-09-14 R/figures/fig_1_revised.R
update fig 1 and 2

@@ -1,0 +1,74 @@
+# Load libraries ---------------------------------------------------------
+library(sf)
+library(ggplot2)
+library(ggthemes)
+# Read data --------------------------------------------------------------
+chem <- readRDS("./data/processed/Hollis_processed_EECO_2022_07_19.rds")
+bio <- readRDS("./data/processed/bio_proxies_2022_08_08.RDS")
+# Merdith2021 shapefiles
+coastline <- "https://gws.gplates.org/reconstruct/coastlines/?time=51&model=MERDITH2021"
+coastline <- sf::read_sf(coastline)
+#Data prep ---------------------------------------------------------------
+# Filter data
+chem <- chem[, c("p_lng", "p_lat", "proxy")]
+chem$shape <- 1
+bio <- bio[, c("p_lng", "p_lat", "type")]
+bio$shape <- 2
+# Rename cols 
+names(chem) <- c("p_lng", "p_lat", "proxy", "shape")
+names(bio) <- c("p_lng", "p_lat", "proxy", "shape")
+# Unique data
+bio <- unique(bio)
+chem <- unique(chem)
+# Bind data
+locs <- rbind.data.frame(chem, bio)
+
+# Set factors
+levels <- c("reef", "mangrove", "d18O", "d47", "TEX86", "MgCa")
+locs$proxy <- factor(locs$proxy, levels = levels)
+
+# Rename data for plotting
+locs[which(locs$type == "reef"), c("type")] <- "Coral reef"
+locs[which(locs$type == "mangrove"), c("type")] <- "Mangrove"
+locs[which(locs$type == "d18O"), c("type")] <- expression(delta^18*"O")
+locs[which(locs$type == "d47"), c("type")] <- expression(Delta[47])
+locs[which(locs$type == "TEX86"), c("type")] <- expression(TEX[86])
+locs[which(locs$type == "MgCa"), c("type")] <- "Mg/Ca"
+
+locs <- sf::st_as_sf(x = locs, coords = c("p_lng", "p_lat"), crs = sf::st_crs(4326))
+
+# Plot ------------------------------------------------------------------------
+# Labels
+labs <- c("Coral reef", "Mangrove", 
+          expression(delta^18*"O"), 
+          expression(Delta[47]), 
+          expression(TEX[86]),
+          "Mg/Ca")
+# Shapes
+#shps <- c(21, 22, 23, 24, 25, 19, 20)
+# Plot
+
+  ggplot() +
[…]

### 2023-09-15 manuscript/main.qmd
update fig 1 and 2
revisions

@@
 ::: {#fig-1}
-![](../figures/fig_1.png){width="15cm"}
+![](../figures/fig_1_revised.png){width="15cm"}
 
@@
 In practice, this has the desirable consequence that locations with few observations and large temperature differences between observations have less influence on the overall regression than well-sampled locations with consistent reconstructed temperatures.
+
+We show an expanded model that includes uncertainties on individual temperature observations in the Supplementary Material.
 
@@
 # modern R2
-R2_modern <- readRDS("results/modern/modern_model_empirical_R2_gelman.rds")
+R2_modern_median <- readRDS("results/modern/modern_model_empirical_R2_gelman.rds")
+R2_modern <- readRDS("results/modern/Rsquared_modern.rds")
 
@@
 ::: {#fig-2}
-![](../figures/fig_2.png){width="15cm"}
+![](../figures/fig_2_revised_final.png){width="15cm"}
 
@@
 ::: {#fig-3}
-![](../figures/fig_3.png){width="10cm"}
+![](../figures/fig_3_revised.png){width="10cm"}
 
 Present-day latitudinal temperature gradient.
-The present-day empirical latitudinal temperature gradient (median sea surface temperature) is depicted as a black line, and the gradient estimated by the Bayesian model is shown in turquoise.
+The present-day empirical latitudinal temperature gradient (median sea surface temperature) is depicted as a black line, and the gradient estimated by the Bayesian model is shown in turquoise ($R^2 = `r round(R2_modern,2)`, N = 42,896)$.
 Grey points depict the individual cell values of the Bio-ORACLE grid of mean sea surface temperatures, which were used to infer the empirical and the modelled gradient.
+The darkness of the background shading indicates the point density.
 :::
@@
 The average, modern temperature gradient can be closely approximated with our model when using the full modern SST dataset (Fig.
-3); almost all of the variation in the empirical median temperatures in bins of 1&deg; absolute latitude is explained by the modelled gradient ($R^2 = `r R2_modern`$).
+3); almost all of the variation in the empirical median temperatures in bins of 1&deg; absolute latitude (black line) is explained by the modelled gradient ($`r 100*as.numeric(R2_modern_median)` \%$).
 The empirical gradient spans `r modern_gradient_empirical`&deg;C from the equator to the poles, the modelled gradient is only slightly higher at `r modern_gradient$"50%"`&deg;C.
@@
 A model run excluding the ecological proxies increases the GMST by `r eoceneproxy_global_diff$"50%"` (`r eoceneproxy_global_diff$"2.5%"` -- `r eoceneproxy_global_diff$"97.5%"`)&deg;C.
-The median latitudinal gradient is similar when excluding the ecological proxies, with a median of `r eocene_proxy_gradient$"50%"`&deg;C, but with a `r proxy_uncertain_diff_percent`% wider 95% CI (Fig.
+The median latitudinal gradient is higher near the equator and in high latitudes when excluding the ecological proxies, with a median of `r eocene_proxy_gradient$"50%"`&deg;C, but with a `r proxy_uncertain_diff_percent`% wider 95% CI (Fig.
 S2).
-This indicates that the ecological proxy data are broadly in agreement with the geochemical proxies, while providing additional constraints on the shape of the early Eocene temperature gradient.
+This indicates that the ecological proxy data provide additional constraints on the shape of the early Eocene temperature gradient.
+In contrast, including ecological proxies, but widening the uncertainty around the low-latitude ecological proxy data does not significantly change the resulting gradient (Fig.
+S4).
 
@@
 The inferred latitudinal gradient is somewhat steeper in the Northern Hemisphere (steeper by `r eocene_north_south_gradient_diff$"50%"`&deg;C, although the 95% CI spans `r eocene_north_south_gradient_diff$"2.5%"` -- `r eocene_north_south_gradient_diff$"97.5%"`&deg;C), but the large uncertainties associated with both gradients, and the lack of polar proxy data in the Southern Hemisphere preclude a more precise statement (see Fig.
-S3).
+S4).
 
 ::: {#fig-4}
-![](../figures/fig_4.png){width="10cm"}
+![](../figures/fig_4_revised.png){width="10cm"}
 

### 2023-09-15 manuscript/supplementary.qmd
revisions

@@
 
+::: {#fig-S2}
+![](../figures/SM/FigS2_geochem_eeco.png){width="10.5cm"}
+
+Estimated latitudinal temperature gradient using only the geochemical proxy data (yellow), showing the median (line) and 95% credible interval (shading).
+Symbols with vertical lines show the median and 95% credible intervals of $\mu_j$.
+The blue line and shading in the background show the latitudinal temperature gradient with the geochemical and ecological proxy data, as in Figure 4.
+:::
+
+::: {#fig-S3}
+![](../figures/SM/FigS_wider-ecol-limits.png){width="10.5cm"}
+
+Estimated latitudinal temperature gradient using only wider ecological limits for the coral data and for the *Avicennia*-Rhizophoraceae mangrove assemblage (orange), showing the median (line) and 95% credible interval (shading).
+The ecological limits were expanded to include $95\%$ of the probability density in the interval of $16.0$ to $35.6$&deg;C, using the minimal monthly temperature experienced by coral reefs (@Kleypas1999), and a theoretical maximum derived by [@jones2022reef].
+Symbols with vertical lines show the median and 95% credible intervals of $\mu_j$.
+The black line and shading in the background show the latitudinal temperature gradient with the narrower ecological limits, as in Figure 4.
+:::
+
+::: {#fig-S4}
+![](../figures/SM/FigS3_north_south_eeco.png){width="12cm"}
+
+Estimated latitudinal temperature gradients for each hemisphere, showing the medians (lines) and 95% credible intervals (shadings) in the Southern (red) and Northern Hemisphere (blue).
+Symbols with vertical lines show the median and 95% credible intervals of $\mu_j$.
+Turquoise symbols in the Northern Hemisphere highlight the ecological proxy data.
+The grey line and shading in the background show the latitudinal temperature gradient with the data from both hemispheres combined, plotted in both hemispheres.
+The dotted lines show the empirical, modern gradient, averaged across bins of 1&deg, for both hemispheres combined (black), and for the Northern hemisphere (blue) and Southern hemisphere (red), separately.
+:::
+
 # EECO gradient with uncertainty on proxy observations 
@@
 
-::: {#fig-S2}
+::: {#fig-S5}
 ![](../figures/SM/FigS_sd_included.png){width="10.5cm"}
@@
 
-::: {#fig-S2}
-![](../figures/SM/FigS2_geochem_eeco.png){width="10.5cm"}
-
-Estimated latitudinal temperature gradient using only the geochemical proxy data (orange), showing the median (line) and 95% credible interval (shading).
-Symbols with vertical lines show the median and 95% credible intervals of $\mu_j$.
-The blue line and shading in the background show the latitudinal temperature gradient with the geochemical and ecological proxy data, as in Figure 4.
-:::
-
-::: {#fig-S3}
-![](../figures/SM/FigS3_north_south_eeco.png){width="12cm"}
-
-Estimated latitudinal temperature gradients for each hemisphere, showing the medians (lines) and 95% credible intervals (shadings) in the Southern (red) and Northern Hemisphere (blue).
-Symbols with vertical lines show the median and 95% credible intervals of $\mu_j$.
-Turquoise symbols in the Northern Hemisphere highlight the ecological proxy data.
-The grey line and shading in the background show the latitudinal temperature gradient with the data from both hemispheres combined, plotted in both hemispheres.
-:::
 <!-- REMOVED

### 2023-09-15 R/analyses/SM/run_EECO_climate_model_different_prior.R

@@ -46,12 +46,12 @@ doParallel::registerDoParallel(cl)
 
 ### Run model
 # source script
-mode3 <- foreach(pc = 1:n_chains) %dopar% { # run 1 chain per cluster
+mode <- foreach(pc = 1:n_chains) %dopar% { # run 1 chain per cluster
   # call model functions
   source("R/functions/model_components/climate_model_EECO.R")
   # set random seed
   set.seed(pc)
-  hierarchical_model(n_iter = 0.1*n_iter, n_thin = 0.1*n_thin,
+  hierarchical_model(n_iter = n_iter, n_thin = n_thin,
                   obsmat = obsmat, distrmat = distrmat, 
                   prior_input = priors,adapt_sd = adapt_sd)
 }
@@ -60,17 +60,23 @@ mode3 <- foreach(pc = 1:n_chains) %dopar% { # run 1 chain per cluster
 parallel::stopCluster(cl)
 
 ### save output
-saveRDS(mode, "results/SM/eeco_climate_model_output_different_M_prior.rds")
+saveRDS(mode, "results/eeco/eeco_climate_model_output.rds")
 
 # combine chains and discard burn-in
 source("R/functions/model_processing/combine_posterior.R")
-mode_all <- combine_posterior(mode,100000*0.1)
-mode_all2 <- combine_posterior(mode2,100000*0.1)
-mode_all3 <- combine_posterior(mode3,100000*0.1)
+source("R/functions/auxiliary_functions.R")
+source("R/functions/model_components/gradient.R")
+
+mode_all <- combine_posterior(mode,100000)
+lat <- seq(from = 0, to = 90, by = 0.1)
+
+source("./R/functions/model_processing/temp_from_gradient.R")
+eeco_temp <- temp_from_gradient(lat = lat, model_out = mode_all)
+
+plot(lat, eeco_temp$median)
+error_polygon(lat, eeco_temp$l_ci_95, eeco_temp$u_ci_95, col = rgb(1,0,0,0.2))
 
 plot_gradient(mode_all, ylim = c(-5,37))
-plot_gradient(mode_all3, add = T, line_col = rgb(0,1,1,.7), confint_col = rgb(0,0,1,0.2))
-plot_gradient(mode_all2, add = T, line_col = rgb(1,0,0,.7), confint_col = rgb(1,0,0,0.2))
 
 plot_chains(mode)
 

### 2023-09-15 R/analyses/SM/run_EECO_climate_model_hemispheres.R

@@ -2,6 +2,8 @@
 ###
 ### seperate gradients- northern hemisphere
 ###
+# source options for analysis
+source("./R/options.R")
 
 ### Read data
 ## Hollis data

### 2023-09-15 R/analyses/SM/run_EECO_climate_model_vary_ecological_constraints.R

@@ -3,6 +3,11 @@
 # source options for analysis
 source("./R/options.R")
 #
+# change prior on M
+priors$f3 <- function(x,log) dunif(x, min = 0, max = 90, log = log) # dnorm(x, mean = 42, sd = 25, log = log) #
+# change prior on K
+priors$f2 <- function(x,lower,log) dtnorm(x, lower, upper = Inf, mean = 28, sd = 15, log = log)
+
 ### Read data
 ## Hollis data
 dat <- readRDS("data/processed/Hollis_processed_EECO_2022_07_19.rds")

### 2023-09-15 R/analyses/processing/modern_gradient_R_squared.R

@@ -1,0 +1,48 @@
+# Load libraries ---------------------------------------------------------
+library(terra)
+library(ggplot2)
+library(ggpubr)
+source("./R/functions/model_processing/temp_from_gradient.R")
+source("./R/functions/model_processing/combine_posterior.R")
+# Load data --------------------------------------------------------------
+temp <- rast("./data/raw/climate/BioOracle_20220711/Present.Surface.Temperature.Mean.asc")
+modern_sample <- readRDS("./results/modern/modern_climate_model_output.rds")
+eocene_sample <- readRDS("./results/modern/modern_sample_eocene_p_lat_gradient.RDS")
+eocene_reps <- readRDS("./results/modern/modern_sample_summary_stats.RDS")
+# Data preparation -------------------------------------------------------
+# Reduce resolution of raster
+r <- rast(res = 1)
+temp <- resample(x = temp, y = r)
+# Extract temperature data
+temp <- terra::as.data.frame(x = temp, xy = TRUE)
+colnames(temp) <- c("lng", "lat", "SST")
+# Remove data from continents
+temp <- na.omit(temp)
+# Make lats absolute values
+temp$lat <- abs(temp$lat)
+eocene_reps$p_lat <- abs(eocene_reps$p_lat)
+# Add 1 deg latitudinal bins
+temp$bin <- ceiling(temp$lat) - 0.5
+# Calculate median per lat bin
+grad <- tapply(temp$SST, temp$bin, median)
+grad <- data.frame(lat = as.numeric(names(grad)), SST = as.vector(grad))
+# save for later use
+saveRDS(grad,"results/modern/empirical_median_1-deg-lat.rds")
+# Calculate modern temperatures from sample
+lat <- seq(from = 0, to = 90, by = 0.1)
+modern_gradient <- temp_from_gradient(lat = lat, model_out = modern_sample)
+# Calculate modern temperatures from palaeodistribution of Eocene data
+#eocene_sample <- combine_posterior(mod = eocene_sample, burnin = 5000)
+#eocene_sample <- temp_from_gradient(lat = lat, model_out = eocene_sample)
+# Plot ------------------------------------------------------------------------
+
+#if(s == 4)     grad <- gradient(lats, four_gradients_model[[g]][[s+1]])
+
+true_grad <- gradient_formulas_static[[g]](lats)
+
+var_fit <- var(modern_gradient$median)
[…]

### 2023-09-15 R/analyses/run_simulated_gradients_revision3.R

@@ -1,379 +1,0 @@
-### Supplementary figure: different gradient shapes
-
-library(dplyr)
-
-# source functions
-source("R/functions/model_components/climate_model_modern.R")
-source("R/functions/model_components/gradient.R")
-source("R/functions/model_processing/combine_posterior.R")
-source("R/functions/auxiliary_functions.R")
-# source options for analysis
-source("R/options.R")
-
-## wider priors on M and K
-# change prior on M
-priors$f3 <- function(x,log) dunif(x, min = 0, max = 90, log = log) # dnorm(x, mean = 42, sd = 25, log = log) #
-# change prior on K
-priors$f2 <- function(x,lower,log) dtnorm(x, lower, upper = Inf, mean = 28, sd = 15, log = log)
-
-
-## Hollis data
-dat <- readRDS("data/processed/Hollis_processed_EECO_2022_07_19.rds")
-#
-# prepare for use in the model
-obsmat <- data.frame(sample = (paste(abs(dat$p_lat),dat$longitude, dat$proxy)),
-                     p_lat = abs(dat$p_lat), temperature = dat$temperature,
-                     sd = dat$temperature_sd,
-                     proxy = dat$proxy)
-# order observation matrix for easily keeping tracks of mean estimates
-obsmat <- obsmat[with(obsmat, order(p_lat, sample)),]
-
-location_sample_size <- as.numeric(c(table(obsmat$sample), c(rep(1,11))))
-
-eocene_lat <- c(-59.42, -59.10, -54.13, -47.84, -47.84, -47.47, -46.14, -45.80, -45.80, -24.25, -8.57, -2.05, -1.86, 0.49, 4.29, 14.67, 27.44, 31.00, 32.50, 32.80, 32.90, 35.54, 36.22, 39.02, 40.39, 40.63, 40.89, 43.79, 44.11, 50.04, 50.39, 58.45, 76.45, 77.13)
-# 
-generate_lat_samples <- function(n) 180/pi * asin(runif(n,0,1))
-
-set.seed(1)
-
-gradient_formulas_static <- rev(list( function(x) 30-x^2/1080, # extreme greenhouse
-                               function(x) 30-x^2/360, 
-                               function(x) {
[…]

### 2023-09-15 R/analyses/run_simulated_gradients_revision_final.R

@@ -1,0 +1,379 @@
+### Supplementary figure: different gradient shapes
+
+library(dplyr)
+
+# source functions
+source("R/functions/model_components/climate_model_modern.R")
+source("R/functions/model_components/gradient.R")
+source("R/functions/model_processing/combine_posterior.R")
+source("R/functions/auxiliary_functions.R")
+# source options for analysis
+source("R/options.R")
+
+## wider priors on M and K
+# change prior on M
+priors$f3 <- function(x,log) dunif(x, min = 0, max = 90, log = log) # dnorm(x, mean = 42, sd = 25, log = log) #
+# change prior on K
+priors$f2 <- function(x,lower,log) dtnorm(x, lower, upper = Inf, mean = 28, sd = 15, log = log)
+
+
+## Hollis data
+dat <- readRDS("data/processed/Hollis_processed_EECO_2022_07_19.rds")
+#
+# prepare for use in the model
+obsmat <- data.frame(sample = (paste(abs(dat$p_lat),dat$longitude, dat$proxy)),
+                     p_lat = abs(dat$p_lat), temperature = dat$temperature,
+                     sd = dat$temperature_sd,
+                     proxy = dat$proxy)
+# order observation matrix for easily keeping tracks of mean estimates
+obsmat <- obsmat[with(obsmat, order(p_lat, sample)),]
+
+location_sample_size <- as.numeric(c(table(obsmat$sample), c(rep(1,11))))
+
+eocene_lat <- c(-59.42, -59.10, -54.13, -47.84, -47.84, -47.47, -46.14, -45.80, -45.80, -24.25, -8.57, -2.05, -1.86, 0.49, 4.29, 14.67, 27.44, 31.00, 32.50, 32.80, 32.90, 35.54, 36.22, 39.02, 40.39, 40.63, 40.89, 43.79, 44.11, 50.04, 50.39, 58.45, 76.45, 77.13)
+# 
+generate_lat_samples <- function(n) 180/pi * asin(runif(n,0,1))
+
+set.seed(1)
+
+gradient_formulas_static <- rev(list( function(x) 30-x^2/1080, # extreme greenhouse
+                               function(x) 30-x^2/360, 
+                               function(x) {
[…]

### 2023-09-15 R/figures/SM/FigS-include-proxy-SD.R

@@ -40,23 +40,33 @@ alpha2 <- 0.75
 col1 <- c(rgb(.9,.8,.5,alpha2), rgb(.8,.7,.5,alpha2), rgb(.7,.6,.5,alpha2), rgb(.65,.6,.6,alpha2))
 
 
+lat <- seq(0,90,1)
+
+all_temp <- temp_from_gradient(lat = lat, model_out = mode_all)
+all_sd_temp <- temp_from_gradient(lat = lat, model_out = mode_all_sd)
+
+
 
 png("figures/SM/FigS_sd_included.png", width = 4.75, height = 3.5, units = "in", res = 400)
 par(mar=c(4.2,4.2,0.5,0.5), mgp = c(2.5,0.8,0)
     ,las = 1)
-plot(obsmat$p_lat,obsmat$temperature, pch = pch1[proxind1], col = col1[proxind1], xlim = c(-2,92), ylim = c(14,41.5),
+plot(obsmat$p_lat,obsmat$temperature, pch = pch1[proxind1], col = col1[proxind1], xlim = c(-2,92), ylim = c(0,41.5),
      xlab = xlab1, ylab = ylab1, cex = cex1[proxind1], type = "n", xaxs = "i", yaxs = "i")
 
-plot_gradient(mode_all,add=T, line_col = rgb(0,0,0,1), confint_col = rgb(0,0,0,0.2))
+error_polygon(lat, all_temp$l_ci_95, all_temp$u_ci_95, col =  rgb(0,0,0,0.2))
 
-plot_gradient(mode_all_sd, add=T, line_col = rgb(0,.6,1,0.75), confint_col = rgb(0,.6,1,0.15))
+error_polygon(lat, all_sd_temp$l_ci_95, all_sd_temp$u_ci_95, col = rgb(0,.6,1,0.15))
+
+
+points(lat, all_temp$median, type = "l", lwd = 2, col = "black")
+points(lat, all_sd_temp$median, type = "l", lwd = 2, col = rgb(0,.6,1,0.75))
 
 plot_posterior(mode_sd[[1]], col_obs = rgb(0,0.6,1,1), col_dist = NA, add=T)# line_col = rgb(0,0,1,0.5), confint_col = rgb(0,0,1,0.1))
 
 plot_posterior(mode[[1]], col_obs = rgb(0,0,0,0.5), col_dist = NA, add=T,
                pch = 19)# line_col = rgb(0,0,1,0.5), confint_col = rgb(0,0,1,0.1))
 
[…]

### 2023-09-15 R/figures/SM/FigS2.R

@@ -22,6 +22,12 @@ mean_t <- c(yobs_mean,distrmat$mu)
 lats <- mode[[1]]$lat # latitudes of locations is saved in model output
 
 
+all_temp <- temp_from_gradient(lat = lat, model_out = mode_all)
+proxy_temp <- temp_from_gradient(lat = lat, model_out = mode_all_prox)
+
+
+
+
 cex1 <- c(1.05,1,1.15,0.92)
 alpha2 <- 0.75
 pch1 <- c(16,17,18,15)
@@ -49,12 +55,17 @@ ind1_mu <- sapply(unique(obsmat$sample), function(x) which(unique(obsmat$proxy[w
 png("figures/SM/FigS2_geochem_eeco.png", width = 4.75, height = 3.5, units = "in", res = 400)
 par(mar=c(4.2,4.2,0.5,0.5), mgp = c(2.5,0.8,0)
     ,las = 1)
-plot(obsmat$p_lat,obsmat$temperature, pch = pch1[proxind1], col = col1[proxind1], xlim = c(-2,92), ylim = c(14,41.5),
+plot(obsmat$p_lat,obsmat$temperature, pch = pch1[proxind1], col = col1[proxind1], xlim = c(-2,92), ylim = c(0,42.5),
      xlab = xlab1, ylab = ylab1, cex = cex1[proxind1], type = "n", xaxs = "i", yaxs = "i")
 
-plot_gradient(mode_all,add=T, line_col = rgb(0,0.3,0.8,0.5), confint_col = rgb(0,0.3,0.8,0.15))
 
-plot_gradient(eecoproxcomb,add=T, line_col = col_proxgrad, confint_col = col_proxgrad_shade)
+
+error_polygon(lat, all_temp$l_ci_95, all_temp$u_ci_95, col = rgb(0,0.3,0.8,0.15))
+error_polygon(lat, proxy_temp$l_ci_95, proxy_temp$u_ci_95, col = col_proxgrad_shade)
+
+
+points(lat, all_temp$median, type = "l", lwd = 2, col = rgb(0,0.3,0.8,0.5))
+points(lat, proxy_temp$median, type = "l", lwd = 2, col = col_proxgrad)
 
 points(lat_obs[1:23], mu_obs[1:23], col = col1[ind1_mu], pch = pch1[ind1_mu], lwd = 3, cex = 0.1+cex1[ind1_mu])
[…]

### 2023-09-15 R/figures/SM/FigS3.R

@@ -62,18 +62,77 @@ n_obs_south <- length(yobs_mean_south)
 mean_t_south <- c(yobs_mean_south)
 lats_south <- mode_south[[1]]$lat # latitudes of locations is saved in model output
 
+lat <- seq(0,90,1)
+
+all_temp <- temp_from_gradient(lat = lat, model_out = mode_all)
+north_temp <- temp_from_gradient(lat = lat, model_out = mode_all_north)
+south_temp <- temp_from_gradient(lat = lat, model_out = mode_all_south)
+
+
+
+
+### Modern
+source("./R/functions/model_processing/temp_from_gradient.R")
+source("./R/functions/model_processing/combine_posterior.R")
+# Load data --------------------------------------------------------------
+temp <- rast("./data/raw/climate/BioOracle_20220711/Present.Surface.Temperature.Mean.asc")
+modern_sample <- readRDS("./results/modern/modern_climate_model_output.rds")
+# Data preparation -------------------------------------------------------
+# Reduce resolution of raster
+r <- rast(res = 1)
+temp <- resample(x = temp, y = r)
+# Extract temperature data
+temp <- terra::as.data.frame(x = temp, xy = TRUE)
+colnames(temp) <- c("lng", "lat", "SST")
+# Remove data from continents
+temp <- na.omit(temp)
+
+# north south seperate
+temp_ns <- temp
+# Add 1 deg latitudinal bins
+temp_ns$bin <- ceiling(temp_ns$lat) - 0.5
+# Calculate median per lat bin
+grad_ns <- tapply(temp_ns$SST, temp_ns$bin, median)
+grad_ns <- data.frame(lat = as.numeric(names(grad_ns)), SST = as.vector(grad_ns))
+
+
+# Make lats absolute values
+temp$lat <- abs(temp$lat)
+# Add 1 deg latitudinal bins
+temp$bin <- ceiling(temp$lat) - 0.5
+# Calculate median per lat bin
+grad <- tapply(temp$SST, temp$bin, median)
+grad <- data.frame(lat = as.numeric(names(grad)), SST = as.vector(grad))
+
+
 # Plot north south
[…]

### 2023-09-15 R/figures/SM/figS-wider-ecol-limits.R

@@ -40,6 +40,14 @@ n_obs <- length(yobs_mean)
 mean_t <- c(yobs_mean,distrmat$mu)
 lats <- mode[[1]]$lat # latitudes of locations is saved in model output
 
+lat <- seq(0,90,0.5)
+
+
+# get temperatures from gradient
+eeco_temp <- temp_from_gradient(lat = lat, model_out = mode_all)
+eeco_wide_temp <- temp_from_gradient(lat = lat, model_out = mode_all_ecowide)
+
+
 
 cex1 <- c(1.05,1,1.15,0.92)
 alpha2 <- 0.75
@@ -56,17 +64,20 @@ col1 <- c(rgb(.9,.8,.5,alpha2), rgb(.8,.7,.5,alpha2), rgb(.7,.6,.5,alpha2), rgb(
 
 
 
+
 png("figures/SM/FigS_wider-ecol-limits.png", width = 4.75, height = 3.5, units = "in", res = 400)
 par(mar=c(4.2,4.2,0.5,0.5), mgp = c(2.5,0.8,0)
     ,las = 1)
-plot(obsmat$p_lat,obsmat$temperature, pch = pch1[proxind1], col = col1[proxind1], xlim = c(-2,92), ylim = c(14,41.5),
+plot(obsmat$p_lat,obsmat$temperature, pch = pch1[proxind1], col = col1[proxind1], xlim = c(-2,92), ylim = c(0,40.5),
      xlab = xlab1, ylab = ylab1, cex = cex1[proxind1], type = "n", xaxs = "i", yaxs = "i")
 
-plot_gradient(mode_all,add=T, line_col = rgb(0,0,0,1), confint_col = rgb(0,0,0,0.2))
 
-plot_gradient(mode_all_ecowide, add=T, line_col = col_proxgrad, confint_col = col_proxgrad_shade)
+error_polygon(lat, eeco_temp$l_ci_95, eeco_temp$u_ci_95, col = rgb(0,0,0,0.2))
+error_polygon(lat, eeco_wide_temp$l_ci_95, eeco_wide_temp$u_ci_95, col = col_proxgrad_shade)
 
 
+points(lat, eeco_temp$median, type = "l", lwd = 2, col = "black")
+points(lat, eeco_wide_temp$median, type = "l", lwd = 2, col = col_proxgrad)
 
[…]

### 2023-09-15 R/figures/fig_3.R

@@ -32,14 +32,18 @@ saveRDS(grad,"results/modern/empirical_median_1-deg-lat.rds")
 lat <- seq(from = 0, to = 90, by = 0.1)
 modern_sample <- temp_from_gradient(lat = lat, model_out = modern_sample)
 # Calculate modern temperatures from palaeodistribution of Eocene data
-eocene_sample <- combine_posterior(mod = eocene_sample, burnin = 5000)
-eocene_sample <- temp_from_gradient(lat = lat, model_out = eocene_sample)
+#eocene_sample <- combine_posterior(mod = eocene_sample, burnin = 5000)
+#eocene_sample <- temp_from_gradient(lat = lat, model_out = eocene_sample)
 # Plot ------------------------------------------------------------------------
 # Modern gradient
 p1 <- ggplot() + 
-  geom_point(data = temp,
-             aes(x = lat, y = SST), 
-             colour = "grey80", size = 0.7, alpha = 0.7) +
+   geom_point(data = temp,
+              aes(x = lat, y = SST), 
+              colour = "grey80", size = 0.7, alpha = 0.7) +
+  stat_density2d(data = temp, aes(x = lat, y = SST, fill = sqrt(..density..)), geom = "tile", contour = FALSE,
+                 show.legend=FALSE) +
+  scale_fill_gradient(low = "transparent", high = "grey20", guide = "none") + 
+  guides(fill = "none",  color = guide_legend(override.aes = list(fill = NA))) +  # Remove fill legend
   geom_line(data = grad,
             aes(x = lat, y = SST, colour = "Empirical gradient"),
             linewidth = 1, alpha = 1) +
@@ -57,6 +61,10 @@ p1 <- ggplot() +
         legend.position = c(0.8, 0.9),
         legend.title = element_blank(),
         legend.background = element_blank(),
+        legend.key = element_blank(), # Remove the key background
+        legend.key.size = unit(1, 'lines'),
+        legend.box.background = element_blank(),
+        legend.box.margin = margin(0, 0, 0, 0),
         legend.text = element_text(size = 12),
         legend.spacing.y = unit(0.5, "mm"),
         axis.text = element_text(size = 14),
@@ -98,6 +106,6 @@ p1
 #         panel.grid.minor = element_blank())
 # # Combine plots and save ------------------------------------------------------
 # ggarrange(p1, p2, ncol = 1, labels = "AUTO", font.label = list(size = 18))
[…]

### 2023-09-15 R/figures/fig_4.R

@@ -57,7 +57,7 @@ ggplot() +
             size = 1.5, alpha = 0.85) +
   geom_ribbon(data = eeco_temp,
               aes(x = lat, ymin = l_ci_95, ymax = u_ci_95),
-              colour = NA, fill = "#54278f", alpha = 0.5) +
+              colour = NA, fill = "#54278f", alpha = 0.35) +
   geom_line(data = eeco_temp,
              aes(x = lat, y = median, colour = "EECO gradient"),
              size = 1.5, alpha = 0.85) +
@@ -82,4 +82,4 @@ ggplot() +
         panel.grid.major = element_line(colour = "grey70", size = 0.2),
         panel.grid.minor = element_blank())
 
-ggsave("./figures/fig_4.png", units = "mm", width = 150, height = 150, dpi = 600)
+ggsave("./figures/fig_4_revised.png", units = "mm", width = 150, height = 150, dpi = 600)

### 2023-09-15 R/functions/model_processing/temp_from_gradient.R

@@ -13,6 +13,16 @@ temp_from_gradient <- function(lat,model_out, burnin = NULL) {
   
   out <- data.frame(lat = lat, median = apply(grad,2,median), 
                     l_ci_95 = apply(grad,2, function(x) quantile(x,0.025)), 
-                    u_ci_95 = apply(grad,2, function(x) quantile(x,0.975)))
+                    u_ci_95 = apply(grad,2, function(x) quantile(x,0.975)),
+                    l_ci_68 = apply(grad,2, function(x) quantile(x,0.16)), 
+                    u_ci_68 = apply(grad,2, function(x) quantile(x,0.84)),
+                    l_ci_90 = apply(grad,2, function(x) quantile(x,0.05)),
+                    u_ci_90 = apply(grad,2, function(x) quantile(x,0.95)),
+                    l_ci_99 = apply(grad,2, function(x) quantile(x,0.005)),
+                    u_ci_99 = apply(grad,2, function(x) quantile(x,0.995)),
+                    l_ci_80 = apply(grad,2, function(x) quantile(x,0.1)),
+                    u_ci_80 = apply(grad,2, function(x) quantile(x,0.9))
+                    
+  )
   return(out)
   }

### 2023-09-21 manuscript/main.qmd
seasonality paragraph, updating results, discussion, SM

@@
 Our results confirm the presence of a flattened latitudinal temperature gradient and unusually high polar temperatures during the EECO, which is supported by high-latitude ecological data.
-We show that integrating multiple types of proxy data, and adequate prior information, has the potential to substantially reduce uncertainty in palaeoclimate reconstructions, allowing for unbiased temperature estimates from sparse data.
+We show that integrating multiple types of proxy data, and adequate prior information, has the potential to enhance quantiative palaeoclimate reconstructions, allowing for more accurate temperature estimates from sparse data.
 
@@
 The blue line depicts the reconstructed gradient represented by the median sea surface temperature value estimated from 1,000 model runs with different random samples (first three columns), and a single run with the EECO latitudinal sampling distribution (fourth column).
+To generate the random samples, noise from a normal distribution with a standard deviation of 3.8 was added to all temperatures.
 The blue shadings depict the 90%, 95%, and 99% credible intervals.
@@
 
-The Eocene temperature gradient reconstructed with our Bayesian model is starkly different from the modern (Fig 4).
+The Eocene temperature gradient reconstructed with our Bayesian model is starkly different from the modern (Fig.
+4).
 Modelled, median equatorial temperatures are `r tdiff_lat0$"50%"` (95% CI: `r tdiff_lat0$"2.5%"` -- `r tdiff_lat0$"97.5%"`)&deg;C higher for the EECO, and polar temperatures are `r tdiff_lat90$"50%"` (`r tdiff_lat90$"2.5%"` -- `r tdiff_lat90$"97.5%"`)&deg;C higher.
@@
 This is reflected in the residual standard deviation ($\sigma$) of the EECO gradient -- `r eeco_sdy$"50%"` (`r eeco_sdy$"2.5%"` -- `r eeco_sdy$"97.5%"`)&deg;C -- which is more than double the  $\sigma$ for the modern gradient, `r modern_sdy$"50%"`.
-This signifies that the early Eocene data does not fit as well to the logistic latitudinal gradient model, which can also be seen from the drastic departure of some of the proxy data from the gradient estimates (Fig.
+This is illustrated by the drastic departure of some of the proxy data from the gradient estimates (Fig.
 4).
@@
 A model run excluding the ecological proxies increases the GMST by `r eoceneproxy_global_diff$"50%"` (`r eoceneproxy_global_diff$"2.5%"` -- `r eoceneproxy_global_diff$"97.5%"`)&deg;C.
-The median latitudinal gradient is higher near the equator and in high latitudes when excluding the ecological proxies, with a median of `r eocene_proxy_gradient$"50%"`&deg;C, but with a `r proxy_uncertain_diff_percent`% wider 95% CI (Fig.
+The median modelled temperature is higher near the equator and in high latitudes when excluding the ecological proxies, with a flattened median gradient of `r eocene_proxy_gradient$"50%"`&deg;C (Fig.
 S2).
-This indicates that the ecological proxy data provide additional constraints on the shape of the early Eocene temperature gradient.
 In contrast, including ecological proxies, but widening the uncertainty around the low-latitude ecological proxy data does not significantly change the resulting gradient (Fig.
-S4).
+S3).
 
@@
 
-Our results show that our Bayesian model can be used to reconstruct different types of latitudinal SST gradients from proxy data, even with small sample sizes (*n* = 10 -- 20) and patchy sampling distributions (Fig.
+Our results show that our Bayesian model can be used to reconstruct different types of latitudinal SST gradients from proxy data with moderate sample sizes (*n* = 10 -- 34) and patchy sampling distributions (Fig.
 2).
@@
 Earlier, reconstructed early Eocene and EECO SST gradients range from 7 -- 21&deg;C (Table 1); a more recent reconstruction that includes terrestrial air and sea surface temperatures arrives at a gradient of ~13&deg;C [@Inglis2020].
-Our polar circle to equatorial gradient estimate is lower than most previous estimates at 7.8&deg;C, although the 95% credible interval extends up to 13.7&deg;C and thus overlaps earlier estimates based on shallow water proxies.
+Our median polar circle to equatorial gradient estimate is similar at `r eocene_gradient$"50%"`&deg;C.
 The confirmation of a very flat gradient by both geochemical and ecological shallow water data indicates that inferred SST gradients based on tropical, shallow water and deep water samples [@Cramwinckel2018; @Evans2018] may overestimate the SST gradient of the early Eocene greenhouse world.
@@
 
-Discrepancies between earlier, proxy-based reconstructions and our modelling results are most pronounced in latitudes beyond the polar circle, as earlier approaches [e.g.
+<!-- OUTDATED Discrepancies between earlier, proxy-based reconstructions and our modelling results are most pronounced in latitudes beyond the polar circle, as earlier approaches [e.g.
 @Tierney2017] predict almost linearly decreasing SSTs towards the poles, whereas our median prediction suggests only a slight decrease beyond the polar circle.
@@
 In contrast, the extrapolated polar temperatures of most previous proxy-based models are likely too low, given the abundance of ecological data indicating temperate or subtropical high-latitude climates during the EECO [@Pross2012; @Popescu2021].
+-->
 
@@
 Despite excluding &delta;^18^O measurements from recrystallised fossils, systematic offsets remain between mostly warm temperatures derived from TEX~86~, and cooler temperatures derived from &delta;^18^O, &Delta;~47~, and the ecological proxies.
-Seasonality [@Keating-Bitonti2011] and temporal changes within the EECO [@Westerhold2018] may also contribute to the large variability of the EECO proxy data.
+
+Temporal changes within the EECO [@Westerhold2018], and seasonality [@Keating-Bitonti2011; @Ivany2022] may also contribute to the large variability of the EECO proxy data.
+Based on the occurrence of heterotrophic carbonates, @Davies2019 suggest that mid- and high-latitude geochemical proxy data from the EECO may be biased towards summer temperatures.
+Some of the geochemical mid-latitude geochemical proxy data from @Hollis2019 may therefore suggest higher than actual mean annual temperatures, and the variability of temperature estimates from individual localities is higher in mid - high latitudes (Fig.
+S6).
+Critically, however, the mangrove data strongly supports our inference of a flattened gradient independent of the geochemical proxy record.
 

### 2023-09-21 manuscript/supplementary.qmd
seasonality paragraph, updating results, discussion, SM

@@
 Estimated latitudinal temperature gradient using only wider ecological limits for the coral data and for the *Avicennia*-Rhizophoraceae mangrove assemblage (orange), showing the median (line) and 95% credible interval (shading).
-The ecological limits were expanded to include $95\%$ of the probability density in the interval of $16.0$ to $35.6$&deg;C, using the minimal monthly temperature experienced by coral reefs (@Kleypas1999), and a theoretical maximum derived by [@jones2022reef].
+The ecological limits were expanded to include $95\%$ of the probability density in the interval of $16.0$ to $35.6$&deg;C, using the minimal monthly temperature experienced by coral reefs (Kleypas et al., 1999), and a theoretical maximum derived by Jones et al.
+(2022).
 Symbols with vertical lines show the median and 95% credible intervals of $\mu_j$.
@@
 
-# EECO gradient with uncertainty on proxy observations 
+## EECO gradient with uncertainty on proxy observations 
 Most of the geochemical proxy data used in the EECO analysis (300 of 308) come with uncertainties around their temperature estimates.
@@
 
+::: {#fig-S6}
+![](../figures/SM/FigS_temp_location_sd.png){width="8.5cm"}
+
+Standard deviation of temperature estimates from individual localities of the EECO geochemical proxy data from Hollis et al.
+(2019) with more than one sample, against the absolute latitude of the locality (n = 17).
+The line shows a linear regression, suggesting a moderate effect of absolute latitude on proxy data variability, with mid- and hight-latitude data being more variable than low-latitude data.
+:::
 <!-- REMOVED

### 2023-09-21 R/analyses/processing/calculate_temperature_gradients_from_models.R

@@ -1,6 +1,7 @@
 ### calculating results
 source("./R/functions/model_processing/temp_from_gradient.R")
 source("./R/functions/model_components/gradient.R")
+source("./R/functions/model_processing/combine_posterior.R")
 
 # modern
 modm <- readRDS("./results/modern/modern_climate_model_output.rds")

### 2023-09-22 manuscript/main.qmd
add Lunt2021 to table

@@
 (2012), as shown in Tierney et al.
-(2017)"),
-  Time = c("EE", "EECO", "EE", "EECO", "EE", "EE"),
-  Gradient = c("7", "13", "12", "21 (±1)", "20 (±3)", "26"),
-  Type_of_gradient = c("equator - polar circle", "equator - polar circle", "equator - polar circle", "equator - deep water", "tropics - deep water", "equator - polar circle"),
+(2017)", "Lunt et al.
+(2021)"),
+  Time = c("EE", "EECO", "EE", "EECO", "EE", "EE", "EECO"),
+  Gradient = c("7", "13", "12", "21 (±1)", "20 (±3)", "26", "18 - 26"), 
+  Type_of_gradient = c("equator - polar circle", "equator - polar circle", "equator - polar circle", "equator - deep water", "tropics - deep water", "equator - polar circle", "tropics - high latitude"),
   Model = c("$2^{nd}$ order polynomial",
@@
             "-",
-            "climate model ensemble (GCM)"),
+            "climate model ensemble",
+            "DeepMIP climate models"),
   Proxy_system = c("$TEX_{86}$, $UK_{37}^{K'}$",
@@
                    "$\\Delta_{47}$, deepwater Mg/Ca",
+                   "none (GCM simulations)",
                    "none (GCM simulations)")
@@
 The confirmation of a very flat gradient by both geochemical and ecological shallow water data indicates that inferred SST gradients based on tropical, shallow water and deep water samples [@Cramwinckel2018; @Evans2018] may overestimate the SST gradient of the early Eocene greenhouse world.
+Likewise, climate simulations using general circulation or Earth system models tend to estimate steeper gradients than most proxy records [Table 1\; @Pross2012; @Lunt2021]
 

### 2023-09-22 manuscript/main.qmd
remove leading spaces from results

@@
 This results in a flattened latitudinal temperature gradient of `r eocene_gradient$"50%"` (`r eocene_gradient$"2.5%"` -- `r eocene_gradient$"97.5%"`)&deg;C for the EECO, as opposed to `r modern_gradient$"50%"`&deg;C for the modern.
-To facilitate the comparison with latitudinal gradients reported in the literature, which sometimes do not report temperatures at very high latitudes, we report also the EECO gradient between the equator and the modern-day polar circle (66.6&deg;), which is slightly lower at `r eocene_gradient_polar_circle$"50%"` (`r eocene_gradient_polar_circle$"2.5%"` -- `r eocene_gradient_polar_circle$"97.5%"`)&deg;C.
+To facilitate the comparison with latitudinal gradients reported in the literature, which sometimes do not report temperatures at very high latitudes, we report also the EECO gradient between the equator and the modern-day polar circle (66.6&deg;), which is a lot lower at `r eocene_gradient_polar_circle$"50%"` (`r eocene_gradient_polar_circle$"2.5%"` -- `r eocene_gradient_polar_circle$"97.5%"`)&deg;C.
 

### 2023-09-22 R/analyses/processing/calculate_temperature_gradients_from_models.R

@@ -50,15 +50,26 @@ n <- length(modern_t0)
 # randomly recombine iterations of modern and eocene gradient and take difference
 set.seed(1)
 tdiff_lat0 <- as.list(format(round(quantile(eocene_t0[sample(1:n,n)] - modern_t0[sample(1:n,n)], probs = c(0.025,0.5,0.975)),1),nsmall = 1))
+tdiff_lat0 <- lapply(tdiff_lat0, function(x) gsub(" ", "", x))
+
 tdiff_lat90 <- as.list(format(round(quantile(eocene_t90[sample(1:n,n)] - modern_t90[sample(1:n,n)], probs = c(0.025,0.5,0.975)),1),nsmall = 1))
+tdiff_lat90 <- lapply(tdiff_lat90, function(x) gsub(" ", "", x))
 
 # calculate overall gradient for modern and eocene (polar - equatorial difference)
 eocene_gradient <- as.list(format(round(quantile(eocene_t0 - eocene_t90, probs = c(0.025,0.5,0.975)),1),nsmall = 1))
+eocene_gradient <- lapply(eocene_gradient, function(x) gsub(" ", "", x))
+
 modern_gradient <- as.list(format(round(quantile(modern_t0 - modern_t90, probs = c(0.025,0.5,0.975)),1),nsmall = 1))
+modern_gradient <- lapply(modern_gradient, function(x) gsub(" ", "", x))
+
 modern_eocene_gradient <- as.list(format(round(quantile(modern_eocene_t0 - modern_eocene_t90, probs = c(0.025,0.5,0.975)),1),nsmall = 1))
+modern_eocene_gradient <- lapply(modern_eocene_gradient, function(x) gsub(" ", "", x))
+
 eocene_proxy_gradient <- as.list(format(round(quantile(eeco_prox_t0 - eeco_prox_t90, probs = c(0.025,0.5,0.975)),1),nsmall = 1))
+eocene_proxy_gradient <- lapply(eocene_proxy_gradient, function(x) gsub(" ", "", x))
 
 eocene_gradient_polar_circle <- as.list(format(round(quantile(eocene_t0 - eocene_t66.6, probs = c(0.025,0.5,0.975)),1),nsmall = 1))
+eocene_gradient_polar_circle <- lapply(eocene_gradient_polar_circle, function(x) gsub(" ", "", x))
 
 
[…]

### 2023-09-22 manuscript/main.qmd
model validation method

@@
 
-We created test data from these gradients as follows: To capture both limited sample size and skewed geographic origins of samples, we randomly sampled (200 iterations) the latitudes of the EECO data set (n = 34) used in this study, at sample sizes of 5, 10 and 20, and also used the full set of 34 latitudes.
-For each latitude, we inferred a location mean temperature from the gradients, adding random noise from a normal distribution with a standard deviation of 2.2, which is the residual standard deviation of modern sea surface temperature averages from the modelled modern gradient (see below).
-We then randomly allocated a sample size $s$ from the sample sizes present in the EECO data set to each location, and randomly generated $s$ temperature data from a normal distribution with mean equal to the previously generated location mean temperatures, and a standard deviation of 3.8.
-This standard deviation is the average standard deviation of the EECO geochemical proxy data, and with that we aim to simulate randomly distributed errors in the proxy data.
-This approach should give a fair assessment of how the model setup can be affected by the known unknowns (quantifiable errors), but we acknowledge that it cannot quantify the potential impact of unknown, systematic offsets that may bias all proxy data from one location in the same direction.
+We created test data from these gradients as follows: We randomly sampled (1000 iterations) latitudes at sample sizes of 5, 10 and 20, with the probability of a latitude being sampled scaling with the decreasing surface area towards higher latitudes; i.e.
+lower latitudes are sampled more frequently.
+For the largest sample size (n = 34), we used the latitudes of the EECO data set of this study in all iterations.
+For each latitude, we inferred a location mean temperature from the gradients, adding random noise from a normal distribution with a standard deviation of 3.8, which is the average standard deviation of the EECO geochemical proxy data.
+With that, we aim to simulate randomly distributed errors in the proxy data, which could arise from miscalibrations, measurement errors, seasonal effects, ect.
+We acknowledge that this approach cannot quantify the potential impact of systematic offsets that may bias all proxy data in the same direction, nor do we know whether a standard deviation of 3.8 is actual, average magnitude of uncertainty that the proxy compilation is afflicted with.
 

### 2023-09-22 manuscript/main.qmd
citation correction
discussion first paragraph improved

@@
 experimentally derived calibrations, influences from seasonality, dissolution effects and differential preservation [e.g.
-@Tierney2017; ], can allow for robust estimates of palaeotemperatures at a local scale.
+@Tierney2017], can allow for robust estimates of palaeotemperatures at a local scale.
 However, recent work has demonstrated that spatial biases in the geochemical proxy record can lead to spurious estimates of regional (e.g.
@@
 Table 1: Inferred latitudinal sea surface temperature (SST) gradients for the early Eocene (EE) or the EECO, as shown in earlier, proxy-based studies.
+The gradient values denote the SST difference between the equator and the polar circle, or other types of gradients.
 For comparison, a gradient derived from an atmosphere-ocean general circulation model (GCM) ensemble is also shown.
@@
 Earlier, reconstructed early Eocene and EECO SST gradients range from 7 -- 21&deg;C (Table 1); a more recent reconstruction that includes terrestrial air and sea surface temperatures arrives at a gradient of ~13&deg;C [@Inglis2020].
-Our median polar circle to equatorial gradient estimate is similar at `r eocene_gradient$"50%"`&deg;C.
-The confirmation of a very flat gradient by both geochemical and ecological shallow water data indicates that inferred SST gradients based on tropical, shallow water and deep water samples [@Cramwinckel2018; @Evans2018] may overestimate the SST gradient of the early Eocene greenhouse world.
+Our median poles-to-equator gradient estimate is similar at `r eocene_gradient$"50%"`&deg;C, but a lot less steep when taking the equator-to-polar-circle estimate, `r eocene_gradient_polar_circle$"50%"`, as the geochemical proxy data suggest high temperatures up to latitudes of ~ 60&deg.
+Both geochemical and ecological shallow water data indicate that inferred SST gradients based on tropical, shallow water and deep water samples [@Cramwinckel2018; @Evans2018] may overestimate the SST gradient of the early Eocene greenhouse world.
 Likewise, climate simulations using general circulation or Earth system models tend to estimate steeper gradients than most proxy records [Table 1\; @Pross2012; @Lunt2021]

### 2023-09-22 manuscript/main.qmd
units of distributions added

@@
 Nevertheless, coral reefs have frequently been recognised as tracers of past (sub-)tropical conditions [@ziegler1984; @kiessling2001].
-During the Eocene, coral communities and reefs expanded across tropical and temperate latitudes, with communities found up to palaeolatitudes of 43 &deg;&nbsp;N [@Zamagni2012].
+During the Eocene, coral communities and reefs expanded across tropical and temperate latitudes, with communities found up to palaeolatitudes of 43&deg;&nbsp;N [@Zamagni2012].
 Using a compilation of Paleocene -- early Eocene coral reefs and community localities [@Zamagni2012], we generated quantitative sea surface temperature estimates for the ECCO.
@@
 Subsequently, we used statistically derived temperature limits (minimum = 21&deg;C, average = 27.6&deg;C, maximum = 29.5&deg;C) from the published literature [@kleypas1999] to define a normal probability distribution of potential temperature values for coral reef localities.
-This normal probability distribution was defined with a mean of 27.6 and a standard deviation of 2.125, placing 97.5% of the probability density above the minimum.
+This normal probability distribution was defined with a mean of 27.6&deg;C and a standard deviation of 2.125&deg;C, placing 97.5% of the probability density above the minimum.
 As the distribution of modern corals is skewed towards warmer temperatures, this approach results in 16.5% of the probability being placed on temperatures \> 29.5&deg;C, allowing for the possibility that Eocene coral reefs were adapted to warmer conditions than present-day coral reefs.
@@
 However, a value of 22.5&deg;C is ascribed as the upper temperature limit here as *Rhizophora* is rare below this temperature.
-We define the *Avicennia*-only temperature distribution as a normal distribution with a mean of 19.05 and a standard deviation of 1.725, resulting in 95% of the probability density being placed within the temperature limits.
+We define the *Avicennia*-only temperature distribution as a normal distribution with a mean of 19.05&deg;C and a standard deviation of 1.725&deg;C, resulting in 95% of the probability density being placed within the temperature limits.
 
@@
 As the upper thermal limits of *Aviciennia* and *Rhizophora* are not well established in @Quisthoudt2012, we assign the same maximum temperature limits (29.5&deg;C) as coral reef localities, because mangroves are also widely distributed throughout tropical regions.
-Consequently, we define the temperature distribution for this locality as a normal distribution with a mean of 25.1 and a standard deviation of 2.2, with 95% probability density within the temperature limits.
+Consequently, we define the temperature distribution for this locality as a normal distribution with a mean of 25.1&deg;C and a standard deviation of 2.2&deg;C, with 95% probability density within the temperature limits.
 

### 2023-09-22 manuscript/main.qmd
add general Bayesian refs

@@
 
-Here, we present a novel Bayesian hierarchical model that combines quantitative proxies and ecological constraints into a fully quantitative model of the latitudinal gradient of sea surface temperatures, bridging spatial gaps in sparsely sampled climate data.
+Here, we present a novel Bayesian hierarchical model [see e.g.
+@Gelman2013; @Mcelreath2018] that combines quantitative proxies and ecological constraints into a fully quantitative model of the latitudinal gradient of sea surface temperatures, bridging spatial gaps in sparsely sampled climate data.
 The Bayesian approach offers a powerful framework for integrating various sources of uncertainty and modelling complex hierarchical relationships, and is increasingly used in palaeoclimatic reconstructions [e.g.

### 2023-09-22 manuscript/main.qmd
remove en-dash spaces

@@
 This compilation provides sea surface temperature data on four different geochemical proxies for reconstructing seawater temperature: &delta;^18^O, &Delta;~47~, Mg/Ca and TEX~86~.
-For our analyses, this dataset was restricted to the EECO (defined as 53.8 -- 49.1 Ma) and samples originating from near the ocean surface or mixed layer.
+For our analyses, this dataset was restricted to the EECO (defined as 53.8--49.1 Ma) and samples originating from near the ocean surface or mixed layer.
 Samples labelled as "thermocline" or "sub-thermocline" were excluded.
@@
 
-**Coral reefs.** Today, shallow warm-water coral reefs are limited to tropical and subtropical latitudes (\~34&deg;&nbsp;N -- 32&deg;&nbsp;S), with minimum sea surface temperature tolerances (\~18&deg;C) being the primary constraint on this distribution [@johannes1983; @kleypas1999; @yamano2001].
+**Coral reefs.** Today, shallow warm-water coral reefs are limited to tropical and subtropical latitudes (\~34&deg;&nbsp;N--32&deg;&nbsp;S), with minimum sea surface temperature tolerances (\~18&deg;C) being the primary constraint on this distribution [@johannes1983; @kleypas1999; @yamano2001].
 As coral reefs reside at the upper thermal limit of the oceans today, their maximum sea surface temperature tolerance is less well-constrained, with some studies suggesting up to 35.6&deg;C in the geological past [@jones2022reef].
@@
 During the Eocene, coral communities and reefs expanded across tropical and temperate latitudes, with communities found up to palaeolatitudes of 43&deg;&nbsp;N [@Zamagni2012].
-Using a compilation of Paleocene -- early Eocene coral reefs and community localities [@Zamagni2012], we generated quantitative sea surface temperature estimates for the ECCO.
+Using a compilation of Paleocene--early Eocene coral reefs and community localities [@Zamagni2012], we generated quantitative sea surface temperature estimates for the ECCO.
 To do so, we extracted localities from the compilation that are inferred to be Ilerdian (early Eocene) coral reefs, and that could be confidently assigned to the EECO.
@@
 The modern, latitudinal gradient, modelled with all data, spans `r modern_gradient$"50%"`&deg;C from the equator to the poles.
-When modelled with only the early Eocene sampling locations, the median temperature gradient is `r modern_eocene_gradient$"50%"` (95% CI: `r modern_eocene_gradient$"2.5%"` -- `r modern_eocene_gradient$"97.5%"`)&deg;C.
+When modelled with only the early Eocene sampling locations, the median temperature gradient is `r modern_eocene_gradient$"50%"` (95% CI: `r modern_eocene_gradient$"2.5%"`--`r modern_eocene_gradient$"97.5%"`)&deg;C.
 A random sampling of modern temperatures at latitudes corresponding to the sampled Eocene palaeolatitudes would thus allow the reconstruction of the modern latitudinal temperature gradient with good accuracy, with expected deviations from the actual gradient as low as `r modern_eocene_gradient_diff$"2.5%"` to `r modern_eocene_gradient_diff$"97.5%"`&deg;C (95 % CI).
@@
 The modern, global mean sea surface temperature (GMsST) estimate is `r motemp_global$"50%"`&deg;C when using the full, modern data set.
-When reduced to the early Eocene sampling distribution, the modern GMsST is estimated at `r moeotemp_global$"50%"` (`r moeotemp_global$"2.5%"` -- `r moeotemp_global$"97.5%"`)&deg;C.
+When reduced to the early Eocene sampling distribution, the modern GMsST is estimated at `r moeotemp_global$"50%"` (`r moeotemp_global$"2.5%"`--`r moeotemp_global$"97.5%"`)&deg;C.
 The deviations from the full modern estimate are again very low, with a 95 % credible interval of `r moeotemp_global_diff$"2.5%"` to `r moeotemp_global_diff$"97.5%"`&deg;C.
@@
 4).
-Modelled, median equatorial temperatures are `r tdiff_lat0$"50%"` (95% CI: `r tdiff_lat0$"2.5%"` -- `r tdiff_lat0$"97.5%"`)&deg;C higher for the EECO, and polar temperatures are `r tdiff_lat90$"50%"` (`r tdiff_lat90$"2.5%"` -- `r tdiff_lat90$"97.5%"`)&deg;C higher.
-This results in a flattened latitudinal temperature gradient of `r eocene_gradient$"50%"` (`r eocene_gradient$"2.5%"` -- `r eocene_gradient$"97.5%"`)&deg;C for the EECO, as opposed to `r modern_gradient$"50%"`&deg;C for the modern.
-To facilitate the comparison with latitudinal gradients reported in the literature, which sometimes do not report temperatures at very high latitudes, we report also the EECO gradient between the equator and the modern-day polar circle (66.6&deg;), which is a lot lower at `r eocene_gradient_polar_circle$"50%"` (`r eocene_gradient_polar_circle$"2.5%"` -- `r eocene_gradient_polar_circle$"97.5%"`)&deg;C.
+Modelled, median equatorial temperatures are `r tdiff_lat0$"50%"` (95% CI: `r tdiff_lat0$"2.5%"`--`r tdiff_lat0$"97.5%"`)&deg;C higher for the EECO, and polar temperatures are `r tdiff_lat90$"50%"` (`r tdiff_lat90$"2.5%"`--`r tdiff_lat90$"97.5%"`)&deg;C higher.
+This results in a flattened latitudinal temperature gradient of `r eocene_gradient$"50%"` (`r eocene_gradient$"2.5%"`--`r eocene_gradient$"97.5%"`)&deg;C for the EECO, as opposed to `r modern_gradient$"50%"`&deg;C for the modern.
+To facilitate the comparison with latitudinal gradients reported in the literature, which sometimes do not report temperatures at very high latitudes, we report also the EECO gradient between the equator and the modern-day polar circle (66.6&deg;), which is a lot lower at `r eocene_gradient_polar_circle$"50%"` (`r eocene_gradient_polar_circle$"2.5%"`--`r eocene_gradient_polar_circle$"97.5%"`)&deg;C.
 
 The high variability of EECO palaeotemperature proxies, particularly in the mid-latitudes, and the scarcity of high-latitude data, result in substantial uncertainties in the modelled temperature gradient.
-This is reflected in the residual standard deviation ($\sigma$) of the EECO gradient -- `r eeco_sdy$"50%"` (`r eeco_sdy$"2.5%"` -- `r eeco_sdy$"97.5%"`)&deg;C -- which is more than double the  $\sigma$ for the modern gradient, `r modern_sdy$"50%"`.
+This is reflected in the residual standard deviation ($\sigma$) of the EECO gradient--`r eeco_sdy$"50%"` (`r eeco_sdy$"2.5%"`--`r eeco_sdy$"97.5%"`)&deg;C--which is more than double the  $\sigma$ for the modern gradient, `r modern_sdy$"50%"`.
 This is illustrated by the drastic departure of some of the proxy data from the gradient estimates (Fig.
@@
 
-The early Eocene GMST is estimated at `r eotemp_global$"50%"` (`r eotemp_global$"2.5%"` -- `r eotemp_global$"97.5%"`)&deg;C, `r  round(as.numeric(eotemp_global$"50%") - as.numeric(motemp_global$"50%"),1)`&deg;C higher than the modern.
-A model run excluding the ecological proxies increases the GMST by `r eoceneproxy_global_diff$"50%"` (`r eoceneproxy_global_diff$"2.5%"` -- `r eoceneproxy_global_diff$"97.5%"`)&deg;C.
+The early Eocene GMST is estimated at `r eotemp_global$"50%"` (`r eotemp_global$"2.5%"`--`r eotemp_global$"97.5%"`)&deg;C, `r  round(as.numeric(eotemp_global$"50%") - as.numeric(motemp_global$"50%"),1)`&deg;C higher than the modern.
+A model run excluding the ecological proxies increases the GMST by `r eoceneproxy_global_diff$"50%"` (`r eoceneproxy_global_diff$"2.5%"`--`r eoceneproxy_global_diff$"97.5%"`)&deg;C.
 The median modelled temperature is higher near the equator and in high latitudes when excluding the ecological proxies, with a flattened median gradient of `r eocene_proxy_gradient$"50%"`&deg;C (Fig.
@@
 Due to the limited spatial coverage of the early Eocene proxy record, and due to the added model complexity of simultaneously estimating a model across both hemispheres, we pooled the proxy data across both hemispheres.
-Applying the model separately within each hemisphere results in substantial differences in hemispherical, average temperatures, with the Southern Hemisphere being warmer by `r eeco_south_north_global_diff$"50%"` (`r eeco_south_north_global_diff$"2.5%"` -- `r eeco_south_north_global_diff$"97.5%"`)&deg;C.
-The inferred latitudinal gradient is somewhat steeper in the Northern Hemisphere (steeper by `r eocene_north_south_gradient_diff$"50%"`&deg;C, although the 95% CI spans `r eocene_north_south_gradient_diff$"2.5%"` -- `r eocene_north_south_gradient_diff$"97.5%"`&deg;C), but the large uncertainties associated with both gradients, and the lack of polar proxy data in the Southern Hemisphere preclude a more precise statement (see Fig.
+Applying the model separately within each hemisphere results in substantial differences in hemispherical, average temperatures, with the Southern Hemisphere being warmer by `r eeco_south_north_global_diff$"50%"` (`r eeco_south_north_global_diff$"2.5%"`--`r eeco_south_north_global_diff$"97.5%"`)&deg;C.
+The inferred latitudinal gradient is somewhat steeper in the Northern Hemisphere (steeper by `r eocene_north_south_gradient_diff$"50%"`&deg;C, although the 95% CI spans `r eocene_north_south_gradient_diff$"2.5%"`--`r eocene_north_south_gradient_diff$"97.5%"`&deg;C), but the large uncertainties associated with both gradients, and the lack of polar proxy data in the Southern Hemisphere preclude a more precise statement (see Fig.
 S4).
@@
 
-Our results show that our Bayesian model can be used to reconstruct different types of latitudinal SST gradients from proxy data with moderate sample sizes (*n* = 10 -- 34) and patchy sampling distributions (Fig.
+Our results show that our Bayesian model can be used to reconstruct different types of latitudinal SST gradients from proxy data with moderate sample sizes (*n* = 10--34) and patchy sampling distributions (Fig.
 2).
@@
 The geochemical proxy record and ecological data indicate that the latitudinal SST gradient of the early Eocene climatic optimum was significantly shallower than the modern [@Huber2011], but beyond that, there is little agreement.
-Earlier, reconstructed early Eocene and EECO SST gradients range from 7 -- 21&deg;C (Table 1); a more recent reconstruction that includes terrestrial air and sea surface temperatures arrives at a gradient of ~13&deg;C [@Inglis2020].
+Earlier, reconstructed early Eocene and EECO SST gradients range from 7--21&deg;C (Table 1); a more recent reconstruction that includes terrestrial air and sea surface temperatures arrives at a gradient of ~13&deg;C [@Inglis2020].
 Our median poles-to-equator gradient estimate is similar at `r eocene_gradient$"50%"`&deg;C, but a lot less steep when taking the equator-to-polar-circle estimate, `r eocene_gradient_polar_circle$"50%"`, as the geochemical proxy data suggest high temperatures up to latitudes of ~ 60&deg.
@@
 
-Recent, marine GMST estimates of the EECO and of the early Eocene range from 23.4 -- 37.1&deg;C, with the lowest GMSTs being derived from &delta;^18^O, and the higher estimates including TEX~86~ [@Inglis2020].
+Recent, marine GMST estimates of the EECO and of the early Eocene range from 23.4--37.1&deg;C, with the lowest GMSTs being derived from &delta;^18^O, and the higher estimates including TEX~86~ [@Inglis2020].
 Many studies include both marine and terrestrial proxies to derive GMST estimates, but despite great differences in proxy selection and in the calculation of global average temperatures, many recent estimates fall in the range of 27 - 29.5&deg;C [@Hansen2013; @Caballero2013; @Cramwinckel2018; @Zhu2019], similar to our median GMST estimate of `r eotemp_global$"50%"`&deg;C.

### 2023-09-22 manuscript/main.qmd
reviewer 2 minor notes corrected

@@
 
-We apply this model to the record of the early Eocene climatic optimum (EECO), combining a compilation of geochemical proxies [@Hollis2019], mangrove communities [@Popescu2021], and coral reefs [@Zamagni2012], using a nearest-living-relative approach [e.g.
+We apply this model to the record of the early Eocene climatic optimum (EECO), combining a compilation of geochemical proxies [@Hollis2019], mangrove communities [@Popescu2021], and coral reefs [@Zamagni2012].
+We use a nearest-living-relative approach [e.g.
 @Greenwood2017] to establish broad temperature ranges for the ecological data.
@@
 This filtering resulted in most &delta;^18^O samples being excluded from the dataset (retaining 8 out of 152).
-After data filtering, `r n_geochem` geochemical proxy samples from `r n_geochem_location` locations remained.
+After data filtering, `r n_geochem` geochemical proxy samples from `r n_geochem_location` locations remained (Fig.
+1).
 For a detailed description of each proxy see @Hollis2019.
@@
 Subsequently, we used statistically derived temperature limits (minimum = 21&deg;C, average = 27.6&deg;C, maximum = 29.5&deg;C) from the published literature [@kleypas1999] to define a normal probability distribution of potential temperature values for coral reef localities.
-This normal probability distribution was defined with a mean of 27.6&deg;C and a standard deviation of 2.125&deg;C, placing 97.5% of the probability density above the minimum.
+This normal probability distribution was defined with a mean of 27.6&deg;C and a standard deviation of 2.125&deg;C, which places the minimum (21&deg;C) at the lower end of the 95% highest density interval of that distribution.
 As the distribution of modern corals is skewed towards warmer temperatures, this approach results in 16.5% of the probability being placed on temperatures \> 29.5&deg;C, allowing for the possibility that Eocene coral reefs were adapted to warmer conditions than present-day coral reefs.
@@
 **Mangroves.** Mangroves are distributed throughout the tropics and subtropics today.
-While factors besides sea surface temperatures (SST) influence the distribution of mangroves, empirical, lower temperature limits have been established for the genera *Avicennia* (15.6&deg;C) and *Rhizophora* (20.7&deg;C) [@Quisthoudt2012].
+While factors besides sea surface temperatures (SST) influence the distribution of mangroves, empirical lower temperature limits have been established for the genera *Avicennia* (15.6&deg;C) and *Rhizophora* (20.7&deg;C) [@Quisthoudt2012].
 Both *Avicennia* and members of the Rhizophoraceae family were widespread and co-occurred across tropical and temperate latitudes in the early Eocene.
@@
 
-1)  *Avicennia*-only assemblages ($n=2$): the absence of Rhizophoraceae is indicative of temperatures being between 15.6&deg;C (lower temperature limt of *Avicennia*) and 20.7&deg;C (lower temperature limit of *Rhizophora*).
-However, a value of 22.5&deg;C is ascribed as the upper temperature limit here as *Rhizophora* is rare below this temperature.
+1)  *Avicennia*-only assemblages ($n=2$): the absence of Rhizophoraceae is indicative of temperatures between 15.6&deg;C (lower temperature limt of *Avicennia*) and 20.7&deg;C (lower temperature limit of *Rhizophora*).
+However, a value of 22.5&deg;C is assumed as the upper temperature limit here as *Rhizophora* is rare below this temperature.
 We define the *Avicennia*-only temperature distribution as a normal distribution with a mean of 19.05&deg;C and a standard deviation of 1.725&deg;C, resulting in 95% of the probability density being placed within the temperature limits.
@@
 ```
-modelled sea surface temperature estimates were generated with [@eq-nu], calculating the sea surface temperatures at any latitude with the parameter estimates of each iteration from the posterior.
-The median and 95% CI of temperatures where then taken from all temperature estimates obtained at the latitudes of interest.
+Modelled sea surface temperature estimates were generated with [@eq-nu], calculating the sea surface temperatures at any latitude with the parameter estimates of each iteration from the posterior.
+The median and 95% CI of temperatures were then taken from all temperature estimates obtained at the latitudes of interest.
 
-The latitudinal gradient is calculated as the difference between the modelled temperature at the equator (0&deg; latitude) and at the poles (90&deg; absolute latitude).
-To facilitate comparison with earlier estimates, we also calculate the gradient with the temperature at the polar circle (66.6&deg; absolute latitude) being used instead of the temperature at the poles.
-Given the sigmoidal shape of the modern as well as the Eocene gradient (see Fig.
-4), these results are broadly comparable to a gradient inferred from the zonal average of equatorial and high-latitude temperatures, as has been done in some earlier studies [@Evans2018].
+The latitudinal gradient was calculated as the difference between the modelled temperature at the equator (0&deg; latitude) and at the poles (90&deg; absolute latitude).
+To facilitate comparison with earlier estimates, we also calculated the gradient with the temperature at the polar circle (66.6&deg; absolute latitude) being used instead of the temperature at the poles.
 

### 2023-09-22 manuscript/main.qmd
add spatial variability in SST to discussion

@@
 Our results confirm the presence of a flattened latitudinal temperature gradient and unusually high polar temperatures during the EECO, which is supported by high-latitude ecological data.
-We show that integrating multiple types of proxy data, and adequate prior information, has the potential to enhance quantiative palaeoclimate reconstructions, allowing for more accurate temperature estimates from sparse data.
+We show that integrating multiple types of proxy data, and adequate prior information, has the potential to enhance quantiative palaeoclimate reconstructions, allowing for more accurate temperature estimates from  data with incomplete spatial sampling.
 
@@
 The very high variability of the proxy record in mid-latitudes results in large uncertainties on the shape of temperature gradient and on the GMST.
-Biases and errors in the proxy reconstructions likely contribute to the observed variability, as geochemical proxies reflect many other factors besides seawater temperature [@Hollis2019].
+Some of this variability may stem from spatial variability in SSTs, as can be observed in the modern (Fig.
+3), e.g.
+due to ocean circulation [@Rahmstorf2002].
+Biases and errors in the proxy reconstructions also likely contribute to the observed variability, as geochemical proxies reflect many other factors besides seawater temperature [@Hollis2019].
 Despite excluding &delta;^18^O measurements from recrystallised fossils, systematic offsets remain between mostly warm temperatures derived from TEX~86~, and cooler temperatures derived from &delta;^18^O, &Delta;~47~, and the ecological proxies.

### 2023-09-22 manuscript/main.qmd
remove degC behind brackets

@@
 4).
-Modelled, median equatorial temperatures are `r tdiff_lat0$"50%"` (95% CI: `r tdiff_lat0$"2.5%"`--`r tdiff_lat0$"97.5%"`)&deg;C higher for the EECO, and polar temperatures are `r tdiff_lat90$"50%"` (`r tdiff_lat90$"2.5%"`--`r tdiff_lat90$"97.5%"`)&deg;C higher.
-This results in a flattened latitudinal temperature gradient of `r eocene_gradient$"50%"` (`r eocene_gradient$"2.5%"`--`r eocene_gradient$"97.5%"`)&deg;C for the EECO, as opposed to `r modern_gradient$"50%"`&deg;C for the modern.
-To facilitate the comparison with latitudinal gradients reported in the literature, which sometimes do not report temperatures at very high latitudes, we report also the EECO gradient between the equator and the modern-day polar circle (66.6&deg;), which is a lot lower at `r eocene_gradient_polar_circle$"50%"` (`r eocene_gradient_polar_circle$"2.5%"`--`r eocene_gradient_polar_circle$"97.5%"`)&deg;C.
+Modelled, median equatorial temperatures are `r tdiff_lat0$"50%"`&deg;C (95% CI: `r tdiff_lat0$"2.5%"`--`r tdiff_lat0$"97.5%"`&deg;C) higher for the EECO, and polar temperatures are `r tdiff_lat90$"50%"`&deg;C (`r tdiff_lat90$"2.5%"`--`r tdiff_lat90$"97.5%"`&deg;C) higher.
+This results in a flattened latitudinal temperature gradient of `r eocene_gradient$"50%"`&deg;C (`r eocene_gradient$"2.5%"`--`r eocene_gradient$"97.5%"`&deg;C) for the EECO, as opposed to `r modern_gradient$"50%"`&deg;C for the modern.
+To facilitate the comparison with latitudinal gradients reported in the literature, which sometimes do not report temperatures at very high latitudes, we report also the EECO gradient between the equator and the modern-day polar circle (66.6&deg;), which is a lot lower at `r eocene_gradient_polar_circle$"50%"`&deg;C (`r eocene_gradient_polar_circle$"2.5%"`--`r eocene_gradient_polar_circle$"97.5%"`&deg;C).
 
 The high variability of EECO palaeotemperature proxies, particularly in the mid-latitudes, and the scarcity of high-latitude data, result in substantial uncertainties in the modelled temperature gradient.
-This is reflected in the residual standard deviation ($\sigma$) of the EECO gradient--`r eeco_sdy$"50%"` (`r eeco_sdy$"2.5%"`--`r eeco_sdy$"97.5%"`)&deg;C--which is more than double the  $\sigma$ for the modern gradient, `r modern_sdy$"50%"`.
+This is reflected in the residual standard deviation ($\sigma$) of the EECO gradient--`r eeco_sdy$"50%"` (`r eeco_sdy$"2.5%"`--`r eeco_sdy$"97.5%"`)--which is more than double the  $\sigma$ for the modern gradient, `r modern_sdy$"50%"`.
 This is illustrated by the drastic departure of some of the proxy data from the gradient estimates (Fig.
@@
 
-The early Eocene GMST is estimated at `r eotemp_global$"50%"` (`r eotemp_global$"2.5%"`--`r eotemp_global$"97.5%"`)&deg;C, `r  round(as.numeric(eotemp_global$"50%") - as.numeric(motemp_global$"50%"),1)`&deg;C higher than the modern.
-A model run excluding the ecological proxies increases the GMST by `r eoceneproxy_global_diff$"50%"` (`r eoceneproxy_global_diff$"2.5%"`--`r eoceneproxy_global_diff$"97.5%"`)&deg;C.
+The early Eocene GMST is estimated at `r eotemp_global$"50%"`&deg;C (`r eotemp_global$"2.5%"`--`r eotemp_global$"97.5%"`&deg;C), `r  round(as.numeric(eotemp_global$"50%") - as.numeric(motemp_global$"50%"),1)`&deg;C higher than the modern.
+A model run excluding the ecological proxies increases the GMST by `r eoceneproxy_global_diff$"50%"`&deg;C (`r eoceneproxy_global_diff$"2.5%"`--`r eoceneproxy_global_diff$"97.5%"`&deg;C).
 The median modelled temperature is higher near the equator and in high latitudes when excluding the ecological proxies, with a flattened median gradient of `r eocene_proxy_gradient$"50%"`&deg;C (Fig.
@@
 Due to the limited spatial coverage of the early Eocene proxy record, and due to the added model complexity of simultaneously estimating a model across both hemispheres, we pooled the proxy data across both hemispheres.
-Applying the model separately within each hemisphere results in substantial differences in hemispherical, average temperatures, with the Southern Hemisphere being warmer by `r eeco_south_north_global_diff$"50%"` (`r eeco_south_north_global_diff$"2.5%"`--`r eeco_south_north_global_diff$"97.5%"`)&deg;C.
-The inferred latitudinal gradient is somewhat steeper in the Northern Hemisphere (steeper by `r eocene_north_south_gradient_diff$"50%"`&deg;C, although the 95% CI spans `r eocene_north_south_gradient_diff$"2.5%"`--`r eocene_north_south_gradient_diff$"97.5%"`&deg;C), but the large uncertainties associated with both gradients, and the lack of polar proxy data in the Southern Hemisphere preclude a more precise statement (see Fig.
+Applying the model separately within each hemisphere results in substantial differences in hemispherical, average temperatures, with the Southern Hemisphere being warmer by `r eeco_south_north_global_diff$"50%"`&deg;C (`r eeco_south_north_global_diff$"2.5%"`--`r eeco_south_north_global_diff$"97.5%"`&deg;C).
+The inferred latitudinal gradient is somewhat steeper in the Northern Hemisphere (steeper by `r eocene_north_south_gradient_diff$"50%"`&deg;C, although the 95% CI of that difference spans `r eocene_north_south_gradient_diff$"2.5%"`--`r eocene_north_south_gradient_diff$"97.5%"`&deg;C), but the large uncertainties associated with both gradients, and the lack of polar proxy data in the Southern Hemisphere preclude a more precise statement (see Fig.
 S4).
@@
 Earlier, reconstructed early Eocene and EECO SST gradients range from 7--21&deg;C (Table 1); a more recent reconstruction that includes terrestrial air and sea surface temperatures arrives at a gradient of ~13&deg;C [@Inglis2020].
-Our median poles-to-equator gradient estimate is similar at `r eocene_gradient$"50%"`&deg;C, but a lot less steep when taking the equator-to-polar-circle estimate, `r eocene_gradient_polar_circle$"50%"`, as the geochemical proxy data suggest high temperatures up to latitudes of ~ 60&deg.
+Our median poles-to-equator gradient estimate is similar at `r eocene_gradient$"50%"`&deg;C, but a lot less steep when taking the equator-to-polar-circle estimate, `r eocene_gradient_polar_circle$"50%"`&deg;C, as the geochemical proxy data suggest high temperatures up to latitudes of ~ 60&deg.
 Both geochemical and ecological shallow water data indicate that inferred SST gradients based on tropical, shallow water and deep water samples [@Cramwinckel2018; @Evans2018] may overestimate the SST gradient of the early Eocene greenhouse world.

### 2023-09-23 manuscript/main.qmd
variability mid and high latitudes[@REFERENCE].
lewis adjustments main

@@
 ```
-Accurately reconstructing large-scale palaeoclimate patterns from sparse local records is critical for understanding the evolution of Earth's climate.
+Accurately reconstructing large-scale palaeoclimatic patterns from sparse local records is critical for understanding the evolution of Earth's climate.
 Particular challenges arise from the patchiness, uneven spatial distribution, and disparate nature of palaeoclimatic proxy records.
@@
 Our results confirm the presence of a flattened latitudinal temperature gradient and unusually high polar temperatures during the EECO, which is supported by high-latitude ecological data.
-We show that integrating multiple types of proxy data, and adequate prior information, has the potential to enhance quantiative palaeoclimate reconstructions, allowing for more accurate temperature estimates from  data with incomplete spatial sampling.
+We show that integrating multiple types of proxy data, and adequate prior information, has the potential to enhance quantitative palaeoclimatic reconstructions, improving temperature estimates from datasets with limited spatial sampling.
 
@@
 
-Palaeoclimate, latitudinal temperature gradients, temperature proxies, Eocene, spatial bias, Bayesian
+Palaeoclimate, latitudinal temperature gradients, temperature proxies, Eocene, sampling bias, Bayesian
 
@@
 
-Understanding the long-term evolution of Earth's climate system and contextualising current global warming relies on accurate reconstructions of past climates [@Royer2004; @Burke2018; @Tierney2020].
-Recent advances in the synthesis of palaeoclimate data [e.g.
+Understanding the long-term evolution of Earth's climate system and contextualising contemporary global warming relies on accurate reconstructions of past climates [@Royer2004; @Burke2018; @Tierney2020].
+Recent advances in the synthesis of palaeoclimatic data [e.g.
 @Veizer2015; @Hollis2019; @song2019; @Grossman2022; @Judd2022] are offering unprecedented insights into the complex and dynamic nature of the Earth's climate system, yet a fundamental challenge remains: the proxy record of past climates is spatially incomplete and afflicted by imperfect preservation and uneven sampling [@judd2020; @Jones2022; @Judd2022].
 
-Acknowledging the assumptions and limitations inherent in geochemical temperature proxies, e.g.
-experimentally derived calibrations, influences from seasonality, dissolution effects and differential preservation [e.g.
-@Tierney2017], can allow for robust estimates of palaeotemperatures at a local scale.
+Acknowledging the assumptions and limitations inherent in geochemical temperature proxies, such as experimentally derived calibrations, influences from seasonality, dissolution effects and differential preservation [e.g.
+@Tierney2017], can enable robust estimates of palaeotemperature at local scales.
 However, recent work has demonstrated that spatial biases in the geochemical proxy record can lead to spurious estimates of regional (e.g.
@@
 However, in order to robustly infer regional or global-scale patterns from an incomplete record, spatial gaps must ultimately be bridged.
-One common approach, which requires no additional computation, is the spatial visualisation of proxy-derived temperatures against latitude, showing broad latitudinal temperature trends [@Hollis2019; @Vickers2021].
-Interpolation is also sometimes used to bridge spatial gaps in palaeoclimate data [@Taylor2004], taking advantage of the autoregressive nature of climatic data: much of the information on the climate of any given location is contained in the climate data of nearby locations [@Reynolds1994].
+One common approach, which requires no additional computation, is the spatial visualisation of proxy-derived temperatures against latitude, showing broad latitudinal temperature trends [e.g.
+@Hollis2019; @Vickers2021].
+Interpolation is also sometimes used to bridge spatial gaps in palaeoclimatic data [e.g.
+@Taylor2004], taking advantage of the autoregressive nature of climatic data: much of the information on the climate of any given location is contained in the climate data of nearby locations [@Reynolds1994].
 Adding to this, some proxy-based reconstructions use statistical modelling to infer palaeoclimatic patterns.
@@
 
-Spatial gaps in the palaeoclimate record can also be addressed through the integration of additional data.
+Spatial gaps in the palaeoclimatic record can also be addressed through the integration of additional data.
 For example, lithological and fossil data can be used to infer past climatic conditions based on analogous modern sediments [@Chandra2021], or based on the premise that the climatic requirements of ancient taxa, biological traits, or ecological communities were similar to those of their nearest modern relatives [@Peppe2011; @Royer2012; @Salonen2019].
@@
 
-Here, we present a novel Bayesian hierarchical model [see e.g.
-@Gelman2013; @Mcelreath2018] that combines quantitative proxies and ecological constraints into a fully quantitative model of the latitudinal gradient of sea surface temperatures, bridging spatial gaps in sparsely sampled climate data.
+Here, we present a novel Bayesian hierarchical model [e.g.
+@Gelman2013; @Mcelreath2018] that combines quantitative proxies and ecological constraints into a fully quantitative model of the latitudinal gradient of sea surface temperatures, bridging spatial gaps in sparsely sampled palaeoclimatic data.
 The Bayesian approach offers a powerful framework for integrating various sources of uncertainty and modelling complex hierarchical relationships, and is increasingly used in palaeoclimatic reconstructions [e.g.
 @Weitzel2019; @Yang2022; @Burgener2023].
-This model expands upon existing, spatially explicit palaeoclimatic reconstructions  by allowing for the integration of (1) prior information based on physical principles and the observed modern sea surface temperature distribution, and of (2) geochemical and ecological climate proxies in a common, quantitative framework.
+This model expands upon existing, spatially explicit palaeoclimatic reconstructions  by allowing for the integration of (1) prior information based on physical principles and the observed modern sea surface temperature distribution, and of (2) geochemical and ecological palaeoclimatic proxies in a common, quantitative framework.
 We chose a generalised logistic function to accurately infer the shape of the temperature gradient despite a patchy latitudinal coverage.
 This choice is motivated by the flexibility and ability of this function to approximate a variety of nonlinear patterns in the underlying temperature gradients, that other parametric approaches, such as lower order polynomials [e.g.
-@Bijl2009; @Keating-Bitoni2011], lack.
-We test the robustness of this method using down-sampled, simulated temperature gradients.
+@Bijl2009; @Keating-Bitonti2011], lack.
+We test the robustness of this method using down-sampled, simulated latitudinal temperature gradients.
 
@@
 The gradient values denote the SST difference between the equator and the polar circle, or other types of gradients.
-For comparison, a gradient derived from an atmosphere-ocean general circulation model (GCM) ensemble is also shown.
+For comparison, a gradient derived from an atmosphere-ocean general circulation model (GCM) ensemble, and a range of gradients from a model intercomparison project, are also shown.
 
@@
 
-Geochemical climate proxy data were extracted from a latest Paleocene and early Eocene compilation [@Hollis2019].
+Geochemical climatic proxy data were extracted from a latest Paleocene and early Eocene compilation [@Hollis2019].
 This compilation provides sea surface temperature data on four different geochemical proxies for reconstructing seawater temperature: &delta;^18^O, &Delta;~47~, Mg/Ca and TEX~86~.
 For our analyses, this dataset was restricted to the EECO (defined as 53.8--49.1 Ma) and samples originating from near the ocean surface or mixed layer.
-Samples labelled as "thermocline" or "sub-thermocline" were excluded.
+Consequently, samples labelled as "thermocline" or "sub-thermocline" were excluded.
 Recrystallised &delta;^18^O samples were also excluded as secondary diagenetic calcite precipitated after deposition can bias isotope measurements and offset temperature values [@Schrag1999].
@@
 
-The palaeogeographic distribution of geochemical and ecological data was reconstructed using the @merdith2021 plate rotation model via the palaeoverse R package [version 1.2.0, @jones2023].
+The palaeogeographic distribution of geochemical and ecological data was reconstructed using the @merdith2021 Global Plate Model via the palaeoverse R package [version 1.2.0, @jones2023].
 The midpoint age of the EECO (51.2 Ma), along with the present-day coordinates of geochemical and ecological data, were used for palaeogeographic reconstruction.
@@
 For example, one limb of a second-order polynomial as in @Bijl2009 can be approximated by increasing $M$ towards high latitudes, and decreasing $B$ to reduce the steepness of the curve.
-The model is designed for modelling the average gradient across both hemispheres, but can also be applied to individual hemispheres, to assess hemispherical differences (Supplementary Material).
+The model is designed for modelling the average gradient across both hemispheres, but can also be applied to individual hemispheres, to assess hemispherical differences (see Fig.
+S4).
 
@@
 
-We show an expanded model that includes uncertainties on individual temperature observations in the Supplementary Material.
+We show an expanded model that includes uncertainties on individual temperature observations in the Supplementary Material (Fig. S5).
 
@@
 
-**A.** Predicted seawater surface temperatures are not allowed to be $<< -2^\circ C$, the freezing point of sea water.
-The highest prior density of $A$ is placed around $0^\circ C$, and it slowly tapers off towards higher temperatures.
+**A.** Predicted seawater surface temperatures are not allowed to be $<< -2$&deg;C, the freezing point of sea water.
+The highest prior density of $A$ is placed around 0&deg;C, and it slowly tapers off towards higher temperatures.
 This shape is achieved by placing a skew-normal prior on the lower asymptote, specified as $$A \sim SN(\xi=-3.0,\omega=12,\alpha_{SN}=30),$${#eq-A}
@@
 To test whether our logistic regression model can adequately describe different latitudinal temperature gradients at various sample sizes, we used the empirical, modern gradient, representative of an icehouse climate, and generated three idealised gradients that emulate potential climatic states throughout Earth's geological history: extreme icehouse, icehouse (modern), greenhouse, and extreme greenhouse [@Frakes1992].
-The idealised gradients serve to test whether our model setup is able to infer gradients that are strongly different from the modern from a varying amount of samples.
+The idealised gradients serve to test whether our model setup is able to infer gradients that are strongly different from the modern from a varying number of samples.
 
-We created test data from these gradients as follows: We randomly sampled (1000 iterations) latitudes at sample sizes of 5, 10 and 20, with the probability of a latitude being sampled scaling with the decreasing surface area towards higher latitudes; i.e.
+We created test data from these gradients as follows: We randomly sampled (1000 iterations) latitudes at sample sizes of 5, 10 and 20, with the probability of a latitude being sampled scaling with the decreasing surface area towards higher latitudes, i.e.
 lower latitudes are sampled more frequently.
@@
 For each latitude, we inferred a location mean temperature from the gradients, adding random noise from a normal distribution with a standard deviation of 3.8, which is the average standard deviation of the EECO geochemical proxy data.
-With that, we aim to simulate randomly distributed errors in the proxy data, which could arise from miscalibrations, measurement errors, seasonal effects, ect.
+With that, we aim to simulate randomly distributed errors in the proxy data, which could arise from miscalibrations, measurement errors, seasonal effects, *ect*.
 We acknowledge that this approach cannot quantify the potential impact of systematic offsets that may bias all proxy data in the same direction, nor do we know whether a standard deviation of 3.8 is actual, average magnitude of uncertainty that the proxy compilation is afflicted with.
@@
 
-To test whether our model can accurately depict the shape of the modern sea surface temperature gradient, and to facilitate comparison with the Eocene gradient, we applied our model to annual sea surface mean temperatures from Bio-Oracle [@Assis2018], aggregated to a $1^\circ \times 1 ^\circ$ raster (*n* = 46,131).
+To test whether our model can accurately depict the shape of the modern sea surface temperature gradient, and to facilitate comparison with the Eocene gradient, we applied our model to mean annual sea surface temperatures from Bio-Oracle [@Assis2018], aggregated to a spatial grid resolution of $1^\circ \times 1 ^\circ$ (*n* = 46,131).
 The $R^2$ for the modern gradient was calculated as above [@Gelman2019], comparing the modelled gradient and the empirical temperature averages in 1&deg; latitude bins.
@@
 Posterior inference on the modern gradient is based on four chains with 60,000 iterations each, 10,000 of which were discarded as burn-in.
-Every 10th iteration was retained, resulting in a total of 20,000 iterations with low autocorrelation.
+Every $10^{th}$ iteration was retained, resulting in a total of 20,000 iterations with low autocorrelation.
 The re-sampled, simulated gradients were modelled in one chain with 10,000 iterations for each of the 1,000 random samples.
-5,000 iterations each were discarded as burn-in, and every 25th iteration was kept, resulting in a total of 200,000 iterations across all 1,000 model runs.
+5,000 iterations each were discarded as burn-in, and every $25^{th}$ iteration was kept, resulting in a total of 200,000 iterations across all 1,000 model runs.
 For the simulated gradients with an Eocene sampling distribution, a single chain with 250,000 iterations was used, thinned to 10,000 iterations after burn-in.
-For the Eocene model, we ran four chains with 600,000 iterations each, discarding 100,000 as burn-in and keeping every 100th iteration, as the hierarchical model structure results in higher autocorrelation of the chains.
+For the Eocene model, we ran four chains with 600,000 iterations each, discarding 100,000 as burn-in and keeping every $100^{th}$ iteration, as the hierarchical model structure results in higher autocorrelation of the chains.
 The Eocene posterior inference is thus based on a total of 20,000 iterations with low autocorrelation (effective multivariate sample size for $A$, $K$, $M$ and $B$ is \> 18,000).
@@
 Each column depicts a different reconstruction for given sample sizes: 5, 10, 20, and 34 (latitudes of EECO samples).
-Each row depicts a different simulated latitudinal temperature gradient that represents idealised climatic states: extreme icehouse, icehouse, greenhouse, and extreme greenhouse.
-The black line illustrates the simulated gradient.
+Each row depicts a different latitudinal temperature gradient that represents idealised or observed climatic states: idealsed extreme icehouse, greenhouse, and extreme greenhouse gradients, and the modern gradient, which represents an icehouse state.
+The black line illustrates the original gradient.
 The blue line depicts the reconstructed gradient represented by the median sea surface temperature value estimated from 1,000 model runs with different random samples (first three columns), and a single run with the EECO latitudinal sampling distribution (fourth column).
@@
 
-Our Bayesian model is able to accurately model a range of idealised temperature gradients, ranging from extreme icehouse to 'super greenhouse' scenarios (Fig.
+Our Bayesian model is able to model a range of idealised temperature gradients, ranging from extreme icehouse to 'super greenhouse' scenarios (Fig.
 2).
-Random latitudinal sampling results in highly accurate reconstructions at a sample sizes as low as 10 for the icehouse scenarios (95 % CI of $R^2$ > 0.9).
-Greenhouse scenarios require additional samples to accurately predict high-latitude temperatures.
-This is because in the absence of high-latitude samples, the modelled gradient is heavily influenced by the priors, which we based on the modern, the only empirically known latitudinal temperature gradient.
-A sampling distribution resembling that of the early Eocene data set used in this study allows for a highly accurate reconstruction of even the extreme greenhouse scenario (95 % CI of $R^2$ > 0.95).
+Random latitudinal sampling results in accurate reconstructions for most random samples at sample sizes of 10 and 20 for the icehouse scenarios (median of $R^2$ > 0.9).
+Greenhouse scenarios perform somewhat worse due to the increased uncertainty at high latitudes (median of $R^2$ > 0.7 at sample sizes 10 and 20).
+A sampling distribution resembling that of the early Eocene dataset used in this study allows for accurate reconstruction of all scenarios, although the $R^2$ is still relatively low in the extreme greenhouse scenario, as a perfectly flat gradient prediction by the model would result in an $R^2$ of 0, despite the original gradient being very flat.
+This also explains the low lower bounds of the 95% credible intervals in the greenhouse scenarios.
 
@@
 The empirical gradient spans `r modern_gradient_empirical`&deg;C from the equator to the poles, the modelled gradient is only slightly higher at `r modern_gradient$"50%"`&deg;C.
-The modern, global mean temperature (GMST) based on our modelled, median gradient is `r motemp_global$"50%"`&deg;C, very similar to the GMST derived from the empirical median gradient (`r motemp_empirical_global`&deg;C).
+The modern, global mean temperature (GMST) based on our modelled, median gradient is `r motemp_global$"50%"`&deg;C, which is nearly equal to the GMST derived from the empirical median gradient (`r motemp_empirical_global`&deg;C).
 
@@
 This results in a flattened latitudinal temperature gradient of `r eocene_gradient$"50%"`&deg;C (`r eocene_gradient$"2.5%"`--`r eocene_gradient$"97.5%"`&deg;C) for the EECO, as opposed to `r modern_gradient$"50%"`&deg;C for the modern.
-To facilitate the comparison with latitudinal gradients reported in the literature, which sometimes do not report temperatures at very high latitudes, we report also the EECO gradient between the equator and the modern-day polar circle (66.6&deg;), which is a lot lower at `r eocene_gradient_polar_circle$"50%"`&deg;C (`r eocene_gradient_polar_circle$"2.5%"`--`r eocene_gradient_polar_circle$"97.5%"`&deg;C).
+To facilitate the comparison with latitudinal gradients reported in the literature, which sometimes do not report temperatures at very high latitudes, we report also the EECO gradient between the equator and the modern-day polar circle (66.6&deg; latitude), which is markedly lower at `r eocene_gradient_polar_circle$"50%"`&deg;C (`r eocene_gradient_polar_circle$"2.5%"`--`r eocene_gradient_polar_circle$"97.5%"`&deg;C).
 
 The high variability of EECO palaeotemperature proxies, particularly in the mid-latitudes, and the scarcity of high-latitude data, result in substantial uncertainties in the modelled temperature gradient.
-This is reflected in the residual standard deviation ($\sigma$) of the EECO gradient--`r eeco_sdy$"50%"` (`r eeco_sdy$"2.5%"`--`r eeco_sdy$"97.5%"`)--which is more than double the  $\sigma$ for the modern gradient, `r modern_sdy$"50%"`.
+This is reflected in the residual standard deviation ($\sigma$) of the EECO gradient--`r eeco_sdy$"50%"`&deg;C (`r eeco_sdy$"2.5%"`--`r eeco_sdy$"97.5%"`&deg;C)--which is more than double the  $\sigma$ for the modern gradient, `r modern_sdy$"50%"`&deg;C.
 This is illustrated by the drastic departure of some of the proxy data from the gradient estimates (Fig.
@@
 Estimates of the median, latitudinal sea surface temperature gradients of the early Eocene climatic optimum (purple line) and of the present-day (turquoise), both estimated with the Bayesian model.
-The purple ribbon (shading) depict the 95% credible interval of the Eocene gradient, the uncertainty of the modern gradient is too low to be visible.
+The purple ribbon (shading) depicts the 95% credible interval of the Eocene gradient, the uncertainty of the modern gradient is too low to be visible.
 Points within the plot depict the geochemical (e.g.
@@
 2).
-This is an advancement over previously used linear, quadratic or Gaussian approximations [e.g.
+This is an advancement over previously used linear, quadratic, or Gaussian approximations [e.g.
 @Bijl2009; @Tierney2017], which can fit only specific types of gradients.
@@
 @Inglis2020].
-This means that the idiosyncrasies of a patchy proxy record, potentially afflicted with measurement errors, calibration errors, and palaeogeographic and temporal uncertainty, dictate the reconstruction of large-scale climate patterns, without the option of including additional knowledge (e.g.
+This means that the idiosyncrasies of a patchy proxy record, potentially afflicted with measurement errors, calibration errors, and palaeogeographic and temporal uncertainty [e.g.
+@Buffan2023], dictate the reconstruction of large-scale climatic patterns, without the option of including additional knowledge (e.g.
 that latitudinal temperature gradients should be broadly negative).
 
-In contrast, our Bayesian, parametric model allows for the inclusion of informative priors on the model parameters.
+In contrast, our Bayesian parametric model allows for the inclusion of informative priors on the model parameters.
 The modelled sea surface temperature gradient thus does not strictly follow the proxy data, but instead represents a compromise between the data and prior knowledge.
@@
 For example, the greater amount of solar radiation per unit area in low latitudes causes Earth's latitudinal temperature gradient to be broadly negative [@Beer2008].
-The ease with which such prior information can be integrated is a major advantage of our method, as the shape of the modelled gradient is controlled by four parameters which clearly relate to its magnitude, steepness and the latitude of its greatest steepness.
+The ease with which such prior information can be integrated is a major advantage of our method, as the shape of the modelled gradient is controlled by four parameters which clearly relate to its magnitude, steepness, and the latitude of its greatest steepness.
 
@@
 By modelling the temperature variation across latitudes, a complete temperature distribution along a latitudinal axis can be obtained, filling in gaps in the proxy record through inter- or extrapolation.
-This eliminates the common problem that specific climate zones dominate the proxy record.
+This eliminates the common problem that specific climatic zones dominate the proxy record.
 Reconstructing the GMST directly from the proxies would lead to an estimate biased towards the well-sampled latitudes.
@@
 
-## The role of ecological constraints in palaeoclimate reconstructions
+## The role of ecological constraints in palaeoclimatic reconstructions
 
-Our results further exemplify how incorporating quantified ecological temperature constraints can provide more precise temperature reconstructions than geochemical proxies alone, adding to the advances in palaeoclimate reconstructions achieved by integrating lithological data [@Scotese2021; @Burgener2023].
-Combining the occurrences of climate-sensitive plant communities [@Greenwood1995], reptiles [@Markwick2007], leaf shapes [@Peppe2011], with geochemical proxies offers substantial potential for improving quantitative palaeoclimate reconstructions across the Phanerozoic.
-Our modelling framework offers a straightforward, efficient way of integrating ecological climate data with other proxy data: The hierarchical model structure accounts for variation of temperature estimates from proxies at individual localities, which is treated equivalent to the uncertainty associated with the ecological temperature proxies.
+Our results further exemplify how incorporating quantified ecological temperature constraints can provide more precise temperature reconstructions than geochemical proxies alone, adding to the advances in palaeoclimatic reconstructions achieved by integrating lithological data [@Scotese2021; @Burgener2023].
+Combining the occurrences of climate-sensitive plant communities [@Greenwood1995], reptiles [@Markwick2007], and leaf shapes [@Peppe2011], with geochemical proxies offers substantial potential for improving quantitative palaeoclimatic reconstructions across the Phanerozoic.
+Our modelling framework offers a straightforward, efficient way of integrating ecological palaeoclimatic data with other proxy data: The hierarchical model structure accounts for variation of temperature estimates from proxies at individual localities, which is treated equivalent to the uncertainty associated with the ecological temperature proxies.
 A local temperature estimate, based on multiple geochemical proxies, thus has the same weight as a local temperature estimate obtained from the occurrence of a climate-sensitive plant community, whilst preserving the uncertainty associated with each estimate.
@@
 
-Our approach for deriving fully quantitative climate reconstructions from ecological data is borrowed from nearest living relative methods, commonly employed in terrestrial, Cenozoic climate reconstructions [@Fauquette2007; @Pross2012].
+Our approach for deriving fully quantitative climate reconstructions from ecological data is borrowed from nearest living relative methods, commonly employed in terrestrial, Cenozoic palaeoclimatic reconstructions [@Fauquette2007; @Pross2012].
 One major limitation to these methods is that the thermal preferences of taxa may have changed over time.
@@
 The absence of coral reefs in higher latitudes in the early Eocene could be due to requirements in irradiance, rather than temperature [@Muir2015].
-Tropical temperatures predicted by the geochemical proxy record indicate hotter-than-modern tropical temperatures for the early Eocene (Fig.&nbsp;S2), suggesting that the modern climate range of coral reefs may underestimate the early Eocene thermal niche for coral reefs.
+Tropical temperatures predicted by the geochemical proxy record indicate hotter-than-modern tropical temperatures for the early Eocene (Fig.&nbsp;S2), suggesting that the modern climatic range of coral reefs may underestimate the early Eocene thermal limits for coral reefs.
 We have tried to account for that possibility by widening the temperature probability distribution for coral reefs, but the predicted temperatures for the reef and mangrove sites still lie below the temperatures indicated by the geochemical proxy record (Fig.
@@
 Earlier, reconstructed early Eocene and EECO SST gradients range from 7--21&deg;C (Table 1); a more recent reconstruction that includes terrestrial air and sea surface temperatures arrives at a gradient of ~13&deg;C [@Inglis2020].
-Our median poles-to-equator gradient estimate is similar at `r eocene_gradient$"50%"`&deg;C, but a lot less steep when taking the equator-to-polar-circle estimate, `r eocene_gradient_polar_circle$"50%"`&deg;C, as the geochemical proxy data suggest high temperatures up to latitudes of ~ 60&deg.
+Our median poles-to-equator gradient estimate is similar at `r eocene_gradient$"50%"`&deg;C, but notably shallower when taking the equator-to-polar-circle estimate, `r eocene_gradient_polar_circle$"50%"`&deg;C, as the geochemical proxy data suggest high temperatures up to latitudes of ~ 60&deg.
 Both geochemical and ecological shallow water data indicate that inferred SST gradients based on tropical, shallow water and deep water samples [@Cramwinckel2018; @Evans2018] may overestimate the SST gradient of the early Eocene greenhouse world.
-Likewise, climate simulations using general circulation or Earth system models tend to estimate steeper gradients than most proxy records [Table 1\; @Pross2012; @Lunt2021]
+Likewise, palaeoclimatic simulations from General Circulation Models tend to estimate steeper gradients than most proxy records [Table 1\; @Pross2012; @Lunt2021]
 
@@
 S6).
+It is difficult to attribute this variability to seasonality alone, as temporal climate variability is also expected to be higher in mid and high latitudes[@Schwartz2008].
 Critically, however, the mangrove data strongly supports our inference of a flattened gradient independent of the geochemical proxy record.
@@
 The Bayesian hierarchical model presented here is able to reconstruct latitudinal gradients from both geochemical and ecological proxy data, while reflecting the uncertainty associated with the ecological temperature proxies, and accounting for the variation of multiple temperature estimates at individual localities.
-Using informative prior information allows for accurate temperature reconstructions from records with geographically incomplete sampling.
+Using informative prior information allows for accurate temperature reconstructions from records with geographically sparse sampling.
 By providing temperature estimates across the entire latitudinal range, this method also facilitates the reconstruction of unbiased global average temperatures.
@@
 Our GMST estimate is in good agreement with most existing estimates, indicating that broadly accurate GMST reconstructions are possible even with substantial deviations in the shape of the latitudinal temperature gradient.
-Our new method opens the door for improving the accuracy of proxy-based palaeoclimate reconstructions and Phanerozoic temperature curves, particularly in intervals with a patchy and unenvenly sampled record.
-Finally, the flexibility of our approach means that estimates can be efficiently updated when new data are made available.
+Our new method opens the door for improving the accuracy of proxy-based palaeoclimatic reconstructions and Phanerozoic temperature curves, particularly in intervals with a patchy and unevenly sampled record.
+Finally, the flexibility of our approach means that estimates can be efficiently updated when new data, or constraints, are made available.
 
@@
 
-The authors are grateful to all those who have enabled this work by collecting, measuring, collating and screening geochemical and fossil data.
+The authors are grateful to all those who have enabled this work by collecting, measuring, collating, and screening geochemical and fossil data.
 The contribution of L.A.J.

### 2023-09-23 manuscript/supplementary.qmd
update supplementary materials and s figures

@@
 Estimated latitudinal temperature gradient using only the geochemical proxy data (yellow), showing the median (line) and 95% credible interval (shading).
-Symbols with vertical lines show the median and 95% credible intervals of $\mu_j$.
+Symbols with vertical lines show the median and 95% credible intervals of the location means $\mu_j$.
 The blue line and shading in the background show the latitudinal temperature gradient with the geochemical and ecological proxy data, as in Figure 4.
@@
 Estimated latitudinal temperature gradient using only wider ecological limits for the coral data and for the *Avicennia*-Rhizophoraceae mangrove assemblage (orange), showing the median (line) and 95% credible interval (shading).
-The ecological limits were expanded to include $95\%$ of the probability density in the interval of $16.0$ to $35.6$&deg;C, using the minimal monthly temperature experienced by coral reefs (Kleypas et al., 1999), and a theoretical maximum derived by Jones et al.
+The ecological limits were expanded to include $95\%$ of the probability density in the interval of $16.0$ to $35.6$&deg;C, using the minimal monthly temperature experienced by coral reefs (Kleypas et al., 1999), and a modelled maximum derived by Jones et al.
 (2022).
@@
 Estimated latitudinal temperature gradients for each hemisphere, showing the medians (lines) and 95% credible intervals (shadings) in the Southern (red) and Northern Hemisphere (blue).
-Symbols with vertical lines show the median and 95% credible intervals of $\mu_j$.
+Symbols with vertical lines show the median and 95% credible intervals of the location means $\mu_j$.
 Turquoise symbols in the Northern Hemisphere highlight the ecological proxy data.
 The grey line and shading in the background show the latitudinal temperature gradient with the data from both hemispheres combined, plotted in both hemispheres.
-The dotted lines show the empirical, modern gradient, averaged across bins of 1&deg, for both hemispheres combined (black), and for the Northern hemisphere (blue) and Southern hemisphere (red), separately.
+The dotted lines show the empirical, modern gradient, averaged across bins of 1&deg;, for both hemispheres combined (black), and for the Northern hemisphere (blue) and Southern hemisphere (red), separately.
 :::
@@
 The resulting gradient estimated with this expanded model is very similar to that estimated with the original model (Fig.
-S2).
+S5).
 The uncertainty around the proxy temperatures allows the estimated location mean temperatures to be drawn closer to the gradient line, resulting in a slightly lower median residual standard deviation (`r eeco_sd_sdy$"50%"` as opposed to `r eeco_sdy$"50%"` in the original model).
@@
 
-Estimated latitudinal temperature gradient using only the geochemical proxy data (orange), showing the median (line) and 95% credible interval (shading).
-Symbols with vertical lines show the median and 95% credible intervals of $\mu_j$.
-The blue line and shading in the background show the latitudinal temperature gradient with the geochemical and ecological proxy data, as in Figure 4.
+Estimated latitudinal temperature gradient with uncertainty on proxy observations (blue), compared to the original model result (black, as in Fig.
+4).
+Gradients are depicted as medians (lines) and 95% credible intervals (shadings).
+Symbols with vertical lines show the median and 95% credible intervals of the location means $\mu_j$ for the geochemical data.
 :::
@@
 (2019) with more than one sample, against the absolute latitude of the locality (n = 17).
-The line shows a linear regression, suggesting a moderate effect of absolute latitude on proxy data variability, with mid- and hight-latitude data being more variable than low-latitude data.
+The line shows a linear regression, suggesting a moderate effect of absolute latitude on proxy data variability, with mid- and high-latitude data being more variable than low-latitude data.
 :::

### 2023-09-23 R/figures/SM/FigS-include-proxy-SD.R

@@ -4,6 +4,8 @@ source("./R/options.R")
 # load figures for plotting, distributions ect.
 source("./R/functions/auxiliary_functions.R")
 source("./R/functions/model_components/gradient.R")
+source("./R/functions/model_processing/temp_from_gradient.R")
+
 # read model output
 mode <- readRDS("./results/eeco/eeco_climate_model_output.rds")
 mode_all <- readRDS("./results/eeco/eeco_climate_model_output_combined.rds")
@@ -33,8 +35,8 @@ proxind1 <- sapply(obsmat$proxy, function(x) which(x == c("d18O","MgCa","TEX86",
 alpha3 <- 0.9
 col_proxgrad <- rgb(0.9,0.3,0,0.9)
 col_proxgrad_shade <- rgb(0.9,0.3,0,0.27)
-xlab1 <- expression("|latitude| ("*degree*")")
-ylab1 <- expression("temperature ("*degree*"C)")
+xlab1 <- expression("absolute latitude ("*degree*")")
+ylab1 <- expression("sea surface temperature ("*degree*"C)")
 lat_obs <- mode[[1]]$lat
 alpha2 <- 0.75
 col1 <- c(rgb(.9,.8,.5,alpha2), rgb(.8,.7,.5,alpha2), rgb(.7,.6,.5,alpha2), rgb(.65,.6,.6,alpha2))

### 2023-09-23 R/figures/SM/FigS2.R

@@ -3,6 +3,8 @@
 # load figures for plotting, distributions ect.
 source("./R/functions/auxiliary_functions.R")
 source("./R/functions/model_components/gradient.R")
+source("./R/functions/model_processing/temp_from_gradient.R")
+
 # read model output
 mode <- readRDS("./results/eeco/eeco_climate_model_output.rds")
 mode_all <- readRDS("./results/eeco/eeco_climate_model_output_combined.rds")
@@ -35,8 +37,8 @@ proxind1 <- sapply(obsmat$proxy, function(x) which(x == c("d18O","MgCa","TEX86",
 alpha3 <- 0.9
 col_proxgrad <- rgb(0.85,0.6,0,0.9)
 col_proxgrad_shade <- rgb(0.85,0.6,0,0.27)
-xlab1 <- expression("|latitude| ("*degree*")")
-ylab1 <- expression("temperature ("*degree*"C)")
+xlab1 <- expression("absolute latitude ("*degree*")")
+ylab1 <- expression("sea surface temperature ("*degree*"C)")
 lat_obs <- mode[[1]]$lat
 alpha2 <- 0.75
 col1 <- c(rgb(.9,.8,.5,alpha2), rgb(.8,.7,.5,alpha2), rgb(.7,.6,.5,alpha2), rgb(.65,.6,.6,alpha2))

### 2023-09-23 R/figures/SM/FigS3.R

@@ -3,6 +3,8 @@
 # load figures for plotting, distributions ect.
 source("./R/functions/auxiliary_functions.R")
 source("./R/functions/model_components/gradient.R")
+source("./R/functions/model_processing/temp_from_gradient.R")
+
 # read model output
 mode <- readRDS("./results/eeco/eeco_climate_model_output.rds")
 mode_all <- readRDS("./results/eeco/eeco_climate_model_output_combined.rds")
@@ -110,7 +112,7 @@ png("figures/SM/FigS3_north_south_eeco.png", width = 5.5, height = 3.5, units = 
 
 par(las = 1, mar = c(4.2,4.2,.5,.5), mgp = c(2.5,0.8,0))
 plot(0,0,type = "n", xlim = c(-94,94), ylim = c(-3,40), yaxs = "i", xaxs = "i",
-     xlab = expression("latitude ("*degree*")"), ylab = expression("temperature ("*degree*"C)"), xaxt = "n")
+     xlab = expression("latitude ("*degree*")"), ylab = expression("sea surface temperature ("*degree*"C)"), xaxt = "n")
 
 error_polygon(lat, all_temp$l_ci_95, all_temp$u_ci_95, col = rgb(0,0,0,0.2))
 error_polygon(-lat, all_temp$l_ci_95, all_temp$u_ci_95, col = rgb(0,0,0,0.2))

### 2023-09-23 R/figures/SM/figS-wider-ecol-limits.R

@@ -4,6 +4,8 @@ source("./R/options.R")
 # load figures for plotting, distributions ect.
 source("./R/functions/auxiliary_functions.R")
 source("./R/functions/model_components/gradient.R")
+source("./R/functions/model_processing/temp_from_gradient.R")
+
 # read model output
 mode <- readRDS("./results/eeco/eeco_climate_model_output.rds")
 mode_all <- readRDS("./results/eeco/eeco_climate_model_output_combined.rds")
@@ -23,6 +25,19 @@ wider_avi_rhi_mean <- wider_coral_mean
 wider_avi_rhi_sd <- wider_coral_sd
 proxy_distributions$mean[2] <- wider_avi_rhi_mean
 proxy_distributions$sd[2] <- wider_avi_rhi_sd
+
+
+### Mangrove and Coral data
+# read and assign proxy type
+bioprox <- readRDS("data/processed/bio_proxies_2022_08_08.RDS")
+bioprox$proxy <- rep("Reef",nrow(bioprox))
+bioprox$proxy[which(bioprox$taxa=="Avicennia")] <- "Avicennia"
+bioprox$proxy[which(bioprox$type=="mangrove" & bioprox$taxa!="Avicennia")] <- 
+  "Avicennia-Rhizophoraceae"
+
+# create distribution matrix for use with model (distributions defined in options)
+proxy_index <- sapply(bioprox$proxy, function(f) which(proxy_distributions$name==f))
+
 
 distrmat2 = data.frame(p_lat = abs(bioprox$p_lat), 
                       mu = proxy_distributions$mean[proxy_index],
@@ -56,8 +71,8 @@ proxind1 <- sapply(obsmat$proxy, function(x) which(x == c("d18O","MgCa","TEX86",
 alpha3 <- 0.9
 col_proxgrad <- rgb(0.9,0.3,0,0.9)
 col_proxgrad_shade <- rgb(0.9,0.3,0,0.27)
-xlab1 <- expression("|latitude| ("*degree*")")
-ylab1 <- expression("temperature ("*degree*"C)")
+xlab1 <- expression("absolute latitude ("*degree*")")
+ylab1 <- expression("sea surface temperature ("*degree*"C)")
 lat_obs <- mode[[1]]$lat
 alpha2 <- 0.75
 col1 <- c(rgb(.9,.8,.5,alpha2), rgb(.8,.7,.5,alpha2), rgb(.7,.6,.5,alpha2), rgb(.65,.6,.6,alpha2))

### 2023-09-24 manuscript/main.qmd
final proof reading & corrections

@@
 We chose a generalised logistic function to accurately infer the shape of the temperature gradient despite a patchy latitudinal coverage.
-This choice is motivated by the flexibility and ability of this function to approximate a variety of nonlinear patterns in the underlying temperature gradients, that other parametric approaches, such as lower order polynomials [e.g.
+This choice is motivated by the flexibility and ability of this function to approximate a variety of nonlinear patterns in the underlying temperature gradients that other parametric approaches, such as lower order polynomials [e.g.
 @Bijl2009; @Keating-Bitonti2011], lack.
@@
 For our analyses, this dataset was restricted to the EECO (defined as 53.8--49.1 Ma) and samples originating from near the ocean surface or mixed layer.
-Consequently, samples labelled as "thermocline" or "sub-thermocline" were excluded.
+Consequently, samples labelled as "thermocline", or "sub-thermocline", were excluded.
 Recrystallised &delta;^18^O samples were also excluded as secondary diagenetic calcite precipitated after deposition can bias isotope measurements and offset temperature values [@Schrag1999].
@@
 
-1)  *Avicennia*-only assemblages ($n=2$): the absence of Rhizophoraceae is indicative of temperatures between 15.6&deg;C (lower temperature limt of *Avicennia*) and 20.7&deg;C (lower temperature limit of *Rhizophora*).
+1)  *Avicennia*-only assemblages ($n=2$): the absence of Rhizophoraceae is indicative of temperatures between 15.6&deg;C (lower temperature limit of *Avicennia*) and 20.7&deg;C (lower temperature limit of *Rhizophora*).
 However, a value of 22.5&deg;C is assumed as the upper temperature limit here as *Rhizophora* is rare below this temperature.
@@
 This is achieved by setting $K \ge A$.
-The prior on the upper asymptote $K$ is a truncated normal distribution with the mean set to $K$ of the modern SST gradient, with a broad standard deviation: $$K \sim TN(\mu_{TN} = 28, \sigma_{TN} = 10, \alpha_{TN} = A, \beta_{TN}= \infty)$${#eq-K}
+The prior on the upper asymptote $K$ is a truncated normal distribution with the mean set to $K$ of the modern SST gradient, with a broad standard deviation: $$K \sim TN(\mu_{TN} = 28, \sigma_{TN} = 15, \alpha_{TN} = A, \beta_{TN}= \infty)$${#eq-K}
 The distribution is truncated to the left at $\alpha_{TN}=A$, but not truncated to the right ($\beta_{TN}$).
 
-**M.** The steepness of the gradient is presumed to be highest in mid-latitudes; this is expressed with a normal prior on $M$ with the mean set to 42, i.e.
-$M$ of the modern SST gradient, and a moderately wide standard deviation of 10: $$M \sim N(42,10)$${#eq-M}
+**M.** A uniform prior is placed on the latitude of greatest steepness of the gradient, allowing it to be steepest anywhere between latitudes 0&deg; and 90&deg; absolute latitude, as this parameter may vary greatly depending on the climate state: $$M \sim Uniform(0,90)$${#eq-M}
 
@@
 The shape and rate parameters $\alpha_G$ and $\beta_G$ were chosen such that the highest prior density is at $B$ of the modern SST gradient, $0.11$.
-We informed the prior distributions on $M$ and $B$ based on a provisional model run with the modern SST data.
+We informed the prior distribution on $B$ based on a provisional model run with the modern SST data.
 
@@
 For the largest sample size (n = 34), we used the latitudes of the EECO data set of this study in all iterations.
-For each latitude, we inferred a location mean temperature from the gradients, adding random noise from a normal distribution with a standard deviation of 3.8, which is the average standard deviation of the EECO geochemical proxy data.
+For each latitude, we took the location mean temperature from the gradients, adding random noise from a normal distribution with a standard deviation of 3.8, which corresponds to the average uncertainty associated with the EECO geochemical proxy data [@Hollis2019] .
 With that, we aim to simulate randomly distributed errors in the proxy data, which could arise from miscalibrations, measurement errors, seasonal effects, *ect*.
-We acknowledge that this approach cannot quantify the potential impact of systematic offsets that may bias all proxy data in the same direction, nor do we know whether a standard deviation of 3.8 is actual, average magnitude of uncertainty that the proxy compilation is afflicted with.
+We acknowledge that this approach cannot quantify the potential impact of systematic offsets that may bias all proxy data in the same direction, nor do we know whether a standard deviation of 3.8 is the actual average magnitude of uncertainty that the proxy compilation is afflicted with.
 
@@
 Every $10^{th}$ iteration was retained, resulting in a total of 20,000 iterations with low autocorrelation.
-The re-sampled, simulated gradients were modelled in one chain with 10,000 iterations for each of the 1,000 random samples.
+The re-sampled, simulated gradients and the re-sampled, modern gradient were modelled in one chain  with 10,000 iterations for each of the 1,000 random samples.
 5,000 iterations each were discarded as burn-in, and every $25^{th}$ iteration was kept, resulting in a total of 200,000 iterations across all 1,000 model runs.
-For the simulated gradients with an Eocene sampling distribution, a single chain with 250,000 iterations was used, thinned to 10,000 iterations after burn-in.
 For the Eocene model, we ran four chains with 600,000 iterations each, discarding 100,000 as burn-in and keeping every $100^{th}$ iteration, as the hierarchical model structure results in higher autocorrelation of the chains.
@@
 ```
-Modelled sea surface temperature estimates were generated with [@eq-nu], calculating the sea surface temperatures at any latitude with the parameter estimates of each iteration from the posterior.
+Modelled sea surface temperature estimates were generated with [@eq-nu], calculating the sea surface temperatures at any latitude with the parameter estimates of each iteration from the posterior samples.
 The median and 95% CI of temperatures were then taken from all temperature estimates obtained at the latitudes of interest.
@@
 Model reconstructions of simulated latitudinal temperature gradients at various sample sizes.
-Each column depicts a different reconstruction for given sample sizes: 5, 10, 20, and 34 (latitudes of EECO samples).
-Each row depicts a different latitudinal temperature gradient that represents idealised or observed climatic states: idealsed extreme icehouse, greenhouse, and extreme greenhouse gradients, and the modern gradient, which represents an icehouse state.
+Each column depicts a different reconstruction for given sample sizes: 5, 10, 20 (randomly sampled latitudes), and 34 (latitudes of EECO samples).
+Each row depicts a different latitudinal temperature gradient that represents idealised or observed climatic states: idealised extreme icehouse, greenhouse, and extreme greenhouse gradients, and the modern gradient, which represents an icehouse state.
 The black line illustrates the original gradient.
-The blue line depicts the reconstructed gradient represented by the median sea surface temperature value estimated from 1,000 model runs with different random samples (first three columns), and a single run with the EECO latitudinal sampling distribution (fourth column).
-To generate the random samples, noise from a normal distribution with a standard deviation of 3.8 was added to all temperatures.
+The blue line depicts the reconstructed gradient represented by the median sea surface temperature value estimated from 1,000 model runs with different random samples.
+To generate the random samples, different random noise from a normal distribution with a standard deviation of 3.8&degC; was added to each temperature.
 The blue shadings depict the 90%, 95%, and 99% credible intervals.
@@
 Grey points depict the individual cell values of the Bio-ORACLE grid of mean sea surface temperatures, which were used to infer the empirical and the modelled gradient.
-The darkness of the background shading indicates the point density.
+Higher opacity of points indicates higher density of data (multiple overlapping points).
 :::
@@
 Greenhouse scenarios perform somewhat worse due to the increased uncertainty at high latitudes (median of $R^2$ > 0.7 at sample sizes 10 and 20).
-A sampling distribution resembling that of the early Eocene dataset used in this study allows for accurate reconstruction of all scenarios, although the $R^2$ is still relatively low in the extreme greenhouse scenario, as a perfectly flat gradient prediction by the model would result in an $R^2$ of 0, despite the original gradient being very flat.
+A sampling distribution resembling that of the early Eocene dataset used in this study allows for accurate reconstruction of all scenarios, although the $R^2$ is still relatively low in the extreme greenhouse scenario, as a perfectly flat gradient, predicted by the model, would result in an $R^2$ of 0, despite the original gradient being very flat.
 This also explains the low lower bounds of the 95% credible intervals in the greenhouse scenarios.
@@
 
-Estimates of the median, latitudinal sea surface temperature gradients of the early Eocene climatic optimum (purple line) and of the present-day (turquoise), both estimated with the Bayesian model.
+Estimates of the median, latitudinal sea surface temperature gradient of the early Eocene climatic optimum (purple line) and of the present-day (turquoise), both estimated with the Bayesian model.
 The purple ribbon (shading) depicts the 95% credible interval of the Eocene gradient, the uncertainty of the modern gradient is too low to be visible.
@@
 A local temperature estimate, based on multiple geochemical proxies, thus has the same weight as a local temperature estimate obtained from the occurrence of a climate-sensitive plant community, whilst preserving the uncertainty associated with each estimate.
-The model could easily be extended to include uncertainties on individual geochemical proxy data, or to variably weight proxy records classified as more or less reliable.
+The model could easily be extended to include uncertainties on individual geochemical proxy data (see Fig.
+S5), or to variably weight proxy records classified as more or less reliable.
 
@@
 Earlier, reconstructed early Eocene and EECO SST gradients range from 7--21&deg;C (Table 1); a more recent reconstruction that includes terrestrial air and sea surface temperatures arrives at a gradient of ~13&deg;C [@Inglis2020].
-Our median poles-to-equator gradient estimate is similar at `r eocene_gradient$"50%"`&deg;C, but notably shallower when taking the equator-to-polar-circle estimate, `r eocene_gradient_polar_circle$"50%"`&deg;C, as the geochemical proxy data suggest high temperatures up to latitudes of ~ 60&deg.
+Our median poles-to-equator gradient estimate is similar at `r eocene_gradient$"50%"`&deg;C, but notably shallower when taking the equator-to-polar-circle estimate, `r eocene_gradient_polar_circle$"50%"`&deg;C, as the geochemical proxy data suggest high temperatures up to latitudes of ~ 60&deg;.
 Both geochemical and ecological shallow water data indicate that inferred SST gradients based on tropical, shallow water and deep water samples [@Cramwinckel2018; @Evans2018] may overestimate the SST gradient of the early Eocene greenhouse world.
@@
 Temporal changes within the EECO [@Westerhold2018], and seasonality [@Keating-Bitonti2011; @Ivany2022] may also contribute to the large variability of the EECO proxy data.
-Based on the occurrence of heterotrophic carbonates, @Davies2019 suggest that mid- and high-latitude geochemical proxy data from the EECO may be biased towards summer temperatures.
+Based on the occurrence of heterotrophic carbonates, @Davies2019 suggested that mid- and high-latitude geochemical proxy data from the EECO may be biased towards summer temperatures.
 Some of the geochemical mid-latitude geochemical proxy data from @Hollis2019 may therefore suggest higher than actual mean annual temperatures, and the variability of temperature estimates from individual localities is higher in mid - high latitudes (Fig.

### 2023-09-24 R/analyses/SM/run_EECO_climate_model_with_proxy_sd.R

@@ -57,7 +57,7 @@ mode_sd <- foreach(pc = 1:n_chains) %dopar% { # run 1 chain per cluster
 parallel::stopCluster(cl)
 
 ### save output
-saveRDS(mode, "results/SM/eeco_climate_model__with_observation_sd_output.rds")
+saveRDS(mode_sd, "results/SM/eeco_climate_model__with_observation_sd_output.rds")
 
 # combine chains and discard burn-in
 source("R/functions/model_processing/combine_posterior.R")
